---
title: "Maximum likelihood"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to use maximum likelihood to estimate unknown
parameters in ecological probability models and describe uncertainty from the
likelihood. In [Chapter 3](../Chapter_3/Theory.qmd#sec-estimates), we
estimated the mean number of seedlings per quadrat by matching a sample mean
to the mean of a Poisson model. But the model tells us more than its mean: it
assigns a probability to every possible count. Can we use that information to
estimate the unknown mean?

We will answer this question using **maximum likelihood estimation** (MLE).
We will first calculate likelihoods for a few counts by hand. Then we will
use R to fit the same model to a larger sample. Tree heights will show what
changes when a model has two parameters, and negative binomial counts will
bring the steps together. All data in this chapter are illustrative or
simulated; they are not field observations.

# Learning goals

After studying this chapter, you should be able to:

- distinguish probability from likelihood and calculate a likelihood and
  log-likelihood for a small sample by hand;
- construct a log-likelihood from independent probability masses or densities;
- find a maximum likelihood estimate from a likelihood curve or with
  `optim()`, including a transformation for a constrained parameter;
- describe finite-sample behaviour, consistency, asymptotic normality,
  asymptotic efficiency, and invariance of maximum likelihood estimators;
- interpret a two-parameter likelihood surface, a slice, and a profile;
- calculate and interpret approximate quadratic confidence intervals
  using a Hessian and profile likelihood confidence intervals;
- compare these intervals and check whether a fitted ecological model
  describes the data reasonably.

# From a probability model to a likelihood {#sec-likelihood}

Likelihood uses a probability model for a different question from probability
itself. After observing data, we hold those observations fixed and compare how
well different candidate parameter values describe them. We now apply that
idea to seedling counts.

Suppose $X_i$ is the number of seedlings in equal-area quadrat $i$. We assume
the quadrat counts are independent and have a common Poisson mean $\lambda$,
measured in seedlings per quadrat:

$$
X_i \sim \operatorname{Poisson}(\lambda).
$$

The [Poisson model](../Supplements/distributions.qmd#sec-poisson)
assigns probability

$$
P(X_i=x_i\mid\lambda)
=\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}
$$

to one possible count $x_i$. If $\lambda$ is fixed, we can ask how probable
different counts are. Once the counts have been observed, we can hold
those counts fixed and compare candidate values of $\lambda$.

::: {.callout-note title="Likelihood"}
The **likelihood** $L(\lambda\mid x_1,\ldots,x_n)$ is the joint
probability of the observed discrete data, viewed as a function of
$\lambda$. For continuous data, it uses joint density instead.
The data are fixed while the parameter varies.
:::

For independent quadrats, multiply the contribution from each count:

$$
L(\lambda\mid x_1,\ldots,x_n)
=\prod_{i=1}^{n}P(X_i=x_i\mid\lambda).
$$

Independence is an assumption that permits this product. Separate
quadrats can still share conditions that make their counts dependent.

## Calculate a likelihood by hand

Consider the illustrative counts $x_1=1$, $x_2=2$, and $x_3=3$. For a
candidate mean of one seedling per quadrat,

$$
\begin{aligned}
L(1\mid 1,2,3)
&=P(X_1=1\mid1)P(X_2=2\mid1)P(X_3=3\mid1)\\
&=\frac{e^{-1}1^1}{1!}
  \frac{e^{-1}1^2}{2!}
  \frac{e^{-1}1^3}{3!}\\
&=\frac{e^{-3}}{12}\approx 0.00415.
\end{aligned}
$$

For a candidate mean of two seedlings per quadrat, the same *observed*
counts give

$$
L(2\mid1,2,3)
=\frac{e^{-6}2^{1+2+3}}{1!\,2!\,3!}
=\frac{64e^{-6}}{12}
\approx 0.01322.
$$

The counts are more likely under $\lambda=2$ than under $\lambda=1$.
This comparison does not give the probability that either value of
$\lambda$ is true.

::: {.callout-warning title="Likelihood is not probability for the parameter"}
Likelihood compares parameter values using fixed data. It need not
integrate to one over parameter values, and $L(2\mid1,2,3)$ is not the
probability that $\lambda=2$.
:::

## Add logs rather than multiply

Products of many small contributions can become too small for a computer
to represent. The **log-likelihood** turns the product into a sum:

$$
\ell(\lambda)
=\log L(\lambda)
=\sum_{i=1}^{n}\log P(X_i=x_i\mid\lambda).
$$

For our three counts, substitute into the Poisson formula:

$$
\begin{aligned}
\ell(\lambda\mid1,2,3)
&=\sum_{i=1}^{3}
  [-\lambda+x_i\log\lambda-\log(x_i!)]\\
&=-3\lambda+6\log\lambda-\log(12).
\end{aligned}
$$

Hence $\ell(1)=-3-\log(12)\approx-5.485$ and
$\ell(2)=-6+6\log(2)-\log(12)\approx-4.326$.
The larger log-likelihood again belongs to $\lambda=2$. Logarithms
preserve the ordering, so likelihood and log-likelihood have the same
maximum. A negative log-likelihood is ordinary: it is the logarithm
of a joint probability smaller than one.

In R, `log = TRUE` asks `dpois()` for each log probability. Sum these
values directly rather than taking the logarithm of a product:

```r
# Store the observed counts and calculate their log-likelihoods.
counts = c(1, 2, 3)
sum(dpois(x = counts, lambda = 1, log = TRUE))
sum(dpois(x = counts, lambda = 2, log = TRUE))
```

:::: {.callout-important title="Practice: a likelihood on paper"}

Three other quadrats contain 0, 1, and 2 seedlings. Without R, write
the three Poisson probability contributions, their product, and the
log-likelihood. Evaluate both at $\lambda=1$ and $\lambda=2$ with a
calculator. Which candidate has greater likelihood?

::: {.callout-tip title="Solution" collapse="true"}

For these counts,
$L(\lambda)=e^{-3\lambda}\lambda^{0+1+2}/(0!\,1!\,2!)
=e^{-3\lambda}\lambda^3/2$ and
$\ell(\lambda)=-3\lambda+3\log\lambda-\log2$.
At $\lambda=1$, the likelihood is $e^{-3}/2\approx0.02489$ and
the log-likelihood is approximately $-3.693$.
At $\lambda=2$, they are $4e^{-6}\approx0.009915$ and
$-6+2\log2\approx-4.614$. Among these candidates, $\lambda=1$ fits
the fixed counts better.

:::
::::

For a continuous model, the same multiplication and log-sum steps use
**density** values. As [Chapter 2](../Chapter_2/Theory.qmd#sec-continuous_measurements)
explains, a density at an exact height is not a probability of that
exact height.

# Find the maximum {#sec-maximum-likelihood}

The **maximum likelihood estimator** is a rule that chooses the
parameter value with the highest likelihood. Applying it to one sample
produces a **maximum likelihood estimate**. For the Poisson mean,

$$
\hat\lambda_{\mathrm{ML}}
=\operatorname*{arg\,max}_{\lambda\geq0}\ell(\lambda).
$$

For independent Poisson counts, the maximum occurs at the sample mean
$\bar x$. We also reached $\bar x$ by the
[method of moments in Chapter 3](../Chapter_3/Theory.qmd#sec-estimates).
The two methods happen to agree for this model; they answer the
estimation problem in different ways. If all counts are zero, the
maximum is at the boundary $\lambda=0$.

Consider a larger illustrative sample from 12 equal-area quadrats:

```r
# Store the illustrative counts and calculate their maximum likelihood estimate.
seedling_counts = c(1, 2, 3, 2, 4, 1, 3, 2, 2, 3, 1, 2)
mean(seedling_counts)
```

The mean is approximately 2.17 seedlings per quadrat. We can evaluate
the log-likelihood at many candidate values and plot it. The highest
point lies at the estimate.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Relative log-likelihood for illustrative seedling counts in 12 quadrats. The vertical line marks the maximum likelihood estimate."
# Store the counts and choose candidate Poisson means.
seedling_counts = c(1, 2, 3, 2, 4, 1, 3, 2, 2, 3, 1, 2)
lambda_grid = seq(from = 0.2, to = 5, length.out = 200)
# Calculate the log-likelihood at each candidate mean.
log_likelihood = vapply(
  X = lambda_grid,
  FUN = function(lambda) {
    sum(dpois(x = seedling_counts, lambda = lambda, log = TRUE))
  },
  FUN.VALUE = numeric(1)
)
# Plot relative log-likelihood and mark the fitted mean.
plot(
  x = lambda_grid, y = log_likelihood - max(log_likelihood),
  type = "l", lwd = 2, xlab = "Mean seedlings per quadrat",
  ylab = "Relative log-likelihood"
)
abline(v = mean(seedling_counts), lty = 2, lwd = 2)
```

The vertical distance from the peak shows how much worse another
candidate fits these observations on the log-likelihood scale. It is
not a probability assigned to the parameter. The relative
log-likelihood has a maximum of zero because we subtract the highest
log-likelihood.

## Let `optim()` find the maximum

R's `optim()` minimises a function by default. We therefore give it
the **negative log-likelihood**. A Poisson mean cannot be negative.
We optimise its logarithm and transform back with `exp()`:

$$
\eta=\log\lambda,\qquad \lambda=e^\eta.
$$

This transformation keeps every value considered by the optimiser
positive. It cannot represent the boundary value $\lambda=0$, so a
sample containing only zeros needs separate treatment.

```r
# Define the negative log-likelihood on a scale that keeps the mean positive.
poisson_nll = function(log_lambda, counts) {
  lambda = exp(log_lambda)
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}

# Fit the model from a valid starting value and return the mean on its original scale.
fit = optim(
  par = log(2), fn = poisson_nll, counts = seedling_counts,
  method = "BFGS"
)
fit$convergence
lambda_hat = exp(fit$par)
c(maximum_likelihood = lambda_hat, sample_mean = mean(seedling_counts))
```

`log(2)` is a starting value, not an estimate. A convergence code of
zero reports that the algorithm terminated successfully. We still
compare the answer with the sample mean and inspect the fitted model.
The estimate is reported on the original scale: seedlings per quadrat.

::: {.callout-note title="A practical fitting sequence"}
State the model and its assumptions, write the contribution from one
observation, sum log contributions, choose a valid parameter scale and
starting value, optimise, and check the returned fit. An optimiser
returning a number does not establish that the ecological model is
appropriate.
:::

:::: {.callout-important title="Practice: read an optimiser result"}

If `optim()` returns `par = log(3.1)` when fitting a Poisson model on
the log scale, what mean count would you report? Why is the returned
`par` value itself not in seedlings per quadrat?

::: {.callout-tip title="Solution" collapse="true"}

The fitted mean is $\exp[\log(3.1)]=3.1$ seedlings per quadrat.
The optimiser worked with $\eta=\log\lambda$, which is a transformed
parameter.

:::
::::

# What an estimate can and cannot guarantee

An MLE changes between repeated samples. The
[sampling-distribution discussion in Chapter 3](../Chapter_3/Theory.qmd)
explains bias, variance, standard error, and confidence interval
coverage. Maximum likelihood does not guarantee zero bias in a
small sample. For example, the normal-model variance MLE divides
the sum of squared deviations by $n$; the related `var()` calculation
divides by $n-1$. Chapter 3 explains the resulting finite-sample bias.
Under the independent Poisson model, $\hat\lambda=\bar X$ is
unbiased and has variance $\lambda/n$. Its mean squared error
therefore also equals $\lambda/n$. This illustrates how an
estimator's bias and spread, rather than the method's name alone,
describe its finite-sample performance.

**Invariance** means that if $\hat\theta$ is an MLE of $\theta$, then
$g(\hat\theta)$ is the MLE of $g(\theta)$ for a parameter quantity
defined by $g$. Thus an estimate of $\log\lambda$ is
$\log\hat\lambda$. We use the same idea when converting an estimate
from the optimiser's log scale back to an ecological scale.

Under suitable model and sampling conditions, MLEs are **consistent**: as
sample size grows, they tend to concentrate near the parameter that generated
the data. Their sampling distributions are often approximately normal in large
samples (**asymptotic normality**). Under additional technical conditions, an
MLE can be **asymptotically efficient**, meaning its large-sample variance
reaches the smallest available value among a suitable class of estimators.
These are long-run properties, not promises that every larger realised sample
gives a closer estimate.

::: {.callout-warning title="A maximum is conditional on the model"}
If quadrat counts have different expected means or are dependent,
a Poisson fit based on a common independent mean may describe them
poorly. A high likelihood within that model cannot repair its
assumptions. Small samples, boundaries, and weakly identified
parameters can also make normal approximations unreliable.
:::

# When there are two parameters {#sec-likelihood-surface}

Many ecological probability models have more than one unknown parameter. The
likelihood then compares combinations of parameter values while holding the
observations fixed. To apply this idea, suppose we measure tree heights $H_i$
in metres. As in [Chapter 2](../Chapter_2/Theory.qmd#sec-continuous_four_faces),
a normal model has mean $\mu$ metres and standard deviation $\sigma$ metres:

$$
H_i\sim\operatorname{Normal}(\mu,\sigma^2).
$$

For independent heights, multiply their densities. The
log-likelihood is easier to calculate:

$$
\ell(\mu,\sigma)
=\sum_{i=1}^{n}
\left[
-\log(\sigma\sqrt{2\pi})
-\frac{(h_i-\mu)^2}{2\sigma^2}
\right].
$$

Here the normal model is parameterised with variance $\sigma^2$,
while R's `dnorm()` takes the standard deviation in its `sd` argument.

## Calculate a density likelihood by hand

Take two illustrative heights, 20 m and 24 m. Candidate A has
$(\mu,\sigma)=(22,2)$ m; candidate B has $(22,4)$ m. Their
log-likelihoods are

$$
\begin{aligned}
\ell_A
&=-2\log(2\sqrt{2\pi})
  -\frac{(20-22)^2+(24-22)^2}{2(2)^2}
 \approx -4.224,\\
\ell_B
&=-2\log(4\sqrt{2\pi})
  -\frac{(20-22)^2+(24-22)^2}{2(4)^2}
 \approx -4.860.
\end{aligned}
$$

The two density contributions for candidate A are
$f(20\mid22,2)$ and $f(24\mid22,2)$; their product is
$L_A=e^{\ell_A}$. Candidate A has the higher likelihood for
these two heights. A wider model does not automatically fit every
short sample better.

:::: {.callout-important title="Practice: two candidate tree-height models"}

Without R, calculate the log-likelihood of heights 20 m and 24 m
under $(\mu,\sigma)=(20,4)$ m. Compare it with candidate B above.
Which has greater likelihood? Write the two individual density
contributions before combining their logs.

::: {.callout-tip title="Solution" collapse="true"}

The contributions are
$f(20\mid20,4)=1/(4\sqrt{2\pi})$ and
$f(24\mid20,4)=e^{-1/2}/(4\sqrt{2\pi})$.
Their log-likelihood is
$-2\log(4\sqrt{2\pi})-1/2\approx-5.110$.
Candidate B, with log-likelihood approximately $-4.860$, fits
these fixed heights better.

:::
::::

## Surface, slice, and profile

With two parameters, the likelihood has a value for every
$(\mu,\sigma)$ pair. A contour plot joins pairs with the same
log-likelihood. To illustrate this general idea, simulate 25 independent
heights from a known normal model. We know the generating parameters only
because this is a simulation.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Relative log-likelihood for 25 simulated tree heights. The filled point is the fitted mean and standard deviation; the cross marks the generating values."
# Simulate tree heights and calculate their normal-model estimates.
set.seed(seed = 505)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)
height_mu_hat = mean(tree_heights)
height_sigma_hat = sqrt(mean((tree_heights - height_mu_hat)^2))
# Evaluate the log-likelihood across a grid of mean and standard-deviation values.
mu_grid = seq(from = 17, to = 27, length.out = 80)
sigma_grid = seq(from = 1.5, to = 7, length.out = 80)
surface = outer(
  X = mu_grid, Y = sigma_grid,
  FUN = Vectorize(function(mu, sigma) {
    sum(dnorm(x = tree_heights, mean = mu, sd = sigma, log = TRUE))
  })
)
# Draw likelihood contours and mark the fitted and generating values.
contour(
  x = mu_grid, y = sigma_grid, z = surface - max(surface),
  xlab = "Mean height (m)", ylab = "Height SD (m)",
  levels = c(-10, -5, -2, -1, -0.5, -0.1)
)
points(x = height_mu_hat, y = height_sigma_hat, pch = 19)
points(x = 22, y = 4, pch = 4, lwd = 2)
legend(
  "topright", legend = c("MLE", "Generating values"),
  pch = c(19, 4), bty = "n"
)
```

The peak gives the joint MLE. The generating values need not sit at
the peak of one finite sample. Contours show *relative fit* across
parameter pairs, not probability regions for the parameters.

A **likelihood slice** varies one parameter while holding the other
fixed. For example, $\ell(\mu,\hat\sigma)$ holds the standard
deviation at its joint estimate. A **profile likelihood** for
$\mu$ instead finds the best $\sigma$ separately for every
candidate $\mu$:

$$
\ell_p(\mu)=\max_{\sigma>0}\ell(\mu,\sigma).
$$

Thus a profile re-estimates the nuisance parameter $\sigma$.
It is at least as high as a slice at every $\mu$, since the slice
offers just one possible value of $\sigma$. A slice should not be
labelled a profile.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Likelihood slice and profile for mean tree height in the same simulated sample. The profile re-estimates the standard deviation for every candidate mean."
# Simulate the tree-height sample and calculate its joint normal-model estimates.
set.seed(seed = 505)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)
mu_hat = mean(tree_heights)
sigma_hat = sqrt(mean((tree_heights - mu_hat)^2))
candidate_means = seq(from = 17, to = 27, length.out = 120)
# Define the normal-model log-likelihood for the fixed simulated heights.
normal_loglik = function(mu, sigma) {
  sum(dnorm(x = tree_heights, mean = mu, sd = sigma, log = TRUE))
}
maximum_loglik = normal_loglik(mu = mu_hat, sigma = sigma_hat)
# Calculate a slice with fixed standard deviation and a profile that refits it.
slice = vapply(
  X = candidate_means,
  FUN = function(mu) normal_loglik(mu = mu, sigma = sigma_hat),
  FUN.VALUE = numeric(1)
)
profile = vapply(
  X = candidate_means,
  FUN = function(mu) {
    sigma_at_mu = sqrt(mean((tree_heights - mu)^2))
    normal_loglik(mu = mu, sigma = sigma_at_mu)
  },
  FUN.VALUE = numeric(1)
)
# Plot both relative log-likelihood curves.
plot(
  x = candidate_means, y = slice - maximum_loglik,
  type = "l", lwd = 2, lty = 2,
  xlab = "Candidate mean height (m)",
  ylab = "Relative log-likelihood"
)
lines(
  x = candidate_means, y = profile - maximum_loglik,
  lwd = 2
)
legend(
  "bottomright", legend = c("Slice", "Profile"),
  lty = c(2, 1), lwd = 2, bty = "n"
)
```

Both curves meet at the joint estimate. Away from it, re-estimating
$\sigma$ can keep the profile higher than the slice.

::: {.callout-note title="Explore a likelihood surface"}
On [Kristoffer Magnusson's interactive likelihood page](https://rpsychologist.com/likelihood/),
change the mean and variance sliders in **Likelihood Calculation**.
Watch how the contributions and their sum change. In **Finding the
Maximum Likelihood Estimates**, locate the highest region of the
two-parameter contour plot and try the optimiser controls. The page
uses variance $\sigma^2$ on a slider; our contour above uses standard
deviation $\sigma$. The later hypothesis-test displays are outside
this chapter's learning goals.
:::

:::: {.callout-important title="Practice: slice or profile?"}

For each proposed mean tree height, you refit the standard deviation
and record the highest log-likelihood. Is this a slice or a profile?
What would you do differently to make the other kind of curve?

::: {.callout-tip title="Solution" collapse="true"}

It is a profile because the standard deviation is re-estimated for
each proposed mean. For a slice, hold the standard deviation at
one chosen value while varying the mean.

:::
::::

# Uncertainty from the likelihood {#sec-likelihood-uncertainty}

A peak shows an estimate. Its width helps describe uncertainty:
a sharply curved log-likelihood changes quickly when we move away
from the estimate; a flatter one changes slowly. Chapter 3 introduced
confidence intervals and their repeated-sampling interpretation.
Here we use two approximations based on the likelihood.

## A quadratic interval from the Hessian

Near a regular maximum, the log-likelihood can often be approximated
by a downward-opening quadratic curve. The **Hessian** describes its
curvature. For several parameters, it is a matrix of second
derivatives of the *negative* log-likelihood. Under suitable
conditions, the inverse Hessian approximates the covariance matrix
of the estimates. The square roots of its diagonal entries are
approximate standard errors, on the scale that was optimised.

We can fit the tree-height model on the $(\mu,\log\sigma)$ scale.
`optim()` returns its Hessian when `hessian = TRUE`.

```r
# Simulate the heights used to illustrate likelihood-based intervals.
set.seed(seed = 505)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)

# Define and fit the negative log-likelihood on the mean and log-SD scales.
normal_nll = function(par, heights) {
  mu = par[1]
  sigma = exp(par[2])
  -sum(dnorm(x = heights, mean = mu, sd = sigma, log = TRUE))
}

normal_fit = optim(
  par = c(mu = mean(tree_heights), log_sigma = log(sd(tree_heights))),
  fn = normal_nll, heights = tree_heights,
  method = "BFGS", hessian = TRUE
)
normal_fit$convergence
# Use the inverse Hessian to obtain approximate standard errors and interval endpoints.
se_on_fit_scale = sqrt(diag(solve(normal_fit$hessian)))
critical_value = qnorm(p = 0.975)

mu_interval = normal_fit$par[1] +
  c(-1, 1) * critical_value * se_on_fit_scale[1]
sigma_interval = exp(
  normal_fit$par[2] +
    c(-1, 1) * critical_value * se_on_fit_scale[2]
)
mu_interval
sigma_interval
```

These are approximate 95% **quadratic**, or Wald, confidence
intervals. The mean interval is in metres. The second interval is
calculated for $\log\sigma$ and then transformed back to metres,
so it cannot include a negative standard deviation. Its endpoints
need not be symmetric around the estimate on the metre scale.
The Hessian should be invertible and indicate a local minimum of
the negative log-likelihood; otherwise this calculation may fail
or be misleading.

::: {.callout-warning title="Curvature is a local approximation"}
A Hessian describes the likelihood close to the fitted point. A
skewed, flat, or boundary-shaped likelihood can make quadratic
intervals inaccurate. Transforming the endpoints keeps them in
the parameter's support but does not guarantee good coverage.
:::

## A profile likelihood interval

For a candidate $\mu$, refit $\sigma$ and compare the resulting
profile log-likelihood with the overall maximum. The
**likelihood-ratio distance** is

$$
D(\mu)=2[\ell(\hat\mu,\hat\sigma)-\ell_p(\mu)].
$$

Under regular, large-sample conditions, $D(\mu)$ is approximately
distributed as $\chi^2_1$ when evaluated at the generating value.
An approximate 95% profile interval includes candidate means for
which $D(\mu)\leq\chi^2_{1,0.95}$. On a relative profile
log-likelihood plot, this is the region above
$-\chi^2_{1,0.95}/2$. The interval can be asymmetric because it
follows the profile's shape.

For a normal sample, the best $\sigma$ at a fixed $\mu$ is the root
mean squared deviation from that candidate mean. The following
function uses this result to evaluate the profile. `uniroot()`
finds the two cutoff crossings.

```r
# Calculate the profile log-likelihood after refitting the standard deviation.
profile_mu = function(mu, heights) {
  sigma_at_mu = sqrt(mean((heights - mu)^2))
  sum(dnorm(
    x = heights, mean = mu, sd = sigma_at_mu, log = TRUE
  ))
}

# Define the profile cutoff and its difference from the maximum log-likelihood.
mu_hat = normal_fit$par[1]
max_log_likelihood = -normal_fit$value
cutoff = qchisq(p = 0.95, df = 1) / 2
crossing = function(mu) {
  profile_mu(mu = mu, heights = tree_heights) -
    max_log_likelihood + cutoff
}

# Locate the two candidate means where the profile reaches its cutoff.
profile_interval = c(
  lower = uniroot(
    f = crossing,
    interval = c(mu_hat - 10 * sd(tree_heights), mu_hat)
  )$root,
  upper = uniroot(
    f = crossing,
    interval = c(mu_hat, mu_hat + 10 * sd(tree_heights))
  )$root
)
profile_interval
```

The code brackets each crossing before calling `uniroot()`.
For another dataset, check that each proposed bracket actually
straddles the cutoff. A general model would re-optimise its
nuisance parameters for each candidate value instead of using
the normal-model shortcut above.

| Interval | What it uses | Main limitation |
|---|---|---|
| Quadratic | Curvature near the maximum | May miss asymmetry or a boundary |
| Profile | Best fit at each candidate parameter value | Requires repeated fitting and an approximate cutoff |

Neither interval is automatically exact. Both rely on the model,
sampling assumptions, and large-sample approximations. As in
Chapter 3, a calculated 95% confidence interval does not give a
95% probability that a fixed parameter lies inside it.

:::: {.callout-important title="Practice: read a profile interval"}

A relative profile log-likelihood reaches zero at the estimate.
At a candidate mean, it equals $-2.5$. The approximate 95% cutoff
is $-qchisq(0.95,1)/2\approx-1.92$. Does this candidate belong
to the profile interval? What is being refitted as $\mu$ varies?

::: {.callout-tip title="Solution" collapse="true"}

No. $-2.5$ is below the cutoff of approximately $-1.92$.
For each candidate $\mu$, the nuisance parameter $\sigma$ is
re-estimated to give the highest log-likelihood.

:::
::::

# Fitting overdispersed seedling counts {#sec-negative-binomial-fit}

The [Negative Binomial model](../Supplements/distributions.qmd#sec-negative_binomial)
allows counts to vary more than expected under a Poisson model with
the same mean. Write its mean as $\mu$ seedlings per quadrat and its
shape parameter as $k>0$:

$$
E[X]=\mu,\qquad
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k}.
$$

We will simulate 80 independent, equal-area quadrats under this
model. The generating values are known because the data are
simulated, not because an ecologist could know them in a field study.
In R, `size` corresponds to $k$ and `mu` to $\mu$.

```r
# Simulate overdispersed seedling counts from a negative binomial model.
set.seed(seed = 530)
seedling_counts_nb = rnbinom(
  n = 80, size = 1.5, mu = 4
)
mean(seedling_counts_nb)
var(seedling_counts_nb)
```

The variance-to-mean comparison is exploratory. It does not
identify why counts vary. To fit the model, use each count's
negative binomial probability mass as its likelihood contribution.
Both parameters must stay positive, so we optimise on their log scales:

```r
# Define the negative binomial negative log-likelihood on two log parameter scales.
negative_binomial_nll = function(par, counts) {
  mu = exp(par[1])
  k = exp(par[2])
  -sum(dnbinom(
    x = counts, size = k, mu = mu, log = TRUE
  ))
}

# Use sample summaries to supply valid starting values for both parameters.
start_mu = max(mean(seedling_counts_nb), 0.1)
sample_variance = var(seedling_counts_nb)
start_k = start_mu^2 /
  max(sample_variance - start_mu, 0.1)

nb_fit = optim(
  par = log(c(mu = start_mu, k = start_k)),
  fn = negative_binomial_nll, counts = seedling_counts_nb,
  method = "BFGS", hessian = TRUE
)
nb_fit$convergence
# Transform the fitted parameters back to the mean and shape scales.
nb_estimates = exp(nb_fit$par)
nb_estimates
```

The starting value for $k$ comes from rearranging the model
mean–variance relation, as in Chapter 3's method of moments. The
small lower bound here only supplies a valid starting value if the
sample variance is close to or below its mean. It is not a final
estimate or evidence of overdispersion.

Check the convergence code, whether estimates are ecologically
plausible, and whether the fitted frequencies describe the sample.
Different starting values can reveal a sensitive optimisation,
although agreement does not prove that a global maximum was found.
For uncertainty, we can use the inverse Hessian on
$(\log\mu,\log k)$ or profile one parameter while refitting the
other. Those intervals inherit the approximation limits described
above.

The comparison below places observed **relative frequencies**
and fitted **probabilities** on the same scale:

```r
# List possible counts and calculate observed relative frequencies and fitted masses.
count_values = 0:max(seedling_counts_nb)
observed_frequency = as.numeric(
  table(factor(seedling_counts_nb, levels = count_values))
) / length(seedling_counts_nb)
fitted_probability = dnbinom(
  x = count_values, size = nb_estimates["k"],
  mu = nb_estimates["mu"]
)
# Plot the sample frequencies against the fitted negative binomial probabilities.
plot(
  x = count_values, y = observed_frequency,
  type = "h", lwd = 5, col = "grey60",
  xlab = "Seedlings per quadrat",
  ylab = "Relative frequency or probability",
  ylim = c(0, max(observed_frequency, fitted_probability))
)
lines(
  x = count_values, y = fitted_probability,
  type = "b", pch = 19, lwd = 2
)
legend(
  "topright",
  legend = c("Observed relative frequency", "Fitted probability"),
  col = c("grey60", "black"), lty = 1, lwd = c(5, 2),
  pch = c(NA, 19), bty = "n"
)
```

Look for systematic differences, especially in the zeros and
largest counts, rather than expecting every bar to touch the line.

::: {.callout-warning title="A fitted model still needs checking"}
The optimiser chooses the best parameter values *within the
specified model*. Compare fitted and observed counts, especially
zeros and the upper tail. A mismatch may indicate that the
distribution or independence assumptions need reconsideration.
:::

:::: {.callout-important title="Practice: mean and shape"}

Under the negative binomial model, what happens to count variance
when $k$ becomes smaller while $\mu$ stays fixed? If a fit returns
`par = c(log(4), log(2))`, what are the estimates on the original
scale?

::: {.callout-tip title="Solution" collapse="true"}

Because $\operatorname{Var}(X)=\mu+\mu^2/k$, a smaller $k$ gives
larger variance for the same mean. The fitted values are
$\hat\mu=4$ seedlings per quadrat and $\hat k=2$.

:::
::::

# Summary

A likelihood uses the probability model to compare parameter values
for **observed** data. For independent observations, multiply their
probability masses or densities; for a log-likelihood, add their
logs. Maximum likelihood chooses the parameter values at the
peak. In R, `optim()` can find the peak by minimising the negative
log-likelihood.

For several parameters, a surface shows their joint fit. A slice
holds other parameters fixed; a profile refits them. Curvature
near the peak gives a quadratic uncertainty approximation, while
a profile interval follows the refitted likelihood. Neither
fitting nor interval calculation removes the need to check the
model's ecological assumptions.

# Chapter quiz

:::: {.callout-important title="Question 1: data or parameter?"}

After observing quadrat counts, what varies along a Poisson
likelihood curve? Does the curve assign probabilities to values
of $\lambda$?

::: {.callout-tip title="Answer" collapse="true"}

The candidate mean $\lambda$ varies while the observed counts
stay fixed. The curve compares their fit; it is not a probability
distribution for $\lambda$.

:::
::::

:::: {.callout-important title="Question 2: calculate a likelihood"}

Two independent quadrats contain 0 and 1 seedlings. Under a
Poisson model with $\lambda=1$, write the two contributions,
their product, and the log-likelihood without R.

::: {.callout-tip title="Answer" collapse="true"}

The contributions are $P(X=0)=e^{-1}$ and $P(X=1)=e^{-1}$.
Their product is $e^{-2}$, so the log-likelihood is $-2$.

:::
::::

:::: {.callout-important title="Question 3: what does `optim()` minimise?"}

Why did we put a minus sign before the sum of log probabilities?
Why did we exponentiate the fitted Poisson parameter?

::: {.callout-tip title="Answer" collapse="true"}

`optim()` minimises by default, so minimising the negative
log-likelihood maximises the log-likelihood. We optimised
$\log\lambda$ to keep $\lambda$ positive and exponentiated it to
report the mean on the count scale.

:::
::::

:::: {.callout-important title="Question 4: slice and profile"}

For each proposed normal-model mean, a researcher re-estimates
the standard deviation. Which likelihood curve results?

::: {.callout-tip title="Answer" collapse="true"}

A profile likelihood for the mean.

:::
::::

:::: {.callout-important title="Question 5: interval meaning"}

Can a quadratic interval and a profile interval have different
endpoints for the same data? Does either interval give the
probability that the fixed parameter lies inside it?

::: {.callout-tip title="Answer" collapse="true"}

Yes. They use different approximations to the likelihood's
shape. Neither assigns a probability to the fixed parameter.
Their confidence levels describe repeated use of the interval
procedures under their assumptions.

:::
::::
