---
title: "Estimation and sampling distributions"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to turn a limited ecological sample into an
estimate of something we actually want to know. In [Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions),
we could specify a count model and calculate its behaviour once we knew its
parameters. [Chapter 3](../Chapter_3/Theory.qmd) did the same for continuous
measurements. Fieldwork gives us the awkward reverse problem: we have a set of
counts from sampled quadrats, but we do not know the mean number of seedlings
per quadrat that the model needs.

It would be comforting if one sample settled the matter. It does not. Another
reasonable sample from the same population would probably give a different
answer, even if nobody made a mistake. This is where estimation earns its place:
we need a rule for extracting a parameter from data, and a way to describe how
much that answer would move between comparable studies. We will use seedling
counts as a running example, then estimate the mean and variation of tree
heights. Whenever we simulate data, we will say which model generated them;
simulation is a way to expose uncertainty, not a substitute for field evidence.

# Learning goals

After studying this chapter, you should be able to:

- distinguish a parameter, an estimator, and an estimate;
- calculate simple method-of-moments estimators;
- describe and simulate the sampling distribution of an estimator;
- explain estimator bias, variance, and mean squared error;
- calculate and interpret an estimated standard error;
- construct and interpret a confidence interval;
- explain consistency and asymptotic normality in practical terms;
- use repeated simulation to evaluate estimators and interval coverage;
- explain why estimation matters when fitting ecological models.

# From a model to an estimate {#sec-estimates}

Suppose we want to describe seedling abundance across a restoration site. Let
$X_i$ be the number of seedlings in quadrat $i$. A deliberately simple starting
model is

$$
X_i \sim \operatorname{Poisson}(\lambda),
$$

where $\lambda$ is the expected count **per quadrat**. We are assuming that
quadrats have the same area and expected count, and that their counts are
independent. Those are model assumptions, not rewards we receive for sampling
at different locations. The [Poisson model](../Supplements/distributions.qmd#sec-poisson)
describes one individual count. Here the interesting quantity, $\lambda$, is
unknown.

::: {.callout-note title="Parameter, estimator, and estimate"}
A **parameter** is a fixed, unknown feature of the assumed model, such as
$\lambda$. An **estimator** is a rule applied to random observations, so its
result can vary between samples. An **estimate** is the number obtained by
applying that rule to one particular sample. A hat marks an estimated
parameter, such as $\hat\lambda$.
:::

Imagine that ten quadrats give us these counts:

```r
# Record the seedling counts from ten illustrative quadrats.
seedling_counts = c(3, 5, 2, 7, 4, 3, 4, 1, 6, 5)
# Calculate their method-of-moments estimate of the Poisson mean.
mean(seedling_counts)
```

Their mean is 4 seedlings per quadrat. That is our answer from these ten
quadrats, not a declaration that the unknown population mean is exactly 4.
Keeping that distinction in view will stop several common mistakes later on.

## Method of moments

The sample mean feels like a natural answer here because the Poisson parameter
is itself a mean. The **method of moments** turns that intuition into a general
recipe: match a sample summary to the corresponding model summary, then solve
for the parameter. In [Chapter 2](../Chapter_2/Theory.qmd#sec-sample_population),
we calculated the mean of an empirical distribution. A *moment* is a model
expectation, such as $E[X]$; the method asks us to make that expectation agree
with what the sample says.

For a Poisson count, $E[X_i]=\lambda$. The corresponding sample moment is
the mean of $n$ counts,

$$
\overline X = \frac{1}{n}\sum_{i=1}^{n}X_i.
$$

Matching them gives the method-of-moments estimator

$$
\hat\lambda = \overline X.
$$

Applied to our illustrative sample, it gives the **estimate**
$\hat\lambda=4$ seedlings per quadrat. The method has not made the sample
perfectly representative. It has supplied a transparent rule for learning from
it. We will return to this rule when a model has two parameters.

:::: {.callout-important title="Practice: estimate from quadrats"}

A second illustrative sample has counts 2, 4, 3, 5, and 1. What are the
parameter, estimator, and estimate under a Poisson model? Give the estimate
in ecological units.

::: {.callout-tip title="Solution" collapse="true"}

The unknown parameter is $\lambda$, the expected number of seedlings per
quadrat. The estimator is $\hat\lambda=\overline X$. For these five counts,
the estimate is $(2+4+3+5+1)/5=3$ seedlings per quadrat. A different sample
could give a different estimate.

:::
::::

# What changes between samples? {#sec-sampling_distribution}

Our illustrative sample gave an estimate of 4, but it cannot show us on its
own how unreliable that estimate might be. To make the repeated-study idea
visible, suppose the Poisson model really has $\lambda=4$ and each study samples
30 independent quadrats. We know the parameter only because **we chose it for
the simulation**.

The **sampling distribution** is the distribution of $\hat\lambda$ across
repeated studies of this kind. Each point in it is a mean calculated from a
whole group of 30 quadrats. That is not the same thing as the Poisson
distribution of counts in *individual* quadrats. The distinction matters:
ecologists often need uncertainty about an estimated mean, not another
description of variation among individual plots.

The code below repeats the entire study 2,000 times. Every repetition returns
one estimate, as if a new team had sampled another set of comparable quadrats.

```r
# Set the model parameters and number of repeated simulated studies.
set.seed(seed = 401)
true_lambda = 4
n_quadrats = 30
n_studies = 2000

# Simulate one study at a time and save its estimated mean count.
lambda_estimates = replicate(
  n = n_studies,
  expr = mean(rpois(n = n_quadrats, lambda = true_lambda))
)

# Summarise the centre and spread of the simulated estimates.
mean(lambda_estimates)
sd(lambda_estimates)
```

The histogram shows the estimates from those imaginary studies. Its horizontal
axis is a mean count per quadrat, not a count from one quadrat. That is why the
values are no longer restricted to whole numbers.

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Estimates of the mean seedling count from 2,000 simulated studies of 30 quadrats each."
# Set the model parameters and simulate one mean estimate per study.
set.seed(seed = 401)
true_lambda = 4
n_quadrats = 30
lambda_estimates = replicate(
  n = 2000,
  expr = mean(rpois(n = n_quadrats, lambda = true_lambda))
)
# Plot the sampling distribution and mark the known model mean.
hist(
  x = lambda_estimates, breaks = 28, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Estimated seedlings per quadrat",
  ylab = "Density", main = ""
)
abline(v = true_lambda, lty = 2, lwd = 2)
legend(
  "topright", legend = "Model mean", lty = 2, lwd = 2, bty = "n"
)
```

## Bias and variance

Before we decide that an estimator is doing a good job, we need to ask two
separate questions. Does it miss the target systematically, and how much does
it move from study to study? The **bias** of an estimator is the difference
between its average over repeated samples and the parameter it aims to estimate.
We use $\theta$ for a general parameter and $\hat\theta$ for its estimator:

$$
\operatorname{Bias}(\hat\theta)=E[\hat\theta]-\theta.
$$

An unbiased estimator has an average estimate equal to the parameter. Under
independent Poisson sampling, $E[\hat\lambda]=\lambda$, so the sample mean is
unbiased for $\lambda$. This does *not* mean each realised estimate lands on
$\lambda$. An estimate above or below it is ordinary sampling variation, not
evidence of bias.

The **variance of an estimator** answers the second question: how much do its
estimates vary across studies? For the Poisson sample mean,

$$
\operatorname{Var}(\hat\lambda)=\frac{\lambda}{n}.
$$

Under this model, larger studies have a narrower sampling distribution. The
variance of individual quadrat counts remains $\lambda$; it is the variation
among their *means* that decreases. More quadrats do not make the site more
uniform. They make our estimate of its mean less variable. We can compare these
theoretical properties with the simulation:

```r
# Compare the simulated bias and variance with the Poisson model variance.
c(
  simulated_bias = mean(lambda_estimates) - true_lambda,
  simulated_variance = var(lambda_estimates),
  model_variance = true_lambda / n_quadrats
)
```

The simulated values will differ slightly from the model values because 2,000
repetitions are still a finite collection of imaginary studies.

## Mean squared error

Bias and spread both matter if we want estimates that tend to stay near the
parameter. The **mean squared error** (MSE) combines them: it is the average
squared distance between estimates and the parameter.

$$
\operatorname{MSE}(\hat\theta)
= E[(\hat\theta-\theta)^2]
= \operatorname{Var}(\hat\theta)
  + \operatorname{Bias}(\hat\theta)^2.
$$

For the unbiased Poisson sample mean, the theoretical MSE equals its variance.
In the simulation, calculate MSE with
`mean((lambda_estimates - true_lambda)^2)`. It need not exactly equal
`var(lambda_estimates)`: the simulation is finite, and `var()` also uses the
$n-1$ divisor across the simulated estimates.

:::: {.callout-important title="Practice: what varies?"}

Suppose two independent studies each sample 30 quadrats and report means of
3.7 and 4.2 seedlings per quadrat. What varies between the studies? Do
these two estimates establish that the estimator is biased?

::: {.callout-tip title="Solution" collapse="true"}

The samples and their estimates vary. Under the assumed frequentist model,
the parameter $\lambda$ is fixed. Two estimates cannot establish bias,
which concerns the estimator's average over repeated samples.

:::
::::

# Standard errors {#sec-standard_errors}

The sampling distribution gives the uncertainty we care about, but its
variance is in squared units. The **standard error** (SE) puts that spread back
into the estimator's original units: it is the standard deviation of an
estimator's sampling distribution.

$$
\operatorname{SE}(\hat\theta)
= \sqrt{\operatorname{Var}(\hat\theta)}.
$$

For the Poisson sample mean, the model gives
$\operatorname{SE}(\hat\lambda)=\sqrt{\lambda/n}$. Of course, in the real
study the whole problem is that $\lambda$ is unknown. We replace it with our
estimate:

$$
\widehat{\operatorname{SE}}(\hat\lambda)
= \sqrt{\frac{\hat\lambda}{n}}.
$$

This is a **plug-in estimate**: we put an estimated parameter into a formula
that calls for the unknown parameter. For the illustrative counts, the estimated
SE is:

```r
# Estimate the Poisson mean and record the number of sampled quadrats.
lambda_hat = mean(seedling_counts)
n_quadrats = length(seedling_counts)
# Use the estimated mean in the Poisson standard-error formula.
estimated_se = sqrt(lambda_hat / n_quadrats)
estimated_se
```

It is approximately 0.63 seedlings per quadrat. This describes the typical
variation in the *estimated mean* between studies of ten quadrats, provided the
Poisson assumptions are a reasonable description of the study.

::: {.callout-warning title="Standard deviation and standard error answer different questions"}
The standard deviation of the ten counts describes variation between
quadrats. The estimated standard error describes variation in their *mean*
between repeated studies. More quadrats can reduce the latter without
making seedlings equally abundant in every quadrat.
If counts are overdispersed relative to the Poisson model, its standard
error formula can understate the sampling variability.
:::

:::: {.callout-important title="Practice: change the study size"}

If the true mean is 4 seedlings per quadrat, what happens to the standard
error of the sample mean when a study grows from 25 to 100 independent
quadrats? Does the standard deviation of individual Poisson counts change?

::: {.callout-tip title="Solution" collapse="true"}

The standard error changes from $\sqrt{4/25}=0.4$ to
$\sqrt{4/100}=0.2$ seedlings per quadrat: it halves. Under the same Poisson
model, the standard deviation of an individual count remains
$\sqrt{4}=2$ seedlings per quadrat.

:::
::::

# Confidence intervals {#sec-confidence_intervals}

An estimate is a useful start, but reporting only one number can make it look
far more settled than the data justify. A **confidence interval** combines an
estimator with a rule for its sampling variation to produce a range of parameter
values. Its confidence level describes how that *procedure* behaves across
repeated studies.

If a sampling distribution is sufficiently close to normal, a common
approximate 95% interval is

$$
\hat\theta \pm 1.96\,\widehat{\operatorname{SE}}(\hat\theta).
$$

The number 1.96 is the 97.5th percentile of a standard normal distribution.
Using the illustrative seedling counts and the Poisson standard error gives a
**Wald interval** for $\lambda$:

```r
# Obtain the standard-normal critical value for a two-sided 95% interval.
critical_value = qnorm(p = 0.975)
# Combine the estimate, standard error, and critical value to find the endpoints.
interval = lambda_hat + c(-1, 1) * critical_value * estimated_se
interval
```

The endpoints are approximately 2.76 and 5.24 seedlings per quadrat. They are
possible values for the *mean count*, not limits on individual quadrat counts.
The calculation relies on the Poisson model, independence, and a normal
approximation for the estimator. We should not let the tidy pair of endpoints
hide those assumptions.

::: {.callout-note title="What 95% confidence means"}
If we repeatedly sampled quadrats and constructed an interval by the same
procedure, about 95% of the intervals would include the fixed true
parameter **when the procedure has 95% coverage under the assumed model**.
The endpoints change between studies; the parameter does not.
:::

::: {.callout-warning title="Check the interval procedure"}
A 95% label does not guarantee 95% coverage in every setting. A normal
approximation to Poisson counts can work poorly for small expected totals.
A Wald interval can even have a negative lower endpoint, although a
Poisson mean cannot be negative. An interval from one study does not give
a 95% probability that the fixed parameter lies between its endpoints.
:::

## Checking coverage by simulation

In a field study, we cannot check whether an interval captured the truth because
the truth is what we are trying to learn. Simulation gives us an honest practice
ground: we set $\lambda$, then check whether each interval includes it. The
proportion that do is the simulated **coverage**. The following code repeats
both the sampling and interval construction:

```r
# Set the model parameters and interval rule for repeated simulated studies.
set.seed(seed = 402)
true_lambda = 4
n_quadrats = 30
n_studies = 2000
critical_value = qnorm(p = 0.975)

# Simulate each study, construct its interval, and record whether it covers the truth.
covered = replicate(
  n = n_studies,
  expr = {
    # Generate one study and calculate its estimated mean and standard error.
    counts = rpois(n = n_quadrats, lambda = true_lambda)
    estimate = mean(counts)
    se = sqrt(estimate / n_quadrats)
    # Construct the interval and test whether it includes the true model mean.
    lower = estimate - critical_value * se
    upper = estimate + critical_value * se
    lower <= true_lambda && true_lambda <= upper
  }
)

# Calculate the simulated coverage across all studies.
mean(covered)
```

Simulated coverage changes with the seed and number of repetitions. It can also
change with sample size, the true parameter, and the interval method. This is a
useful reminder that “95% confidence interval” is a claim about a procedure,
not a magic label we attach after fitting a model. We cannot calculate coverage
from one field study because its true parameter is unknown.

:::: {.callout-important title="Practice: interpret an interval"}

A study reports a 95% confidence interval of 2.8 to 5.2 seedlings per
quadrat. What does the 95% refer to? What would need to be checked before
trusting this interval for an ecological conclusion?

::: {.callout-tip title="Solution" collapse="true"}

The 95% refers to the long-run coverage of the interval-making procedure,
not a probability assigned to the fixed parameter after seeing this one
interval. We would examine whether the Poisson model and independence are
reasonable and whether the approximation used for the interval is adequate.

:::
::::

# Estimating two tree-height parameters {#sec-two_parameters}

Counts are not special. The same estimation questions appear when we measure
continuous variables. In [Chapter 3](../Chapter_3/Theory.qmd#sec-continuous_four_faces),
we used a normal model for tree height. Suppose $X_i$ is the height in metres
of tree $i$ and independent trees follow

$$
X_i \sim \operatorname{Normal}(\mu,\sigma^2).
$$

The model mean is $\mu$ metres and its variance is $\sigma^2$ square metres.
R's `rnorm()` takes the standard deviation $\sigma$ in its `sd` argument. A
normal model is an approximation, not an ecological law. It allows negative
heights, so we need its tails to be implausible only in a part of the scale that
does not matter for our trees.

## Method of moments for mean and variance

Now the model has two unknown parameters, so one sample mean is not enough. The
first model moment is $E[X]=\mu$, giving the method-of-moments estimator
$\hat\mu=\overline X$. A second relationship is $E[X^2]=\mu^2+\sigma^2$.
Matching it to the sample mean of the squared heights gives

$$
\widehat{\sigma^2}_{\mathrm{MM}}
= \frac{1}{n}\sum_{i=1}^{n}(X_i-\overline X)^2.
$$

The method now gives us a rule for both parameters. You do not need to memorise
the algebraic derivation to use it: calculate the mean height, then average the
squared deviations from that mean.

```r
# Simulate tree heights from an assumed normal population.
set.seed(seed = 403)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)

# Calculate the mean and compare the two variance divisors.
height_mean = mean(tree_heights)
variance_mm = mean((tree_heights - height_mean)^2)
variance_with_var = var(tree_heights)

c(
  mean_m = height_mean,
  variance_mm_m2 = variance_mm,
  variance_with_var_m2 = variance_with_var
)
```

The two variance estimates differ. `variance_mm` divides by $n$; `var()` divides
by $n-1$. This is not a quirk that R invented. Under independent sampling from
the same distribution with finite variance,

$$
E[\widehat{\sigma^2}_{\mathrm{MM}}]
= \frac{n-1}{n}\sigma^2,
\qquad
E[s^2]=\sigma^2,
$$

where $s^2$ denotes the result of `var()`. The method-of-moments variance
estimator is therefore biased downward by $\sigma^2/n$. It is still useful, and
its bias becomes smaller as the sample size grows. Bias alone does not tell us
which estimator has smaller MSE, so “unbiased” is not a complete scorecard.

::: {.callout-note title="Why the divisor matters"}
The empirical variance introduced in Chapter 2 uses the divisor $n$.
R's `var()` uses $n-1$ to estimate the variance of the population or
model from which the sample came. Name the quantity you are calculating
before choosing the divisor.
:::

## A confidence interval for mean height

The standard error of the sample mean is $\sigma/\sqrt n$. Once again, the
unknown parameter appears in the formula, so we estimate it by $s/\sqrt n$,
where $s=\texttt{sd(tree_heights)}$. For independent observations from a normal
model, a 95% interval for $\mu$ uses a $t$ quantile:

$$
\overline X \pm t_{n-1,0.975}\frac{s}{\sqrt n}.
$$

The $t$ distribution accounts for estimating $\sigma$ from the same sample.
Under the stated normal model, this procedure has 95% coverage. Applied to the
simulated heights:

```r
# Record the sample size and calculate the standard error of the mean height.
n_trees = length(tree_heights)
height_se = sd(tree_heights) / sqrt(n_trees)
# Use the appropriate t critical value to construct the confidence interval.
t_critical = qt(p = 0.975, df = n_trees - 1)
height_interval = height_mean + c(-1, 1) * t_critical * height_se
height_interval
```

The interval is for the population **mean height in metres**, not a range that
contains 95% of individual trees. That interpretation is tempting because it
sounds similar, but it answers a different question. With skewed measurements,
such as body mass, the same $t$ interval for the original-scale mean is only
approximate; we will investigate that in the practical.

:::: {.callout-important title="Practice: two kinds of spread"}

Which R expression estimates variation among the 25 tree heights, and
which estimates the standard error of their mean? Why does an interval
for the mean use the second quantity?

::: {.callout-tip title="Solution" collapse="true"}

`sd(tree_heights)` estimates variation among trees.
`sd(tree_heights) / sqrt(length(tree_heights))` estimates the
standard error of the sample mean. The interval describes uncertainty
in the mean height, so it uses the sampling variation of the mean.

:::
::::

# What larger samples change {#sec-large_samples}

We saw in [Chapter 2](../Chapter_2/Theory.qmd#sec-sample_population) that
empirical summaries can settle down as sample size increases. That is the
comforting part of collecting more data, but it needs a careful interpretation.
For an estimator, **consistency** means that the probability of being far from
its target parameter tends to zero as sample size grows. Under the independent
Poisson model, the sample mean is consistent for $\lambda$.

Its standard error is $\sqrt{\lambda/n}$, so multiplying the sample size by
four halves the standard error. This describes the *distribution across
possible studies*. It does not promise that every larger realised sample will
be closer to the truth. A larger sample can still be unlucky, and a beautifully
large sample cannot repair a sampling design that misses the population of
interest.

Many sample means also have an **approximately normal sampling distribution**
when sample size is sufficiently large, even if individual observations are not
normal. This large-sample behaviour is called **asymptotic normality**. It is
the reason the normal approximation above is often useful, not a reason to use
it without looking at the setting. For the Poisson mean,

$$
\hat\lambda \overset{\mathrm{approx}}{\sim}
\operatorname{Normal}\left(\lambda,\frac{\lambda}{n}\right)
\quad\text{for sufficiently large }n.
$$

The approximation depends on the setting. A small expected total count
$n\lambda$ can leave the sampling distribution discrete and asymmetric. This is
why checking an interval procedure's coverage is more informative than assuming
that a familiar formula must work.

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Sampling distributions of mean seedling counts for studies of 5 and 100 quadrats. Curves show normal approximations under the assumed Poisson model."
# Set the Poisson model and the study sizes to compare.
set.seed(seed = 404)
true_lambda = 4
sample_sizes = c(5, 100)
# Save the current graphics settings before drawing two panels.
old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))
for (sample_size in sample_sizes) {
  # Simulate repeated studies and calculate one estimated mean per study.
  estimates = replicate(
    n = 2000,
    expr = mean(rpois(n = sample_size, lambda = true_lambda))
  )
  # Use the histogram density to choose a vertical scale for the plot.
  histogram = hist(x = estimates, breaks = 25, plot = FALSE)
  # Plot the sampling distribution and overlay its normal approximation.
  hist(
    x = estimates, breaks = 25, probability = TRUE,
    col = "grey85", border = "white",
    ylim = c(0, max(histogram$density) * 1.25),
    xlab = "Estimated seedlings per quadrat",
    ylab = "Density", main = paste("Quadrats:", sample_size)
  )
  curve(
    expr = dnorm(
      x = x, mean = true_lambda,
      sd = sqrt(true_lambda / sample_size)
    ),
    add = TRUE, lwd = 2, lty = 2
  )
  legend(
    "topleft", legend = "Normal approximation",
    lty = 2, lwd = 2, bty = "n", cex = 0.8
  )
}
# Restore the graphics settings used before this figure.
par(old_par)
```

The larger-sample estimates are more concentrated around 4, and the normal
curve describes their shape more closely. Individual quadrat counts have not
become normal; they are still Poisson counts. We have changed the distribution
of their *mean*, not the process in each quadrat.

:::: {.callout-important title="Practice: bigger is not guaranteed closer"}

A study of 20 quadrats estimates 3.9 seedlings per quadrat; another
study of 100 estimates 4.2. The true model mean is 4. Did the larger
study disprove consistency?

::: {.callout-tip title="Solution" collapse="true"}

No. The smaller study happened to give an estimate nearer 4. Consistency
concerns how the *probability* of large errors changes as sample size
increases, not the ordering of two realised estimates.

:::
::::

# What simulation can tell us {#sec-simulation_limits}

Simulation lets us specify a model and known parameters, generate samples, apply
an estimator to each one, and compare the repeated results with the truth. That
is how we studied bias, standard errors, MSE, and interval coverage. It is a
particularly useful teaching tool because field data rarely come with the true
parameter written in the margin. More repetitions usually make a simulated
summary more stable; they do not make an unrealistic ecological model realistic.

Our Poisson studies assumed independent quadrats of equal area with the same
mean count. A real restoration site may have spatial clustering, changing
habitat, or imperfect detection. Any of these could change both the count
distribution and the performance of our estimates and intervals. The calculation
is therefore only one part of the work. We also need to examine the sampling
design, the observations, and the ecological story before drawing conclusions.

This chapter has treated parameters as fixed and samples as variable. In
[Chapter 5](../Chapter_5/Theory.qmd), we will use the probability model more
fully through **maximum likelihood estimation**. The questions stay remarkably
similar: What does the parameter mean? What estimate did we obtain? How much
should we trust it?

# Summary

Estimation starts with a model and a limited sample, then asks what that sample
can tell us about an unknown parameter. Method of moments matches sample
summaries to model moments. An estimator is the rule; an estimate is the number
that rule gives for one sample. Repeated samples reveal an estimator's bias and
spread, while a standard error puts that spread in the parameter's units.

A confidence level describes coverage of an interval-making procedure across
repeated studies; it is not a probability attached to a fixed parameter after
we have seen the data. Simulation lets us check these ideas under assumptions
we specify. Whether those assumptions are defensible still depends on the
ecological setting and the way observations were collected.

# Chapter quiz

The questions below ask you to keep the model, the sample, and the ecological
meaning of an estimate separate. Try each one before opening its answer.

:::: {.callout-important title="Question 1: an estimate"}

The mean of counts from 20 quadrats is 5.1 seedlings per quadrat.
Under a Poisson model, name the parameter, estimator, and estimate.

::: {.callout-tip title="Answer" collapse="true"}

The parameter is the unknown model mean $\lambda$. The estimator is
$\hat\lambda=\overline X$. Its estimate for these counts is 5.1
seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 2: what does the histogram show?"}

A histogram contains one sample mean from each of 1,000 simulated studies.
Is it an empirical distribution of individual observations or an
approximation to a sampling distribution?

::: {.callout-tip title="Answer" collapse="true"}

It approximates the sampling distribution of the sample mean. Each
histogram entry is one estimate from a whole study.

:::
::::

:::: {.callout-important title="Question 3: bias and spread"}

An estimator's average over repeated studies equals its target, but
individual estimates vary considerably. Is it biased? Is its variance zero?

::: {.callout-tip title="Answer" collapse="true"}

It is unbiased because its repeated-sampling mean equals the target.
Its variance is greater than zero because its estimates vary.

:::
::::

:::: {.callout-important title="Question 4: standard error"}

For 100 independent quadrats under a Poisson model with $\lambda=4$,
what is the standard error of the estimated mean count? What would
the standard deviation of one quadrat count be?

::: {.callout-tip title="Answer" collapse="true"}

The standard error is $\sqrt{4/100}=0.2$ seedlings per quadrat. The
standard deviation of one count is $\sqrt{4}=2$ seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 5: interval meaning"}

Does a calculated 95% confidence interval assign 95% probability to
the fixed parameter lying inside it? Explain what 95% describes.

::: {.callout-tip title="Answer" collapse="true"}

No. Under the assumed model, 95% describes the proportion of intervals
that the procedure would expect to cover the fixed parameter over
repeated studies, if the procedure achieves its nominal coverage.

:::
::::

:::: {.callout-important title="Question 6: estimating variation"}

Which divisor does the method-of-moments variance estimator use, and
which does R's `var()` use? What happens to the former's bias as
sample size grows?

::: {.callout-tip title="Answer" collapse="true"}

The method-of-moments estimate divides by $n$; `var()` divides by
$n-1$. Under the stated independent sampling assumptions, the
method-of-moments estimator's bias is $-\sigma^2/n$, which approaches
zero as $n$ grows.

:::
::::
