---
title: "Estimation and sampling distributions"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to estimate unknown parameters in ecological
probability models and describe the uncertainty in those estimates. In
[Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions), we used
probability models to describe counts when their parameters were known. In
[Chapter 3](../Chapter_3/Theory.qmd), we did the same for continuous
measurements. An ecologist usually faces the question in the other direction:
given counts from sampled quadrats, what can we learn about the unknown mean
number of seedlings per quadrat?

One sample can give us an **estimate**. Another sample from the same population
would probably give a different estimate. This chapter explains how to
calculate an estimate and how to describe that variation between samples. We
will use seedling counts as a running example, then estimate the mean and
variation of tree heights. Whenever we simulate data, we will specify the
model that generated them; simulated data are not field observations.

# Learning goals

After studying this chapter, you should be able to:

- distinguish a parameter, an estimator, and an estimate;
- calculate simple method-of-moments estimators;
- describe and simulate the sampling distribution of an estimator;
- explain estimator bias, variance, and mean squared error;
- calculate and interpret an estimated standard error;
- construct and interpret a confidence interval;
- explain consistency and asymptotic normality in practical terms;
- use repeated simulation to evaluate estimators and interval coverage;
- explain why estimation matters when fitting ecological models.

# From a model to an estimate {#sec-estimates}

Suppose $X_i$ is the number of seedlings in quadrat $i$. A starting model is

$$
X_i \sim \operatorname{Poisson}(\lambda),
$$

where $\lambda$ is the expected count **per quadrat**. We assume quadrats
have the same area and expected count, and that their counts are independent.
Those assumptions belong to the model; sampling different locations does not
establish them. The [Poisson model in Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions)
describes the distribution of an individual count. Here, $\lambda$ is unknown.

::: {.callout-note title="Parameter, estimator, and estimate"}
A **parameter** is a fixed, unknown feature of the assumed model, such as
$\lambda$. An **estimator** is a rule applied to random observations, so its
result can vary between samples. An **estimate** is the number obtained by
applying that rule to one particular sample. A hat marks an estimated
parameter, such as $\hat\lambda$.
:::

For illustration, consider these counts from ten quadrats:

```r
# Record the seedling counts from ten illustrative quadrats.
seedling_counts = c(3, 5, 2, 7, 4, 3, 4, 1, 6, 5)
# Calculate their method-of-moments estimate of the Poisson mean.
mean(seedling_counts)
```

Their mean is 4 seedlings per quadrat. This is an illustrative sample; it
does not tell us that the unknown population mean is exactly 4.

## Method of moments

In [Chapter 2](../Chapter_2/Theory.qmd#sec-sample_population), we calculated
the mean of an empirical distribution. The **method of moments** uses such a
sample summary to estimate a model parameter: match a sample moment to its
corresponding model moment, then solve for the parameter. A moment is an
expected value, such as the mean $E[X]$.

For a Poisson count, $E[X_i]=\lambda$. The corresponding sample moment is
the mean of $n$ counts,

$$
\overline X = \frac{1}{n}\sum_{i=1}^{n}X_i.
$$

Matching them gives the method-of-moments estimator

$$
\hat\lambda = \overline X.
$$

Applied to our illustrative sample, it gives the **estimate**
$\hat\lambda=4$ seedlings per quadrat. We will return to the method of
moments when a model has two parameters.

:::: {.callout-important title="Practice: estimate from quadrats"}

A second illustrative sample has counts 2, 4, 3, 5, and 1. What are the
parameter, estimator, and estimate under a Poisson model? Give the estimate
in ecological units.

::: {.callout-tip title="Solution" collapse="true"}

The unknown parameter is $\lambda$, the expected number of seedlings per
quadrat. The estimator is $\hat\lambda=\overline X$. For these five counts,
the estimate is $(2+4+3+5+1)/5=3$ seedlings per quadrat. A different sample
could give a different estimate.

:::
::::

# What changes between samples? {#sec-sampling_distribution}

The illustrative sample gave an estimate of 4. To see what could happen in
other studies, suppose the Poisson model really has $\lambda=4$ and each
study samples 30 independent quadrats. We know the parameter here because
**we chose it for the simulation**.

The **sampling distribution** is the distribution of $\hat\lambda$ across
repeated studies of this kind. It describes means calculated from groups of
30 quadrats. The Poisson distribution describes counts in *individual*
quadrats. These are different quantities.

The code below repeats the whole study 2,000 times. Each repetition returns
one estimate.

```r
# Set the model parameters and number of repeated simulated studies.
set.seed(seed = 401)
true_lambda = 4
n_quadrats = 30
n_studies = 2000

# Simulate one study at a time and save its estimated mean count.
lambda_estimates = replicate(
  n = n_studies,
  expr = mean(rpois(n = n_quadrats, lambda = true_lambda))
)

# Summarise the centre and spread of the simulated estimates.
mean(lambda_estimates)
sd(lambda_estimates)
```

The histogram displays the estimates. Its horizontal axis is a mean count
per quadrat, not a count from one quadrat.

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Estimates of the mean seedling count from 2,000 simulated studies of 30 quadrats each."
# Set the model parameters and simulate one mean estimate per study.
set.seed(seed = 401)
true_lambda = 4
n_quadrats = 30
lambda_estimates = replicate(
  n = 2000,
  expr = mean(rpois(n = n_quadrats, lambda = true_lambda))
)
# Plot the sampling distribution and mark the known model mean.
hist(
  x = lambda_estimates, breaks = 28, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Estimated seedlings per quadrat",
  ylab = "Density", main = ""
)
abline(v = true_lambda, lty = 2, lwd = 2)
legend(
  "topright", legend = "Model mean", lty = 2, lwd = 2, bty = "n"
)
```

## Bias and variance

The **bias** of an estimator is the difference between its average over
repeated samples and the parameter it aims to estimate. We use $\theta$
for a general parameter and $\hat\theta$ for its estimator:

$$
\operatorname{Bias}(\hat\theta)=E[\hat\theta]-\theta.
$$

An unbiased estimator has an average estimate equal to the parameter. Under
independent Poisson sampling, $E[\hat\lambda]=\lambda$, so the sample mean
is unbiased for $\lambda$. One estimate above or below $\lambda$ is normal
sampling variation; it does not by itself show bias.

The **variance of an estimator** describes how much its estimates vary
across studies. For the Poisson sample mean,

$$
\operatorname{Var}(\hat\lambda)=\frac{\lambda}{n}.
$$

Larger studies therefore have a narrower sampling distribution under this
model. The variance of individual quadrat counts remains $\lambda$;
it is the variation among their *means* that decreases. We can compare the
theoretical properties with a simulation:

```r
# Compare the simulated bias and variance with the Poisson model variance.
c(
  simulated_bias = mean(lambda_estimates) - true_lambda,
  simulated_variance = var(lambda_estimates),
  model_variance = true_lambda / n_quadrats
)
```

The simulated values will differ slightly from the model values because
2,000 repetitions are still a finite number of studies.

## Mean squared error

Bias and spread both matter if we want an estimate near the parameter. The
**mean squared error** (MSE) is the average squared distance between
estimates and the parameter:

$$
\operatorname{MSE}(\hat\theta)
= E[(\hat\theta-\theta)^2]
= \operatorname{Var}(\hat\theta)
  + \operatorname{Bias}(\hat\theta)^2.
$$

For the unbiased Poisson sample mean, its theoretical MSE equals its
variance. Simulated MSE can be calculated with
`mean((lambda_estimates - true_lambda)^2)`; it need not exactly equal
`var(lambda_estimates)` in a finite simulation. The latter also uses
the $n-1$ divisor across simulated estimates.

:::: {.callout-important title="Practice: what varies?"}

Suppose two independent studies each sample 30 quadrats and report means of
3.7 and 4.2 seedlings per quadrat. What varies between the studies? Do
these two estimates establish that the estimator is biased?

::: {.callout-tip title="Solution" collapse="true"}

The samples and their estimates vary. Under the assumed frequentist model,
the parameter $\lambda$ is fixed. Two estimates cannot establish bias,
which concerns the estimator's average over repeated samples.

:::
::::

# Standard errors {#sec-standard_errors}

The **standard error** (SE) is the standard deviation of an estimator's
sampling distribution:

$$
\operatorname{SE}(\hat\theta)
= \sqrt{\operatorname{Var}(\hat\theta)}.
$$

For the Poisson sample mean, the model gives
$\operatorname{SE}(\hat\lambda)=\sqrt{\lambda/n}$. In a real study
$\lambda$ is unknown, so we replace it with our estimate:

$$
\widehat{\operatorname{SE}}(\hat\lambda)
= \sqrt{\frac{\hat\lambda}{n}}.
$$

This is a **plug-in estimate**: an estimated parameter is used in a formula
that calls for the unknown parameter. For the illustrative counts, the
estimated SE is:

```r
# Estimate the Poisson mean and record the number of sampled quadrats.
lambda_hat = mean(seedling_counts)
n_quadrats = length(seedling_counts)
# Use the estimated mean in the Poisson standard-error formula.
estimated_se = sqrt(lambda_hat / n_quadrats)
estimated_se
```

It is approximately 0.63 seedlings per quadrat. This describes typical
variation in the *estimated mean* between studies of ten quadrats, if the
Poisson assumptions are appropriate.

::: {.callout-warning title="Standard deviation and standard error answer different questions"}
The standard deviation of the ten counts describes variation between
quadrats. The estimated standard error describes variation in their *mean*
between repeated studies. More quadrats can reduce the latter without
making seedlings equally abundant in every quadrat.
If counts are overdispersed relative to the Poisson model, its standard
error formula can understate the sampling variability.
:::

:::: {.callout-important title="Practice: change the study size"}

If the true mean is 4 seedlings per quadrat, what happens to the standard
error of the sample mean when a study grows from 25 to 100 independent
quadrats? Does the standard deviation of individual Poisson counts change?

::: {.callout-tip title="Solution" collapse="true"}

The standard error changes from $\sqrt{4/25}=0.4$ to
$\sqrt{4/100}=0.2$ seedlings per quadrat: it halves. Under the same Poisson
model, the standard deviation of an individual count remains
$\sqrt{4}=2$ seedlings per quadrat.

:::
::::

# Confidence intervals {#sec-confidence_intervals}

An estimate is one number, but we also want to communicate its uncertainty.
A **confidence interval** uses an estimator and a rule for its sampling
variation to produce a range of parameter values. Its confidence level
describes the procedure over repeated studies.

If a sampling distribution is sufficiently close to normal, a common
approximate 95% interval is

$$
\hat\theta \pm 1.96\,\widehat{\operatorname{SE}}(\hat\theta).
$$

The number 1.96 is the 97.5th percentile of a standard normal distribution.
Using the illustrative seedling counts and the Poisson standard error gives
a **Wald interval** for $\lambda$:

```r
# Obtain the standard-normal critical value for a two-sided 95% interval.
critical_value = qnorm(p = 0.975)
# Combine the estimate, standard error, and critical value to find the endpoints.
interval = lambda_hat + c(-1, 1) * critical_value * estimated_se
interval
```

The endpoints are approximately 2.76 and 5.24 seedlings per quadrat.
They are possible values for the *mean count*, not limits on individual
quadrat counts. The interval calculation relies on the Poisson model,
independence, and a normal approximation for the estimator.

::: {.callout-note title="What 95% confidence means"}
If we repeatedly sampled quadrats and constructed an interval by the same
procedure, about 95% of the intervals would include the fixed true
parameter **when the procedure has 95% coverage under the assumed model**.
The endpoints change between studies; the parameter does not.
:::

::: {.callout-warning title="Check the interval procedure"}
A 95% label does not guarantee 95% coverage in every setting. A normal
approximation to Poisson counts can work poorly for small expected totals.
A Wald interval can even have a negative lower endpoint, although a
Poisson mean cannot be negative. An interval from one study does not give
a 95% probability that the fixed parameter lies between its endpoints.
:::

## Checking coverage by simulation

Because a simulation gives us the true $\lambda$, we can check whether each
interval includes it. The proportion that do is the simulated
**coverage**. The following code repeats both sampling and interval
construction:

```r
# Set the model parameters and interval rule for repeated simulated studies.
set.seed(seed = 402)
true_lambda = 4
n_quadrats = 30
n_studies = 2000
critical_value = qnorm(p = 0.975)

# Simulate each study, construct its interval, and record whether it covers the truth.
covered = replicate(
  n = n_studies,
  expr = {
    # Generate one study and calculate its estimated mean and standard error.
    counts = rpois(n = n_quadrats, lambda = true_lambda)
    estimate = mean(counts)
    se = sqrt(estimate / n_quadrats)
    # Construct the interval and test whether it includes the true model mean.
    lower = estimate - critical_value * se
    upper = estimate + critical_value * se
    lower <= true_lambda && true_lambda <= upper
  }
)

# Calculate the simulated coverage across all studies.
mean(covered)
```

Simulated coverage changes with the seed and number of repetitions. It can
also change with sample size, the true parameter, and the interval method.
We cannot calculate coverage from a single field study because its true
parameter is unknown.

:::: {.callout-important title="Practice: interpret an interval"}

A study reports a 95% confidence interval of 2.8 to 5.2 seedlings per
quadrat. What does the 95% refer to? What would need to be checked before
trusting this interval for an ecological conclusion?

::: {.callout-tip title="Solution" collapse="true"}

The 95% refers to the long-run coverage of the interval-making procedure,
not a probability assigned to the fixed parameter after seeing this one
interval. We would examine whether the Poisson model and independence are
reasonable and whether the approximation used for the interval is adequate.

:::
::::

# Estimating two tree-height parameters {#sec-two_parameters}

The same ideas apply to continuous measurements. In
[Chapter 3](../Chapter_3/Theory.qmd#sec-normal_distribution), we used a
normal model for tree height. Suppose $X_i$ is the height in metres of
tree $i$ and independent trees follow

$$
X_i \sim \operatorname{Normal}(\mu,\sigma^2).
$$

The model mean is $\mu$ metres and its variance is $\sigma^2$ square
metres. R's `rnorm()` takes the standard deviation $\sigma$ in its
`sd` argument. This model is an approximation: a normal distribution
allows negative heights, so its tails must be plausible for the ecological
setting.

## Method of moments for mean and variance

The first model moment is $E[X]=\mu$, so the method-of-moments estimator of
the mean is $\hat\mu=\overline X$. A second relationship is
$E[X^2]=\mu^2+\sigma^2$. Matching it to the sample mean of the squared
heights gives

$$
\widehat{\sigma^2}_{\mathrm{MM}}
= \frac{1}{n}\sum_{i=1}^{n}(X_i-\overline X)^2.
$$

The method gives a rule for both parameters. Students do not need the
algebraic derivation to use it: take the mean height, then average the
squared deviations from that mean.

```r
# Simulate tree heights from an assumed normal population.
set.seed(seed = 403)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)

# Calculate the mean and compare the two variance divisors.
height_mean = mean(tree_heights)
variance_mm = mean((tree_heights - height_mean)^2)
variance_with_var = var(tree_heights)

c(
  mean_m = height_mean,
  variance_mm_m2 = variance_mm,
  variance_with_var_m2 = variance_with_var
)
```

The two variance estimates differ. `variance_mm` divides by $n$;
`var()` divides by $n-1$. Under independent sampling from the same
distribution with finite variance,

$$
E[\widehat{\sigma^2}_{\mathrm{MM}}]
= \frac{n-1}{n}\sigma^2,
\qquad
E[s^2]=\sigma^2,
$$

where $s^2$ denotes the result of `var()`. The method-of-moments
variance estimator is therefore biased downward by $\sigma^2/n$.
It is still a useful estimator, and its bias becomes smaller as the
sample size grows. Bias alone does not tell us which estimator has
smaller MSE.

::: {.callout-note title="Why the divisor matters"}
The empirical variance introduced in Chapter 2 uses the divisor $n$.
R's `var()` uses $n-1$ to estimate the variance of the population or
model from which the sample came. Name the quantity you are calculating
before choosing the divisor.
:::

## A confidence interval for mean height

The standard error of the sample mean is $\sigma/\sqrt n$. With
$\sigma$ unknown, we estimate it by $s/\sqrt n$, where
$s=\texttt{sd(tree_heights)}$. For independent observations from a normal
model, a 95% interval for $\mu$ uses a $t$ quantile:

$$
\overline X \pm t_{n-1,0.975}\frac{s}{\sqrt n}.
$$

The $t$ distribution allows for estimating $\sigma$ from the same sample.
Under the stated normal model, this procedure has 95% coverage. Applied
to the simulated heights:

```r
# Record the sample size and calculate the standard error of the mean height.
n_trees = length(tree_heights)
height_se = sd(tree_heights) / sqrt(n_trees)
# Use the appropriate t critical value to construct the confidence interval.
t_critical = qt(p = 0.975, df = n_trees - 1)
height_interval = height_mean + c(-1, 1) * t_critical * height_se
height_interval
```

The interval is for the population **mean height in metres**, not the
heights of 95% of individual trees. With skewed measurements, such as
body mass, the same $t$ interval for the original-scale mean is only
approximate; we will investigate that in the practical.

:::: {.callout-important title="Practice: two kinds of spread"}

Which R expression estimates variation among the 25 tree heights, and
which estimates the standard error of their mean? Why does an interval
for the mean use the second quantity?

::: {.callout-tip title="Solution" collapse="true"}

`sd(tree_heights)` estimates variation among trees.
`sd(tree_heights) / sqrt(length(tree_heights))` estimates the
standard error of the sample mean. The interval describes uncertainty
in the mean height, so it uses the sampling variation of the mean.

:::
::::

# What larger samples change {#sec-large_samples}

We saw in [Chapter 2](../Chapter_2/Theory.qmd#sec-sample_population)
that empirical summaries can approach their model counterparts as sample
size increases. For an estimator, **consistency** means that the
probability of being far from its target parameter tends to zero as the
sample size grows. Under the independent Poisson model, the sample mean
is consistent for $\lambda$.

Its standard error is $\sqrt{\lambda/n}$, so multiplying the sample size
by four halves the standard error. This is a statement about the
*distribution across possible studies*. It does not promise that every
larger realised sample is closer to the truth.

Many sample means also have an **approximately normal sampling
distribution** when sample size is sufficiently large, even if individual
observations are not normal. This large-sample behaviour is called
**asymptotic normality**. For the Poisson mean,

$$
\hat\lambda \overset{\mathrm{approx}}{\sim}
\operatorname{Normal}\left(\lambda,\frac{\lambda}{n}\right)
\quad\text{for sufficiently large }n.
$$

The approximation depends on the setting; a small expected total count
$n\lambda$ can leave the sampling distribution discrete and asymmetric.
This is why checking an interval's coverage is useful.

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Sampling distributions of mean seedling counts for studies of 5 and 100 quadrats. Curves show normal approximations under the assumed Poisson model."
# Set the Poisson model and the study sizes to compare.
set.seed(seed = 404)
true_lambda = 4
sample_sizes = c(5, 100)
# Save the current graphics settings before drawing two panels.
old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))
for (sample_size in sample_sizes) {
  # Simulate repeated studies and calculate one estimated mean per study.
  estimates = replicate(
    n = 2000,
    expr = mean(rpois(n = sample_size, lambda = true_lambda))
  )
  # Use the histogram density to choose a vertical scale for the plot.
  histogram = hist(x = estimates, breaks = 25, plot = FALSE)
  # Plot the sampling distribution and overlay its normal approximation.
  hist(
    x = estimates, breaks = 25, probability = TRUE,
    col = "grey85", border = "white",
    ylim = c(0, max(histogram$density) * 1.25),
    xlab = "Estimated seedlings per quadrat",
    ylab = "Density", main = paste("Quadrats:", sample_size)
  )
  curve(
    expr = dnorm(
      x = x, mean = true_lambda,
      sd = sqrt(true_lambda / sample_size)
    ),
    add = TRUE, lwd = 2, lty = 2
  )
  legend(
    "topleft", legend = "Normal approximation",
    lty = 2, lwd = 2, bty = "n", cex = 0.8
  )
}
# Restore the graphics settings used before this figure.
par(old_par)
```

The larger-sample estimates are more concentrated around 4, and the
normal curve describes their shape more closely. The individual
quadrat counts are still Poisson counts.

:::: {.callout-important title="Practice: bigger is not guaranteed closer"}

A study of 20 quadrats estimates 3.9 seedlings per quadrat; another
study of 100 estimates 4.2. The true model mean is 4. Did the larger
study disprove consistency?

::: {.callout-tip title="Solution" collapse="true"}

No. The smaller study happened to give an estimate nearer 4. Consistency
concerns how the *probability* of large errors changes as sample size
increases, not the ordering of two realised estimates.

:::
::::

# What simulation can tell us {#sec-simulation_limits}

Simulation lets us specify a model and known parameters, generate samples,
apply an estimator to each sample, and compare its repeated results with
the truth. That is how we studied bias, standard errors, MSE, and interval
coverage. More repetitions usually make a simulated summary more stable;
they do not make an unrealistic ecological model realistic.

Our Poisson studies assumed independent quadrats of equal area with the
same mean count. Spatial clustering, varying habitat, or imperfect
detection could change both the count distribution and the performance
of our estimates and intervals. These are reasons to examine the
sampling design and observations before drawing ecological conclusions.

This chapter has treated parameters as fixed and samples as variable.
In [Chapter 5](../Chapter_5/Theory.qmd), we will use the probability
model more fully through **maximum likelihood estimation**. The same
questions remain: What does the parameter mean, what estimate did we
get, and how uncertain is it?

# Summary

Method of moments matches sample summaries to model moments. An
estimator is a rule, while an estimate is its result for one sample.
Repeated samples show an estimator's bias and spread; a standard error
describes the latter in the parameter's units. A confidence level
describes an interval procedure's coverage across repeated studies.
Simulation can check these properties under specified assumptions,
which must still be judged against the ecological setting.

# Chapter quiz

Try each question before opening its answer.

:::: {.callout-important title="Question 1: an estimate"}

The mean of counts from 20 quadrats is 5.1 seedlings per quadrat.
Under a Poisson model, name the parameter, estimator, and estimate.

::: {.callout-tip title="Answer" collapse="true"}

The parameter is the unknown model mean $\lambda$. The estimator is
$\hat\lambda=\overline X$. Its estimate for these counts is 5.1
seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 2: what does the histogram show?"}

A histogram contains one sample mean from each of 1,000 simulated studies.
Is it an empirical distribution of individual observations or an
approximation to a sampling distribution?

::: {.callout-tip title="Answer" collapse="true"}

It approximates the sampling distribution of the sample mean. Each
histogram entry is one estimate from a whole study.

:::
::::

:::: {.callout-important title="Question 3: bias and spread"}

An estimator's average over repeated studies equals its target, but
individual estimates vary considerably. Is it biased? Is its variance zero?

::: {.callout-tip title="Answer" collapse="true"}

It is unbiased because its repeated-sampling mean equals the target.
Its variance is greater than zero because its estimates vary.

:::
::::

:::: {.callout-important title="Question 4: standard error"}

For 100 independent quadrats under a Poisson model with $\lambda=4$,
what is the standard error of the estimated mean count? What would
the standard deviation of one quadrat count be?

::: {.callout-tip title="Answer" collapse="true"}

The standard error is $\sqrt{4/100}=0.2$ seedlings per quadrat. The
standard deviation of one count is $\sqrt{4}=2$ seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 5: interval meaning"}

Does a calculated 95% confidence interval assign 95% probability to
the fixed parameter lying inside it? Explain what 95% describes.

::: {.callout-tip title="Answer" collapse="true"}

No. Under the assumed model, 95% describes the proportion of intervals
that the procedure would expect to cover the fixed parameter over
repeated studies, if the procedure achieves its nominal coverage.

:::
::::

:::: {.callout-important title="Question 6: estimating variation"}

Which divisor does the method-of-moments variance estimator use, and
which does R's `var()` use? What happens to the former's bias as
sample size grows?

::: {.callout-tip title="Answer" collapse="true"}

The method-of-moments estimate divides by $n$; `var()` divides by
$n-1$. Under the stated independent sampling assumptions, the
method-of-moments estimator's bias is $-\sigma^2/n$, which approaches
zero as $n$ grows.

:::
::::
