---
title: "Maximum likelihood estimation"
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures in this chapter.
library(ggplot2)
```

# Introduction

In this chapter, we learn how to estimate the parameters of a model and their uncertainty
by maximum likelihood. In [Chapter 3](../Chapter_3/Theory.qmd) we estimated parameters with
the method of moments, which only uses the mean and the variance of a sample. Maximum likelihood uses the
probability that the model assigns to every observation, and it applies to any model for
which we can write a probability mass or density function. This is why it is the method of
estimation in the rest of the course. The concepts of Chapter 3 (estimator, sampling
distribution, standard error and confidence interval) remain the same; we now apply them to
a different estimator.

The chapter uses two examples. The first is the counts of balsam fir seedlings in 50
quadrats of Chapter 3, under a Poisson model, so that we can compare every result with
Chapter 3. Where the shape of the likelihood matters we also use the first ten of those
quadrats, as a smaller sample from the same data. The second is the number of calls made
by barn owl nestlings, the field data of the Chapter 3 practical (the data set `Owls` of
the package `glmmTMB`), under a Negative Binomial model with two parameters. The chapter
ends with the use of information criteria to compare different models fitted to the same data.

The chapter has two parts. The first part builds the method: the likelihood function,
maximum likelihood estimation with `optim()` and the package RTMB, and models with
several parameters. The
second part is about what we can say once a model is fitted: the uncertainty of the
estimates, the properties of maximum likelihood estimators and the comparison of models.

# Learning goals

In this chapter, we will cover:

- the likelihood function and the log-likelihood;
- maximum likelihood estimation by numerical optimisation;
- constraints on parameters and transformations;
- likelihood surfaces, slices and profiles for models with several parameters;
- Fisher information, standard errors and Wald intervals;
- profile likelihood intervals and joint confidence regions;
- the properties of maximum likelihood estimators in large and small samples;
- the comparison of models with the Akaike information criterion.

We will use the [Poisson](../Supplements/distributions.qmd#sec-poisson) and
[Negative Binomial](../Supplements/distributions.qmd#sec-negative_binomial) distributions
throughout the chapter. They are described in the
[Supplement](../Supplements/distributions.qmd). We write and fit the models with the
packages RTMB and RTMBdist, whose distribution functions we have used since Chapter 2.

The table below is a roadmap of the chapter. Each row is one step, with the question that
it answers, the R tools that we use for it and the example on which we show it.

| Step | Question it answers | In R | Example |
|---|---|---|---|
| **Part 1** | | | |
| The likelihood | How probable are the data for a given value of the parameter? | `dpois(..., log = TRUE)`, `sum()` | Fir seedlings |
| Estimation | Which value of the parameter makes the data most probable? | NLL with `getAll()`, `cmb()`, `MakeADFun()`, `optim()` | Fir seedlings |
| Transformations | How do we keep a parameter within its range? | `exp()`, `plogis()` | Fir seedlings |
| Several parameters | What does the likelihood look like, and what does it say about one parameter? | `dnbinom2()`, `for` loops with `obj$fn()`, `contour()`, profiles | Owls |
| **Part 2** | | | |
| Standard errors | How precise is the estimate? | The Hessian by hand, then `sdreport()` and `ADREPORT()` | Fir seedlings and owls |
| Confidence intervals | Which parameter values do the data support? | Wald interval on the fitted scale; profile with `TMB::tmbprofile()` and `confint()` | Fir seedlings and owls |
| Properties | Why does the method work, and when does it fail? | Simulation with `obj$simulate()` | Owls |
| Model comparison | Which of two models is better supported by the data? | AIC | Owls |

: Roadmap of the chapter. NLL is the negative log-likelihood, which is defined in @sec-maximum-likelihood.

# The likelihood function {#sec-likelihood}

The **likelihood function** is the joint probability of the observed sample, read as a
function of the parameters of the model with the data held fixed. For continuous data it is
the joint density of the sample. Let $x_1,\ldots,x_n$ be the observations and assume that
they are i.i.d. with a distribution that has a parameter $\theta$ (as in Chapter 3,
$\theta$ stands for a parameter in general). The likelihood of $\theta$ is

$$
L(\theta\mid x_1,\ldots,x_n)
= P(X_1=x_1,\ldots,X_n=x_n\mid\theta)
= \prod_{i=1}^{n}P(X_i=x_i\mid\theta),
$$

where $P(X_i=x_i\mid\theta)$ is the probability mass that the model assigns to the
observation $x_i$ when the parameter is $\theta$,^[The vertical bar reads "for a given
value of": $P(X=x\mid\theta)$ is the probability of $x$ for a given value of $\theta$, and
$L(\theta\mid x_1,\ldots,x_n)$ is the likelihood of $\theta$ for a given sample. The same
symbol is used for conditional distributions, which are introduced in
[Chapter 7](../Chapter_7/Theory.qmd).] and the symbol $\prod$ denotes a product over the
$n$ observations, in the same way as $\sum$ denotes a sum. The second equality holds
because the observations are independent, so their joint probability is the product of
their individual probabilities (see
[Chapter 2](../Chapter_2/Theory.qmd#sec-joint-independence), where we announced that the
i.i.d. assumption would be fundamental for estimation). For continuous data we replace each
probability mass by the density $f_X(x_i\mid\theta)$.

This is the formula of a joint distribution from Chapter 2. What changes is what we hold
fixed. In Chapter 2 we fixed the parameters and calculated the probability of different
possible samples. Now the sample has been observed, so $x_1,\ldots,x_n$ are fixed numbers,
and we look at how the probability of that sample changes as we vary $\theta$. A parameter
value under which the observed sample is more probable is better supported by the data,
that is the crucial idea behind likelihood.^[Fisher introduced the maximum likelihood method in 1912, in a
paper that he wrote as an undergraduate student and in which he called it an "absolute
criterion". He introduced the word likelihood in 1921, to keep this quantity apart from a
probability, and the name "maximum likelihood" in 1922, in the paper that set out the
theory of the method. Choosing the parameter value under which the observations are most
probable had been anticipated in the eighteenth and early nineteenth centuries by Lambert,
Daniel Bernoulli and Gauss, although without a reasoned defence of the method. If you want
to read more about this history, see Aldrich (1997), "R. A. Fisher and the making of
maximum likelihood 1912–1922", *Statistical Science* 12: 162–176, and Stigler (2007), "The
epic story of maximum likelihood", *Statistical Science* 22: 598–620.] The likelihood is
the basis of everything that follows in this book: the estimates, their confidence
intervals and the comparison of models are all calculated from it.

Let's illustrate it with the counts of balsam fir seedlings in 50 quadrats of
[Chapter 3](../Chapter_3/Theory.qmd#sec-estimates), where $X_i$ is the number of seedlings
in quadrat $i$ and $X_i \sim \operatorname{Poisson}(\lambda)$. The probability mass
function of the Poisson distribution is
$P(X_i=x_i\mid\lambda)=e^{-\lambda}\lambda^{x_i}/x_i!$. We start with a case that is small
enough to do by hand, the counts of the first three quadrats (0, 1 and 2). Their
likelihood is the product of three contributions (remember that $0!=1$):

$$
L(\lambda\mid 0,1,2)
= \frac{e^{-\lambda}\lambda^{0}}{0!}\times
  \frac{e^{-\lambda}\lambda^{1}}{1!}\times
  \frac{e^{-\lambda}\lambda^{2}}{2!}
= \frac{e^{-3\lambda}\lambda^{3}}{2}.
$$

For $\lambda=1$ this gives $e^{-3}/2\approx 0.0249$ and for $\lambda=2$ it gives
$e^{-6}\,2^{3}/2\approx 0.0099$.

To verify our maths with R, we first load the packages RTMB and RTMBdist and the counts of
the 50 quadrats, as in Chapter 3:

```{r}
#| message: false
# Load RTMB first and RTMBdist second.
library(RTMB)
library(RTMBdist)

# Counts of balsam fir seedlings in 50 quadrats (see ?boot::fir), as in Chapter 3,
# and the number of quadrats.
data(fir, package = "boot")
seedling_counts = fir$count
n_quadrats = length(seedling_counts)
```

`dpois()` returns the contribution of each of the three counts for a given value of
$\lambda$, and `prod()` multiplies them:

```{r}
# The counts of the first three quadrats.
first_counts = seedling_counts[1:3]
first_counts

# Probability of each of the three counts at lambda = 1.
dpois(x = first_counts, lambda = 1)

# Likelihood of the three counts: prod() multiplies their probabilities.
prod(dpois(x = first_counts, lambda = 1))
prod(dpois(x = first_counts, lambda = 2))
```

The two products are the values that we calculated by hand. These three counts are about 2.5
times more probable under a mean of 1 than under a mean of 2 (their own mean is 1). The
likelihood of the whole sample is calculated in the same way:

```{r}
# Likelihood of all 50 counts at lambda = 1 and at lambda = 2.
prod(dpois(x = seedling_counts, lambda = 1))
prod(dpois(x = seedling_counts, lambda = 2))
```

With the 50 counts the order is reversed: the likelihood is approximately
$4.2\times10^{-50}$ for $\lambda=1$ and $1.3\times10^{-39}$ for $\lambda=2$. That is, the
whole sample is about $3\times10^{10}$ times more probable under a mean of 2 than under a
mean of 1, so 50 quadrats discriminate between these two values far more strongly than
three.
Note how small both numbers are. Every observation multiplies the likelihood by a
probability smaller than one, so **its value depends on the number of observations and has
no meaning by itself**. What we use is the comparison between the likelihoods of different
parameter values for the same data.

A likelihood is calculated from probabilities but it is not a probability distribution for
the parameter, for two reasons:

1. We work with aleatoric probability ([Chapter 2](../Chapter_2/Theory.qmd)): probabilities describe how
the data or estimates vary across replications of the experiment, while the parameter of the model is a
fixed quantity^[The likelihood is also at the core of Bayesian inference. There it is multiplied
by a prior distribution, which describes what is known about the parameter before seeing
the data, and the product is rescaled to give a probability distribution for the parameter
(the posterior distribution). [Chapter 1](../Chapter_1/Theory.qmd) explains why this book
stays with the Frequentist interpretation.]. So in the model there is no variation or error
about $\lambda$ to which a probability
could refer.

2. For a fixed $\lambda$ the probabilities of all the
possible samples add up to one, but nothing forces the likelihoods of all the possible
values of $\lambda$ to do so for a fixed sample. For the 50 counts, the area under the
likelihood across all the possible values of $\lambda$ is approximately
$8.7\times10^{-40}$, which is very far from one.

```{r}
#| include: false
# Closed-form Poisson likelihood of a sample of counts. Unlike a product of dpois() over
# the data, this function accepts a vector of lambda values, which integrate() needs.
# It is calculated on the log scale and then exponentiated, because lambda^sum(counts)
# is too large for the computer at the large values of lambda that integrate() tries.
poisson_likelihood = function(lambda, counts) {
  exp(-length(counts) * lambda + sum(counts) * log(lambda) - sum(lfactorial(counts)))
}

# Check that it gives the same likelihood as the product of probabilities.
poisson_likelihood(lambda = 2, counts = seedling_counts)
prod(dpois(x = seedling_counts, lambda = 2))

# Area under the likelihood across all possible values of lambda. The likelihood is
# tiny, so the default absolute tolerance is too coarse: we set it to zero and ask
# for a small relative error instead.
integrate(
  f = poisson_likelihood,
  lower = 0,
  upper = Inf,
  counts = seedling_counts,
  rel.tol = 1e-8,
  abs.tol = 0
)

# The integral has a closed form, which confirms the numerical result:
# (sum of counts)! / n^(sum of counts + 1) / product of the factorials of the counts.
exp(
  lgamma(sum(seedling_counts) + 1) -
    (sum(seedling_counts) + 1) * log(n_quadrats) -
    sum(lfactorial(seedling_counts))
)
```

::: {.callout-warning title="A likelihood is not the probability of a parameter value"}
The likelihood $L(\theta\mid x_1,\ldots,x_n)$ is the probability (or density) of the
observed data for a given value of the parameter, read as a function of the parameter. It
does not add up or integrate to one over the values of the parameter, so it does not tell
us how probable a parameter value is. We use it to compare parameter values: the data are
better supported by the value with the higher likelihood.
:::

## The log-likelihood

The likelihood of 50 counts is already of the order of $10^{-39}$. With a larger sample the
product can become smaller than the smallest positive number that the computer can
represent, and R then returns zero. For this reason, and because sums are easier to work
with than products, we use the **log-likelihood**, the natural logarithm of the likelihood:

$$
\ell(\theta\mid x_1,\ldots,x_n)
= \log L(\theta\mid x_1,\ldots,x_n)
= \sum_{i=1}^{n}\log P(X_i=x_i\mid\theta).
$$

The logarithm turns the product into a sum. It is an increasing function, so a parameter
value with a higher likelihood also has a higher log-likelihood, and both functions have
their maximum at the same value of $\theta$. Once the data are fixed we will write
$L(\theta)$ and $\ell(\theta)$ for short.

In R, all the `d*` functions have an argument `log`. With `log = TRUE` they return the
logarithm of the probability mass (or density), which is more accurate than calculating
the probability first and taking its logarithm afterwards. The log-likelihood is the sum
of these values:

```{r}
# Log-likelihood of the 50 counts at lambda = 2: sum the log probabilities,
# using the argument log = TRUE of dpois().
sum(dpois(x = seedling_counts, lambda = 2, log = TRUE))

# The same number, from the logarithm of the product.
log(prod(dpois(x = seedling_counts, lambda = 2)))
```

Both calculations give the same value. With 50 counts the product still works.
To see when it fails, we simulate 1,000 counts from a Poisson model with $\lambda=4$:

```{r}
# Likelihood and log-likelihood of a large simulated sample of 1,000 counts at lambda = 4.
set.seed(seed = 501)
many_counts = rpois(n = 1000, lambda = 4)

# The product is smaller than the smallest positive number R can represent.
prod(dpois(x = many_counts, lambda = 4))

# The sum of log probabilities is an ordinary number.
sum(dpois(x = many_counts, lambda = 4, log = TRUE))
```

R returns zero for the product, although none of the probabilities in it is zero, whereas
the log-likelihood is an ordinary number. From here on we always work with
log-likelihoods.

For continuous data the contributions to the likelihood are densities. Recall from
[Chapter 2](../Chapter_2/Theory.qmd#sec-continuous_measurements) that a density has units
(one over the unit of the measurement), that it changes with those units and that it can
be larger than one. The log-likelihood inherits these properties: **it can be positive**, and
**its value changes if we express the same measurements in other units**. For example, if body
masses are converted from grams to kilograms, the density of every mass is multiplied by
1,000 (the same probability is spread over a thousand times fewer units), so the
log-likelihood of $n$ masses increases by $n\log(1000)$. This shift is a constant, so the
differences between the log-likelihoods of two parameter values, or of two models, are the
same in both units. In general, the value of a log-likelihood means nothing by itself, and
only differences between log-likelihoods calculated from the same data are meaningful. You
will check this with measurements of body mass in the practical.

:::: {.callout-important title="Practice: crab burrows"}

Three plots on a mudflat contain 3, 1 and 2 crab burrows. Assume that the counts are i.i.d.
with a Poisson distribution. On paper, write the likelihood as a function of $\lambda$ and
calculate the likelihood and the log-likelihood for $\lambda=2$ and for $\lambda=3$ burrows
per plot. Which of the two values is better supported by the data?

::: {.callout-tip title="Solution" collapse="true"}

The likelihood is the product of the three contributions:

$$
L(\lambda\mid 3,1,2)
= \frac{e^{-\lambda}\lambda^{3}}{3!}\times
  \frac{e^{-\lambda}\lambda^{1}}{1!}\times
  \frac{e^{-\lambda}\lambda^{2}}{2!}
= \frac{e^{-3\lambda}\lambda^{6}}{12}.
$$

For $\lambda=2$ the likelihood is $64e^{-6}/12\approx 0.0132$ and the log-likelihood is
$-6+\log(64/12)\approx -4.33$. For $\lambda=3$ the likelihood is $729e^{-9}/12\approx 0.0075$
and the log-likelihood is $-9+\log(729/12)\approx -4.89$. The data are better supported by
$\lambda=2$, which is the mean of the three counts. The difference between the two
log-likelihoods is 0.57, so the counts are about 1.8 times more probable ($e^{0.57}$)
under $\lambda=2$ than under $\lambda=3$. In R:

```{r}
# Log-likelihood of the three counts at lambda = 2 and at lambda = 3.
burrow_counts = c(3, 1, 2)
sum(dpois(x = burrow_counts, lambda = 2, log = TRUE))
sum(dpois(x = burrow_counts, lambda = 3, log = TRUE))
```

:::
::::

# Maximum likelihood estimation {#sec-maximum-likelihood}

The **maximum likelihood estimator** (MLE) of a parameter is the value of the parameter for
which the likelihood of the observed sample is highest:

$$
\hat\theta
= \underset{\theta}{\arg\max}\;L(\theta\mid x_1,\ldots,x_n)
= \underset{\theta}{\arg\max}\;\ell(\theta\mid x_1,\ldots,x_n),
$$

where $\arg\max$ reads "the value of $\theta$ at which the function reaches its maximum".
That is, $\max$ would give the height of the highest point and $\arg\max$ gives its
location. In plain English, the MLE is the parameter value under which the model assigns
the highest probability to the data that we observed. As we saw in [Chapter 3](../Chapter_3/Theory.qmd#sec-estimates),
the procedure of calculation is called estimator and the value that it returns for a given
sample is the estimate; we use the abbreviation MLE for both.

For a few models the MLE can be written as a formula, and we start with one of them, the
Poisson model. In general there is no such formula, and we find the maximum numerically,
with an algorithm that searches for it. That is the method that we use in the rest of the
book, because it works for any model for which we can write a likelihood.

At the highest point of the log-likelihood its slope is zero (if you need to refresh how
a derivative is calculated, see the
[rules of differentiation](../Supplements/calculus.qmd#sec-calc-rules) of the calculus
refresher). For i.i.d. Poisson counts,
setting the derivative of the log-likelihood with respect to $\lambda$ to zero and solving
for $\lambda$ gives

$$
\hat\lambda=\frac{1}{n}\sum_{i=1}^{n}x_i=\overline x.
$$

So the MLE of $\lambda$ is the mean of the sample, which is also the method-of-moments
estimator of [Chapter 3](../Chapter_3/Theory.qmd#sec-estimates). For the 50 counts:

```{r}
# The maximum likelihood estimate of the Poisson mean is the mean of the counts.
lambda_hat = mean(seedling_counts)
lambda_hat
```

The estimate is 2.14 seedlings per quadrat, as in Chapter 3. A solution that can be written as a formula, like
this one, is called a **closed-form** solution, and it exists only for a few
models.^[The Normal model is another of them. The MLE of $\mu$ is
the mean of the sample and the MLE of $\sigma^2$ is the variance with divisor $n$, that is,
the method-of-moments estimators of
[Chapter 3](../Chapter_3/Theory.qmd#sec-two_parameters). The MLE of the variance is
therefore biased.] For the others, setting the derivatives to zero gives equations that
cannot be solved with algebra.

::: {.callout-caution collapse="true" icon=false title="Statistical detail: deriving the Poisson estimator"}
At the maximum of a smooth curve
the slope is zero, so we differentiate the log-likelihood with respect to the parameter,
set the derivative to zero and solve for the parameter. For i.i.d. Poisson counts:

$$
\begin{aligned}
\ell(\lambda)
&= \sum_{i=1}^{n}\log\left(\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}\right)
= -n\lambda+\log(\lambda)\sum_{i=1}^{n}x_i-\sum_{i=1}^{n}\log(x_i!),\\
\frac{d\ell}{d\lambda}
&= -n+\frac{1}{\lambda}\sum_{i=1}^{n}x_i,\\
0 &= -n+\frac{1}{\hat\lambda}\sum_{i=1}^{n}x_i
\quad\Longrightarrow\quad
\hat\lambda=\frac{1}{n}\sum_{i=1}^{n}x_i=\overline x.
\end{aligned}
$$

The last term of the log-likelihood does not contain $\lambda$, so it disappears when we
differentiate.^[The derivative of the log-likelihood with respect to a parameter is called
the **score**. The MLE is the value of the parameter(s) at which the score is zero.] The second
derivative, $-\sum x_i/\lambda^2$, is negative, which confirms that the point where the
slope is zero is a maximum (derivatives, and what they tell us about the shape of a
function, return in [Chapter 5](../Chapter_5/Theory.qmd)). In R:

```{r}
# The derivative of the log-likelihood, sum(x) / lambda - n, is zero at that value.
sum(seedling_counts) / lambda_hat - n_quadrats
```
:::

## The likelihood curve

A **likelihood curve** is the log-likelihood of a model with one parameter, drawn as a
function of that parameter. It shows where the maximum is and how quickly the
log-likelihood falls on either side of it, which is what we will use in
@sec-likelihood-uncertainty to calculate confidence intervals. To draw it, we calculate
the log-likelihood for many values of the parameter. We plot the result
as a **relative log-likelihood**, which is the difference between the log-likelihood and
its maximum,

$$
\ell(\theta)-\ell(\hat\theta),
$$

so that the curve is zero at the estimate and negative everywhere else. We draw all the
likelihood figures of this chapter on this scale, because only differences in
log-likelihood are meaningful.

We draw the curve for a smaller sample from the same data, the first ten quadrats, because
the shape of a likelihood curve is easier to see when the sample is small (in
@sec-likelihood-uncertainty we compare what ten and 50 quadrats say about $\lambda$). From
here on we say each time which of the two samples a result refers to. The ten quadrats
contain 22 seedlings, so their estimate is 2.2 seedlings per quadrat:

```{r}
# The counts of the first ten quadrats and their maximum likelihood estimate.
ten_counts = seedling_counts[1:10]
ten_counts
lambda_hat_ten = mean(ten_counts)
lambda_hat_ten
```

We define a grid of values of $\lambda$ and fill a vector with the log-likelihood at each
of them, using a `for` loop:

```{r}
# Candidate values of lambda (seedlings per quadrat).
lambda_grid = seq(from = 0.8, to = 5, by = 0.05)

# Pre-allocate a vector for the log-likelihood and fill it with a loop over the grid.
loglik_grid = rep(NA, times = length(lambda_grid))
for (i in seq_along(lambda_grid)) {
  loglik_grid[i] = sum(dpois(x = ten_counts, lambda = lambda_grid[i], log = TRUE))
}
```

We then subtract the log-likelihood at the estimate from every value of the grid. We also
calculate the relative log-likelihood at two values of $\lambda$ that are one seedling
away from the estimate, on either side of it:

```{r}
# Relative log-likelihood: the difference from the log-likelihood at the estimate.
loglik_lambda_hat_ten = sum(dpois(x = ten_counts, lambda = lambda_hat_ten, log = TRUE))
relative_loglik = loglik_grid - loglik_lambda_hat_ten

# Relative log-likelihood at 1.2 and at 3.2 seedlings per quadrat.
sum(dpois(x = ten_counts, lambda = 1.2, log = TRUE)) - loglik_lambda_hat_ten
sum(dpois(x = ten_counts, lambda = 3.2, log = TRUE)) - loglik_lambda_hat_ten
```

The curve reaches its maximum at $\hat\lambda=2.2$ seedlings per quadrat
(@fig-likelihood_curve). Note that the curve is not symmetric: it falls more steeply to
the left of the maximum than to the right. The relative log-likelihood is $-3.33$ at
$\lambda=1.2$ and $-1.76$ at $\lambda=3.2$, so a mean of 3.2 is better supported by these
data than a mean of 1.2, although both are one seedling away from the estimate. We come
back to this asymmetry in @sec-likelihood-uncertainty.

```{r}
#| echo: false
#| label: fig-likelihood_curve
#| fig-cap: "Relative log-likelihood of the counts of fir seedlings in the first ten quadrats under a Poisson model, as a function of the mean number of seedlings per quadrat. The vertical line marks the maximum."
# Plot the relative log-likelihood curve and mark its maximum.
curve_data = data.frame(lambda = lambda_grid, relative = relative_loglik)
ggplot(curve_data, aes(x = lambda, y = relative)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_vline(
    xintercept = lambda_hat_ten,
    linetype = "dashed",
    colour = "grey40"
  ) +
  coord_cartesian(ylim = c(-8, 0)) +
  labs(
    x = expression(paste("Mean seedlings per quadrat, ", lambda)),
    y = "Relative log-likelihood"
  ) +
  theme_classic()
```

A grid of values shows us the curve. To find its highest point for any model, with any
number of parameters, we use an optimiser.

## Numerical optimisation

An **optimiser** is an algorithm that searches for the parameter values at which a function
reaches its minimum. It starts from values that we provide, evaluates the function, and
moves step by step towards lower values until it cannot improve any further. In R we use
the function `optim()`. Since it minimises, we do not give it the log-likelihood but the
**negative log-likelihood** (NLL), $-\ell(\theta)$: the parameter value with the lowest NLL
is the one with the highest likelihood.

In this book every NLL is written in the same form, which the package RTMB needs. It is a
function with two arguments, both of them named lists: the first holds the parameters
(`parms`) and the second the data (`data`). Its first line calls `getAll()`, which makes
the elements of both lists available by their names, so that the rest of the function
reads like the model. For the Poisson model the list of parameters has one element,
`lambda`, and the list of data has one element, `counts`:

```{r}
# Negative log-likelihood of Poisson counts. The first argument holds the parameters
# and the second the data, both as named lists.
poisson_nll_lambda = function(parms, data) {
  # Make the elements of both lists available by name.
  getAll(parms, data)
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}
```

Three steps take us from this function to the estimate. First, we connect the likelihood
to a data set with a small helper function, `cmb()`, which takes an NLL and a list of data
and returns the NLL of those data as a function of the parameters only. This way we can
later calculate the same likelihood on other data, as long as the variables in the list
have the same names.

```{r}
# Connect a likelihood to a data set.
cmb = function(f, d) function(p) f(p, d)
```

Second, `MakeADFun()` processes the NLL so that the tools that we use
in the rest of the course can work with it (the details are explained progressively
through the course). We give it the function returned by `cmb()` and a list with the
starting value of every parameter, and it returns the object that the optimiser needs:

```{r}
# Process the NLL of the 50 counts, with a starting value of 1 seedling per quadrat.
# silent = TRUE stops MakeADFun() from printing its progress.
fir_lambda_obj = MakeADFun(
  cmb(poisson_nll_lambda, list(counts = seedling_counts)),
  list(lambda = 1),
  silent = TRUE
)
```

We take three elements from this object that are passed to the `optim()` function:

- `par`, the starting values as a vector;
- `fn`, the NLL as a function of that vector;
- and `gr`, the gradients of the NLL with respect to parameters (details on why we need it in [Chapter 6](../Chapter_6/Theory.qmd) and [Chapter 8](../Chapter_8/Theory.qmd)).

```{r}
# optim() minimises fn, starting from par and using the slope gr.
fir_lambda_fit = optim(
  par = fir_lambda_obj$par,
  fn = fir_lambda_obj$fn,
  gr = fir_lambda_obj$gr,
  method = "BFGS"
)
```

The fourth argument (`method`) determines which specific algorithm is used for optimisation.
We will use `"BFGS"` in most cases as it is quite efficient. `optim()` returns a list and
its most relevant elements are

- `par`, the value of the parameter at the minimum (i.e., the maximum likelihood estimate);
- `value`,  the value of NLL at the maximum likelihood estimate;
- `convergence`, a code that is zero when the algorithm stopped normally, which is the first thing to
check after every fit. If you get a value different from 0 that means something went wrong.

```{r}
# Estimate, minimum of the negative log-likelihood, and convergence code (0 = normal stop).
fir_lambda_fit$par
fir_lambda_fit$value
fir_lambda_fit$convergence
```

## Constraints and transformations {#sec-constraints}

A **constraint** is a restriction on the values that a parameter can take. The mean
$\lambda$ of a Poisson distribution must be positive, and a probability must lie between
zero and one. An optimiser such as BFGS does not know about these restrictions. It can try
any real number, and for a negative $\lambda$ the function `dpois()` returns `NaN` (not a
number). In the fit above the search stayed among positive values, but that is not
guaranteed. In this book we deal with constraints by **transforming** the parameter: the
optimiser works with a quantity that can take any real value, and the NLL function
converts it into the parameter of the model.

For a positive parameter we use the logarithm. The optimiser searches over
$\log\lambda$, which can be any real number, and inside the function `exp()` turns it into
a positive mean:

```{r}
# Negative log-likelihood on the log scale: log_lambda can take any real value
# and exp() turns it into a positive mean.
poisson_nll = function(parms, data) {
  getAll(parms, data)
  # Transform the parameter to the original scale.
  lambda = exp(log_lambda)
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}
```

The steps of the fit are the same as before. The parameter in the list of starting values
is now `log_lambda`, so the search starts at $\log 1=0$:

```{r}
# Process the NLL of the 50 counts, starting from log(1).
fir_obj = MakeADFun(
  cmb(poisson_nll, list(counts = seedling_counts)),
  list(log_lambda = log(1)),
  silent = TRUE
)

# Fit and check the convergence code.
fir_fit = optim(par = fir_obj$par, fn = fir_obj$fn, gr = fir_obj$gr, method = "BFGS")
fir_fit$convergence
```

The element `par` of the fit is the estimate of $\log\lambda$, the quantity that the
optimiser searched over. We transform it back with `exp()` to report it in seedlings per
quadrat (`unname()` drops the name `log_lambda`, which no longer describes the value):

```{r}
# Estimate on the log scale and transformed back to seedlings per quadrat.
fir_fit$par
lambda_mle = unname(exp(fir_fit$par))
lambda_mle
```

On the original scale the estimate is again 2.14 seedlings per quadrat.

The same NLL can be connected to another data set, as long as its list uses the same
name, `counts`. For the first ten quadrats, only the list of data changes:

```{r}
# The same likelihood connected to the counts of the first ten quadrats.
fir_ten_obj = MakeADFun(
  cmb(poisson_nll, list(counts = ten_counts)),
  list(log_lambda = log(1)),
  silent = TRUE
)
fir_ten_fit = optim(
  par = fir_ten_obj$par,
  fn = fir_ten_obj$fn,
  gr = fir_ten_obj$gr,
  method = "BFGS"
)
fir_ten_fit$convergence

# Estimate in seedlings per quadrat.
unname(exp(fir_ten_fit$par))
```

The estimate is 2.2 seedlings per quadrat, the mean of the ten counts. We will use both
objects again in @sec-likelihood-uncertainty.

For a probability $p$ the corresponding transformation is the **logit**,
$\operatorname{logit}(p)=\log\left[p/(1-p)\right]$, which covers all the real numbers as
$p$ goes from zero to one. In R, `qlogis()` calculates the logit and `plogis()` its
inverse, which turns any real number into a value between zero and one (the names come
from the logistic distribution, of which they are the quantile and cumulative probability
functions):

```{r}
# The logit transformation for a probability, and its inverse.
qlogis(p = 0.25)
plogis(q = qlogis(p = 0.25))
```

The logit of 0.25 is a negative number, and `plogis()` returns the 0.25 that we started
with. We will use this transformation in @sec-likelihood-surface.

Fitting on a transformed scale and reporting on the original scale relies on a property of
maximum likelihood called **invariance**: the MLE of a function of a parameter is that
function of the MLE.^[See Zehna (1966), "Invariance of maximum likelihood estimators",
*Annals of Mathematical Statistics* 37: 744, for the general statement.] That is, for a
function $g$ of the parameter,

$$
\widehat{g(\theta)}=g(\hat\theta).
$$

Here the optimiser found the MLE of $\log\lambda$, and invariance tells us that its
exponential is the MLE of $\lambda$. The reason is that the transformation only relabels
the horizontal axis of the likelihood curve. Every value of $\log\lambda$ corresponds to
one value of $\lambda$ with the same likelihood, so the highest point is the same on both
scales.

Invariance also gives us the MLE of other quantities that are calculated from the
parameter. For example, under the Poisson model the probability that a quadrat has no
seedlings is $P(X=0\mid\lambda)=e^{-\lambda}$, and its MLE is $e^{-\hat\lambda}$:

```{r}
# By invariance, the estimate of the probability of an empty quadrat, exp(-lambda),
# is calculated from the estimate of lambda.
exp(-lambda_mle)
```

That is, we estimate that approximately 12% of the quadrats are empty (in the sample, 7 of
the 50 quadrats have no seedlings).
[Chapter 3](../Chapter_3/Theory.qmd#sec-sampling_distribution) used this same quantity as
an example of a biased estimator: $\hat\lambda$ is an unbiased estimator of $\lambda$, but
$e^{-\hat\lambda}$ is a biased estimator of $e^{-\lambda}$. So invariance carries over the
estimate, but it does not carry over unbiasedness.

::: {.callout-note title="When a fit fails"}
Check the element `convergence` after every fit. A code other than zero means that the
search did not finish normally, and the values in `par` are then not estimates.

While it searches, `optim()` can try parameter values for which the NLL cannot be
calculated, and `fn` then returns `NaN` without a warning. BFGS treats such a value as a
failed step and tries a shorter one, so this is harmless when the search ends with
convergence code 0 and the estimates are sensible. When the NLL is `NaN` at the starting
values, `optim()` stops with the error "initial value in 'vmmin' is not finite".

If a fit fails, check first that every constrained parameter is transformed inside the
NLL. Then try other starting values (the method-of-moments estimates are a good choice),
look at the data, and calculate the NLL at the starting values, for example with
`fir_obj$fn(fir_obj$par)`, to see whether it returns an ordinary number.
[Chapter 8](../Chapter_8/Theory.qmd) explains why optimisers fail and what else can be
done.
:::

:::: {.callout-important title="Practice: back to the ecological scale"}

A colleague fits two models with `optim()`. In the first, the parameter is the logarithm of
the mean number of ticks per lizard and the result is `par = 0.92`. In the second, the
parameter is the logit of the probability that a seed germinates and the result is
`par = -0.85`. Report both estimates on the ecological scale. Are they the maximum
likelihood estimates of the mean and of the probability?

::: {.callout-tip title="Solution" collapse="true"}

We apply the inverse of each transformation:

```{r}
# Inverse of the logarithm and inverse of the logit.
exp(0.92)
plogis(q = -0.85)
```

The estimates are approximately 2.5 ticks per lizard and a germination probability of
0.30. By invariance, they are the maximum likelihood estimates of the mean and of the
probability, because the optimiser found the maximum likelihood estimates of their
logarithm and logit.

:::
::::

# Models with several parameters {#sec-likelihood-surface}

With several parameters the likelihood is a function of all of them jointly, and the MLE is
the combination of values with the highest likelihood. The definitions of the previous
sections do not change, but the likelihood is now a surface over the parameters instead of
a curve. This raises questions that did not exist with one parameter: how to look at the
surface, how the estimates of the different parameters depend on each other, and how to
describe what the data say about one of them. These questions matter for the rest of the
book, because the models of Chapters 6 and 7 always have several parameters.

We will use the barn owl data of the Chapter 3 practical to illustrate these concepts. The
data set `Owls` in the package `glmmTMB` contains observations of barn owl nests (Roulin
and Bersier 2007). For each visit of a parent to a nest, the variable `SiblingNegotiation`
is the number of calls made by the nestlings. As in that practical, we keep the visits of
the female parent to nests where the nestlings had been fed (`Satiated`), so that all
counts can reasonably share one mean:

```{r}
# Load the Owls data from the glmmTMB package without attaching the package.
data(Owls, package = "glmmTMB")

# Keep the visits of the female parent to nests with satiated nestlings.
satiated_female = Owls[Owls$FoodTreatment == "Satiated" & Owls$SexParent == "Female", ]

# Number of calls at each of these visits, and the number of visits.
call_counts = satiated_female$SiblingNegotiation
n_visits = length(call_counts)
```

We also calculate the mean and the variance of the counts and, from them, the
method-of-moments estimate of the shape of a Negative Binomial distribution (the model is
introduced below):

```{r}
# Method-of-moments estimates, as in the practical of Chapter 3.
mu_hat = mean(call_counts)
variance_hat = mean((call_counts - mu_hat)^2)
k_mom = mu_hat^2 / (variance_hat - mu_hat)

c(n_visits = n_visits, mu_hat = mu_hat, variance_hat = variance_hat, k_mom = k_mom)
```

There are 122 visits, with a mean of 4.75 calls per visit and a variance of 42.3, almost
nine times the mean. Let $X_i$ be the number of calls at visit $i$. Because of this
overdispersion we use a
[Negative Binomial](../Supplements/distributions.qmd#sec-negative_binomial) model with mean
$\mu$ and shape $k$,

$$
X_i \sim \operatorname{NegBin}(\mu, k), \qquad \operatorname{Var}(X_i)=\mu+\frac{\mu^2}{k},
$$

and we assume that the visits are i.i.d. (a simplification to which we return in
@sec-misspecified). The method of moments gives $\hat\mu=\overline x\approx 4.75$ calls
per visit and $\hat k_{\mathrm{MM}}=\hat\mu^2/(\hat\sigma^2-\hat\mu)\approx 0.60$, where
$\hat\sigma^2$ is the variance of the sample with divisor $n$ and the subscript marks the
method-of-moments estimate, as in Chapter 3.

The NLL is written as before. The list of parameters now has two elements, `log_mu` and
`log_k` (both parameters are positive, so we use the logarithm for both), and the list of
data has one, `calls`. For the probability mass we use `dnbinom2()` of the package
RTMBdist, which is written in terms of the mean (`mu`) and the shape (`size`):

```{r}
# Negative log-likelihood of the Negative Binomial model with mean mu and shape k.
nb_nll = function(parms, data) {
  getAll(parms, data)
  # Transform the parameters to the original scale.
  mu = exp(log_mu)
  k = exp(log_k)
  -sum(dnbinom2(x = calls, mu = mu, size = k, log = TRUE))
}
```

::: {.callout-warning title="Two functions called `dnbinom2()`"}
The packages RTMB and RTMBdist both have a function called `dnbinom2()`, with different
arguments: in RTMBdist they are the mean and the shape (`mu` and `size`), and in RTMB the
mean and the variance (`mu` and `var`). R uses the function of the package that was
attached last, which is RTMBdist when we load `library(RTMB)` first and
`library(RTMBdist)` second, as at the start of this chapter. So always load both packages
in that order and name the arguments `mu` and `size`. If only RTMB is loaded, R stops
with an error about an unused argument `size`.
:::

The optimiser needs a starting value for each parameter. The method-of-moments estimates
are a sensible choice: they are quick to calculate and, being estimates of the same
parameters, they should not be far from the maximum. We give them to `MakeADFun()` in a
list with the names that the NLL uses, and `optim()` keeps those names in its result:

```{r}
# Process the NLL of the owl counts, starting from the method-of-moments estimates
# on the log scale.
owl_obj = MakeADFun(
  cmb(nb_nll, list(calls = call_counts)),
  list(log_mu = log(mu_hat), log_k = log(k_mom)),
  silent = TRUE
)

# Fit the model and check the convergence code.
owl_fit = optim(par = owl_obj$par, fn = owl_obj$fn, gr = owl_obj$gr, method = "BFGS")
owl_fit$convergence
```

The element `par` of the fit is now a vector with two named elements, `log_mu` and
`log_k`. We select each of them by its name and transform it back with `exp()`:

```{r}
# Maximum likelihood estimates on the original scale.
mu_mle = unname(exp(owl_fit$par["log_mu"]))
k_mle = unname(exp(owl_fit$par["log_k"]))
c(mu_mle = mu_mle, k_mle = k_mle)
```

The estimate of the mean is $\hat\mu\approx 4.75$ calls per
visit, which is the mean of the sample: for a Negative Binomial model the MLE of
the mean is the mean of the sample, whatever the value of $k$. The estimate of the shape
is $\hat k\approx 0.33$, far from the method-of-moments estimate of 0.60 at which the
search started. We return to the difference between the two estimates in
@sec-misspecified.

## The fitted model

`optim()` returns estimates for any model that we give it, so after a fit we check whether
the fitted model describes the data. For counts we can compare the relative frequency of
each count in the sample with the probability that the fitted model assigns to it. As a
reference we add a Poisson model with the same mean, which is the model that we would use
if there were no overdispersion (its MLE is the mean of the sample, as we saw in
@sec-maximum-likelihood). As a specific check, we calculate the fraction of visits without
calls in the sample and the probability that each fitted model assigns to a count of zero:

```{r}
# Fraction of visits without calls: observed, and under each fitted model.
c(
  observed = mean(call_counts == 0),
  nb_mle = dnbinom2(x = 0, mu = mu_mle, size = k_mle),
  poisson = dpois(x = 0, lambda = mu_hat)
)
```

The visits without calls are 44% of the sample (54 of 122). The fitted Negative Binomial
model assigns a probability of 0.41 to a count of zero, whereas under the Poisson model a
visit without calls would be very rare (a probability of 0.009). The log-likelihoods of
the two models compare them on the whole sample. The element `value` of a fit is the NLL
at the estimates, so the maximum log-likelihood of the Negative Binomial model is minus
that value:

```{r}
# Log-likelihood at the maximum likelihood estimates and for a Poisson model with the
# sample mean.
loglik_mle = -owl_fit$value
loglik_poisson = sum(dpois(x = call_counts, lambda = mu_hat, log = TRUE))
c(mle = loglik_mle, poisson = loglik_poisson)
```

The log-likelihood of the Negative Binomial model is about 335 units higher than that of
the Poisson model. @fig-owl_fits shows the two fitted distributions with the relative
frequencies of the sample. The Negative Binomial model reproduces the high frequency of
visits without calls and spreads the rest of its probability over a wide range of counts,
whereas the Poisson model concentrates its probability around the mean. We compare the two
models with a formal criterion in @sec-aic, and we return in @sec-misspecified to what the
Negative Binomial model does not describe well.

```{r}
#| echo: false
#| label: fig-owl_fits
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of calls per visit (bars) and probabilities under the fitted models: Negative Binomial with maximum likelihood estimates, and Poisson with the sample mean."
# Relative frequencies of each count in the sample.
count_values = 0:max(call_counts)
observed_data = data.frame(
  count = count_values,
  frequency = tabulate(call_counts + 1, nbins = length(count_values)) / n_visits
)

# Fitted probabilities for each model at every possible count.
fitted_data = rbind(
  data.frame(
    count = count_values,
    probability = dnbinom2(x = count_values, mu = mu_mle, size = k_mle),
    model = "Negative Binomial (maximum likelihood)"
  ),
  data.frame(
    count = count_values,
    probability = dpois(x = count_values, lambda = mu_hat),
    model = "Poisson"
  )
)

# Bars for the sample and lines with points for the models.
ggplot() +
  geom_col(
    data = observed_data,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(
    values = c("#D55E00", "grey20")
  ) +
  labs(
    x = "Number of calls per visit",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top", legend.direction = "vertical")
```

## The likelihood surface

With two parameters we can still evaluate the log-likelihood on a grid, which now has two
dimensions: one for the values of $\mu$ and one for the values of $k$. We store the results
in a matrix with one row per value of $\mu$ and one column per value of $k$, and fill it
with two nested `for` loops. Inside the loops we call `owl_obj$fn()`, which returns the
NLL for a vector of parameters on the scale that was optimised. So we give it the
logarithms of the candidate values and change the sign of the result:

```{r}
# Grids of candidate values for the mean (calls per visit) and the shape.
mu_grid = seq(from = 2.5, to = 9, length.out = 91)
k_grid = seq(from = 0.12, to = 0.75, length.out = 91)

# Pre-allocate a matrix and fill it with two nested loops: row i for the i-th mean
# and column j for the j-th shape. fn() takes log(mu) and log(k) and returns the NLL.
loglik_surface = matrix(NA, nrow = length(mu_grid), ncol = length(k_grid))
for (i in seq_along(mu_grid)) {
  for (j in seq_along(k_grid)) {
    loglik_surface[i, j] = -owl_obj$fn(c(log(mu_grid[i]), log(k_grid[j])))
  }
}

# Relative log-likelihood: the difference from the maximum found by optim().
relative_surface = loglik_surface - loglik_mle
```

A matrix like this one can be drawn with `contour()`, as we did in the Chapter 1 practical:

```{r}
#| eval: false
# The matrix can be drawn with contour(), as in the practical of Chapter 1.
contour(x = mu_grid, y = k_grid, z = relative_surface, levels = c(-6, -4, -3, -2, -1))
```

@fig-likelihood_surface shows the same contours. Each line joins the combinations of $\mu$
and $k$ that have the same relative log-likelihood, like the contours of the joint density
in [Chapter 2](../Chapter_2/Theory.qmd#sec-two_continuous_variables), and the MLE is at the
top of the hill. The contours
are not tilted: the value of $\mu$ with the highest log-likelihood is the same for every
value of $k$ (the mean of the sample), and the best value of $k$ hardly changes with
$\mu$. The surface is not symmetric, however, as it falls more slowly towards large values
of $\mu$ and of $k$ than towards small ones.

```{r}
#| echo: false
#| label: fig-likelihood_surface
#| fig-cap: "Contours of the relative log-likelihood of the owl data as a function of the mean and the shape of the Negative Binomial model. The circle is the maximum likelihood estimate."
# Arrange the matrix as a data frame: mu varies fastest, as in as.vector() of the matrix.
surface_data = expand.grid(mu = mu_grid, k = k_grid)
surface_data$relative = as.vector(relative_surface)

# Contours of the relative log-likelihood with the maximum likelihood estimate.
ggplot() +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_point(
    data = data.frame(mu = mu_mle, k = k_mle),
    aes(x = mu, y = k),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    x = expression(paste("Mean number of calls per visit, ", mu)),
    y = expression(paste("Shape, ", italic(k)))
  ) +
  theme_classic()
```

## Parameterisation and the shape of the surface

The same model can be written with different sets of parameters, each of which is called a
**parameterisation**. For the Negative Binomial distribution, R's own parameters are the
shape $k$ (`size`) and a probability $p$ (`prob`), which is related to the mean through
$p=k/(k+\mu)$ (the "Inverse" line of the
[supplement entry](../Supplements/distributions.qmd#sec-negative_binomial)).

```{r}
#| include: false
# Negative log-likelihood in R's own parameters: shape k (size) and probability p (prob),
# with log(k) and logit(p) as the transformed parameters.
nb_prob_nll = function(parms, data) {
  getAll(parms, data)
  k = exp(log_k)
  p = plogis(logit_p)
  -sum(dnbinom(x = calls, size = k, prob = p, log = TRUE))
}

# Starting values: the method-of-moments shape and p = k / (k + mu).
owl_prob_obj = MakeADFun(
  cmb(nb_prob_nll, list(calls = call_counts)),
  list(log_k = log(k_mom), logit_p = qlogis(k_mom / (k_mom + mu_hat))),
  silent = TRUE
)
owl_prob_fit = optim(
  par = owl_prob_obj$par,
  fn = owl_prob_obj$fn,
  gr = owl_prob_obj$gr,
  method = "BFGS"
)
owl_prob_fit$convergence

# Estimates on the original scale and the maximum log-likelihood.
k_mle_prob = unname(exp(owl_prob_fit$par["log_k"]))
p_mle = unname(plogis(owl_prob_fit$par["logit_p"]))
c(k_mle_prob = k_mle_prob, p_mle = p_mle, loglik = -owl_prob_fit$value)

# By invariance, p equals k / (k + mu) evaluated at the first fit.
k_mle / (k_mle + mu_mle)
```

I fitted the same model to the owl data in these parameters, with
$\log k$ and $\operatorname{logit}p$ as the transformed parameters, and calculated the
log-likelihood on a grid of values of $k$ and $p$ (the code repeats the steps above with
the arguments `size` and `prob` of `dnbinom()`, so it is not shown). The maximum
log-likelihood is $-299.44$, the same as before, and so is
$\hat k\approx 0.33$. The estimate of the probability is $\hat p\approx 0.064$, which is
the value of $\hat k/(\hat k+\hat\mu)$ from the first fit, as invariance says (the two
numbers differ in the fifth decimal because `optim()` always has a bit of numerical error,
nothing to really worry about). It is the same model at the same maximum. What changes is
the shape of the surface around the maximum (@fig-surface_size_prob).

```{r}
#| include: false
# Grids of candidate values for the shape and the probability.
k_grid_prob = seq(from = 0.12, to = 0.75, length.out = 91)
p_grid = seq(from = 0.02, to = 0.16, length.out = 91)

# Fill the matrix of log-likelihoods with two nested loops; fn() takes log(k) and logit(p).
loglik_surface_prob = matrix(NA, nrow = length(k_grid_prob), ncol = length(p_grid))
for (i in seq_along(k_grid_prob)) {
  for (j in seq_along(p_grid)) {
    loglik_surface_prob[i, j] =
      -owl_prob_obj$fn(c(log(k_grid_prob[i]), qlogis(p_grid[j])))
  }
}
relative_surface_prob = loglik_surface_prob - loglik_mle
```

```{r}
#| echo: false
#| label: fig-surface_size_prob
#| fig-cap: "Contours of the relative log-likelihood of the owl data as a function of the shape k and the probability p of the Negative Binomial model. The circle is the maximum likelihood estimate."
# Arrange the matrix as a data frame: k varies fastest.
surface_prob_data = expand.grid(k = k_grid_prob, p = p_grid)
surface_prob_data$relative = as.vector(relative_surface_prob)

# Contours with the maximum likelihood estimate.
ggplot() +
  geom_contour(
    data = surface_prob_data,
    aes(x = k, y = p, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_point(
    data = data.frame(k = k_mle_prob, p = p_mle),
    aes(x = k, y = p),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    x = expression(paste("Shape, ", italic(k))),
    y = expression(paste("Probability, ", italic(p)))
  ) +
  theme_classic()
```

The contours are now tilted: a larger $k$ goes with a larger $p$. In
these parameters the mean of the distribution is $k(1-p)/p$, and the data determine the
mean number of calls well. So a higher value of $k$ can be compensated by a higher value of
$p$ that keeps the mean close to 4.75, with little loss of likelihood. Tilted contours
describe two parameters whose estimates depend on
each other, which has consequences for slices and profiles (below) and for the uncertainty
of the estimates (@sec-likelihood-uncertainty).

This book writes distributions in terms of their mean where possible, because we normally
model the effect of covariates on the mean (see the
[Supplement](../Supplements/distributions.qmd)). For the Negative Binomial distribution
that parameterisation also gives estimates that hardly depend on each other, as
@fig-likelihood_surface showed. The choice of parameters also matters for numerical
optimisation, which is a topic of [Chapter 8](../Chapter_8/Theory.qmd).

## Slices and profiles

A surface can only be drawn for two parameters, and even then we usually want to know what
the data say about one parameter at a time. There are two ways of reducing a likelihood
surface to a curve for one parameter, and they answer different questions.

A **likelihood slice** fixes the other parameters at one value, normally their estimates,
and varies the parameter of interest. For the shape $k$ in the parameters $k$ and $p$, the
slice is $\ell(k,\hat p)$, a cut through the surface along a straight line.

A **profile likelihood** re-estimates the other parameters for every value of the parameter
of interest:

$$
\ell_{\mathrm{prof}}(k)=\max_{p}\,\ell(k,p).
$$

That is, for each value of $k$ we find the value of $p$ that maximises the log-likelihood
with $k$ held fixed, and we record that maximum. The parameters that are re-estimated are
called **nuisance parameters**: they are needed to define the model, but they are not the
parameter that the question is about. Which parameter is of interest and which is a
nuisance depends on the question, and we will profile each parameter in turn.

We calculate both curves on a grid of values of $k$: the slice evaluates the
log-likelihood with $p$ fixed at $\hat p$, and the profile runs `optim()` at each value of
$k$ to re-estimate $p$ (the code of a profile is shown in @sec-profile_intervals, where we
need it to calculate confidence intervals).

```{r}
#| include: false
# Fine grid of shape values on which the slice and the profile are evaluated.
k_fine = seq(from = 0.12, to = 0.75, length.out = 127)

# Slice: p is fixed at its estimate. Profile: for each k, p is re-estimated with optim(),
# starting from the joint estimate. The argument map of MakeADFun() keeps log(k) fixed
# at the value given in the list of parameters, so that only logit(p) is estimated.
slice_loglik = rep(NA, times = length(k_fine))
profile_loglik = rep(NA, times = length(k_fine))
profile_p = rep(NA, times = length(k_fine))
profile_convergence = rep(NA, times = length(k_fine))
for (i in seq_along(k_fine)) {
  slice_loglik[i] = -owl_prob_obj$fn(c(log(k_fine[i]), qlogis(p_mle)))
  inner_obj = MakeADFun(
    cmb(nb_prob_nll, list(calls = call_counts)),
    list(log_k = log(k_fine[i]), logit_p = qlogis(p_mle)),
    map = list(log_k = factor(NA)),
    silent = TRUE
  )
  inner_fit = optim(
    par = inner_obj$par,
    fn = inner_obj$fn,
    gr = inner_obj$gr,
    method = "BFGS"
  )
  profile_loglik[i] = -inner_fit$value
  profile_p[i] = plogis(inner_fit$par)
  profile_convergence[i] = inner_fit$convergence
}

# Relative log-likelihoods and the range of k within 1.92 of the maximum
# (the cutoff is explained in the section on profile likelihood intervals).
slice_relative = slice_loglik - loglik_mle
profile_relative = profile_loglik - loglik_mle
range(k_fine[slice_relative >= -1.92])
range(k_fine[profile_relative >= -1.92])

# All inner fits stopped normally.
all(profile_convergence == 0)
```

```{r}
#| include: false
# In the parameters mu and k, the best mu for any k is the sample mean, so the
# profile of k equals the slice at the estimate of mu. Check this on the same grid.
slice_k_mu = rep(NA, times = length(k_fine))
profile_k_mu = rep(NA, times = length(k_fine))
for (i in seq_along(k_fine)) {
  slice_k_mu[i] = -owl_obj$fn(c(log(mu_mle), log(k_fine[i])))
  inner_obj = MakeADFun(
    cmb(nb_nll, list(calls = call_counts)),
    list(log_mu = log(mu_mle), log_k = log(k_fine[i])),
    map = list(log_k = factor(NA)),
    silent = TRUE
  )
  inner_fit = optim(
    par = inner_obj$par,
    fn = inner_obj$fn,
    gr = inner_obj$gr,
    method = "BFGS"
  )
  profile_k_mu[i] = -inner_fit$value
}
max(abs(slice_k_mu - profile_k_mu))
```

```{r}
#| echo: false
#| label: fig-slice_profile
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Slice and profile of the log-likelihood for the shape k of the Negative Binomial model fitted to the owl data. Left: contours in k and p with the slice (dashed line) and the values of p that maximise the log-likelihood for each k (solid line). Right: relative log-likelihood along the slice and along the profile; the horizontal line is at -1.92."
library(patchwork)

# Left panel: contours with the slice and the path of the profile.
path_data = data.frame(k = k_fine, p = profile_p)
left_panel = ggplot() +
  geom_contour(
    data = surface_prob_data,
    aes(x = k, y = p, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_hline(yintercept = p_mle, linetype = "dashed", colour = "#D55E00") +
  geom_line(data = path_data, aes(x = k, y = p), colour = "#D55E00", linewidth = 0.8) +
  geom_point(
    data = data.frame(k = k_mle_prob, p = p_mle),
    aes(x = k, y = p),
    colour = "#D55E00",
    size = 3
  ) +
  labs(x = expression(paste("Shape, ", italic(k))), y = expression(paste("Probability, ", italic(p)))) +
  theme_classic()

# Right panel: relative log-likelihood along the slice and along the profile.
curve_data = rbind(
  data.frame(k = k_fine, relative = slice_relative, curve = "Slice"),
  data.frame(k = k_fine, relative = profile_relative, curve = "Profile")
)
right_panel = ggplot(curve_data, aes(x = k, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -1.92, colour = "grey40") +
  coord_cartesian(ylim = c(-6, 0)) +
  labs(x = expression(paste("Shape, ", italic(k))), y = "Relative log-likelihood", linetype = NULL) +
  theme_classic() +
  theme(legend.position = "top")

left_panel | right_panel
```

@fig-slice_profile shows why the two curves differ. In the left panel, the slice follows
the horizontal line at $\hat p$, which leaves the high part of the surface quickly because
the contours are tilted. The profile follows the ridge of the surface: for each $k$ it
moves to the best $p$. Both pass through the MLE, and everywhere else the profile is
higher than the slice, since it is the maximum over all the values of $p$ (right panel).
The slice is therefore narrower: the values of $k$ within 1.92 units of the maximum, a
cutoff that we justify in @sec-profile_intervals, go from 0.26 to 0.41 for the slice and
from 0.24 to 0.45 for the profile. The slice describes what the data would say about $k$
if we knew that $p$ is exactly 0.064, whereas the profile takes into account that $p$ is
estimated from the same data.

How much the two curves differ depends on the parameterisation. With the mean as the
second parameter, the slice and the profile of $k$ coincide, because the best $\mu$ is the
mean of the sample for every value of $k$. The profile itself does not depend on how the
nuisance parameter is written: maximising over $p$ or over $\mu$ gives the same curve for
$k$. A slice does depend on it.

We will use profiles to calculate confidence intervals (@sec-profile_intervals). Slices
need no optimisation, so they are a quick way to inspect a fitted model, and they return
as a diagnostic in [Chapter 8](../Chapter_8/Theory.qmd).

::: {.callout-note title="Explore a likelihood surface"}
Kristoffer Magnusson's
[interactive page on maximum likelihood](https://rpsychologist.com/likelihood/) lets you
do by hand what an optimiser does. It uses a Normal model with two parameters, a mean and a
variance, and a sample of ten observations. In the part "Likelihood Calculation", move the
sliders of the mean and the variance and watch how the contribution of each observation
and the log-likelihood change. In "Finding the Maximum Likelihood Estimates", follow how an
algorithm climbs the log-likelihood step by step. The last part of the page is about
hypothesis tests, which we do not use in this course.
:::

# Uncertainty from the likelihood {#sec-likelihood-uncertainty}

In [Chapter 3](../Chapter_3/Theory.qmd#sec-standard_errors) we reported the uncertainty of
an estimate with a standard error and a confidence interval, and for the mean of Poisson
counts we had a formula for the standard error. For most models there is no such formula.
In this section we calculate standard errors and confidence intervals from the likelihood
itself, which we can do for any model that we fit with `optim()`.

A confidence interval for a parameter can be calculated from a likelihood in two ways. The
**Wald interval** uses the asymptotic normality of the MLE and the curvature of the
log-likelihood at its maximum, which gives the standard error. The **profile likelihood
interval** uses the whole shape of the log-likelihood around the maximum. In this book the
profile likelihood interval is the better method in general, but the Wald interval is the fast
approximation that software usually reports. This section explains both, shows when they
differ and checks their coverage by simulation. The Wald interval relies on a result for
large samples that we state below and use straight away. In @sec-mle_properties we then
examine that result by simulation, together with the other properties of maximum
likelihood estimators.

## Standard errors and Wald intervals

The standard error of an MLE follows from a result for large samples.
[Chapter 3](../Chapter_3/Theory.qmd#sec-large_samples) ended with the central limit
theorem: the mean of $n$ i.i.d. observations is asymptotically normal. It also announced
that this chapter would generalise the result to other parameters. The generalisation is
that the MLE of any parameter is **asymptotically normal**: as the sample grows, its
sampling distribution approaches a Normal distribution centred on the
parameter.^[Fisher used the large-sample standard error of maximum likelihood estimators
in 1922 and developed the concept of information in 1925, with arguments that were not
rigorous. The result was made rigorous by Wald (1943) and by Cramér (1946), whose
treatment states the conditions explicitly.] For one parameter:

$$
\hat\theta \overset{\mathrm{approx}}{\sim}
\operatorname{Normal}\left(\theta,\frac{1}{\mathcal I(\theta)}\right)
\quad\text{for sufficiently large }n,
$$

where $\mathcal I(\theta)$ is the **Fisher information** that the sample contains about
$\theta$. In plain English, across hypothetical experiments with a large sample^[These results require some conditions on the model, known as regularity
conditions. In words: the parameter is not on the boundary of its range (e.g., a variance
of exactly zero), the support of the distribution does not depend on the parameter, the
log-likelihood is a smooth function of the parameters, and the number of parameters does
not grow with the sample size.] the
estimates fall around the parameter in the shape of a bell (the shape of a Normal
distribution), and the more information the sample contains, the narrower the bell.

::: {.callout-note title="Asymptotic normality of maximum likelihood estimators"}
For a large sample of i.i.d. observations from the model, the sampling distribution of a
maximum likelihood estimator is approximately Normal, centred on the parameter, with a
variance equal to the inverse of the Fisher information. The observed information (minus
the second derivative of the log-likelihood at the estimate) estimates it from the sample.
:::

We build the Wald interval in four steps, which are the subsections below. We first
calculate the information and the standard error for a model with one parameter, where
everything can be done by hand ("Fisher information"). We then generalise the calculation
to several parameters with the Hessian matrix ("The Hessian matrix"). Because the fits
of this chapter search for unconstrained parameters, we next see how the standard error
and the interval change when a parameter is transformed ("Constrained and unconstrained
parameters"). Finally, we let TMB do all these calculations ("Automation with TMB").

### Fisher information

The information measures how sharply the log-likelihood is curved around its maximum. We
calculate it from our sample as the **observed information**, which is the negative of the second
derivative of the log-likelihood at the estimate,^[Strictly, the Fisher information in the
result above is the **expected information**: the expectation of $-\ell''(\theta)$ across
hypothetical experiments. The observed information is calculated from the log-likelihood of
the sample that we have, and the two agree in large samples. Efron and Hinkley (1978)
compared them and argued for the observed information.] and from it we obtain an estimate
of the variance and of the standard error of the estimator:

$$
\hat{\mathcal I}=-\ell''(\hat\theta),
\qquad
\widehat{\operatorname{Var}}(\hat\theta)=\frac{1}{\hat{\mathcal I}},
\qquad
\widehat{\operatorname{SE}}(\hat\theta)=\frac{1}{\sqrt{\hat{\mathcal I}}}.
$$

The second derivative measures how fast the slope of a curve changes. At a maximum it is
negative, and the more negative it is, the faster the log-likelihood falls as we move away
from the estimate. A sharply curved log-likelihood therefore means that the data
discriminate well among parameter values, which corresponds to a small variance of the
estimator. A flat log-likelihood means that many parameter values are almost equally well
supported, which corresponds to a large variance.

For the Poisson model we can do the calculation by hand. The second derivative of its
log-likelihood with respect to $\lambda$ is $\ell''(\lambda)=-\sum x_i/\lambda^2$ (the
derivation is in the box of @sec-maximum-likelihood). At $\hat\lambda=\overline x$ the sum
of the counts is $n\hat\lambda$, so the observed information is $n/\hat\lambda$. We do the
calculation for the ten quadrats of @fig-likelihood_curve, where $\hat\lambda=2.2$, and
keep this small sample until we have calculated the three intervals of this section. At
the end of @sec-profile_intervals we repeat them for the 50 quadrats.

$$
\hat{\mathcal I}=\frac{n\hat\lambda}{\hat\lambda^2}=\frac{n}{\hat\lambda}=\frac{10}{2.2}\approx 4.55,
\qquad
\widehat{\operatorname{Var}}(\hat\lambda)=\frac{\hat\lambda}{n}=0.22,
\qquad
\widehat{\operatorname{SE}}(\hat\lambda)=\sqrt{0.22}\approx 0.47.
$$

The result, $\sqrt{\hat\lambda/n}$, is exactly the formula of the plug-in standard error
that we obtained in [Chapter 3](../Chapter_3/Theory.qmd#sec-standard_errors) from the
variance of a mean. Maximum likelihood gives it to us from the curvature of the
log-likelihood, which we can calculate for any model, also when there is no formula for
the variance of the estimator. We can verify our maths with R:

```{r}
# Observed information of the Poisson model at the estimate, for the ten quadrats:
# -l''(lambda_hat) = sum(x) / lambda_hat^2, which equals n / lambda_hat at lambda_hat = mean(x).
observed_information = sum(ten_counts) / lambda_hat_ten^2
observed_information

# Approximate variance of the estimator and its standard error (seedlings per quadrat).
approximate_variance = 1 / observed_information
standard_error = sqrt(approximate_variance)
c(approximate_variance = approximate_variance, standard_error = standard_error)
```

The same formula gives the standard error for all 50 quadrats:

```{r}
# The same standard error for all 50 quadrats: the value of Chapter 3.
sqrt(lambda_hat / n_quadrats)
```

With all 50 quadrats the standard error is 0.207 seedlings per quadrat, the value that we
calculated in Chapter 3. With ten quadrats it is more than twice as large (0.47), because
the sample contains about five times less information.

In [Chapter 3](../Chapter_3/Theory.qmd#sec-confidence_intervals) we defined the Wald
interval^[The interval is named after Abraham Wald, who developed in 1943 the large-sample
test on which it is based.] as
$\hat\theta\pm z_{0.975}\,\widehat{\operatorname{SE}}(\hat\theta)$ and justified it with
the approximate normality of the estimator. For an MLE the standard error comes from the
observed information, as we have just seen. For the ten quadrats we calculated it by
hand, $\widehat{\operatorname{SE}}(\hat\lambda)=\sqrt{\hat\lambda/n}\approx 0.47$, so the
interval is:

```{r}
# Wald interval for the Poisson mean of the ten quadrats, on the original scale.
critical_value = qnorm(p = 0.975)
wald_original = lambda_hat_ten + c(-1, 1) * critical_value * standard_error
wald_original
```

That is, 1.28 to 3.12 seedlings per quadrat. This interval is symmetric around the
estimate and it is calculated on the original scale of the parameter, with the standard
error of $\hat\lambda$.

### The Hessian matrix

With several parameters, the information is a matrix that contains the second derivatives
of the log-likelihood with respect to every pair of parameters. The matrix of second
derivatives of a function is called its **Hessian**. The Hessian of the NLL at the
estimates is the observed information matrix, and its inverse approximates the variances
of the estimates (on its diagonal) and the covariances between them (outside the
diagonal). We use it first for the Poisson model and then for the owl model.

We do not need to derive the second derivatives by hand. With `hessian = TRUE`, `optim()`
also returns the Hessian of the function that it minimised, calculated numerically at the
estimate. It is the Hessian with respect to the parameter that `optim()` searched over,
which in our fit is $\log\lambda$ (see @sec-constraints). So the standard error that we
obtain from it is the standard error of the estimate of $\log\lambda$:

```{r}
# Fit the Poisson model to the ten quadrats again, now asking for the Hessian.
fir_ten_fit = optim(
  par = fir_ten_obj$par, fn = fir_ten_obj$fn, gr = fir_ten_obj$gr,
  method = "BFGS", hessian = TRUE
)

# The Hessian is a 1 by 1 matrix; take its single element. Its inverse is the
# approximate variance of the estimate of log(lambda).
hessian_log = fir_ten_fit$hessian[1, 1]
se_log_lambda = sqrt(1 / hessian_log)
c(hessian_log = hessian_log, se_log_lambda = se_log_lambda)
```

A fit with one parameter returns a matrix with one row and one column, from which we take
the single element. The Hessian is 22 (you can check that the second derivative of the NLL
with respect to $\log\lambda$ is $n\lambda$, which at the estimate is the total number of
seedlings) and the standard error of the estimate of
$\log\lambda$ is $1/\sqrt{22}\approx 0.213$.

The owl model has two parameters, so its Hessian is a matrix with two rows and two
columns, and we need its inverse, which approximates the covariance matrix of the
estimates of $\log\mu$ and $\log k$. We fit the model of the previous section again, now
asking `optim()` for the Hessian. The model is the same, so the estimates do not change:

```{r}
# Fit the owl model again, now asking optim() for the Hessian.
owl_fit = optim(
  par = owl_obj$par, fn = owl_obj$fn, gr = owl_obj$gr,
  method = "BFGS", hessian = TRUE
)
```

Three functions take us from the Hessian to the standard errors:

- `solve()` applied to a matrix returns its inverse (matrices were introduced in the
  Chapter 1 practical);
- `diag()` extracts the diagonal of a matrix, which here holds the variances of the two
  estimates;
- `sqrt()` turns the variances into standard errors.

```{r}
# Invert the Hessian, take the variances from its diagonal and their square roots.
owl_covariance = solve(owl_fit$hessian)
sqrt(diag(owl_covariance))
```

These are the standard errors of the estimates of $\log\mu$ and $\log k$, which are the
parameters that `optim()` searched over, and not of $\mu$ and $k$ themselves. We now look
at why that matters.

### Constrained and unconstrained parameters

Both the mean $\lambda$ of the Poisson model and the shape $k$ of the Negative Binomial
model are constrained to be positive, and in @sec-constraints we removed the constraint
by fitting their logarithms, which can take any real value. The Hessian that we have just
used belongs to the unconstrained parameter. For the ten quadrats we therefore have two
standard errors, one for the estimator $\hat\lambda$ (calculated by hand, in seedlings per
quadrat) and one for the estimator of $\log\lambda$ (from the Hessian). The second is
approximately the first one divided by the estimate, that is, the standard error relative
to the size of the estimate ($0.47/2.2\approx 0.213$). We compare them with two ways of
moving the second one back to the original scale:

```{r}
c(
  # By hand, from the observed information of lambda (seedlings per quadrat).
  original_scale = standard_error,
  # From the Hessian with respect to log(lambda).
  log_scale = se_log_lambda,
  # exp() applied to the standard error on the log scale.
  exp_of_se = exp(se_log_lambda),
  # The standard error on the log scale multiplied by the estimate.
  se_times_estimate = lambda_hat_ten * se_log_lambda
)
```

Applying `exp()` to the standard error gives a number that is far from the standard error
of $\hat\lambda$, whereas multiplying it by the estimate reproduces it. A standard error
cannot be back-transformed as we back-transform a value of the parameter, because it is
calculated from expectations and expectations do not pass through non-linear functions
(this is known as Jensen's inequality).^[A standard error is the square root of an expected
squared deviation. **Jensen's inequality** states that, if $g$ is a convex function (as
$\exp$ is), then $E[g(X)]\ge g(E[X])$, with equality only when $g$ is linear. A non-linear
transformation of an estimator therefore does not transform its mean, and the standard
error, which is built from expectations, is not transformed either: the transformation
stretches the sampling distribution by different amounts in different places. For a
function $g$ of an estimate, the delta method corrects for this by multiplying the
standard error by the slope of the function at the estimate,
$\widehat{\operatorname{SE}}(g(\hat\theta))\approx
|g'(\hat\theta)|\,\widehat{\operatorname{SE}}(\hat\theta)$, and for $g=\exp$ the slope is
$\exp(\hat\theta)=\hat\lambda$, which is the multiplication in the last element above.]
The correction that gives the right standard error is the **delta method**, which
[Chapter 6](../Chapter_6/Theory.qmd#sec-delta-method) explains.

The endpoints of an interval are different. We can transform them because the logarithm is
an increasing function: $\log\lambda$ lies between the two endpoints on the log scale
exactly when $\lambda$ lies between their exponentials. So the transformed interval
contains the parameter in the same hypothetical experiments as the interval on the log
scale, and it has the same coverage. We calculate the Wald interval on the log scale,
where the standard error from the Hessian applies, and transform its endpoints back with
`exp()`:

```{r}
# Wald interval on the log scale, with the endpoints transformed back to seedlings
# per quadrat.
wald_log = exp(
  fir_ten_fit$par[["log_lambda"]] + c(-1, 1) * critical_value * se_log_lambda
)
wald_log
```

This interval (1.45 to 3.34 seedlings per quadrat) differs from the first one (1.28 to
3.12), although both are 95% Wald intervals for the same parameter and the same data. A
Wald interval depends on the scale on which it is calculated, because the Normal
approximation is applied to a different quantity in each case ($\hat\lambda$ or its
logarithm). The interval from the log scale is better here because it captures the
asymmetry around the estimate and it cannot include negative values, which addresses two
shortcomings of the Wald interval on the original scale.

The Wald interval amounts to replacing the log-likelihood by a parabola with the same
curvature at the maximum,

$$
\ell(\theta)-\ell(\hat\theta)\approx
-\frac{(\theta-\hat\theta)^2}{2\,\widehat{\operatorname{SE}}(\hat\theta)^2},
$$

which is why it is also called a quadratic approximation, and the interval a quadratic
interval. The approximation is good when the log-likelihood is close to a parabola on the
scale that we use, and we saw in @fig-likelihood_curve that for the ten quadrats it is not
symmetric.

### Automation with TMB

The steps above are the same for every model: obtain the Hessian at the estimates, invert
it, take the square roots of its diagonal and, for a quantity calculated from the
parameters, apply the delta method. TMB does all of this for us. The function
`sdreport()` of RTMB takes the object returned by `MakeADFun()` after the fit (it reads
the estimates from the object), calculates the Hessian itself, and `summary()` of its
result is a table with one row per parameter, on the scale that was optimised, and two
columns:

- `Estimate`, the estimate of the parameter;
- `Std. Error`, its standard error.

```{r}
# sdreport() calculates the standard errors from the Hessian at the estimates.
fir_ten_sd = sdreport(fir_ten_obj)
fir_ten_summary = summary(fir_ten_sd)
fir_ten_summary
```

The standard error of `log_lambda` is the one that we obtained from the Hessian, and from
here on we use `sdreport()`. For a quantity calculated from the parameters, such as
$\lambda$ or the shape $k$ of the owl model, we ask for its standard error with
`ADREPORT()` inside the NLL. TMB then applies the delta method for us and adds a row for
that quantity to the table. For the owl model we add one line to the NLL, `ADREPORT(k)`,
which asks for the standard error of $k$ itself, next to that of $\log k$:

```{r}
# The Negative Binomial NLL with one more line.
nb_nll = function(parms, data) {
  getAll(parms, data)
  mu = exp(log_mu)
  k = exp(log_k)
  # Ask for the standard error of k itself.
  ADREPORT(k)
  -sum(dnbinom2(x = calls, mu = mu, size = k, log = TRUE))
}
```

We process the new NLL and fit it again, because `sdreport()` reads the estimates from
the processed object. The model is the same, so the estimates do not change:

```{r}
# Process the new NLL and fit it again.
owl_obj = MakeADFun(
  cmb(nb_nll, list(calls = call_counts)),
  list(log_mu = log(mu_hat), log_k = log(k_mom)),
  silent = TRUE
)
owl_fit = optim(par = owl_obj$par, fn = owl_obj$fn, gr = owl_obj$gr, method = "BFGS")
```

The table of `sdreport()` has the standard errors that we obtained from the inverse of the
Hessian, and one more row for the quantity in `ADREPORT()`:

```{r}
# The standard errors of log(mu) and log(k), and the row of k from ADREPORT().
owl_sd = sdreport(owl_obj)
owl_summary = summary(owl_sd)
owl_summary
```

The last row is the quantity that we asked for with `ADREPORT()`: $\hat k\approx 0.327$,
with a standard error of 0.054. It agrees with the delta method for the exponential
that we applied above, $0.327\times 0.165$, the estimate multiplied by the standard error
of $\log k$. We can now calculate a Wald interval for $k$ on each of the two scales:

```{r}
# Lower and upper endpoints of a Wald interval for every row of the table.
owl_lower = owl_summary[, "Estimate"] - critical_value * owl_summary[, "Std. Error"]
owl_upper = owl_summary[, "Estimate"] + critical_value * owl_summary[, "Std. Error"]

# Rows log_mu and log_k: intervals on the log scales, with the endpoints transformed back.
owl_wald_mu = exp(c(owl_lower[["log_mu"]], owl_upper[["log_mu"]]))
owl_wald_k = exp(c(owl_lower[["log_k"]], owl_upper[["log_k"]]))

# Row k, from ADREPORT(): interval for k on the original scale.
owl_wald_k_original = c(owl_lower[["k"]], owl_upper[["k"]])

rbind(
  "mu, from the log scale" = owl_wald_mu,
  "k, from the log scale" = owl_wald_k,
  "k, original scale" = owl_wald_k_original
)
```

The Wald intervals from the log scales are 3.45 to 6.55 calls per visit for $\mu$ and
0.237 to 0.453 for $k$. On the original scale the interval for $k$ is 0.221 to 0.434: it
is symmetric around the estimate and lies lower than the first one. As for the ten
quadrats, the same fit gives two different Wald intervals, and in @sec-profile_intervals
we compare both with the profile likelihood interval.

:::: {.callout-important title="Practice: an interval from the log scale"}

In the practice about ticks on lizards, the fit on the log scale returned `par = 0.92`.
For that parameter, `sdreport()` reports a standard error of 0.141. Calculate a 95%
Wald interval for the mean number of ticks per lizard. Why is it not symmetric around the
estimate?

::: {.callout-tip title="Solution" collapse="true"}

The standard error is that of the estimate of the logarithm of the mean, so we calculate
the interval on the log scale and transform its endpoints:

```{r}
# Wald interval on the log scale, back-transformed to ticks per lizard.
exp(0.92 + c(-1, 1) * qnorm(p = 0.975) * 0.141)
```

The interval goes from approximately 1.9 to 3.3 ticks per lizard, around an estimate of
2.5. It is symmetric on the log scale, and `exp()` stretches the part above the estimate
more than the part below it.

:::
::::

## Covariance between estimates

The inverse of the Hessian contains more than variances. Its off-diagonal elements are the
**covariances** between the estimates, which describe how the estimates of two parameters
vary together across hypothetical experiments. A covariance divided by the two standard
errors is the correlation between the estimates
([Chapter 2](../Chapter_2/Theory.qmd#sec-two_continuous_variables) introduced the
correlation as a parameter of the bivariate Normal distribution), and `cov2cor()` does this
conversion for a whole matrix. `sdreport()` keeps the inverse of the Hessian in its
element `cov.fixed`:

```{r}
# Correlations between the estimates, from the covariance matrix kept by sdreport().
cov2cor(owl_sd$cov.fixed)

# The same for the parameters k and p, whose estimates are correlated.
cov2cor(sdreport(owl_prob_obj)$cov.fixed)
```

For the estimates of $\log\mu$ and $\log k$ the correlation is zero to four decimals. For
those of $\log k$ and $\operatorname{logit}p$ it is 0.71: hypothetical experiments that
overestimate $k$ would also tend to overestimate $p$. This is the tilt of
@fig-surface_size_prob expressed as a number. In large samples the estimates of two
parameters jointly follow a bivariate Normal distribution, and the contours of its density
have the shape of the contours of the likelihood near the maximum.

## Profile likelihood intervals {#sec-profile_intervals}

The Wald interval uses the curvature of the log-likelihood at one point, its maximum. The
profile likelihood interval uses the whole curve: it contains all the values of the
parameter whose relative profile log-likelihood is above a cutoff, that is, the values
that the data support almost as well as the estimate. The cutoff is chosen so that, in a
large sample, the interval contains the parameter in approximately 95% of hypothetical
experiments. It is half the 0.95 quantile of a **chi-squared distribution** with one
degree of freedom, which is the distribution of the square of a standard Normal variable.
In R, the quantiles of a chi-squared distribution are calculated with `qchisq()`:

```{r}
# Cutoff for the relative profile log-likelihood: half the 95% quantile of a
# chi-squared distribution with one degree of freedom.
qchisq(p = 0.95, df = 1)
sqrt(qchisq(p = 0.95, df = 1))
cutoff = qchisq(p = 0.95, df = 1) / 2
cutoff
```

The 0.95 quantile is 3.84, which is the square of the 1.96 of the Wald interval, so the
cutoff is 1.92. That is, an approximate 95% profile likelihood interval contains all the
values of the parameter whose relative profile log-likelihood is above $-1.92$. This is
the cutoff that we used in @sec-likelihood-surface to compare the slice and the profile.

::: {.callout-note title="Profile likelihood interval"}
An approximate 95% confidence interval for a parameter contains all the values whose
profile log-likelihood is within 1.92 units of the maximum log-likelihood (half the 0.95
quantile of a chi-squared distribution with one degree of freedom). The other parameters
of the model are re-estimated for each value.
:::

::: {.callout-caution collapse="true" icon=false title="Statistical detail: Wilks' theorem"}
The cutoff of the profile likelihood interval comes from
**Wilks' theorem**,^[Wilks (1938). The theorem is about the ratio between two maximised
likelihoods (a difference between two log-likelihoods is the logarithm of a likelihood
ratio), and it holds under the regularity conditions that asymptotic normality requires. The profile
likelihood interval follows from it.] which describes how far below its maximum the
log-likelihood is at the parameter value that generated the data. Let $\theta_0$ be that
value for the parameter of interest. Then

$$
2\left[\ell(\hat\theta)-\ell_{\mathrm{prof}}(\theta_0)\right]
\overset{\mathrm{approx}}{\sim}\chi^2_1
\quad\text{for sufficiently large }n,
$$

where $\ell(\hat\theta)$ is the maximum of the log-likelihood,
$\ell_{\mathrm{prof}}(\theta_0)$ is the profile log-likelihood at $\theta_0$ (with a single
parameter it is the log-likelihood itself) and $\chi^2_1$ is the chi-squared distribution
with one degree of freedom, which is the distribution of the square of a standard Normal
variable. In plain English, across hypothetical experiments the parameter value that
generated the data is rarely far down the slope of the profile: twice its distance to the
top behaves like the square of a standard Normal variable.^[The same result is the basis
of the likelihood ratio test, which compares two nested models by calculating a p-value
from the chi-squared distribution. As with p-values in general, we do not use it in this
course.] So in
about 95% of hypothetical experiments the profile log-likelihood at $\theta_0$ is within
$3.84/2=1.92$ units of the maximum. The profile likelihood interval contains
$\theta_0$ in exactly those experiments, which gives it a coverage of
approximately 95%.
:::

We calculate the profile likelihood interval in three steps. We first calculate it by hand
for a model with one parameter, the Poisson model, and then for a model with two
parameters, the Negative Binomial model of the owls, to see what the calculation involves.
In practice we do not do these calculations ourselves: TMB automates them for any model
that was processed with `MakeADFun()`, and the third subsection shows the two functions
that do it. From then on, this is how we calculate profile likelihood intervals.

### One parameter

The Poisson model has one parameter, so the profile is the log-likelihood itself, and the
endpoints of the interval are the two values of $\lambda$ at which the relative
log-likelihood equals $-1.92$. We calculate them by hand for the ten quadrats
(@fig-likelihood_curve). To find them we use `uniroot()`,
which searches for the value at which a function crosses zero. Its main arguments are:

- `f`, the function;
- `interval`, the range in which to search, whose two ends must give values of opposite
  sign;
- `tol`, the precision with which the crossing point is located.

It returns a list, and the crossing point is the element called `root`. We give it the
relative log-likelihood plus the cutoff, which is zero at
the endpoints of the confidence interval, positive at the estimate (where it equals 1.92)
and negative far from it. There is one crossing on each side of the estimate, so we call
`uniroot()` twice: between 0.8 and $\hat\lambda$, and between $\hat\lambda$ and 5. These
are the two ends of the curve in @fig-likelihood_curve, where the relative log-likelihood
is far below the cutoff.

```{r}
# Relative log-likelihood of the ten quadrats plus the cutoff, as a function of lambda.
# With one parameter the log-likelihood is also the profile.
ten_crossing = function(lambda) {
  sum(dpois(x = ten_counts, lambda = lambda, log = TRUE)) - loglik_lambda_hat_ten + cutoff
}

# uniroot() finds where a function crosses zero inside an interval whose ends give
# values of opposite sign: once below the estimate and once above it.
lambda_lower = uniroot(f = ten_crossing, interval = c(0.8, lambda_hat_ten), tol = 1e-8)$root
lambda_upper = uniroot(f = ten_crossing, interval = c(lambda_hat_ten, 5), tol = 1e-8)$root
c(lambda_lower, lambda_upper)
```

The likelihood interval goes from 1.40 to 3.25 seedlings per quadrat. We will not repeat
this calculation for other models: it is what `tmbprofile()` and `confint()` do for us
in the third subsection.

### Two parameters

The owl model has two parameters, so the interval of each of them is calculated from its
profile, as defined in @sec-likelihood-surface. To see what the calculation involves, we
calculate the profile of $k$ once by
hand, on a grid of values of $k$. For each value we process the NLL with $\log k$ held
fixed and fit it with `optim()`, which then re-estimates only $\log\mu$. The argument
`map = list(log_k = factor(NA))` of `MakeADFun()` keeps `log_k` fixed at the value given
in the list of parameters:

```{r}
# Profile of k by hand: for each value of k on a grid, the mean is re-estimated with k
# held fixed, starting from the joint estimate.
k_values = seq(from = 0.15, to = 0.60, by = 0.005)
profile_k = rep(NA, times = length(k_values))
for (i in seq_along(k_values)) {
  inner_obj = MakeADFun(
    cmb(nb_nll, list(calls = call_counts)),
    list(log_mu = owl_fit$par[["log_mu"]], log_k = log(k_values[i])),
    map = list(log_k = factor(NA)),
    silent = TRUE
  )
  inner_fit = optim(par = inner_obj$par, fn = inner_obj$fn, gr = inner_obj$gr, method = "BFGS")
  # Relative profile log-likelihood (both values are NLLs, so the order is reversed).
  profile_k[i] = owl_fit$value - inner_fit$value
}

# Values of k on the grid whose relative profile log-likelihood is above the cutoff.
range(k_values[profile_k >= -cutoff])
```

On this grid the values of $k$ above the cutoff go from 0.24 to 0.45, to the precision of
its step.

### Automation with TMB {#sec-profile_automation}

The two calculations above follow the same steps for every model and parameter: move the
parameter away from its estimate, re-estimate the other parameters at each step (the inner
fits), record the NLL and find the two values at which the relative profile
log-likelihood crosses the cutoff. TMB does all of this for us, so in practice we never
write the grid, the inner fits or the calls to `uniroot()` ourselves. Two functions are
enough:

- `TMB::tmbprofile()` calculates the profile of one parameter of a model that was
  processed with `MakeADFun()`. Its main arguments are the object and `name`, the name of
  the parameter to profile as it is written in the list of parameters; `trace = FALSE`
  stops the printing of progress. It returns a data frame with the values of the parameter
  and the NLL at each of them. The function belongs to the package TMB, on which RTMB is
  built and which is installed with it, so we write `TMB::tmbprofile()`.
- `confint()` takes the result of `tmbprofile()` and finds the two crossings of the
  cutoff, that is, the endpoints of the 95% profile likelihood interval.

The profile is calculated on the scale that was optimised, so we transform the endpoints
back. For the Poisson model with one parameter there are no other parameters to
re-estimate, and the result is the interval that we calculated by hand:

```{r}
# Profile of log(lambda) for the ten quadrats (trace = FALSE stops the printing of
# progress) and the endpoints of its 95% interval, transformed back.
fir_ten_profile = TMB::tmbprofile(fir_ten_obj, name = "log_lambda", trace = FALSE)
likelihood_interval = exp(confint(fir_ten_profile))
likelihood_interval
```

We now have three intervals for $\lambda$ from the ten quadrats, and we collect them in a
matrix to compare them:

```{r}
# The three intervals for lambda (seedlings per quadrat) from the ten quadrats.
intervals_ten = rbind(wald_original, wald_log, likelihood_interval)
dimnames(intervals_ten) = list(
  c("Wald, original scale", "Wald, log scale", "Likelihood"),
  c("lower", "upper")
)
intervals_ten
```

The likelihood interval is not
symmetric: it extends 0.80 below the estimate and 1.05 above it, following the asymmetry of
the curve.
It also does not depend on the scale, because a value of $\lambda$ and its logarithm have
the same likelihood, so we obtain the same interval whether we search over $\lambda$, as
we did by hand, or over $\log\lambda$, as `tmbprofile()` did. As expected (see above), the
Wald interval from the log scale is closer to it than the Wald interval on the original
scale.

We now repeat the three intervals for all 50 quadrats, with `sdreport()` and
`tmbprofile()`:

```{r}
# Wald interval on the original scale for the 50 quadrats (the interval of Chapter 3).
wald_original_all = lambda_hat + c(-1, 1) * critical_value * sqrt(lambda_hat / n_quadrats)

# Wald interval from the log scale, with the standard error from sdreport().
fir_summary = summary(sdreport(fir_obj))
wald_log_all = exp(
  fir_summary["log_lambda", "Estimate"] +
    c(-1, 1) * critical_value * fir_summary["log_lambda", "Std. Error"]
)

# Likelihood interval from the profile.
fir_profile = TMB::tmbprofile(fir_obj, name = "log_lambda", trace = FALSE)

# The three intervals for lambda (seedlings per quadrat) from the 50 quadrats.
intervals_all = rbind(wald_original_all, wald_log_all, exp(confint(fir_profile)))
dimnames(intervals_all) = dimnames(intervals_ten)
intervals_all
```

With 50 quadrats the three intervals nearly coincide: 1.73 to 2.55 seedlings per quadrat
for the Wald interval on the original scale (the interval of
[Chapter 3](../Chapter_3/Theory.qmd#sec-confidence_intervals)), 1.77 to 2.59 for the Wald
interval from the log scale and 1.76 to 2.57 for the likelihood interval. They are also
much narrower than those of the ten quadrats. The three intervals agree because the
log-likelihood of a larger sample is closer to a parabola around its maximum, so the
quadratic approximation behind the Wald interval is better. With ten quadrats the lower
endpoint was 1.28, 1.45 or 1.40 depending on the method, whereas with 50 quadrats it hardly
matters which interval we report.

For the owl model, `tmbprofile()` makes the inner fits that we wrote as a loop, with
the other parameter re-estimated at each step, and `confint()` locates the endpoints. We
use them for both parameters, in two lines each instead of the loop:

```{r}
# Profiles of log(k) and log(mu), and their 95% intervals transformed back.
owl_profile_k = TMB::tmbprofile(owl_obj, name = "log_k", trace = FALSE)
owl_profile_mu = TMB::tmbprofile(owl_obj, name = "log_mu", trace = FALSE)
owl_interval_k = exp(confint(owl_profile_k))
owl_interval_mu = exp(confint(owl_profile_mu))

# Profile intervals next to the Wald intervals.
owl_intervals = rbind(
  owl_interval_mu, owl_wald_mu, owl_interval_k, owl_wald_k, owl_wald_k_original
)
dimnames(owl_intervals) = list(
  c("mu, profile", "mu, Wald from the log scale",
    "k, profile", "k, Wald from the log scale", "k, Wald on the original scale"),
  c("lower", "upper")
)
owl_intervals
```

The profile likelihood intervals are 3.49 to 6.68 calls per visit for $\mu$ and 0.235 to
0.451 for $k$. For $k$ the Wald interval from the log scale is almost the same (0.237 to
0.453). The Wald interval on the original scale (0.221 to 0.434) is not: it is symmetric
around the estimate, whereas the profile likelihood interval extends 0.09 below the
estimate and 0.12 above it, so both of its endpoints are too low. This is why we calculate
Wald intervals on the scale that was optimised and transform their endpoints, and why the
profile likelihood interval is the reference with which we compare them. For $\mu$ the
profile likelihood interval reaches further
above the estimate than the Wald interval from the log scale (6.68 against 6.55).
@fig-profile_interval shows that the profile of $\mu$ is not a parabola on the original
scale. The parabola in the figure is the quadratic approximation on that scale, drawn with
the standard error of the mean under the fitted model,
$\sqrt{(\hat\mu+\hat\mu^2/\hat k)/n}\approx 0.78$ calls per visit (the variance of the
counts divided by the number of visits, as for any mean). To the right of the estimate the
profile falls more slowly than the parabola, and to the left it falls faster, so an
interval that is symmetric around the estimate cannot follow it.

```{r}
#| echo: false
#| label: fig-profile_interval
#| fig-cap: "Relative profile log-likelihood of the mean number of calls per visit for the owl data (solid line) and the quadratic approximation that corresponds to the Wald interval on the original scale (dashed line). The horizontal line is the cutoff of the 95% interval, at -1.92, and the vertical lines mark the endpoints of the profile likelihood interval."
# The profile of mu over a wider range than the default of tmbprofile() (ytol is the
# drop in log-likelihood at which the profile stops), as a relative log-likelihood.
profile_mu_wide = TMB::tmbprofile(owl_obj, name = "log_mu", ytol = 6, trace = FALSE)
mu_lower = owl_interval_mu[1, "lower"]
mu_upper = owl_interval_mu[1, "upper"]
k_lower = owl_interval_k[1, "lower"]
k_upper = owl_interval_k[1, "upper"]

# Quadratic approximation around the estimate, with the standard error of mu from the
# Negative Binomial variance: sqrt((mu + mu^2 / k) / n), in calls per visit.
mu_fine = seq(from = 2.5, to = 9, length.out = 131)
se_mu_nb = sqrt((mu_mle + mu_mle^2 / k_mle) / n_visits)
quadratic_values = -(mu_fine - mu_mle)^2 / (2 * se_mu_nb^2)

curve_data = rbind(
  data.frame(
    mu = exp(profile_mu_wide$log_mu),
    relative = -(profile_mu_wide$value - min(profile_mu_wide$value)),
    curve = "Profile"
  ),
  data.frame(mu = mu_fine, relative = quadratic_values, curve = "Quadratic approximation")
)
ggplot(curve_data, aes(x = mu, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -cutoff, colour = "grey40") +
  geom_vline(xintercept = c(mu_lower, mu_upper), colour = "grey40", linetype = "dotted") +
  coord_cartesian(xlim = c(2.5, 9), ylim = c(-6, 0)) +
  labs(
    x = expression(paste("Mean number of calls per visit, ", mu)),
    y = "Relative log-likelihood",
    linetype = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

## Joint confidence regions

A **joint confidence region** is the counterpart of a confidence interval for two
parameters at once. It is defined in the same way as the profile likelihood interval, with
a cutoff of 3.00 units of log-likelihood instead of 1.92 (half the 0.95 quantile of a
chi-squared distribution with two degrees of freedom, `qchisq(p = 0.95, df = 2) / 2`). The
combinations of $\mu$ and $k$ whose relative log-likelihood is above $-3.00$ form an
approximate 95% joint confidence region, which is the area inside one contour of the
likelihood surface (@fig-joint_region). The region extends beyond the profile likelihood
interval of each parameter, because it answers a question about both parameters at once:
in about 95% of hypothetical experiments it contains the pair of values that generated the
data, which is more demanding than containing one of them. We use an interval when a
question is about one parameter and a region when it is about a combination of parameters.

```{r}
#| echo: false
#| label: fig-joint_region
#| fig-cap: "Approximate joint 95% confidence region for the mean and the shape of the Negative Binomial model fitted to the owl data (contour in vermillion). The segments along the axes are the 95% profile likelihood intervals of each parameter."
# Cutoff for a joint region of two parameters: half the 95% quantile of a chi-squared
# distribution with two degrees of freedom (qchisq(p = 0.95, df = 2) is 5.99).
joint_cutoff = qchisq(p = 0.95, df = 2) / 2

# Contours of the surface, the contour of the joint region, the estimate and the
# profile intervals of the two parameters drawn along the axes.
ggplot() +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = c(-12, -9, -6, -4, -2, -1),
    colour = "grey40"
  ) +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = -joint_cutoff,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  geom_point(
    data = data.frame(mu = mu_mle, k = k_mle),
    aes(x = mu, y = k),
    colour = "#D55E00",
    size = 3
  ) +
  annotate(
    "segment",
    x = mu_lower, xend = mu_upper, y = min(k_grid), yend = min(k_grid),
    linewidth = 1.2
  ) +
  annotate(
    "segment",
    x = min(mu_grid), xend = min(mu_grid), y = k_lower, yend = k_upper,
    linewidth = 1.2
  ) +
  labs(
    x = expression(paste("Mean number of calls per visit, ", mu)),
    y = expression(paste("Shape, ", italic(k)))
  ) +
  theme_classic()
```

::: {.callout-caution collapse="true" icon=false title="Statistical detail: the cutoff for a joint region"}
Wilks' theorem also applies to several parameters at once. If we fix two parameters at the
values that generated the data, twice the drop in log-likelihood from the maximum follows
approximately a chi-squared distribution with two degrees of freedom (the distribution of
the sum of the squares of two independent standard Normal variables). The cutoff for the
joint region is half its 0.95 quantile (5.99), `qchisq(p = 0.95, df = 2) / 2`.

A profile likelihood interval is a statement about one parameter, whatever the value of
the other: it is the range that the contour at $-1.92$ covers along one axis. Capturing
two parameters simultaneously is more demanding than capturing one, so the cutoff of the
joint region is further from the maximum, and the rectangle formed by the two separate
intervals is not a 95% region for the pair.
:::

## Coverage checked by simulation

As in [Chapter 3](../Chapter_3/Theory.qmd#sec-confidence_intervals), we check the coverage
of the intervals with a simulation in which we know the parameters. To simulate from a
model that is written as an NLL, we mark the data inside it with `OBS()`. The object
returned by `MakeADFun()` then has a function `simulate()`, which draws one hypothetical
experiment of the size of the data for the parameter values that we give it, on the scale
that was optimised. With this line the NLL of the owl model is complete:

```{r}
# The complete NLL of the Negative Binomial model.
nb_nll = function(parms, data) {
  getAll(parms, data)
  # Mark the data as observations, so that they can be simulated.
  calls = OBS(calls)
  mu = exp(log_mu)
  k = exp(log_k)
  ADREPORT(k)
  -sum(dnbinom2(x = calls, mu = mu, size = k, log = TRUE))
}
```

We process and fit it as before, and the estimates do not change:

```{r}
# Process and fit as before (the estimates do not change).
owl_obj = MakeADFun(
  cmb(nb_nll, list(calls = call_counts)),
  list(log_mu = log(mu_hat), log_k = log(k_mom)),
  silent = TRUE
)
owl_fit = optim(par = owl_obj$par, fn = owl_obj$fn, gr = owl_obj$gr, method = "BFGS")
```

We give `simulate()` the estimates on the log scale, which are in the element `par` of the
fit. It returns a list in which the simulated data have the name that they have in the NLL
(`calls`). For example, these are the first 20 counts of one hypothetical experiment of 122
visits:

```{r}
# One hypothetical experiment of 122 visits at the estimates: its first 20 counts.
set.seed(seed = 700)
owl_obj$simulate(owl_fit$par)$calls[1:20]
```

For each hypothetical experiment we calculate both types of interval and record whether
each of them contains the parameter used to simulate (the code is not shown; it repeats
the loop of @sec-mle_properties). We consider two cases with small samples, where the
large-sample results are most at risk (@fig-coverage).

The first case uses a Poisson model and compares the Wald interval on the original scale
with the likelihood interval, with 5,000 hypothetical experiments. Chapter 3 warned that
the Wald interval does badly when the expected total count $n\lambda$ is small, so we take
$\lambda=0.5$ and 10 quadrats, an expected total of five seedlings. The coverage of the
Wald interval is 0.878, far from the intended 0.95, and that of the likelihood interval is
0.924.^[A sample in which all the quadrats are empty
gives an estimate of zero. The Wald interval on the original scale then collapses to
$[0,0]$, and a likelihood interval can still be calculated at the boundary (it runs from
zero to the value at which the log-likelihood has dropped by 1.92). The simulation counts
such a sample as a miss for both intervals; it happens in less than 1% of these
experiments.]

The second case uses the Negative Binomial model with the owl estimates as parameters and
20 visits per experiment. It compares, for the shape $k$, the Wald interval calculated on
the log scale with the profile likelihood interval, with 2,000 hypothetical experiments.
The coverages are 0.945 and 0.942, so here the two intervals perform equally well: a
coverage simulated with 2,000 experiments has itself a standard error of about 0.005
($\sqrt{0.95\times 0.05/2000}$), so a difference of this size is within the noise of the
simulation.

```{r}
#| include: false
#| warning: false
# Poisson model: coverage of the Wald interval on the original scale and of the
# likelihood interval, for a mean of 0.5 seedlings per quadrat and 10 quadrats.
poisson_sim_nll = function(parms, data) {
  getAll(parms, data)
  counts = OBS(counts)
  -sum(dpois(x = counts, lambda = exp(log_lambda), log = TRUE))
}
set.seed(seed = 701)
lambda_sim = 0.5
n_quadrats_sim = 10
# Placeholder data of the right length: the values are not used.
poisson_sim_obj = MakeADFun(
  cmb(poisson_sim_nll, list(counts = rep(0, times = n_quadrats_sim))),
  list(log_lambda = 0),
  silent = TRUE
)
# The estimate is the mean of the counts, so only the total of each experiment is kept.
totals = rep(NA, times = 5000)
for (e in seq_along(totals)) {
  totals[e] = sum(poisson_sim_obj$simulate(log(lambda_sim))$counts)
}
estimates = totals / n_quadrats_sim

# Wald interval on the original scale.
se = sqrt(estimates / n_quadrats_sim)
wald_covered = estimates - critical_value * se <= lambda_sim &
  lambda_sim <= estimates + critical_value * se

# The likelihood interval contains lambda when its relative log-likelihood is within
# the cutoff. A sample of zeros has no finite estimate of the interval: a miss.
relative = -n_quadrats_sim * (lambda_sim - estimates) + totals * log(lambda_sim / estimates)
likelihood_covered = totals > 0 & relative >= -cutoff
coverage_poisson_small = c(wald = mean(wald_covered), likelihood = mean(likelihood_covered))
coverage_poisson_small
mean(totals == 0)

# Negative Binomial model (owl estimates), 20 visits: coverage of the Wald interval for k
# on the log scale and of the profile interval. The profile interval contains the
# generating k when the relative profile log-likelihood at that k is within the cutoff,
# which takes one extra fit with log(k) fixed in place of a whole profile.
set.seed(seed = 702)
n_visits_sim = 20
nb_sim_obj = MakeADFun(
  cmb(nb_nll, list(calls = rep(0, times = n_visits_sim))),
  list(log_mu = 0, log_k = 0),
  silent = TRUE
)
wald_covered = rep(NA, times = 2000)
profile_covered = rep(NA, times = 2000)
for (e in seq_along(wald_covered)) {
  sim_calls = nb_sim_obj$simulate(owl_fit$par)$calls
  # Starting values: the method-of-moments estimates when they exist.
  sim_mu = mean(sim_calls)
  sim_variance = mean((sim_calls - sim_mu)^2)
  k_start = ifelse(sim_variance > sim_mu, sim_mu^2 / (sim_variance - sim_mu), 1)
  sim_obj = MakeADFun(
    cmb(nb_nll, list(calls = sim_calls)),
    list(log_mu = log(sim_mu), log_k = log(k_start)),
    silent = TRUE
  )
  sim_fit = optim(par = sim_obj$par, fn = sim_obj$fn, gr = sim_obj$gr, method = "BFGS")
  se_log_k = tryCatch(
    summary(sdreport(sim_obj))["log_k", "Std. Error"],
    error = function(err) NA
  )
  if (sim_fit$convergence == 0 && !is.na(se_log_k)) {
    wald_covered[e] = abs(log(k_mle) - sim_fit$par[["log_k"]]) <= critical_value * se_log_k
    fixed_obj = MakeADFun(
      cmb(nb_nll, list(calls = sim_calls)),
      list(log_mu = log(sim_mu), log_k = log(k_mle)),
      map = list(log_k = factor(NA)),
      silent = TRUE
    )
    fixed_fit = optim(par = fixed_obj$par, fn = fixed_obj$fn, gr = fixed_obj$gr, method = "BFGS")
    profile_covered[e] = sim_fit$value - fixed_fit$value >= -cutoff
  }
}
coverage_nb_small = c(
  wald = mean(wald_covered, na.rm = TRUE),
  profile = mean(profile_covered, na.rm = TRUE),
  failed = sum(is.na(wald_covered))
)
coverage_nb_small

# Collect the results for the figure.
coverage_results = data.frame(
  case = rep(
    c(
      "Poisson, λ = 0.5,\n10 quadrats",
      "Negative Binomial, k,\n20 visits"
    ),
    each = 2
  ),
  interval = rep(c("Wald", "Likelihood"), times = 2),
  coverage = c(
    coverage_poisson_small[["wald"]], coverage_poisson_small[["likelihood"]],
    coverage_nb_small[["wald"]], coverage_nb_small[["profile"]]
  )
)
coverage_results$case = factor(coverage_results$case, levels = unique(coverage_results$case))
coverage_results
```

```{r}
#| echo: false
#| label: fig-coverage
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Simulated coverage of nominal 95% Wald and likelihood intervals. Poisson case: 5,000 hypothetical experiments, interval for the mean on the original scale. Negative Binomial case: 2,000 hypothetical experiments, interval for the shape k, with the Wald interval calculated on the log scale. The dashed line is the nominal level."
# Coverage by case and type of interval; the dashed line is the intended 0.95.
ggplot(coverage_results, aes(x = case, y = coverage, shape = interval, linetype = interval)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey40") +
  geom_point(
    position = position_dodge(width = 0.4),
    colour = "#D55E00",
    size = 3
  ) +
  scale_linetype_manual(values = c("solid", "solid")) +
  coord_cartesian(ylim = c(0.8, 1)) +
  labs(x = NULL, y = "Simulated coverage", shape = NULL) +
  guides(linetype = "none") +
  theme_classic() +
  theme(legend.position = "top")
```

So the Wald interval fails on the original scale when the counts are few, where the
log-likelihood is far from a parabola, and the likelihood interval is clearly better
there. A Wald interval calculated on a suitable scale, such as the logarithm of a positive
parameter, can be just as good as the profile likelihood interval, and it needs one fit
instead of many. In the practical you will run a simulation
of this kind for another model.

# Properties of maximum likelihood estimators {#sec-mle_properties}

An MLE is an estimator, so everything that we learnt in
[Chapter 3](../Chapter_3/Theory.qmd#sec-sampling_distribution) applies to it: it has a
sampling distribution across hypothetical experiments, a bias, a variance and a standard
error. In a finite sample we do not know that sampling distribution in general. For large
samples, however, there are results that hold for the MLEs of a very wide class of
models, and they are the reason why this book uses maximum
likelihood. The standard errors and the intervals of @sec-likelihood-uncertainty relied on
them. This section shows three of these properties at work (consistency, asymptotic
normality and efficiency) and then looks at what happens in small samples and when the model
is not a good description of the data generating process.

We can check all these properties with the same simulation. As in the Chapter 3 practical, we simulate
hypothetical experiments from the model using the plug-in estimates: a Negative Binomial model
with the maximum likelihood estimates for the owls as parameters ($\mu=4.75$ and
$k=0.327$). We simulate 2,000 hypothetical experiments for each of three sample sizes: 20,
122 (the size of the owl sample) and 500 visits. The function `simulate()` of an object
draws experiments of the size of its data, so for each sample size we process the NLL
with placeholder data of that length, whose values are not used. Each simulated sample is
then fitted like the data, with `cmb()`, `MakeADFun()` and `optim()`, and we keep the
maximum likelihood estimate of $k$ and the method-of-moments estimate:

```{r}
#| warning: false
# Hypothetical experiments from a Negative Binomial model whose parameters are the
# maximum likelihood estimates for the owls, for three numbers of visits.
set.seed(seed = 601)
n_experiments = 2000
visit_sizes = c(20, 122, 500)
sim_k_ml = matrix(NA, nrow = n_experiments, ncol = length(visit_sizes))
sim_k_mom = matrix(NA, nrow = n_experiments, ncol = length(visit_sizes))
for (s in seq_along(visit_sizes)) {
  # Placeholder data of the right length: the values are not used.
  placeholder_obj = MakeADFun(
    cmb(nb_nll, list(calls = rep(0, times = visit_sizes[s]))),
    list(log_mu = 0, log_k = 0),
    silent = TRUE
  )
  for (e in seq_len(n_experiments)) {
    # One hypothetical experiment at the estimates for the owls.
    sim_calls = placeholder_obj$simulate(owl_fit$par)$calls

    # Method of moments: k exists only when the variance is larger than the mean.
    sim_mu = mean(sim_calls)
    sim_variance = mean((sim_calls - sim_mu)^2)
    if (sim_variance > sim_mu) {
      sim_k_mom[e, s] = sim_mu^2 / (sim_variance - sim_mu)
    }

    # Maximum likelihood, starting from the method-of-moments estimates (or k = 1).
    k_start = ifelse(is.na(sim_k_mom[e, s]), 1, sim_k_mom[e, s])
    sim_obj = MakeADFun(
      cmb(nb_nll, list(calls = sim_calls)),
      list(log_mu = log(sim_mu), log_k = log(k_start)),
      silent = TRUE
    )
    sim_fit = optim(par = sim_obj$par, fn = sim_obj$fn, gr = sim_obj$gr, method = "BFGS")
    sim_k_ml[e, s] = exp(sim_fit$par[["log_k"]])
  }
}
```

## Consistency

An estimator is **consistent** when its estimates get closer to the parameter as the
sample grows: the chance of missing the parameter by more than any given margin becomes
small ([Chapter 3](../Chapter_3/Theory.qmd#sec-large_samples) introduced this property).
Maximum likelihood estimators are consistent when the observations are i.i.d. from the
model. In the simulation we can calculate the chance that the estimate misses the shape
$k$ by a margin of 0.1 or more. The matrix `sim_k_ml` has one column per sample size, and
`colMeans()` calculates the mean of each column. As in Chapter 3, the mean of `TRUE` and
`FALSE` values is the fraction of `TRUE` values:

```{r}
# Fraction of the estimates that differ by 0.1 or more from the shape used to simulate,
# for 20, 122 and 500 visits.
colMeans(abs(sim_k_ml - k_mle) >= 0.1)
```

With 20 visits, 45% of the estimates differ from the $k$ used to simulate by 0.1 or more.
With 122 visits it is 7%, and with 500 visits none of the 2,000 estimates does. For your
own analysis, consistency means that more data bring the estimate closer to the parameter
of the model, so a larger sample is worth its cost. How close the estimate from a sample
of a given size is likely to be is what the standard error tells us.

::: {.callout-caution collapse="true" icon=false title="Statistical detail: consistency"}
The box on consistency in [Chapter 3](../Chapter_3/Theory.qmd#sec-large_samples) defined a
consistent estimator as one that converges in probability to the parameter as the sample
size grows. Maximum
likelihood estimators are consistent when the observations are i.i.d. from the
model.^[The consistency of maximum likelihood estimators was proved rigorously, under
regularity conditions, by Abraham Wald (1949).] For
the shape of the Negative Binomial model, with $\hat k_n$ the MLE calculated from $n$
visits and any positive margin $\varepsilon$:

$$
\lim_{n \to \infty}
P\left(\left|\hat k_n - k\right| \ge \varepsilon\right)=0.
$$

In plain English, the chance that the estimate misses $k$ by $\varepsilon$ or more becomes
small as we observe more visits.

Why should the maximum of the likelihood be close to the parameter in a large sample?
Chapter 3 gives us an intuition. If we divide the log-likelihood by $n$, which does not
move its maximum, we obtain

$$
\frac{1}{n}\ell(\theta)=\frac{1}{n}\sum_{i=1}^{n}\log P(X_i=x_i\mid\theta),
$$

the mean of $\log P(X=x\mid\theta)$ over the sample. This is an expectation under the
empirical distribution of the sample
([Chapter 3](../Chapter_3/Theory.qmd#sec-samples)). As the sample grows, the empirical
distribution converges to the distribution that generated the data (the Glivenko–Cantelli
theorem), so the function that we maximise converges to the expectation of
$\log P(X=x\mid\theta)$ under that distribution. It can be shown that this expectation is
highest when $\theta$ is the parameter that generated the data (the box on the
Kullback–Leibler divergence in @sec-misspecified says why).
:::

## Asymptotic normality

In @sec-likelihood-uncertainty we stated that, in a large sample, the sampling
distribution of an MLE is approximately Normal and centred on the parameter, with a
variance equal to the inverse of the Fisher information, and we used that result to
calculate standard errors and Wald intervals. How large the sample needs to be depends on the model and on the parameter, so we look at
the simulation (@fig-sampling_distribution_k). With 20 visits the estimates of $k$ are
strongly skewed. Half of them are below 0.36, but a long tail to the right reaches values
above 1, and their mean is 0.40, against the 0.327 used to simulate. A Normal curve with
the same mean and standard deviation is a poor description of this distribution. With 500
visits the estimates have a mean of 0.329, their distribution is nearly symmetric and the
Normal curve is close to it.

```{r}
#| echo: false
#| label: fig-sampling_distribution_k
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Maximum likelihood estimates of the shape k from 2,000 hypothetical experiments with 20 and with 500 visits, simulated from a Negative Binomial model with the estimates for the owls as parameters. The vertical line is the shape used in the simulation and the curves are Normal distributions with the mean and standard deviation of the estimates."
library(patchwork)

# Histogram of the estimates for one number of visits, with its Normal approximation.
plot_sampling_k = function(column, bin_width) {
  estimates = sim_k_ml[, column]
  estimate_grid = seq(from = min(estimates), to = max(estimates), length.out = 200)
  approximation = data.frame(
    estimate = estimate_grid,
    density = dnorm(x = estimate_grid, mean = mean(estimates), sd = sd(estimates))
  )
  ggplot(data.frame(estimate = estimates), aes(x = estimate)) +
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = bin_width,
      fill = "grey90",
      colour = "grey40"
    ) +
    geom_line(
      data = approximation,
      aes(x = estimate, y = density),
      colour = "#D55E00",
      linewidth = 0.8
    ) +
    geom_vline(xintercept = k_mle, linetype = "dashed") +
    labs(
      title = paste(visit_sizes[column], "visits"),
      x = expression(paste("Estimated shape, ", italic(k))),
      y = "Density"
    ) +
    theme_classic()
}

plot_sampling_k(column = 1, bin_width = 0.04) |
  plot_sampling_k(column = 3, bin_width = 0.01)
```

The simulation also allows us to check the standard error that we calculated from the
Hessian of the owl fit. The hypothetical experiments with 122 visits have the sample size
of the owl data and were simulated with the estimates of that fit, so the standard
deviation of their estimates of $\log k$ is the quantity that the standard error of the
estimate of $\log k$ approximates:

```{r}
# Standard deviation of the estimates of log(k) across the hypothetical experiments
# with 122 visits (second column of sim_k_ml), to compare with the standard error of
# log(k) obtained from the Hessian of the owl fit (0.165).
sd(log(sim_k_ml[, 2]))
```

The standard deviation of the estimates of $\log k$ across the 2,000 hypothetical
experiments with 122 visits is 0.162, close to the 0.165 that `sdreport()` gave us from
a single sample. For your own analysis this means that, with a sample of this size, the
standard error from one fit describes the variation across hypothetical experiments well.
With 20 visits the sampling distribution of $\hat k$ is far from Normal, which is why the
scale matters in a small sample: for the Wald interval calculated on the log scale, the
coverage with 20 visits was close to the intended 95% (@fig-coverage).

## Efficiency

An estimator is more **efficient** than another estimator of the same parameter when its
estimates vary less across hypothetical experiments. This is the property that separates
maximum likelihood from the method of moments. The two methods use different
information from the sample. The method of moments chooses the
parameters that reproduce the mean and the variance of the sample. Maximum likelihood
chooses the parameters under which the observed counts are most probable, so every
count contributes with the probability that the model assigns to it. Both estimators of
$k$ are consistent, so with a very large sample from the model they would give nearly the
same value. However, they are not equally precise.

The mean squared error (MSE) of
[Chapter 3](../Chapter_3/Theory.qmd#sec-sampling_distribution) allows us to compare
estimators of the same parameter. In large samples, where the bias is negligible, this
comes down to comparing their variances. We compare the two estimators of $k$ with the
estimates stored in the simulation. The method of moments gives no estimate of $k$ when
the variance of a sample is not larger than its mean, and the simulation left an `NA` in
that case, so we first count the samples in which it happened:

```{r}
# Number of simulated samples without a method-of-moments estimate of k.
sum(is.na(sim_k_mom))
```

For each number of visits we then calculate the mean, the standard deviation and the MSE
of both estimators, and collect them in a table with one row per number of visits. In its
column names, `ml` stands for maximum likelihood and `mom` for the method of moments:

```{r}
# Mean, standard deviation and mean squared error of both estimators of k for each
# number of visits. Column s of sim_k_ml and sim_k_mom holds the 2,000 estimates for
# the s-th number of visits.
efficiency_table = data.frame(
  visits = visit_sizes,
  mean_ml = NA, sd_ml = NA, mse_ml = NA,
  mean_mom = NA, sd_mom = NA, mse_mom = NA
)
for (s in seq_along(visit_sizes)) {
  ml = sim_k_ml[, s]
  mom = sim_k_mom[, s]
  efficiency_table$mean_ml[s] = mean(ml)
  efficiency_table$sd_ml[s] = sd(ml)
  efficiency_table$mse_ml[s] = mean((ml - k_mle)^2)
  efficiency_table$mean_mom[s] = mean(mom)
  efficiency_table$sd_mom[s] = sd(mom)
  efficiency_table$mse_mom[s] = mean((mom - k_mle)^2)
}

# Show the table with three significant digits.
# ml = maximum likelihood, mom = method of moments
signif(efficiency_table, digits = 3)
```

With 122 visits the standard deviation of the estimates (i.e., standard error) is 0.055 for
maximum likelihood against 0.086 for the method of moments, and the MSE is 0.0031
against 0.0085. The difference does not disappear in a larger sample: with 500 visits the
standard errors are 0.024 and 0.042, respectively. Standard errors
decrease with the square root of the sample size, so the method of moments would need
about $(0.086/0.055)^2\approx 2.5$ times as many visits to estimate $k$ with the precision
that maximum likelihood achieves with 122. With 20 visits maximum likelihood also has the
smaller MSE (0.044 against 0.117).

What the simulation shows for this model is a general property. An MLE is **asymptotically
efficient**: in large samples no other regular estimator of the same parameter is more
precise. The variance that an MLE approaches as the sample grows, the inverse of the Fisher
information (@sec-likelihood-uncertainty), is the smallest that such an estimator can
have. For your own analysis, this is the reason to fit a model by maximum likelihood
whenever you can write its likelihood: with a large sample, no other method of estimation
obtains more precise estimates from the same data.

::: {.callout-caution collapse="true" icon=false title="Statistical detail: the Cramér–Rao bound"}
How small can the variance of an estimator be?
The **Cramér–Rao bound** states that no unbiased estimator has a variance smaller than the
inverse of the Fisher information:^[The bound was derived independently in the 1940s by
Fréchet, Darmois, Rao and Cramér, and it carries the names of the last two. Fisher had
already obtained a version for large samples in 1925.]

$$
\operatorname{Var}(\tilde\theta)\ge\frac{1}{\mathcal I(\theta)}
\quad\text{for any unbiased estimator }\tilde\theta.
$$

In plain English, the information in a sample sets a limit to the precision with which a
parameter can be estimated from it, whatever the method. The variance of an MLE approaches
this limit as the sample grows, which is what asymptotic efficiency means.
:::

## Bias in finite samples

Asymptotic normality tells us that the sampling distribution of an MLE is centred on the
parameter in a large sample. In a small sample this need not be the case, and the table
above already suggests it: the mean of the estimates of $k$ from 20 visits is 0.40. The
bias relative to the $k$ used to simulate is:

```{r}
# Relative bias of the maximum likelihood estimator of k for each number of visits.
data.frame(
  visits = efficiency_table$visits,
  relative_bias = efficiency_table$mean_ml / k_mle - 1
)
```

With 20 visits, $\hat k$ is on average 21% larger than the $k$ used to simulate. With 122
visits the bias is 2.6% and with 500 visits 0.6%. This is the general behaviour: MLEs are
in general biased in finite samples, and the bias decreases as the sample grows.^[Corrections
of the bias exist for particular models. Dividing by $n-1$ in the estimator of a variance
is one of them ([Chapter 3](../Chapter_3/Theory.qmd#sec-two_parameters)). Restricted
maximum likelihood (REML) is the general approach to reduce the bias of the estimates of
variance parameters, and it is used with the models for grouped observations of
[Chapter 7](../Chapter_7/Theory.qmd).] It decreases faster than the standard error, so in
a large sample it is negligible compared with the variation of the estimates across
hypothetical experiments. The warning about finite samples in
[Chapter 3](../Chapter_3/Theory.qmd#sec-large_samples) applies to maximum likelihood as
much as to the method of moments: for a small sample the properties of this section are
approximations, and simulation is the way to check them.

RTMB has a function that runs such a simulation for a fitted model, `checkConsistency()`.
It draws `n` hypothetical experiments of the size of the data from the model at its
estimates, using the data marked with `OBS()`, and with `estimate = TRUE` it fits the
model to each of them. `summary()` of its result holds the estimates of every experiment
on the scale that was optimised, from which we calculate the relative bias of $\hat k$:

```{r}
#| warning: false
# 2,000 hypothetical experiments of 122 visits at the estimates, each of them fitted.
set.seed(seed = 801)
owl_check = checkConsistency(owl_obj, n = 2000, estimate = TRUE)

# Estimates of log(mu) and log(k) in every experiment, and the relative bias of k.
check_estimates = summary(owl_check)$estimate$par
mean(exp(check_estimates$log_k)) / k_mle - 1
```

The relative bias is 3.0%, close to the 2.6% of our own loop for 122 visits (the two
numbers come from different sets of 2,000 experiments). `checkConsistency()` also reports
a p-value for a test that the simulation and the NLL agree, which we do not use here.

## Misspecified models {#sec-misspecified}

This section can be skipped at a first reading: the rest of the chapter does not depend on
it. All the properties above assume that the observations are a random sample from the model.
[Chapter 1](../Chapter_1/Theory.qmd) described the model as an approximation of the data
generating process, so this assumption is never exactly right. A model is **misspecified**
when it does not contain the distribution of the data generating process, whatever the
values of its parameters. What does the MLE estimate in that case?

The answer is that, as the sample grows, the MLE converges to the parameter values for
which the model distribution is closest to the true distribution of the data generating
process. That is, the fitted model is the best approximation that the model can offer, and
the estimates are the parameters of that approximation (the box at the end of this section
says how closeness is measured).

For the owls there are two signs that the Negative Binomial model with i.i.d. visits is an
approximation. The first is in the fitted model itself. We saw in @sec-likelihood-surface
that it reproduces the fraction of visits without calls. We can also calculate the
variance that it implies, $\hat\mu+\hat\mu^2/\hat k$, and compare it with the variance of
the sample:

```{r}
# Variance implied by the maximum likelihood fit, against the variance of the sample.
c(implied = mu_mle + mu_mle^2 / k_mle, sample = variance_hat)
```

The fitted model implies a variance of about 74, much larger than the variance of the
sample (42). The model has two parameters, and no pair of values reproduces the mean, the
variance and the fraction of visits without calls of this sample at the same time. Maximum
likelihood settles for the compromise under which the 122 observed counts are most
probable.

This is also the reason for a difference that we left open in @sec-likelihood-surface: for
the shape of the owl model the method of moments gave 0.60 and maximum likelihood 0.33, a
gap of 0.27. The two estimators are not equally precise (see "Efficiency" above), so some
difference between them is expected in any sample. To see how large a difference to expect
when the model generated the data, we use the simulation of this section, in which the
hypothetical experiments with 122 visits were simulated from the fitted model:

```{r}
# Observed gap between the method-of-moments and the maximum likelihood estimates of k.
observed_gap = k_mom - k_mle
observed_gap

# Difference between the two estimates in each of the 2,000 hypothetical experiments with
# 122 visits (second column of sim_k_mom and sim_k_ml), its largest value, and the
# fraction of experiments in which it is at least as large as the observed gap.
simulated_gap = sim_k_mom[, 2] - sim_k_ml[, 2]
max(simulated_gap)
mean(simulated_gap >= observed_gap)
```

In none of the 2,000 hypothetical experiments does the method-of-moments estimate exceed
the maximum likelihood estimate by 0.27 or more (the largest difference is 0.23). A gap of
this size is therefore unusual when the model generated the data, so the two methods
disagree mainly because the model is an approximation, and each of them settles for a
different compromise. To see the compromise of the method of moments, we calculate the
probability of a visit without calls that its estimates imply:

```{r}
# Fraction of visits without calls under the method-of-moments estimates.
dnbinom2(x = 0, mu = mu_hat, size = k_mom)
```

The method of moments reproduces the mean and the variance of the
sample by construction, but its estimates imply that 27% of the visits have no calls,
against 44% in the sample. Maximum likelihood reproduces the mean and comes close to that
fraction (41%), and implies too large a variance.

The second sign is the design of the study. The 122 visits come from a limited number of
nests:

```{r}
# The visits come from a limited number of nests.
length(unique(satiated_female$Nest))
```

There are 24 nests, and visits to the same nest are likely to resemble each other, so the
visits are not independent. The Chapter 3 practical already described the i.i.d. assumption
as a simplification for these data, and the models for observations nested in groups are
the subject of [Chapter 7](../Chapter_7/Theory.qmd). The estimates that we obtained are
still the best that this model can do for these data. The standard errors and confidence
intervals of @sec-likelihood-uncertainty are calculated as if the model had generated the data, so
with a misspecified model they are approximations too.

::: {.callout-caution collapse="true" icon=false title="Statistical detail: the Kullback–Leibler divergence"}
The closeness between the model and the data generating process is measured by the
**Kullback–Leibler divergence**,^[Kullback and Leibler (1951) introduced this measure of "information for
discrimination" between two distributions, which is now called the Kullback–Leibler
divergence. Maximum likelihood estimation under a misspecified model is studied by White
(1982), "Maximum likelihood estimation of misspecified models", *Econometrica*; Huber
(1967) is an earlier treatment.] which for a discrete variable is

$$
D_{\mathrm{KL}}(\theta)
= E_g\left[\log\frac{g(X)}{P(X\mid\theta)}\right]
= \sum_x g(x)\log\frac{g(x)}{P(X=x\mid\theta)},
$$

where $g(x)$ is the probability with which the data generating process produces the value
$x$, $P(X=x\mid\theta)$ is the probability that the model assigns to it, and $E_g$ is an
expectation ([Chapter 2](../Chapter_2/Theory.qmd#sec-expectations)) calculated with the
probabilities $g$. In plain English, the divergence is the average, over the values that
the data generating process produces, of the logarithm of the ratio between the
probability of a value and the probability that the model gives to it. It is zero when the
model assigns the right probability to every value, and positive otherwise.

This completes the intuition given in the box on consistency. The expectation of
$\log P(X\mid\theta)$, which the log-likelihood divided by $n$ approaches in a large
sample, is the only part of the divergence that depends on $\theta$, and it enters with a
minus sign. Maximising the likelihood therefore amounts to minimising the divergence
between the model and the data generating process. When the model contains the
distribution that generated the data, the minimum is zero and it is reached at the
parameter that generated them, which is consistency. When it does not, the MLE settles on
the best approximation that the model can offer.
:::

# Model comparison with AIC {#sec-aic}

More than one model can be plausible for the same data.
[Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions) gave heuristics to choose
a distribution and announced formal tools, and the likelihood provides the first of them.
We have compared parameter values of one model through their log-likelihoods, so it is
natural to compare two models through their maximum log-likelihoods. That comparison is
not fair when the models have different numbers of parameters. A model with more
parameters has more freedom to adapt to the particular sample that we collected, including
its accidental features. When one model contains the other as a special case (the Negative
Binomial distribution approaches the Poisson distribution as $k$ grows), its maximum
log-likelihood can never be lower.

The **Akaike information criterion** (AIC) corrects the maximum log-likelihood for the
number of parameters:

$$
\mathrm{AIC}=2K-2\,\ell(\hat\theta),
$$

where $\ell(\hat\theta)$ is the maximum log-likelihood of the model and $K$ is the number
of parameters that were estimated (we write $K$ because $k$ is the shape of the Negative
Binomial distribution). The log-likelihood enters with a minus sign, so among models
fitted to the same data the one with the lowest AIC is preferred. Like the value of a
log-likelihood, the AIC of one model means nothing by itself: only differences between
models fitted to the same observations, in the same units, are meaningful.

A model is an approximation of the data generating process
([Chapter 1](../Chapter_1/Theory.qmd)), and the criterion measures how good that
approximation is. Apart from a constant that is
the same for all the models, AIC estimates how far the fitted model is from the data
generating process when the fitted model is used to predict new data from the same
process. The distance is the Kullback–Leibler divergence, which is defined in the box
"Statistical detail" of @sec-misspecified.^[Akaike (1973, 1974) derived the criterion
as an estimate of a loss based on the Kullback–Leibler divergence. Reading the term $2K$
as a correction for using the same data twice is the usual interpretation of the
criterion.] The maximum log-likelihood alone is too optimistic for that purpose, because
the same data were used to fit the model and to evaluate it, and the term $2K$ corrects
for this. In plain English, AIC rewards a model for assigning a high probability to the
data and penalises it for each parameter that it needed to do so.

For the owls we have fitted two models: the Poisson model, with one parameter, and the
Negative Binomial model, with two. We collect the number of parameters and the maximum
log-likelihood of each model in a table (for a model fitted with `optim()`, the maximum
log-likelihood is minus the element `value` of the fit) and calculate the AIC from them.
Only differences in AIC matter, so we also calculate the difference between each model and
the one with the lowest AIC:

```{r}
# Maximised log-likelihood, number of estimated parameters K and AIC for the two models
# fitted to the owl data. AIC = 2K - 2 * log-likelihood.
aic_table = data.frame(
  model = c("Poisson", "Negative Binomial"),
  K = c(1, 2),
  loglik = c(loglik_poisson, -owl_fit$value)
)
aic_table$aic = 2 * aic_table$K - 2 * aic_table$loglik

# Only differences in AIC matter: difference from the lowest value.
aic_table$delta_aic = aic_table$aic - min(aic_table$aic)
aic_table
```

The difference of 667 units in favour of the Negative Binomial model is very large
compared with the two units that its additional parameter costs, in agreement with what
@fig-owl_fits showed.

The justification of AIC is asymptotic, and when the number of observations $n$
is small compared with the number of parameters $K$ the term $2K$ is too small to properly
penalise complex models: the criterion then tends to prefer models with too many parameters.
The **corrected AIC**
(AICc) adds a penalty that is large when $n$ is close to $K$ and disappears as $n$ grows:^[The
correction was derived by Sugiura (1978) for linear regression with Normal errors and
extended by Hurvich and Tsai (1989), "Regression and time series model selection in small
samples", *Biometrika*. Burnham and Anderson (2002), *Model selection and multimodel
inference*, recommend AICc whenever $n/K$ is smaller than about 40.]

$$
\mathrm{AICc}
=\mathrm{AIC}+\frac{2K(K+1)}{n-K-1}
=2K\,\frac{n}{n-K-1}-2\,\ell(\hat\theta),
$$

where $n$ is the number of observations and $K$ and $\ell(\hat\theta)$ are the same as in
AIC. The correction needs $n>K+1$. In plain English, each parameter costs $2n/(n-K-1)$
units instead of two: slightly more than two in a moderate sample, and a lot more when the
sample has hardly more observations than parameters. AICc is used in the same way as AIC,
as a difference between models fitted to the same observations.

The expression above comes from
linear regression with Normal errors, and for the non-linear models with other
distributions that we use in this course it is an approximation: there is no general
formula for the small-sample bias of AIC. AICc is nevertheless common in the ecological
literature, where samples are often small, and a correction that is approximately right
is better than none, so you will meet it often. For the owls the correction is negligible:

```{r}
# Additional penalty of AICc, 2K(K + 1) / (n - K - 1), for the Poisson (K = 1) and
# Negative Binomial (K = 2) models, with n the number of visits.
k_models = c(1, 2)
2 * k_models * (k_models + 1) / (n_visits - k_models - 1)
```

It is about 0.03 units for the Poisson model and 0.10 for the Negative Binomial model,
compared with a difference of 667 units between them, because the 122 visits are many more
than the one or two parameters.

AIC ranks the candidate models that we give it. It does not tell us whether the best of
them describes the data well: the Negative Binomial model is far better than the Poisson
model for the owls, and it is still an approximation of the data generating process (for
one thing, we treated visits to the same nest as i.i.d.). [Chapter 6](../Chapter_6/Theory.qmd)
returns to model comparison with further tools (BIC, cross-validation) and with checks of
goodness of fit.

Many authors use AIC to reject models that differ largely from the best model and this may
be used as a form of hypothesis testing. That is, if adding a new term to your model does
not lower the AIC by more than two units, that new model is rejected.
This is a form of null hypothesis testing in disguise and the new term is treated as if it
was "not significant". This carries over all the issues of null hypothesis testing discussed
in [Chapter 1](../Chapter_1/Theory.qmd) and it may actually be worse. Traditional statistics
has developed several techniques to control false positive error rates (e.g. when testing
many hypotheses at the same time, as in post-hoc tests) which do not exist for AIC. Instead,
we can have a more nuanced view of model comparison, where AIC is just one of the criteria
that we use to contrast different models and in many cases there might not be a true reason
to make any choice (i.e., scientific research may not be about choices but rather about
evaluating the strength of evidence in favour of alternative hypotheses, and about making
predictions supported by evidence).



:::: {.callout-important title="Practice: waiting for a pollinator"}

The waiting times between visits of pollinators to a flower are fitted with an Exponential
model, which has one parameter, and with a Gamma model, which has two. The maximum
log-likelihoods are $-231.0$ and $-230.4$. Calculate the AIC of both models. Which one has
the lower AIC? Could the log-likelihood of the Gamma model have been the lower of the two?

::: {.callout-tip title="Solution" collapse="true"}

For the Exponential model $\mathrm{AIC}=2\times 1+2\times 231.0=464.0$ and for the Gamma
model $\mathrm{AIC}=2\times 2+2\times 230.4=464.8$. The Exponential model has the lower
AIC, although its log-likelihood is lower: the improvement of 0.6 units of log-likelihood
does not compensate for the additional parameter. A difference of 0.8 units of AIC is very
small, so these data hardly distinguish the two models. The log-likelihood of the Gamma
model could not have been the lower one, because the Exponential distribution is a Gamma
distribution with a shape of one, and a model that contains another as a special case fits
the same data at least as well.

:::
::::

The steps of this chapter are the same for every model that we fit in the rest of the
book, so we collect them as a recipe:

::: {.callout-note title="A recipe for a fit"}
1. Write the model: the random variable, its distribution and the assumptions (for
   example, that the observations are i.i.d.).
2. Write the NLL as an R function of two lists, the parameters and the data, that starts
   with `getAll()` and uses `log = TRUE`. Transform every constrained parameter inside it
   (the logarithm for a positive parameter, the logit for a probability).
3. Connect the NLL to the data with `cmb()` and process it with `MakeADFun()`, with a
   list of starting values, for example the method-of-moments estimates.
4. Minimise the NLL with `optim()`, using the elements `par`, `fn` and `gr` of the object
   and `method = "BFGS"`, and check that `convergence` is zero.
5. Transform the estimates back to the original scale and compare the fitted model with
   the data.
6. Calculate standard errors with `sdreport()`, and Wald intervals on the scale that was
   optimised, transforming their endpoints back. For the standard error of a quantity
   calculated from the parameters, add `ADREPORT()` to the NLL.
7. Calculate profile likelihood intervals with `TMB::tmbprofile()` and `confint()` for the
   parameters that the question is about.
8. Compare the candidate models with AIC.
:::

# Summary

The likelihood is the joint probability (or density) of the observed sample, read as a
function of the parameters with the data held fixed. For i.i.d. observations it is a
product of contributions, and its logarithm, the log-likelihood, is a sum. The maximum
likelihood estimate is the parameter value with the highest likelihood. We find it by
minimising the negative log-likelihood with `optim()`, after connecting it to the data
with `cmb()` and processing it with `MakeADFun()`, because a closed-form solution
exists only for a few models. Constraints on parameters are handled with transformations
(the logarithm for positive parameters and the logit for probabilities), and invariance
allows us to transform the estimates back.

With several parameters the likelihood is a surface. Its shape depends on the
parameterisation, although the maximum and the fitted model do not. A slice fixes the
other parameters, whereas a profile re-estimates them for every value of the parameter of
interest, so the profile takes their uncertainty into account.

Uncertainty is calculated from the same likelihood. In a large sample the sampling
distribution of a maximum likelihood estimator is approximately Normal, with a variance
equal to the inverse of the Fisher information, which we calculate from the curvature of
the log-likelihood at its maximum. From the inverse of the Hessian, `sdreport()` gives
standard errors and covariances on the scale that was optimised, and Wald
intervals are calculated on that scale and transformed back. The standard error of a
quantity calculated from the parameters is requested with `ADREPORT()`. Profile likelihood
intervals, which we calculate with `tmbprofile()`,
contain the values whose profile log-likelihood is within 1.92 units of the maximum; they
do not depend on the scale and they follow the asymmetry of the likelihood. A joint
confidence region for two parameters uses a cutoff of 3.00. Simulation shows how close the
coverage of each interval is to the confidence level.

Maximum likelihood estimators are consistent, asymptotically normal and asymptotically
efficient: in large samples they are centred on the parameter and no other regular
estimator is more precise. In finite samples they are in general biased, and their
sampling distribution can be far from Normal. When the model does not contain the
distribution of the data generating process (a misspecified model, @sec-misspecified), the
estimator converges to the parameter values that bring the model closest to that
process. Finally, AIC compares models fitted to the same data by penalising
the maximum log-likelihood for the number of parameters.

So far every observation had the same distribution. In
[Chapter 5](../Chapter_5/Theory.qmd) we study deterministic functions that describe how an
expected response changes with a predictor, and in [Chapter 6](../Chapter_6/Theory.qmd) we
combine them with the likelihood of this chapter to fit ecological response curves.

# Chapter quiz

The questions below follow the order of the chapter. Try each question before opening its
solution.

:::: {.callout-important title="Question 1: frog egg masses"}

You count the egg masses of a frog in eight ponds and write the likelihood
$L(\lambda\mid x_1,\ldots,x_8)$ of a Poisson model. Which quantities are fixed in this
function and which one varies? Does the likelihood give the probability that $\lambda$ has
a certain value?

::: {.callout-tip title="Answer" collapse="true"}

The eight counts are fixed and $\lambda$ varies. The likelihood is the probability of the
observed counts for a given value of $\lambda$. It does not integrate to one over the
values of $\lambda$, and in the model $\lambda$ is a fixed quantity, so the likelihood is
not a probability distribution for $\lambda$. We use it to compare values of $\lambda$.

:::
::::

:::: {.callout-important title="Question 2: mayfly nymphs"}

Two samples from a stream contain 2 and 0 mayfly nymphs. Assuming i.i.d. counts with a
Poisson distribution, write the likelihood as a function of $\lambda$ and evaluate it at
$\lambda=1$ and at $\lambda=2$. Which of the two values is better supported by the data?

::: {.callout-tip title="Answer" collapse="true"}

The likelihood is

$$
L(\lambda\mid 2,0)=\frac{e^{-\lambda}\lambda^{2}}{2!}\times e^{-\lambda}
=\frac{e^{-2\lambda}\lambda^{2}}{2}.
$$

At $\lambda=1$ it is $e^{-2}/2\approx 0.068$ and at $\lambda=2$ it is
$4e^{-4}/2\approx 0.037$. The data are better supported by $\lambda=1$, which is the mean
of the two counts.

:::
::::

:::: {.callout-important title="Question 3: bat passes"}

A detector records the number of bat passes on each of 300 nights. Why would you calculate
the log-likelihood of a model for these counts and not the likelihood? Write an R
expression for the log-likelihood of a Poisson model with a mean of 12 passes/night,
for counts stored in a vector called `passes`.

::: {.callout-tip title="Answer" collapse="true"}

The likelihood is a product of 300 probabilities, which can be smaller than the smallest
number that the computer can represent, so R may return zero. The log-likelihood is a sum
of log probabilities and does not have this problem:
`sum(dpois(x = passes, lambda = 12, log = TRUE))`.

:::
::::

:::: {.callout-important title="Question 4: germinated seeds"}

You minimise an NLL for the mean number of germinated seeds per tray, in which the
parameter is the logarithm of the mean. `optim()` returns `par = 1.7`, `value = 96.4` and
`convergence = 0`. What is the estimate of the mean? What is the maximum log-likelihood?
What does the convergence code tell you?

::: {.callout-tip title="Answer" collapse="true"}

The estimate of the mean is $e^{1.7}\approx 5.5$ germinated seeds per tray. `value` is the
minimum of the NLL, so the maximum log-likelihood is $-96.4$. A convergence code of zero
means that the algorithm stopped normally.

:::
::::

:::: {.callout-important title="Question 5: empty nest boxes"}

Under a Poisson model, the MLE of the mean number of eggs per nest box is 2.3. What is the
MLE of the probability that a nest box has no eggs? Which property of maximum likelihood
did you use?

::: {.callout-tip title="Answer" collapse="true"}

Under a Poisson model the probability of a count of zero is $e^{-\lambda}$. By invariance,
its MLE is $e^{-2.3}\approx 0.10$.

:::
::::

:::: {.callout-important title="Question 6: two curves for one parameter"}

For the half-saturation parameter of a model with three parameters, the range of values
within 1.92 units of the maximum log-likelihood is 2.1 to 2.9 when the other two
parameters are fixed at their estimates, and 1.8 to 3.4 when they are re-estimated for
each value. Which of the two is a slice and which is a profile? Which would you report as
a confidence interval, and why is the other one narrower?

::: {.callout-tip title="Answer" collapse="true"}

The first is a slice and the second a profile. The confidence interval is the one from the
profile, 1.8 to 3.4. The slice is narrower because it treats the other two parameters as
if they were known, whereas the profile takes into account that they are estimated from
the same data.

:::
::::

:::: {.callout-important title="Question 7: two surveys"}

Two surveys estimate the same detection probability by maximum likelihood on the logit
scale. The Hessian of the NLL is 16 in the first survey and 64 in the second. Calculate the
standard error of the estimate on the logit scale in each survey. In which survey is the
log-likelihood more sharply curved, and what does that say about the data?

::: {.callout-tip title="Answer" collapse="true"}

The standard errors are $1/\sqrt{16}=0.25$ and $1/\sqrt{64}=0.125$. The log-likelihood of
the second survey is more sharply curved: it contains four times as much information about
the parameter (for example, because it has four times as many observations), which halves
the standard error.

:::
::::

:::: {.callout-important title="Question 8: seed predation"}

The relative profile log-likelihood of a seed predation rate is $-0.8$ at a rate of
0.10/day and $-2.6$ at a rate of 0.20/day. Which of these two values is inside the 95%
profile likelihood interval? Which cutoff would you use for a joint 95% confidence region
of two parameters?

::: {.callout-tip title="Answer" collapse="true"}

A rate of 0.10/day is inside the interval, because its relative profile log-likelihood
is above $-1.92$, and a rate of 0.20/day is outside. For a joint region of two
parameters the cutoff is half the 0.95 quantile of a chi-squared distribution with two
degrees of freedom, `qchisq(p = 0.95, df = 2) / 2`, which is approximately 3.00.

:::
::::

:::: {.callout-important title="Question 9: parasites on fish"}

For counts of parasites per fish under a Negative Binomial model, the method of moments
and maximum likelihood give the same estimate of the mean but different estimates of the
shape $k$. How can two methods give different estimates from the same sample? Which of the
two estimators is more precise in a large sample?

::: {.callout-tip title="Answer" collapse="true"}

The two methods use different information. The method of moments chooses the parameters
that reproduce the mean and the variance of the sample, whereas maximum likelihood chooses
those under which all the observed counts are most probable. In a large sample the MLE is
the more precise of the two, because it is asymptotically efficient.

:::
::::

:::: {.callout-important title="Question 10: flower visits"}

A Poisson model fitted to counts of flower visits has a maximum log-likelihood of
$-412.6$, and a Negative Binomial model fitted to the same counts has $-389.2$. Calculate
the AIC of both models. Which model is preferred? Does this tell you that the preferred
model describes the counts well?

::: {.callout-tip title="Answer" collapse="true"}

The Poisson model has one parameter, so $\mathrm{AIC}=2+825.2=827.2$. The Negative
Binomial model has two, so $\mathrm{AIC}=4+778.4=782.4$. The Negative Binomial model is
preferred, with an AIC that is 44.8 units lower. AIC only ranks the models that we
compare, so it does not tell us whether the better of the two describes the counts well.

:::
::::
