---
title: "Maximum likelihood estimation"
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures in this chapter.
library(ggplot2)
```

# Introduction

# Learning goals

# The likelihood function {#sec-likelihood}

```{r}
# Seedling counts from the ten quadrats of Chapter 3, and the number of quadrats.
seedling_counts = c(3, 5, 2, 7, 4, 3, 4, 1, 6, 5)
n_quadrats = length(seedling_counts)

# The first three counts and two candidate values of lambda.
first_counts = c(3, 5, 2)
lambda_candidates = c(3, 4)

# Probability of each of the three counts at lambda = 3.
dpois(x = first_counts, lambda = 3)

# Likelihood of the three counts: the product of their probabilities, at each lambda.
prod(dpois(x = first_counts, lambda = lambda_candidates[1]))
prod(dpois(x = first_counts, lambda = lambda_candidates[2]))

# Likelihood of all ten counts at lambda = 4.
prod(dpois(x = seedling_counts, lambda = 4))
```

```{r}
# Closed-form Poisson likelihood of a sample of counts. Unlike a product of dpois() over
# the data, this function accepts a vector of lambda values, which integrate() needs.
poisson_likelihood = function(lambda, counts) {
  exp(-length(counts) * lambda) * lambda^sum(counts) / prod(factorial(counts))
}

# Check that it gives the same likelihood as the product of probabilities.
poisson_likelihood(lambda = 4, counts = seedling_counts)

# Area under the likelihood across all possible values of lambda. The likelihood is
# tiny, so the default absolute tolerance is too coarse: we set it to zero and ask
# for a small relative error instead.
integrate(
  f = poisson_likelihood,
  lower = 0,
  upper = Inf,
  counts = seedling_counts,
  rel.tol = 1e-8,
  abs.tol = 0
)
```

<!-- VALUES: likelihood of the counts (3, 5, 2) at lambda = 3: 0.00506; at lambda = 4: 0.00447 (probabilities, no unit); likelihood of the ten counts at lambda = 4: 2.37e-9; area under the likelihood: 3.76e-9 (exact value 3.765e-9; the default tolerances of integrate() return 3.80e-9 with a stated error larger than the value, so the tolerances are set in the code) -->

## The log-likelihood

```{r}
# Log-likelihood of the ten counts at lambda = 4: sum the log probabilities,
# using the argument log = TRUE of dpois().
sum(dpois(x = seedling_counts, lambda = 4, log = TRUE))

# The same number, from the logarithm of the product.
log(prod(dpois(x = seedling_counts, lambda = 4)))
```

```{r}
# Likelihood and log-likelihood of a large simulated sample of 1,000 counts at lambda = 4.
set.seed(seed = 501)
many_counts = rpois(n = 1000, lambda = 4)

# The product is smaller than the smallest positive number R can represent.
prod(dpois(x = many_counts, lambda = 4))

# The sum of log probabilities is an ordinary number.
sum(dpois(x = many_counts, lambda = 4, log = TRUE))
```

<!-- VALUES: log-likelihood of the ten counts at lambda = 4: -19.860 (same from the log of the product); product for 1,000 simulated counts at lambda = 4: 0; log-likelihood of those 1,000 counts: -2094.6 -->

# Maximum likelihood estimation {#sec-maximum-likelihood}

## The likelihood curve

```{r}
# Candidate values of lambda (seedlings per quadrat).
lambda_grid = seq(from = 1, to = 8, by = 0.05)

# Pre-allocate a vector for the log-likelihood and fill it with a loop over the grid.
loglik_grid = rep(NA, times = length(lambda_grid))
for (i in seq_along(lambda_grid)) {
  loglik_grid[i] = sum(dpois(x = seedling_counts, lambda = lambda_grid[i], log = TRUE))
}

# Relative log-likelihood: the difference from the highest value on the grid.
relative_loglik = loglik_grid - max(loglik_grid)

# Value of lambda on the grid with the highest log-likelihood.
lambda_grid[which.max(loglik_grid)]
```

```{r}
#| echo: false
#| label: fig-likelihood_curve
#| fig-cap: "Relative log-likelihood of the ten seedling counts under a Poisson model, as a function of the mean count lambda. The vertical line marks the maximum."
# Plot the relative log-likelihood curve and mark its maximum.
curve_data = data.frame(lambda = lambda_grid, relative = relative_loglik)
ggplot(curve_data, aes(x = lambda, y = relative)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_vline(
    xintercept = lambda_grid[which.max(loglik_grid)],
    linetype = "dashed",
    colour = "grey40"
  ) +
  coord_cartesian(ylim = c(-8, 0)) +
  labs(
    x = "Mean seedlings per quadrat (lambda)",
    y = "Relative log-likelihood"
  ) +
  theme_classic()
```

<!-- VALUES: maximum of the grid: lambda = 4 seedlings per quadrat -->

## Analytical solution for the Poisson model

```{r}
# The maximum likelihood estimate from the derivation is the mean of the counts.
lambda_hat = mean(seedling_counts)
lambda_hat

# The derivative of the log-likelihood, sum(x) / lambda - n, is zero at that value.
sum(seedling_counts) / lambda_hat - n_quadrats
```

<!-- VALUES: lambda_hat = 4 seedlings per quadrat; derivative of the log-likelihood at lambda_hat: 0 -->

## Numerical optimisation

```{r}
# Negative log-likelihood of Poisson counts as a function of lambda.
poisson_nll_lambda = function(lambda, counts) {
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}

# optim() minimises fn, starting from par. Further named arguments (here counts)
# are passed on to fn.
fit_lambda = optim(
  par = 2,
  fn = poisson_nll_lambda,
  counts = seedling_counts,
  method = "BFGS"
)

# Estimate, minimum of the negative log-likelihood, and convergence code (0 = normal stop).
fit_lambda$par
fit_lambda$value
fit_lambda$convergence
```

<!-- VALUES: optim() estimate of lambda: 4 seedlings per quadrat; minimum of the NLL: 19.860; convergence code 0 -->

## Constraints and transformations

```{r}
# Negative log-likelihood on the log scale: log_lambda can take any real value
# and exp() turns it into a positive mean.
poisson_nll = function(log_lambda, counts) {
  lambda = exp(log_lambda)
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}

# Fit starting from log(2) and transform the estimate back to the original scale.
poisson_fit = optim(
  par = log(2),
  fn = poisson_nll,
  counts = seedling_counts,
  method = "BFGS"
)
poisson_fit$convergence
exp(poisson_fit$par)

# The logit transformation for a probability, and its inverse.
qlogis(p = 0.25)
plogis(q = qlogis(p = 0.25))

# By invariance, the estimate of the probability of an empty quadrat, exp(-lambda),
# is calculated from the estimate of lambda.
exp(-exp(poisson_fit$par))
```

<!-- VALUES: estimate of lambda on the original scale: 4 seedlings per quadrat (convergence code 0); logit of 0.25: -1.0986 (plogis returns 0.25); estimated probability of an empty quadrat exp(-4): 0.0183 -->

# Models with several parameters {#sec-likelihood-surface}

```{r}
# Load the Owls data from the glmmTMB package without attaching the package.
data(Owls, package = "glmmTMB")

# Keep the visits of the female parent to nests with satiated nestlings.
satiated_female = Owls[Owls$FoodTreatment == "Satiated" & Owls$SexParent == "Female", ]

# Number of calls at each of these visits, and the number of visits.
call_counts = satiated_female$SiblingNegotiation
n_visits = length(call_counts)

# Method-of-moments estimates, as in the practical of Chapter 3.
mu_hat = mean(call_counts)
variance_hat = mean((call_counts - mu_hat)^2)
k_mom = mu_hat^2 / (variance_hat - mu_hat)

c(n_visits = n_visits, mu_hat = mu_hat, variance_hat = variance_hat, k_mom = k_mom)
```

```{r}
# Negative log-likelihood of the Negative Binomial model with mean mu and shape k.
# par holds the two parameters on the log scale: par[1] is log(mu) and par[2] is log(k).
nb_nll = function(par, counts) {
  mu = exp(par[1])
  k = exp(par[2])
  -sum(dnbinom(x = counts, mu = mu, size = k, log = TRUE))
}

# Start from the method-of-moments estimates, on the log scale.
start = c(log_mu = log(mu_hat), log_k = log(k_mom))

# Fit the model; hessian = TRUE also returns the matrix of second derivatives,
# which we use later for standard errors.
owl_fit = optim(
  par = start,
  fn = nb_nll,
  counts = call_counts,
  method = "BFGS",
  hessian = TRUE
)
owl_fit$convergence

# Maximum likelihood estimates on the original scale.
mu_mle = unname(exp(owl_fit$par["log_mu"]))
k_mle = unname(exp(owl_fit$par["log_k"]))
c(mu_mle = mu_mle, k_mle = k_mle)
```

<!-- VALUES: n_visits = 122; mu_hat = 4.754 calls per visit; variance_hat = 42.35 calls^2; k_mom = 0.601; mu_mle = 4.754 calls per visit; k_mle = 0.327; convergence code 0 -->

## Comparison with the method of moments

```{r}
# Log-likelihood at the maximum likelihood estimates, at the method-of-moments
# estimates, and for a Poisson model with the sample mean.
loglik_mle = -owl_fit$value
loglik_mom = -nb_nll(par = log(c(mu_hat, k_mom)), counts = call_counts)
loglik_poisson = sum(dpois(x = call_counts, lambda = mu_hat, log = TRUE))
c(mle = loglik_mle, mom = loglik_mom, poisson = loglik_poisson)

# Fraction of visits without calls: observed, and under each fitted model.
c(
  observed = mean(call_counts == 0),
  nb_mle = dnbinom(x = 0, mu = mu_mle, size = k_mle),
  nb_mom = dnbinom(x = 0, mu = mu_hat, size = k_mom),
  poisson = dpois(x = 0, lambda = mu_hat)
)

# Variance implied by the maximum likelihood fit, against the variance of the sample.
c(implied = mu_mle + mu_mle^2 / k_mle, sample = variance_hat)
```

```{r}
#| echo: false
#| label: fig-owl_fits
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of calls per visit (bars) and probabilities under the fitted models: Negative Binomial with maximum likelihood estimates, Negative Binomial with method-of-moments estimates, and Poisson."
# Relative frequencies of each count in the sample.
count_values = 0:max(call_counts)
observed_data = data.frame(
  count = count_values,
  frequency = tabulate(call_counts + 1, nbins = length(count_values)) / n_visits
)

# Fitted probabilities for each model at every possible count.
fitted_data = rbind(
  data.frame(
    count = count_values,
    probability = dnbinom(x = count_values, mu = mu_mle, size = k_mle),
    model = "Negative Binomial (maximum likelihood)"
  ),
  data.frame(
    count = count_values,
    probability = dnbinom(x = count_values, mu = mu_hat, size = k_mom),
    model = "Negative Binomial (method of moments)"
  ),
  data.frame(
    count = count_values,
    probability = dpois(x = count_values, lambda = mu_hat),
    model = "Poisson"
  )
)

# Bars for the sample and lines with points for the models.
ggplot() +
  geom_col(
    data = observed_data,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(
    values = c("#D55E00", "#009E73", "grey20")
  ) +
  labs(
    x = "Number of calls per visit",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top", legend.direction = "vertical")
```

<!-- VALUES: log-likelihood at the MLE: -299.44; at the method-of-moments estimates: -306.56; Poisson: -634.16; fraction of visits without calls: observed 0.443, Negative Binomial (MLE) 0.407, Negative Binomial (method of moments) 0.269, Poisson 0.009; variance implied by the ML fit: 73.8 calls^2 against 42.3 calls^2 in the sample; maximum count in the sample: 26 -->

## The likelihood surface

```{r}
# Grids of candidate values for the mean (calls per visit) and the shape.
mu_grid = seq(from = 2.5, to = 9, length.out = 91)
k_grid = seq(from = 0.12, to = 0.75, length.out = 91)

# Pre-allocate a matrix and fill it with two nested loops: row i for the i-th mean
# and column j for the j-th shape.
loglik_surface = matrix(NA, nrow = length(mu_grid), ncol = length(k_grid))
for (i in seq_along(mu_grid)) {
  for (j in seq_along(k_grid)) {
    loglik_surface[i, j] = sum(
      dnbinom(x = call_counts, mu = mu_grid[i], size = k_grid[j], log = TRUE)
    )
  }
}

# Relative log-likelihood: the difference from the maximum found by optim().
relative_surface = loglik_surface - loglik_mle
```

```{r}
#| eval: false
# The matrix can be drawn with contour(), as in the practical of Chapter 1.
contour(x = mu_grid, y = k_grid, z = relative_surface, levels = c(-6, -4, -3, -2, -1))
```

```{r}
#| echo: false
#| label: fig-likelihood_surface
#| fig-cap: "Contours of the relative log-likelihood of the owl data as a function of the mean and the shape of the Negative Binomial model. The circle is the maximum likelihood estimate and the triangle is the method-of-moments estimate."
# Arrange the matrix as a data frame: mu varies fastest, as in as.vector() of the matrix.
surface_data = expand.grid(mu = mu_grid, k = k_grid)
surface_data$relative = as.vector(relative_surface)

# The two estimates, to be drawn as points.
estimate_data = data.frame(
  mu = c(mu_mle, mu_hat),
  k = c(k_mle, k_mom),
  method = c("Maximum likelihood", "Method of moments")
)

# Contours of the relative log-likelihood with both estimates.
ggplot() +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_point(
    data = estimate_data,
    aes(x = mu, y = k, shape = method),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    x = "Mean number of calls per visit (mu)",
    y = "Shape (k)",
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: highest value of the relative log-likelihood on the grid: -0.0013 (the MLE lies between grid points); grid of 91 values of mu (2.5 to 9) and of k (0.12 to 0.75) -->

## Parameterisation and the shape of the surface

```{r}
# Negative log-likelihood in R's own parameters: shape k (size) and probability p (prob).
# par[1] is log(k) and par[2] is logit(p).
nb_prob_nll = function(par, counts) {
  k = exp(par[1])
  p = plogis(par[2])
  -sum(dnbinom(x = counts, size = k, prob = p, log = TRUE))
}

# Starting values: the method-of-moments shape and p = k / (k + mu).
start_prob = c(log_k = log(k_mom), logit_p = qlogis(k_mom / (k_mom + mu_hat)))

owl_fit_prob = optim(
  par = start_prob,
  fn = nb_prob_nll,
  counts = call_counts,
  method = "BFGS",
  hessian = TRUE
)
owl_fit_prob$convergence

# Estimates on the original scale and the maximum log-likelihood.
k_mle_prob = unname(exp(owl_fit_prob$par["log_k"]))
p_mle = unname(plogis(owl_fit_prob$par["logit_p"]))
c(k_mle_prob = k_mle_prob, p_mle = p_mle, loglik = -owl_fit_prob$value)

# By invariance, p equals k / (k + mu) evaluated at the first fit.
k_mle / (k_mle + mu_mle)
```

```{r}
# Grids of candidate values for the shape and the probability.
k_grid_prob = seq(from = 0.12, to = 0.75, length.out = 91)
p_grid = seq(from = 0.02, to = 0.16, length.out = 91)

# Fill the matrix of log-likelihoods with two nested loops.
loglik_surface_prob = matrix(NA, nrow = length(k_grid_prob), ncol = length(p_grid))
for (i in seq_along(k_grid_prob)) {
  for (j in seq_along(p_grid)) {
    loglik_surface_prob[i, j] = sum(
      dnbinom(x = call_counts, size = k_grid_prob[i], prob = p_grid[j], log = TRUE)
    )
  }
}
relative_surface_prob = loglik_surface_prob - loglik_mle
```

```{r}
#| echo: false
#| label: fig-surface_size_prob
#| fig-cap: "Contours of the relative log-likelihood of the owl data as a function of the shape k and the probability p of the Negative Binomial model. The circle is the maximum likelihood estimate."
# Arrange the matrix as a data frame: k varies fastest.
surface_prob_data = expand.grid(k = k_grid_prob, p = p_grid)
surface_prob_data$relative = as.vector(relative_surface_prob)

# Contours with the maximum likelihood estimate.
ggplot() +
  geom_contour(
    data = surface_prob_data,
    aes(x = k, y = p, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_point(
    data = data.frame(k = k_mle_prob, p = p_mle),
    aes(x = k, y = p),
    colour = "#D55E00",
    size = 3
  ) +
  labs(
    x = "Shape (k)",
    y = "Probability (p)"
  ) +
  theme_classic()
```

<!-- VALUES: convergence code 0; k = 0.327 and p = 0.0644 (no unit); maximum log-likelihood -299.44, the same as in the (mu, k) parameters; k / (k + mu) from the first fit: 0.0644 -->

## Slices and profiles

```{r}
# Fine grid of shape values on which the slice and the profile are evaluated.
k_fine = seq(from = 0.12, to = 0.75, length.out = 127)

# Negative log-likelihood as a function of logit(p) only, for a given shape k.
nll_given_k = function(logit_p, k, counts) {
  -sum(dnbinom(x = counts, size = k, prob = plogis(logit_p), log = TRUE))
}

# Slice: p is fixed at its estimate. Profile: for each k, p is re-estimated with optim(),
# starting from the joint estimate.
slice_loglik = rep(NA, times = length(k_fine))
profile_loglik = rep(NA, times = length(k_fine))
profile_p = rep(NA, times = length(k_fine))
profile_convergence = rep(NA, times = length(k_fine))
for (i in seq_along(k_fine)) {
  slice_loglik[i] = sum(
    dnbinom(x = call_counts, size = k_fine[i], prob = p_mle, log = TRUE)
  )
  inner_fit = optim(
    par = owl_fit_prob$par["logit_p"],
    fn = nll_given_k,
    k = k_fine[i],
    counts = call_counts,
    method = "BFGS"
  )
  profile_loglik[i] = -inner_fit$value
  profile_p[i] = plogis(inner_fit$par)
  profile_convergence[i] = inner_fit$convergence
}

# Relative log-likelihoods and the range of k within 1.92 of the maximum
# (the cutoff is explained in the section on profile likelihood intervals).
slice_relative = slice_loglik - loglik_mle
profile_relative = profile_loglik - loglik_mle
range(k_fine[slice_relative >= -1.92])
range(k_fine[profile_relative >= -1.92])

# All inner fits stopped normally.
all(profile_convergence == 0)
```

```{r}
#| echo: false
# In the parameters mu and k, the best mu for any k is the sample mean, so the
# profile of k equals the slice at the estimate of mu. Check this on the same grid.
slice_k_mu = rep(NA, times = length(k_fine))
profile_k_mu = rep(NA, times = length(k_fine))
for (i in seq_along(k_fine)) {
  slice_k_mu[i] = sum(
    dnbinom(x = call_counts, mu = mu_mle, size = k_fine[i], log = TRUE)
  )
  inner_fit = optim(
    par = log(mu_mle),
    fn = function(log_mu, k, counts) {
      -sum(dnbinom(x = counts, mu = exp(log_mu), size = k, log = TRUE))
    },
    k = k_fine[i],
    counts = call_counts,
    method = "BFGS"
  )
  profile_k_mu[i] = -inner_fit$value
}
max(abs(slice_k_mu - profile_k_mu))
```

```{r}
#| echo: false
#| label: fig-slice_profile
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Slice and profile of the log-likelihood for the shape k of the Negative Binomial model fitted to the owl data. Left: contours in k and p with the slice (dashed line) and the values of p that maximise the log-likelihood for each k (solid line). Right: relative log-likelihood along the slice and along the profile; the horizontal line is at -1.92."
library(patchwork)

# Left panel: contours with the slice and the path of the profile.
path_data = data.frame(k = k_fine, p = profile_p)
left_panel = ggplot() +
  geom_contour(
    data = surface_prob_data,
    aes(x = k, y = p, z = relative),
    breaks = c(-12, -9, -6, -4, -3, -2, -1),
    colour = "grey40"
  ) +
  geom_hline(yintercept = p_mle, linetype = "dashed", colour = "#D55E00") +
  geom_line(data = path_data, aes(x = k, y = p), colour = "#D55E00", linewidth = 0.8) +
  geom_point(
    data = data.frame(k = k_mle_prob, p = p_mle),
    aes(x = k, y = p),
    colour = "#D55E00",
    size = 3
  ) +
  labs(x = "Shape (k)", y = "Probability (p)") +
  theme_classic()

# Right panel: relative log-likelihood along the slice and along the profile.
curve_data = rbind(
  data.frame(k = k_fine, relative = slice_relative, curve = "Slice"),
  data.frame(k = k_fine, relative = profile_relative, curve = "Profile")
)
right_panel = ggplot(curve_data, aes(x = k, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -1.92, colour = "grey40") +
  coord_cartesian(ylim = c(-6, 0)) +
  labs(x = "Shape (k)", y = "Relative log-likelihood", linetype = NULL) +
  theme_classic() +
  theme(legend.position = "top")

left_panel | right_panel
```

<!-- VALUES: k within 1.92 of the maximum, on a grid with step 0.005: slice 0.260 to 0.405, profile 0.240 to 0.450 (no unit); all 127 inner fits converged; profile of k in the (mu, k) parameters against the slice at mu_mle: maximum absolute difference 1.2e-7 -->

# Properties of maximum likelihood estimators {#sec-mle_properties}

## Consistency

## Asymptotic normality and Fisher information

```{r}
# Observed information of the Poisson model at the estimate: -l''(lambda_hat) = sum(x) / lambda_hat^2,
# which equals n / lambda_hat at lambda_hat = mean(x).
observed_information = sum(seedling_counts) / lambda_hat^2
observed_information

# Approximate variance of the estimator and its standard error (seedlings per quadrat).
approximate_variance = 1 / observed_information
standard_error = sqrt(approximate_variance)
c(approximate_variance = approximate_variance, standard_error = standard_error)

# The plug-in standard error of Chapter 3.
sqrt(lambda_hat / n_quadrats)
```

```{r}
#| echo: false
#| warning: false
# Fit the Negative Binomial model to a sample of counts by maximum likelihood.
# Returns the estimates, the method-of-moments shape, and whether the fit converged.
# The starting values are the method-of-moments estimates when they exist.
fit_negbin = function(counts, hessian = FALSE) {
  mu_start = mean(counts)
  variance_start = mean((counts - mu_start)^2)
  k_start_mom = ifelse(
    variance_start > mu_start,
    mu_start^2 / (variance_start - mu_start),
    NA
  )
  fit = optim(
    par = log(c(mu_start, ifelse(is.na(k_start_mom), 1, k_start_mom))),
    fn = nb_nll,
    counts = counts,
    method = "BFGS",
    hessian = hessian
  )
  list(
    mu = exp(fit$par[1]),
    k = exp(fit$par[2]),
    k_mom = k_start_mom,
    loglik = -fit$value,
    convergence = fit$convergence,
    hessian = fit$hessian
  )
}

# Hypothetical experiments from a Negative Binomial model whose parameters are the
# maximum likelihood estimates for the owls, for three numbers of visits.
set.seed(seed = 601)
n_experiments = 2000
visit_sizes = c(20, 122, 500)
sim_k_ml = matrix(NA, nrow = n_experiments, ncol = length(visit_sizes))
sim_k_mom = matrix(NA, nrow = n_experiments, ncol = length(visit_sizes))
sim_converged = matrix(NA, nrow = n_experiments, ncol = length(visit_sizes))
for (s in seq_along(visit_sizes)) {
  for (e in seq_len(n_experiments)) {
    counts = rnbinom(n = visit_sizes[s], mu = mu_mle, size = k_mle)
    fit = fit_negbin(counts = counts)
    sim_k_ml[e, s] = fit$k
    sim_k_mom[e, s] = fit$k_mom
    sim_converged[e, s] = fit$convergence == 0
  }
}
```

```{r}
#| echo: false
#| label: fig-sampling_distribution_k
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Maximum likelihood estimates of the shape k from 2,000 hypothetical experiments with 20 and with 500 visits, simulated from a Negative Binomial model with the estimates for the owls as parameters. The vertical line is the shape used in the simulation and the curves are Normal distributions with the mean and standard deviation of the estimates."
library(patchwork)

# Histogram of the estimates for one number of visits, with its Normal approximation.
plot_sampling_k = function(column, bin_width) {
  estimates = sim_k_ml[, column]
  estimate_grid = seq(from = min(estimates), to = max(estimates), length.out = 200)
  approximation = data.frame(
    estimate = estimate_grid,
    density = dnorm(x = estimate_grid, mean = mean(estimates), sd = sd(estimates))
  )
  ggplot(data.frame(estimate = estimates), aes(x = estimate)) +
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = bin_width,
      fill = "grey90",
      colour = "grey40"
    ) +
    geom_line(
      data = approximation,
      aes(x = estimate, y = density),
      colour = "#D55E00",
      linewidth = 0.8
    ) +
    geom_vline(xintercept = k_mle, linetype = "dashed") +
    labs(
      title = paste(visit_sizes[column], "visits"),
      x = "Estimated shape (k)",
      y = "Density"
    ) +
    theme_classic()
}

plot_sampling_k(column = 1, bin_width = 0.04) |
  plot_sampling_k(column = 3, bin_width = 0.01)
```

<!-- VALUES: Poisson: observed information 2.5, approximate variance 0.4 and standard error 0.632 seedlings per quadrat, equal to the Chapter 3 plug-in standard error; simulation from a Negative Binomial model with mu = 4.754 and k = 0.327 (the MLEs), 2,000 hypothetical experiments for each of 20, 122 and 500 visits; all 6,000 fits converged -->

## Efficiency

```{r}
# Skewness of a vector of estimates (third standardised moment).
skewness = function(x) {
  mean((x - mean(x))^3) / mean((x - mean(x))^2)^1.5
}

# Mean, standard deviation and mean squared error of both estimators of k for each
# number of visits, from the hypothetical experiments simulated above. The method of
# moments gives no estimate when the sample variance does not exceed the mean (NA).
efficiency_table = data.frame(
  visits = visit_sizes,
  mean_ml = NA, sd_ml = NA, mse_ml = NA, skewness_ml = NA,
  mean_mom = NA, sd_mom = NA, mse_mom = NA,
  mom_missing = NA
)
for (s in seq_along(visit_sizes)) {
  ml = sim_k_ml[, s]
  mom = sim_k_mom[, s]
  efficiency_table$mean_ml[s] = mean(ml)
  efficiency_table$sd_ml[s] = sd(ml)
  efficiency_table$mse_ml[s] = mean((ml - k_mle)^2)
  efficiency_table$skewness_ml[s] = skewness(ml)
  efficiency_table$mean_mom[s] = mean(mom, na.rm = TRUE)
  efficiency_table$sd_mom[s] = sd(mom, na.rm = TRUE)
  efficiency_table$mse_mom[s] = mean((mom - k_mle)^2, na.rm = TRUE)
  efficiency_table$mom_missing[s] = sum(is.na(mom))
}
efficiency_table
```

<!-- VALUES: ML estimate of k, 20 | 122 | 500 visits: mean 0.397 | 0.336 | 0.329; SD 0.194 | 0.0546 | 0.0242; MSE 0.0423 | 0.00306 | 0.000588; skewness 2.59 | 0.50 | 0.22. Method of moments: mean 0.520 | 0.360 | 0.337; SD 0.274 | 0.0862 | 0.0423; MSE 0.112 | 0.00851 | 0.00187. Shape used in the simulation: 0.327. Experiments without a method-of-moments estimate: 0 in each case -->

## Bias in finite samples

```{r}
# Relative bias of the maximum likelihood estimator of k for each number of visits.
data.frame(
  visits = efficiency_table$visits,
  relative_bias = efficiency_table$mean_ml / k_mle - 1
)
```

<!-- VALUES: relative bias of the ML estimate of k, 20 | 122 | 500 visits: 0.211 | 0.0265 | 0.0056 (fractions) -->

## Misspecified models

```{r}
# The visits come from a limited number of nests.
length(unique(satiated_female$Nest))
```

<!-- VALUES: number of nests among the 122 visits: 24; variance implied by the ML fit 73.8 calls^2 and variance of the sample 42.3 calls^2 (from the owl comparison above) -->

# Uncertainty from the likelihood {#sec-likelihood-uncertainty}

## Standard errors and Wald intervals

```{r}
# Wald interval for the Poisson mean, on the original scale (as in Chapter 3).
critical_value = qnorm(p = 0.975)
wald_original = lambda_hat + c(-1, 1) * critical_value * sqrt(lambda_hat / n_quadrats)
wald_original
```

```{r}
# Refit the Poisson model on the log scale, now with the Hessian.
poisson_fit = optim(
  par = log(2),
  fn = poisson_nll,
  counts = seedling_counts,
  method = "BFGS",
  hessian = TRUE
)

# The Hessian is a 1 by 1 matrix; take its single element. Its inverse is the
# approximate variance of the estimate of log(lambda).
hessian_log = poisson_fit$hessian[1, 1]
se_log_lambda = sqrt(1 / hessian_log)
c(hessian_log = hessian_log, se_log_lambda = se_log_lambda)

# Wald interval on the log scale, back-transformed to seedlings per quadrat.
wald_log = exp(poisson_fit$par + c(-1, 1) * critical_value * se_log_lambda)
wald_log
```

```{r}
# Two parameters: solve() returns the inverse of the Hessian, which approximates the
# covariance matrix of the estimates of log(mu) and log(k). diag() takes its diagonal
# (the variances) and sqrt() gives the standard errors.
owl_covariance = solve(owl_fit$hessian)
owl_se = sqrt(diag(owl_covariance))
owl_se

# Wald intervals on the log scales, back-transformed to the original scales.
owl_wald_lower = exp(owl_fit$par - critical_value * owl_se)
owl_wald_upper = exp(owl_fit$par + critical_value * owl_se)
owl_wald = rbind(
  mu = c(estimate = mu_mle, lower = owl_wald_lower[[1]], upper = owl_wald_upper[[1]]),
  k = c(estimate = k_mle, lower = owl_wald_lower[[2]], upper = owl_wald_upper[[2]])
)
owl_wald

# Standard error of the mean in calls per visit, with the formula of the practical of
# Chapter 3 and the maximum likelihood estimate of k, and the Chapter 3 value.
se_mu_nb = sqrt((mu_mle + mu_mle^2 / k_mle) / n_visits)
se_mu_mom = sqrt((mu_hat + mu_hat^2 / k_mom) / n_visits)
c(se_mu_nb = se_mu_nb, se_mu_mom = se_mu_mom)

# Wald interval for mu from the Chapter 3 formula, with the maximum likelihood estimate of k.
mu_mle + c(-1, 1) * critical_value * se_mu_nb
```

<!-- VALUES: Poisson Wald interval, original scale: 2.76 to 5.24 seedlings per quadrat; log-scale Hessian 40.0 and SE 0.158; Wald interval from the log scale, back-transformed: 2.93 to 5.45; owls: SEs 0.1636 for log(mu) and 0.1655 for log(k); Wald intervals mu = 3.45 to 6.55 calls per visit, k = 0.237 to 0.453; SE of mu 0.778 calls per visit with the Negative Binomial formula and 0.589 with the method-of-moments value of Chapter 3; interval for mu from the formula: 3.23 to 6.28 calls per visit -->

## Covariance between estimates

```{r}
# Correlations between the estimates, from the inverse Hessian.
cov2cor(owl_covariance)

# The same for the parameters k and p, whose estimates are correlated.
cov2cor(solve(owl_fit_prob$hessian))
```

<!-- VALUES: correlation between the estimates of log(mu) and log(k): 0.00006 (5.7e-5); between log(k) and logit(p): 0.711 -->

## Profile likelihood intervals

```{r}
# Cutoff for the relative profile log-likelihood: half the 95% quantile of a
# chi-squared distribution with one degree of freedom.
qchisq(p = 0.95, df = 1)
sqrt(qchisq(p = 0.95, df = 1))
cutoff = qchisq(p = 0.95, df = 1) / 2
cutoff
```

```{r}
# Relative log-likelihood of the Poisson model: with one parameter it is also the profile.
poisson_relative_loglik = function(lambda, counts) {
  sum(dpois(x = counts, lambda = lambda, log = TRUE)) -
    sum(dpois(x = counts, lambda = mean(counts), log = TRUE))
}

# uniroot() finds where a function crosses zero inside an interval whose ends give
# values of opposite sign. We look for the two values of lambda where the relative
# log-likelihood equals -cutoff, one below the estimate and one above it.
lambda_lower = uniroot(
  f = function(lambda) poisson_relative_loglik(lambda, counts = seedling_counts) + cutoff,
  interval = c(0.5, lambda_hat),
  tol = 1e-8
)$root
lambda_upper = uniroot(
  f = function(lambda) poisson_relative_loglik(lambda, counts = seedling_counts) + cutoff,
  interval = c(lambda_hat, 20),
  tol = 1e-8
)$root
likelihood_interval = c(lambda_lower, lambda_upper)
likelihood_interval

# The three intervals for lambda (seedlings per quadrat).
poisson_intervals = rbind(
  "Wald, original scale" = wald_original,
  "Wald, log scale" = wald_log,
  "Likelihood" = likelihood_interval
)
colnames(poisson_intervals) = c("lower", "upper")
poisson_intervals
```

```{r}
# Profile log-likelihood of mu relative to the maximum: for each mu the shape is
# re-estimated with optim() on log(k), starting from the joint estimate.
nll_given_mu = function(log_k, mu, counts) {
  -sum(dnbinom(x = counts, mu = mu, size = exp(log_k), log = TRUE))
}
profile_relative_mu = function(mu, counts) {
  inner_fit = optim(
    par = owl_fit$par["log_k"],
    fn = nll_given_mu,
    mu = mu,
    counts = counts,
    method = "BFGS"
  )
  -inner_fit$value - loglik_mle
}

# The same for k, re-estimating the mean on the log scale.
nll_given_k_mu = function(log_mu, k, counts) {
  -sum(dnbinom(x = counts, mu = exp(log_mu), size = k, log = TRUE))
}
profile_relative_k = function(k, counts) {
  inner_fit = optim(
    par = owl_fit$par["log_mu"],
    fn = nll_given_k_mu,
    k = k,
    counts = counts,
    method = "BFGS"
  )
  -inner_fit$value - loglik_mle
}

# Both profiles are zero at the joint estimate.
c(profile_relative_mu(mu = mu_mle, counts = call_counts),
  profile_relative_k(k = k_mle, counts = call_counts))

# Endpoints from uniroot(). The ends of each search interval were chosen from the
# profile curves: below and above the estimate, where the profile is under -cutoff.
mu_lower = uniroot(
  f = function(mu) profile_relative_mu(mu, counts = call_counts) + cutoff,
  interval = c(1, mu_mle),
  tol = 1e-8
)$root
mu_upper = uniroot(
  f = function(mu) profile_relative_mu(mu, counts = call_counts) + cutoff,
  interval = c(mu_mle, 20),
  tol = 1e-8
)$root
k_lower = uniroot(
  f = function(k) profile_relative_k(k, counts = call_counts) + cutoff,
  interval = c(0.05, k_mle),
  tol = 1e-8
)$root
k_upper = uniroot(
  f = function(k) profile_relative_k(k, counts = call_counts) + cutoff,
  interval = c(k_mle, 2),
  tol = 1e-8
)$root

# Profile intervals next to the Wald intervals.
owl_intervals = rbind(
  "mu, profile" = c(mu_lower, mu_upper),
  "mu, Wald" = owl_wald["mu", c("lower", "upper")],
  "k, profile" = c(k_lower, k_upper),
  "k, Wald" = owl_wald["k", c("lower", "upper")]
)
colnames(owl_intervals) = c("lower", "upper")
owl_intervals
```

```{r}
#| echo: false
#| label: fig-profile_interval
#| fig-cap: "Profile relative log-likelihood of the mean number of calls per visit for the owl data (solid line) and the quadratic approximation used by the Wald interval (dashed line). The horizontal line is at minus half the 95% quantile of the chi-squared distribution with one degree of freedom, and the vertical lines mark the profile interval."
# Evaluate the profile on a grid of means.
mu_fine = seq(from = 2.5, to = 9, length.out = 131)
profile_mu_values = rep(NA, times = length(mu_fine))
for (i in seq_along(mu_fine)) {
  profile_mu_values[i] = profile_relative_mu(mu = mu_fine[i], counts = call_counts)
}

# Quadratic approximation around the estimate, with the standard error of mu from the
# Negative Binomial variance (se_mu_nb).
quadratic_values = -(mu_fine - mu_mle)^2 / (2 * se_mu_nb^2)

curve_data = rbind(
  data.frame(mu = mu_fine, relative = profile_mu_values, curve = "Profile"),
  data.frame(mu = mu_fine, relative = quadratic_values, curve = "Quadratic approximation")
)
ggplot(curve_data, aes(x = mu, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -cutoff, colour = "grey40") +
  geom_vline(xintercept = c(mu_lower, mu_upper), colour = "grey40", linetype = "dotted") +
  coord_cartesian(ylim = c(-6, 0)) +
  labs(
    x = "Mean number of calls per visit (mu)",
    y = "Relative log-likelihood",
    linetype = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: qchisq(0.95, 1) = 3.841, its square root 1.960; cutoff 1.921; Poisson intervals for lambda (seedlings per quadrat): Wald original 2.76 to 5.24, Wald log 2.93 to 5.45, likelihood 2.89 to 5.37; owls: profile interval for mu 3.49 to 6.68 calls per visit (Wald 3.45 to 6.55), for k 0.235 to 0.451 (Wald 0.237 to 0.453); the profile values at the estimates are about 1e-7 (zero up to optimiser tolerance) -->

## Joint confidence regions

```{r}
# Cutoff for a joint region of two parameters: half the 95% quantile of a
# chi-squared distribution with two degrees of freedom.
qchisq(p = 0.95, df = 2)
joint_cutoff = qchisq(p = 0.95, df = 2) / 2
joint_cutoff
```

```{r}
#| echo: false
#| label: fig-joint_region
#| fig-cap: "Approximate joint 95% confidence region for the mean and the shape of the Negative Binomial model fitted to the owl data (contour in vermillion). The segments along the axes are the 95% profile likelihood intervals of each parameter."
# Contours of the surface, the contour of the joint region, the estimate and the
# profile intervals of the two parameters drawn along the axes.
ggplot() +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = c(-12, -9, -6, -4, -2, -1),
    colour = "grey40"
  ) +
  geom_contour(
    data = surface_data,
    aes(x = mu, y = k, z = relative),
    breaks = -joint_cutoff,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  geom_point(
    data = data.frame(mu = mu_mle, k = k_mle),
    aes(x = mu, y = k),
    colour = "#D55E00",
    size = 3
  ) +
  annotate(
    "segment",
    x = mu_lower, xend = mu_upper, y = min(k_grid), yend = min(k_grid),
    linewidth = 1.2
  ) +
  annotate(
    "segment",
    x = min(mu_grid), xend = min(mu_grid), y = k_lower, yend = k_upper,
    linewidth = 1.2
  ) +
  labs(
    x = "Mean number of calls per visit (mu)",
    y = "Shape (k)"
  ) +
  theme_classic()
```

<!-- VALUES: qchisq(0.95, 2) = 5.991 and the joint cutoff 2.996 (3.00), against 1.921 for one parameter -->

## Coverage checked by simulation

```{r}
#| echo: false
#| warning: false
# Cutoff of the relative log-likelihood for 95% profile intervals (one parameter).
cutoff = qchisq(p = 0.95, df = 1) / 2
critical_value = qnorm(p = 0.975)

# Poisson model: coverage of the Wald interval on the original scale and of the
# likelihood interval, for a mean lambda and a number of quadrats.
poisson_coverage = function(lambda, n_quadrats, n_experiments) {
  # One row per hypothetical experiment, one column per quadrat.
  counts = matrix(
    rpois(n = n_experiments * n_quadrats, lambda = lambda),
    nrow = n_experiments,
    ncol = n_quadrats
  )
  totals = rowSums(counts)
  estimates = totals / n_quadrats

  # Wald interval on the original scale.
  se = sqrt(estimates / n_quadrats)
  wald_covered = estimates - critical_value * se <= lambda &
    lambda <= estimates + critical_value * se

  # The likelihood interval contains lambda when its relative log-likelihood is within
  # the cutoff. A sample of zeros has no finite estimate of the interval: a miss.
  relative = -n_quadrats * (lambda - estimates) + totals * log(lambda / estimates)
  likelihood_covered = totals > 0 & relative >= -cutoff

  c(wald = mean(wald_covered), likelihood = mean(likelihood_covered))
}

set.seed(seed = 701)
coverage_poisson_small = poisson_coverage(lambda = 0.5, n_quadrats = 10, n_experiments = 5000)
coverage_poisson_large = poisson_coverage(lambda = 4, n_quadrats = 30, n_experiments = 5000)

# Negative Binomial model (owl estimates): coverage of the Wald interval for k on the log
# scale and of the profile interval. The profile at the generating k is the log-likelihood
# with mu set to the sample mean, which is its maximum for any k.
negbin_coverage = function(n_visits, n_experiments) {
  wald_covered = rep(NA, times = n_experiments)
  profile_covered = rep(NA, times = n_experiments)
  for (e in seq_len(n_experiments)) {
    counts = rnbinom(n = n_visits, mu = mu_mle, size = k_mle)
    fit = fit_negbin(counts = counts, hessian = TRUE)
    se_log_k = tryCatch(sqrt(solve(fit$hessian)[2, 2]), error = function(e) NA)
    if (fit$convergence == 0 && !is.na(se_log_k)) {
      wald_covered[e] = abs(log(k_mle) - log(fit$k)) <= critical_value * se_log_k
      profile_loglik_k0 = sum(
        dnbinom(x = counts, mu = mean(counts), size = k_mle, log = TRUE)
      )
      profile_covered[e] = profile_loglik_k0 - fit$loglik >= -cutoff
    }
  }
  c(
    wald = mean(wald_covered, na.rm = TRUE),
    profile = mean(profile_covered, na.rm = TRUE),
    failed = sum(is.na(wald_covered))
  )
}

set.seed(seed = 702)
coverage_nb_small = negbin_coverage(n_visits = 20, n_experiments = 2000)
coverage_nb_large = negbin_coverage(n_visits = 122, n_experiments = 2000)

# Collect the results for the figure.
coverage_results = data.frame(
  case = rep(
    c(
      "Poisson, lambda = 0.5,\n10 quadrats",
      "Poisson, lambda = 4,\n30 quadrats",
      "Negative Binomial, k,\n20 visits",
      "Negative Binomial, k,\n122 visits"
    ),
    each = 2
  ),
  interval = rep(c("Wald", "Likelihood"), times = 4),
  coverage = c(
    coverage_poisson_small[["wald"]], coverage_poisson_small[["likelihood"]],
    coverage_poisson_large[["wald"]], coverage_poisson_large[["likelihood"]],
    coverage_nb_small[["wald"]], coverage_nb_small[["profile"]],
    coverage_nb_large[["wald"]], coverage_nb_large[["profile"]]
  )
)
coverage_results$case = factor(coverage_results$case, levels = unique(coverage_results$case))
coverage_results
```

```{r}
#| echo: false
#| label: fig-coverage
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Simulated coverage of nominal 95% Wald and likelihood intervals. Poisson cases: 5,000 hypothetical experiments, interval for the mean on the original scale. Negative Binomial cases: 2,000 hypothetical experiments, interval for the shape k, with the Wald interval calculated on the log scale. The dashed line is the nominal level."
# Coverage by case and type of interval; the dashed line is the intended 0.95.
ggplot(coverage_results, aes(x = case, y = coverage, shape = interval, linetype = interval)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey40") +
  geom_point(
    position = position_dodge(width = 0.4),
    colour = "#D55E00",
    size = 3
  ) +
  scale_linetype_manual(values = c("solid", "solid")) +
  coord_cartesian(ylim = c(0.8, 1)) +
  labs(x = NULL, y = "Simulated coverage", shape = NULL) +
  guides(linetype = "none") +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: simulated coverage (Wald, likelihood): Poisson lambda = 0.5 with 10 quadrats (5,000 experiments) 0.871 and 0.931; Poisson lambda = 4 with 30 quadrats 0.952 and 0.954; Negative Binomial k with 20 visits (2,000 experiments) 0.946 and 0.942 (Wald on the log scale, profile); with 122 visits 0.959 and 0.960; failed fits: 0 in both Negative Binomial cases -->

# Model comparison with AIC {#sec-aic}

```{r}
# Maximised log-likelihood, number of estimated parameters K and AIC for the two models
# fitted to the owl data. AIC = 2K - 2 * log-likelihood.
aic_table = data.frame(
  model = c("Poisson", "Negative Binomial"),
  K = c(1, 2),
  loglik = c(loglik_poisson, loglik_mle)
)
aic_table$aic = 2 * aic_table$K - 2 * aic_table$loglik

# Only differences in AIC matter: difference from the lowest value.
aic_table$delta_aic = aic_table$aic - min(aic_table$aic)
aic_table
```

<!-- VALUES: Poisson: log-likelihood -634.16, AIC 1270.33, difference 667.4; Negative Binomial: log-likelihood -299.44, AIC 602.88, difference 0 -->

# Summary

# Chapter quiz
