---
title: "Chapter 3: Probability for continuous data"
date: today
---

These exercises use assumed probability models and simulated measurements. They
do not represent observations from a field study. Work through the first example
before trying the remaining exercises. The distribution functions are introduced
in the [Chapter 3 theory](../Theory.qmd).

# Exercise 1: Tree heights under a normal model

Suppose the height $X$ of a randomly selected tree follows a normal model with
mean 22 m and standard deviation 4 m:

$$
X\sim\operatorname{Normal}(22,4^2).
$$

1. Calculate the probability that a tree is between 18 m and 26 m tall.
2. Find the height below which 95% of trees fall under this model.
3. Simulate 1,000 tree heights. Compare a density-scaled histogram with the
   theoretical density and explain any differences.

## Solution

The observational unit is one tree. Height is continuous, so we calculate
probabilities for **intervals**, not for an exact height. The notation above
uses the variance $4^2$; R's `sd` argument takes the standard deviation, 4 m.

First, subtract the cumulative probability below 18 m from the cumulative
probability below 26 m:

$$
P(18<X\leq26)=P(X\leq26)-P(X\leq18).
$$

```{r}
height_mean = 22
height_sd = 4

prob_18_to_26 = pnorm(q = 26, mean = height_mean, sd = height_sd) -
  pnorm(q = 18, mean = height_mean, sd = height_sd)
prob_18_to_26
```

The probability is approximately 0.683. Under this model, about 68% of trees
have heights in this interval. Including or excluding the endpoints makes no
difference for a continuous model.

Next, a 95th percentile is a height with 95% of the model's probability below it:

```{r}
height_95 = qnorm(p = 0.95, mean = height_mean, sd = height_sd)
height_95
```

The height is approximately 28.6 m. This is a model quantile, not the height
of a particular observed tree.

To compare a sample with the model, set a seed and generate 1,000 possible
tree heights. The seed makes this example reproducible.

```{r}
set.seed(seed = 101)
tree_heights = rnorm(n = 1000, mean = height_mean, sd = height_sd)

hist(
  x = tree_heights, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Density per metre",
  main = "Simulated tree heights and assumed model"
)
curve(
  dnorm(x = x, mean = height_mean, sd = height_sd),
  add = TRUE, lwd = 2
)
legend(
  "topright", legend = "Assumed normal density",
  lty = 1, lwd = 2, bty = "n"
)
```

The histogram uses density scaling because the curve also shows density.
The bars will not match the curve exactly: they summarise one finite sample,
and their appearance also depends on the bin widths. The curve is the model
that generated the sample.

# Exercise 2: Body mass under a lognormal model

Suppose body mass $X$ in grams follows a lognormal model with
`meanlog = log(30)` and `sdlog = 0.5`. These parameters describe **log body
mass**, not body mass in grams.

1. Calculate the model median and mean body mass. Explain why they differ.
2. Calculate the probability that body mass exceeds 50 g and the 95th percentile.
3. Simulate 500 body masses with seed 202. Make density-scaled histograms on the
   original and log scales. Describe their shapes.
4. Add the theoretical lognormal density to the original-scale histogram.
   Explain what the comparison can and cannot tell you about field data.

# Exercise 3: Vegetation cover under a beta model

Suppose $X$ is the continuously measured fraction of a quadrat covered by
vegetation, and $X$ follows a beta model with $a=2$ and $b=5$. In R these
parameters are `shape1` and `shape2`.

1. Calculate the model mean and variance. Express the mean as a percentage of
   the quadrat.
2. Calculate $P(X<0.20)$ and the 95th percentile of cover. Explain both results.
3. Simulate 500 cover fractions with seed 303. Plot a density-scaled histogram
   and add the theoretical beta density.
4. A field survey also records some quadrats with exactly zero cover. Explain
   why the proposed beta model cannot describe those observations directly.

# Exercise 4: Two measurements on each tree

Use this code to simulate heights in metres and diameters in centimetres for
300 trees. Each position in the two vectors describes the **same** tree. The
simulation assumes that height and diameter tend to increase together.

```{r}
set.seed(seed = 404)
n_trees = 300
z_height = rnorm(n = n_trees)
z_other = rnorm(n = n_trees)

tree_height = 22 + 4 * z_height
tree_diameter = 30 + 5 * (0.7 * z_height + sqrt(1 - 0.7^2) * z_other)
```

1. Plot height against diameter as a scatterplot. Describe the relationship
   and one feature that the plot does not establish.
2. Plot a histogram of all heights. Does it show a joint, marginal, or
   conditional distribution?
3. Compare height histograms for trees with diameters below 27 cm and above
   33 cm. What type of distributions are these, and how do they differ?
4. Use the density grid below to make a contour plot with `contour()`. Describe
   what the tilted contours say about the model. Does a contour enclose a
   specified probability?

The grid gives model density for combinations of diameter and height. You
only need to plot and interpret it; deriving its formula is beyond this chapter.

```{r}
diameter_grid = seq(from = 10, to = 50, length.out = 100)
height_grid = seq(from = 8, to = 36, length.out = 100)
rho = 0.7

joint_density = outer(
  X = diameter_grid, Y = height_grid,
  FUN = function(diameter, height) {
    dx = (diameter - 30) / 5
    hy = (height - 22) / 4
    exp(-(dx^2 - 2 * rho * dx * hy + hy^2) / (2 * (1 - rho^2))) /
      (2 * pi * 5 * 4 * sqrt(1 - rho^2))
  }
)
```
