---
title: "Chapter 3: From samples to estimates"
date: today
---

Exercise 1 uses field data on barn owl nests and the remaining exercises use
simulated ecological measurements. Work through the first example before trying
the remaining exercises. The ideas and notation are introduced in the
[Chapter 3 theory](../Theory.qmd).

# Exercise 1: Calls of barn owl nestlings

The data set `Owls` in the package `glmmTMB` contains observations of barn owl
nests (Roulin and Bersier 2007). For each visit of a parent to a nest, the variable
`SiblingNegotiation` is the number of calls made by the nestlings. We keep the visits
of the female parent to nests where the nestlings had been fed (`Satiated`), so that
all counts can reasonably share one mean:

```{r}
# Load the Owls data from the glmmTMB package without attaching the package.
data(Owls, package = "glmmTMB")

# Keep the visits of the female parent to nests with satiated nestlings.
satiated_female = Owls[Owls$FoodTreatment == "Satiated" & Owls$SexParent == "Female", ]

# Number of calls at each of these visits.
call_counts = satiated_female$SiblingNegotiation
```

Let $X_i$ be the number of calls at visit $i$. The counts are much more variable than a
Poisson distribution allows, so we use a
[Negative Binomial](../../Supplements/distributions.qmd#sec-negative_binomial) model with
mean $\mu$ and shape $k$,

$$
X_i \sim \operatorname{NegBin}(\mu, k), \qquad \operatorname{Var}(X_i)=\mu+\frac{\mu^2}{k},
$$

and assume that the visits are i.i.d.

1. Estimate the parameters of the Negative Binomial using the method of moments.
2. Calculate the estimated standard error of the mean. Explain how it differs from the
   standard deviation of the individual counts.
3. Construct an approximate 95% Wald confidence interval for $\mu$ using `qnorm()`.
   Interpret the interval and its confidence level. Name one model assumption that matters
   for this interpretation.

## Solution

### Question 1

The observational unit is one visit of the female parent to a nest. The model has two
parameters. The mean $\mu=E[X_i]$ is the expected number of calls per visit, and the shape
$k$ controls the overdispersion: the smaller $k$ is, the larger the variance is relative to
the mean. The method of moments needs two equations, so we equate the mean and the variance
of the model with those of the empirical distribution. The first gives $\hat\mu=\overline X$.
The second is $\hat\sigma^2=\hat\mu+\hat\mu^2/\hat k$, which we solve for the shape:

$$
\hat k = \frac{\hat\mu^2}{\hat\sigma^2-\hat\mu},
$$

where $\hat\sigma^2=\frac{1}{n}\sum_{i=1}^{n}(X_i-\overline X)^2$ is the variance of the
empirical distribution. The estimate of $k$ exists only if the variance of the sample is
larger than its mean as it cannot be negative.

```{r}
# Number of visits and the mean of the sample.
n_visits = length(call_counts)
mu_hat = mean(call_counts)

# Variance of the empirical distribution (divisor n), as in the method of moments.
variance_hat = mean((call_counts - mu_hat)^2)

# Solve variance = mu + mu^2 / k for the shape k.
k_hat = mu_hat^2 / (variance_hat - mu_hat)

c(n_visits = n_visits, mu_hat = mu_hat, variance_hat = variance_hat, k_hat = k_hat)
```

The 122 counts sum to 580, so the estimate of the mean is $580/122\approx 4.75$ calls per
visit. The variance of the sample is approximately 42.3, almost nine times the mean, which
gives $\hat k\approx 0.60$.

### Question 2

For i.i.d. observations, $\operatorname{Var}(\overline X)=\operatorname{Var}(X_i)/n$. Under
the Negative Binomial model $\operatorname{Var}(X_i)=\mu+\mu^2/k$, and we do not know either
parameter, so we insert the estimates:

$$
\widehat{\operatorname{SE}}(\hat\mu)
= \sqrt{\frac{\hat\mu+\hat\mu^2/\hat k}{n}}.
$$

```{r}
# Plug the estimates of the mean and the shape into the standard error.
estimated_se = sqrt((mu_hat + mu_hat^2 / k_hat) / n_visits)

# Standard deviation of the individual counts.
count_sd = sd(call_counts)

# Standard error that a Poisson model would give, for comparison.
poisson_se = sqrt(mu_hat / n_visits)

c(estimated_se = estimated_se, count_sd = count_sd, poisson_se = poisson_se)
```

The estimated standard error is approximately 0.59 calls per visit. Because $\hat k$ comes
from the method of moments, $\hat\mu+\hat\mu^2/\hat k$ is the variance of the sample, so
this value is the same as $\sqrt{\hat\sigma^2/n}$. It describes the variation of the
estimated **mean count** across hypothetical experiments with 122 visits each. The standard
deviation of the counts, approximately 6.5 calls, describes the variation among the
**individual visits**; it does not decrease when we observe more visits, whereas the
standard error does. The Poisson formula would give 0.20 calls per visit, about a third of
the value above, because it assumes that the variance equals the mean and ignores the
overdispersion.

### Question 3

For a Wald interval, use the 97.5th percentile of a standard Normal distribution on each
side of the estimate:

```{r}
# 0.975 quantile of the standard Normal distribution.
critical_value = qnorm(p = 0.975)

# Lower and upper endpoints of the interval for the mean number of calls.
mu_interval = c(
  lower = mu_hat - critical_value * estimated_se,
  upper = mu_hat + critical_value * estimated_se
)
mu_interval
```

The endpoints are approximately 3.60 and 5.91 calls per visit. This is a range for the
**mean count** per visit, and most individual visits fall outside it. The 95% refers to
the procedure: if we repeated the study many times and calculated an interval from each
sample in this way, about 95% of the intervals would contain the fixed $\mu$, provided that
the model and the Normal approximation are adequate. The Normal approximation is less
accurate when the counts are as skewed as these.

The assumption that matters most here is independence. The 122 visits come from only 24
nests, and visits to the same nest are likely to resemble each other. Treating the
visits as i.i.d. is a simplification that we accept in this chapter; the models for
observations nested in groups are the subject of [Chapter 7](../../Chapter_7/Theory.qmd).

# Exercise 2: Asymptotic convergence under a Beta model

Let $X_i$ be the fraction of the area of leaf $i$ that has been eaten by herbivores, a
continuous variable between 0 and 1. Assume that the leaves are i.i.d. and that
$X_i \sim \operatorname{Beta}(a, b)$ with $a=2$ and $b=6$ (`shape1` and `shape2` in R). You
will simulate samples from this model with two sample sizes, 10 and 1,000 leaves, to see how
the properties of a sample behave as the sample size grows (see
[Samples and distributions](../Theory.qmd#sec-samples)). The mean, variance and R functions
of the Beta distribution are in the
[Beta entry](../../Supplements/distributions.qmd#sec-beta) of the supplement.

1. Set seed 123 and simulate one sample of 10 leaves and one sample of 1,000 leaves.
   Calculate the mean, median and variance of each sample.
2. Repeat question 1 four times, using the seeds 11, 22, 33 and 44.
   How much do the sample properties vary between repetitions?
3. Calculate the mean, median and variance of the model. Compare the sample properties with these model
   values at the two sample sizes. Do the sample mean and the sample median approach the
   same value? What happens to the variation across repeated samples as the sample size
   increases?

# Exercise 3: Variation in tree heights

Suppose heights of independently sampled trees follow a normal model
with mean $\mu=22$ m and standard deviation $\sigma=4$ m. These are
the simulation parameters; in a field study they would be unknown.

1. Set seed 433 and simulate 25 heights. Estimate $\mu$ by method of
   moments. Estimate $\sigma^2$ using the mean squared deviation
   from the sample mean, with divisor $n$. Compare this variance
   estimate with `var()`. Give their units.
2. Set seed 434 and repeat the 25-tree study 2,000 times. For each
   study, calculate both variance estimates. Compare their average
   values and simulated biases with the true variance. The expected
   bias of the method-of-moments estimate is $-\sigma^2/n$.
3. Calculate the mean squared error of each variance estimator in
   the simulation. Does knowing which estimator is unbiased, by
   itself, tell you which has the smaller mean squared error?

Hint: `replicate()` can return two named estimates for each study.

# Exercise 4: Mean body mass and interval coverage

Suppose independent small-mammal body masses in grams follow a
lognormal model with `meanlog = log(30)` and `sdlog = 0.8`.
As described in the [LogNormal entry](../../Supplements/distributions.qmd#sec-lognormal) of the supplement,
the model median on the original scale is $e^{\mu}$ and the
**mean mass** is $e^{\mu+\sigma^2/2}$, where $\mu$ and $\sigma$
describe log mass. The parameters are known because you will simulate
the studies.

For the mean of independent Normal observations we know the sampling
distribution analytically, which makes this a special case. It gives an
interval with exactly 95% coverage, the **$t$ interval**:

$$
\overline X \pm t_{n-1,\,0.975}\frac{s}{\sqrt n},
$$

where $\overline X$ is the sample mean, $s$ the sample standard
deviation (`sd()`), $n$ the sample size and $t_{n-1,\,0.975}$ the $0.975$
quantile of a Student's $t$ distribution with $n-1$ degrees of freedom
(`qt()` in R). This distribution resembles a standard Normal distribution
with heavier tails, so the interval is wider than a Wald interval for
small $n$, and the two coincide as $n$ grows. The body masses below are
not Normal, so here the coverage will only be approximate.

1. Calculate the model mean and median mass in grams. Set seed 443,
   simulate one study of 12 animals, and calculate its sample mean,
   sample median, and estimated standard error of the sample mean.
2. For this study, form a 95% $t$ interval for the **original-scale
   mean mass** using the sample mean, `sd(mass) / sqrt(n)`, and
   `qt(p = 0.975, df = n - 1)`. Interpret the interval. Why is its
   coverage only approximate for these lognormal masses?
3. Set seed 444. Simulate 2,000 studies of 12 animals and 2,000
   studies of 100 animals. Recalculate the $t$ interval for every
   study and find the fraction covering the true original-scale mean
   at each size. Compare the fractions with 0.95 and explain why
   coverage might change with sample size.
4. State one reason why coverage in this simulation does not prove
   that the same procedure will work equally well for field data.

Hint: `replicate()` can return `TRUE` or `FALSE` for whether one
interval contains the model mean. `mean()` then gives the coverage
fraction.
