---
title: "Chapter 3: From samples to estimates — solutions"
date: today
---

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions use the
same models, tasks, and seeds as Exercises 2–4 on that page.

# Exercise 2: Asymptotic convergence under a Beta model

Let $X_i$ be the fraction of the area of leaf $i$ that has been eaten by herbivores, a
continuous variable between 0 and 1. Assume that the leaves are i.i.d. and that
$X_i \sim \operatorname{Beta}(a, b)$ with $a=2$ and $b=6$ (`shape1` and `shape2` in R). You
will simulate samples from this model with two sample sizes, 10 and 1,000 leaves, to see how
the properties of a sample behave as the sample size grows (see
[Samples and distributions](../Theory.qmd#sec-samples)). The mean, variance and R functions
of the Beta distribution are in the
[Beta entry](../../Supplements/distributions.qmd#sec-beta) of the supplement.

1. Set seed 123 and simulate one sample of 10 leaves and one sample of 1,000 leaves.
   Calculate the mean, median and variance of each sample (use the variance of the
   empirical distribution, with divisor $n$, not `var()`).
2. Repeat question 1 four times, using the seeds 11, 22, 33 and 44 (tip: use a for loop).
   How much do the sample properties vary between repetitions?
3. Calculate the mean, median and variance of the model (the median is the 0.5 quantile,
   which you can obtain with `qbeta()`). Compare the sample properties with these model
   values at the two sample sizes. Do the sample mean and the sample median approach the
   same value? What happens to the variation across repeated samples as the sample size
   increases?

## Solution

### Question 1

The observational unit is one leaf. We simulate the fraction of eaten area of each leaf with
`rbeta()` and summarise the sample through its empirical distribution: the mean, the median
(the 0.5 quantile of the sample) and the variance with divisor $n$,
$\frac{1}{n}\sum_{i=1}^{n}(x_i-\bar x)^2$. The function `var()` would use $n-1$ instead.

```{r}
# Shape parameters of the Beta model and the two sample sizes.
shape_a = 2
shape_b = 6
sample_sizes = c(10, 1000)

# Table that will store one row of summaries per sample.
first_results = data.frame()

set.seed(seed = 123)

# Simulate one sample per sample size and store its mean, median and variance.
for (sample_size in sample_sizes) {
  leaf_damage = rbeta(n = sample_size, shape1 = shape_a, shape2 = shape_b)
  sample_mean = mean(leaf_damage)

  first_results = rbind(
    first_results,
    data.frame(
      n = sample_size,
      mean = sample_mean,
      median = median(leaf_damage),
      variance = mean((leaf_damage - sample_mean)^2)
    )
  )
}

first_results
```

The sample of 10 leaves has a mean of approximately 0.263, a median of 0.237 and a variance
of 0.0224. The sample of 1,000 leaves has a mean of approximately 0.253, a median of 0.233
and a variance of 0.0210. Both variances are fractions squared, so they have no units.

### Question 2

Repeat the simulation with each of the four seeds. Setting the seed before each repetition
makes the result reproducible, and a different seed gives a different (but equally valid)
sample:

```{r}
seeds = c(11, 22, 33, 44)

# Table that will store one row per seed and sample size.
repeated_results = data.frame()

# The outer loop repeats the whole exercise; the inner loop covers the two sample sizes.
for (seed in seeds) {
  set.seed(seed = seed)

  for (sample_size in sample_sizes) {
    leaf_damage = rbeta(n = sample_size, shape1 = shape_a, shape2 = shape_b)
    sample_mean = mean(leaf_damage)

    repeated_results = rbind(
      repeated_results,
      data.frame(
        seed = seed,
        n = sample_size,
        mean = sample_mean,
        median = median(leaf_damage),
        variance = mean((leaf_damage - sample_mean)^2)
      )
    )
  }
}

repeated_results
```

With 10 leaves the sample properties change considerably between repetitions: the mean
ranges from approximately 0.19 to 0.26, the median from 0.17 to 0.23 and the variance from
0.0085 to 0.0209. With 1,000 leaves they are almost the same in every repetition: the mean
ranges from 0.246 to 0.251, the median from 0.223 to 0.233 and the variance from 0.0197 to
0.0206.

### Question 3

The mean and the variance of the Beta model are $E[X]=a/(a+b)$ and
$\operatorname{Var}(X)=ab/((a+b)^2(a+b+1))$. The median has no simple formula, so we
calculate it as the 0.5 quantile. We summarise the variation across the four repetitions
with the standard deviation of each sample property:

```{r}
# Model values, which do not depend on the sample size.
model_mean = shape_a / (shape_a + shape_b)
model_median = qbeta(p = 0.5, shape1 = shape_a, shape2 = shape_b)
model_variance = shape_a * shape_b /
  ((shape_a + shape_b)^2 * (shape_a + shape_b + 1))

model_values = data.frame(
  n = sample_sizes,
  mean = model_mean,
  median = model_median,
  variance = model_variance
)
model_values

# Standard deviation across the four repetitions for each sample size.
aggregate(
  cbind(mean, median, variance) ~ n,
  data = repeated_results,
  FUN = sd
)
```

The model values are the same for both sample sizes: a mean of 0.25, a median of
approximately 0.228 and a variance of approximately 0.0208 (a standard deviation of 0.144).
The Beta model with $a<b$ is skewed to the right, so its mean is larger than its median.
Because of that, the sample mean and the sample median do not approach the same value. As
the sample size grows, the sample mean converges to 0.25 and the sample median to 0.228,
which is why every median in the previous question, even for 1,000 leaves, is clearly below
0.25. Both statistics converge, but to different features of the model.

The variation across repetitions decreases with the sample size for all three properties.
The standard deviation of the sample mean across the four repetitions falls from
approximately 0.032 for 10 leaves to 0.002 for 1,000 leaves, that of the median from 0.034
to 0.005 and that of the variance from 0.0055 to 0.0004. In these repetitions the median
varies more than the mean at 1,000 leaves, but four repetitions are too few to conclude
that this holds in general. Finally, the variance with divisor $n$ has an expected value of
$\frac{n-1}{n}\operatorname{Var}(X)$, which is approximately 0.0188 for $n=10$ and 0.0208
for $n=1{,}000$. This explains why the variances of the small samples tend to fall slightly
below the model value, a difference that becomes negligible as the sample size increases.

# Exercise 3: Variation in tree heights

The data set `trees`, which is part of the `datasets` package that comes with R,
contains measurements of 31 felled black cherry trees. We use the column `Height`, the height
of each tree in feet (ft). Let $X_i$ be the height of tree $i$ and assume that the trees are
i.i.d. with $X_i \sim \operatorname{Normal}(\mu,\sigma^2)$, where $\mu$ is the mean of the
model (in ft) and $\sigma^2$ its variance (in ft²). Remember that the argument `sd` of
`rnorm()` takes the standard deviation $\sigma$. See
[Estimation of two parameters](../Theory.qmd#sec-two_parameters) for the method of moments
with this model.

With a single sample we cannot tell how far an estimate is from the parameter, because the
parameter is unknown. To study the behaviour of an estimator we can simulate hypothetical
experiments from a model whose parameters we know. Here we use the estimates from the
observed trees as the parameters of that model (a plug-in estimate, as for the standard
error in [Standard errors](../Theory.qmd#sec-standard_errors)). The simulation then shows how the estimators would behave if
the fitted model had generated the data. The results approximate the behaviour of the
estimators for the real trees, and the approximation is better the closer the estimates are
to the parameters.

1. Load the data with `data(trees, package = "datasets")`. Estimate $\mu$ by method of
   moments. Estimate $\sigma^2$ in two ways: as the variance of the empirical distribution
   (i.e. with $n$ in the denominator) and with `var()` (i.e., with $n-1$ in the denominator).
2. Set seed 434 and simulate 2,000 hypothetical experiments, each with 31 trees from a Normal
   model whose mean and variance are the method-of-moments estimates of question 1. For each
   experiment, calculate both variance estimates. Compare their average values and simulated
   biases with the variance that you used to simulate.
3. Calculate the mean squared error of each variance estimator in the simulation. Does
   knowing which estimator is unbiased, by itself, tell you which has the smaller mean
   squared error?

## Solution

### Question 1

The observational unit is one tree, and the data are the 31 values of `Height`, measured in
feet. The method of moments matches the mean of the model, $E[X]=\mu$, to the sample mean,
and the variance of the model, $\operatorname{Var}(X)=\sigma^2$, to the mean squared
deviation from the sample mean, which uses divisor $n$. R's `var()` uses $n-1$ instead.

```{r}
# Heights (in ft) of the 31 trees and the sample size.
data(trees, package = "datasets")
tree_heights = trees$Height
n_trees = length(tree_heights)

# Method-of-moments estimates, and the variance calculated by var().
mu_hat = mean(tree_heights)
variance_mm = mean((tree_heights - mu_hat)^2)
variance_with_var = var(tree_heights)

c(
  estimated_mean_ft = mu_hat,
  variance_mm_ft2 = variance_mm,
  variance_with_var_ft2 = variance_with_var
)
```

The estimated mean is 76 ft. The method-of-moments estimate of the variance is
approximately 39.29 ft² and the one from `var()` is 40.60 ft². Both variances are in square
feet, and they differ because `var()` divides by $n-1=30$ instead of $n=31$ (so the second
is the first multiplied by $31/30$). With a single sample we cannot say which of them is
closer to the unknown variance of the model.

### Question 2

We now use the estimates of question 1 as the parameters of the model that generates the
hypothetical experiments. In this simulation, the "true" mean is `mu_hat` and the "true"
variance is `variance_mm`, which is the value that the estimates are compared with. Note
that `rnorm()` takes the standard deviation, so we pass the square root of the variance.
Each hypothetical experiment simulates 31 trees and returns a pair of variance estimates.

```{r}
set.seed(seed = 434)
n_experiments = 2000

# Each run simulates one experiment of 31 trees and returns both variance estimates.
variance_estimates = replicate(
  n = n_experiments,
  expr = {
    heights = rnorm(
      n = n_trees, mean = mu_hat, sd = sqrt(variance_mm)
    )
    sample_mean = mean(heights)
    c(
      method_of_moments = mean((heights - sample_mean)^2),
      with_var = var(heights)
    )
  }
)

# Average of each estimator across experiments and its simulated bias.
average_variance = rowMeans(variance_estimates)
simulated_bias = average_variance - variance_mm

data.frame(
  estimator = rownames(variance_estimates),
  average_variance_ft2 = average_variance,
  simulated_bias_ft2 = simulated_bias,
  row.names = NULL
)
```

The method-of-moments estimator averages approximately 37.56 ft², below the 39.29 ft² used
to simulate. Its simulated bias is approximately $-1.73$ ft², while the expected bias is
$-39.29/31 = -1.27$ ft². The estimator from `var()` averages approximately 38.81 ft², with
a simulated bias of $-0.48$ ft² rather than exactly zero.

The differences from the expected values are due to the finite number of hypothetical
experiments: each average is itself an estimate, and it would change with another seed. We
can calculate how much such an average varies. The standard error of the average of 2,000
estimates is the standard deviation of those estimates divided by $\sqrt{2000}$ (see
[Standard errors](../Theory.qmd#sec-standard_errors)), and the simulated bias has the same
standard error because it only subtracts a constant:

```{r}
# Standard deviation of each estimator across the hypothetical experiments.
sd_estimates = c(
  method_of_moments = sd(variance_estimates["method_of_moments", ]),
  with_var = sd(variance_estimates["with_var", ])
)

# Standard error of each average, and therefore of each simulated bias.
se_simulated_bias = sd_estimates / sqrt(n_experiments)
se_simulated_bias

# Distance between the simulated and the expected bias, in standard errors.
expected_bias = c(method_of_moments = -variance_mm / n_trees, with_var = 0)
(simulated_bias - expected_bias) / se_simulated_bias
```

The standard error is approximately 0.22 ft² for the method-of-moments estimator and
0.23 ft² for `var()`. So both simulated biases are about two standard errors (2.1) below
their expected values. A deviation of this size is unusual, but it is still within what
2,000 experiments can produce, and a simulation with more experiments would come closer to
the expected values. The simulation shows the bias of the
estimators for the fitted model; the real model has an unknown variance that is probably
somewhat different from 39.29 ft², so the bias of the real estimators will also differ.

### Question 3

The mean squared error (MSE) of an estimator is the average squared distance between its
estimates and the parameter. It includes both the variance of the estimator across
experiments and its squared bias. We calculate it from the same simulated estimates:

```{r}
# Average squared distance between each estimate and the variance used to simulate.
simulated_mse = rowMeans((variance_estimates - variance_mm)^2)
simulated_mse
```

The method-of-moments estimator has a simulated MSE of approximately 101.1 ft⁴, compared
with approximately 105.0 ft⁴ for `var()` (the units are feet to the fourth power because the
estimates are squared feet). So the estimator from `var()` is unbiased but has the larger
MSE, because the method-of-moments estimator varies less across experiments by more than
enough to compensate for its bias. Knowing that one estimator is unbiased does not, by
itself, tell us which has the smaller MSE.

# Exercise 4: Mean body mass and interval coverage

Suppose that the body masses (in grams) of small mammals are independent and follow a
LogNormal model with `meanlog = log(30)` and `sdlog = 0.8`. As described in the
[LogNormal entry](../../Supplements/distributions.qmd#sec-lognormal) of the supplement, the
model median on the original scale is $e^{\mu}$ and the **mean mass** is
$e^{\mu+\sigma^2/2}$, where $\mu$ and $\sigma$ describe log mass.  In this exercise, you
calculate approximate 95% Wald confidence intervals for the mean mass (see
[Confidence intervals](../Theory.qmd#sec-confidence_intervals)) and check how often they
contain it.

1. Calculate the model mean and median mass in grams. Set seed 443, simulate one study of 12
   animals, and calculate its sample mean, sample median, and estimated standard error of the
   sample mean (`sd(mass) / sqrt(n)`).
2. For this study, calculate a 95% Wald interval for the **original-scale mean mass** using
   `qnorm()`. Interpret the interval. Why is its coverage only approximate for these
   LogNormal masses?
3. Set seed 444. Simulate 2,000 studies of 12 animals and 2,000 studies of 100 animals.
   Calculate the Wald interval for every study and find the fraction of intervals that
   contain the model mean at each sample size. Compare the fractions with 0.95 and explain
   why coverage might change with sample size.
4. State one reason why coverage in this simulation does not prove that the same procedure
   will work equally well for field data.

## Solution

### Question 1

The parameters of the LogNormal model describe log mass, so on the gram scale the model median
is $e^{\mu}$ and the model mean is $e^{\mu+\sigma^2/2}$. The distribution is skewed to the
right, which makes the mean larger than the median.

```{r}
# Parameters of the LogNormal model (on the log scale).
mass_meanlog = log(30)
mass_sdlog = 0.8

# Median and mean of the model on the gram scale.
true_median = exp(mass_meanlog)
true_mean = exp(mass_meanlog + mass_sdlog^2 / 2)
c(median_g = true_median, mean_g = true_mean)
```

The model median is 30 g and the model mean is approximately 41.3 g. Next, we simulate one
study of 12 animals. For i.i.d. observations $\operatorname{Var}(\overline X)=
\operatorname{Var}(X_i)/n$, so we estimate the standard error of the sample mean as
$s/\sqrt n$, where $s$ is the standard deviation calculated by `sd()`.

```{r}
set.seed(seed = 443)
n_animals = 12
body_mass = rlnorm(
  n = n_animals,
  meanlog = mass_meanlog, sdlog = mass_sdlog
)

# Sample mean, sample median and estimated standard error of the sample mean.
sample_mean = mean(body_mass)
sample_median = median(body_mass)
estimated_se = sd(body_mass) / sqrt(n_animals)
c(
  sample_mean_g = sample_mean,
  sample_median_g = sample_median,
  estimated_se_g = estimated_se
)
```

The sample mean is approximately 34.4 g and the sample median approximately 25.4 g, both below
the model values. The estimated standard error is approximately 9.2 g. It describes the
variation of the sample mean across hypothetical experiments with 12 animals each, whereas the
standard deviation of the 12 masses describes the variation among individual animals and does
not decrease when more animals are measured.

### Question 2

A Wald interval adds and subtracts $z_{0.975}$ estimated standard errors from the estimate,
where $z_{0.975}\approx 1.96$ is the 0.975 quantile of the standard Normal distribution:

```{r}
# 0.975 quantile of the standard Normal distribution.
critical_value = qnorm(p = 0.975)

# Lower and upper endpoints of the interval for the mean mass.
mass_interval = c(
  lower = sample_mean - critical_value * estimated_se,
  upper = sample_mean + critical_value * estimated_se
)
mass_interval
```

The interval runs from approximately 16.3 to 52.4 g. It is a range of candidate values for the
**mean mass in grams** and does not describe the spread of individual masses. The 95% refers
to the procedure: if we repeated the study many times and calculated an interval from each
sample in this way, about 95% of the intervals would contain the model mean, provided that
the sampling distribution of the sample mean is approximately Normal. Here the interval
contains the model mean of 41.3 g, which we know only because we simulated the data.

LogNormal masses are skewed to the right, so with only 12 animals the sampling distribution of
the sample mean is also skewed and the Normal approximation can be poor. The coverage of the
interval is therefore only approximate, and we check it by simulation.

### Question 3

For each simulated study, we calculate a new interval and record whether it contains the model
mean. The fraction of `TRUE` values is the simulated coverage.

```{r}
# Simulated coverage of the 95% Wald interval for a given number of animals.
coverage_for_size = function(n_animals, n_studies = 2000) {
  covered = replicate(
    n = n_studies,
    expr = {
      # One hypothetical study: simulate masses and calculate the interval.
      mass = rlnorm(
        n = n_animals,
        meanlog = mass_meanlog, sdlog = mass_sdlog
      )
      estimate = mean(mass)
      se = sd(mass) / sqrt(n_animals)
      lower = estimate - critical_value * se
      upper = estimate + critical_value * se
      # TRUE if the interval contains the model mean.
      lower <= true_mean && true_mean <= upper
    }
  )
  mean(covered)
}

set.seed(seed = 444)
simulated_coverage = c(
  n_12 = coverage_for_size(n_animals = 12),
  n_100 = coverage_for_size(n_animals = 100)
)
simulated_coverage
```

The simulated coverage is approximately 0.864 for 12 animals and 0.924 for 100 animals, both
below the nominal 0.95. These fractions depend on the seed: with 2,000 studies, the
simulation error of a coverage near 0.9 is approximately 0.007. With 12 animals, a sample that
lacks the few very large masses has both a small sample mean and a small estimated standard
error. Its interval is then too narrow and lies below the model mean, which is how most of
the misses occur. With 100 animals the sampling distribution of the sample mean is closer to
Normal (see [Consistency and asymptotic normality](../Theory.qmd#sec-large_samples)), so the
coverage is closer to 0.95, although at this sample size it is still not equal to it.

### Question 4

These coverages hold under the specified LogNormal model with independent animals. Real body
masses may follow a different distribution, so the shape of the sampling distribution of the
mean, and therefore the coverage of the Wald interval, can differ from the simulated one.
Animals from the same site or litter would also violate the independence assumption, which
the simulation does not explore.
