---
title: "Chapter 3: Probability for continuous data — solutions"
date: today
---

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions follow the same
assumed models, tasks, and seeds as Exercises 2–4 on that page. The formulae used
below are in the LogNormal and Beta entries of the
[distribution supplement](../../Supplements/distributions.qmd).

# Exercise 2: Body mass under a lognormal model


The lognormal parameters describe $\log(X)$. On the original scale, the
median is $e^\mu$ and the mean is $e^{\mu+\sigma^2/2}$, where `meanlog` is
$\mu$ and `sdlog` is $\sigma$.

```{r}
mass_meanlog = log(30)
mass_sdlog = 0.5

median_mass = exp(mass_meanlog)
mean_mass = exp(mass_meanlog + mass_sdlog^2 / 2)
c(median_g = median_mass, mean_g = mean_mass)
```

The median is 30 g and the mean is approximately 34 g. A few possible
large masses in the right tail pull the mean above the median.

`plnorm()` gives the cumulative probability. For mass *above* 50 g, we
request its upper tail. `qlnorm()` gives the mass below which 95% of the
model's probability falls.

```{r}
prob_over_50 = plnorm(
  q = 50, meanlog = mass_meanlog, sdlog = mass_sdlog,
  lower.tail = FALSE
)
mass_95 = qlnorm(p = 0.95, meanlog = mass_meanlog, sdlog = mass_sdlog)
c(prob_over_50 = prob_over_50, percentile_95_g = mass_95)
```

These are predictions under the assumed model for a randomly selected
animal's mass. The quantile is a mass in grams, not a probability.

Generate a sample and compare its shape on the two scales. The model density
must be placed over a **density-scaled** histogram on the original scale.

```{r}
set.seed(seed = 202)
body_mass = rlnorm(n = 500, meanlog = mass_meanlog, sdlog = mass_sdlog)

old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

hist(
  x = body_mass, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Body mass (g)", ylab = "Density per gram",
  main = "Original scale"
)
curve(
  dlnorm(x = x, meanlog = mass_meanlog, sdlog = mass_sdlog),
  add = TRUE, lwd = 2
)

hist(
  x = log(body_mass), breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Log body mass", ylab = "Density",
  main = "Log scale"
)

par(old_par)
```

The original masses are positive and right-skewed; their logarithms are
more symmetric. The histogram and curve need not agree exactly for 500
simulated values. Because the data were generated from this very model, this
comparison checks what sampling variation looks like. It cannot establish
that a lognormal model fits body masses measured in a field study.

# Exercise 3: Vegetation cover under a beta model


For beta shape parameters $a=2$ and $b=5$, the model mean and variance are
$a/(a+b)$ and $ab/[(a+b)^2(a+b+1)]$.

```{r}
cover_a = 2
cover_b = 5

mean_cover = cover_a / (cover_a + cover_b)
variance_cover = cover_a * cover_b /
  ((cover_a + cover_b)^2 * (cover_a + cover_b + 1))

c(mean_fraction = mean_cover,
  mean_percent = 100 * mean_cover,
  variance_fraction_squared = variance_cover)
```

The expected fraction covered is approximately 0.286, or 28.6% of a
quadrat. The variance describes variation in squared fraction units.

```{r}
prob_below_20_percent = pbeta(
  q = 0.20, shape1 = cover_a, shape2 = cover_b
)
cover_95 = qbeta(p = 0.95, shape1 = cover_a, shape2 = cover_b)

c(prob_below_20_percent = prob_below_20_percent,
  percentile_95_fraction = cover_95)
```

The first result is the model probability of less than 20% cover. The
second is a cover fraction below which 95% of modelled quadrats fall.
For a continuous model, using $<$ or $\leq$ at 0.20 gives the same
probability.

```{r}
set.seed(seed = 303)
cover_fraction = rbeta(n = 500, shape1 = cover_a, shape2 = cover_b)

hist(
  x = cover_fraction, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Vegetation cover (fraction)", ylab = "Density",
  main = "Simulated cover and beta model"
)
curve(
  dbeta(x = x, shape1 = cover_a, shape2 = cover_b),
  add = TRUE, lwd = 2
)
```

The histogram shows one simulated sample. Its bars need not follow the
theoretical curve exactly. A standard continuous beta model assigns zero
probability to cover of exactly zero or one. Before modelling field
observations at those boundaries, we would ask whether they represent
genuine bare or fully covered quadrats and account for how they arose.

# Exercise 4: Two measurements on each tree


Recreate the paired measurements with the exercise's seed. The values at
the same position in the vectors belong to one simulated tree.

```{r}
set.seed(seed = 404)
n_trees = 300
z_height = rnorm(n = n_trees)
z_other = rnorm(n = n_trees)

tree_height = 22 + 4 * z_height
tree_diameter = 30 + 5 * (0.7 * z_height + sqrt(1 - 0.7^2) * z_other)

plot(
  x = tree_diameter, y = tree_height, pch = 16,
  col = grDevices::adjustcolor("black", alpha.f = 0.35),
  xlab = "Tree diameter (cm)", ylab = "Tree height (m)",
  main = "Simulated trees"
)
```

Heights and diameters tend to increase together, although there is
variation among trees. The scatterplot shows an association in a simulated
sample. It does not establish that changing a tree's diameter would cause
its height to change.

A histogram of *all* heights estimates the marginal distribution of height:

```{r}
hist(
  x = tree_height, breaks = 15,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Number of trees",
  main = "All simulated trees"
)
```

Now select trees by diameter and compare their height distributions. These
are **conditional on diameter ranges**, rather than height distributions
for all trees.

```{r}
small_trees = tree_diameter < 27
large_trees = tree_diameter > 33
height_limits = range(tree_height)

old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

hist(
  x = tree_height[small_trees], breaks = 10,
  xlim = height_limits, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Density",
  main = "Diameter below 27 cm"
)
hist(
  x = tree_height[large_trees], breaks = 10,
  xlim = height_limits, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Density",
  main = "Diameter above 33 cm"
)

par(old_par)

c(
  mean_height_small_m = mean(tree_height[small_trees]),
  mean_height_large_m = mean(tree_height[large_trees])
)
```

The larger-diameter group tends to contain taller trees in this simulation.
The common horizontal scale makes their positions easier to compare. The
histograms are sample summaries, so their shapes also vary with sample size
and bin choice.

Finally, use the supplied grid to display the **model's joint density**.

```{r}
diameter_grid = seq(from = 10, to = 50, length.out = 100)
height_grid = seq(from = 8, to = 36, length.out = 100)
rho = 0.7

joint_density = outer(
  X = diameter_grid, Y = height_grid,
  FUN = function(diameter, height) {
    dx = (diameter - 30) / 5
    hy = (height - 22) / 4
    exp(-(dx^2 - 2 * rho * dx * hy + hy^2) / (2 * (1 - rho^2))) /
      (2 * pi * 5 * 4 * sqrt(1 - rho^2))
  }
)

contour(
  x = diameter_grid, y = height_grid, z = joint_density,
  nlevels = 8,
  xlab = "Tree diameter (cm)", ylab = "Tree height (m)",
  main = "Assumed joint density"
)
```

Each line connects combinations with equal **density**. The tilted
contours show that greater diameter tends to accompany greater height
under the model. A line or the region inside it does not automatically
correspond to a specified probability; probability for a region depends
on the density across its area.
