---
title: "Chapter 3: From samples to estimates — solutions"
date: today
---

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions use the
same models, tasks, and seeds as Exercises 2–4 on that page.

# Exercise 2: Asymptotic convergence under a Beta model

Let $X_i$ be the fraction of the area of leaf $i$ that has been eaten by herbivores, a
continuous variable between 0 and 1. Assume that the leaves are i.i.d. and that
$X_i \sim \operatorname{Beta}(a, b)$ with $a=2$ and $b=6$ (`shape1` and `shape2` in R). You
will simulate samples from this model with two sample sizes, 10 and 1,000 leaves, to see how
the properties of a sample behave as the sample size grows (see
[Samples and distributions](../Theory.qmd#sec-samples)). The mean, variance and R functions
of the Beta distribution are in the
[Beta entry](../../Supplements/distributions.qmd#sec-beta) of the supplement.

1. Set seed 123 and simulate one sample of 10 leaves and one sample of 1,000 leaves.
   Calculate the mean, median and variance of each sample (use the variance of the
   empirical distribution, with divisor $n$, not `var()`).
2. Repeat question 1 four times, using the seeds 11, 22, 33 and 44 (tip: use a for loop).
   How much do the sample properties vary between repetitions?
3. Calculate the mean, median and variance of the model (the median is the 0.5 quantile,
   which you can obtain with `qbeta()`). Compare the sample properties with these model
   values at the two sample sizes. Do the sample mean and the sample median approach the
   same value? What happens to the variation across repeated samples as the sample size
   increases?

## Solution

### Question 1

The observational unit is one leaf. We simulate the fraction of eaten area of each leaf with
`rbeta()` and summarise the sample through its empirical distribution: the mean, the median
(the 0.5 quantile of the sample) and the variance with divisor $n$,
$\frac{1}{n}\sum_{i=1}^{n}(x_i-\bar x)^2$. The function `var()` would use $n-1$ instead.

```{r}
# Shape parameters of the Beta model and the two sample sizes.
shape_a = 2
shape_b = 6
sample_sizes = c(10, 1000)

# Table that will store one row of summaries per sample.
first_results = data.frame()

set.seed(seed = 123)

# Simulate one sample per sample size and store its mean, median and variance.
for (sample_size in sample_sizes) {
  leaf_damage = rbeta(n = sample_size, shape1 = shape_a, shape2 = shape_b)
  sample_mean = mean(leaf_damage)

  first_results = rbind(
    first_results,
    data.frame(
      n = sample_size,
      mean = sample_mean,
      median = median(leaf_damage),
      variance = mean((leaf_damage - sample_mean)^2)
    )
  )
}

first_results
```

The sample of 10 leaves has a mean of approximately 0.263, a median of 0.237 and a variance
of 0.0224. The sample of 1,000 leaves has a mean of approximately 0.253, a median of 0.233
and a variance of 0.0210. Both variances are fractions squared, so they have no units.

### Question 2

Repeat the simulation with each of the four seeds. Setting the seed before each repetition
makes the result reproducible, and a different seed gives a different (but equally valid)
sample:

```{r}
seeds = c(11, 22, 33, 44)

# Table that will store one row per seed and sample size.
repeated_results = data.frame()

# The outer loop repeats the whole exercise; the inner loop covers the two sample sizes.
for (seed in seeds) {
  set.seed(seed = seed)

  for (sample_size in sample_sizes) {
    leaf_damage = rbeta(n = sample_size, shape1 = shape_a, shape2 = shape_b)
    sample_mean = mean(leaf_damage)

    repeated_results = rbind(
      repeated_results,
      data.frame(
        seed = seed,
        n = sample_size,
        mean = sample_mean,
        median = median(leaf_damage),
        variance = mean((leaf_damage - sample_mean)^2)
      )
    )
  }
}

repeated_results
```

With 10 leaves the sample properties change considerably between repetitions: the mean
ranges from approximately 0.19 to 0.26, the median from 0.17 to 0.23 and the variance from
0.0085 to 0.0209. With 1,000 leaves they are almost the same in every repetition: the mean
ranges from 0.246 to 0.251, the median from 0.223 to 0.233 and the variance from 0.0197 to
0.0206.

### Question 3

The mean and the variance of the Beta model are $E[X]=a/(a+b)$ and
$\operatorname{Var}(X)=ab/((a+b)^2(a+b+1))$. The median has no simple formula, so we
calculate it as the 0.5 quantile. We summarise the variation across the four repetitions
with the standard deviation of each sample property:

```{r}
# Model values, which do not depend on the sample size.
model_mean = shape_a / (shape_a + shape_b)
model_median = qbeta(p = 0.5, shape1 = shape_a, shape2 = shape_b)
model_variance = shape_a * shape_b /
  ((shape_a + shape_b)^2 * (shape_a + shape_b + 1))

model_values = data.frame(
  n = sample_sizes,
  mean = model_mean,
  median = model_median,
  variance = model_variance
)
model_values

# Standard deviation across the four repetitions for each sample size.
aggregate(
  cbind(mean, median, variance) ~ n,
  data = repeated_results,
  FUN = sd
)
```

The model values are the same for both sample sizes: a mean of 0.25, a median of
approximately 0.228 and a variance of approximately 0.0208 (a standard deviation of 0.144).
The Beta model with $a<b$ is skewed to the right, so its mean is larger than its median.
Because of that, the sample mean and the sample median do not approach the same value. As
the sample size grows, the sample mean converges to 0.25 and the sample median to 0.228,
which is why every median in the previous question, even for 1,000 leaves, is clearly below
0.25. Both statistics converge, but to different features of the model.

The variation across repetitions decreases with the sample size for all three properties.
The standard deviation of the sample mean across the four repetitions falls from
approximately 0.032 for 10 leaves to 0.002 for 1,000 leaves, that of the median from 0.034
to 0.005 and that of the variance from 0.0055 to 0.0004. In these repetitions the median
varies more than the mean at 1,000 leaves, but four repetitions are too few to conclude
that this holds in general. Finally, the variance with divisor $n$ has an expected value of
$\frac{n-1}{n}\operatorname{Var}(X)$, which is approximately 0.0188 for $n=10$ and 0.0208
for $n=1{,}000$. This explains why the variances of the small samples tend to fall slightly
below the model value, a difference that becomes negligible as the sample size increases.

# Exercise 3: Variation in tree heights

The method of moments matches the mean of the model, $E[X]=\mu$, to the
sample mean, and the variance of the model, $\operatorname{Var}(X)=\sigma^2$,
to the mean squared deviation from the sample mean, which uses divisor $n$.
R's `var()` uses $n-1$ instead.

```{r}
set.seed(seed = 433)
n_trees = 25
true_mu = 22
true_sigma = 4
true_variance = true_sigma^2
tree_heights = rnorm(
  n = n_trees, mean = true_mu, sd = true_sigma
)

mu_hat = mean(tree_heights)
variance_mm = mean((tree_heights - mu_hat)^2)
variance_with_var = var(tree_heights)
c(
  estimated_mean_m = mu_hat,
  variance_mm_m2 = variance_mm,
  variance_with_var_m2 = variance_with_var
)
```

The estimated mean is in metres; both variance estimates are in
square metres. They differ because `var()` adjusts the divisor.
Neither single-sample estimate needs to equal the model variance
of $4^2=16$ square metres.

To study bias, repeat the entire 25-tree study. Each repetition
returns a pair of variance estimates.

```{r}
set.seed(seed = 434)
n_studies = 2000
variance_estimates = replicate(
  n = n_studies,
  expr = {
    heights = rnorm(
      n = n_trees, mean = true_mu, sd = true_sigma
    )
    sample_mean = mean(heights)
    c(
      method_of_moments = mean((heights - sample_mean)^2),
      with_var = var(heights)
    )
  }
)

average_variance = rowMeans(variance_estimates)
simulated_bias = average_variance - true_variance
simulated_mse = rowMeans(
  (variance_estimates - true_variance)^2
)

variance_summary = data.frame(
  estimator = rownames(variance_estimates),
  average_variance_m2 = average_variance,
  simulated_bias_m2 = simulated_bias,
  simulated_mse_m4 = simulated_mse,
  row.names = NULL
)
variance_summary
```

The method-of-moments estimate averages about 15.14 square metres,
below the true 16. Its simulated bias is about $-0.86$ square
metres; its theoretical bias is $-16/25=-0.64$ square metres.
The estimate from `var()` averages about 15.77, with simulated
bias about $-0.23$ rather than exactly zero. The differences from
theoretical values reflect finite simulation.

MSE includes both variance across studies and squared bias. In
this simulation, the method-of-moments estimator has MSE about
$19.84\ \mathrm{m}^4$, compared with about $20.77\ \mathrm{m}^4$
for `var()`. Knowing that one estimator is unbiased does not,
by itself, settle which has smaller MSE.

# Exercise 4: Mean body mass and interval coverage

The lognormal parameters describe **log mass**. On the original
gram scale, the median is $e^\mu$ and the mean is
$e^{\mu+\sigma^2/2}$. Right skew makes the mean larger than the
median.

```{r}
mass_meanlog = log(30)
mass_sdlog = 0.8
true_median = exp(mass_meanlog)
true_mean = exp(mass_meanlog + mass_sdlog^2 / 2)
c(median_g = true_median, mean_g = true_mean)
```

The model median is 30 g and the model mean is approximately
41.3 g. Simulate one study and calculate the standard error of
its original-scale sample mean:

```{r}
set.seed(seed = 443)
n_animals = 12
body_mass = rlnorm(
  n = n_animals,
  meanlog = mass_meanlog, sdlog = mass_sdlog
)

sample_mean = mean(body_mass)
sample_median = median(body_mass)
estimated_se = sd(body_mass) / sqrt(n_animals)
c(
  sample_mean_g = sample_mean,
  sample_median_g = sample_median,
  estimated_se_g = estimated_se
)
```

This standard error describes uncertainty in the sample mean, not
variation among the 12 individual masses. The $t$ interval uses
the estimated standard error and $n-1$ degrees of freedom:

```{r}
critical_value = qt(p = 0.975, df = n_animals - 1)
mass_interval = c(
  lower = sample_mean - critical_value * estimated_se,
  upper = sample_mean + critical_value * estimated_se
)
mass_interval
```

The interval runs from approximately 14.1 to 54.6 g. Its endpoints
are candidate values for the **population mean mass in grams**,
not an interval containing 95% of animal masses.
This $t$ interval has exact 95% coverage for independent normal
observations. Lognormal masses are skewed, so here it is an
approximation; its actual coverage must be checked.

For each simulated study below, calculate a new interval and
record whether it contains the known model mean. The proportion
of `TRUE` results estimates coverage.

```{r}
coverage_for_size = function(n_animals, n_studies = 2000) {
  critical_value = qt(p = 0.975, df = n_animals - 1)
  covered = replicate(
    n = n_studies,
    expr = {
      mass = rlnorm(
        n = n_animals,
        meanlog = mass_meanlog, sdlog = mass_sdlog
      )
      estimate = mean(mass)
      se = sd(mass) / sqrt(n_animals)
      lower = estimate - critical_value * se
      upper = estimate + critical_value * se
      lower <= true_mean && true_mean <= upper
    }
  )
  mean(covered)
}

set.seed(seed = 444)
simulated_coverage = c(
  n_12 = coverage_for_size(n_animals = 12),
  n_100 = coverage_for_size(n_animals = 100)
)
simulated_coverage
```

The simulated coverage is 0.889 for 12 animals and 0.927 for 100
animals, both below 0.95. These fractions depend on the seed and
2,000 repetitions. With the more skewed sampling distribution at
12 animals, the usual $t$ interval misses the true mean more often
than its nominal 5% rate. At 100 animals, the sample mean tends
to have a more nearly normal sampling distribution, making this
approximation more plausible. These results describe performance
**under the specified lognormal model and independent sampling**.
Actual body masses or a field sampling design could depart from
both assumptions, so simulated coverage alone does not establish
coverage for a real study.
