---
title: "Deterministic functions in ecology"
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures in this chapter.
library(ggplot2)
```

# Introduction

In this chapter, we learn how to describe the relationship between an ecological response
and a predictor with a deterministic function, and how to find out the shape of such a
function from its formula. To work with a function we need to know where it starts and
where it ends, how steep it is, where it peaks or bends, and what each of its parameters
does to the curve. We answer these questions with limits and derivatives, and we check
the answers with R.

This chapter differs from the previous ones in that it contains no
random variables and no data analysis. [Chapter 2](../Chapter_2/Theory.qmd),
[Chapter 3](../Chapter_3/Theory.qmd) and [Chapter 4](../Chapter_4/Theory.qmd) built the
first pillar of the book, probability and likelihood, for samples in which every
observation had the same distribution. However, ecological questions are usually about how a
response changes with a predictor, so that the distribution of an observation depends on
the conditions under which it was made. This chapter builds the second pillar, the
functions that describe that change. The two
come together in [Chapter 6](../Chapter_6/Theory.qmd), where a parameter of a distribution
(usually its mean) follows one of these deterministic functions and the parameters of the
deterministic function are estimated by maximum likelihood. To fit such a model we first
need to know the candidate curves, what their parameters mean and which starting values to
give to the optimiser. That is why this chapter comes now, and why it is mostly pen and
paper.

This chapter analyses two functions in full, the Michaelis–Menten function and the Ricker
function. It also uses two data sets from the
predation experiments with reed frog tadpoles of Vonesh and Bolker (2005).^[Vonesh, J. R.
and Bolker, B. M. (2005), "Compensatory larval responses shift trade-offs associated with
predator-induced hatching plasticity", *Ecology* 86: 1580–1591. The data are in the
package `emdbook`.] The practical of Chapter 4 analysed the survival of tadpoles in one of
those experiments; here we use two others, in which the number of tadpoles killed was
recorded for different densities and for different body lengths of the tadpoles. The data
are only a backdrop for the functions. The parameter values of every curve in this chapter
were eyeballed, since how to estimate them from the data is the topic
of Chapter 6.

This chapter makes use of calculus and the analysis of functions. If the rules for exponents,
logarithms, limits and derivatives are no longer
familiar, work through the [calculus refresher](../Supplements/calculus.qmd) first.

# Learning goals

In this chapter, we will cover:

- deterministic functions, their domain, range, parameters and units;
- phenomenological and mechanistic functions;
- how to write and plot functions in R, including functions of two inputs;
- limits and asymptotes;
- derivatives, slopes and curvature;
- maxima, inflection points, half-maxima and characteristic scales;
- the meaning of parameters, reparameterisation and derivatives with respect to
  parameters;
- how to build new functions by shifting, scaling, composing and generalising;
- how to choose a function and read starting values for its parameters from a plot.

We will use the
[Michaelis–Menten](../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten) and
[Ricker](../Supplements/deterministic-functions.qmd#sec-fn-ricker) functions
throughout the chapter. Many other functions are described in the
[Supplement](../Supplements/deterministic-functions.qmd), which you will need for the
practical. The rules of calculus that the chapter relies on are in the
[calculus refresher](../Supplements/calculus.qmd).

# Deterministic functions {#sec-deterministic-functions}

A **deterministic function** assigns exactly one output to each input, for given values of
its parameters: the same input always gives the same output. We write $y=f(x)$, where $x$
is the input, $y$ is the output and $f$ is the rule that turns one into the other. The
**parameters** of the function are the constants in that rule. They stay fixed while $x$
varies, and changing them changes the curve. The **domain** of the function is the set of
values of $x$ for which we use it, and its **range** is the set of values that $y$ can
take.

Let's illustrate these concepts with the reed frog experiments. In the first one, tanks
were stocked with different numbers of tadpoles (from 5 to 100) and with three dragonfly
larvae, and the number of tadpoles killed was counted after 14 days. In the second one,
each tank received 10 tadpoles of a given body length (from 9 to 37 mm) and the number
killed was counted after 3 days. In R:

```{r}
# Load the two reed frog experiments of the package emdbook
data(ReedfrogFuncresp, package = "emdbook")
data(ReedfrogSizepred, package = "emdbook")

# Tadpoles killed (Killed) against the initial number of tadpoles per tank (Initial)
head(ReedfrogFuncresp)

# Tadpoles killed out of 10 (Kill) against tadpole body length in mm (TBL)
head(ReedfrogSizepred)
```

A function that is often used for data like those of the first experiment is the
**Michaelis–Menten** function,

$$
f(x)=\frac{ax}{b+x},
$$

where $x$ is the initial number of tadpoles per tank, $f(x)$ is the number of tadpoles
killed, and $a$ and $b$ are the parameters. For positive $a$ and $b$ and for $x\ge0$ the
function starts at zero, increases, and levels off below $a$ (we will derive these
properties in the next sections). So its domain is $x\ge0$ and its range is $0\le y<a$.

In this book we will generally assume that parameters are positive and that inputs are non-negative,
unless stated otherwise, as this is the most common situation in ecology.^[Strictly speaking,
the domain is part of the definition of a function, and the same formula with another domain
is another function. Ecological texts usually leave the domain and the conditions on the
parameters implicit. When a formula in this chapter is used with negative inputs or
negative parameters, the text says so.]

The formula also tells us about the **units** of the parameters, which is the first step towards
interpreting them. Only quantities with the same units can be added, so in $b+x$ the
parameter $b$ must have the units of $x$ (tadpoles per tank). The ratio $x/(b+x)$ then has
no units, so $a$ must have the units of $y$ (tadpoles killed). We can use two rules for the
kind of models we will see in the course:

- Variables that are added or substracted must have the same units.

- The arguments of functions like exponential or logarithm^[In this book $\log$
is always the natural logarithm (base $e$), as in R and in the previous chapters.] has no
units, so a parameter that multiplies $x$ inside an exponential, as in $e^{-bx}$, has the
inverse units of $x$, whereas a parameter that divides $x$ has the units of $x$.

We always need to state the units of the parameters and of the results, as they are what
turns a number such as $b=120$ into an ecologically meaningful statement.

@fig-candidate_curves shows both experiments with a candidate function for each. The
curve in the left panel is a Michaelis–Menten function with $a=60$ tadpoles killed and
$b=120$ tadpoles per tank. The curve in the right panel is a **Ricker** function,
$f(x)=axe^{-bx}$, which rises from zero to a peak and then declines, with $a=1$ tadpole
killed/mm and $b=0.15$/mm. Note that the Ricker function describes the second
experiment poorly: it peaks at a body length below those of the experiment and it declines
too slowly. With two parameters it cannot place a narrow peak at 12 mm, where most
tadpoles were killed. In @sec-nested-functions we will add a third parameter that can.

```{r}
#| echo: false
#| label: fig-candidate_curves
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Two reed frog predation experiments with curves whose parameter values were chosen by eye. Left: number of tadpoles killed against the initial number of tadpoles per tank, with a Michaelis–Menten function (a = 60, b = 120). Right: number of tadpoles killed out of 10 against body length, with a Ricker function (a = 1, b = 0.15)."
library(patchwork)

# Curves written out in full: the functions are defined later in the chapter
density_curve = data.frame(Initial = seq(0, 100, length.out = 200))
density_curve$Killed = 60 * density_curve$Initial / (120 + density_curve$Initial)
size_curve = data.frame(TBL = seq(0, 40, length.out = 200))
size_curve$Kill = 1 * size_curve$TBL * exp(-0.15 * size_curve$TBL)

panel_density = ggplot() +
  geom_point(data = ReedfrogFuncresp, aes(x = Initial, y = Killed),
             colour = "#0072B2", alpha = 0.7, size = 2) +
  geom_line(data = density_curve, aes(x = Initial, y = Killed),
            colour = "#D55E00", linewidth = 0.8) +
  labs(x = "Initial number of tadpoles per tank", y = "Tadpoles killed in 14 days") +
  theme_classic()

panel_size = ggplot() +
  geom_point(data = ReedfrogSizepred, aes(x = TBL, y = Kill),
             colour = "#0072B2", alpha = 0.7, size = 2) +
  geom_line(data = size_curve, aes(x = TBL, y = Kill),
            colour = "#D55E00", linewidth = 0.8) +
  labs(x = "Tadpole body length (mm)", y = "Tadpoles killed out of 10 in 3 days") +
  theme_classic()

panel_density + panel_size
```

:::: {.callout-important title="Practice: units of the parameters of a growth curve"}

The height of a plant (cm) as a function of its age (days) is described with the function
$f(x)=\dfrac{a}{1+e^{-b(x-c)}}$. What are the units of $a$, $b$ and $c$?

::: {.callout-tip title="Solution" collapse="true"}

The parameter $c$ is subtracted from $x$, so it has the units of $x$ (days). The argument
of the exponential, $b(x-c)$, has no units, so $b$ has the inverse units of $x$ (1/day).
The denominator then has no units, so $a$ has the units of $y$ (cm).

:::
::::

## Phenomenological and mechanistic functions {#sec-mechanistic}

A **phenomenological** function describes the pattern in the data: it has the right shape,
and its parameters describe that shape (a height, a location, a steepness). A
**mechanistic** function is derived from assumptions about the process that generates the
pattern, so its parameters measure something in the system (a rate, a time, a
probability). The difference lies in how we arrive at the formula, and the same formula
can be either. Above we chose the Michaelis–Menten function only because it increases and
levels off. It can also be derived from the way in which a predator spends its time, and
the derivation is short enough to show it here as an example.

Suppose that a predator hunts for a total time $T$ in a tank with $N$ prey. While it
searches, it encounters a fraction $\alpha$ of the prey per unit of time (the **attack
rate**). Every prey that it catches costs it a **handling time** $H$ (to subdue, eat and
digest the prey) during which it does not search. If the predator searches for a time
$T_s$, the number of prey that it eats is $N_e=\alpha NT_s$, and the total time is the sum
of the time spent searching and the time spent handling, $T=T_s+HN_e$. We substitute
$T_s=T-HN_e$ in the expression for $N_e$ and solve for $N_e$:

$$
\begin{aligned}
N_e &= \alpha N\left(T-HN_e\right) \\
N_e\left(1+\alpha HN\right) &= \alpha NT \\
\frac{N_e}{T} &= \frac{\alpha N}{1+\alpha HN},
\end{aligned}
$$

where $N_e/T$ is the number of prey eaten per unit of time. This is the **disc equation**
of Holling,^[The equation is due to Holling, who analysed it for a predator that searches
for and handles prey; it is known as the disc equation. See Holling (1959), "Some
characteristics of simple types of predation and parasitism", *The Canadian Entomologist*
91: 385–398.] better known in ecology as the **Holling type II** functional response. As
for the units, $\alpha$ is a fraction per unit of time (e.g., per day) and $H$ is a time
per prey (e.g., days per tadpole), so $\alpha HN$ has no units, as it must if we add it to the
number 1. When prey are scarce the predator spends all of its time searching and eats $\alpha N$
prey per unit of time. When prey are abundant it spends all of its time handling them and
eats $1/H$ prey per unit of time.

If we divide the numerator and the denominator by $\alpha H$, we find that the disc
equation is the Michaelis–Menten function^[Michaelis and Menten (1913) showed that the
rate of a reaction catalysed by an enzyme follows this function of the concentration of
the substrate ("Die Kinetik der Invertinwirkung", *Biochemische Zeitschrift* 49: 333–369).
The same function is known as the Monod equation for the growth rate of micro-organisms on
a limiting nutrient (Monod 1949, "The growth of bacterial cultures", *Annual Review of
Microbiology* 3: 371–394) and as the Beverton–Holt stock–recruitment curve in fisheries
(Beverton and Holt 1957, *On the Dynamics of Exploited Fish Populations*).] with $x=N$,
$a=1/H$ and $b=1/(\alpha H)$. @sec-reparameterisation works out this conversion with
numbers. So, if we accept the assumptions of the derivation, the asymptote of the curve in
the left panel of @fig-candidate_curves measures a handling time, which is a quantity that
could also be measured directly by watching a predator. The parameters of a
phenomenological function have no such second meaning. Many of the functions in the
[Supplement](../Supplements/deterministic-functions.qmd) can be derived from a mechanism,
and each entry says whether that is the case.

Note that a mechanistic derivation tells us what the parameters would measure if its
assumptions held. A curve that fits the data well does not show that they hold, because
other processes can produce the same shape (we chose the Michaelis–Menten shape above only
because it increases and levels off).

<!-- AUTHOR: your view on when a phenomenological description is enough and when a
mechanistic function is worth the effort. -->

## The two functions of this chapter {#sec-two-functions}

The chapter analyses two functions in full, each with a shape that is common in ecological
data (@fig-two_functions). To describe the shapes of functions we need four terms. A
function is **monotonic** when it only increases or only decreases over its domain,
**saturating** when it increases and levels off at an upper limit, **unimodal** (or
hump-shaped) when it rises to a single maximum and declines after it, and **sigmoid** (or
S-shaped) when it increases slowly at first, then fast, and then slowly again. The two
functions are:

- **Michaelis–Menten**, $f(x)=\dfrac{ax}{b+x}$. It is monotonic and saturating: it
  increases from zero towards $a$. It describes rates that are limited by something other
  than $x$ when $x$ is abundant, such as a feeding rate as a function of the density of
  prey or the uptake of a nutrient as a function of its concentration
  ([entry](../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten)).
- **Ricker**, $f(x)=axe^{-bx}$.^[The function is named after Ricker, who proposed it as a
  curve relating the recruits of a fish stock to the size of the spawning stock. See
  Ricker (1954), "Stock and recruitment", *Journal of the Fisheries Research Board of
  Canada* 11: 559–623.] It is unimodal: it rises from zero to a maximum and declines
  towards zero. Besides the number of recruits as a function of the number of spawners, it
  is used for any response that is highest at intermediate values of the predictor
  ([entry](../Supplements/deterministic-functions.qmd#sec-fn-ricker)).

Two other shapes return in the practical and in later chapters, and the supplement has a
full entry for a function of each. The
[exponential decay](../Supplements/deterministic-functions.qmd#sec-fn-exponential),
$f(x)=ae^{-bx}$, is monotonic: it decreases from $a$ at $x=0$ towards zero, as a quantity
does that loses a constant fraction per unit of $x$. The
[logistic](../Supplements/deterministic-functions.qmd#sec-fn-logistic) function,
$f(x)=a/(1+e^{-b(x-c)})$, is sigmoid: it increases from zero towards $a$ and it is
steepest half-way.

The formulas of this book share the letters $a$, $b$, $c$ and $d$. These letters are positions in a
formula and they control something different in each function, so for every function we
will say what each of them does. Symbols with an ecological meaning, such as $\alpha$ and
$H$ above, appear only when we discuss a particular parameterisation. Note also that some
of these letters had other meanings for the distributions of the previous chapters ($a$
and $b$ are the parameters of the Beta distribution, $k$ is the shape of the Negative
Binomial distribution and $K$ the number of parameters in AIC).

```{r}
#| echo: false
#| label: fig-two_functions
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "The two functions of this chapter. Left: Michaelis–Menten function with a = 10 and b = 5; the dashed line marks its asymptote. Right: Ricker function with a = 1 and b = 0.2; the dashed line marks the location of its peak."
library(patchwork)

# Each function is drawn over a range of x that shows its shape
mm_data = data.frame(x = seq(0, 50, length.out = 200))
mm_data$y = 10 * mm_data$x / (5 + mm_data$x)
ricker_data = data.frame(x = seq(0, 30, length.out = 200))
ricker_data$y = 1 * ricker_data$x * exp(-0.2 * ricker_data$x)

panel_mm = ggplot(mm_data, aes(x = x, y = y)) +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Michaelis–Menten", x = "x", y = "y") +
  theme_classic()

panel_ricker = ggplot(ricker_data, aes(x = x, y = y)) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Ricker", x = "x", y = "y") +
  theme_classic()

panel_mm + panel_ricker
```

# Functions in R {#sec-functions-r}

To explore a function we first write it in R. An R function for $f(x)$ takes the input as
its first argument and the parameters as further arguments. We give each parameter a
**default value**, written as `a = 1` in the definition, which R uses when a call does not
mention that argument. For the Ricker function:

```{r}
# A function of x with two parameters; a and b have default values of 1,
# which R uses when the call does not give them
ricker = function(x, a = 1, b = 1) {
  a * x * exp(-b * x)
}

# One value of x with the default parameters
ricker(x = 2)

# One value of x with other values of the parameters
ricker(x = 2, a = 1, b = 0.15)

# Several values of x at once: the arithmetic is vectorised
ricker(x = c(0, 2, 5, 10), a = 1, b = 0.15)

# One value of x and several values of a
ricker(x = 2, a = c(1, 2, 3), b = 0.15)
```

The first call uses the default values and returns $2e^{-2}\approx0.271$. The arithmetic of
R is vectorised (see the practical of Chapter 1), so the same function accepts a vector of
inputs and returns one output for each of them, which is what we need to draw a curve. It
also accepts a vector of values for a parameter, which shows how the output at a fixed $x$
changes with that parameter: in the last call the output is proportional to $a$. The
Michaelis–Menten function is written in the same way:

```{r}
# The Michaelis–Menten function, with the same layout
michaelis_menten = function(x, a = 1, b = 1) {
  a * x / (b + x)
}

michaelis_menten(x = c(5, 20, 50, 100), a = 60, b = 120)
```

For a quick look at a function we use `curve()`. Its first argument is an expression
written in terms of `x`, which `curve()` evaluates between the values `from` and `to`. With
`add = TRUE` it draws on top of the existing plot, which is a fast way of seeing what a
parameter does. Here, halving $b$ makes the Michaelis–Menten function rise faster towards
the same asymptote:

```{r}
# Quick exploration with curve(): the first argument is an expression in x
curve(michaelis_menten(x, a = 60, b = 120), from = 0, to = 100,
      xlab = "x", ylab = "y", ylim = c(0,40))

# add = TRUE draws another curve on the same plot
curve(michaelis_menten(x, a = 60, b = 60), add = TRUE, lty = 2)
```

The figures of this book are drawn with `ggplot2`. For those we can calculate the function on
a sequence of values of $x$, store the result in a data frame and draw it as a line
(@fig-ricker_two_values):

```{r}
#| label: fig-ricker_two_values
#| fig-cap: "Ricker function with a = 1 for two values of b."
# The pattern used for the figures of this book: a sequence of x values,
# a data frame with the function values, and ggplot2
x_values = seq(0, 40, length.out = 200)
ricker_curves = data.frame(
  x = rep(x_values, times = 2),
  y = c(ricker(x_values, a = 1, b = 0.15), ricker(x_values, a = 1, b = 0.3)),
  b = rep(c("b = 0.15", "b = 0.3"), each = length(x_values))
)

ggplot(ricker_curves, aes(x = x, y = y, linetype = b)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(x = "x", y = "y", linetype = NULL) +
  theme_classic()
```

Doubling $b$ moves the peak of the Ricker function to the left and lowers it. We will see
why in @sec-function-maxima.

Some functions are defined by different formulas on different parts of their domain
(**piecewise** functions). In R we write them with `ifelse()`, which takes a condition and
two expressions: for each element of `x` it returns the value of the first expression
where the condition is `TRUE` and the value of the second one where it is `FALSE`. Two
examples are the
[threshold function](../Supplements/deterministic-functions.qmd#sec-fn-threshold), which
jumps from $a$ to $b$ at $x=c$, and the
[hockey stick](../Supplements/deterministic-functions.qmd#sec-fn-hockey-stick), which
increases as a straight line with slope $a$ up to $x=c$ and stays at $ac$ from there on
(@fig-piecewise). The hockey stick can also be written with `pmin()`, which returns the
smaller of two values for each element of `x`:

```{r}
# Functions defined in pieces: ifelse() chooses between two expressions
# for each value of x. The third parameter is called c, as in the formula. R keeps
# functions and other objects apart, so c() below still combines values
threshold = function(x, a = 0, b = 1, c = 1) {
  ifelse(x < c, a, b)
}

hockey_stick = function(x, a = 1, c = 1) {
  ifelse(x < c, a * x, a * c)
}

# pmin() takes the smaller of the two values for each x: a shorter hockey stick
hockey_stick_short = function(x, a = 1, c = 1) {
  pmin(a * x, a * c)
}

x_values = c(0, 2.5, 5, 7.5, 10)
hockey_stick(x_values, a = 2, c = 5)
hockey_stick_short(x_values, a = 2, c = 5)
```

Both versions return the same values: the function increases by $a=2$ per unit of $x$ up
to $x=5$ and stays at $ac=10$ beyond it. Note that these functions have a parameter called
`c`, as in their formulas. This does not interfere with the function `c()` that combines
values, because R looks for a function whenever a name is followed by parentheses.

```{r}
#| echo: false
#| label: fig-piecewise
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Two functions defined in pieces. Left: threshold function with a = 1, b = 4 and c = 5. Right: hockey stick function with a = 0.8 and c = 5."
library(patchwork)

x_values = seq(0, 10, length.out = 400)
threshold_data = data.frame(x = x_values, y = threshold(x_values, a = 1, b = 4, c = 5))
hockey_data = data.frame(x = x_values, y = hockey_stick(x_values, a = 0.8, c = 5))

panel_threshold = ggplot(threshold_data, aes(x = x, y = y)) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Threshold", x = "x", y = "y") +
  theme_classic()

panel_hockey = ggplot(hockey_data, aes(x = x, y = y)) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Hockey stick", x = "x", y = "y") +
  theme_classic()

panel_threshold + panel_hockey
```

## Functions of two inputs {#sec-function-surfaces}

A **function of two inputs** assigns one output to each pair of values of its inputs,
$y=f(x_1,x_2)$. Its graph is a surface over the plane of the two inputs. We need such
functions when a response depends on two predictors, and we have already met one: the
log-likelihood of a model with two parameters in
[Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-surface) was a surface over the plane
of the two parameters.

As an example we combine the two reed frog experiments in one function. According to the
disc equation, a predator eats $\alpha N/(1+\alpha HN)$ prey per unit of time, so each of
the $N$ tadpoles in a tank is eaten at a rate $\alpha/(1+\alpha HN)$, which we will call
its mortality risk. Suppose now that the attack rate depends on the body length $s$ of the
tadpoles as a Ricker function, $\alpha(s)=ase^{-bs}$, so that tadpoles of intermediate
length are attacked the most. The mortality risk is then a function of body length and of
density:

$$
f(s,N)=\frac{\alpha(s)}{1+\alpha(s)HN}.
$$

We take $a=0.02$/day/mm and $b=(1/12)$/mm (so the attack rate is highest for
tadpoles of 12 mm, where it is $0.02\times12\times e^{-1}\approx0.088$/day) and a
handling time $H=0.5$ days per tadpole. For a tadpole of 12 mm in a tank with 20 tadpoles
the risk is $0.088/(1+0.088\times0.5\times20)\approx0.047$/day. To evaluate the
function for many combinations of the two inputs we use `outer()`, and we draw the surface
with `contour()`, as in the practical of Chapter 1:

```{r}
# Attack rate of the predator as a function of tadpole body length (mm)
attack_rate = function(size, a = 0.02, b = 1 / 12) {
  a * size * exp(-b * size)
}

# Per capita mortality risk: attack rate divided by 1 + attack rate x handling time x density
mortality_risk = function(size, density, H = 0.5) {
  alpha = attack_rate(size)
  alpha / (1 + alpha * H * density)
}

# One pair of values
mortality_risk(size = 12, density = 20)

# All combinations of 50 sizes and 50 densities: outer() needs a vectorised function
size_values = seq(5, 40, length.out = 50)
density_values = seq(1, 100, length.out = 50)
risk_matrix = outer(size_values, density_values, mortality_risk)

# Contour lines of the surface
contour(size_values, density_values, risk_matrix,
        xlab = "Body length (mm)", ylab = "Tadpoles per tank")
```

The function `outer()` calls `mortality_risk()` once, with all the combinations of the two
vectors, so it requires a function that is vectorised. A likelihood function takes a whole
sample and returns a single number for one set of parameter values, which is why in
Chapter 4 we filled the grid with loops.

@fig-mortality_surface shows the same surface drawn with `ggplot2`. The risk is highest
for tadpoles of about 12 mm at low density, and it decreases with density because the
predators have less time available per tadpole. If we read the figure along a horizontal
line (a fixed density) we see how the risk changes with body length alone, and along a
vertical line we see how it changes with density alone. Looking at one input while the
other is held fixed is the idea behind the partial derivative of
@sec-parameter-derivatives. Note also that we built this function by placing one function
inside another, which is called composition (@sec-function-composition).

```{r}
#| echo: false
#| label: fig-mortality_surface
#| fig-cap: "Per capita mortality risk of a tadpole as a function of its body length and of the number of tadpoles per tank. Shading shows the band of risk; each boundary joins the combinations of body length and density with the same risk."
risk_grid = expand.grid(size = size_values, density = density_values)
risk_grid$risk = mortality_risk(risk_grid$size, risk_grid$density)

# Filled contours in greys: darker means higher risk
ggplot(risk_grid, aes(x = size, y = density, z = risk)) +
  geom_contour_filled(breaks = c(0, 0.02, 0.04, 0.06, 0.08, 0.1)) +
  scale_fill_grey(start = 0.95, end = 0.3) +
  labs(x = "Tadpole body length (mm)", y = "Tadpoles per tank",
       fill = "Risk (1/day)") +
  theme_classic()
```

# Limits and asymptotes {#sec-function-limits}

The **limit** of a function as $x$ approaches a value is the value that the function
approaches as $x$ gets closer and closer to it. We write $\lim_{x\to\infty}f(x)=a$, or
$f(x)\to a$ as $x\to\infty$, when the function approaches $a$ as $x$ grows without bound.
A finite limit as $x\to\infty$ is a horizontal **asymptote**: a horizontal line that the
curve approaches (a curve may cross its asymptote, as the
[Holling type IV](../Supplements/deterministic-functions.qmd#sec-fn-holling4) function
does). The first thing to ask of a candidate function is
where it starts and where it ends, since that must agree with what we know about the
response. It turns out that those two values are usually parameters of the function or
simple expressions of them, and this is how the parameters start to acquire a meaning.

Finding these limits rarely takes a long calculation. When a function is defined at a
point and has no jump there, its limit is its value at that point, so the left end of our
functions is found by setting $x=0$. For the right end the working rule is to keep the
term that grows fastest and to ignore the others. As $x\to\infty$, an exponential $e^{bx}$
grows faster than any power of $x$, a higher power of $x$ grows faster than a lower one,
and any power grows faster than a constant. As $x\to0$ the ranking is reversed: the
constant, or else the lowest power of $x$, is what remains. In plain English, at each end
of a function only one term of every sum matters. The
[calculus refresher](../Supplements/calculus.qmd#sec-calc-limits) explains where this
ranking comes from (L'Hôpital's rule) and has more examples. For our two functions:

- **Michaelis–Menten.** At $x=0$ the numerator is zero, so $f(0)=0$. For large $x$ the
  constant $b$ in the denominator is negligible next to $x$. Formally, we divide the
  numerator and the denominator by $x$ and obtain $f(x)=a/(b/x+1)\to a$, because
  $b/x\to0$. So $a$ is the asymptote. For small $x$ it is $x$ that is negligible next to
  $b$, so $f(x)\approx(a/b)x$ when $x\ll b$ (the symbol $\ll$ reads "much smaller than"):
  the function starts as a straight line with slope $a/b$.
- **Ricker.** At $x=0$, $f(0)=0$. As $x\to\infty$ the function is the product of a term
  that grows ($x$) and a term that approaches zero ($e^{-bx}=1/e^{bx}$). The exponential
  dominates, so $f(x)\to0$. For small $x$ the exponential is close to 1 and
  $f(x)\approx ax$.

A limit can always be checked by evaluating the function at a very large or a very small
value of $x$. In R, for the curves of the previous sections:

```{r}
# Fraction of the asymptote reached at the largest density in the data (100 tadpoles per tank)
michaelis_menten(x = 100, a = 60, b = 120) / 60

# Fraction of the asymptote at x = 9b
michaelis_menten(x = 9 * 120, a = 60, b = 120) / 60

# For x much smaller than b the function is close to (a / b) x
michaelis_menten(x = 1, a = 60, b = 120)
(60 / 120) * 1

# The exponential dominates x in the Ricker function: values at large x
ricker(x = c(100, 1000), a = 1, b = 0.15)
```

With the values that we chose by eye for the tadpoles ($a=60$ tadpoles killed and $b=120$
tadpoles per tank), the curve has reached 45% of its asymptote at the highest density of
the experiment, 100 tadpoles per tank. So the experiment covers less than half of the
rise of the curve, which will matter when we read starting values from the plot
(@sec-starting-values). In general, the Michaelis–Menten function approaches its
asymptote slowly: it reaches 90% of it only at $x=9b$. At the other end, the function
gives 0.496 tadpoles killed for one tadpole per tank, close to the 0.5 of the straight
line $(a/b)x$. The Ricker function of the size experiment is already at $3\times10^{-5}$
for $x=100$.

:::: {.callout-important title="Practice: the two ends of two functions"}

1. The number of seeds produced by a plant decreases with the density of plants around
   it, $x$, as the hyperbolic function $f(x)=\dfrac{a}{b+x}$. Calculate $f(0)$ and
   $\lim_{x\to\infty}f(x)$.
2. The feeding rate of a predator that ignores scarce prey is described by the Holling
   type III function $f(x)=\dfrac{ax^2}{b^2+x^2}$. Calculate $f(0)$ and
   $\lim_{x\to\infty}f(x)$. How does the function behave when $x\ll b$?

::: {.callout-tip title="Solution" collapse="true"}

1. At zero, $f(0)=a/b$: this is the seed production of a plant without neighbours. For
   large $x$ the constant $b$ is negligible next to $x$, so $f(x)\approx a/x\to0$.
2. At zero the numerator is zero, so $f(0)=0$. For large $x$ we divide the numerator and
   the denominator by $x^2$, which gives $a/(b^2/x^2+1)\to a$. For $x\ll b$ the term $x^2$
   in the denominator is negligible next to $b^2$, so $f(x)\approx(a/b^2)x^2$. The function
   starts as a parabola, with a slope of zero, whereas the Michaelis–Menten function
   starts as a straight line.

:::
::::

# Derivatives, slopes and curvature {#sec-function-derivatives}

The **derivative** of a function at a point, written $f'(x)$ or
$\dfrac{df}{dx}$,^[The notation $df/dx$ is due to Leibniz (1675) and $f'$ to Lagrange
(1797); both are in use. We write $f'(x)$ in running text and $df/dx$ when it matters with
respect to which quantity we differentiate.] is the slope of the function at that point:
the change in $y$ per unit change in $x$, measured over an interval so short that the
curve is indistinguishable from a straight line. It follows that the units of a derivative
are the units of $y$ divided by the units of $x$. We have already used a derivative in
[Chapter 4](../Chapter_4/Theory.qmd#sec-maximum-likelihood) and they will come back in
[Chapter 6](../Chapter_6/Theory.qmd).

Derivatives are found with a small set of rules, listed in the table below, where $c$, $n$
and $b$ are constants and $g$ and $h$ are functions of $x$. This chapter recalls the
rules and applies them. If they are not familiar, work through
[Rules of differentiation](../Supplements/calculus.qmd#sec-calc-rules) in the calculus
refresher first.

| Rule | Function | Derivative |
|:---|:---|:---|
| Constant multiple | $c\,g(x)$ | $c\,g'(x)$ |
| Sum | $g(x)+h(x)$ | $g'(x)+h'(x)$ |
| Power | $x^n$ | $n\,x^{n-1}$ |
| Exponential | $e^{bx}$ | $b\,e^{bx}$ |
| Logarithm | $\log x$ | $1/x$ |
| Product | $g(x)\,h(x)$ | $g'(x)\,h(x)+g(x)\,h'(x)$ |
| Quotient | $g(x)/h(x)$ | $\dfrac{g'(x)\,h(x)-g(x)\,h'(x)}{h(x)^2}$ |
| Chain | $g(h(x))$ | $g'(h(x))\,h'(x)$ |

The two functions require different rules. For the Michaelis–Menten function we use the
quotient rule with $g(x)=ax$ and $h(x)=b+x$:

$$
\frac{d}{dx}\left(\frac{ax}{b+x}\right) = \frac{a(b+x)-ax}{(b+x)^2} = \frac{ab}{(b+x)^2}.
$$

For the Ricker function we use the product rule with $g(x)=ax$ and $h(x)=e^{-bx}$, whose
derivative is $-b\,e^{-bx}$ by the rule for the exponential:

$$
\frac{d}{dx}\left(axe^{-bx}\right) = ae^{-bx}-abx\,e^{-bx} = a(1-bx)\,e^{-bx}.
$$

Note also the signs. The derivative of the Michaelis–Menten function is positive for every
$x$, whereas the derivative of the Ricker function changes sign.

If we now take the limits of these derivatives, we obtain the slope of each function at
its two ends:

| Function | Slope at $x=0$ | Slope as $x\to\infty$ |
|:---|:---|:---|
| Michaelis–Menten | $a/b$ | $0$ |
| Ricker | $a$ | $0$ |

Both functions become flat for large $x$. At $x=0$ the slope of the Michaelis–Menten
function is $a/b$, which is the straight line that we found with limits, and the slope of
the Ricker function is $a$. So the parameter $a$ of the Ricker function is its initial
slope, and in the Michaelis–Menten function the initial slope is a ratio of the two
parameters.

We can verify our math with R. The function `D()` differentiates a formula with the rules
of the table. It needs the formula as an expression that has not been evaluated, which
`expression()` creates, and the name of the variable in quotes. Its result is another
expression, which `eval()` evaluates for the values that we supply in a list:

```{r}
# D() differentiates an expression: expression() stores the formula without evaluating it
mm_expression = expression(a * x / (b + x))
mm_derivative = D(mm_expression, "x")
mm_derivative

# eval() evaluates the derivative for given values of the parameters and of x
eval(mm_derivative, list(a = 60, b = 120, x = 50))

# The formula derived on paper: a * b / (b + x)^2
60 * 120 / (120 + 50)^2

# A finite-difference slope: the change in y over a small change in x
h = 0.00001
(michaelis_menten(50 + h, a = 60, b = 120) - michaelis_menten(50 - h, a = 60, b = 120)) / (2 * h)
```

`D()` applies the rules mechanically and does not simplify the result, so its output
looks different from our formula, but it is the same function: both give a slope of 0.249
at $x=50$ for the curve of the tadpoles. The units are tadpoles killed per tadpole per
tank. That is, at a density of 50 tadpoles per tank, one more tadpole in the tank adds
about a quarter of a tadpole to the expected number killed, against half a tadpole at very
low density ($a/b=0.5$). The last lines of the chunk calculate the slope in a third way, as
a **finite difference**: the change in the function between two points very close to
$x=50$, divided by the distance between them. This requires no calculus, and it is how
software approximates a derivative when no formula is available
([Chapter 8](../Chapter_8/Theory.qmd)). The same check for the Ricker function, with
$a=10$ and $b=0.5$ at $x=4$:

```{r}
# The same check for the Ricker function, with D() and with the formula
# derived above, at x = 4
ricker_derivative = D(expression(a * x * exp(-b * x)), "x")
eval(ricker_derivative, list(a = 10, b = 0.5, x = 4))

# The formula derived on paper: a * (1 - b * x) * exp(-b * x)
10 * (1 - 0.5 * 4) * exp(-0.5 * 4)
```

Again, `D()` and the formula that we derived return the same slope. It is negative,
because $x=4$ lies beyond the peak of this function.

The **second derivative**, written $f''(x)$ or $\dfrac{d^2f}{dx^2}$, is the derivative of
the derivative: the rate at which the slope changes. Its sign describes the curvature of
the function. Where $f''(x)<0$ the slope decreases as $x$ increases and the function is
**concave** (it bends downwards). Where $f''(x)>0$ the slope increases and the function is
**convex** (it bends upwards). For the Michaelis–Menten function we differentiate
$f'(x)=ab(b+x)^{-2}$ with the power rule and the chain rule:

$$
f''(x)=-\frac{2ab}{(b+x)^3},
$$

which is negative for every $x\ge0$. So the function is concave everywhere: every
additional tadpole in the tank adds less to the number killed than the previous one, which
is what saturating means. We also met the second derivative in Chapter 4, where the
curvature of the log-likelihood at its maximum gave the standard error of an estimate
(see [Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-uncertainty)). In R, we apply
`D()` twice:

```{r}
# Second derivative of the Michaelis–Menten function: D() applied twice
mm_second = D(mm_derivative, "x")
eval(mm_second, list(a = 60, b = 120, x = 50))

# The formula derived on paper: -2 * a * b / (b + x)^3
-2 * 60 * 120 / (120 + 50)^3
```

:::: {.callout-important title="Practice: growth rate of a fish and metabolic rate of a mammal"}

1. The length of a fish (cm) at age $x$ (years) follows the saturating exponential
   function $f(x)=a\left(1-e^{-bx}\right)$ with $a=60$ cm and $b=0.2$/year. Derive
   $f'(x)$. How fast does the fish grow at birth and at an age of 5 years?
2. The metabolic rate of a mammal (W) as a function of its body mass $x$ (kg) follows the
   power law $f(x)=ax^{b}$ with $a=3.4$ and $b=0.75$. Derive $f'(x)$ and calculate it for
   a mammal of 16 kg. What are its units?

::: {.callout-tip title="Solution" collapse="true"}

1. The derivative of the constant $a$ is zero, and the rule for the exponential gives
   $f'(x)=-a\times(-b)\,e^{-bx}=ab\,e^{-bx}$. At birth the growth rate is $ab=12$
   cm/year, and at 5 years it is $12e^{-1}\approx4.4$ cm/year.
2. The power rule gives $f'(x)=ab\,x^{b-1}=2.55\,x^{-0.25}$. For $x=16$ kg we have
   $16^{-0.25}=1/2$, so the slope is about 1.3 W/kg: a mammal that is one kg heavier
   than 16 kg has a metabolic rate that is about 1.3 W higher.

:::
::::

# Special points of a function {#sec-function-special-points}

Limits and derivatives lead to a few points that summarise a curve: where it peaks, where
it bends and how wide it is. These special points are what we can see in a plot of data.
If we write them as expressions of the parameters, we can go from a plot to values of the
parameters, which is what we will do at the end of the chapter (@sec-starting-values).

## Maxima and minima {#sec-function-maxima}

A **critical point** of a function is a value of $x$ at which $f'(x)=0$, so that the
function is flat there. It is a maximum when the slope changes from positive to negative
at that point, which is the case when $f''(x)<0$ there, and a minimum when the slope
changes from negative to positive ($f''(x)>0$).^[The derivative can also be zero at a
point where the function flattens on its way up or down without turning, as $x^3$ does at
$x=0$. Such a point is called a saddle point, and it is the reason why the sign of the
slope on both sides has to be checked.] We saw above that the derivative of the
Michaelis–Menten function never changes sign, which confirms that this function is
monotonic and has no critical point.

The Ricker function has one. Its derivative is $f'(x)=a(1-bx)e^{-bx}$, and since $a$ and
$e^{-bx}$ are positive, it is zero only where $1-bx=0$:

$$
x_{\max}=\frac{1}{b}, \qquad f(x_{\max})=\frac{a}{b}\,e^{-1}=\frac{a}{be},
$$

where $x_{\max}$ is the location of the critical point and $f(x_{\max})$ the value of the
function there. The slope is positive for $x<1/b$ and negative for $x>1/b$, so the point
is a maximum. That is, $b$ alone decides where the peak is (note that $1/b$ has the units
of $x$), and for a given $b$ the height of the peak is proportional to $a$. This is why
doubling $b$ in @fig-ricker_two_values halved both the location and the height of the
peak. With the values of the size experiment, $a=1$ tadpole killed/mm and
$b=0.15$/mm, the peak is at $1/0.15\approx6.7$ mm with a height of $1/(0.15e)\approx2.45$ tadpoles
killed. We can verify our math with R by looking for the highest value on a fine grid with
`which.max()`, which returns the position of the highest value of a vector:

```{r}
# Check the peak of the Ricker function on a fine grid of body lengths
size_grid = seq(0, 40, by = 0.01)
ricker_values = ricker(size_grid, a = 1, b = 0.15)

# Location of the maximum and its height
size_grid[which.max(ricker_values)]
max(ricker_values)

# The values found on paper: 1 / b and a / (b * e)
1 / 0.15
1 / (0.15 * exp(1))
```

The smallest tadpoles of the experiment were 9 mm long and most tadpoles were killed at
12 mm, so a peak at 6.7 mm lies outside the data, as we saw in the right panel of
@fig-candidate_curves. Moving the peak to 12 mm requires $b=(1/12)$/mm, and the function
then declines even more slowly at larger body lengths, because the same parameter controls
the location of the peak and the rate of the decline.

The derivative of the Ricker function also shows a shortcut that we will use often. A
product is zero only where one of its factors is zero, and a quotient is zero only where
its numerator is zero. So to find the critical points of a function whose derivative is a
quotient we only need the numerator of the derivative (e.g., for the
[Holling type IV](../Supplements/deterministic-functions.qmd#sec-fn-holling4) function in
the supplement).

## Inflection points {#sec-function-inflection}

An **inflection point** is a value of $x$ at which $f''(x)$ changes sign, so that the
function changes from convex to concave or the other way around. At that point the slope
stops increasing and starts to decrease (or the reverse), so the function is at its
steepest or at its shallowest. For a sigmoid function the inflection point is where the
response changes fastest, and together with the slope at that point it is what we read
off a plot of data (for the
[logistic](../Supplements/deterministic-functions.qmd#sec-fn-logistic) function it is the
midpoint $c$, as its entry in the supplement shows).

The second derivative of the Michaelis–Menten function is negative for every $x\ge0$, so
this function has no inflection point. For the Ricker function the product rule applied
to $f'(x)=a(1-bx)e^{-bx}$ gives

$$
f''(x)=-ab\,e^{-bx}-ab(1-bx)\,e^{-bx}=-ab(2-bx)\,e^{-bx},
$$

which is zero at $x=2/b$, that is, at twice the location of the peak. The function is
concave before that point and convex after it, so this is where the decline after the
peak is steepest. The slope there is $f'(2/b)=-a/e^{2}$, which is $-0.135$ tadpoles
killed/mm for the size experiment, where the inflection point is at $2/0.15\approx13.3$ mm.
In R:

```{r}
# The Ricker function changes from concave to convex at 2 / b: the second
# derivative is zero there, negative before it and positive after it
ricker_second = D(ricker_derivative, "x")
eval(ricker_second, list(a = 1, b = 0.15, x = c(10, 2 / 0.15, 20)))

# Slope at the inflection point, and the formula derived on paper: -a / e^2
eval(ricker_derivative, list(a = 1, b = 0.15, x = 2 / 0.15))
-1 / exp(2)
```

The second derivative is negative at 10 mm, zero at the inflection point and positive at
20 mm, and the slope at the inflection point is $-0.135$ tadpoles killed/mm.

## Half-maxima and characteristic scales {#sec-function-scales}

The **half-maximum** of a saturating function is the value of $x$ at which the function
reaches half of its asymptote. To find the half-maximum we set the function equal to $a/2$
and solve for $x$. For the Michaelis–Menten function:

$$
\begin{aligned}
\frac{a\,x_{1/2}}{b+x_{1/2}} &= \frac{a}{2} \\
2x_{1/2} &= b+x_{1/2} \\
x_{1/2} &= b,
\end{aligned}
$$

where $x_{1/2}$ denotes the half-maximum. So the parameter $b$ of the Michaelis–Menten
function is its half-maximum. With the values for the tadpoles, half of the asymptote (30
tadpoles killed) is reached at 120 tadpoles per tank, which is beyond the densities of the
experiment. The same algebra shows that the function reaches 2/3 of its asymptote at
$x=2b$ and 3/4 at $x=3b$ (@fig-scales), which is the slow approach that we
found with limits.

More generally, a parameter that multiplies or divides $x$ wherever $x$ appears in a
formula is a **scale parameter**: changing it stretches or compresses the curve along the
$x$ axis without changing its shape. The parameter $b$ of the Michaelis–Menten function is
one, since $ax/(b+x)=a(x/b)/(1+x/b)$. A scale parameter that divides $x$ has the units of
$x$ and is the **characteristic scale** of the function: the distance along the $x$ axis
over which the function changes by a fixed proportion. A scale parameter that multiplies
$x$, as $b$ does in $e^{-bx}$, is a rate with the inverse units of $x$, and the
characteristic scale is then $1/b$.

The Ricker function has such a rate: it can be written as $(a/b)\,(bx)\,e^{-bx}$, and its
characteristic scale $1/b$ is the location of its peak. The simplest example is the
[exponential decay](../Supplements/deterministic-functions.qmd#sec-fn-exponential),
$ae^{-bx}$, which falls by a factor $e\approx2.72$ over every distance $1/b$, wherever we
start, and is halved over every distance $\log 2/b\approx0.69/b$. When $x$ is time the
latter is the **half-life**, $t_{1/2}=\log 2/b$ (its entry in the supplement has the
details).

```{r}
#| echo: false
#| label: fig-scales
#| fig-cap: "Michaelis–Menten function with a = 10 and b = 5. The function reaches half of its asymptote at x = b, 2/3 at 2b and 3/4 at 3b."
# Michaelis–Menten: marks at b, 2b and 3b
mm_scale = data.frame(x = seq(0, 25, length.out = 200))
mm_scale$y = 10 * mm_scale$x / (5 + mm_scale$x)
mm_marks = data.frame(x = c(5, 10, 15), y = c(5, 10 * 10 / 15, 7.5))

ggplot(mm_scale, aes(x = x, y = y)) +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "grey40") +
  geom_segment(data = mm_marks, aes(x = x, xend = x, y = 0, yend = y),
               linetype = "dashed", colour = "grey40") +
  geom_segment(data = mm_marks, aes(x = 0, xend = x, y = y, yend = y),
               linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  scale_x_continuous(breaks = c(5, 10, 15), labels = c("b", "2b", "3b")) +
  scale_y_continuous(breaks = c(0, 5, 10 * 10 / 15, 7.5, 10),
                     labels = c("0", "a / 2", "2a / 3", "3a / 4", "a")) +
  labs(x = "x", y = "y") +
  theme_classic() +
  theme(axis.title.y = element_text(margin = margin(r = 12)))
```

We can verify our math with R for the Michaelis–Menten function of the tadpoles, which
returns 30, 40 and 45 tadpoles killed (1/2, 2/3 and 3/4 of its asymptote of 60):

```{r}
# Michaelis–Menten function at b, 2b and 3b
michaelis_menten(x = c(120, 240, 360), a = 60, b = 120)
```

**Some special points have no formula**. For example, the body length at which the Ricker
function has fallen back to half of its peak is the solution of $axe^{-bx}=a/(2be)$, an
equation that cannot be solved for $x$ with algebra. In that case we find the point
numerically with `uniroot()`, as we did for the endpoints of profile likelihood intervals
in Chapter 4. `uniroot()` needs an interval at whose ends the function has opposite signs,
and we choose it from the plot: at the peak (6.67 mm) the Ricker function is above half of
its peak, and at 40 mm it is far below it.

```{r}
# A point without a formula: where does the Ricker function fall back to half of its peak?
peak_height = 1 / (0.15 * exp(1))

# uniroot() needs an interval with a sign change; from the plot, the peak is at
# 6.67 mm and the function is small at 40 mm
half_peak = uniroot(
  function(x) ricker(x, a = 1, b = 0.15) - peak_height / 2,
  interval = c(1 / 0.15, 40)
)
half_peak$root
```

So the Ricker function of the size experiment falls back to half of its peak at a body
length of 17.9 mm.

:::: {.callout-important title="Practice: a half-life and a half-maximum"}

1. The mass of leaf litter in a mesh bag decreases as $ae^{-bx}$, where $x$ is the time in
   days and $b=0.02$/day. Calculate the half-life of the litter and the time over
   which its mass falls by a factor $e$.
2. Find the half-maximum of the Holling type III function
   $f(x)=\dfrac{ax^2}{b^2+x^2}$.

::: {.callout-tip title="Solution" collapse="true"}

1. The half-life is $\log 2/b=0.693/0.02\approx34.7$ days, and the mass falls by a factor
   $e$ every $1/b=50$ days. Both are times, as they must be, because $b$ has the inverse
   units of $x$. Neither depends on the initial mass $a$.
2. The asymptote is $a$ (see the practice of @sec-function-limits), so we solve
   $ax^2/(b^2+x^2)=a/2$. This gives $2x^2=b^2+x^2$, so $x^2=b^2$ and $x_{1/2}=b$. Writing
   the function with $b^2$ is what makes $b$ the half-maximum, with the units of $x$.

:::
::::

# Parameters {#sec-function-parameters}

In the previous sections we started from a formula and found the properties of its curve.
This section looks at the same results from the side of the parameters: what each
parameter means, how a function can be rewritten with other parameters, and how the
function responds when a parameter changes.

## The meaning of parameters {#sec-meaning-parameters}

Once the special points and trends of a function are written as expressions of its parameters,
the parameters can be related to the shape of the curve. The table below collects the results
of the previous sections for the two functions. We will use it in the practical to interpret
parameters and, in Chapter 6, to interpret estimates and to choose starting values. Every
entry of the [Supplement](../Supplements/deterministic-functions.qmd) opens with the same
summary for its function.

| Function | Value at $x=0$ | Limit as $x\to\infty$ | Slope at $x=0$ | Special point |
|:---|:---|:---|:---|:---|
| Michaelis–Menten, $\dfrac{ax}{b+x}$ | $0$ | $a$ | $a/b$ | half-maximum at $x=b$ |
| Ricker, $axe^{-bx}$ | $0$ | $0$ | $a$ | peak at $x=1/b$ with height $a/(be)$; inflection at $2/b$ |

## Reparameterisation {#sec-reparameterisation}

A **reparameterisation** writes the same function with a different set of parameters. Each
new parameter is an expression of the old ones, and the curve does not change: for
corresponding values of the parameters, the two formulas return the same $y$ for every
$x$. What changes is the meaning of the parameters, and with it their units and what we
can read from them. To make sure that you are dealing with two parameterisation of the same
function, you can check with numbers that they produce the same curve, as we do below.

The disc equation of @sec-mechanistic is a reparameterisation of the Michaelis–Menten
function. If we divide its numerator and denominator by $\alpha H$:

$$
\frac{\alpha x}{1+\alpha Hx}=\frac{(1/H)\,x}{1/(\alpha H)+x},
\qquad\text{so}\qquad
a=\frac{1}{H},\quad b=\frac{1}{\alpha H}
\qquad\text{and back}\qquad
H=\frac{1}{a},\quad \alpha=\frac{a}{b}.
$$

That is, the asymptote is the inverse of the handling time and the attack rate is the
initial slope of the function.^[Biochemists write the same function with the maximum rate
of the reaction, $v_{\max}$, in the place of $a$ and the Michaelis constant, $K_m$, in the
place of $b$.] For the units we need to know over which time the response was measured.
In the reed frog experiment the tadpoles killed were counted over the whole experiment
(14 days, with three dragonfly larvae per tank), so the unit of time is one experiment.
With $a=60$ tadpoles killed and $b=120$ tadpoles per tank, the attack rate is
$\alpha=60/120=0.5$ per experiment: at low density, half of the tadpoles are killed in the
course of the experiment. The handling time is $H=1/60$ of an experiment per tadpole
killed, that is, the three larvae together cannot kill more than one tadpole every
$14/60\approx0.23$ days. In R:

```{r}
# The Holling type II form of the Michaelis–Menten function
holling_2 = function(x, alpha = 1, H = 1) {
  alpha * x / (1 + alpha * H * x)
}

# a = 60 and b = 120 correspond to alpha = a / b and H = 1 / a
x_values = c(5, 20, 50, 100)
michaelis_menten(x_values, a = 60, b = 120)
holling_2(x_values, alpha = 60 / 120, H = 1 / 60)

# and back: a = 1 / H and b = 1 / (alpha * H)
1 / (1 / 60)
1 / (0.5 * (1 / 60))
```

Both forms return the same four values, and converting back gives $a=60$ and $b=120$.

The Ricker function can also be written with the location of its peak as a parameter,
$f(x)=axe^{-x/c}$ with $c=1/b$, so that the parameter has the units of $x$ and can be read
from a plot:

```{r}
# Ricker function with the peak location c as a parameter: b = 1 / c
ricker_peak = function(x, a = 1, c = 1) {
  a * x * exp(-x / c)
}

ricker(x = c(5, 10, 20), a = 1, b = 0.2)
ricker_peak(x = c(5, 10, 20), a = 1, c = 5)
```

Both calls return the same values. The entries of the supplement list the other forms of
each function under "Also written". The
[logistic](../Supplements/deterministic-functions.qmd#sec-fn-logistic) function, for
example, has three forms in common use, and its entry shows how to convert between them.

There can be multiple reasons why several parameterisations of a function are in use. For
example, when derived mechanistically ($\alpha$ and $H$ measure something very
specific, ecologically speaking). Sometimes,  a parameter is easier to eyebal from a plot or
to compare among studies (a peak location, a midpoint, a half-maximum). Another important
reason is that the estimates of some sets of parameters are less correlated than those of others,
which makes a model easier to fit (we saw a similar example for Negative Binomial distribution
in [Chapter 4](../Chapter_4/Theory.qmd)).

## Derivatives with respect to parameters {#sec-parameter-derivatives}

The **partial derivative** of a function with respect to one of its parameters is the
derivative taken with $x$ and the other parameters held fixed. To show the parameters we
write the function as $f(x;a,b)$ and the partial derivative with respect to $a$ as
$\dfrac{\partial f}{\partial a}$. It tells us how much the function moves, at a given $x$,
when $a$ changes by a small amount, so it is itself a function of $x$. The rules of
differentiation are the same as before; what differs is the quantity that varies (see
[Partial derivatives](../Supplements/calculus.qmd#sec-calc-partial) in the calculus
refresher).

For the Michaelis–Menten function, $a$ multiplies the ratio $x/(b+x)$, which does not
contain $a$. For $b$ we write the function as $ax(b+x)^{-1}$ and use the power rule and
the chain rule:

$$
\frac{\partial f}{\partial a}=\frac{x}{b+x},
\qquad
\frac{\partial f}{\partial b}=-\frac{ax}{(b+x)^2}.
$$

In R we use `D()` with the name of the parameter:

```{r}
# Partial derivatives of the Michaelis–Menten function with respect to a and b
mm_wrt_a = D(mm_expression, "a")
mm_wrt_b = D(mm_expression, "b")
mm_wrt_a
mm_wrt_b

# Evaluated at x = 120 = b with a = 60 and b = 120
eval(mm_wrt_a, list(a = 60, b = 120, x = 120))
eval(mm_wrt_b, list(a = 60, b = 120, x = 120))

# The formulas derived on paper
120 / (120 + 120)
-60 * 120 / (120 + 120)^2
```

At a density of 120 tadpoles per tank, with $a=60$ and $b=120$, the two partial
derivatives are 0.5 and $-0.125$. That is, raising the asymptote by one tadpole killed
raises the curve by half a tadpole at that density, and raising the half-maximum by one
tadpole per tank lowers the curve by 0.125 tadpoles.

@fig-parameter_derivatives shows the two partial derivatives over a range of densities.
The derivative with respect to $a$ increases from 0 towards 1, so a change in $a$ moves
the curve most at large $x$, where the curve is close to its asymptote. The derivative
with respect to $b$ is negative (a larger half-maximum means a lower curve). It is zero at
both ends and largest in size at $x=b$, where it equals $-a/(4b)$. So a change in $b$
leaves the two ends of the curve where they are and moves the curve most around its
half-maximum. This has a consequence for data: the observations that can tell us about a
parameter are those made where the curve responds to it (densities around $b$ for $b$,
high densities for $a$). How much they tell us also depends on the variation of the
observations there and on the other parameters, which is a topic for Chapter 6.

Partial derivatives with respect to parameters return twice in the book. The delta method
of [Chapter 6](../Chapter_6/Theory.qmd) uses them to calculate the uncertainty of a fitted
curve from the uncertainty of its parameters, and the optimisers of
[Chapter 8](../Chapter_8/Theory.qmd) use them as the gradient that tells them in which
direction to move.

```{r}
#| echo: false
#| label: fig-parameter_derivatives
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Partial derivatives of the Michaelis–Menten function with respect to a (left) and to b (right), for a = 60 and b = 120. Each curve shows how much the function changes at a given x when the parameter increases by one unit."
library(patchwork)

x_values = seq(0, 600, length.out = 300)
wrt_a = data.frame(x = x_values, y = x_values / (120 + x_values))
wrt_b = data.frame(x = x_values, y = -60 * x_values / (120 + x_values)^2)

panel_wrt_a = ggplot(wrt_a, aes(x = x, y = y)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(x = "x", y = "Partial derivative with respect to a") +
  theme_classic()

panel_wrt_b = ggplot(wrt_b, aes(x = x, y = y)) +
  geom_vline(xintercept = 120, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(x = "x", y = "Partial derivative with respect to b") +
  theme_classic()

panel_wrt_a + panel_wrt_b
```

# Building new functions {#sec-function-building}

The functions that we have seen can be modified when none of them has the properties that
a response requires. This section presents three ways of doing so: shifting and scaling a
function, composing two functions, and adding a parameter that generalises a function.

## Shifting and scaling {#sec-function-shifts}

Any function can be stretched and moved along the two axes. There are four operations,
and each of them adds one parameter:

- scaling along the $x$ axis: replace $x$ by $x/b$, which stretches the curve
  horizontally by a factor $b$ (a scale parameter, as in @sec-function-scales);
- scaling along the $y$ axis: multiply the function by $a$, which stretches the curve
  vertically by a factor $a$;
- shifting along the $x$ axis: replace $x$ by $x-c$, which moves the curve to the right by
  $c$. Such a parameter is called a **location parameter**;
- shifting along the $y$ axis: add $d$ to the function, which moves the curve up by $d$.

Let's apply them one by one to the function $x/(1+x)$, which starts at zero, has an
asymptote of 1 and a half-maximum at $x=1$. Scaling $x$ gives $(x/b)/(1+x/b)=x/(b+x)$, and
scaling $y$ gives $ax/(b+x)$. So the Michaelis–Menten function is the simplest saturating
ratio with one scale parameter for each axis. Shifting it along both axes gives

$$
f(x)=d+\frac{a(x-c)}{b+(x-c)}, \qquad x\ge c,
$$

where the landmarks have moved with the curve: the function starts at the point $(c,d)$,
it reaches half of its rise at $x=c+b$, where its value is $d+a/2$, and it approaches
$d+a$ (@fig-shift_scale). The domain has moved as well. To the left of $c$ the formula
still returns numbers, but they are below $d$ and they become infinite at $x=c-b$, so the
shifted function is used from $c$ onwards only. In R, with $a=4$, $b=2$, $c=1$ and $d=1$:

```{r}
# Start from x / (1 + x), scale x by b, scale y by a, then shift by c and d
shifted_mm = function(x, a = 1, b = 1, c = 0, d = 0) {
  d + a * (x - c) / (b + (x - c))
}

# Values at c, c + b and for large x with a = 4, b = 2, c = 1, d = 1
shifted_mm(x = c(1, 3, 1000), a = 4, b = 2, c = 1, d = 1)
```

The function returns $d=1$ at $x=c$, $d+a/2=3$ at $x=c+b$, and a value close to $d+a=5$
for a large $x$. We have already seen a function of this kind: the curve for brain mass
against body mass in [Chapter 1](../Chapter_1/Theory.qmd#fig-parameter-uncertainty) was a
Michaelis–Menten function, shifted and scaled along both axes.

```{r}
#| echo: false
#| label: fig-shift_scale
#| fig-cap: "Michaelis–Menten function after scaling and shifting, d + a(x − c) / (b + (x − c)) with a = 4, b = 2, c = 1 and d = 1 for x ≥ c. The function starts at (c, d), reaches d + a/2 at x = c + b and approaches d + a."
shift_data = data.frame(x = seq(1, 15, length.out = 200))
shift_data$y = shifted_mm(shift_data$x, a = 4, b = 2, c = 1, d = 1)

ggplot() +
  geom_hline(yintercept = c(1, 3, 5), linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(1, 3), linetype = "dashed", colour = "grey40") +
  geom_line(data = shift_data, aes(x = x, y = y),
            colour = "#D55E00", linewidth = 0.8) +
  scale_x_continuous(breaks = c(1, 3), labels = c("c", "c + b")) +
  scale_y_continuous(breaks = c(1, 3, 5), labels = c("d", "d + a / 2", "d + a")) +
  labs(x = "x", y = "y") +
  theme_classic() +
  theme(axis.title.y = element_text(margin = margin(r = 12)))
```

## Range and composition {#sec-function-composition}

The **composition** of two functions applies one function to the output of the other:
$f(x)=g(h(x))$, where $h$ is the inner function and $g$ the outer function. In this book
we use composition to give a function the range that a quantity requires. The mean of a
count must be positive and a probability must lie between 0 and 1, and a straight line
$a+bx$ respects neither of these restrictions. If we pass the line through an outer
function whose range is the required one, the result respects them for any value of $x$
and of the parameters.

For a positive quantity the outer function is the exponential: $e^{a+bx}$ is positive
whatever the values of $a$, $b$ and $x$. Since $e^{a+bx}=e^{a}e^{bx}$, this is an
exponential function that starts at $e^{a}$ and grows when $b>0$ or decays when $b<0$, so
here $a$ and $b$ can take any real value. Every increase of $x$ by one unit multiplies the
function by $e^{b}$.

For a quantity between 0 and 1 the outer function is the **inverse logit**,
$\operatorname{logit}^{-1}(z)=1/(1+e^{-z})$, which is `plogis()` in R. It lies between 0
and 1 for any $z$. The inverse logit of a line, $1/(1+e^{-(a+bx)})$, is the
[logistic](../Supplements/deterministic-functions.qmd#sec-fn-logistic) function with an
upper asymptote of 1. It is a sigmoid curve that is steepest at its midpoint, $x=-a/b$,
where the line equals zero, the function equals 0.5 and its slope is $b/4$. The exponential
and the inverse logit are the two functions that we used in
[Chapter 4](../Chapter_4/Theory.qmd#sec-maximum-likelihood) (section "Constraints and
transformations") to keep a parameter within its constraints during optimisation, with
`exp()` for a positive parameter and `plogis()` for a probability. Here the same idea is
applied to a function of $x$.

The inner function does not have to be a straight line. If it is a parabola that opens
downwards, the composition is a hump between 0 and 1:

$$
f(x)=\operatorname{logit}^{-1}\left[d-b(x-c)^2\right]=\frac{1}{1+e^{-\left(d-b(x-c)^2\right)}}.
$$

To find its peak we use the chain rule, $f'(x)=g'(h(x))\,h'(x)$. The inverse logit
increases everywhere, so $g'$ is positive and $f'(x)$ is zero only where
$h'(x)=-2b(x-c)=0$, that is, at $x=c$. In other words, an increasing outer function
leaves the peaks of the inner function where they are. The height of the peak is
$\operatorname{logit}^{-1}(d)$. In R, with $d=1$, $b=0.1$ and $c=5$:

```{r}
# The inverse logit of a line: plogis() and the formula written out
x_values = seq(0, 10, by = 2.5)
plogis(-4 + 0.8 * x_values)
1 / (1 + exp(-(-4 + 0.8 * x_values)))

# The inverse logit of a quadratic: a hump between 0 and 1 with its peak at c
hump = function(x, b = 1, c = 0, d = 0) {
  plogis(d - b * (x - c)^2)
}

hump(x = c(0, 5, 10), b = 0.1, c = 5, d = 1)
plogis(1)
```

The first two lines of output are identical: `plogis()` is the inverse logit, and for the
line $-4+0.8x$ it returns 0.5 at the midpoint, $x=4/0.8=5$. The hump has its highest value, 0.731, at
$x=5$, and that value is $\operatorname{logit}^{-1}(1)$. @fig-composition shows the three
compositions. In [Chapter 6](../Chapter_6/Theory.qmd) we will use them to link a predictor
to the mean of a Poisson distribution and to the probability of a Bernoulli distribution.

:::: {.callout-important title="Practice: seedlings along a moisture gradient"}

The expected number of seedlings in a quadrat is described as $e^{a+bx}$, where $x$ is the
soil moisture (%), $a=-2$ and $b=0.4$ per unit of moisture. What is the expected number at
$x=5$? By which factor does it change when the moisture increases by one unit, and by 5
units?

::: {.callout-tip title="Solution" collapse="true"}

At $x=5$ the exponent is $-2+0.4\times5=0$, so the expected number is $e^{0}=1$ seedling
per quadrat. An increase of one unit multiplies the function by $e^{0.4}\approx1.49$, and
an increase of 5 units by $e^{0.4\times5}=e^{2}\approx7.4$. The change is a factor, the
same at any moisture, and the function stays positive.

:::
::::

```{r}
#| echo: false
#| label: fig-composition
#| fig-width: 10
#| fig-height: 3.5
#| fig-cap: "Three compositions that give a function the range it needs. Left: exp(a + bx) is positive (a = −2, b = 0.4). Centre: the inverse logit of a line lies between 0 and 1 (a = −4, b = 0.8). Right: the inverse logit of a quadratic is a hump between 0 and 1 with its peak at c (d = 1, b = 0.1, c = 5)."
library(patchwork)

x_values = seq(0, 10, length.out = 200)
positive_data = data.frame(x = x_values, y = exp(-2 + 0.4 * x_values))
sigmoid_data = data.frame(x = x_values, y = plogis(-4 + 0.8 * x_values))
hump_data = data.frame(x = x_values, y = hump(x_values, b = 0.1, c = 5, d = 1))

panel_positive = ggplot(positive_data, aes(x = x, y = y)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Positive", x = "x", y = "y") +
  theme_classic()

panel_sigmoid = ggplot(sigmoid_data, aes(x = x, y = y)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Between 0 and 1", x = "x", y = "y") +
  theme_classic()

panel_hump = ggplot(hump_data, aes(x = x, y = y)) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(title = "Hump between 0 and 1", x = "x", y = "y") +
  theme_classic()

panel_positive + panel_sigmoid + panel_hump
```

## Nested functions {#sec-nested-functions}

A function is **nested** in another function when it is the special case that we obtain
by fixing one of the parameters of the other at a particular value. Adding a parameter is
the usual way of making a function more flexible, and comparing a function with its
generalisation is the most common comparison of models.
[Chapter 4](../Chapter_4/Theory.qmd#sec-aic) introduced AIC for comparing models, and in
[Chapter 6](../Chapter_6/Theory.qmd) we will use it to compare response curves.

Let's go back to the size experiment, for which the Ricker function was too wide. The
**generalised Ricker** function^[Persson et al. (1998) used this function for the attack
rate of a predator as a function of its body mass: "Ontogenetic scaling of foraging rates
and the dynamics of a size-structured consumer–resource model", *Theoretical Population
Biology* 54: 270–293.] is

$$
f(x)=a\left[\frac{x}{b}\,e^{1-x/b}\right]^{c},
$$

where the term in brackets is a Ricker function written so that its peak is at $x=b$ and
has a height of 1. Raising it to the power $c$ leaves the peak where it is, so for any $c$
the parameter $b$ is the location of the peak (in mm) and $a$ is its height (in tadpoles
killed). The exponent $c$ has no units and controls the width: away from the peak the term
in brackets is below 1, and a number below 1 becomes smaller when it is raised to a higher
power, so a larger $c$ gives a narrower peak. With $c=1$ the function is
$(ae/b)\,xe^{-x/b}$, a Ricker function with parameters $ae/b$ and $1/b$. So the Ricker
function is nested in the generalised Ricker function. In R:

```{r}
generalised_ricker = function(x, a = 1, b = 1, c = 1) {
  a * ((x / b) * exp(1 - x / b))^c
}

# With c = 1 it is a Ricker function with parameters a * e / b and 1 / b
x_values = c(5, 12, 20, 30)
generalised_ricker(x_values, a = 4, b = 12, c = 1)
ricker(x_values, a = 4 * exp(1) / 12, b = 1 / 12)

# With c = 30 the peak at b = 12 is much narrower
generalised_ricker(x_values, a = 4, b = 12, c = 30)
```

The first two lines of output are identical, as they should be. With $c=30$ the function
still returns 4 tadpoles killed at 12 mm, but at 20 mm it has fallen to 0.04, against 3.4
for $c=1$. @fig-nested shows the data with the Ricker function of @fig-candidate_curves
and with this generalised Ricker function, which places a narrow peak at 12 mm as the data
require. Note that a narrow peak needs a large exponent.

Other functions are generalised in the same way. The
[Shepherd](../Supplements/deterministic-functions.qmd#sec-fn-shepherd) function,
$ax/(b+x^{c})$, and the
[Hassell](../Supplements/deterministic-functions.qmd#sec-fn-hassell) function,
$ax/(b+x)^{c}$, both contain the Michaelis–Menten function as the case $c=1$, and the
[Richards](../Supplements/deterministic-functions.qmd#sec-fn-richards) function contains
the logistic function. The cost of the additional flexibility is that exponents such as
$c$ are hard to estimate from data, a point that we take up in Chapters 6 and 8.

```{r}
#| echo: false
#| label: fig-nested
#| fig-cap: "Tadpoles killed out of 10 against body length, with the Ricker function of the first figure (dashed line, a = 1, b = 0.15) and a generalised Ricker function (solid line, a = 4, b = 12 and c = 30). Setting c = 1 in the generalised function gives a Ricker function."
size_grid = seq(0, 40, length.out = 400)
curve_data = data.frame(
  TBL = rep(size_grid, times = 2),
  Kill = c(ricker(size_grid, a = 1, b = 0.15),
           generalised_ricker(size_grid, a = 4, b = 12, c = 30)),
  curve = rep(c("Ricker", "Generalised Ricker"), each = length(size_grid))
)

ggplot() +
  geom_point(data = ReedfrogSizepred, aes(x = TBL, y = Kill),
             colour = "#0072B2", alpha = 0.7, size = 2) +
  geom_line(data = curve_data, aes(x = TBL, y = Kill, linetype = curve),
            colour = "#D55E00", linewidth = 0.8) +
  scale_linetype_manual(values = c("Ricker" = "dashed", "Generalised Ricker" = "solid")) +
  labs(x = "Tadpole body length (mm)", y = "Tadpoles killed out of 10 in 3 days",
       linetype = NULL) +
  theme_classic()
```

# Choosing a function {#sec-choosing-function}

The two functions of this chapter are a small part of the functions that are used in
ecology. The others are described in the
[Supplement](../Supplements/deterministic-functions.qmd), with the formula, the landmarks,
a worked example and a practice exercise for each of them. You should read at least the
entries of the functions that the practical names.

To choose a function for a response we go through the questions that the tools of this
chapter answer. Where does the response start: at zero or at a positive value, and with a
slope or flat? Where does it end: at an asymptote, at zero, or does it keep increasing? Is
it monotonic in between, or does it have a maximum? If it saturates, does it bend
downwards from the start (concave everywhere) or does it accelerate first (sigmoid)? And
is there a mechanism that suggests a formula? The table below gives the answers for the
functions of the supplement.

| Function | Domain | Left end | Right end | Middle |
|:---|:---|:---|:---|:---|
| [Linear](../Supplements/deterministic-functions.qmd#sec-fn-linear) | all $x$ | straight line | straight line | monotonic |
| [Quadratic](../Supplements/deterministic-functions.qmd#sec-fn-quadratic) | all $x$ | grows without bound | grows without bound | one maximum or minimum |
| [Threshold](../Supplements/deterministic-functions.qmd#sec-fn-threshold) | all $x$ | flat | flat | one jump |
| [Hockey stick](../Supplements/deterministic-functions.qmd#sec-fn-hockey-stick) | all $x$ | straight line | flat | one breakpoint |
| [Hyperbolic](../Supplements/deterministic-functions.qmd#sec-fn-hyperbolic) | $x\ge0$ | finite value | tends to 0 | decreasing |
| [Michaelis–Menten](../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten) | $x\ge0$ | 0, straight line | asymptote | saturating |
| [Holling type III](../Supplements/deterministic-functions.qmd#sec-fn-holling3) | $x\ge0$ | 0, accelerating | asymptote | sigmoid |
| [Holling type IV](../Supplements/deterministic-functions.qmd#sec-fn-holling4) | $x\ge0$ | 0, accelerating | asymptote | hump before the asymptote when $c<0$ |
| [Exponential decay](../Supplements/deterministic-functions.qmd#sec-fn-exponential) | $x\ge0$ | finite value | tends to 0 | decreasing |
| [Saturating exponential](../Supplements/deterministic-functions.qmd#sec-fn-monomolecular) | $x\ge0$ | 0, straight line | asymptote | saturating |
| [Ricker](../Supplements/deterministic-functions.qmd#sec-fn-ricker) | $x\ge0$ | 0, straight line | tends to 0 | one maximum |
| [Logistic](../Supplements/deterministic-functions.qmd#sec-fn-logistic) | all $x$ | close to 0, accelerating | asymptote | sigmoid |
| [Gompertz](../Supplements/deterministic-functions.qmd#sec-fn-gompertz) | all $x$ | close to 0, accelerating | asymptote | sigmoid, asymmetric |
| [Bell-shaped](../Supplements/deterministic-functions.qmd#sec-fn-gaussian) | all $x$ | tends to 0 | tends to 0 | one maximum, symmetric |
| [Power law](../Supplements/deterministic-functions.qmd#sec-fn-power) | $x>0$ | 0 or unbounded | 0 or unbounded | monotonic |
| [Shepherd](../Supplements/deterministic-functions.qmd#sec-fn-shepherd) and [Hassell](../Supplements/deterministic-functions.qmd#sec-fn-hassell) | $x\ge0$ | 0, straight line | asymptote if $c=1$, tends to 0 if $c>1$ | saturating, or one maximum if $c>1$ |
| [Generalised Ricker](../Supplements/deterministic-functions.qmd#sec-fn-generalised-ricker) | $x\ge0$ | 0 | tends to 0 | one maximum of adjustable width |
| [Richards](../Supplements/deterministic-functions.qmd#sec-fn-richards) | all $x$ | close to 0 | asymptote | sigmoid of adjustable asymmetry |

The supplement also has entries for functions that join two straight lines
([piecewise linear](../Supplements/deterministic-functions.qmd#sec-fn-piecewise-linear)),
for growth curves
([von Bertalanffy](../Supplements/deterministic-functions.qmd#sec-fn-von-bertalanffy)) and
for a saturating function whose bend can be made sharper or smoother (the
[non-rectangular hyperbola](../Supplements/deterministic-functions.qmd#sec-fn-nonrectangular)).

Note that several functions give the same answers. The saturating exponential and the
Michaelis–Menten function both start at zero as a straight line and level off at an
asymptote, and the logistic and the Holling type III functions are both sigmoid. Within
each pair, the function built from an exponential approaches its asymptote faster than the
one built from a ratio, but over the range of a typical set of data the two differ
little. The choice between them then rests on a mechanism, on what is customary in a
field, or on which parameters are easier to interpret. As we said for distributions in
[Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions), there is no correct
function for a given set of data. When several candidates remain, we compare the fitted
models formally (see [Chapter 4](../Chapter_4/Theory.qmd#sec-aic) and
[Chapter 6](../Chapter_6/Theory.qmd)).

<!-- AUTHOR: how you choose between candidate functions in your own work. -->

## Starting values by eye {#sec-starting-values}

The optimiser of Chapter 4 needs a starting value for every parameter. For the parameters
of a distribution we used the method-of-moments estimates. For the parameters of a
function we read them from a plot of the data. We look in the plot for the landmarks of
the function (a value at zero, an initial slope, an asymptote, a peak, a half-maximum),
we calculate the parameters from them with the expressions of @sec-meaning-parameters,
and then we draw the curve over the data and adjust the values. Let's do this for the two
reed frog experiments.

For the experiment on density we use the Michaelis–Menten function, whose landmarks are
the initial slope $a/b$ and the asymptote $a$. The initial slope can be read from the
data: at the two lowest densities about half of the tadpoles were killed (1 and 2 out of
5, 5 and 6 out of 10), so $a/b\approx0.5$. The asymptote cannot be read, because the
numbers killed are still increasing at 100 tadpoles per tank. All that we can say is that
$a$ lies above the highest number killed (35). If we take $a=60$ tadpoles killed, the
initial slope gives $b=60/0.5=120$ tadpoles per tank. We draw this curve over the data,
together with two curves that have the same initial slope and an asymptote that is two and
four times higher:

```{r}
#| label: fig-start_density
#| fig-cap: "Reed frog functional response with three Michaelis–Menten curves that share the initial slope a/b = 0.5 and differ in the asymptote: a = 60 (solid), a = 120 (dashed) and a = 240 (dotted)."
# Tadpoles killed against initial density, with candidate curves added one by one
plot(Killed ~ Initial, data = ReedfrogFuncresp,
     xlab = "Initial number of tadpoles per tank", ylab = "Tadpoles killed in 14 days")
curve(michaelis_menten(x, a = 60, b = 120), add = TRUE, lty = 1)
curve(michaelis_menten(x, a = 120, b = 240), add = TRUE, lty = 2)
curve(michaelis_menten(x, a = 240, b = 480), add = TRUE, lty = 3)
```

The three curves of @fig-start_density predict 27, 35 and 41 tadpoles killed at the
highest density, a difference of the same size as the one between the two tanks at that
density (25 and 35 tadpoles killed). So these data tell us the ratio $a/b$ but little
about $a$ and $b$ separately, which is what we expected from an experiment that covers
less than half of the rise of the curve (@sec-function-limits). Any of the three pairs is
a reasonable starting value. What this means for the estimates and for their uncertainty
is a question for [Chapter 6](../Chapter_6/Theory.qmd).

For the experiment on body length we use the generalised Ricker function, whose peak
location and peak height are parameters. Most tadpoles were killed at 12 mm, where the
three tanks lost 3, 4 and 5 tadpoles, so we read $b=12$ mm and $a=4$ tadpoles killed. The
exponent $c$ has no landmark that we can read, so we find it by trying values:

```{r}
#| label: fig-start_size
#| fig-cap: "Reed frog size experiment with three generalised Ricker curves that share the peak read from the plot (a = 4, b = 12) and differ in the exponent: c = 5 (solid), c = 20 (dashed) and c = 50 (dotted)."
# Tadpoles killed against body length: the peak is at about 12 mm with a height
# of about 4; the exponent c is found by trying values
plot(Kill ~ TBL, data = ReedfrogSizepred,
     xlab = "Tadpole body length (mm)", ylab = "Tadpoles killed out of 10 in 3 days")
curve(generalised_ricker(x, a = 4, b = 12, c = 5), add = TRUE, lty = 1)
curve(generalised_ricker(x, a = 4, b = 12, c = 20), add = TRUE, lty = 2)
curve(generalised_ricker(x, a = 4, b = 12, c = 50), add = TRUE, lty = 3)
```

With $c=5$ the peak is too wide (@fig-start_size): the curve predicts 3.3 tadpoles killed
at 9 mm and 1.5 at 21 mm, where the tanks lost 0 to 2 tadpoles and none. With $c=20$ and
$c=50$ the curve is close to zero from 21 mm onwards, as in the data, and the two values
differ mostly in what they predict at 9 mm (1.9 and 0.6 tadpoles killed). A value between
20 and 50 is therefore a reasonable start, and it is the reason why we used $c=30$ in
@fig-nested.

Values read in this way are where a fit starts. In Chapter 6 the optimiser takes them and
improves them by maximum likelihood, and the better we understand the function, the closer
our starting values are to the estimates.

# Summary

A deterministic function assigns one output to each input for given values of its
parameters. In an ecological model it describes how an expected response changes with a
predictor. A function is phenomenological when it only describes the shape of the response
and mechanistic when it is derived from assumptions about the process, as the disc
equation is from the time that a predator spends searching and handling prey. The units of
the parameters follow from the formula. We analysed two functions in full, the
Michaelis–Menten function (saturating) and the Ricker function (unimodal), and the
supplement describes the others.

The shape of a function is found from its formula with a few tools. Limits give the two
ends of the function and its asymptotes. The derivative gives the slope at any point and
the second derivative the curvature, concave or convex. From them we obtain the special
points of a function: a maximum where the derivative is zero and changes sign, an
inflection point where the second derivative changes sign, and a half-maximum or a
characteristic scale by solving the function for $x$. Each of these results can be checked
in R, with `D()`, with a finite difference, on a grid or with `uniroot()`.

The special points give the parameters a meaning. The asymptote and the half-maximum are
the parameters of the Michaelis–Menten function, and the initial slope and the location of
the peak of the Ricker function are $a$ and $1/b$. A reparameterisation writes the same curve with other parameters, such as
the attack rate and the handling time. A partial derivative with respect to a parameter
shows where along $x$ the curve responds to that parameter.

New functions are built from existing ones by shifting and scaling, by composition, which
gives a function the range that a quantity requires, and by adding a parameter, so that
the original function is nested in the new one.

A function is chosen by asking where the response starts and ends, whether it has a
maximum and whether a mechanism suggests a formula, and the starting values of its
parameters are read from a plot through its landmarks. In
[Chapter 6](../Chapter_6/Theory.qmd) the two pillars of the book come together: the mean
of a distribution follows one of these functions, and its parameters are estimated by
maximum likelihood, starting from values read by eye.

# Chapter quiz

The questions below follow the order of the chapter. Try each question before opening its
solution.

:::: {.callout-important title="Question 1: seed dispersal"}

The density of seeds on the ground (seeds/m²) at a distance $x$ (m) from a parent tree
is described by $f(x)=ae^{-bx}$. What are the domain and the range of this function? What
are the units of $a$ and $b$?

::: {.callout-tip title="Answer" collapse="true"}

A distance cannot be negative, so the domain is $x\ge0$. The function equals $a$ at $x=0$
and decreases towards zero without reaching it, so its range is $0<y\le a$. The parameter
$a$ has the units of $y$ (seeds/m²). The argument of the exponential has no units, so
$b$ has the inverse units of $x$ (1/m).

:::
::::

:::: {.callout-important title="Question 2: ladybirds and aphids"}

A ladybird on a plant with $N$ aphids encounters a fraction $\alpha=0.1$ of the
aphids per hour of searching and needs $H=0.25$ hours to handle each aphid. According to the disc
equation, how many aphids does it eat per hour on a plant with 40 aphids? What is the
highest feeding rate that it can reach, and why can it not exceed it?

::: {.callout-tip title="Answer" collapse="true"}

The feeding rate is $\alpha N/(1+\alpha HN)=0.1\times40/(1+0.1\times0.25\times40)=4/2=2$
aphids/hour. The highest rate is $1/H=4$ aphids/hour: a ladybird that spent all of
its time handling aphids, and none searching, would eat one aphid every quarter of an
hour. At 40 aphids per plant it eats half of that, so 40 aphids is the half-maximum,
$1/(\alpha H)$.

:::
::::

:::: {.callout-important title="Question 3: pitfall traps"}

The number of beetle species found with $x$ pitfall traps is described by the saturating
exponential function $f(x)=40\left(1-e^{-0.1x}\right)$. What is the value of the function
at $x=0$ and what is its limit as $x\to\infty$? How many species does it give for 10
traps?

::: {.callout-tip title="Answer" collapse="true"}

At $x=0$, $e^{0}=1$ and $f(0)=0$: without traps no species are found. As $x\to\infty$,
$e^{-0.1x}\to0$ and $f(x)\to40$ species, which is the asymptote. For 10 traps the function
gives $40(1-e^{-1})\approx25.3$ species, that is, 63% of the asymptote.

:::
::::

:::: {.callout-important title="Question 4: nitrate uptake"}

The uptake of nitrate by roots (µmol/g/hour) as a function of the concentration of
nitrate $x$ (µmol/L) is $f(x)=8x/(2+x)$. Calculate the slope of the function at
$x=0$ and at $x=2$. What are the units of the slope?

::: {.callout-tip title="Answer" collapse="true"}

The derivative of the Michaelis–Menten function is $f'(x)=ab/(b+x)^2=16/(2+x)^2$. At $x=0$
it is $16/4=4$ and at $x=2$ it is $16/16=1$. The units are those of $y$ divided by those
of $x$: (µmol/g/hour)/(µmol/L). That is, at a concentration of 2
µmol/L an additional µmol/L raises the uptake by about 1 µmol/g/hour, a
quarter of the increase at very low concentration.

:::
::::

:::: {.callout-important title="Question 5: salmon recruitment"}

The number of recruits of a salmon stock (in thousands) as a function of the number of
spawners $x$ (in thousands) is described by the Ricker function $f(x)=3xe^{-0.01x}$. At
which number of spawners is recruitment highest, and how many recruits does the function
give there?

::: {.callout-tip title="Answer" collapse="true"}

The derivative $f'(x)=a(1-bx)e^{-bx}$ is zero at $x=1/b=1/0.01=100$ thousand spawners. It
is positive below that value and negative above it, so this is a maximum. The number of
recruits there is $a/(be)=3/(0.01e)\approx110$ thousand.

:::
::::

:::: {.callout-important title="Question 6: bumblebees and flowers"}

The number of flowers that a bumblebee visits per minute as a function of the density of
flowers $x$ (flowers/m²) is $f(x)=12x/(30+x)$. At which density does the bumblebee
reach half of its highest visiting rate? Which fraction of that rate does it reach at 90
flowers/m², and which density does it need to reach 90% of it?

::: {.callout-tip title="Answer" collapse="true"}

The asymptote is $a=12$ flowers/min and the half-maximum is $b=30$ flowers/m²,
where the function gives 6 flowers/min. A density of 90 flowers/m² is $3b$, so
the function reaches 3/4 of its asymptote there: $12\times90/120=9$ flowers/min. For
90% we solve $x/(30+x)=0.9$, which gives $x=9b=270$ flowers/m². The approach to the
asymptote is slow: nine times the half-maximum is needed for the last 40% of the rise.

:::
::::

:::: {.callout-important title="Question 7: mayflies in a light trap"}

The number of mayflies caught per night in a light trap, $x$ nights after the first
mayflies emerged, is given in a paper as $f(x)=5xe^{-x/8}$. Write this function in the
form $axe^{-bx}$, with the units of $a$ and $b$. On which night is the catch highest, and
how many mayflies does the function give for that night? On which night does the catch
decline fastest?

::: {.callout-tip title="Answer" collapse="true"}

The paper uses the form of the Ricker function with the location of the peak as a
parameter, $axe^{-x/c}$ with $c=8$ nights. So $a=5$ mayflies/night/night (the
initial slope) and $b=1/c=0.125$/night. The catch is highest on night $1/b=8$, with
$a/(be)=40/e\approx14.7$ mayflies. It declines fastest at the inflection point,
$2/b=16$ nights. Both formulas describe the same curve.

:::
::::

:::: {.callout-important title="Question 8: photosynthesis of a seedling"}

The net photosynthesis of a seedling is $-2$ µmol/m²/s in the dark. It
increases with light and approaches 18 µmol/m²/s, and it is half-way between
those two values at a light intensity of 150 µmol/m²/s. Write a shifted and
scaled Michaelis–Menten function with these properties.

::: {.callout-tip title="Answer" collapse="true"}

The function starts at $x=0$, so there is no shift along $x$ ($c=0$), and it starts at
$-2$, so $d=-2$. The rise from $-2$ to 18 is $a=20$, and half of that rise is reached at
$b=150$. So

$$
f(x)=-2+\frac{20x}{150+x}.
$$

We can check it at $x=150$: $-2+20\times150/300=8$, which is half-way between $-2$ and 18.

:::
::::

:::: {.callout-important title="Question 9: occurrence of a forest herb"}

The probability that a forest herb occurs in a plot is described as
$\operatorname{logit}^{-1}(-3+0.2x)$, where $x$ is the canopy cover (%). Which two
functions are composed here? What is the range of the composition, and why can the
straight line $-3+0.2x$ not be used directly as a probability? At which canopy cover is
the probability 0.5?

::: {.callout-tip title="Answer" collapse="true"}

The inner function is the line $-3+0.2x$ and the outer function is the inverse logit. The
line is negative for $x<15$ and exceeds 1 for $x>20$, values that a probability cannot
take. The inverse logit lies between 0 and 1 for any input, so the composition does too.
The probability is 0.5 where the line equals zero, at a canopy cover of $3/0.2=15$%.

:::
::::

:::: {.callout-important title="Question 10: beetles and bait"}

In a plot of the number of beetles caught per trap against the amount of bait (g), the
points rise from zero, bend, and level off at about 40 beetles. They pass through 20
beetles at about 5 g of bait. Which function would you try first, and with which starting
values? Which initial slope do those values imply?

::: {.callout-tip title="Answer" collapse="true"}

The response starts at zero and saturates without accelerating first, so a
Michaelis–Menten function is a natural candidate. The asymptote is read on the vertical
axis, $a=40$ beetles per trap, and the half-maximum on the horizontal axis, $b=5$ g, as the
amount of bait at which half of the asymptote is reached. These values imply an initial
slope of $a/b=8$ beetles/g of bait, which can be compared with the first points of the
plot.

:::
::::
