---
title: "Chapter 6: Deterministic functions in ecology — solutions"
date: today
---

The [student practical](no_solution.qmd) includes the fully worked
Exercise 1. These solutions answer Exercises 2–8 with the paper
reasoning first and R checks afterwards. The examples are illustrative,
not field observations.

# Exercise 2: Values at the ends of curves

For the light-intensity curve, $D(0)=20e^0=20$ light units.
As $z\to\infty$, $e^{-0.4z}\to0$, so the intensity approaches
zero. To find the half-depth, solve

$$
\begin{aligned}
20e^{-0.4z}&=10,\\
e^{-0.4z}&=\frac12,\\
-0.4z&=\log(1/2)=-\log2,\\
z&=\frac{\log2}{0.4}\approx1.73\text{ m}.
\end{aligned}
$$

For $H(N)=15N/(5+N)$, $H(0)=0$. Dividing top and bottom by
$N$ gives $H(N)=15/(5/N+1)\to15$. If $N$ is much smaller than
5, then $5+N\approx5$ and $H(N)\approx3N$. At $N=5$,
$H(5)=15(5)/10=7.5$, half the asymptote.

For $C(N)=8/(2+N)$, $C(0)=4$ and $C(N)\to0$ as $N$ grows.
Its derivative, $C'(N)=-8/(2+N)^2$, is negative throughout
$N\geq0$, confirming that the predicted per-plant seed production
falls with planting density.

For $S(N)=3N^2/(4+N^2)$, $S(0)=0$. Dividing top and bottom
by $N^2$ shows that $S(N)\to3$. When $N$ is small relative
to 2, $N^2\ll4$, so $S(N)\approx(3/4)N^2$. It initially
rises quadratically rather than linearly.

The following plots check the limits and the different initial shapes:

```{r}
old_par = par(mfrow = c(2, 2))
curve(20 * exp(-0.4 * x), from = 0, to = 12,
      xlab = "Canopy depth (m)", ylab = "Light intensity (arbitrary units)")
curve(15 * x / (5 + x), from = 0, to = 40,
      xlab = "Density N", ylab = "Consumption H(N)")
curve(8 / (2 + x), from = 0, to = 20,
      xlab = "Planting density N", ylab = "Seeds per plant C(N)")
curve(3 * x^2 / (4 + x^2), from = 0, to = 15,
      xlab = "Density N", ylab = "Response S(N)")
par(old_par)
```

# Exercise 3: Derivative workshop

For $f(x)=3+2x^2$, the constant has derivative zero and the
power rule gives

$$
f'(x)=0+2(2x)=4x,\qquad f'(2)=8.
$$

For $g(t)=12e^{-0.3t}$, the chain rule contributes the
derivative $-0.3$ of the exponent:

$$
g'(t)=12e^{-0.3t}(-0.3)=-3.6e^{-0.3t},
\qquad g'(0)=-3.6.
$$

The negative slope says light or another declining quantity
decreases as $t$ increases. Its numerical units depend on the
units assigned to $g$ and $t$.

For $h(N)=10N/(4+N)$, let $u=10N$ and $v=4+N$.
Then $u'=10$ and $v'=1$. The quotient rule gives

$$
\begin{aligned}
h'(N)
&=\frac{10(4+N)-10N}{(4+N)^2}\\
&=\frac{40}{(4+N)^2},
\qquad h'(0)=\frac{40}{16}=2.5.
\end{aligned}
$$

The numerator stays fixed while the positive denominator grows,
so the slope decreases but remains positive.

For $R(N)=5Ne^{-0.2N}$, the product rule differentiates $5N$
and the exponential separately. The chain rule contributes
$-0.2$ when differentiating the exponential:

$$
\begin{aligned}
R'(N)
&=5e^{-0.2N}+5N(-0.2)e^{-0.2N}\\
&=5e^{-0.2N}(1-0.2N).
\end{aligned}
$$

Finally, for $j(N)=N^2/(9+N^2)$, use the quotient rule with
$u=N^2$, $u'=2N$, $v=9+N^2$, and $v'=2N$:

$$
\begin{aligned}
j'(N)
&=\frac{2N(9+N^2)-N^2(2N)}{(9+N^2)^2}\\
&=\frac{18N}{(9+N^2)^2},
\qquad j'(0)=0.
\end{aligned}
$$

The power rule supplies the derivatives in the numerator and
denominator; no nested function needs an additional chain-rule
step in this example.

One numerical check evaluates the functions immediately either
side of $x=1$. It should be close to each analytical derivative
at $x=1$:

```{r}
central_slope = function(fun, x, step = 0.0001) {
  (fun(x + step) - fun(x - step)) / (2 * step)
}
f = function(x) 3 + 2 * x^2
g = function(t) 12 * exp(-0.3 * t)
h = function(N) 10 * N / (4 + N)
ricker = function(N) 5 * N * exp(-0.2 * N)
j = function(N) N^2 / (9 + N^2)

analytical = c(
  f = 4,
  g = -3.6 * exp(-0.3),
  h = 40 / 5^2,
  ricker = 5 * exp(-0.2) * (1 - 0.2),
  j = 18 / 10^2
)
numerical = c(
  f = central_slope(f, 1),
  g = central_slope(g, 1),
  h = central_slope(h, 1),
  ricker = central_slope(ricker, 1),
  j = central_slope(j, 1)
)
data.frame(analytical, numerical)
```

Agreement at one point checks the arithmetic locally. The
paper derivations establish how the slope depends on the
whole input range.

# Exercise 4: Peaks and an inflection point

For $Q(N)=6N-N^2$,
$Q'(N)=6-2N$. This derivative is positive for $N<3$,
zero at 3, and negative for $N>3$. Thus $Q$ rises then falls,
with a maximum of $Q(3)=18-9=9$. At both boundaries of
$0\leq N\leq6$, $Q=0$. The boundary check confirms that
9 is the global maximum on the specified domain.

For the Ricker curve,
$R'(N)=5e^{-0.2N}(1-0.2N)$. Since the first factor is
positive, the slope changes from positive to negative when
$1-0.2N=0$, or $N=5$. The peak height is

$$
R(5)=5(5)e^{-1}=\frac{25}{e}\approx9.20.
$$

$R(0)=0$ and $R(N)\to0$ as $N\to\infty$, so the peak is
the global maximum for $N\geq0$.

For the logistic, write $k=0.8$ and $L'=kL(1-L)$.
Apply the product and chain rules:

$$
\begin{aligned}
L''
&=k\left[L'(1-L)-LL'\right]\\
&=kL'(1-2L)\\
&=k^2L(1-L)(1-2L).
\end{aligned}
$$

At $x=5$, the exponent is zero and $L(5)=1/2$. The
factor $1-2L$ changes from positive to negative there,
so $(5,1/2)$ is the inflection point. Its slope is
$L'(5)=0.8(1/2)(1/2)=0.2$ response units per input unit.

```{r}
#| fig-width: 11
#| fig-height: 4
old_par = par(mfrow = c(1, 3), mar = c(4, 4, 1, 1))
curve(6 * x - x^2, from = 0, to = 6, xaxt = "n",
      xlab = "Input N", ylab = "Recruitment index Q(N)")
axis(side = 1, at = c(0, 3, 6))
points(x = 3, y = 9, pch = 19)
curve(5 * x * exp(-0.2 * x), from = 0, to = 25,
      xaxt = "n",
      xlab = "Adult density N", ylab = "Recruitment R(N)")
axis(side = 1, at = c(0, 5, 10, 15, 20, 25))
points(x = 5, y = 25 / exp(1), pch = 19)
curve(1 / (1 + exp(-0.8 * (x - 5))), from = 0, to = 10,
      ylim = c(0, 1), xaxt = "n",
      xlab = "Input x", ylab = "Logistic response L(x)")
axis(side = 1, at = c(0, 5, 10))
points(x = 5, y = 0.5, pch = 19)
abline(h = 1, lty = 2)
par(old_par)
```

The logistic rises towards 1 as $x\to\infty$ but never equals
1 at a finite $x$. It therefore has an upper asymptote,
not a finite maximum on the whole real line.

# Exercise 5: Half-saturation and parameterisation

Dividing numerator and denominator by $\alpha H$ gives

$$
\frac{\alpha N}{1+\alpha HN}
=\frac{(1/H)N}{1/(\alpha H)+N}.
$$

Thus $a=1/H$ and $b=1/(\alpha H)$. Solving these equations
in the other direction gives $H=1/a$ and $\alpha=a/b$.
This algebra shows that the two parameterisations produce
the same response at every valid $N$.

For $\alpha=0.6$ square metres per hour and $H=0.25$
hours per prey,

$$
a=\frac1{0.25}=4\ \text{prey per hour},
\qquad
b=\frac1{0.6(0.25)}
=\frac{20}{3}\approx6.67\ \text{prey per square metre}.
$$

The asymptote is 4 prey per hour for one predator. The initial
slope is $a/b=\alpha=0.6$ square metres per hour. The
half-saturation density is $b\approx6.67$ prey per square
metre, where the predicted rate is 2 prey per hour.

For the second curve, $b=a/(a/b)=9/1.5=6$ prey per square
metre. Therefore $\alpha=a/b=1.5$ square metres per hour
and $H=1/a=1/9\approx0.111$ hours per prey.

```{r}
form_ab = function(N, a, b) {
  a * N / (b + N)
}
form_alpha_h = function(N, alpha, H) {
  alpha * N / (1 + alpha * H * N)
}
curve(
  form_ab(N = x, a = 4, b = 20 / 3),
  from = 0, to = 30,
  xlab = "Prey density (prey per square metre)",
  ylab = "Prey eaten by one predator per hour",
  lty = 1
)
curve(
  form_alpha_h(N = x, alpha = 0.6, H = 0.25),
  from = 0, to = 30, add = TRUE, lty = 2
)
legend(
  "bottomright", legend = c("a, b form", "alpha, H form"),
  lty = c(1, 2), bty = "n"
)
```

The two curves overlap. A graph checks selected plotted values;
the algebra establishes their equality throughout the domain.

# Exercise 6: Shift and scale

Substitute $z=(N-3)/4$ into $g(z)=z/(1+z)$ and multiply
top and bottom by 4:

$$
F(N)
=2+10\frac{(N-3)/4}{1+(N-3)/4}
=2+\frac{10(N-3)}{4+N-3}
=2+\frac{10(N-3)}{N+1}.
$$

Relative to $g$, the new curve shifts right by 3 input units,
stretches horizontally by a factor of 4, scales vertically
by 10, and shifts up by 2 output units. Its stated domain is
$N\geq3$.

At the start, $F(3)=2$. As $N\to\infty$, the fraction tends
to 10, so $F(N)\to12$. Halfway between 2 and 12 is 7:

$$
2+\frac{10(N-3)}{N+1}=7
\quad\Longrightarrow\quad
2(N-3)=N+1
\quad\Longrightarrow\quad N=7.
$$

The quotient rule gives

$$
F'(N)
=10\frac{(N+1)-(N-3)}{(N+1)^2}
=\frac{40}{(N+1)^2},
\qquad F'(3)=\frac{40}{16}=2.5.
$$

The input $z=(N-3)/4$ was defined for $z\geq0$, so the
transformed ecological curve is intended for $N\geq3$.
Evaluating the algebraic expression at lower densities would
extend the model beyond that domain; it even has a pole at
$N=-1$.

```{r}
base_curve = function(z) z / (1 + z)
shifted_curve = function(N) {
  2 + 10 * (N - 3) / (4 + N - 3)
}
old_par = par(mfrow = c(1, 2))
curve(base_curve(x), from = 0, to = 6,
      ylim = c(0, 1), xlab = "Input z", ylab = "Base response g(z)")
abline(h = 1, lty = 2)
curve(shifted_curve(x), from = 3, to = 27,
      ylim = c(0, 12), xlab = "Input N",
      ylab = "Shifted response F(N)")
points(x = c(3, 7), y = c(2, 7), pch = 19)
abline(h = 12, lty = 2)
par(old_par)
```

# Exercise 7: Transformations and a local approximation

Because $D(t)>0$, we can take logs:

$$
\log D(t)=\log(30e^{-0.2t})
=\log30-0.2t.
$$

The line has intercept $\log30\approx3.401$ and slope
$-0.2$ per unit time. The logarithm is defined only for
positive $D$; zero and negative observations cannot be
put directly on this log scale.

For $Y(x)=2x^{1.5}$ with $x>0$,

$$
\log Y=\log2+1.5\log x.
$$

The log-log slope is 1.5. Doubling $x$ multiplies $Y$
by $2^{1.5}\approx2.83$, regardless of the starting
positive $x$.

For $F(N)=12N/(6+N)$, the derivatives from the theory give

$$
F(0)=0,\qquad
F'(N)=\frac{72}{(6+N)^2},\qquad
F''(N)=-\frac{144}{(6+N)^3}.
$$

Therefore $F'(0)=2$ and $F''(0)=-2/3$. The second-order
Taylor approximation around zero is

$$
T_2(N)
=F(0)+F'(0)N+\frac{F''(0)}2N^2
=2N-\frac13N^2.
$$

At $N=1$, the exact value is $12/7\approx1.714$ and
the approximation is $5/3\approx1.667$. At $N=10$,
the exact value is $120/16=7.5$, while the approximation
is $20-100/3=-40/3\approx-13.33$. The approximation is
useful near zero and becomes ecologically implausible
farther away.

```{r}
N = c(1, 10)
exact = 12 * N / (6 + N)
approximation = 2 * N - N^2 / 3
data.frame(N, exact, approximation)
```

A straight line on a transformed scale is an identity
for the proposed deterministic function. A Taylor
polynomial is a local description of its shape. Neither
specifies the distribution of observations around the
function, so neither supplies a likelihood.

# Exercise 8: Choose and explore curve shapes

The Ricker form $R(N)=rNe^{-qN}$ matches the stated
pattern: it is zero at $N=0$, rises, has an interior peak,
and tends back towards zero at high $N$. A straight line
does not have an interior peak, while a Holling type II
curve keeps rising towards a positive asymptote.

For a Ricker curve the peak is at $N=1/q$ and has height
$r/(qe)$. A peak at 5 adults per hectare gives
$q=1/5=0.2$ hectares per adult. Setting the height to
10 recruits per hectare gives

$$
\frac{r}{qe}=10
\quad\Longrightarrow\quad
r=10qe=2e\approx5.44
\text{ recruits per adult}.
$$

These units also make $qN$ dimensionless and
$rN$ a recruit density. A mathematical match to the
stated pattern does not establish the process responsible
for the decline.

For $M(x)=12x/(6+x)$ and $E(x)=12(1-e^{-x/6})$,
both values at zero are zero and both limits at large
$x$ are 12. Their derivatives are

$$
M'(x)=\frac{72}{(6+x)^2},
\qquad
E'(x)=2e^{-x/6},
$$

so both initial slopes are 2. At $x=6$,

$$
M(6)=6,\qquad
E(6)=12(1-e^{-1})\approx7.59.
$$

The exponential form approaches the shared asymptote
more rapidly over this part of the input range, even
though the two curves share the stated landmarks.

```{r}
q = 1 / 5
r = 2 * exp(1)
old_par = par(mfrow = c(1, 2))
curve(r * x * exp(-q * x), from = 0, to = 30,
      xlab = "Adult density (adults per hectare)",
      ylab = "Recruit density (recruits per hectare)")
points(x = 5, y = 10, pch = 19)
curve(12 * x / (6 + x), from = 0, to = 30,
      ylim = c(0, 12),
      xlab = "Input x", ylab = "Saturating response",
      lty = 1)
curve(12 * (1 - exp(-x / 6)), from = 0, to = 30,
      add = TRUE, lty = 2)
legend("bottomright",
       legend = c("Rational", "Exponential"),
       lty = c(1, 2), bty = "n")
par(old_par)
```

The contour calculation evaluates the same response for
combinations of prey density and handling time:

```{r}
response = function(N, H) {
  0.6 * N / (1 + 0.6 * H * N)
}
density_values = seq(from = 0, to = 30, length.out = 60)
handling_values = seq(from = 0.15, to = 0.5, length.out = 60)
response_grid = outer(
  X = density_values, Y = handling_values,
  FUN = response
)
contour(
  x = density_values, y = handling_values, z = response_grid,
  xlab = "Prey density (prey per square metre)",
  ylab = "Handling time (hours per prey)"
)
```

At any fixed positive prey density, longer handling time
reduces the predicted consumption rate. For $N=0$ the
rate is zero whatever $H$ is. To fit any of these
deterministic curves to observed counts, we would also
need a probability model describing the observations
conditional on the predictor and curve parameters.
