---
title: "Chapter 5: Deterministic functions in ecology"
date: today
---

Exercise 1 analyses one function from end to end, with data on the amount of virus in the
blood of infected rabbits. Exercises 2 to 5 are drills without data, one for each skill of
the chapter: limits, derivatives, special points and parameters. Exercises 6 and 7 are
cases with real data: the rate of an enzymatic reaction (the data set `Puromycin`, which
comes with R), the fecundity of fir trees and the recruitment of a reef fish. The data of
Exercises 1 and 7 are in the package `emdbook`, which you need to install once with
`install.packages("emdbook")`. As in the theory, the curves are placed over the data by
eye and nothing is fitted, which is the topic of [Chapter 6](../../Chapter_6/Theory.qmd).

Before you start, read the entries of the
[function supplement](../../Supplements/deterministic-functions.qmd) that the exercises
name. If derivatives are no longer familiar, work through the
[calculus refresher](../../Supplements/calculus.qmd) first. Questions that start with "On
paper" and the drills of Exercises 2 to 5 are meant to be done with pen and paper before
you open R (a calculator is fine for logarithms and exponentials). Work through the first
exercise before trying the remaining ones. The ideas and notation are introduced in the
[Chapter 5 theory](../Theory.qmd).

# Exercise 1: Virus titre in infected rabbits

The data set `MyxoTiter_sum` in the package `emdbook` contains the titre of the myxoma
virus in the blood of European rabbits on different days after infection, for viruses of
four grades (Fenner and McIntyre 1956; the values were read from a figure in Dwyer et al.
1990). The titre measures the concentration of virus and is expressed as the logarithm in
base 10 of the number of rabbit infectious doses. We keep the 27 observations of grade 1,
which were taken between 2 and 9 days after infection:

```{r}
# Load the myxomatosis data from the emdbook package without attaching the package.
data(MyxoTiter_sum, package = "emdbook")

# Keep the observations of virus grade 1: their number and the days that they cover.
grade_1 = MyxoTiter_sum[MyxoTiter_sum$grade == 1, ]
nrow(grade_1)
range(grade_1$day)
```

A virus multiplies in its host until the immune response of the host catches up with it,
so the titre rises from a low value, reaches a maximum and then declines. We describe this
shape with the [Ricker](../../Supplements/deterministic-functions.qmd#sec-fn-ricker)
function,

$$
f(x)=a\,x\,e^{-bx},
$$

where $x$ is the time since infection in days, $f(x)$ is the titre, and $a$ and $b$ are
positive parameters. In this exercise we find out everything that the formula can tell us
about the curve, and then we use it to place the curve over the data by eye.

1. On paper, find the units of $a$ and $b$, the value of the function at $x=0$ and its
   limit as $x\to\infty$.
2. On paper, derive $f'(x)$. What is the slope of the function at $x=0$ and as
   $x\to\infty$? Find the location and the height of the peak.
3. On paper, derive $f''(x)$ and find the inflection point. Where is the function concave
   and where is it convex?
4. Plot the titre against the day. Read the location and the height of the peak from the
   plot and calculate $a$ and $b$ from them, with their units.
5. Write the function in R and draw it over the data, marking the peak and the inflection
   point. Check the peak with `which.max()` and the derivative at day 2 with `D()`.

## Solution

### Question 1

The units follow from the two rules of the
[theory](../Theory.qmd#sec-deterministic-functions). The argument of an exponential has
no units, so in $e^{-bx}$ the parameter $b$ must have the inverse units of $x$: per day.
The factor $e^{-bx}$ then has no units either, so the product $ax$ must have the units of
$f(x)$. As $x$ is in days, $a$ is in units of titre per day.

At $x=0$ the function is $f(0)=a\times0\times e^{0}=0$. As $x\to\infty$ the factor $x$
grows without bound while $e^{-bx}$ goes to zero, so we have to find out which of the two
wins. An exponential dominates any power of $x$ (see
[Limits and asymptotes](../Theory.qmd#sec-function-limits)), so

$$
\lim_{x\to\infty}a\,x\,e^{-bx}=\lim_{x\to\infty}\frac{a\,x}{e^{bx}}=0.
$$

The same result follows from L'Hôpital's rule: the derivative of the numerator is $a$ and
that of the denominator is $b\,e^{bx}$, and $a/(b\,e^{bx})\to0$. So the function starts at
zero and returns to zero, which is what a hump-shaped function has to do.

### Question 2

The function is the product of $g(x)=ax$ and $h(x)=e^{-bx}$, with $g'(x)=a$ and, by the
rule for the exponential, $h'(x)=-b\,e^{-bx}$. The product rule gives

$$
\begin{aligned}
f'(x) &= g'(x)\,h(x)+g(x)\,h'(x)\\
&= a\,e^{-bx}-ab\,x\,e^{-bx}\\
&= a\,(1-bx)\,e^{-bx}.
\end{aligned}
$$

At $x=0$ the slope is $f'(0)=a$. So $a$ is the initial slope of the function, which agrees
with its units (titre per day). As $x\to\infty$ the slope goes to zero, by the same
argument as in question 1, because the exponential dominates the factor $1-bx$.

At the peak the slope is zero. Since $a$ and $e^{-bx}$ are positive, $f'(x)=0$ only where
$1-bx=0$:

$$
x_{\max}=\frac{1}{b}, \qquad
f(x_{\max})=a\,\frac{1}{b}\,e^{-1}=\frac{a}{b\,e}.
$$

The slope is positive for $x<1/b$ and negative for $x>1/b$, so this critical point is a
maximum. The location of the peak, $1/b$, is in days, and its height is in units of titre,
as it should be.

### Question 3

We differentiate $f'(x)=a(1-bx)e^{-bx}$, again with the product rule. The derivative of
$a(1-bx)$ is $-ab$ and that of $e^{-bx}$ is $-b\,e^{-bx}$:

$$
\begin{aligned}
f''(x) &= -ab\,e^{-bx}-ab\,(1-bx)\,e^{-bx}\\
&= -ab\,(2-bx)\,e^{-bx}.
\end{aligned}
$$

The second derivative is zero where $2-bx=0$, that is, at $x=2/b$, which is twice the
location of the peak. For $x<2/b$ it is negative, so the function is concave (it bends
downwards), and for $x>2/b$ it is positive, so the function is convex (it bends upwards).
The sign changes at $x=2/b$, which makes it an inflection point. The value of the function
and its slope at that point are

$$
f\!\left(\frac{2}{b}\right)=\frac{2a}{b\,e^{2}}, \qquad
f'\!\left(\frac{2}{b}\right)=a\,(1-2)\,e^{-2}=-\frac{a}{e^{2}}.
$$

This is the steepest decline of the function. Note that the value at the inflection point
is $2/e\approx0.74$ times the height of the peak.

### Question 4

We plot the 27 observations (@fig-titre_data):

```{r}
#| label: fig-titre_data
#| fig-cap: "Titre of the myxoma virus of grade 1 in the blood of rabbits against the time since infection."
library(ggplot2)

# Titre against the day of the observation.
ggplot(grade_1, aes(x = day, y = titer)) +
  geom_point(colour = "#0072B2", alpha = 0.7, size = 2) +
  labs(x = "Time since infection (days)", y = "Virus titre (log10 infectious doses)") +
  theme_classic()
```

The titre increases from about 5.5 on day 2 and levels off between 7 and 8 from day 6
onwards. The observations do not show a clear decline, so the location of the peak can only
be read roughly. We take a peak on day 6 with a titre of 7.6. With the two results of
question 2 we can go from the peak to the parameters. From $x_{\max}=1/b$,

$$
b=\frac{1}{x_{\max}}=\frac{1}{6}\approx0.167 \text{ per day},
$$

and from $f(x_{\max})=a/(be)$,

$$
a=f(x_{\max})\,b\,e=\frac{7.6\times2.718}{6}\approx3.44 \text{ units of titre per day}.
$$

That is, a curve with this peak starts with a slope of 3.44 units of titre per day. Its
inflection point is at $2/b=12$ days, where the titre is $2\times7.6/e\approx5.59$ and the
slope is $-3.44/e^2\approx-0.47$ units of titre per day. The observations end on day 9, so
the inflection point lies outside the data.

### Question 5

We write the function with the input first and the parameters after it, and draw it with
the values of question 4 (@fig-titre_curve). The dashed lines mark the peak and the
triangle marks the inflection point:

```{r}
#| label: fig-titre_curve
#| fig-cap: "The same observations with a Ricker function whose parameters were calculated from a peak read by eye (a = 3.44, b = 1/6). Dashed lines mark the peak and the triangle marks the inflection point."
# Ricker function: initial slope a and rate b.
ricker = function(x, a = 1, b = 1) {
  a * x * exp(-b * x)
}

# Parameters calculated from a peak on day 6 with a titre of 7.6.
b_titre = 1 / 6
a_titre = 7.6 * exp(1) / 6

# The function on a sequence of days, for the curve.
titre_curve = data.frame(day = seq(0, 14, length.out = 200))
titre_curve$titer = ricker(titre_curve$day, a = a_titre, b = b_titre)

# The inflection point: twice the location of the peak.
inflection = data.frame(day = 2 / b_titre,
                        titer = ricker(2 / b_titre, a = a_titre, b = b_titre))

ggplot(grade_1, aes(x = day, y = titer)) +
  geom_vline(xintercept = 1 / b_titre, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 7.6, linetype = "dashed", colour = "grey40") +
  geom_point(colour = "#0072B2", alpha = 0.7, size = 2) +
  geom_line(data = titre_curve, colour = "#D55E00", linewidth = 0.8) +
  geom_point(data = inflection, shape = 17, size = 3, colour = "#D55E00") +
  labs(x = "Time since infection (days)", y = "Virus titre (log10 infectious doses)") +
  theme_classic()
```

We can verify our math with R:

```{r}
# Peak of the function on a fine grid of days: location and height.
day_grid = seq(0, 14, by = 0.001)
titre_values = ricker(day_grid, a = a_titre, b = b_titre)
day_grid[which.max(titre_values)]
max(titre_values)

# Derivative with D() and with the formula of question 2, on day 2.
ricker_derivative = D(expression(a * x * exp(-b * x)), "x")
ricker_derivative
eval(ricker_derivative, list(a = a_titre, b = b_titre, x = 2))
a_titre * (1 - b_titre * 2) * exp(-b_titre * 2)

# Value of the function on days 2, 9 and 12 (the inflection point).
ricker(c(2, 9, 12), a = a_titre, b = b_titre)
```

The highest value on the grid is 7.6 on day 6, as intended. `D()` does not simplify its
result, but it returns the same slope as our formula: on day 2 the titre increases by 1.64
units per day, about half of the initial slope. The value on day 12 is 5.59, as calculated
on paper.

The curve passes through the middle of the observations between days 3 and 7. It gives a
titre of 4.9 on day 2, a little below most of the observations of that day, and of 6.9 on
day 9, where the observed titre is 7.7. The reason is the one that we met in the
theory for the size of the tadpoles: in the Ricker function the parameter $b$ sets both the
location of the peak and the rate of the decline after it, so with a peak on day 6 the
curve has to decline at a given rate, whether the observations do or not. A function with a
third parameter, such as the
[generalised Ricker](../../Supplements/deterministic-functions.qmd#sec-fn-generalised-ricker)
function, could separate the two. Whether the extra parameter is supported by the data is
a question for [Chapter 6](../../Chapter_6/Theory.qmd), where the curves are fitted.

# Exercise 2: Limits and end behaviour

The first thing to find out about a function is where it starts and where it ends. For
each of the five functions below:

- on paper, calculate the value of the function at the lower end of its domain and its
  limit as $x\to\infty$, with their units;
- on paper, find the simpler function that it resembles close to the lower end of its
  domain;
- sketch the function by hand, and only then draw it with `curve()` to check your sketch.

The first three questions are the core of this exercise and the last two are extra
practice. The entries of the supplement that you need are the
[hyperbolic](../../Supplements/deterministic-functions.qmd#sec-fn-hyperbolic),
[saturating exponential](../../Supplements/deterministic-functions.qmd#sec-fn-monomolecular),
[Holling type III](../../Supplements/deterministic-functions.qmd#sec-fn-holling3) and
[Michaelis–Menten](../../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten)
functions.

1. The mean mass of a seedling (mg) as a function of the number of seedlings in a pot,
   $x$: $f(x)=\dfrac{480}{12+x}$.
2. The amount of nitrogen released from a crop residue (kg per ha) as a function of the
   time since it was ploughed in, $x$ (weeks): $f(x)=90\left(1-e^{-0.2x}\right)$.
3. The number of seeds removed by rodents in one night as a function of the density of
   seeds, $x$ (seeds per m²): $f(x)=\dfrac{30x^2}{100+x^2}$.
4. The number of eggs in the clutch of an insect as a function of the body mass of the
   female, $x$ (mg), for $x\ge5$: $f(x)=2+\dfrac{10\,(x-5)}{15+(x-5)}$.
5. A function in which a power and an exponential compete, for $x\ge0$:
   $f(x)=0.5\,x^2e^{-0.4x}$. Compare its behaviour close to zero with that of a Ricker
   function.

# Exercise 3: Derivatives and slopes

A derivative is the slope of a function, and it has the units of $y$ divided by the units
of $x$. Do all the questions on paper, name the rule that you use at each step (the rules
are in the [calculus refresher](../../Supplements/calculus.qmd#sec-calc-rules)), and give
the units of every slope. The first three questions are the core of this exercise and the
last two are extra practice. Questions 2 and 3 continue with functions of Exercise 2.

1. The body mass of a fish (g) as a function of its length $x$ (cm) follows the
   [power law](../../Supplements/deterministic-functions.qmd#sec-fn-power) $f(x)=0.01\,x^3$.
   Derive $f'(x)$ and calculate the mass and the slope for a fish of 20 cm. What are the
   units of the parameter 0.01?
2. For the saturating exponential function $f(x)=a\left(1-e^{-bx}\right)$, derive $f'(x)$
   and show that it can be written as $f'(x)=b\,(a-f(x))$. What does this say about how the
   function approaches its asymptote? For the nitrogen released from a crop residue
   ($a=90$ kg per ha, $b=0.2$ per week), calculate the slope at the start and at the time
   when half of the nitrogen has been released.
3. For the Holling type III function $f(x)=\dfrac{ax^2}{b^2+x^2}$, derive $f'(x)$. What is
   the slope at $x=0$? For the seeds removed by rodents ($a=30$ seeds per night, $b=10$
   seeds per m²), calculate the slope at $x=10$ seeds per m².
4. The shell length of a snail (mm) as a function of its age $x$ (months) follows the
   [Gompertz](../../Supplements/deterministic-functions.qmd#sec-fn-gompertz) function
   $f(x)=a\exp\left(-e^{-b(x-c)}\right)$ with $a=24$ mm, $b=0.3$ per month and $c=4$
   months. Derive $f'(x)$ and write it in terms of $f(x)$. Calculate the length and the
   growth rate of the snail at 4 months.
5. Derive the partial derivatives of the Ricker function $f(x;a,b)=a\,x\,e^{-bx}$ with
   respect to $a$ and with respect to $b$, and calculate both for $a=2$, $b=0.5$ and
   $x=2$. At which value of $x$ does a small change in $b$ move the curve most?

When you have finished, check your answers to questions 3 and 5 with `D()`.

# Exercise 4: Special points

The special points of a function (a peak, an inflection point, a half-maximum) are what we
can read from a plot, so we need them as expressions of the parameters. Do all the
questions on paper and give the units of every result. The first three questions are the
core of this exercise and the last two are extra practice.

1. For the saturating exponential function $f(x)=a\left(1-e^{-bx}\right)$, derive the
   half-maximum $x_{1/2}$. Then show that the function reaches three quarters of its
   asymptote at $2x_{1/2}$. Calculate both times for the nitrogen released from a crop
   residue in Exercise 2 ($b=0.2$ per week).
2. The concentration of a pesticide in a soil falls from 8 to 2 mg per kg in 30 days.
   Assuming an [exponential](../../Supplements/deterministic-functions.qmd#sec-fn-exponential)
   decay, $f(x)=a\,e^{-bx}$, find its half-life, the rate $b$ and the characteristic scale
   $1/b$, with their units. What is the concentration after one characteristic scale?
3. The number of adult parasitoids that emerge from a caterpillar as a function of the
   number of eggs laid in it, $x$, follows the
   [Hassell](../../Supplements/deterministic-functions.qmd#sec-fn-hassell) function with
   an exponent of 2, $f(x)=\dfrac{ax}{(b+x)^2}$. Derive $f'(x)$ and find the location and
   the height of the peak as expressions of $a$ and $b$. Calculate them for $a=1280$ and
   $b=40$ eggs. What is the initial slope?
4. For the Holling type III function $f(x)=\dfrac{ax^2}{b^2+x^2}$, start from the
   derivative that you found in Exercise 3 and derive $f''(x)$. Find the inflection point,
   and the value and the slope of the function there. Calculate them for the seeds removed
   by rodents ($a=30$ seeds per night, $b=10$ seeds per m²).
5. The intake of a herbivore (g per minute) as a function of the biomass of the
   vegetation, $x$ (g per m²), follows the
   [Holling type IV](../../Supplements/deterministic-functions.qmd#sec-fn-holling4)
   function $f(x)=\dfrac{ax^2}{b+cx+x^2}$ with $a=12$, $b=100$ and $c=-10$. Find the
   location and the height of the peak, using the fact that a quotient is zero where its
   numerator is zero. Compare the height of the peak with the limit of the function as
   $x\to\infty$.

When you have finished, check the peaks of questions 3 and 5 with `which.max()` on a fine
grid of values of $x$.

# Exercise 5: Parameters

The same curve can be written with different parameters, and a curve can be moved,
stretched and bounded by changing its formula (see
[Parameters](../Theory.qmd#sec-function-parameters) and
[Building new functions](../Theory.qmd#sec-function-building) in the theory). Do all the
questions on paper and give the units of every parameter. The first three questions are
the core of this exercise and the last two are extra practice.

1. A wolf spider searches an area of $\alpha=2$ m² per day for prey and needs $H=0.05$
   days to handle each prey. Write its functional response in the Holling type II form,
   $f(x)=\dfrac{\alpha x}{1+\alpha Hx}$, where $x$ is the density of prey (prey per m²),
   and convert it to the Michaelis–Menten form, $f(x)=\dfrac{ax}{b+x}$. What are $a$ and
   $b$, and what do they mean? A second species has $a=50$ prey per day and $b=25$ prey
   per m². What are its $\alpha$ and $H$?
2. Rewrite the Ricker function $f(x)=a\,x\,e^{-bx}$ so that its two parameters are the
   location of the peak, $x_{\max}$, and the height of the peak, $y_{\max}$. What are
   $x_{\max}$ and $y_{\max}$ for $f(x)=8\,x\,e^{-0.25x}$?
3. A plant does not flower until it has a mass of 2 g, at which it carries 3 flowers. The
   number of flowers then increases with the mass of the plant, $x$ (g), and approaches 27
   for the heaviest plants. It is halfway between 3 and 27 at a mass of 10 g. Write the
   formula of a shifted and scaled Michaelis–Menten function with these properties, for
   $x\ge2$.
4. The probability that a plant species occurs at a site as a function of the pH of the
   soil, $x$, is described with the composition
   $f(x)=\operatorname{logit}^{-1}\left[d-b\,(x-c)^2\right]$ with $d=2$, $b=0.5$ and $c=6$.
   What is the range of the function? Where is its peak and how high is it? At which
   values of the pH is the probability 0.5?
5. Show that the [Shepherd](../../Supplements/deterministic-functions.qmd#sec-fn-shepherd)
   function $f(x)=\dfrac{ax}{b+x^c}$ contains the Michaelis–Menten function, and that the
   [Richards](../../Supplements/deterministic-functions.qmd#sec-fn-richards) function
   $f(x)=\dfrac{a}{\left[1+d\,e^{-b(x-c)}\right]^{1/d}}$ contains the logistic function.
   Which function does the Richards function become for $d=-1$?

When you have finished, check your answers to questions 1 and 2 in R by evaluating both
forms of each function at a few values of $x$.

# Exercise 6: Enzyme kinetics

The data set `Puromycin` of base R contains the rate of an enzymatic reaction (`rate`, in
counts/min/min) at six concentrations of the substrate (`conc`, in parts per million,
ppm), measured in cells that were treated with the antibiotic puromycin (12 observations)
and in untreated cells (11 observations) (Treloar 1974, in Bates and Watts 1988). The
column `state` says to which of the two groups an observation belongs. Load the data with
`data(Puromycin, package = "datasets")`.

This is the kind of experiment for which Michaelis and Menten proposed their function,
$f(x)=\dfrac{ax}{b+x}$, where $x$ is the concentration of the substrate and $f(x)$ is the
rate of the reaction (see the
[Michaelis–Menten](../../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten)
entry of the supplement). In this exercise you place the curve by eye and compare it
with an alternative function.

1. Plot the rate against the concentration, with a different symbol for each state. Read
   $a$ and $b$ from the plot for each state, with their units, and draw the two curves
   over the data. Which of the two parameters differs between the states?
2. On paper, calculate the initial slope of the curve of the treated cells, with its units.
   The saturating exponential function $g(x)=a\left(1-e^{-b_gx}\right)$ has the same shape
   as the Michaelis–Menten function. Find the rate $b_g$ for which it has the same
   asymptote and the same initial slope as the curve of the treated cells. Which fraction
   of the asymptote does each function reach at $x=b$? In R, draw both curves and find the
   concentration at which they differ most.

# Exercise 7: Choosing functions for fecundity and for recruitment

This exercise uses two data sets of the package `emdbook`, with a different shape each.

**Part A.** The data set `FirDBHFec_sum` contains the diameter at breast height (`DBH`, in
cm) and the fecundity (`fecundity`) of 333 balsam fir trees (*Abies balsamea*) on Whiteface
Mountain, New York (Silvertown and Dodd 1999; Dodd and Silvertown 2000). Fecundity was
measured as the number of cone rachises on a tree: the rachis is the axis of a cone, which
stays on the branch after the cone has fallen apart. Load the data with
`data(FirDBHFec_sum, package = "emdbook")`.

1. The diameter of some trees is missing. Count those trees and remove them explicitly.
   Plot fecundity against diameter and describe the shape of the relationship.
2. On paper, explain why a [power law](../../Supplements/deterministic-functions.qmd#sec-fn-power),
   $f(x)=ax^b$, is a candidate for these data. Read two points from the curve that you
   would draw through the middle of the cloud and calculate $b$ and $a$ from them. What
   are the units of $a$? Draw the curve over the data.
3. On paper, calculate by which factor the fecundity changes when the diameter doubles.
   Which measure of the size of a tree is proportional to the square of its diameter?

**Part B.** The data set `DamselRecruitment_sum` comes from a field experiment with the
three-spot damselfish *Dascyllus trimaculatus*, whose larvae settle on sea anemones
(Schmitt, Holbrook and Osenberg 1999). Each of the 11 rows is a target density of settlers:
`settler.den` is the density of settlers and `surv.den` is the mean density of survivors
after six months, both as the number of fish per 0.1 m² of anemone. Load the data with
`data(DamselRecruitment_sum, package = "emdbook")`.

4. Plot the density of survivors against the density of settlers. Read starting values
   for a Michaelis–Menten function from the plot, with their units, and draw the curve. Compare its initial
   slope with the fraction of the settlers that survived at the lowest densities.
5. Write a [hockey stick](../../Supplements/deterministic-functions.qmd#sec-fn-hockey-stick)
   function with `ifelse()`, choose its two parameters by eye and add it to the plot.
   Where does it describe the data differently from the Michaelis–Menten function?
6. The [Shepherd](../../Supplements/deterministic-functions.qmd#sec-fn-shepherd) function
   can be written as $f(x)=\dfrac{sx}{1+(x/k)^c}$, where $s$ is the initial slope, $k$ is a
   scale parameter and $c$ is an exponent. Write it in R, give $s$ and $k$ the values that
   correspond to your Michaelis–Menten function, and draw it for $c=0.8$, $c=1$ and
   $c=1.2$. What does $c$ change, and which part of the data can tell us about it?
7. Which of the three functions are nested? What would you need in order to choose among
   them?
