---
title: "Chapter 5: Maximum likelihood — solutions"
date: today
---

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions use the
same models, tasks, and seeds as Exercises 2–4 on that page.

# Exercise 2: Tree heights and two parameters

For a normal model, the density contribution from height $h$ is

$$
f(h\mid\mu,\sigma)
=\frac{1}{\sigma\sqrt{2\pi}}
  \exp\left[-\frac{(h-\mu)^2}{2\sigma^2}\right].
$$

For heights 21 m and 23 m, each contribution under candidate A
$(\mu,\sigma)=(22,1)$ is
$e^{-1/2}/\sqrt{2\pi}$. Under candidate B $(22,3)$, each is
$e^{-1/18}/(3\sqrt{2\pi})$. Multiply within each pair, then
take logs:

$$
\begin{aligned}
L_A&=\frac{e^{-1}}{2\pi},
&\ell_A&=-\log(2\pi)-1\approx-2.838,\\
L_B&=\frac{e^{-1/9}}{18\pi},
&\ell_B&=-\log(18\pi)-\frac19\approx-4.146.
\end{aligned}
$$

Candidate A has the greater likelihood for these **fixed**
measurements. Each factor is a density, not the probability of
observing an exact continuous height. Probabilities under this
model apply to height intervals.

For the larger simulated sample, define $\sigma=\exp(\log\sigma)$
inside the negative log-likelihood. This keeps the standard
deviation positive while `optim()` varies the fitted parameters.

```{r}
set.seed(seed = 522)
heights = rnorm(n = 30, mean = 22, sd = 4)

normal_nll = function(par, heights) {
  mu = par[1]
  sigma = exp(par[2])
  -sum(dnorm(x = heights, mean = mu, sd = sigma, log = TRUE))
}

normal_fit = optim(
  par = c(mu = mean(heights), log_sigma = log(sd(heights))),
  fn = normal_nll, heights = heights, method = "BFGS"
)
normal_fit$convergence
estimates = c(
  mu_m = unname(normal_fit$par[1]),
  sigma_m = exp(unname(normal_fit$par[2]))
)
estimates
```

The fitted mean is approximately 22.97 m and the fitted standard
deviation is approximately 4.08 m. They need not equal 22 m and
4 m: those values generated one random sample, whose realised
mean and spread can differ.

Evaluate the same log-likelihood across pairs of parameter
values. Subtracting the largest grid value makes the plotted
log-likelihood relative to its grid maximum.

```{r}
mu_grid = seq(from = 17, to = 27, length.out = 80)
sigma_grid = seq(from = 1.5, to = 7, length.out = 80)
loglik_surface = outer(
  X = mu_grid, Y = sigma_grid,
  FUN = Vectorize(function(mu, sigma) {
    sum(dnorm(
      x = heights, mean = mu, sd = sigma, log = TRUE
    ))
  })
)
contour(
  x = mu_grid, y = sigma_grid,
  z = loglik_surface - max(loglik_surface),
  levels = c(-10, -5, -2, -1, -0.5, -0.1),
  xlab = "Mean tree height (m)",
  ylab = "Tree-height SD (m)"
)
points(x = estimates["mu_m"], y = estimates["sigma_m"], pch = 19)
points(x = 22, y = 4, pch = 4, lwd = 2)
legend(
  "topright", legend = c("MLE", "Generating values"),
  pch = c(19, 4), bty = "n"
)
```

A slice for $\mu$ evaluates $\ell(\mu,\hat\sigma)$ while keeping
$\sigma$ fixed at its joint estimate. A profile for $\mu$
re-estimates $\sigma$ at **each** candidate $\mu$. Both curves
describe fit to these data; neither is a probability distribution
for $\mu$.

# Exercise 3: Overdispersed seedling counts

For $k=1$, the supplied probability mass gives
$P(X=0\mid\mu,1)=1/(1+\mu)$ and
$P(X=1\mid\mu,1)=\mu/(1+\mu)^2$. Assuming independent quadrats,

$$
L(\mu\mid0,1;k=1)
=\frac{\mu}{(1+\mu)^3},
\qquad
\ell(\mu)=\log\mu-3\log(1+\mu).
$$

At $\mu=1$, the likelihood is $1/8=0.125$ and the
log-likelihood is $-\log8\approx-2.079$. At $\mu=2$, they
are $2/27\approx0.07407$ and
$\log2-3\log3\approx-2.603$. Thus $\mu=1$ gives these
two counts greater likelihood among the proposed means.

The model used to simulate the larger sample has mean 4 seedlings
per quadrat and variance
$4+4^2/1.5\approx14.67$ squared seedlings per quadrat.

```{r}
set.seed(seed = 533)
counts = rnbinom(n = 80, size = 1.5, mu = 4)
c(
  model_mean = 4,
  model_variance = 4 + 4^2 / 1.5,
  sample_mean = mean(counts),
  sample_variance = var(counts)
)
```

With this seed, the sample mean is 4.65 and its sample variance
is approximately 13.45. The sample summaries vary between
simulations. A variance larger than the mean is exploratory
evidence of overdispersion
relative to a common-mean Poisson model; it cannot establish
whether habitat differences, clustering, dependence, or another
process caused it.

Both $\mu$ and $k$ must be positive. Their logarithms are
unconstrained optimisation parameters. The starting mean uses
the sample mean; a starting $k$ comes from rearranging the
negative binomial mean–variance relation. If the sample variance
is too close to or below its mean, the small positive floor in
the denominator still supplies a valid starting value. It does
not change the fitted likelihood.

```{r}
nb_nll = function(par, counts) {
  mu = exp(par[1])
  k = exp(par[2])
  -sum(dnbinom(x = counts, size = k, mu = mu, log = TRUE))
}

start_mu = max(mean(counts), 0.1)
start_k = start_mu^2 / max(var(counts) - start_mu, 0.1)
nb_fit = optim(
  par = log(c(mu = start_mu, k = start_k)),
  fn = nb_nll, counts = counts, method = "BFGS"
)
nb_fit$convergence
nb_estimates = exp(nb_fit$par)
nb_estimates
```

The fit returns a convergence code of zero, an estimated mean
of approximately 4.65 seedlings per quadrat, and an estimated
shape of approximately 2.64. These positive estimates need
not reproduce the known generating values exactly. We also
need to check whether the fitted probabilities reasonably
describe the observed counts.

```{r}
count_values = 0:max(
  counts,
  qnbinom(
    p = 0.995, size = nb_estimates["k"], mu = nb_estimates["mu"]
  )
)
observed_frequency = as.numeric(
  table(factor(counts, levels = count_values))
) / length(counts)
fitted_probability = dnbinom(
  x = count_values,
  size = nb_estimates["k"], mu = nb_estimates["mu"]
)

plot(
  x = count_values, y = observed_frequency,
  type = "h", lwd = 5, col = "grey60",
  xlab = "Seedlings per quadrat",
  ylab = "Relative frequency or probability",
  ylim = c(0, max(observed_frequency, fitted_probability))
)
lines(
  x = count_values, y = fitted_probability,
  type = "b", pch = 19, lwd = 2
)
legend(
  "topright",
  legend = c("Observed relative frequency", "Fitted probability"),
  col = c("grey60", "black"), lty = 1, lwd = c(5, 2),
  pch = c(NA, 19), bty = "n"
)
```

Large, systematic differences at zero or in the upper tail
would call the model into question. Some differences are expected
from a finite sample, so a plot is a check rather than proof that
the model is correct.

# Exercise 4: Uncertainty in mean tree height

Fit the normal model on the $(\mu,\log\sigma)$ scale. Inverting
the Hessian gives an approximate covariance matrix on that
same fitted scale. Its diagonal supplies approximate squared
standard errors.

```{r}
set.seed(seed = 544)
heights = rnorm(n = 30, mean = 22, sd = 4)

normal_nll = function(par, heights) {
  mu = par[1]
  sigma = exp(par[2])
  -sum(dnorm(x = heights, mean = mu, sd = sigma, log = TRUE))
}

fit = optim(
  par = c(mu = mean(heights), log_sigma = log(sd(heights))),
  fn = normal_nll, heights = heights,
  method = "BFGS", hessian = TRUE
)
fit$convergence
c(
  mu_m = unname(fit$par[1]),
  sigma_m = exp(unname(fit$par[2]))
)

covariance_on_fit_scale = solve(fit$hessian)
se_on_fit_scale = sqrt(diag(covariance_on_fit_scale))
se_on_fit_scale

z_critical = qnorm(p = 0.975)
mu_quadratic_ci = fit$par[1] +
  c(-1, 1) * z_critical * se_on_fit_scale[1]
sigma_quadratic_ci = exp(
  fit$par[2] +
    c(-1, 1) * z_critical * se_on_fit_scale[2]
)
mu_quadratic_ci
sigma_quadratic_ci
```

The first standard error and interval refer to mean height
in metres. The second standard error refers to
$\log\sigma$; exponentiating its interval endpoints gives
an interval of approximately 3.13–5.20 m for the standard
deviation.
The quadratic intervals rely on a local normal approximation
to the estimator's sampling distribution. Check that the Hessian
can be inverted and gives sensible positive variances.

For a candidate $\mu$, re-estimate $\sigma$ using the root
mean squared deviation from that candidate mean. The
relative profile log-likelihood must be at least
$-qchisq(0.95,1)/2$ to lie inside the approximate 95%
profile interval. The function `crossing()` is zero
at either interval endpoint.

```{r}
profile_mu = function(mu, heights) {
  sigma_at_mu = sqrt(mean((heights - mu)^2))
  sum(dnorm(
    x = heights, mean = mu, sd = sigma_at_mu, log = TRUE
  ))
}

mu_hat = fit$par[1]
max_log_likelihood = -fit$value
cutoff = qchisq(p = 0.95, df = 1) / 2
crossing = function(mu) {
  profile_mu(mu = mu, heights = heights) -
    max_log_likelihood + cutoff
}

mu_profile_ci = c(
  lower = uniroot(
    f = crossing,
    interval = c(mu_hat - 10 * sd(heights), mu_hat)
  )$root,
  upper = uniroot(
    f = crossing,
    interval = c(mu_hat, mu_hat + 10 * sd(heights))
  )$root
)
rbind(
  quadratic_mu = mu_quadratic_ci,
  profile_mu = mu_profile_ci
)
```

The quadratic interval for $\mu$ is approximately 19.84–22.73 m;
the profile interval is approximately 19.80–22.78 m. They differ
because the profile follows the log-likelihood while re-estimating
$\sigma$, whereas the quadratic interval uses curvature near
the fitted point. Both describe uncertainty about **mean tree
height in metres**, not the spread of individual heights.
Their coverage depends on the normal model, independence,
and the approximations used. Neither assigns 95% probability
to the fixed population mean being inside the already
calculated endpoints.
