---
title: "Grouped ecological response curves"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to model repeated measurements with grouped
ecological response curves. A grouped model lets us estimate a common response
and variation among groups while recognising that measurements from one group
need not behave as independent rows.

In Chapter 7, we fitted one light-response curve to net-photosynthesis
measurements. The original experimental record now adds grouping information:
at each light level, the five rows belong to the same five leaves, in the same
order. This lets us ask how net photosynthesis changes with light and whether
all leaves follow the same response curve.

That information changes the model. One observational unit is a net
photosynthesis measurement on a leaf at one light intensity. The grouping unit
is a leaf, which contributes several observations. Measurements on a leaf can
share a tendency to lie above or below the common curve.

This chapter compares ways of sharing information between leaves. It then uses
a grouped likelihood to estimate both the common light response and variation
among leaves. The practical applies the same ideas to repeated tree-height
measurements in `datasets::Loblolly`.

# Learning goals

After this chapter, you should be able to:

- identify repeated measurements and distinguish observational and grouping
  units;
- explain fixed effects, random effects, mixed-effects, hierarchical, and
  multilevel models in a concrete example;
- compare complete pooling, no pooling, and partial pooling for a response
  curve;
- interpret shrinkage, regularisation, and the variance components
  $\sigma^2_{\mathrm{leaf}}$ and $\sigma^2_{\mathrm{obs}}$;
- explain why a grouped model can use maximum marginal likelihood and the
  intuition and limitations of Laplace approximation; and
- fit a grouped RTMB model with `optim()` and interpret its shared parameters,
  variance components, and conditional group-specific curves.

# Identify the repeated measurements

The light-response data are the same supplied data used in Chapter 7. The CSV
file alone does not identify a leaf. The author supplied the mapping below from
the original experimental record: within every set of five rows at a light
level, rows 1 to 5 came from leaves 1 to 5 respectively. This is not a general
rule for reconstructing repeated measurements from row order.

```r
# Load the supplied light-response data.
light = read.csv("../Chapter_7/Practicals/shapes1.csv")
# Attach the leaf mapping supplied from the original experimental record.
light$leaf_id = factor(rep(1:5, times = 11))

# Check the number of measurements at each light level and for each leaf.
table(light$x)
table(light$leaf_id)
```

The two tables check the supplied structure: there are five measurements at
each of 11 light levels, and 11 measurements for each derived leaf identifier.
The following plot joins only observations with the author-supplied same leaf
identifier. Symbols and line types distinguish leaves as well as colour.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Net photosynthesis measurements by author-supplied leaf identifier. Lines join repeated measurements from the same leaf."
# Load the data and apply the author-supplied leaf mapping.
light = read.csv("../Chapter_7/Practicals/shapes1.csv")
light$leaf_id = factor(rep(1:5, times = 11))
# Use both colour and symbol to distinguish the leaf trajectories.
leaf_colours = c("black", "#0072B2", "#D55E00", "#009E73", "#CC79A7")
leaf_symbols = c(16, 17, 15, 18, 8)
old_par = par(mar = c(4, 4, 1, 1))
# Draw the measurements, then join observations from the same leaf.
plot(light$x, light$y, type = "n",
     xlab = "Light intensity (recorded units)",
     ylab = "Net photosynthesis (recorded units)")
for (j in seq_along(levels(light$leaf_id))) {
  leaf_data = light[light$leaf_id == levels(light$leaf_id)[j], ]
  lines(leaf_data$x, leaf_data$y, lty = j, col = leaf_colours[j])
  points(leaf_data$x, leaf_data$y, pch = leaf_symbols[j],
         col = leaf_colours[j])
}
legend("bottomright", legend = paste("Leaf", 1:5),
       col = leaf_colours, pch = leaf_symbols, lty = 1:5, bty = "n")
# Restore the graphical settings used before this figure.
par(old_par)
```

:::: {.callout-important title="Question: what does row order establish?"}
If this identifier had not been supplied, would repeated point patterns at each
light level establish five leaf trajectories?

::: {.callout-tip title="Answer" collapse="true"}
No. Repeated values at each light level show replication at that level, but
they do not say which observation at one light level belongs with an
observation at another. The trajectories in the figure depend on the
author-supplied experimental record.
:::
::::

# Three ways to share information

We retain Chapter 7's net-photosynthesis response function,

$$
m(x;r,a,k)=r+a(1-e^{-kx}), \qquad a>0,\quad k>0.
$$

Here $x$ is light intensity, $r$ is expected net photosynthesis at zero light,
$r+a$ is its limiting expected value, and $k$ controls how quickly the curve
increases. Net photosynthesis can be negative at low light when respiration
exceeds carbon fixation.

## Names for the parts of the model

The concrete model helps make common terminology less mysterious:

- **Fixed effects** are the shared, population-level parameters estimated from
  these data, such as $r$, $a$, and $k$. Fixed does not mean known or causal.
- **Random effects** are group deviations, such as $u_j$, modelled with a
  distribution whose variance is estimated.
- A **mixed-effects model** contains both shared parameters and random effects.
- A **hierarchical** or **multilevel model** represents observations nested
  within leaves.

"Random effect" does not mean that leaves were randomly allocated to a
treatment. It means that the model represents among-leaf variation with a
distribution, conditional on the chosen curve and the other assumptions.

The table separates three choices about leaf information while keeping the
same curve and normal observation model.

| Model | Leaf component | Question |
|---|---|---|
| Complete pooling | $u_j=0$ for every leaf | What common curve describes the recorded leaves? |
| No pooling | A separately estimated $r_j$ for every leaf | How do these leaves differ when no information is shared? |
| Partial pooling | $r_j=r+u_j$, with $u_j\sim\operatorname{Normal}(0,\sigma^2_{\mathrm{leaf}})$ | What common curve and among-leaf variation describe the data? |

The following conceptual figure uses chosen parameter values, not estimates
fitted to the light data.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Illustrative complete-pooling, no-pooling, and partial-pooling response curves. The partial-pooling deviations are closer to the common intercept than the no-pooling intercepts."
# Define the light values and response curve used in this conceptual figure.
x_grid = seq(from = 0, to = 1000, length.out = 200)
light_curve = function(x, r, a, k) {
  r + a * (1 - exp(-k * x))
}
# Choose illustrative, rather than fitted, intercepts for five leaves.
common_r = -2
leaf_r = c(-4, -3, -2, -1, 0)
partial_r = common_r + 0.6 * (leaf_r - common_r)
leaf_colours = c("black", "#0072B2", "#D55E00", "#009E73", "#CC79A7")
old_par = par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
# Draw one shared curve for complete pooling.
plot(x_grid, light_curve(x_grid, common_r, 11, 0.004), type = "l",
     ylim = c(-5, 10), lwd = 2,
     xlab = "Light intensity (recorded units)",
     ylab = "Expected net photosynthesis (recorded units)",
     main = "Complete pooling")
# Draw separately fitted intercepts for no pooling.
plot(x_grid, light_curve(x_grid, leaf_r[1], 11, 0.004), type = "n",
     ylim = c(-5, 10), xlab = "Light intensity (recorded units)",
     ylab = "Expected net photosynthesis (recorded units)",
     main = "No pooling")
for (j in 1:5) {
  lines(x_grid, light_curve(x_grid, leaf_r[j], 11, 0.004),
        lty = j, col = leaf_colours[j])
}
# Draw intercepts pulled towards the common value for partial pooling.
plot(x_grid, light_curve(x_grid, partial_r[1], 11, 0.004), type = "n",
     ylim = c(-5, 10), xlab = "Light intensity (recorded units)",
     ylab = "Expected net photosynthesis (recorded units)",
     main = "Partial pooling")
for (j in 1:5) {
  lines(x_grid, light_curve(x_grid, partial_r[j], 11, 0.004),
        lty = j, col = leaf_colours[j])
}
# Identify the five curves and restore the original graphical settings.
legend("bottomright", legend = paste("Leaf", 1:5),
       col = leaf_colours, lty = 1:5, bty = "n", cex = 0.7)
par(old_par)
```

With complete pooling, every leaf has the same expected curve. With no
pooling, each leaf receives its own intercept but the curve's increase and
shape are still shared. These $r_j$ values are ordinary separately fitted
parameters, not random effects. No pooling is useful for describing these
recorded leaves, but a lightly measured leaf has no direct way to borrow
information from the others.

For partial pooling, use the model

$$
\begin{aligned}
u_j &\sim \operatorname{Normal}(0,\sigma^2_{\mathrm{leaf}}),\\
Y_{ij}\mid u_j,x_{ij} &\sim
\operatorname{Normal}\{m(x_{ij};r,a,k)+u_j,\sigma^2_{\mathrm{obs}}\}.
\end{aligned}
$$

$Y_{ij}$ is measurement $i$ on leaf $j$, $x_{ij}$ is its light intensity, and
$u_j$ is leaf $j$'s deviation from the common curve. Conditional on the
predictor, shared parameters, and its leaf effect, the normal observation model
assumes the measurements are independent. It would be wrong to describe all 55
rows as independent without that qualification.

::: {.callout-warning title="More parameters do not remove every dependence"}
Separate leaf intercepts do not automatically make the normal observation
model adequate. For example, measurements can still be correlated over time
within a leaf, or leaves can differ in curve shape rather than only intercept.
The grouping structure helps us state a better model; it does not prove its
remaining assumptions.
:::

# Shrinkage and variance components

Partial pooling is a compromise between one common curve and wholly separate
curves. The normal distribution for $u_j$ makes very large leaf deviations less
compatible with the group model unless the leaf's data support them. This is
**regularisation**: it is part of the model and likelihood, not a correction
applied after fitting.

The result is **shrinkage**. A noisy or lightly measured leaf is generally
pulled further towards the common curve than a precisely measured leaf. How
much it is pulled depends on its measurements and on the estimated
$\sigma^2_{\mathrm{leaf}}$. Shrinkage does not show that a leaf equals the
population mean; it expresses uncertainty about its individual deviation.

::: {.callout-note title="The two variance components"}
$\sigma^2_{\mathrm{leaf}}$ describes variation in leaf deviations around the
common curve. $\sigma^2_{\mathrm{obs}}$ describes variation of measurements
around a leaf's shifted curve. Both have squared net-photosynthesis units.
Their standard deviations, $\sigma_{\mathrm{leaf}}$ and
$\sigma_{\mathrm{obs}}$, have net-photosynthesis units, as do the deviations
$u_j$. R's `dnorm(sd = ...)` takes a standard deviation, so code that estimates
a standard deviation on a log scale squares it when reporting model variance.
:::

If the fitted $\sigma^2_{\mathrm{leaf}}$ is near zero, partial pooling
approaches complete pooling. A variance estimate on this boundary makes usual
large-sample confidence intervals delicate. The five balanced leaves here also
give limited opportunity to see unequal shrinkage visually.

# Maximum marginal likelihood

For one candidate leaf effect $u_j$, the joint contribution has two parts: the
normal densities for that leaf's measurements around its shifted curve and the
normal density that says how compatible $u_j$ is with the leaf population. The
leaf effect is unobserved, so maximum marginal likelihood compares shared
parameters after integrating it out:

$$
L_{j,\mathrm{marg}}(\boldsymbol\theta\mid\mathbf y_j)=
\int_{-\infty}^{\infty}
L_{j,\mathrm{joint}}(\boldsymbol\theta,u_j\mid\mathbf y_j)\,du_j.
$$

Here $\mathbf y_j$ is all observations from leaf $j$, and
$\boldsymbol\theta$ contains the shared curve parameters and both variance
components. The integrand is the joint likelihood contribution from the leaf's
measurements and its group distribution. The integral considers all possible
leaf deviations, weighting each by how well it agrees with both the leaf's
observations and the population model. The full marginal likelihood combines
these contributions across leaves.

Fitting only the single best $u_j$ as an ordinary fixed parameter is a different
calculation. Marginal likelihood incorporates uncertainty in unobserved leaf
effects before comparing shared parameter values. You will interpret this idea;
you will not calculate the integral by hand.

# Laplace approximation and RTMB

The integral is usually impractical to solve algebraically for a nonlinear
response curve. **Laplace approximation** uses the fact that the integrand is
often largest near the leaf effect that balances that leaf's observations and
the group distribution. Near this peak, a smooth log integrand can be
approximated by a quadratic. A quadratic log curve has a normal-shaped peak with
a calculable area, which gives an approximation to the integral.

This approximation can be unreliable when conditional shapes are strongly
non-normal, information is weak, a parameter is on a boundary, or the
observation model is unsuitable. In a Gaussian linear random-intercept model,
the calculation can be exact; that fact does not automatically extend to this
nonlinear light-response model.

RTMB records a grouped model and performs this numerical work. In the following
outline, `u` is a vector of candidate leaf deviations. The `u %~%` line records
their group distribution, and `random = "u"` tells `MakeADFun()` to integrate
them out using Laplace approximation. RTMB returns the resulting marginal
objective and derivatives for `optim()`.

```r
# Load RTMB before recording the grouped likelihood calculation.
library(RTMB)

# Create consecutive indices that select each leaf's deviation.
light$leaf_index = as.integer(light$leaf_id)
# Define the grouped normal model and its two sources of variation.
leaf_model = function(par) {
  a = exp(par$log_a)
  k = exp(par$log_k)
  sigma_leaf = exp(par$log_sigma_leaf)
  sigma_obs = exp(par$log_sigma_obs)
  mu = par$r + a * (1 - exp(-k * light$x)) + par$u[light$leaf_index]
  par$u %~% dnorm(mean = 0, sd = sigma_leaf)
  light$y %~% dnorm(mean = mu, sd = sigma_obs)
}

# Supply starting values, including one initial deviation per leaf.
leaf_parameters = list(
  r = 0,
  log_a = log(10),
  log_k = log(0.01),
  log_sigma_leaf = log(1),
  log_sigma_obs = log(1),
  u = rep(0, 5)
)
# Record the marginal objective, then minimise it using its gradient.
leaf_obj = MakeADFun(leaf_model, leaf_parameters,
                      random = "u", silent = TRUE)
leaf_fit = optim(par = leaf_obj$par, fn = leaf_obj$fn,
                 gr = leaf_obj$gr, method = "BFGS")
```

The fitted positive values `sigma_leaf` and `sigma_obs` are standard
deviations. Their squares are the model variances
$\sigma^2_{\mathrm{leaf}}$ and $\sigma^2_{\mathrm{obs}}$. Conditional fitted
leaf effects can help draw curves, but they are not independent leaf maximum
likelihood estimates. RTMB does not guarantee convergence or establish that the
model describes the data suitably. Its [introduction vignette](https://cran.r-project.org/web/packages/RTMB/vignettes/RTMB-introduction.html)
and the [Template Model Builder paper](https://www.jstatsoft.org/article/view/v070i05)
give technical background.

# Interpretation and limits

A fitted curve for an observed leaf conditions on its estimated leaf effect. A
prediction for a new leaf also includes among-leaf variation, so it should be
more variable. Five leaves provide limited information about a population
distribution. The normal leaf distribution and common curve shape are model
assumptions, and a random intercept can miss genuine leaf differences in the
rate or shape of response.

The practical now applies this workflow to a distinct dataset of repeated tree
heights. It fits complete pooling, no pooling, and partial pooling models in
RTMB, while letting RTMB perform the marginalisation.

# Chapter quiz

:::: {.callout-important title="Question 1: units"}
What is the observational unit and what is the grouping unit in the light data?

::: {.callout-tip title="Answer" collapse="true"}
An observational unit is one net-photosynthesis measurement at one light
intensity on a leaf. A grouping unit is one leaf, which contributes repeated
measurements.
:::
::::

:::: {.callout-important title="Question 2: pooling"}
How do complete pooling, no pooling, and partial pooling differ?

::: {.callout-tip title="Answer" collapse="true"}
Complete pooling uses one common curve. No pooling estimates a separate leaf
intercept without sharing information between those intercepts. Partial pooling
models leaf deviations from a common intercept with an estimated variance.
:::
::::

:::: {.callout-important title="Question 3: shrinkage"}
Does shrinkage show that a leaf is the same as the population mean?

::: {.callout-tip title="Answer" collapse="true"}
No. It reflects the compromise between a leaf's data and the estimated
among-leaf distribution, especially when the leaf provides limited information.
:::
::::

:::: {.callout-important title="Question 4: two variances"}
What do $\sigma^2_{\mathrm{leaf}}$ and $\sigma^2_{\mathrm{obs}}$ describe?

::: {.callout-tip title="Answer" collapse="true"}
The first describes variation in leaf deviations around the common curve. The
second describes measurement-to-measurement variation around a leaf's curve.
:::
::::

:::: {.callout-important title="Question 5: marginal likelihood"}
What is integrated over in the marginal likelihood for one leaf?

::: {.callout-tip title="Answer" collapse="true"}
All possible unobserved values of that leaf's deviation are integrated over,
with their compatibility with the observations and group distribution included.
:::
::::

:::: {.callout-important title="Question 6: Laplace approximation"}
What is the main intuition behind Laplace approximation?

::: {.callout-tip title="Answer" collapse="true"}
It approximates the log integrand near its important peak by a quadratic, then
uses the area of the resulting normal-shaped peak to approximate the integral.
:::
::::

:::: {.callout-important title="Question 7: RTMB"}
What does `random = "u"` mean in `MakeADFun()`?

::: {.callout-tip title="Answer" collapse="true"}
It tells RTMB to treat `u` as random effects and calculate a marginal objective
by integrating them out with Laplace approximation.
:::
::::
