---
title: "Numerical optimisation: local and global search for ecological models"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to find and diagnose parameter values that
minimise a likelihood-based objective function. Numerical optimisation is the
search a computer performs after we have specified a probability model and its
likelihood. We examine how an optimiser proposes parameter values, why searches
can end at different places, and what evidence supports a numerical result.

In [Chapter 4](../Chapter_4/Theory.qmd#sec-maximum-likelihood), we constructed
negative log-likelihoods and supplied them to `optim()`. Chapters 6 and 7 used
the same idea for nonlinear response curves and grouped models. This chapter
looks inside that numerical search.

An optimiser answers a numerical question: which allowed parameter values give
the smallest objective value it can find? Successful termination does not show
that the probability model, its independence assumptions, or its ecological
mechanism are suitable. Numerical convergence and model adequacy are separate
questions.

# Learning goals

After studying this chapter, you should be able to:

- define an objective function and parameter space, and distinguish local from
  global minima and a stopping rule;
- explore a low-dimensional objective with a grid and use `optimize()` for a
  bounded one-dimensional problem;
- choose and use suitable local methods in `optim()` or `nlminb()`;
- explain how gradients, Hessians, transformations, bounds, and scaling affect
  a numerical search;
- recognise difficult objective surfaces and use simulated annealing,
  differential evolution, or a real-valued genetic algorithm when appropriate;
- combine broad exploration with local refinement; and
- diagnose numerical convergence without confusing it with parameter
  identification or model adequacy.

# The numerical optimisation problem

Let $\boldsymbol\theta$ be a vector of model parameters. We use the objective

$$
Q(\boldsymbol\theta)=-\ell(\boldsymbol\theta),
$$

where $\ell$ is the log-likelihood for the observed data. Minimising $Q$ is
equivalent to maximising the likelihood. The likelihood construction is already
covered in [Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood); here our subject
is the search across the allowed **parameter space**.

The **objective value** is $Q$ at one candidate parameter vector. A **local
minimum** is lower than nearby points, whereas a **global minimum** is no higher
than any allowed point. An **iteration** is one update of an algorithm. It may
require one or many objective or derivative evaluations, so iteration counts
from different algorithms are not directly comparable.

A **stopping rule** ends the search when, for example, changes in parameters or
the objective become sufficiently small, or an evaluation limit is reached.
The word *convergence* usually means that such a numerical rule was satisfied.
It does not mean that the global minimum was proved or that the fitted model is
scientifically convincing.

::: {.callout-warning title="Convergence is not model checking"}
A convergence code describes the numerical search, not the ecological model. A
fit can converge to a local minimum, estimate a weakly identified parameter, or
rely on an unsuitable observation model.
:::

We will first use a smooth simulated example. One observational unit is the
number of grazing insects recorded at one nutrient concentration. Conditional
on the recorded concentration $x_i$, let the count $Y_i$, which can take values
$0,1,2,\ldots$, have the model

$$
Y_i\sim\operatorname{Poisson}(\mu_i),\qquad
\mu_i=\frac{Ax_i}{h+x_i},
$$

where $A>0$ is the asymptotic expected insect count and $h>0$ is the nutrient
concentration at which the expected count is half of $A$. The Poisson model
assumes that the counts are conditionally independent and that both their mean
and variance equal $\mu_i$. This resembles the response curves fitted in
Chapter 6, but our new question concerns the search algorithm. The data below
are simulated, not field observations.

```r
sat_mean = function(x, A, h) {
  A * x / (h + x)
}

sat_nll = function(par, x, y) {
  # Transform both positive response-curve parameters to their natural scales.
  A = exp(par[1])
  h = exp(par[2])
  mu = sat_mean(x = x, A = A, h = h)
  # Sum log probability masses because `optim()` minimises this objective.
  -sum(dpois(x = y, lambda = mu, log = TRUE))
}

# Simulate one insect count at each recorded nutrient concentration.
set.seed(seed = 901)
nutrient = seq(0.5, 12, length.out = 36)
expected_count = sat_mean(x = nutrient, A = 22, h = 3)
insect_count = rpois(n = length(expected_count), lambda = expected_count)
```

```{r}
#| echo: false
#| eval: true
#| fig-cap: "A one-dimensional grid through a smooth negative log-likelihood. The filled point is the lowest value on the fine grid."
sat_mean = function(x, A, h) {
  A * x / (h + x)
}
sat_nll = function(par, x, y) {
  A = exp(par[1])
  h = exp(par[2])
  mu = sat_mean(x = x, A = A, h = h)
  -sum(dpois(x = y, lambda = mu, log = TRUE))
}
set.seed(seed = 901)
nutrient = seq(0.5, 12, length.out = 36)
# Recreate the simulated insect counts used to draw this likelihood slice.
expected_count = sat_mean(x = nutrient, A = 22, h = 3)
insect_count = rpois(n = length(expected_count), lambda = expected_count)
# Hold h fixed to show how the objective changes along the A direction.
a_grid = seq(12, 32, length.out = 300)
grid_value = vapply(a_grid, function(A) {
  sat_nll(par = c(log(A), log(3)), x = nutrient, y = insect_count)
}, 0.0)
plot(a_grid, grid_value, type = "l",
     xlab = "Asymptotic expected insect count, A",
     ylab = "Negative log-likelihood")
best_grid = which.min(grid_value)
points(a_grid[best_grid], grid_value[best_grid], pch = 16)
```

# Direct exploration and one-dimensional search

A **grid search** evaluates the objective at selected parameter values. A coarse
grid reveals the broad shape; a fine grid then searches a promising interval.
The range must be wide enough to include scientifically plausible values, while
the spacing must be fine enough for the required accuracy. If each of $p$
parameters has $m$ grid values, a full grid requires $m^p$ evaluations. That
rapid growth makes grids most useful for one or two parameters and for
diagnosis rather than routine high-dimensional fitting.

For one continuous parameter, `optimize()` searches a supplied interval using a
combination of golden-section search and parabolic interpolation. Golden-section
search repeatedly reduces a bracket containing the minimum. Parabolic
interpolation uses recent objective values to propose a faster step when their
curvature is informative.

```r
one_parameter_nll = function(log_A, h, x, y) {
  sat_nll(par = c(log_A, log(h)), x = x, y = y)
}

# Search the log-scale interval while holding the half-saturation value fixed.
one_fit = optimize(
  f = one_parameter_nll,
  interval = log(c(5, 60)),
  h = 3, x = nutrient, y = insect_count
)
c(A = exp(one_fit$minimum), nll = one_fit$objective)
```

The bounds are part of the problem: `optimize()` does not discover a sensible
ecological interval for us. `optim(method = "Brent")` provides the same kind of
bounded one-dimensional search. When there are few parameters, a likelihood
slice or profile can provide an additional diagnostic; [Chapter 4 explains the
distinction](../Chapter_4/Theory.qmd#sec-likelihood-surface).

# Smooth local optimisation

For a differentiable objective, the **gradient**
$\mathbf g(\boldsymbol\theta)$ is the vector of local slopes. At an interior,
well-identified minimum it should be close to zero. The **Hessian**
$\mathbf H(\boldsymbol\theta)$ is the matrix of second derivatives, describing
local curvature and interactions between parameter directions.

Newton's method uses both:

$$
\boldsymbol\theta_{k+1}
=\boldsymbol\theta_k-
\mathbf H(\boldsymbol\theta_k)^{-1}\mathbf g(\boldsymbol\theta_k).
$$

Near a regular minimum, the local quadratic approximation can make this step
fast. Far from the minimum, the curvature may point towards a poor step, the
Hessian may be nearly singular, or the objective may not be approximately
quadratic.

**BFGS** is a quasi-Newton method. It evaluates gradients but builds an
approximation to inverse curvature from successive steps rather than requiring
the full Hessian at every iteration. It is a useful default for a smooth,
unconstrained, reasonably scaled likelihood, but it remains a local search: its
result can depend on the starting value.

# Derivatives in practice

An analytical derivative is derived from the objective's equation. For the
saturating Poisson model on the log-parameter scale, one such gradient is:

```r
sat_gradient = function(par, x, y) {
  # Evaluate the gradient on the log-A and log-h fitting scale.
  A = exp(par[1])
  h = exp(par[2])
  mu = sat_mean(x = x, A = A, h = h)
  c(log_A = sum(mu - y),
    log_h = sum((y - mu) * h / (h + x)))
}

# Give BFGS the analytical gradient and request its local curvature matrix.
bfgs_fit = optim(
  par = c(log_A = log(15), log_h = log(5)),
  fn = sat_nll, gr = sat_gradient,
  x = nutrient, y = insect_count,
  method = "BFGS", hessian = TRUE
)
```

If `gr` is omitted, `optim()` can approximate the gradient with finite
differences. In one dimension, a centred approximation is

$$
Q'(\theta)\approx
\frac{Q(\theta+\varepsilon)-Q(\theta-\varepsilon)}{2\varepsilon}.
$$

A large $\varepsilon$ gives a poor local approximation. An extremely small
$\varepsilon$ subtracts nearly equal floating-point numbers and can expose
rounding error. Function noise makes the trade-off worse.

RTMB supplied `obj$gr` and `obj$he` in Chapter 6 by **automatic
differentiation**. It applies derivative rules to the recorded calculations and
avoids choosing a finite-difference step. It does not make a discontinuous model
smooth or repair an incorrect objective. Whether a gradient is analytical,
numerical, or automatic, compare it with an independent numerical check while
developing a model.

The Hessian can guide optimisation and can support the local uncertainty
approximation taught in
[Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-uncertainty). Here we use it
only as evidence about local geometry: a nearly zero eigenvalue can signal a
flat direction, while negative curvature would contradict a regular local
minimum.

# Derivative-free local optimisation

Nelder–Mead maintains a **simplex**: three vertices in two dimensions, four in
three dimensions, and generally one more vertex than the number of parameters.
It compares objective values and changes the simplex through:

- **reflection**, moving the worst vertex across the others;
- **expansion**, moving farther when reflection is promising;
- **contraction**, taking a smaller step when reflection disappoints; and
- **shrinkage**, pulling the simplex towards its best vertex.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "A two-dimensional simplex and illustrative reflection of its worst vertex. Nelder–Mead uses objective values at the vertices, not gradients."
# Define an illustrative simplex and reflect its worst vertex across the centroid.
simplex = rbind(c(1, 1), c(4, 1.5), c(2.5, 4))
centroid = colMeans(simplex[1:2, , drop = FALSE])
reflected = 2 * centroid - simplex[3, ]
# Draw the current simplex and the proposed reflection.
plot(rbind(simplex, reflected), type = "n", asp = 1,
     xlab = "Parameter 1", ylab = "Parameter 2")
polygon(simplex, border = "#0072B2", lwd = 2)
segments(simplex[3, 1], simplex[3, 2], reflected[1], reflected[2],
         lty = 2)
points(simplex, pch = c(16, 16, 4), col = "#0072B2", cex = 1.2)
points(reflected[1], reflected[2], pch = 17, col = "#D55E00")
legend("topright", c("Current simplex", "Worst vertex", "Reflection"),
       pch = c(NA, 4, 17), lty = c(1, NA, NA),
       col = c("#0072B2", "#0072B2", "#D55E00"), bty = "n")
```

Because it needs only objective values, Nelder–Mead can tolerate some
irregularity and is useful when reliable gradients are unavailable. It is often
less efficient than BFGS on a smooth problem and remains vulnerable to flat
regions and competing minima.

```r
# Start both local methods at the same log-scale parameter values.
nelder_fit = optim(
  par = c(log_A = log(15), log_h = log(5)),
  fn = sat_nll, x = nutrient, y = insect_count,
  method = "Nelder-Mead"
)
c(bfgs = bfgs_fit$value, nelder_mead = nelder_fit$value)
```

For these simulated data, both searches give $A\approx20.47$,
$h\approx2.29$, and the same negative log-likelihood to the displayed
precision. Agreement supports this smooth example, but it is not a universal
ranking of the algorithms.

# Constraints, transformations, and scaling

Parameter constraints come from the model and ecological question. Two common
approaches are:

- transform a positive parameter as $\theta=\exp(\eta)$ or a probability as
  $p=\operatorname{logit}^{-1}(\eta)$, then search over an unconstrained
  parameter; or
- give scientifically defensible lower and upper bounds to a box-constrained
  method.

`optim(method = "L-BFGS-B")` uses box constraints. `nlminb()` is another local
optimiser that accepts lower and upper bounds and optional gradients and
Hessians.

```r
# Define the same defensible parameter region on the log scale.
lower = log(c(A = 1, h = 0.1))
upper = log(c(A = 60, h = 20))

# Fit within those bounds using a gradient-based local method.
bounded_fit = optim(
  par = c(log_A = log(15), log_h = log(5)),
  fn = sat_nll, gr = sat_gradient,
  x = nutrient, y = insect_count,
  method = "L-BFGS-B", lower = lower, upper = upper
)

port_fit = nlminb(
  start = c(log_A = log(15), log_h = log(5)),
  objective = sat_nll, gradient = sat_gradient,
  lower = lower, upper = upper,
  x = nutrient, y = insect_count
)

sat_nll_natural = function(par, x, y) {
  # This version receives the positive parameters on their natural scales.
  mu = sat_mean(x = x, A = par[1], h = par[2])
  -sum(dpois(x = y, lambda = mu, log = TRUE))
}

# Fit the natural-scale version over the corresponding bounded region.
natural_bounded_fit = optim(
  par = c(A = 15, h = 5),
  fn = sat_nll_natural,
  x = nutrient, y = insect_count,
  method = "L-BFGS-B",
  lower = c(A = 1, h = 0.1),
  upper = c(A = 60, h = 20)
)

rbind(
  transformed_bfgs = exp(bfgs_fit$par),
  bounded_log_scale = exp(bounded_fit$par),
  nlminb_log_scale = exp(port_fit$par),
  natural_scale_bounds = natural_bounded_fit$par
)
```

A boundary estimate deserves attention. It may be ecologically plausible, but
it may also show that a bound is controlling the answer or that the parameter
is weakly identified. Widening a bound is informative only when the wider range
remains scientifically meaningful. Here the transformation-based BFGS fit and
the natural-scale bounded fit describe the same scientific parameter space and
reach the same ecological estimates. The table compares all four fits on the
ecological parameter scale. That agreement need not hold when a bound is active.

Optimisation also becomes harder when a one-unit step is enormous for one
parameter and negligible for another. Transformations often improve scaling.
For `optim()`, `control = list(parscale = ...)` expresses typical parameter
changes on each coordinate. Tolerances and evaluation limits should be adjusted
only after checking the objective and scaling; they cannot rescue an invalid
start or non-finite likelihood. Always report final estimates on their
ecological scales.

# Difficult objective surfaces

Local methods can struggle with separated basins, flat ridges, discontinuities,
invalid regions, expensive calculations, and many parameters. A zero convergence
code can still refer to the wrong basin or to a flat plateau.

Consider simulated insect counts before and after an unknown environmental
threshold $c$. One observation is again a count at one recorded predictor value.
The expected count is $\lambda_{\mathrm L}$ when $x_i<c$ and
$\lambda_{\mathrm H}$ otherwise. Moving $c$ without crossing an observed
predictor value changes no fitted means, so the likelihood is flat between
consecutive predictor values and jumps at the data. This verified piecewise
structure is a reason to use broad exploration; different local answers alone
would not prove it.

```r
threshold_nll = function(par, x, y) {
  # Transform the two positive Poisson means from their log scales.
  cut = par[1]
  low = exp(par[2])
  high = exp(par[3])
  # Return a finite penalty outside the scientifically defined search region.
  if (!is.finite(cut) || !is.finite(low) || !is.finite(high) ||
      cut < 10 || cut > 90 || low <= 0 || high <= 0) {
    return(1e10)
  }
  mu = ifelse(x < cut, low, high)
  value = -sum(dpois(x = y, lambda = mu, log = TRUE))
  if (is.finite(value)) value else 1e10
}

# Simulate the threshold-count study used by the global search examples.
set.seed(seed = 902)
threshold_x = seq(0, 100, length.out = 60)
threshold_mu = ifelse(threshold_x < 55, 3, 12)
threshold_y = rpois(n = length(threshold_x), lambda = threshold_mu)
```

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Negative log-likelihood over a threshold grid after fitting the two Poisson means at each threshold. Flat steps arise because group membership changes only when the threshold crosses an observed predictor value."
# Recreate the simulated threshold data and evaluate one fitted value per cut.
set.seed(seed = 902)
threshold_x = seq(0, 100, length.out = 60)
threshold_mu = ifelse(threshold_x < 55, 3, 12)
threshold_y = rpois(n = length(threshold_x), lambda = threshold_mu)
cut_grid = seq(10, 90, length.out = 500)
cut_nll = vapply(cut_grid, function(cut) {
  # Refit the two group means for each candidate threshold.
  low_y = threshold_y[threshold_x < cut]
  high_y = threshold_y[threshold_x >= cut]
  low_mean = mean(low_y)
  high_mean = mean(high_y)
  -sum(dpois(low_y, lambda = low_mean, log = TRUE)) -
    sum(dpois(high_y, lambda = high_mean, log = TRUE))
}, 0.0)
plot(cut_grid, cut_nll, type = "s",
     xlab = "Environmental threshold (recorded units)",
     ylab = "Fitted negative log-likelihood")
```

Useful checks include several plausible starts, objective plots in low
dimensions, comparison of final objective values, gradient checks where
derivatives exist, and a restart from the fitted values. Complex marginal
objectives such as those in Chapter 7 make these checks more important; we do
not need to repeat their construction here.

# Simulated annealing

`optim(method = "SANN")` uses **simulated annealing** (SANN) to propose
stochastic moves. It accepts improvements and can accept worse moves,
particularly at a high temperature. Cooling gradually shifts the balance from
broad exploration towards local exploitation. This is an optimisation procedure,
not posterior sampling or Markov chain Monte Carlo (MCMC).

```r
# Set a seed so this stochastic search can be reproduced.
set.seed(seed = 903)
sann_fit = optim(
  par = c(cut = 25, log_low = log(3), log_high = log(10)),
  fn = threshold_nll,
  x = threshold_x, y = threshold_y,
  method = "SANN",
  control = list(maxit = 20000, temp = 10,
                 parscale = c(15, 1, 1))
)
```

The objective must return a finite penalty for proposals outside the intended
domain because SANN does not apply `lower` and `upper` bounds. In R, SANN's
`maxit` is its number of function evaluations and its reported convergence code
does not demonstrate stability. Set a seed for reproducibility and compare
multiple runs.

# Differential evolution with DEoptim

Differential evolution maintains a population of real-valued parameter vectors.
One common mutation rule creates

$$
\mathbf v_i=
\mathbf x_{r_0}+F(\mathbf x_{r_1}-\mathbf x_{r_2}),
$$

from three distinct population members. Crossover combines the mutant with the
current member, and selection retains the candidate with the lower objective.
Population differences therefore determine both direction and step size.

`DEoptim::DEoptim()` minimises a scalar objective between supplied bounds. Its
main controls here are population size `NP`, differential weight `F`, crossover
probability `CR`, and generation limit `itermax`. `DEoptim` is an add-on R
package, so install it before working through this example.

```r
# Define the scientifically defensible region for every population member.
lower = c(cut = 10, log_low = log(0.2), log_high = log(0.2))
upper = c(cut = 90, log_low = log(30), log_high = log(30))

# Set a seed, then evolve a population within those bounds.
set.seed(seed = 904)
de_fit = DEoptim::DEoptim(
  fn = threshold_nll,
  lower = lower, upper = upper,
  x = threshold_x, y = threshold_y,
  control = DEoptim::DEoptim.control(
    NP = 60, F = 0.8, CR = 0.9, itermax = 120,
    trace = FALSE
  )
)

de_fit$optim$bestmem
de_fit$optim$bestval
de_fit$optim$nfeval
de_fit$member$bestvalit
```

The bounds define the search region and must be scientifically defensible.
Differential evolution needs no derivative and can cross irregular regions, but
it repeatedly evaluates the objective. Keep the function efficient and ensure
it returns one finite scalar throughout the expected search region. Results can
vary across seeds, and the label *global search* does not prove that a global
minimum was found.

# Genetic algorithms with GA

A genetic algorithm also evolves a population. **Selection** favours candidates
with high fitness, **crossover** combines candidates, **mutation** introduces
variation, and **elitism** retains some of the strongest candidates. Search can
stop at a generation limit or after a specified run without improvement.

`GA::ga()` comes from the GA add-on package, which must be installed before you
run this example.

The sign convention matters:

- `optim()`, `nlminb()`, and `DEoptim::DEoptim()` minimise our negative
  log-likelihood;
- `GA::ga()` maximises a fitness value;
- therefore the fitness must be the negative of the negative log-likelihood.

```r
threshold_fitness = function(par, x, y) {
  # Convert the objective to a fitness because GA maximises.
  -threshold_nll(par = par, x = x, y = y)
}

# Set a seed, then search with real-valued candidate parameter vectors.
set.seed(seed = 905)
ga_fit = GA::ga(
  type = "real-valued", fitness = threshold_fitness,
  lower = lower, upper = upper,
  x = threshold_x, y = threshold_y,
  popSize = 60, maxiter = 150, run = 40,
  pcrossover = 0.8, pmutation = 0.1,
  monitor = FALSE
)

ga_candidate = as.numeric(ga_fit@solution[1, ])
ga_nll = -ga_fit@fitnessValue
ga_fit@summary
```

The package also supports binary and permutation encodings, but this chapter
uses only real-valued parameter vectors. `optim = TRUE` requests a hybrid
genetic algorithm that launches local searches from promising solutions. We
will keep the global and local calls separate below so their objective values
can be compared directly.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Progress of differential evolution and a real-valued genetic algorithm on the same simulated threshold likelihood. GA fitness is negated here to show both traces as negative log-likelihoods."
# Define and simulate the threshold example used by both population methods.
threshold_nll = function(par, x, y) {
  cut = par[1]
  low = exp(par[2])
  high = exp(par[3])
  if (!is.finite(cut) || !is.finite(low) || !is.finite(high) ||
      cut < 10 || cut > 90 || low <= 0 || high <= 0) {
    return(1e10)
  }
  mu = ifelse(x < cut, low, high)
  value = -sum(dpois(x = y, lambda = mu, log = TRUE))
  if (is.finite(value)) value else 1e10
}
set.seed(seed = 902)
threshold_x = seq(0, 100, length.out = 60)
threshold_mu = ifelse(threshold_x < 55, 3, 12)
threshold_y = rpois(n = length(threshold_x), lambda = threshold_mu)
lower = c(cut = 10, log_low = log(0.2), log_high = log(0.2))
upper = c(cut = 90, log_low = log(30), log_high = log(30))
# Run differential evolution and calculate GA fitness from the same objective.
set.seed(seed = 904)
de_fit = DEoptim::DEoptim(fn = threshold_nll, lower = lower, upper = upper,
  x = threshold_x, y = threshold_y,
  control = DEoptim::DEoptim.control(NP = 60, F = 0.8, CR = 0.9,
    itermax = 120, trace = FALSE))
threshold_fitness = function(par, x, y) {
  -threshold_nll(par = par, x = x, y = y)
}
set.seed(seed = 905)
ga_fit = GA::ga(type = "real-valued", fitness = threshold_fitness,
  lower = lower, upper = upper, x = threshold_x, y = threshold_y,
  popSize = 60, maxiter = 150, run = 40,
  pcrossover = 0.8, pmutation = 0.1, monitor = FALSE)
# Plot both progress traces as negative log-likelihoods.
plot(seq_along(de_fit$member$bestvalit), de_fit$member$bestvalit,
     type = "l", lwd = 2, col = "#0072B2",
     xlab = "Generation", ylab = "Best negative log-likelihood",
     ylim = range(c(de_fit$member$bestvalit, -ga_fit@summary[, "max"])))
lines(seq_len(nrow(ga_fit@summary)), -ga_fit@summary[, "max"],
      lty = 2, lwd = 2, col = "#D55E00")
legend("topright", c("DEoptim", "GA"), lty = c(1, 2), lwd = 2,
       col = c("#0072B2", "#D55E00"), bty = "n")
```

# Hybrid global and local optimisation

Global and local methods are complementary. A useful workflow is:

1. Define and test one negative log-likelihood and its ecological domain.
2. Explore broadly with multiple starts or a population method.
3. Pass the strongest candidate to BFGS, L-BFGS-B, or `nlminb()`.
4. Inspect the refined objective, gradient, convergence information, and nearby
   surface.
5. Compare results across algorithms and stochastic seeds.
6. Transform estimates back to ecological units and check the fitted model.

```r
# Refine each population candidate with the same bounded local method.
refined_de = optim(
  par = de_fit$optim$bestmem,
  fn = threshold_nll,
  x = threshold_x, y = threshold_y,
  method = "L-BFGS-B", lower = lower, upper = upper
)

refined_ga = optim(
  par = ga_candidate,
  fn = threshold_nll,
  x = threshold_x, y = threshold_y,
  method = "L-BFGS-B", lower = lower, upper = upper
)

c(de_before = de_fit$optim$bestval,
  de_after = refined_de$value,
  ga_before = ga_nll,
  ga_after = refined_ga$value)
```

For the stated seeds, both population methods reach a threshold plateau near
55 and a negative log-likelihood of approximately 145.565. Local refinement
does not worsen that value and refines the two mean parameters. The exact
threshold within a plateau is not identified by these data. Agreement across
methods is reassuring, but it is not a proof of global optimality or model
adequacy.

# Comparing methods fairly

| Method | Exploration | Derivatives | Bounds | Reproducibility | Typical strength | Important limitation |
|---|---|---|---|---|---|---|
| Grid | Global over the grid | No | Grid range | Deterministic | Reveals low-dimensional geometry | Cost grows rapidly with dimension |
| Brent / `optimize()` | One-dimensional | No | Required | Deterministic | Efficient bounded scalar search | Only one parameter |
| BFGS | Local | Gradient; can be numerical | No box bounds | Deterministic | Efficient on smooth, scaled problems | Start and surface matter |
| Nelder–Mead | Local | No | No box bounds | Deterministic | Works without a gradient | Often slower; can stall |
| L-BFGS-B | Local | Gradient; can be numerical | Box bounds | Deterministic | Smooth constrained problems | Needs finite values and reliable local geometry |
| `nlminb()` | Local | Optional gradient and Hessian | Box bounds | Deterministic | Flexible bounded local fitting | Still start-dependent |
| SANN | Broad stochastic search | No | Must be handled in the objective | Stochastic | Can leave a local basin | Tuning and many evaluations |
| DEoptim | Population global search | No | Required | Stochastic | Irregular, bounded real problems | Computational cost; no proof of global optimum |
| GA | Population global search | No | Required here | Stochastic | Flexible population exploration | Fitness sign and tuning require care |

Fair comparisons use the same objective definition, comparable parameter
domains, and either the same parameterisation or an explicit reason for changing
it. Compare final objective values as well as estimates, record objective and
derivative evaluations where available, and repeat stochastic searches with
several seeds. A generation, a simplex iteration, and a BFGS iteration represent
different amounts of work. Timing comparisons are meaningful only on the same
machine and implementation and should be interpreted cautiously.

# A diagnostic workflow

- Use `optimize()` for a genuinely one-dimensional bounded problem.
- Use BFGS with reliable derivatives for a smooth, reasonably scaled local
  problem.
- Use Nelder–Mead when a gradient is unavailable or mildly unreliable.
- Use L-BFGS-B or `nlminb()` for box constraints.
- Use several plausible starts when more than one basin is possible.
- Consider SANN, DEoptim, or GA when local fits disagree or the objective is
  strongly irregular or multimodal.
- Refine a global candidate locally where the objective permits it.
- Check objective values, convergence messages, gradients, boundaries, and the
  surrounding surface.
- Finally, return to the ecological scale and assess the fitted observation
  model. Numerical convergence is not a model check.

# Summary

Optimisation turns a likelihood into a numerical search over allowed parameter
values. Smooth local methods use objective geometry efficiently, while
derivative-free and stochastic methods trade additional evaluations for less
dependence on local derivatives. Transformations, bounds, scaling, starts, and
stopping rules are part of the analysis, not afterthoughts. Difficult problems
often benefit from global exploration followed by local refinement. No
algorithm removes the need to inspect the objective and interpret the fitted
model ecologically.

The discussion of direct search, quasi-Newton ideas, numerical derivatives,
simplex search, simulated annealing, scaling, discontinuities, and fitting
challenges draws on Benjamin Bolker's *Optimization and all that* (supplied 2007
draft, especially Sections 1, 2, and 4). The R interfaces follow the current
[`optim()` documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/optim.html),
[`nlminb()` documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/nlminb.html),
[`DEoptim` reference manual](https://cran.r-project.org/web/packages/DEoptim/refman/DEoptim.html),
and [GA vignette](https://luca-scr.github.io/GA/articles/GA.html).

# Chapter quiz

:::: {.callout-important title="Question 1: local or global?"}

Three BFGS fits from plausible starts end at substantially different negative
log-likelihood values. What should you conclude and do next?

::: {.callout-tip title="Answer" collapse="true"}

The local searches have not established a common solution. Check the objective
and parameter domains, plot low-dimensional regions where possible, try more
starts, and consider broad stochastic exploration followed by local refinement.
Different results alone do not prove which minimum is global.

:::
::::

:::: {.callout-important title="Question 2: sign convention"}

Your function returns a negative log-likelihood. What should `GA::ga()` use as
fitness?

::: {.callout-tip title="Answer" collapse="true"}

Use the negative of the negative log-likelihood. `GA::ga()` maximises fitness,
whereas the other optimisers in this chapter minimise the objective.

:::
::::

:::: {.callout-important title="Question 3: derivative source"}

Why might an automatic gradient improve a BFGS fit without guaranteeing a
valid result?

::: {.callout-tip title="Answer" collapse="true"}

It avoids finite-difference step-size error for a smooth recorded objective.
It cannot repair an incorrect model, a discontinuity, weak identification, or
a poor ecological assumption.

:::
::::

:::: {.callout-important title="Question 4: boundary estimate"}

An L-BFGS-B fit places a rate at its upper bound. What should you inspect?

::: {.callout-tip title="Answer" collapse="true"}

Check whether the bound is scientifically justified, inspect the objective near
the boundary, and determine whether a wider defensible range changes the fit.
Do not silently report the boundary value as an unconstrained estimate.

:::
::::

:::: {.callout-important title="Question 5: stochastic reproducibility"}

Why are a seed and repeated runs both useful for DEoptim or GA?

::: {.callout-tip title="Answer" collapse="true"}

The seed makes one run reproducible. Distinct seeded runs reveal whether the
conclusion is stable under the algorithm's stochastic variation.

:::
::::

:::: {.callout-important title="Question 6: convergence"}

What does a zero convergence code establish?

::: {.callout-tip title="Answer" collapse="true"}

It indicates that the algorithm satisfied its numerical stopping rule. It does
not prove a global optimum, strong parameter identification, or an adequate
ecological probability model.

:::
::::
