# Learning outcomes

After completing the tutorial you will be able to:

1. Create, run, and source R scripts, and organize work in an RStudio project.

2. Use basic R syntax, functions, variables, and common scalar types.

3. Create and manipulate vectors, factors, and data frames, including selecting and summarizing data.

4. Make basic plots with base R graphics and ggplot2.

5. Use if statements, for loops, and vectorized operations.

6. Write and use simple functions.

7. Import tabular data and use dplyr to filter, transform, and summarize it.

8. Install and load packages, and find help for R functions.

# On the use of AI

This tutorial was written in collaboration with AI tools, including GPT-5.6 Luna
and Kimi K3, but it was designed and edited by a human.

**You may use AI** as a tutor to help you:

- Translate the material
- Explain concepts
- Provide examples
- Fix bugs in your code

If you use AI to generate the solutions to the exercises without even trying
yourself first, you will not learn the material and the course will become much
harder for you (especially the exam when AI is not there to help you).

::: {.callout-warning title="A note about AI"}
AI tools hallucinate and make mistakes, **you are responsible** for your learning
and assignments, **AI tools are not liable**.
:::

# How to use this tutorial

Please go through the tutorial in order, as each section builds on the previous
one. Run all the code examples in your own `R` session (**do not just read**),
and try to understand what each line of code does.

**I expect this tutorial to take less than 5 h**, depending on your prior knowledge
of R and programming.

We will assume that you have installed `R` and `RStudio` on your computer. If
you haven't done so, please follow the [installation instructions](https://posit.co/download/RStudio-desktop/) for your operating system.

In most of the sections we will use built-in datasets or small data examples. In
one of the latest sections we will learn how to import external data into an R
session. That will be the norm for the rest of the course. You will get these
files during the course.

# Introduction to R and RStudio

R is a programming language designed mostly for statistics and data analysis. It
is free and open source, and has a large community of users and developers. Much
functionality comes pre-installed with R, but there are also thousands of additional **packages** that can be installed to extend R's functionality.

R is interactive, meaning that you can type commands on a **console** and get
immediate feedback. This is useful for quickly testing ideas, but for any real
work you should write your code in a script file and run it from there. This way
you can save your work and re-run it later, and you can also share your code with
others.

RStudio is a program to help you write R code. It is also free and open source,
and it is the most popular (but not only) way to use R. RStudio offers two main
features:

  - **Projects**: To organize all the files for your data analysis

  - **Panels**: To allow you write R code and display output

## RStudio projects

Your analyses will also involve multiple data files, output figures and possibly
multiple script files. To keep your work organized, you should create a new
RStudio project for each project (e.g., your thesis, a class assignment, etc.).

Instructions on how to create an RStudio project can be found [here](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)

:::::: {.callout-important title="Exercise" collapse="true"}

Following the instructions in the link above, create a new RStudio project that you will
use for this tutorial. You can use any name you like and save it anywhere on your drive.

Ignore the different Project options just go with the defaults. Make sure that
you are working within the project for the remainder of this tutorial.

::::::

## RStudio panels

RStudio contains [multiple panels](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html).

- Script panel: This is where you write and edit your R scripts (it is closed if
no files are opened).
- Console panel: This is where you run your R code and see the output.
- Environment panel: This shows the variables in your current R session.
- Plotting panel: This is where your plots are displayed.
- Files panel: This is where you can navigate your file system and open files.
- Help panel: This is where you can access the R documentation and help files.

:::::: {.callout-important title="Exercise" collapse="true"}

Find all the panels mentioned above and check if they contain any information (most will not).

Try to resize the panels.

If you can't see a panel, you can open it from the `View` menu.

::::::

## Basic R syntax

We will start typing very simple R code in the console, to get a feeling of the syntax and
the execution loop of R.

### R commands

R code is made of **commands**. An R command is an instruction you give the computer to
perform some operation. If you type the R command in the console and press [Enter] you will
get back the result of the operation. For example, at the command prompt `>`, type in `2+2`
and hit `Enter`:

```{r}
2 + 2
```
To do anything useful, you have to store the results from calculations by
*assigning* them to variables using `=`. We always use `=` for assignment in this course.
For example:

```{r}
a = 2 + 2
```

`R` automatically creates the variable `a` and stores the result (4) in it, but
it doesn't print anything. To ask `R` to print the value, just type the variable name:

```{r}
a
```

The `[1]` at the beginning of the line is just `R` printing an index of the answer. This
is more useful for answers with multiple elements that span multiple lines.

### Multiline commands

Try the following:

```{r}
1:40
```

This created a vector of numbers from 1 to 40 and now you can see the index at the beginning
of each line.

You can break a command into multiple lines anywhere that `R` can tell you
haven't finished your command. In that case, `R` will give you a "continuation" prompt
(`+`) to let you know that it expects more input. Try the following incomplete command:

```{r}
#| error: true
a = 3*(4 -
```

`R` expects a second number after the `-` symbol and to close the parenthesis. It will keep
giving you the continuation prompt until you finish the command:

```{r}
#| error: true
2)
```

You will sometimes see the continuation prompt when you don't expect it. This indicates a bug in
your code (often you forgot to close a parenthesis). You can always hit the `Escape` key to
cancel the command.

### Variable names

Variable names in `R` can only contain:

- letters (a-z, A-Z)
- digits (0-9)
- periods (.)
- underscores (_)

And also

- cannot start with a digit (`x1` is valid, but `1x` is not)
- are case sensitive (`Abc` and `abc` are different variables)
- some words are reserved for the language (`NA`, `TRUE`, `FALSE`, `if`, `else`, `for`, `while`, etc.) and cannot be used as variable names.

### Basic math and function syntax

`R` does basic math calculations like in a calculator with the symbols:

- `+` for addition
- `-` for subtraction
- `*` for multiplication
- `/` for division
- `^` for exponentiation

For example:

```{r}
x = 5
y = 2
z1 = x*y
z2 = x/y
z3 = x^y
z1
z2
z3
```

There is a consistent order of preference of operations in `R`, but when in doubt just use
parenthesis.

`R` also has many *built-in mathematical functions* that operate on
variables such as:

* `abs`: Absolute values.
* `cos`, `sin` and `tan`: For trigonometry (arguments always in radians).
* `exp`: Exponential function.
* `log`, `log10`: Natural logarithm and logarithm base 10.
* `sqrt`: Square root.

All functions in R follow the same syntax: the function name followed by parentheses
containing the inputs to the function (called **arguments**). For example:

```{r}
abs(-1)
sqrt(4)
```

Many functions accept more than one argument, separated by commas. Some arguments are
*optional* and have a **default value** that is used unless you specify otherwise. For
example, `log` computes the natural logarithm by default, but you can pass a second
argument to use a different base:

```{r}
log(10)         # natural logarithm (default)
log(10, base = 10)
```

You can find the available arguments of a function (and their defaults) in the help
pages, see [R documentation](#r-documentation).

:::::: {.callout-important title="Exercise" collapse="true"}

Compute the following operations in R:

1. $\frac{2^7}{2^7 - 1}$ and compare it with $\left({1 - \frac{1}{2^7}}\right)^{-1}$

2.
  * $1 + 0.2$
  * $1 + 0.2 + 0.2^2/2 + \cos(2.3)$
  * $\log(1)$
  * $\exp(0.2)$


::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
(2^7)/(2^7-1)
(1-2^7)^(-1)
```


2.

```{r}
1 + 0.2
1 + 0.2 + 0.2^2/2 + cos(2.3)
log(1)
exp(0.2)
```

::::
:::::
::::::

## Script files

Please create a new R script file in RStudio using the available menu options and save it
inside the project folder. We will write the rest of the code of this tutorial in this file.

Script files often have comments. These are lines that start with a `#` symbol and are ignored
by R (they are meant for humans). You can use comments to explain what your code does, or to
temporarily disable a line of code. For example, go to the top of the file and type the
following:

```{r}
# Chapter 1 of Ecological Modelling and Data
```

You will see that the line is colored (by default in green) to highlight that it is a comment.
Let's add a simple arithmetic operation in the next line:

```{r}
a = 2 + 2
```

How do we execute this R command? There are multiple ways to do this:

1. Select the code with the mouse and (i) click the `Run` button in the top right corner of
the script panel or (ii) press `Ctrl + Enter` (or `Cmd + Enter` on an Apple device).

2. Place the cursor on the line and click the `Run` button or press `Ctrl + Enter`.

The first approach will send the select code to the console and execute it (as if you were
to copy and paste it there). The second approach will automatically recognize the entire
command (even if it is multiline) and send it to the console for execution.

There is a third option: to execute the entire script file, in the order in which it is
written. This is called **sourcing** the script. To do this, click the `Source` button in
the top right corner of the toolbar.

## R packages

A package is an R project made of:

  - A collection of R functions and sample data

  - Written documentation for functions and data (see [R documentation](#r-documentation) below)

There are two different steps when using a package:

  - Installing a package: This is done only once, and it downloads the package from the
internet and installs it in your computer.

  - Loading a package: This is done every time you start a new R session, and it makes the
functions and data in the package available for you to use.

The best way to install a package is to use the Packages panel in RStudio. Go to the panel,
click on `Install`, and the select the package of interest by typing the name.

:::::: {.callout-important title="Exercise" collapse="true"}

Install the package `ggplot2` using the Packages panel in RStudio.

::::::

Loading a package is done by using the `library` function with the name of the package as
argument. For example, to load the `ggplot2` package, you would type:

```{r}
library(ggplot2)
```

I recommend to always load the packages you will use at the beginning of your script file,
before any actual analysis code, so you can write all the `library()` commands at the start
of the script.

# Data types

So far we have only worked with numbers, but `R` can store and manipulate other kinds of
data. Every value in `R` has a **data type**, which determines what operations you can do
with it. Here we introduce the most basic types: values that hold a single element (often
called *scalars*). Later we will see how to combine them into vectors and data frames.

## Numeric

Numeric values are numbers, either with or without decimals. `R` does not make an important
distinction between integers and decimal numbers for everyday work:

```{r}
x = 5
y = 2.5
```

You can check the type of any value with the `class` function:

```{r}
class(x)
class(y)
```

Both are `"numeric"`, and all the mathematical operations and functions we saw earlier work
on them.

## Character (strings)

Text data is stored as **character** values (often called *strings*). Strings are always
enclosed in quotes, either single (`'`) or double (`"`):

```{r}
species = "Quercus robur"
class(species)
```

You cannot do math with strings, but there are many functions to manipulate them. A useful
one is `paste`, which joins strings together:

```{r}
genus = "Quercus"
species = "robur"
paste(genus, species)
```

Note that a number inside quotes is a string, not a number:

```{r}
#| error: true
a = "5"
a + 1    # this produces an error
```

## Logical

Logical values can only be `TRUE` or `FALSE` (always in capital letters, and without quotes).
They typically result from **comparisons** using the operators:

- `==` equal to (note the double `=`; a single `=` is assignment!)
- `!=` not equal to
- `<`, `<=` less than, less than or equal to
- `>`, `>=` greater than, greater than or equal to

For example:

```{r}
x = 5
x > 3
x == 3
class(x > 3)
```

Logical values can be combined with `&` (and), `|` (or) and `!` (not):

```{r}
(x > 3) & (x < 10)
(x < 3) | (x > 10)
!(x == 5)
```

Logical values will become very useful later, for example in if-statements and for
selecting parts of your data.

## Factors

Factors are a special data type for **categorical** data: values that can only take a
limited set of possibilities, called *levels*. This is very common in ecological data,
for example a treatment in an experiment or a habitat type:

```{r}
treatment = factor("control")
treatment
```

Factors behave in some ways like strings, but `R` keeps track of the allowed levels. We
will come back to factors when we discuss vectors and data frames, where they are most
useful. For now, just remember: use strings for free text, and factors for variables with
a fixed set of categories.

## Missing values: `NA`

Real ecological data almost always contains missing observations: a sensor failed, a
sample was lost, a measurement was not taken. `R` represents missing values with the
special value `NA` (without quotes):

```{r}
biomass = NA
```

`NA` is contagious: almost any operation involving `NA` returns `NA`, because the result
is unknown:

```{r}
biomass + 1
```

This also affects comparisons. To test whether a value is missing you cannot use
`== NA`; you must use the function `is.na`:

```{r}
biomass == NA   # does not work as you might expect
is.na(biomass)  # the correct way
```

Many functions, such as `mean`, return `NA` if the data contains missing values, and
offer an argument `na.rm = TRUE` to ignore them. We will see examples of this later.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Create a variable with the name of your study species (as a string) and a variable
with its average body mass (as a number). Check the type of both with `class`.

2. Use `paste` to print the sentence: "The average body mass of *species* is *mass* g".

3. Define `a = 10` and `b = 4`. Use comparison operators to check:
   * whether `a` is greater than `b`
   * whether `a` is equal to `b`
   * whether `a` is between 5 and 15

4. Assign `NA` to a variable `temp`. What is the result of `temp > 20`? What is the
result of `is.na(temp)`?

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
species = "Daphnia magna"
mass = 3.2
class(species)
class(mass)
```

2.

```{r}
paste("The average body mass of", species, "is", mass, "g")
```

3.

```{r}
a = 10
b = 4
a > b
a == b
(a > 5) & (a < 15)
```

4.

```{r}
temp = NA
temp > 20    # NA, the comparison is unknown
is.na(temp)  # TRUE
```

::::
:::::
::::::

# Vectors

The data types we saw before hold a single value, but in practice we always work with
collections of values: all the trees in a plot, all the dates in a time series, etc. The most
basic data structure in `R` is the **vector**: a sequence of values *of the same
type*.

## Creating vectors

You create a vector with the function `c` (for *combine*):

```{r}
mass = c(2.1, 3.4, 1.8, 4.0)
mass
class(mass)
```

Note that the `class` is still `numeric`. In fact, R treats variables with single values as
vectors of length 1.

We have already seen another way to create numeric vectors: the `:` operator creates a
sequence of integers:

```{r}
1:10
```

For more control use `seq`, which creates a sequence with a given step size:

```{r}
seq(from = 0, to = 1, by = 0.25)
```

And `rep` repeats values:

```{r}
rep("control", times = 4)
```

Vectors can be of any of the data types we saw before:

```{r}
species = c("Quercus robur", "Fagus sylvatica", "Quercus robur")
alive = c(TRUE, TRUE, FALSE)
class(species)
class(alive)
```

If you combine values of different types, `R` silently **coerces** everything to a single
type. This is a common source of bugs:

```{r}
c(1, 2, "3")
```

You can find the length of a vector with `length`:

```{r}
length(mass)
```

## Vectorized operations

One of the most useful features of `R` is that mathematical operations and functions work
on entire vectors at once, applied **element by element**:

```{r}
mass * 2
mass + 1
log(mass)
```

Operations between two vectors of the same length are also done element by element:

```{r}
height = c(10, 12, 8, 15)
mass / height
```

There are many useful functions that take a vector and return a single summary value:

```{r}
sum(mass)
mean(mass)
sd(mass)     # standard deviation
min(mass)
max(mass)
```

If the vector contains missing values, these functions return `NA`. Use the argument
`na.rm = TRUE` to ignore them:

```{r}
mass2 = c(2.1, NA, 1.8, 4.0)
mean(mass2)
mean(mass2, na.rm = TRUE)
```

## Indexing: selecting elements

You can select elements of a vector with square brackets `[ ]`. Indexing in `R` starts at
1 (not 0 like in some other languages):

```{r}
mass
mass[1]       # first element
mass[2:3]     # elements 2 and 3
```

A negative index *excludes* elements:

```{r}
mass[-1]      # everything except the first element
```

You can also index with a logical vector. This is extremely useful for selecting the parts
of your data that meet some condition:

```{r}
mass > 2              # a logical vector
mass[mass > 2]        # only the elements where the condition is TRUE
```

Combined with `is.na`, this lets you remove missing values:

```{r}
mass2
mass2[!is.na(mass2)]
```

## Factors in vectors

Factors become really useful as vectors. Imagine a vector of treatments in an experiment:

```{r}
treatment = factor(c("control", "fertilized", "control", "fertilized"))
treatment
levels(treatment)
```

`R` knows that this variable can only take the two levels, and statistical functions will
treat it appropriately later on.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Create a vector `dbh` with the diameter at breast height of 6 trees:
12.5, 8.3, 15.1, NA, 9.7, 11.2. Check its length.

2. Compute the mean and standard deviation of `dbh`, ignoring the missing value.

3. Use indexing to select:
   * the third tree
   * all trees except the first
   * all trees with a dbh larger than 10 (be careful with the `NA`!)

4. Create a factor vector `plot` indicating that the first three trees are in plot "A"
and the last three in plot "B". What are its levels?

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
dbh = c(12.5, 8.3, 15.1, NA, 9.7, 11.2)
length(dbh)
```

2.

```{r}
mean(dbh, na.rm = TRUE)
sd(dbh, na.rm = TRUE)
```

3.

```{r}
dbh[3]
dbh[-1]
dbh[dbh > 10]                    # includes an NA!
dbh[!is.na(dbh) & dbh > 10]      # the correct way
```

4.

```{r}
plot = factor(c("A", "A", "A", "B", "B", "B"))
levels(plot)
```

::::
:::::
::::::

# Matrices

A **matrix** is a rectangular, two-dimensional data structure. Like a vector, all
elements of a matrix must have the same type. Matrices are useful for ecological
data such as species abundances across several sites.

You create a matrix with the `matrix` function:

```{r}
abundance = matrix(
  c(12, 8, 15,
    5, 11, 9,
    20, 13, 18),
  nrow = 3,
  byrow = TRUE
)
abundance
```

The argument `nrow` specifies the number of rows. By default, `matrix` fills the
matrix by columns; `byrow = TRUE` makes it fill the matrix by rows.

You can add row and column names:

```{r}
rownames(abundance) = c("Quercus robur", "Fagus sylvatica", "Pinus sylvestris")
colnames(abundance) = c("plot_A", "plot_B", "plot_C")
abundance
```

Useful functions for inspecting a matrix include:

```{r}
dim(abundance)       # number of rows and columns
nrow(abundance)
ncol(abundance)
length(abundance)    # total number of elements
```

Matrices use the same indexing system as data frames, with a row index and a
column index separated by a comma:

```{r}
abundance[1, 2]                    # row 1, column 2
abundance["Quercus robur", ]       # one entire row
abundance[, "plot_B"]              # one entire column
abundance["Fagus sylvatica", "plot_C"]
```

Leaving one index empty selects all rows or all columns. Matrix operations are
usually vectorized:

```{r}
abundance * 2
abundance + 1
abundance > 10
```

The functions `rowSums` and `colSums` calculate totals for each row or column:

```{r}
rowSums(abundance)
colSums(abundance)
```

You can transpose a matrix, exchanging its rows and columns, with `t`:

```{r}
t(abundance)
```

Unlike a data frame, all elements in a matrix must have the same type. If you
combine numbers and character values, the numbers are converted to character
values, just as they are in a vector.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Create a matrix `cover` containing the percentage cover of three plant species
   in three plots:

| species | plot_A | plot_B | plot_C |
|---------|--------|--------|--------|
| grass   | 35     | 42     | 28     |
| moss    | 12     | 18     | 15     |
| fern    | 5      | 7      | 9      |

2. Add suitable row and column names.

3. Calculate the total cover of each species and the total cover in each plot.

4. Select all species with a cover greater than 20 in any plot.

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

```{r}
cover = matrix(
  c(35, 42, 28,
    12, 18, 15,
    5, 7, 9),
  nrow = 3,
  byrow = TRUE
)

rownames(cover) = c("grass", "moss", "fern")
colnames(cover) = c("plot_A", "plot_B", "plot_C")

rowSums(cover)
colSums(cover)

cover[cover > 20]
```

::::
:::::
::::::

# Lists

A **list** is a collection of objects. Unlike a vector or matrix, the elements of
a list do not have to be the same type or length. A list can contain numbers,
character strings, logical values, vectors, matrices, data frames, or even other
lists.

For example, we can store information about one tree in a list:

```{r}
tree = list(
  species = "Quercus robur",
  dbh = 25.4,
  alive = TRUE,
  measurements = c(24.8, 25.1, 25.4)
)

tree
class(tree)
length(tree)
names(tree)
str(tree)
```

The components of a named list can be accessed with `$`:

```{r}
tree$species
tree$dbh
tree$measurements
```

They can also be accessed with double square brackets:

```{r}
tree[["species"]]
tree[["dbh"]]
```

The single-square-bracket form returns a list containing the selected component,
whereas `$` and `[[ ]]` return the component itself:

```{r}
tree["dbh"]       # a list containing dbh
tree[["dbh"]]     # the numeric value of dbh
```

You can add new components or modify existing ones using `$`:

```{r}
tree$site = "plot_A"
tree$dbh = 26.0
tree
```

Lists are especially useful for grouping related results. For example, a model
run might contain its parameters, the time points, and the simulated population
in one object:

```{r}
model_run = list(
  parameters = c(N0 = 20, r = 1.1),
  time = 0:5,
  population = c(20, 22, 24.2, 26.62, 29.282, 32.2102)
)

model_run$parameters
model_run$time
model_run$population
```

Lists are commonly used when a function needs to return several related results
at once. They can also contain other lists, allowing more complex objects to be
organized hierarchically.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Create a list called `survey` containing:

   - a character vector of three species
   - a numeric vector of their masses
   - a factor vector indicating their sites
   - a logical value indicating whether the survey was complete

2. Inspect the list with `str` and `names`.

3. Select the mass of the second species and calculate the mean mass.

4. Add the survey date as a new component.

5. Create a list called `model_run` with a named vector of parameters, a vector
   of time points, and a vector of population sizes. Compare
   `model_run["population"]` with `model_run[["population"]]`.

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

```{r}
survey = list(
  species = c("Daphnia magna", "Gammarus pulex", "Asellus aquaticus"),
  mass = c(3.2, 5.1, 2.8),
  site = factor(c("A", "B", "A")),
  complete = TRUE
)

str(survey)
names(survey)

survey$mass[2]
mean(survey$mass)

survey$date = "2026-09-27"
survey
```

```{r}
model_run = list(
  parameters = c(N0 = 20, r = 1.1),
  time = 0:3,
  population = c(20, 22, 24.2, 26.62)
)

model_run["population"]       # returns a list
model_run[["population"]]     # returns the population vector
```

::::
:::::
::::::

# Data frames

Vectors can only hold one type of data, but an ecological dataset typically mixes numbers
(measurements) with text (species names) and categories (treatments, sites). The data
structure for this is the **data frame**: a table where each column is a vector (of
possibly different type) and each row is an observation. Most of the data you will load
into `R` will be stored as a data frame.

## Creating a data frame

You create a data frame with the function `data.frame`, giving each column a name:

```{r}
trees = data.frame(
  species = c("Quercus robur", "Fagus sylvatica", "Quercus robur", "Pinus sylvestris"),
  dbh = c(12.5, 8.3, 15.1, 9.7),
  alive = c(TRUE, TRUE, FALSE, TRUE)
)
trees
```

You can inspect the structure of a data frame with a few useful functions:

```{r}
nrow(trees)       # number of rows
ncol(trees)       # number of columns
names(trees)      # column names
str(trees)        # structure: types of all columns at once
head(trees)       # first rows (useful for large datasets)
```

## Selecting parts of a data frame

The most common way to get a column is with the `$` operator, which returns it as a
vector:

```{r}
trees$dbh
mean(trees$dbh)
```

You can also index data frames with square brackets, but now you need two indices
separated by a comma: `[rows, columns]`. Leaving one of them empty selects everything
along that dimension:

```{r}
trees[1, 2]       # row 1, column 2
trees[1, ]        # all columns of row 1
trees[, 2]        # all rows of column 2
trees[, "dbh"]    # columns can also be selected by name
```

Just like with vectors, you can select rows with a logical condition. This is how you
will most often subset your data:

```{r}
trees[trees$dbh > 10, ]
trees[trees$species == "Quercus robur", ]
trees[trees$alive == TRUE, ]
```

## Modifying a data frame

You can add a new column simply by assigning to a new name with `$`:

```{r}
trees$basal_area = pi * (trees$dbh / 2)^2
trees
```

And you can modify existing columns in the same way, for example to convert units:

```{r}
trees$dbh = trees$dbh / 100   # from cm to m
```

:::::: {.callout-important title="Exercise" collapse="true"}

1. Create a data frame `birds` with the following data:

| species    | mass (g) | site |
|------------|------|------|
| Parus major | 18.5 | A    |
| Parus major | 17.9 | B    |
| Cyanistes caeruleus | 11.2 | A |
| Cyanistes caeruleus | NA   | B    |

Make `site` a factor.

2. How many rows and columns does the data frame have? Check its structure with `str`.

3. Compute the mean body mass, ignoring missing values.

4. Select all rows for *Parus major* and site "A".

5. Add a column `mass_kg` with the body mass in kilograms.

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
birds = data.frame(
  species = c("Parus major", "Parus major",
              "Cyanistes caeruleus", "Cyanistes caeruleus"),
  mass = c(18.5, 17.9, 11.2, NA),
  site = factor(c("A", "B", "A", "B"))
)
```

2.

```{r}
nrow(birds)
ncol(birds)
str(birds)
```

3.

```{r}
mean(birds$mass, na.rm = TRUE)
```

4.

```{r}
birds[birds$species == "Parus major", ]
birds[birds$site == "A", ]
```

5.

```{r}
birds$mass_kg = birds$mass / 1000
birds
```

::::
:::::
::::::

# Visualization

Visualizing your data is one of the first things you should do in any analysis: plots
reveal patterns, outliers and data entry errors that summary statistics can hide. `R` has
two main systems for making plots:

- **Base R graphics**: the plotting functions that come built into `R`. They are quick to
use and ideal for *exploratory* plots: you type one command and immediately see your data.
They also work with different data types (vectors, matrices, data frames) and are very
flexible, but the syntax is not always consistent and it can be hard to make complex plots.

- **ggplot2**: a package that implements a systematic "grammar of graphics". Plots are
built by adding components layer by layer. It requires a bit more code and are meant to work
with data frames, but produces publication-quality figures and makes complex plots (groups,
facets, etc.) much easier.

If you want to make sense of a complex dataset, I recommend to use `ggplot2` from the start
as it is more productive. When you are working with simulations or ad-hoc calculations, base
R graphics will work better. But at the end of the day, it is a matter of personal preference
and you should use whatever works best for you (the more you use it, the better you'll get).

Make sure the `ggplot2` package is loaded (you may have to install first!):

```{r}
library(ggplot2)
```

For the examples we will use the built-in `iris` dataframe (these are datasets loaded in the
background, you bring them to your session by just referring to their names). You can
summarise the content of a dataframe with `summary`:

```{r}
summary(iris)
```
This represents biometric measurements on flowers from three species.

## Base R graphics

All base R plots are made with a single function call, and the plot appears immediately in
the Plots panel.

### Scatter plot

The workhorse function is `plot`, which shows the relationship between two numeric
variables. They use the *formula* syntax `y ~ x` (read as "y as a function of x").
You can assign groups to color (`col`) and point shape (`pch`) if they are stored
as factors and you convert them to numbers:

```{r}
plot(Sepal.Length~Petal.Length, data = iris, col = as.numeric(Species), pch = as.numeric(Species))
```

You can customize the axis labels with `xlab` and `ylab`:

```{r}
plot(Sepal.Length~Petal.Length, data = iris,
     col = as.numeric(Species), pch = as.numeric(Species),
     xlab = "Petal Length (cm)",
     ylab = "Sepal length (cm)")
```

### Histograms

To see the distribution of a single numeric variable, use `hist`:

```{r}
hist(iris$Petal.Length, xlab = "Petal length (cm)")
```

### Boxplots

Boxplots are the standard way to compare a numeric variable across groups. You also use a
formula but the `x` is always the grouping variable:

```{r}
boxplot(Petal.Length ~ Species, data = iris, ylab = "Petal length (cm)")
```

### Bar plots

Bar plots display counts or values per category. For example, we can plot the number
of values per species (we can do this with the `table` function):

```{r}
counts = table(iris$Species)
barplot(counts, ylab = "Number of flowers")
```

Base R plots can be customized extensively with many arguments (colors, point
shapes, axis limits...), but every customization requires you to know the right argument.
The easiest way to learn is to just ask an AI too (see note on AI usage early on) or a search
engine to find a blog.

Insert this after the **Bar plots** subsection and before the paragraph beginning “Base R plots can be customized…”:

### Vertical-line and staircase plots

**TODO**: Introduce this in Chapter 2
**TODO**: Turn into a Programming Card

The `type` argument of `plot` controls how observations are displayed in a base R plot.
For example, `type = "h"` draws a vertical line from the x-axis to each
observation. We use this to visualize the empirical distribution of a sample of discrete
data (when the sample size is large this will resemble the probability mass of the true
distribution):

```{r}
set.seed(123)
N = 100
sample_values = rpois(N, 5)
frequency = as.numeric(table(sample_values)/N)

plot(1:10, dpois(1:10, 5), type = "h",
     xlab = "Observation",
     ylab = "Probability mass")
points(frequency, col = 2, pch = 16)
```

**TODO**: Introduce this in Chapter 2
**TODO**: Turn into a Programming Card

A staircase plot uses `type = "s"`, this is useful to visualize empirical cumulative
distribution of a sample (when the sample size is large this will resemble the probability
mass of the true distribution):

```{r}
plot(cumsum(frequency),
    type = "s",
     xlab = "Value",
     ylab = "Cumulative proportion",
     main = "Empirical cumulative distribution")
```

The details of empirical cumulative distributions will be introduced later.

### Plotting functions with `curve`

The `curve` function provides a quick way to visualize a mathematical function.
The function is given as an expression, together with the range of values over
which it should be plotted.

For example, we can plot a logistic growth curve by specifying the initial
population size `N0`, carrying capacity `K`, and growth rate `r`:

```{r}
N0 = 10
K = 100
r = 0.4

curve(K / (1 + ((K - N0) / N0) * exp(-r * t)),
      from = 0, to = 20,
      xname = "t",
      xlab = "Time",
      ylab = "Population size",
      main = "Logistic growth")
```

Changing the parameter values changes the shape of the curve:

```{r}
r = 0.2

curve(K / (1 + ((K - N0) / N0) * exp(-r * t)),
      from = 0, to = 20,
      xname = "t",
      xlab = "Time",
      ylab = "Population size",
      main = "Logistic growth with a lower growth rate")
```

### Pairwise plots with `pairs`

The `pairs` function produces a grid of plots showing the relationships between
several variables. This is useful for quickly exploring correlations in a
dataset.

The first four columns of the built-in `iris` dataset contain numeric
measurements:

```{r}
pairs(iris[, 1:4],
      col = as.numeric(iris$Species),
      pch = 19,
      main = "Pairwise relationships among iris measurements")
```

Each off-diagonal panel shows the relationship between two variables. The
diagonal panels show the distribution of each variable. In this example, the
point colors correspond to the three species.

## ggplot2 graphics

`ggplot2` takes a different approach. Instead of one function with many arguments, you
build a plot by *adding layers* with the `+` operator. Every plot has three basic
ingredients:

1. **Data**: a data frame, given to the `ggplot` function.
2. **Aesthetics** (`aes`): the mapping from columns of the data to visual properties
   of the plot (x and y position, color, shape...).
3. **Geometry** (`geom_*`): the type of graphical object to draw (points, bars, boxes...).

The basic template is always the same:

```{r}
#| eval: false
ggplot(data, aes(x = ..., y = ...)) +
  geom_...()
```

Inside the function `aes` you can map columns to the `x` and `y` axis, the `color`, `shape`
and `linetype` arguments, etc. This may feel verbose for simple plots, but the payoff is that
all plots, from a simple scatter plot to a complex multi-panel figure, follow the same logic.

Learning how to master `ggplot2` is a long-term goal and here I am just giving a very short
crash course (you will learn more throughout the course). If you need more context and
details to undersand it, I suggest to look here: https://ggplot2.tidyverse.org/ or use
AI/search engines as indicated before.

### Scatter plot

We create scatter plots with the layer `geom_point`:

```{r}
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length, color = Species, shape = Species)) +
  geom_point()
```

Labels are added as another layer with `labs`:

```{r}
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length, color = Species, shape = Species)) +
  geom_point() +
  labs(x = "Sepal Length (cm)",
       y = "Petal Length (cm)")
```

### Histograms

We create histograms with the layer `geom_histogram`:

```{r}
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(bins = 10) +
  labs(x = "Petal length (cm)")
```

The `bins` argument controls how many bars the data is split into; try changing it to see
how it affects the impression the plot gives.

### Boxplots

We create boxplots with the layer `geom_boxplot`:

```{r}
ggplot(iris, aes(x = Species, y = Petal.Length)) +
  geom_boxplot() +
  labs(y = "Petal length (cm)")
```

### Bar plots

`geom_bar` counts the observations in each category automatically, so no `table` is
needed:

```{r}
ggplot(iris, aes(x = Species)) +
  geom_bar() +
  labs(y = "Number of flowers")
```
If you want to plot something else (e.g., a `mean`) you should calculate it first
(see below section on Data Transformation) and use `geom_col` instead.

:::::: {.callout-important title="Exercise" collapse="true"}

Use the `ChickWeight` data frame that comes with R. This represents the weight of individual
chickens as a function of time (days) and given different diets (column `Chick` identifies the
individual chickens):

```{r}
summary(ChickWeight)
```

1. Using base R, create:
   * a scatter plot of body weight as a function of time, color by diet
   * a histogram of body weight on day 10, edit the X axis title
   * a boxplot of final (day 21) obdy weight per diet

2. Recreate the same three plots with `ggplot2`. Which version do you prefer for each
plot, and why?

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
# The scatter plot
plot(weight~Time, data = ChickWeight, col = as.integer(Diet),
     ylab = "Weight (g)", xlab = "Time (days)")
# The histogram
hist(ChickWeight$weight[ChickWeight$Time == 10], xlab = "Weight (g)")
# The boxplot
boxplot(weight~Diet, data = ChickWeight[ChickWeight$Time == 21,])
```

2.

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight, col = Diet)) +
  geom_point()

ggplot(data = ChickWeight[ChickWeight$Time == 10,], aes(x = weight)) +
  geom_histogram()

ggplot(data = ChickWeight[ChickWeight$Time == 21,], aes(x = Diet, y = weight)) +
  geom_boxplot()
```
::::
:::::
::::::

# Control flow

So far our scripts execute one command after another, from top to bottom. For real analysis
and simulation work we need more flexibility: sometimes we want to execute code *only if a
condition is met*, and sometimes we want to *repeat* the same code many times. This is
called **control flow**.

## Conditional branching

The `if` statement executes a block of code only when a condition (a logical value) is
`TRUE`. The block is delimited by curly braces `{ }`:

```{r}
x = 4

if (x > 0) {
  sqrt(x)
}
```

If the condition is `FALSE`, the block is simply skipped and nothing happens. An optional
`else` block runs instead when the condition is `FALSE`:

```{r}
x = -4

if (x > 0) {
  sqrt(x)
} else {
  print("x is negative: cannot take the square root")
}
```

You can chain several conditions with `else if`. This is useful to classify values into
categories, for example size classes of a tree:

```{r}
dbh = 35

if (dbh < 10) {
  size_class = "seedling"
} else if (dbh < 30) {
  size_class = "sapling"
} else {
  size_class = "adult"
}
size_class
```

Only the *first* condition that is `TRUE` is executed, so the order matters.

Note that the condition of an `if` statement must be a *single* logical value. If you want
to test many values at once (a vector), use `ifelse` (several ones can be nested):

```{r}
dbh = c(35, 9, 22)
ifelse(dbh < 10,  # First argument is condition
       "seedling", # Second is the value returned if true
       ifelse(dbh < 30, # Third is the value returned if false (can be nested ifelse)
              "sapling",
              "adult"))
```

## Iterations

A `for` loop repeats a block of code for each element of a vector (the **index variable**
takes each value in turn):

```{r}
for (i in 1:5) {
  print(i)
}
```

A very common pattern in simulation work is: create an empty vector to store the results,
then fill it element by element inside the loop:

```{r}
n = 10
squares = numeric(n)      # an empty numeric vector of length n

for (i in 1:n) {
  squares[i] = i^2
}
squares
```

Note that we created the result vector *before* the loop with the right size using `numeric(n)`.

## When not to use a loop: vectorization

Loops are essential for simulations, where each iteration depends on a new random draw or
depends on a previous value. But for many tasks that *look* like they need a loop, `R` already
offers a vectorized alternative that is shorter and faster. Remember that arithmetic and
most functions work element by element on vectors:

```{r}
# with a loop (works, but unnecessary)
squares = numeric(10)
for (i in 1:10) {
  squares[i] = i^2
}

# vectorized: one line
squares = (1:10)^2
```

Note that there are also specific functions to avoid for loops in R, such as the `apply`
family of functions or functional maps. You may use them if you want but many early R
programmers find it a bit obscure and difficult to understand and this can be a source of
bugs. Writing for loops when needed will make your code much more readable and explicit and
I recommend you start there even if seasoned R programmers sneered at that.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Write an `if`-`else` statement that takes a variable `count` (number of individuals
observed) and prints `"no individuals"` if it is 0, `"few individuals"` if it is between
1 and 5, and `"many individuals"` if it is larger than 5. Test it with different values.

2. A population grows geometrically: $N_{t+1} = r \cdot N_t$, with growth rate
$r = 1.1$ and initial population size $N_0 = 20$. Use a `for` loop to simulate the
population size over 50 years, storing the size of each year in a vector (remember to
create the vector before the loop and to include $N_0$ in it). Plot the result.
This loop cannot be vectorized — do you see why?

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
count = 3

if (count == 0) {
  print("no individuals")
} else if (count <= 5) {
  print("few individuals")
} else {
  print("many individuals")
}
```

2.

```{r}
n_years = 50
r = 1.1

N = numeric(n_years + 1)
N[1] = 20                 # initial population size

for (t in 2:(n_years + 1)) {
  N[t] = r * N[t - 1]     # each year depends on the previous one: cannot vectorize
}

plot(0:n_years, N, type = "l",
     xlab = "Year", ylab = "Population size")
```
::::
:::::
::::::

# Creating your own functions

So far we have only used functions that come with `R` or with a package. But you can also
define **your own functions**. This is essential in this course: when we fit models to
data, we will have to write our own functions.

A function is defined with the keyword `function`, followed by the arguments in
parentheses and the code to execute (the *body*) in curly braces. They follow this
pattern

```{r}
my_function = function(argument1, argument2) {
  # code that does something with the arguments
  result = ...
  return(result)
}
```

- The **arguments** are the inputs of the function. Inside the body they behave like
normal variables.
- The **body** is any `R` code, which can use the arguments and create new variables.
- The `return` statement defines the **output** of the function: the value you get when
you call it.

For example, here is a function that computes the basal area of a tree from its diameter:

```{r}
basal_area = function(dbh) {
  area = pi * (dbh / 2)^2
  return(area)
}
```

Once defined, you use your function exactly like any built-in function:

```{r}
basal_area(10)
basal_area(c(10, 20, 30))   # it works on vectors too
```

Note that variables created *inside* a function (like `area`) only exist while the
function is running: after the call, `area` is destroyed. This is a good
thing: it means the code inside a function cannot accidentally mess up the rest of your
script. The only way to get information out of a function is through its return value.

Functions can take several arguments, and arguments can have **default values** exactly
like in the R functions we have used. For example, here is the exponential population growth
model from the Control flow section, wrapped into a reusable function:

```{r}
population_growth = function(N0, r, years) {
  N = numeric(years + 1)
  N[1] = N0
  for (t in 2:(years + 1)) {
    N[t] = r * N[t - 1]
  }
  return(N)
}
```

Now we can simulate different scenarios with a single line each:

```{r}
population_growth(N0 = 20, r = 1.1, years = 50)
population_growth(N0 = 20, r = 0.9, years = 50)
```

Wrapping your model in a function is a habit worth developing early: it makes it trivial
to run the model with different parameter values (something we will do constantly when
estimating parameters), and it keeps your script readable.

:::::: {.callout-important title="Exercise" collapse="true"}

1. Write a function `sphere_volume` that takes the radius of a sphere and returns its
volume ($V = \frac{4}{3}\pi r^3$). Test it with $r = 1$ and with a vector of radii.

2. Write a function `se` that computes the standard error of the mean of a vector,
$\text{se} = \text{sd}(x) / \sqrt{n}$. Test it with `c(3.2, 4.1, 2.8, 3.9, 3.5)`. What
happens if the vector contains an `NA`? Add the argument `na.rm = TRUE` to your function
and pass it on to `sd` and `length` appropriately (hint: for the sample size you will
need `sum(!is.na(x))` when `na.rm` is `TRUE`).

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
sphere_volume = function(r) {
  V = (4 / 3) * pi * r^3
  return(V)
}
sphere_volume(1)
sphere_volume(c(1, 2, 3))
```

2.

```{r}
se = function(x, na.rm = FALSE) {
  if (na.rm) {
    n = sum(!is.na(x))
  } else {
    n = length(x)
  }
  return(sd(x, na.rm = na.rm) / sqrt(n))
}
se(c(3.2, 4.1, 2.8, 3.9, 3.5))
se(c(3.2, 4.1, NA, 3.9, 3.5))               # NA
se(c(3.2, 4.1, NA, 3.9, 3.5), na.rm = TRUE) # works
```
::::
:::::
::::::

# Loading data

In practice you will rarely type your data into `R` by hand: you will read it from a file,
usually a text file (like `.csv`) or an Excel file (`.xlsx`). The result is always a data
frame.

To illustrate this we will use the dataset `seedpred.dat`, from a seed predation
experiment by Duncan & Duncan (2000). In the experiment, seeds of 8 plant species were
exposed to seed predators and checked repeatedly over time. Make sure the file is inside
your RStudio project folder (you can check this in the Files panel: if you see the file
there, `R` will find it).

## Text files (csv and tab-separated)

The function `read.csv` reads comma-separated text files which is the most common format:

```{r}
seedpred = read.csv("seedpred.csv", header = TRUE)
```

- The first argument is the **path to the file**. A simple filename works because the file
  is in the project folder; otherwise you would need to give the full path.
- `header = TRUE` tells `R` that the first line contains the column names.

Always check what you just read before doing anything else:

```{r}
head(seedpred)
str(seedpred)
```

The dataset has 5 columns: `species` (the plant species), `tcum` (cumulative time since
the start of the experiment), `tint` (time since the last check), `remaining` (seeds
remaining at the check) and `taken` (seeds taken during the interval). Note that `species`
was read as a character column; we can convert it to a factor:

```{r}
seedpred$species = factor(seedpred$species)
levels(seedpred$species)
```
# Transforming data

Once loaded, data is rarely in exactly the form you need for analysis or plotting. Common
operations are: (i) selecting rows that meet a condition, (ii) adding derived columns, and
(iii) computing summaries per group. The `dplyr` package offers a set of functions with a
consistent and readable syntax that has become the standard for this kind of work. Install it
and load it:

```{r}
library(dplyr) # Install it first!
```

## Filtering rows

`filter` selects rows that meet a condition (equivalent to logical indexing):

```{r}
seedpred_taken = filter(seedpred, taken > 0)
head(seedpred_taken)
cor_taken = filter(seedpred, species == "cor" & taken > 0)
head(cor_taken)
```

## Adding columns

`mutate` adds new columns (or modifies existing ones), computed from other columns:

```{r}
seedpred = mutate(seedpred, prop_taken = taken / (remaining + taken))
head(seedpred)
```

## Grouped summaries

A very powerful combination is `group_by` followed by `summarise`: it splits the data
frame into groups (here, one per species) and computes a summary for each group. For
example, the total number of seeds taken per species:

```{r}
by_species = group_by(seedpred, species)
summarise(by_species, total_taken = sum(taken))
```

You can compute several summaries at once:

```{r}
summarise(by_species,
          total_taken = sum(taken),
          n_checks = n(),                 # number of rows in the group
          mean_remaining = mean(remaining))
```

## Sorting

You can sort your dataframe by any column or combination of columns using `arrange`:

```{r}
summary_pred = summarise(by_species,
                         total_taken = sum(taken),
                         n_checks = n(),                 # number of rows in the group
                         mean_remaining = mean(remaining))
arrange(summary_pred, total_taken)
```

By default it will sort in increasing order. If you want a decreasing order just use the
negative of the column:

```{r}
arrange(summary_pred, -total_taken)
```

These few functions (`filter`, `mutate`, `group_by`, `summarise`, `arrange`) cover most of
the data manipulation you will need in this course. If we want to learn more, you may take a
look at https://dplyr.tidyverse.org/.

:::::: {.callout-important title="Exercise" collapse="true"}

1. How many observations are there per species in `seedpred` dataset?

2. How many seeds were taken in total during the experiment? Which species lost the most
seeds?

3. Create a data frame with only the observations of species `"psd"` where predation
occurred. Use it to make a scatter plot of seeds taken against cumulative time.

4. Compute, for each species, the proportion of checks in which at least one seed was
taken. Which species is most and least attractive to seed predators?

::::: {.content-hidden unless-meta="show_solution"}
:::: {.callout-tip title="Solution" collapse="true"}

1.

```{r}
table(seedpred$species)

# or with dplyr:
summarise(group_by(seedpred, species), n = n())
```

2.

```{r}
sum(seedpred$taken)

summarise(group_by(seedpred, species), total_taken = sum(taken))

```
The species "cd" lost the most seeds, "uva" the fewest.

3.

```{r}
psd = filter(seedpred, species == "psd" & taken > 0)
plot(taken ~ tcum, data = psd,
     xlab = "Time (days)", ylab = "Seeds taken")
```

4.

```{r}
summarise(group_by(seedpred, species),
          prop_checks = mean(taken > 0))
```

Species "cd" and "pol" are predated most often, while "psd" is rarely touched.

::::
:::::
::::::

# R documentation

R has a help system, meant for reference or reminding you how to use a function. This is useful
if you already know which function to use, but you need to check some details.

You can get help on any R function by entering a `?` followed by the function name:

```{r}
#| eval: false
?plot
```

A help page for an R function always has the same structure.

  - *Description*: Brief description of what the function does

  - *Usage*: Name of the function with the different arguments and default values (if any)

  - *Arguments*: Detailed description of each argument of a function

  - *Details*: Further details on what the function does. This may include further details on arguments as well as technical details concerning statistics or methodology

  - *Value*: A description of the type of data returned by the function, specially for complex objects

  - *See also*: A list of related functions (e.g. similar or related methods)

  - *Examples*: A short R script illustrating how to use the function

# Using AI to learn R (optional)

## Example 1: Deal with warnings and errors

Imagine we are adding two vectors and we are getting a warning

```{r}
#| eval: false
a = c(1,2,3)
b = c(4, 5)
a + b
```

You can safely ignore warnings **if you understand what they mean**. Otherwise treat them
as errors. Assuming you don't what this is, let's ask an AI. I type this prompt to the AI:

::: {.callout-note title="Prompt" collapse="true"}
I am learning how to program in R. I just had an exercise with code that looks like this:

```{r}
#| eval: false
a = c(1,2,3)
b = c(4, 5)
a + b
```

And I got this warning: "In a + b : longer object length is not a multiple of shorter object length"

I am not sure what it means and why it was generated. I suspect I did something wrong and I am not getting the result that I want to (even though the code runs). Could you help me understand:

1. Why was the warning triggered.
2. How can I fix it.

Note that I do not just care about the solution to the problem but I want to understand why it happens and learn any R related concept that is relevant so that I can be a better programmer.

Please be concise in your answer.
:::

Notice several things:

1. I include code within *fenced quotes* (the "```") which tells the AI that what I type
inside those quotes is code.

2. I also quote the warning message to make sure it understand this is a literal quote.

3. I give it context (I am student, I am learning R).

4. I do not just ask for a solution to my problem but explain why it is happening so that
I can actually learn.

5. I am very explicit about what I want to get back (in bullet points).

6. Finally I ask the AI to be concise. You may or may not need to do this depending on which
AI you are using and your own preferences. I find they all tend to talk too much...

This turns my interaction with the AI into an opportunity to learn. An AI can be an amazing
personalized tutor (especially at this level). The answer that it give is actually very good,
I do not think I could do much better!

::: {.callout-note title="Answer" collapse="true"}
Yes — this is an important R concept called **vector recycling**.

**1. Why did the warning occur?**

In R, `+` operates **element-by-element**:

```{r}
#| eval: false
a + b
```

means roughly:

```text
a:  1  2  3
b:  4  5
    ↓  ↓  ↓
    5  7  ?
```

But `b` only has two elements. R therefore **recycles** `b`:

```text
a:  1  2  3
b:  4  5  4
    ↓  ↓  ↓
    5  7  7
```

So R actually returns:

```{r}
#| eval: false
[1] 5 7 7
```

The warning occurs because the lengths are **not compatible for clean recycling**:

```{r}
#| eval: false
length(a)  # 3
length(b)  # 2
```

R's recycling rule is:

> If the longer vector's length is a multiple of the shorter vector's length, recycling happens without a warning.

For example:

```{r}
c(1, 2, 3, 4) + c(10, 20)
```

gives:

```text
11 22 13 24
```

because 4 is a multiple of 2.

But 3 is **not** a multiple of 2, so R warns you that it is doing something potentially unintended.

**2. How do you fix it?**

That depends on what you intended.

If you wanted to add only corresponding elements, make the vectors the same length:

```{r}
#| eval: false
a = c(1, 2)
b = c(4, 5)

a + b
# 5 7
```

If you intended `b` to repeat, make that explicit:

```{r}
#| eval: false
a + rep(b, length.out = length(a))
# 5 7 7
```

**The important R concept**

**R is vectorized.** Operations such as `+`, `-`, `*`, `/`, and many functions operate on whole vectors rather than requiring loops:

```{r}
#| eval: false
c(1, 2, 3) * 10
# 10 20 30
```

And when vector lengths differ, R uses **recycling**. This is very convenient, but it can also silently produce results you didn't intend.

A good programming habit is therefore: **when you see a recycling warning, don't just suppress it — check whether you actually intended the recycling.**
:::

## Example 2: Learning a new feature

We have learned how to make scatterplots with ggplot2:

```{r}
#| eval: false
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length, color = Species, shape = Species)) +
  geom_point()
```

I can see that there are fairly linear responses between Sepal Length and Petal Length. I
would like to add some trendlines to the plot describing those linear trends. I have no idea
how. I could search online, read some blog or book, or I can ask AI.


::: {.callout-note title="Prompt" collapse="true"}
I am a student learning R and I am interesting in becoming better at using ggplot2. I have
produced a scatterplot of Sepal Length vs Petal Length for different species. Here is my
current code:

```{r}
#| eval: false
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length, color = Species, shape = Species)) +
  geom_point()
```

I can see three clear clusters of data (one per species) and also a strong linear relationship
between the two variables for each cluster. I would like to add a linear trendline for each
cluster to visualize better this relationship. I would like to know:

1. What is the procedure to add linear trends to an existing scatter plot in ggplot2 with
differents groups? Is there a layer to do that or do I need to fit a model first?

2. How would it apply to my example

Note that my intent is to learn how to do this by myself in the future, I do not want just a
solution specific to this problem.

Please be concise in your answer.
:::

We use a similar strategy to before but notice that here I am also using my basic knowledge
of ggplot2. Because of how this package works, I suspect (or rather hope) that there is a
layer (like `geom_point`) that actually does the job. Something like `geom_trendline` or so.
I also specify that I would rather not fit a model just to show the trendline. The rest is
as before, I want to make sure that the answer I get is not "AI does my homework" but rather
"I am going to learn something new".

::: {.callout-note title="Prompt" collapse="true"}
Yes—`ggplot2` has a layer specifically for this, so you **do not need to fit the models yourself first**.

**1. General procedure**

The layer is:

```{r}
#| eval: false
geom_smooth(method = "lm")
```

`geom_smooth()` fits a model and draws the fitted line. The important idea for grouped data is:

> **If you map a grouping variable to an aesthetic such as `color`, `ggplot2` fits a separate model for each group.**

So the general pattern is:

```{r}
#| eval: false
ggplot(data, aes(x = x_variable, y = y_variable, color = group)) +
  geom_point() +
  geom_smooth(method = "lm")
```

You can think of this as:

**data → points → group-wise model fitting → fitted lines**

**2. Your example**

You already have `Species` mapped to `color`, so simply add the layer:

```{r}
#| eval: false
ggplot(iris, aes(x = Petal.Length,
                 y = Sepal.Length,
                 color = Species,
                 shape = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

`ggplot2` will fit **three separate linear models**, one for each species, and draw the corresponding lines.

A useful thing to experiment with is:

```{r}
#| eval: false
geom_smooth(method = "lm", se = FALSE)
```

Here `se = FALSE` removes the shaded confidence interval, which can make the three trendlines easier to see.

**Key principle to remember:** `geom_smooth()` does the model fitting for you, and the grouping comes from your aesthetic mappings (here, `color = Species`).
:::

## Example 3: Why the human still matters

As an example of why the human still matters, I did a little test. I gave the same AI as in
previous exercises the `seedpred` dataset and simply ask "Can you tell which species was predated the most? And the least?".

AI has advanced so much that it actually loaded the data, wrote a little Python script to
analyze the data, generated a table of results and interpreted the results correctly, saying
that `cd` experience the highes predation and `uva` the least.

I check the numbers and realized that what the AI had done is add up the number of seeds
taken for each species (it did not explain it to me, I had to check the numbers). But is
that we meant? If you look at the earlier exercise, we also calculated the proportion of
instances when at least one seed was taken and the results are different.

It turns that  predation data is a bit more complicated as it usually involves two processes:
(i) the predator visits the location with seeds and (ii) the predator consumes the seeds. So
what do we mean when we ask "most predated"?  Also, is it enough to just do a simple
calculation from the data? Are we going to get the same result if we repeat the experiment?
What kind of statistics is relevant here?

Turning an ecological question into an acceptable scientific answer requires understanding
ecology, statistics and often it also requires figuring out what we really want to know.
AI understands nothing and generally does not ask, it just assumes and runs with it. We still
matter :).
