---
title: "Introduction to Probability (Discrete)"
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures in this chapter.
library(ggplot2)
```

# Introduction

In this chapter, we learn how to introduce probability distributions as the central pillar
to build out statistical  models. We focus on count data first. You may count, for example,
surviving seedlings, flowering plants, or wildlife camera detections, but a count from one
location or night is never the whole ecological story. We represent these counts using
**discrete random variables** which take separate, countable values such as $0, 1, 2, \ldots$.
[Chapter 3](../Chapter_3/Theory.qmd) extends these ideas to continuous measurements

# Learning goals

In this chapter, we will cover:

- discrete random variables and probability;
- the four faces of a distribution (mass, cumulative, quantile and samplers);
- the assumption of independence;
- means and variances from distributions and samples;
- how samples and distributions are related;
- how to simulate possible observations from a model;
- the concept of overdispersion.

We will learn more about the Binomial, Poisson and Negative Binomial distributions and
practice with them. Additional distributions are available in the [Supplement](../Supplements/distributions.qmd).

# Random variables and probability distributions

A **random variable** is the quantity whose value is uncertain before we make that observation.
For a count, its possible values are non-negative integers. We will use a hypothetical
experiment on seed survival to illustrate how to model ecological count data. Suppose we
randomly distribute several quadrats in a field and in each of them we place four seeds of
a target species. We come back after a pre-specified period of time and we measure how many
seeds survived inside the quadrat.

The random variable $X$ keeps track of the possible counts of seeds that would survive
in a quadrat. If the quadrat contains four seeds, then $X$ can take any integer value from
0 to 4. Mathematically, this is expressed as:

$$
X \in \{0,1,2,3,4\}.
$$

The **support** of a random variable is the set of values that the variable can take. In this
case $X$ can take a **finite** set of values. In other cases, there is no pre-specified maximum,
such as the number of spores or insects that could fall into a trap. In those cases we will
model count data with random variables that can take any positive integer value (i.e.,
$\{1, 2, 3, ...\}$, formally known as **countably infinite**). Technically, this accounts for an
infinite number of counts, though in practice the counts will always be finite and
constrained to the relevant order of magnitude.

::: {.callout-note title="Discrete random variable"}
A random variable is discrete when it takes values from a finite or countably infinite set.
:::

Note that $X$ does not keep track of the actual seeds that survived in a particular quadrat,
that is what we call an **observation** of the real world. Rather, $X$ represents how likely
we are to observe a particular number. That is measured by a **probability**. That is, a
random variable will have a probability distribution associated with it that assigns a value
between 0 and 1 to every possible observation or set of observations. For example,
it might assign these probabilities:

| $x$ | $P(X=x)$ |
|---:|---:|
| 0 | 0.05 |
| 1 | 0.15 |
| 2 | 0.30 |
| 3 | 0.30 |
| 4 | 0.20 |

We can also define the probability of any set of values by addition of individual probabilities.
For example, the probability that at least 4 seeds survive is:

$$
P(X \geq 4) = P(X = 4) + P(x = 5).
$$

The probability that any value occurs is the sum of all the individual probabilities, and
that sum must always be 1:

$$
\sum_{\forall x} P(X = x) = 1,
$$

where $\forall x$ means "for all possible values $x$".

Note that the capital letter $X$ refers to the random variable, while the lower-case $x$
refers to a possible observation. This distinction looks fussy at first, but it helps to keep
the model and the data separate once we start estimating parameters later in the course.

## What do these probabilities mean?

We saw above that we could assign probabilities to the different possible observations and
that probabilities obey some basic mathematical properties. What does it mean to say "the
probability that 2 seeds survive is 0.15"? There are two complementary perspectives that one
can take depending on how probability is being applied (see [Chapter 1](../Chapter_1/Theory.qmd)
for further details on applications of probability):

- **Epistemic probability**^[In some texts this is called subjective probability, but this
has an unwarranted pejorative connotation.]: If the goal is to predict the number of seeds that survive
in a specific quadrat, probability represents how certain we are about each possible
predicted value (e.g., in the table above I am twice as confident that the value will be 2
compared to 1). This relates to how much knowledge or information we have about the system
we are predicting.


- **Aleatoric probability**^[This is sometimes called objective probability, which implies
the other must be subjective.]: Probability can also represent the relative frequency at which
different counts will be observed across all the possible replications of the same experiment
or observation protocol. This relates to errors in measurements, and stochasticity and
variability in ecological systems.

For many reasons, this course will focus on aleatoric probability as implemented by the
Frequentist statistical paradigm[^3], so we will use probabilities to model the relative
frequencies at which different observations (or, in [Chapter 4](../Chapter_4/Theory.qmd),
different parameter estimates) can occur.

[^3]: Sometimes we may want to say something about uncertainty of predictions and then we will muddle the waters and step out of a strict Frequentist viewpoint and borrow some ideas from the competing camp, i.e., the Bayesian statistical paradigm. This is very common in practice (though it tends to work better with larger datasets). If the focus of your research is on forecasting and uncertainty quantification, you should do it in a Bayesian way and, instead of reading this book, read [Ecological Forecasting](https://press.princeton.edu/books/hardcover/9780691160573/ecological-forecasting?srsltid=AU7gw4WqqIymK1Rnltv3sGazFOGdtKm1mR_bg1rz7cSucqUn3c_8nAxz) by Michael Dietze.

## Probability distributions {#sec-four_faces}

When the set of possible observations is small we can keep track of probabilities in tables.
However, as the number of possible observations increases this is not feasible. Instead, we
use a **probability mass function** which is a formula that assigns a probability to every
possible $x$. From the mass function we can compute by countable addition the probability
of any possible set of observations, which is what we call a **probability distribution**.


::: {.callout-note title="Probability distribution"}
A probability distribution assigns probabilities to all possible sets of values of a random
variable. Its foundation is a probability mass function.
:::

There are many probability distributions that already exist, and we represent them by their
probability mass function. The oldest and most basic distribution for discrete data is the Binomial
distribution, discovered by Jacob Bernoulli in 1713[^2]. This distribution assumes that
we are observing a fixed number of events, each of which can have two values (0 or 1) and
each event has the same fixed probability of having the value 1. It makes sense to use this
distribution for our seed survival example: the event is whether a seed survives (1) or not
(0) and we assume all seeds have the same survival probability.^[The classic example is
flipping a coin a fixed number of times and counting the number of heads. Many classical texts
describe the binomial in terms of failures and successes, but this language has strange
connotations.]

The probability mass function of the Binomial is given by:

$$
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x},
$$

where $n$ is the number of events being observed, and $p$ is the probability of an event having the
value 1. I provide more details on this distribution (and others) later in the chapter.

[^2]: Bernoulli, J. (1713). *Ars Conjectandi*. Impensis Thurnisiorum, Fratrum.

Every distribution has four functions that are interrelated^[I call them the four faces; this
is terminology that I completely made up] @fig-four_faces:

- Probability mass;
- Cumulative probability;
- Quantile;
- Random sampler.

In R, all of these functions are implemented for each distribution. The name of each function
consists of a prefix letter followed by a shortened name for the distribution. The four faces of
the Binomial distribution are implemented by functions `dbinom`, `pbinom`, `qbinom` and `rbinom`,
respectively.

To illustrate the four faces, we will assume in our seed survival example that each seed has
a probability of survival of $0.4$ and that the variation in observations across different
quadrants follows a Binomial distribution. In R:

```{r}
#| eval: false
# size = n, prob = p in formula for Binomia
seed_survival_probability = 0.4
seed_count  = 4
dbinom(x = 2,   size = seed_count, prob = seed_survival_probability)
pbinom(q = 2,   size = seed_count, prob = seed_survival_probability)
qbinom(p = 0.5, size = seed_count, prob = seed_survival_probability)
rbinom(n = 10,  size = seed_count, prob = seed_survival_probability)
```

Note that the first argument (`x`, `q`, `p`, or `n`) is always the same for all distributions
in R. The other arguments are the parameters for a given distribution (in this case, `size`
is $n$ and `prob` is $p$ in the formula above).

::: {.callout-note title="The four faces of distributions in R"}
For a distribution in R:

- `d*` functions calculate probability mass
- `p*` functions calculate cumulative probabilities
- `q*` functions calculate quantiles
- `r*` functions generate random samples

The first argument is conventionally named `x`, `q`, `p`, or `n`, respectively.
:::

A **cumulative probability function** ($F_X$) is the result of adding probability masses up
to and including a value $x$ (`q` in `pbinom`):

$$
F_X(x)=P(X\leq x).
$$

For example, $F_X(3)$ is the probability that at most three seeds survive. The cumulative
probability is always between zero and one, and it never decreases as $x$ increases. We
will not use it much in this course^[it is the basis for calculating p-values, which we will
completely ignore in this course].

A **quantile function** is the inverse of the cumulative probability function. That is, we
supply a cumulative probability and ask at which value of $x$ it is reached. Because we are
dealing with discrete variables, we define the $\alpha$-quantile as the smallest $x$ whose
cumulative probability (`p` in `qbinom`) reaches or exceeds $\alpha$, which mathematically
is written as:

$$
q_\alpha=\min\{x:F_X(x)\geq\alpha\}.
$$

We will use quantiles to describe a distribution numerically (e.g., the median of a
distribution is the $0.5$ quantile) as complements to expectations and moments (see later).
We will also use quantiles to quantify parameter uncertainty as confidence intervals (see
[Chapter 4](../Chapter_4/Theory.qmd#sec-confidence_intervals)).


The fourth face is not so much a function as a procedure. A random sampler is a procedure
that generates possible samples from the distribution^[technically, independent and
identically distributed samples; this is explained later]. This means a series of possible
observations that are compatible with the distribution. For example, if we want to simulate
a collection of 10 quadrats from the example above, we would generate 10 random samples from
the Binomial distribution.

We will use random sampling throughout the course (we will call it **simulation**) as it is
a versatile way of analysing the statistical properties of a model being fitted to data. We
will discuss in more detail how samples relate to their distributions in @sec-samples.


```{r}
#| echo: false
#| label: fig-four_faces
#| fig-cap: "The four faces of a Binomial distribution with $n = 4$ and $p = 0.4$."
library(ggplot2)
library(patchwork)

# Distribution parameters and a chosen cutoff
size = 4
prob = 0.4
cutoff = 2

# Probabilities at each possible number of successes
binomial = data.frame(
  x = 0:size,
  probability = dbinom(0:size, size = size, prob = prob),
  cumulative_probability = pbinom(0:size, size = size, prob = prob)
)

# Cumulative probability at the chosen cutoff
p_cutoff = pbinom(cutoff, size = size, prob = prob)

# Panel 1: probability mass function
p1 = ggplot(binomial, aes(x = x, y = probability)) +
  geom_col(fill = "#34495e", width = 0.7) +
  geom_vline(xintercept = cutoff, linetype = "dashed", colour = "#e74c3c") +
  geom_point(
    data = subset(binomial, x == cutoff),
    colour = "#e74c3c",
    size = 3
  ) +
  annotate(
    "text", x = 2.15, y = 0.36,
    label = paste0(
      "P(X = ", cutoff, ") = ",
      round(dbinom(cutoff, size, prob), 2)
    ),
    colour = "#e74c3c",
    hjust = 0
  ) +
  scale_x_continuous(breaks = 0:size) +
  labs(
    title = "Probability mass function",
    x = "Count",
    y = "Probability"
  ) +
  theme_minimal()

# Panel 2: cumulative probability
p2 = ggplot(binomial, aes(x = x, y = probability)) +
  geom_col(
    aes(fill = x <= cutoff),
    width = 0.7,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = cutoff + 0.35, colour = "#2980b9") +
  annotate(
    "text", x = 3, y = 0.35,
    label = paste0(
      "P(X \u2264 ", cutoff, ") = ",
      round(p_cutoff, 2)
    ),
    colour = "#2980b9"
  ) +
  scale_fill_manual(values = c("TRUE" = "#85c1e9", "FALSE" = "grey80")) +
  scale_x_continuous(breaks = 0:size) +
  labs(
    title = "Cumulative probability function",
    x = "Count",
    y = "Probability mass"
  ) +
  theme_minimal()

# Panel 3: quantile function
probabilities = seq(0, 1, length.out = 1001)

quantiles = data.frame(
  p = probabilities,
  quantile = qbinom(probabilities, size = size, prob = prob)
)

p3 = ggplot(quantiles, aes(x = p, y = quantile)) +
  geom_step(colour = "#16a085", linewidth = 1.1) +
  geom_point(
    data = data.frame(p = p_cutoff, quantile = cutoff),
    colour = "#16a085",
    size = 3
  ) +
  annotate(
    "text", x = 0.58, y = 2.45,
    label = paste0(
      "qbinom(", round(p_cutoff, 2), ") = ", cutoff
    ),
    colour = "#16a085"
  ) +
  scale_y_continuous(breaks = 0:size) +
  labs(
    title = "Quantile function",
    x = "Cumulative probability",
    y = "Quantile"
  ) +
  theme_minimal()

# Panel 4: simulated random sample
set.seed(42)

sample_data = data.frame(
  x = rbinom(20, size = size, prob = prob)
)

# Calculate the relative frequency of each possible outcome
sample_frequency = data.frame(
  x = 0:size,
  relative_frequency = tabulate(
    sample_data$x + 1,
    nbins = size + 1
  ) / nrow(sample_data)
)

p4 = ggplot(sample_frequency, aes(x = x, y = relative_frequency)) +
  geom_col(fill = "#af7ac5", colour = "#8e44ad", width = 0.7) +
  scale_x_continuous(breaks = 0:size) +
  labs(
    title = "20 random samples",
    x = "Count",
    y = "Relative frequency"
  ) +
  theme_minimal()

(p1 | p2) / (p3 | p4)
```


When we simulate random samples in R, the computer does not observe a truly random
physical process. It uses a deterministic algorithm to generate a **pseudo-random
sequence**: a sequence designed to have the mathematical properties needed for random
sampling. This is enough for the probability simulations we use in this course, while also
making them reproducible.

A **seed** specifies the starting state of that algorithm. If you set the same seed before
running the same simulation, R produces the same pseudo-random sequence again. We set a seed
with `set.seed()`, usually once before a simulation or a batch of repeated simulations:

```{r}
#| eval: false
# Make a later random simulation reproducible.
set.seed(42)
```

Changing the seed changes the sequence. Always set the seed at the start of your script or
(even better) before any random sampling, to make sure that you generate reproducible
results.

:::: {.callout-important title="Practice: orchid fruit set"}

In a restoration plot, we tag six orchid flowers. Each of them has a probability of setting
fruit of $0.5$. Let $X$ be the number of these flowers that set fruit.

1. What values can $X$ take?
2. Use R to calculate (i) the probability that exactly four flowers set fruit and (ii) the
   probability that at most four set fruit, assuming a Binomial distribution.
3. What is the meaning of the second probability calculated in point 2?
4. Find the 0.75 quantile of $X$. In words, what does this quantile tell you?
5. Which R function prefix (`d`, `p`, `q`, or `r`) did you use for each calculation?

::: {.callout-tip title="Solution" collapse="true"}

The count can be any integer from 0 to 6. The exact-count probability uses `dbinom()`,
the cumulative probability uses `pbinom()`, and the quantile uses `qbinom()`.

```{r}
flower_count = 6
fruit_set_probability = 0.5

# Calculate an exact probability, a cumulative probability, and a quantile.
dbinom(x = 4, size = flower_count, prob = fruit_set_probability) # 0.23
pbinom(q = 4, size = flower_count, prob = fruit_set_probability) # 0.89
qbinom(p = 0.75, size = flower_count, prob = fruit_set_probability) # 4
```

The quantile tells me at least 75% of the probability is on counts of four or fewer, but note
that cumulative probability for four is 89%. The difference is due to the discrete nature of
the counts.

The cumulative probability means that if we repeat this 6-plant experimental unit many, many
times, 89% of them will have four flowers or less. Alternatively, it could mean that we are
89% sure that a particular 6-plant experimental unit will yield 4 flowers or less.
:::
::::

## Joint distributions and independence {#sec-joint-independence}

Two random variables are said to be **independent** if knowledge about one of them does not
change the probabilities for the other. Understanding this concept requires understanding
joint distributions of multiple random variables. Let's first illustrate it with a simple
example involving two binary random variables (for variety, I will use seed germination).

Let's consider germination of two planted seeds. Let $X_1$ and $X_2$ be random variables to
track germination (`1`) or not (`0`) for each of the seeds, after a pre-specified period of
time. The **joint distribution** of $X_1$ and $X_2$ assigns a probability to every combination
of values for $X_1$ and $X_2$ (this is known as the **product space** of $X_1$ and $X_2$).

| $X_2 \backslash X_1$ | $x_1 = 0$ | $x_1 = 1$ | Marginal $X_2$ |
|---|---:|---:|---:|
| $x_2 = 0$ | $P(X_1 = 0, X_2 = 0)$ | $P(X_1 = 1, X_2 = 0)$ | $P(X_2 = 0)$ |
| $x_2 = 1$ | $P(X_1 = 0, X_2 = 1)$ | $P(X_1 = 1, X_2 = 1)$ | $P(X_2 = 1)$ |
| Marginal $X_1$ | $P(X_1 = 0)$ | $P(X_1 = 1)$ | $1$ |


For example, $P(X_1 = 0, X_2 = 0)$ is the probability that neither seed germinates, whereas
$P(X_1 = 1, X_2 = 0)$ indicates that the first seed germinates. The original distributions
for $X_1$ and $X_2$ are the **marginal distributions** of the joint distribution because
you can recover them by adding over rows or columns. The marginal distribution of $X_2$ tells
us about the probability of the second seed germinating, regardless of what is happening to the
first seed. By the rule of addition, we just need to add all the joint probabilities where the
second seed is germinating.

Now, if $X_1$ and $X_2$ are independent, the probability of any particular combination is
the product of their marginal distributions, that is:

$$
P(X_1=x_1, X_2=x_2)=P(X_1=x_1)P(X_2=x_2).
$$

If we now assume $P(X_1 = 1) = 0.6$ and $P(X_2 = 1) = 0.3$, then the table becomes

| $X_2 \backslash X_1$ | $x_1 = 0$ | $x_1 = 1$ | Marginal $X_2$ |
|---|---:|---:|---:|
| $x_2 = 0$ | $0.4 \times 0.7 = 0.28$ | $0.6 \times 0.7 = 0.42$ | $0.70$ |
| $x_2 = 1$ | $0.4 \times 0.3 = 0.12$ | $0.6 \times 0.3 = 0.18$ | $0.30$ |
| Marginal $X_1$ | $0.40$ | $0.60$ | $1.00$ |

Real seeds often share a parent plant, a soil patch, a drought, or a hungry herbivore. In
general, lack of independence can arise through spatial clustering, shared weather, competition,
or repeated measurements of the same individual. [Chapter 8](../Chapter_8/Theory.qmd) explores
how to relax the assumption of independence to build more advanced models.

::: {.callout-warning title="Independence is a modelling assumption"}
Independence means that knowing the value of one random variable does not change the
probability distribution of another. It is not guaranteed merely because observations
were collected at different locations or times.
:::

#### Independent and identically distributed random variables

We often make a stronger working assumption: observations are both independent and **identically
distributed**, meaning that each follows the same probability distribution. We call this
**independent and identically distributed (i.i.d.)**. When we take repeated observations of
what we consider to be the same ecological phenomenon, we will treat these observations as
i.i.d. Also, the random samples that are generated by R functions such as `rbinom` are i.i.d.

Going back to the two-seed germination trial we just imagined, if the two seeds are
from the same species and origin, have been treated in the same manner and are buried in the same
soil under the same environmental conditions, we will have no reason to believe that their
behaviour differs. So we will treat them as i.i.d. If we then define $P(X_1 = 1) = P(X_2 = 1) = 0.3$
the table becomes:

| $X_2 \backslash X_1$ | $x_1 = 0$ | $x_1 = 1$ | Marginal $X_2$ |
|---|---:|---:|---:|
| $x_2 = 0$ | $0.7 \times 0.7 = 0.49$ | $0.3 \times 0.7 = 0.21$ | $0.70$ |
| $x_2 = 1$ | $0.7 \times 0.3 = 0.21$ | $0.3 \times 0.3 = 0.09$ | $0.30$ |
| Marginal $X_1$ | $0.70$ | $0.30$ | $1.00$ |

We will see that the i.i.d. assumption is going to be fundamental for the methods of parameter
estimation in [Chapter 4](../Chapter_4/Theory.qmd) and [Chapter 5](../Chapter_5/Theory.qmd).

## Expectations

An **expectation** is the procedure of taking a probability-weighted average of any quantity
calculated from a random variable. That is, for any function of a random variable $f(X)$, you
can define its expectation:

$$
E[f(x)] = \sum_x f(x) P(X=x).
$$

The **mean** of a distribution ($\mu$, also known as the **expected value**) is the expectation for
$f(X) = X$, that is:

$$
\mu = E[X] = \sum_x x P(X=x).
$$

We can compute the Binomial distribution of seed survival as:

```{r}
# Use the binomial probability mass to calculate a weighted average.
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
seed_probability = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
expected_value = sum(seed_values * seed_probability)
expected_value
```


The **variance** of a distribution ($Var(X)$) is the expectation for $f(X) = (X-E[X])^2$,
that is the expected squared distance from the mean:

$$
\operatorname{Var}(X) = E[(X-E[X])^2] = \sum_x P(X=x)(x-E[X])^2.
$$

The variance measures how widely a distribution spreads its probability across possible
values. For two distributions with comparable means, a smaller variance generally means
counts are more concentrated near the mean.

We can also calculate the variance for the Binomial model:

```{r}
# Weight each squared distance from the mean by its probability.
variance = sum((seed_values - expected_value)^2 * seed_probability)
variance
```

The **standard deviation** is the square root of the variance:

$$
\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.
$$

::: {.callout-note title="Mean, variance and standard deviation"}
The mean is a property of a probability distribution. It need not be an observable value.
For example, a quadrat cannot contain 1.6 surviving seeds, even
though that is the mean for the seed survival example. Variance measures the spread of a
distribution in squared units. Standard deviation is the square root of variance and has the
same units as the original variable, making it usually easier to interpret.
:::

:::: {.callout-important title="Practice: fruit set on orchid plants"}

Suppose we tag 16 orchid flowers (2 plants, 8 flowers per plant) and each of the flowers
has a probability of fruit set of 0.6 and we assume a Binomial model. Let $T$ be the total
number of flowers setting fruit across both plants and assume that the fruit-setting outcomes of the
16 flowers are independent events.

1. Find the probability of every possible value of $T$.
2. Calculate the mean and variance of $T$.
3. Give one ecological reason why independence among the flowers might be doubtful.

::: {.callout-tip title="Solution" collapse="true"}

The possible counts range from 0 to 16. First calculate the probability of each count,
then use the weighted-sum formulas for the mean and variance:

```{r}
# Calculate the probability mass for all possible totals.
fruit_count_values = seq.int(from = 0, to = 16)
fruit_probabilities = dbinom(
  x = fruit_count_values,
  size = 16,
  prob = 0.6
)

expected_fruit_count = sum(fruit_count_values * fruit_probabilities)
fruit_count_variance = sum(
  (fruit_count_values - expected_fruit_count)^2 * fruit_probabilities
)

expected_fruit_count
fruit_count_variance
```

Flowers on the same plant may share physiological conditions or pollinator
visits, making their outcomes dependent. The binomial calculation is conditional
on the stated independence assumption; [Chapter 8](../Chapter_8/Theory.qmd)
introduces models for grouped observations.
:::
::::

# Samples and distributions {#sec-samples}

So far we have been working with mathematical models: we calculate probabilities and
expectations of distributions. But in research we collect limited, often noisy, samples of
data and want to generalize from these observations to the broader context in which they
were collected. As discuss in Chapter 1, we use statistical models as approximations for the
data generating process. This means that we treat the observations from our experiments as
if they were random samples from our statistical model. In this section we learn in more
detail about this relationship as it will be the basis for [Chapter 4](../Chapter_4/Theory.qmd)

For example, we might want to learn about seed survival across a restoration site while observing
only some quadrats. We cannot quietly turn those observed quadrats into a claim about the whole
site. To make that connection, we need assumptions about how quadrats were selected and how their
counts vary from quadrat to quadrat. We model this data generating process with a distribution
that is (at least qualitatively) compatible with that process.

## From frequency to probability

For a sample of counts, the **relative frequency** of a value is the number of times it appears
divided by the sample size. Let's simulate ten quadrats from the binomial model used
earlier:

```{r}
# Simulate ten quadrats from the binomial model.
set.seed(seed = 123)
n_quadrats = 10
sample_survivors = rbinom(
  n = n_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
```

`table()` counts how often each value appears. Divide by the number of quadrats to
calculate relative frequencies:

```{r}
# Count each possible value and divide by the sample size.
seed_values = seq.int(from = 0, to = seed_count)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = count_survivors / length(sample_survivors)
relative_frequency
```

Some possible counts may be absent from a small sample. Setting all possible
values as `factor()` levels makes `table()` include those counts with frequency zero.

The relative frequencies lie between zero and one and sum to one. Compare them
with the probabilities predicted by the Binomial model that we use to approximate the
data generating process:

```{r}
# Put the sample frequencies beside the model probabilities for comparison.
data.frame(
  seed_count = seed_values,
  relative_frequency = as.numeric(x = relative_frequency),
  probability = dbinom(
    x = seed_values,
    size = seed_count,
    prob = seed_survival_probability
  )
)
```

The two columns differ because the sample contains only ten quadrats. That is not a failure
of the model: it is what sampling variation looks like. For small sample sizes, the relative
frequencies and model robabilities can differ greatly. However, for independent and identically
distributed (i.i.d.) observations, the relative frequency of each count approaches its model
probability as sample size grows. This is an application of the **weak law of large numbers**.

For example, let $I_i=1$ when the $i$th quadrat has exactly three surviving seeds and
$I_i=0$ otherwise. The relative frequency of three survivors is
$\bar{I}_n=(I_1+\cdots+I_n)/n$. For any positive margin $\varepsilon$,

$$
\lim_{n \to \infty}
P\left(\left|\bar{I}_n - P(X=3)\right| \ge \varepsilon\right)=0.
$$

In plain English, as $n$ grows, the chance that the relative frequency differs from the
model probability by at least $\varepsilon$ becomes small. There are two probabilities here:
the probability assigned to a count by the model, and the probability of seeing a
relative frequency that is far from the model one. This convergence is
not deterministic: a larger realised sample need not be closer to the model probability
than every smaller realised sample.

::: {.callout-note title="Frequencies converge to probabilities"}
For independent observations from the same distribution, the relative frequency of any
specified value converges in probability to its probability mass as sample size grows.
:::

We can push the simulation further and compare relative frequencies at sample sizes
from 10 to 100,000 to see this convergence in action:
```{r}
# Compare relative frequencies for several sample sizes with the model.
set.seed(123)
sample_sizes = 10^(seq_len(length.out = 5))
seed_values = seq.int(from = 0, to = seed_count)

all_frequencies = matrix(NA, nrow = length(seed_values),
                         ncol = length(sample_sizes) + 1)
colnames(all_frequencies) = c(as.character(sample_sizes), "theory")

for (i in seq_along(sample_sizes)) {
  sample_survivors = rbinom(
    n = sample_sizes[i],
    size = seed_count,
    prob = seed_survival_probability
  )
  count_survivors = table(factor(
    x = sample_survivors,
    levels = seed_values
  ))
  all_frequencies[, i] = as.numeric(count_survivors) / sample_sizes[i]
}
all_frequencies[, length(sample_sizes) + 1] = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
round(all_frequencies, 3)
```

One sequence of samples cannot show convergence directly. For example, the relative frequency
of three surviving seeds at 10,000 quadrats can happen to be closer to the model probability
than its value at 100,000 quadrats. To show what usually changes with sample size, repeat the
simulation 100 times at each size. This is shown in @fig-relative-frequency-convergence.
At small sample sizes, the repeated relative frequencies cover a wide range. As the sample
size grows, the relative frequencies concentrate around the probabilities of the model.
Individual samples can still generate values far from the line, but large deviations become
less likely.

```{r}
#| echo: false
#| eval: true
#| label: fig-relative-frequency-convergence
#| fig-cap: "Relative frequencies of exactly three surviving seeds for simulated samples if difference sizes. The horizontal line is the probability assigned by the Binomial distribution."
# Repeatedly simulate samples and retain the relative frequency of three survivors.
set.seed(seed = 20261003)
seed_count = 4
seed_survival_probability = 0.4
target_count = 3
sample_sizes = 10^(seq_len(length.out = 5))
n_repetitions = 100

repeated_frequencies = data.frame(
  sample_size = numeric(),
  relative_frequency = numeric()
)

for (sample_size in sample_sizes) {
  relative_frequency = numeric(length = n_repetitions)

  for (repetition in seq_len(n_repetitions)) {
    sample_survivors = rbinom(
      n = sample_size,
      size = seed_count,
      prob = seed_survival_probability
    )
    relative_frequency[repetition] = mean(sample_survivors == target_count)
  }

  repeated_frequencies = rbind(
    repeated_frequencies,
    data.frame(
      sample_size = sample_size,
      relative_frequency = relative_frequency
    )
  )
}

# Calculate the model probability that the figure uses as a reference line.
probability = dbinom(
  x = target_count,
  size = seed_count,
  prob = seed_survival_probability
)

ggplot(
  repeated_frequencies,
  aes(
    x = sample_size,
    y = relative_frequency,
    group = sample_size
  )
) +
  geom_point(width = 0.35, fill = "grey90", colour = "grey40",
             position = position_jitter(width = 0.05, height = 0.005)) +
  geom_hline(
    yintercept = probability,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  scale_x_log10(
    breaks = sample_sizes,
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    x = "Sample size (log10 scale)",
    y = "Relative frequency of three surviving seeds"
  ) +
  theme_classic()
```

:::: {.callout-important title="Practice: fruit-set frequency"}
In a sample of 24 tagged orchid flowers, 15 set fruit. Calculate the relative frequency
of fruit set. Explain what this value estimates and why it need not equal the model
fruit-set probability exactly.

::: {.callout-tip title="Solution" collapse="true"}

The relative frequency is $15/24=0.625$. It is not equal to the probability assigned by the
model to that count, but it is probably not too different for a sample of 24.
:::
::::

## Empirical distribution

Because of this relationship between the relative frequencies and probabilities, we often
define the **empirical distribution** of a sample. This is a procedure where we assign each
observed count its relative frequency and we treat it as if it were a probability mass
function @fig-ecdf. In principle it is just a way to tabulate a (potentially long) collection of
observations, but we will see that this also facilitates calculating properties of a sample
such as quantiles or expectations that we can related to the equivalent properties of the
model.

::: {.callout-note title="Empirical distribution"}
A distribution whose mass at each count equals the relative frequency in the sample.
:::

The empirical distribution is also important because, generally speaking^[this is not a
formal statistical statement, more of a heuristic derived from experience and common sense,
to build intuition], a sample which empirical distributions matches well the model
distribution will generate reasonable estimates of the model's parameters with relatively
low uncertainty. In this we would say that the "sample is representative" of the process we
are trying to model.This is why it is crucial to avoid introducing biases in the way we
collect data from experiment (motivating procedures such as randomization).

Like the any distribution, an empirical distribution has a cumulative function: the fraction
of observations at or below each count. When comparing the empirical and model cumulative
probability functions, we often use step-plots, which easily allow both functions side by
side @fig-ecdf. These types of visualisations are useful if you want to compare how well a
distribution matches a given sample of data (i.e., we want to test if a sample of observations
could be treated as a random sample from a given distribution). We will expand on this notion
in later chapters.

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 4
#| label: fig-ecdf
#| fig-cap: "Empirical and model distribution for a sample of ten quadrats in the example on seed survival."
# Simulate a sample and calculate its empirical distribution.
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
n_quadrats = 10
seed_values = seq.int(from = 0, to = seed_count)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
cumulative_relative_frequency = cumsum(relative_frequency)

model_comparison_data = rbind(
  data.frame(
    surviving_seeds = seed_values,
    value = model_mass,
    quantity = "Probability mass",
    series = "Model"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = relative_frequency,
    quantity = "Probability mass",
    series = "Sample"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = model_cumulative,
    quantity = "Cumulative probability",
    series = "Model"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = cumulative_relative_frequency,
    quantity = "Cumulative probability",
    series = "Sample"
  )
)
model_comparison_data$quantity = factor(
  model_comparison_data$quantity,
  levels = c("Probability mass", "Cumulative probability")
)

# Use different geometries as well as colours to distinguish the model and sample.
ggplot() +
  geom_linerange(
    data = subset(
      model_comparison_data,
      quantity == "Probability mass" & series == "Model"
    ),
    aes(x = surviving_seeds, ymin = 0, ymax = value, colour = series),
    linewidth = 0.8
  ) +
  geom_point(
    data = subset(
      model_comparison_data,
      quantity == "Probability mass" & series == "Sample"
    ),
    aes(x = surviving_seeds, y = value, colour = series),
    alpha = 0.7,
    size = 2
  ) +
  geom_step(
    data = subset(model_comparison_data, quantity == "Cumulative probability"),
    aes(x = surviving_seeds, y = value, colour = series, linetype = series),
    direction = "hv",
    linewidth = 0.8
  ) +
  facet_wrap(~quantity, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = c(Model = "#D55E00", Sample = "#0072B2")) +
  scale_linetype_manual(values = c(Model = "solid", Sample = "dashed")) +
  scale_x_continuous(breaks = seed_values) +
  labs(
    x = "Surviving seeds per quadrat",
    y = "Probability or relative frequency",
    colour = NULL,
    linetype = NULL
  ) +
  theme_classic()
```

### Mean, variance and quantiles of a data sample

Just like for any probability distribution described earlier, we can compute
expectations such as means and variances from an empirical distribution, as well as quantiles
by inverting the empirical cumulative distribution. We will simply replace probabilities in
the formulae by relative frequencies of the sample.

Note that, strictly speaking, a data sample is just a collection of values. It does not have
a mean or a variance; those are properties of distributions. However, language in applied
statistics is often used lazily and inconsistently, which can lead to a lot of confusion.
Whenever we say *sample mean* or *sample variance* we actually imply the mean or variance of
the empirical distribution (but see below that the variance of a sample is an even more
confusing term).

It is trivial to see that the mean of the empirical distribution equals the arithmetic
mean of the sample:

```{r}
# Simulate a small sample and calculate its weighted and ordinary means.
set.seed(123)
n_quadrats = 10
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
weighted_average = sum(relative_frequency * seed_values)
sample_mean = mean(sample_survivors)
c(weighted_average, sample_mean)
```

However, the variance of the empirical distribution is not exactly the same as what we normally
call *sample variance*:

```{r}
# Calculate the variance of the empirical distribution
empirical_variance = sum((seed_values - weighted_average)^2 * relative_frequency)
sample_variance = var(sample_survivors)
c(empirical_variance, sample_variance)
```

It turns out that what most people call sample variance is actually a statistical estimator
of the variance of the model distribution (i.e., in this case the variance of the Binomial
distribution), which is a completely different concept. For reasons that will become clear in
[Chapter 4](../Chapter_4/Theory.qmd), this
variance is calculated with a different formula^[To make matters worse, many individuals will
claim dogmatically that this second way of calculating the sample variance is the correct one,
without realizing what the differences are between the two. Unfortunately, this type of
confusion and misunderstanding from sloppy use of language is very common in applied
statistics... And this has also polluted AI chatbots, so sometimes trying to understand
proper statistics becomes a very uphill battle.].

A quantile from the empirical distribution can be derived from the `quantile()` function. Note
that when dealing with discrete data you need to use `type = 1` as an argument. For example,
if we calculate the quantile for a cumulative probability of 0.5 (i.e. the median) it will
return the smallest observed count whose empirical cumulative probability reaches $0.5$:

```{r}
# Use type = 1 so the empirical quantile is one of the observed counts.
quantile(x = sample_survivors, probs = 0.5, type = 1)
```

The function `quantile()` appears in many areas of statistics, and there are up to nine different
ways of computing it in R. So the purpose for which we calculate the quantile really matters.

### Asymptotic convergence of the empirical distribution

::: {.callout-note title="Asymptotic convergence of empirical distribution"}
For independent observations from the same fixed distribution, the empirical
probability mass function converges to the model probability mass function as sample size
grows towards infinity. All the moments (means, variances, and others), as well as all
quantiles from the empirical distribution, will also converge to their equivalent values
in the model. With bounded counts, as in our binomial example, the empirical mean and
variance also converge to the values in the model. A finite sample can sometimes agree
exactly; agreement is not guaranteed at any finite size.
:::

By defining the empirical distribution we can now better understand how samples and
distributions are related. That is, a collection of observations can be treated as a random
sample from a model if the empirical distribution converges to that model as sample size
increases. This does not imply that the observations are actual random samples from some
mathematical stochastic representation; it simply means that we have modelled the variation
across different observations correctly.

Note that in practice we can never verify this relationship as we will never have infinite
samples and for any finite size there can be deviations. But it does mean that the larger the
sample, the better we can judge if we have a good model (@fig-ecdf-binom). Also, larger
samples are more **stable**, meaning that there will be less variation across repeated samples,
which will increase our confidence (reduce our uncertainty) in the results (we cover this
in more detail in [Chapter 4](../Chapter_4/Theory.qmd)).

```{r}
#| echo: false
#| eval: true
#| label: fig-ecdf-binom
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Model probabilities and sample frequencies for 10 quadrats (top) and 1,000 quadrats (bottom)."
# Compare empirical distributions at two sample sizes.
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_sizes = c(10, 1000)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)

# Store the model and empirical summaries for both sample sizes before plotting.
sample_comparison_data = data.frame()
for (n_quadrats in sample_sizes) {
  sample_survivors = rbinom(
    n = n_quadrats, size = seed_count, prob = seed_survival_probability
  )
  count_survivors = table(factor(x = sample_survivors, levels = seed_values))
  relative_frequency = as.numeric(count_survivors) / n_quadrats
  cumulative_relative_frequency = cumsum(relative_frequency)

  sample_size_label = paste(format(n_quadrats, big.mark = ","), "quadrats")
  sample_comparison_data = rbind(
    sample_comparison_data,
    data.frame(
      surviving_seeds = seed_values,
      value = model_mass,
      quantity = "Probability mass",
      series = "Model",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = relative_frequency,
      quantity = "Probability mass",
      series = "Sample",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = model_cumulative,
      quantity = "Cumulative probability",
      series = "Model",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = cumulative_relative_frequency,
      quantity = "Cumulative probability",
      series = "Sample",
      sample_size = sample_size_label
    )
  )
}

sample_comparison_data$sample_size = factor(
  sample_comparison_data$sample_size,
  levels = c("10 quadrats", "1,000 quadrats")
)
sample_comparison_data$quantity = factor(
  sample_comparison_data$quantity,
  levels = c("Probability mass", "Cumulative probability")
)

# Compare each empirical distribution with the same model using matching panels.
ggplot() +
  geom_linerange(
    data = subset(
      sample_comparison_data,
      quantity == "Probability mass" & series == "Model"
    ),
    aes(x = surviving_seeds, ymin = 0, ymax = value, colour = series),
    linewidth = 0.8
  ) +
  geom_point(
    data = subset(
      sample_comparison_data,
      quantity == "Probability mass" & series == "Sample"
    ),
    aes(x = surviving_seeds, y = value, colour = series),
    alpha = 0.7,
    size = 2
  ) +
  geom_step(
    data = subset(sample_comparison_data, quantity == "Cumulative probability"),
    aes(x = surviving_seeds, y = value, colour = series, linetype = series),
    direction = "hv",
    linewidth = 0.8
  ) +
  facet_grid(sample_size ~ quantity, scales = "free_y") +
  scale_colour_manual(values = c(Model = "#D55E00", Sample = "#0072B2")) +
  scale_linetype_manual(values = c(Model = "solid", Sample = "dashed")) +
  scale_x_continuous(breaks = seed_values) +
  labs(
    x = "Surviving seeds per quadrat",
    y = "Probability or relative frequency",
    colour = NULL,
    linetype = NULL
  ) +
  theme_bw() + theme(panel.grid = element_blank())
```

:::: {.callout-important title="Practice: orchid flowering across plots"}

Suppose each restoration plot contains 12 orchid plants. Each plant flowers with
probability 0.25, independently of the others. Let $X$ be the number of flowering
plants in a plot. This is a binomial model. Simulate counts for 12 plots and for 500
plots. For each sample, calculate the empirical mean and plot the empirical mass function.
Compare both distributions with the model probability mass function. What changes in
the result from the larger sample? How much can you rely on a single simulation?

::: {.callout-tip title="Solution" collapse="true"}

One possible simulation is:

```{r}
# Simulate the two sample sizes and compare each with the same model.
set.seed(246)

plants_per_plot = 12
flowering_probability = 0.25

small_plot_counts = rbinom(
  n = 12,
  size = plants_per_plot,
  prob = flowering_probability
)
large_plot_counts = rbinom(
  n = 500,
  size = plants_per_plot,
  prob = flowering_probability
)

mean(small_plot_counts)
mean(large_plot_counts)

count_values = seq.int(from = 0, to = plants_per_plot)
small_plot_table = table(factor(
  x = small_plot_counts,
  levels = count_values
))
large_plot_table = table(factor(
  x = large_plot_counts,
  levels = count_values
))

small_frequency = as.numeric(small_plot_table) / length(small_plot_counts)
large_frequency = as.numeric(large_plot_table) / length(large_plot_counts)
model_mass = dbinom(
  x = count_values,
  size = plants_per_plot,
  prob = flowering_probability
)

small_frequency_data = data.frame(
  flowering_plants = count_values,
  probability = small_frequency,
  series = "12 plots"
)
large_frequency_data = data.frame(
  flowering_plants = count_values,
  probability = large_frequency,
  series = "500 plots"
)
model_mass_data = data.frame(
  flowering_plants = count_values,
  probability = model_mass,
  series = "Binomial model"
)

# Use bars for one empirical distribution, points for the other, and lines for the model.
# Set common axis limits
all_probabilities = c(
  small_frequency_data$probability,
  large_frequency_data$probability,
  model_mass_data$probability
)

plot(
  NA,
  xlim = range(count_values),
  ylim = c(0, max(all_probabilities)),
  xaxt = "n",
  xlab = "Flowering plants per plot",
  ylab = "Probability or relative frequency"
)

axis(1, at = count_values)

# Draw bars for the small empirical distribution
with(
  small_frequency_data,
  rect(
    flowering_plants - 0.35,
    0,
    flowering_plants + 0.35,
    probability,
    col = "grey",
    border = 1
  )
)

# Add points for the large empirical distribution
with(
  large_frequency_data,
  points(
    flowering_plants,
    probability,
    pch = 16,
    col = 2
  )
)

# Add vertical lines for the model distribution
with(
  model_mass_data,
  segments(
    flowering_plants,
    0,
    flowering_plants,
    probability,
    col = 3
  )
)

# Add one legend for all three series
legend(
  "topright",
  legend = c("12 plots", "500 plots", "Binomial model"),
  col = c(1, 2, 3),
  fill = c("grey", NA, NA),
  border = c(1, NA, NA),
  pch = c(NA, 16, NA),
  lty = c(0, 0, 1)
)
```

The larger sample will generally give a more stable empirical distribution and mean
closer to the values in the model. A particular simulation need not be closer in every
summary, and the empirical distribution need not exactly match the model.

:::
::::

# Probability distributions for count data {#sec-discrete_distributions}

In all the examples above we have used the Binomial distribution, and I made sure that the
examples were adequate for such a distribution. However, there are many different distributions
for counts and other discrete data, each with their own properties (e.g. how means and
variances are related). Below I go over three important examples, though more can be found
in the [distribution supplement](../Supplements/distributions.qmd).

Remember that there is no *correct* distribution for a given sample of data. The range of
possible values and the ecological processes that are relevant give us plausible candidates,
but often we are left with multiple candidate models. In [Chapter 7](../Chapter_7/Theory.qmd)
we will learn formal methods for model comparison, but for now we will use some basic
heuristics to decide which distributions are reasonable to use.

::: {.callout-warning title="Choosing a probability distribution"}
A distribution should be chosen according to the type and range of observed values, its
mean–variance relationship, and how well it describes the observed data. Ecological
interpretation is useful, but a distribution is not restricted to only the process from
which it was originally derived.
:::

## The binomial distribution

The binomial distribution is useful when every sampling unit contains a fixed number of binary
events or outcomes. It models the number of times a particular outcome occurs (e.g., survival,
germination, success, etc.). Each event is assumed to be independent of the others and has the same
probability of obtaining the outcome of interest.

As seen before, the probability mass function of the Binomial distribution is:

$$
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x}
=\frac{n!}{x!(n-x)!}p^x(1-p)^{n-x},
$$

where $\binom{n}{x}$ counts the ways to choose $x$ outcomes of interest (each with probability
$p$ of occurring) from $n$ events and $n!$ is a factorial.

Applying the relevant formulae, we can derive the mean and variance of a binomial distribution as:

$$
\mu=E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$


### Example: seedling survival

Suppose 20 seedlings are planted in each plot and each survives independently with probability
$0.7$. Let $X$ be the number surviving in one plot. The probability that exactly 14 survive
is:

$$
P(X=14)
=
\binom{20}{14}(0.7)^{14}(1-0.7)^{20-14}
=
\frac{20!}{14! (20 - 14)!}(0.7)^{14}(0.3)^6
\approx 0.1916.
$$

We can verify our math with R:

```{r}
# Set the binomial model parameters and the count we want to check.
seedling_count = 20
seedling_survival_probability = 0.7
survivors_to_check = 14

dbinom(
  x = survivors_to_check,
  size = seedling_count,
  prob = seedling_survival_probability
)
```

The model gives an expected count of $20(0.7)=14$ seedlings per plot and
a variance of $20(0.7)(0.3)=4.2$ seedlings squared:

```{r}
# Calculate the expected count and variance from the binomial formulas.
expected_survivors = seedling_count * seedling_survival_probability
variance_survivors = seedling_count *
  seedling_survival_probability *
  (1 - seedling_survival_probability)
```

If a binomial mean and variance are known, we can invert their formulas.
Writing $m=E[X]$ and $v=\operatorname{Var}(X)$ gives

$$
\begin{aligned}
p &= 1-\frac{v}{m}, \\
n &= \frac{m}{p}.
\end{aligned}
$$

This conversion requires $m>0$, a probability $0<p\leq 1$, and a positive
integer result for $n$. For the model above, it recovers $p=0.7$ and $n=20$:

```{r}
# Recover the binomial parameters from the model mean and variance.
mean_survivors = expected_survivors
success_probability = 1 - variance_survivors / mean_survivors
estimated_trial_count = mean_survivors / success_probability
```

With real data, sample means and variances will not generally recover exact model parameters.
That is not bad algebra; it is parameter uncertainty. [Chapter 4](../Chapter_4/Theory.qmd#sec-estimates)
uses sample summaries to construct estimates and explains why estimates vary between samples.

## The Poisson distribution

The Poisson distribution is a common starting point when we count events in a fixed unit of time,
area, or sampling effort: insects caught per trap-night, seedlings per quadrat, or camera detections
per night. It is often the default distribution for count data when we do not have a fixed
upper limit to the number of observations. Its most important property is to always have the
same variance and mean, such that when this is not appropriate, we need to switch to other
distributions for count data.
The Poisson distribution has one parameter:

$$
\lambda=\text{expected count per sampling unit}, \qquad \lambda\geq 0.
$$

Its probability mass function is

$$
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!},
$$

The mean and variance are both equal to $\lambda$:

$$
\mu = E[X]=\lambda
$$

and

$$
\operatorname{Var}(X)=\lambda.
$$


### Example: wildlife detections

Suppose a camera records an average of $0.8$ detections per night. If the expected detection
count scales with observation time, three nights have expected count

$$
\lambda = 3(0.8) = 2.4.
$$

Under an assumed Poisson model, the probability of exactly two detections
over those three nights is

$$
P(X=2)
=
\frac{e^{-2.4}(2.4)^2}{2!}
\approx 0.2613.
$$

We can calculate it in R:

```{r}
# Calculate the Poisson mean for three nights and one exact probability.
rate_per_night = 0.8
n_nights = 3
expected_detections = rate_per_night * n_nights
detections_to_check = 2

dpois(x = detections_to_check, lambda = expected_detections)
```

We can also ask the probability of at least five detections in three nights. We can compute
this from the cumulative probability (note that we calculate the cumulative probability for
four and then take the complement):

```{r}
# Calculate the probability of at least the requested number of detections.
minimum_detections = 5
1 - ppois(
  q = minimum_detections - 1,
  lambda = expected_detections
)
```

Unlike a binomial count, a Poisson count has no fixed upper limit. A count of 100 detections is
possible under this model, although extremely unlikely when $\lambda=2.4$.

```{r}
#| echo: false
#| fig-cap: "Poisson probability mass for detections during three camera nights."
# Calculate and plot the Poisson probability mass over a useful range of counts.
expected_detections = 2.4
maximum_detections_to_plot = 15
detection_values = seq.int(from = 0, to = maximum_detections_to_plot)
detection_mass = dpois(x = detection_values, lambda = expected_detections)

detection_mass_data = data.frame(
  detections = detection_values,
  probability = detection_mass
)

# Plot each possible detection count as a vertical probability line.
plot(
  detection_mass_data$detections,
  detection_mass_data$probability,
  type = "h",
  xaxt = "n",
  xlab = "Number of detections",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: flowering stems per orchid plot"}

A survey covers two square metres of a restoration site. Assume flowering stems occur
at an average rate of 1.2 per square metre. Calculate the probability of finding exactly
three stems in the surveyed area. Why does this model allow counts greater than three,
even though three is the count in the question?

::: {.callout-tip title="Solution" collapse="true"}

The expected count across two square metres is $\lambda=2(1.2)=2.4$. The probability is:

```{r}
# Convert the observation effort into the Poisson mean before calculating a probability.
rate_per_square_metre = 1.2
survey_area = 2
expected_stems = rate_per_square_metre * survey_area

dpois(x = 3, lambda = expected_stems)
```

:::
::::

## Overdispersion

The Poisson model is characterised by having its mean and variance equal. When the variance
is higher than the mean, we say that such data has *overdispersion* relative to the Poisson
distribution^[The "relative to the Poisson distribution" bit is often omitted; it is generally
implied that this is the reference].

For example, seedling counts may differ strongly among quadrats because some have better
habitat. But extra variation can also reflect differences in detection or dependence among
observations. The distribution itself will not tell you why overdispersion happens. Note that
if you could identify and measure the sources of overdispersion you could use the techniques
from [Chapter 8](../Chapter_8/Theory.qmd) to model it.

::: {.callout-warning title="Overdispersion is a statistical description"}
Overdispersion means that the observed variation is greater than the variation expected
under a reference, often the Poisson distribution. It does not identify the ecological cause.
Clustering, habitat differences, omitted predictors, detection variation, and dependence
can all produce overdispersion.
:::

### A simple diagnostic for overdispersion

A first exploratory check compares the sample variance with the sample mean (for convenience
just use `var()`; be aware of the subtle differences mentioned earlier). If the observations
are reasonably described as independent Poisson counts with the same mean, these quantities
tend to be similar. The **dispersion ratio** measures to what extent this holds true:

$$
\text{dispersion ratio}
=
\frac{\text{sample variance}}{\text{sample mean}}.
$$

Here we simulate quadrat counts from a negative binomial model to see what
the ratio can look like when variation exceeds a Poisson model's prediction:

```{r}
# Simulate overdispersed quadrat counts from a negative binomial model.
set.seed(seed = 123)

n_quadrats = 100
negative_binomial_size = 1.5
negative_binomial_mean = 4

counts = rnbinom(
  n = n_quadrats,
  size = negative_binomial_size,
  mu = negative_binomial_mean
)

sample_mean = mean(counts)
sample_variance = var(counts)
dispersion_ratio = sample_variance / sample_mean
c(sample_mean, sample_variance, dispersion_ratio)
```

A ratio close to one is consistent with the Poisson mean–variance relationship;
a substantially larger ratio suggests overdispersion relative to that model.
It is possible to obtain a ratio substantially lower than one, although it is rare. This
is called *underdispersion*. The Conway-Maxwell-Poisson distribution described in the
[Supplement](../Supplements/distributions.qmd) can handle underdispersion.


:::: {.callout-important title="Practice: variation in orchid stem counts"}

Across a sample of restoration plots, the mean flowering-stem count is 3 and the sample
variance is 12. Calculate the dispersion ratio and describe what it suggests relative to
a Poisson model. Can this ratio tell you why the counts vary so much?

::: {.callout-tip title="Solution" collapse="true"}

The dispersion ratio is $12/3=4$. It suggests that the observed variation is greater
than expected under a Poisson model with the same mean. The ratio alone cannot identify
the cause; for example, plots may differ in habitat, or counts may be clustered.

:::
::::

## The negative binomial distribution

When overdispersion occurs, the negative binomial distribution is a useful starting model.
It has support $\{0, 1, 2, \ldots\}$ and can accommodate more variation
than a Poisson model with the same mean. The negative binomial model has two parameters:

- $\mu>0$: the mean count per sampling unit;
- $k>0$: a shape or dispersion parameter that controls the additional variation

The probability mass function is

$$
P(X=x)
=
\frac{\Gamma(x+k)}{\Gamma(k)x!}
\left(\frac{k}{k+\mu}\right)^k
\left(\frac{\mu}{k+\mu}\right)^x,
$$

where $\Gamma$ is the gamma function, which extends the factorial to real values.

The mean and variance are:

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k}.
$$

For $\mu>0$ and finite $k$, the variance exceeds the mean. At a fixed $\mu$, smaller $k$ gives
greater variance. As $k$ increases, the extra variation fades and the variance approaches the mean:

$$
\mu+\frac{\mu^2}{k}\longrightarrow\mu.
$$

Thus, the negative binomial distribution approaches the Poisson distribution as $k$ becomes
large.

In R, `dnbinom()` and related functions, the argument `size` corresponds to $k$. This naming
can be confusing because `size` does not mean the sample size here; it is the negative
binomial shape parameter. The *Details* section of the documentation page for `dnbinom`
explains why it has that name.

### Example: seedlings in quadrats

Suppose the mean number of seedlings per quadrat is $\mu=4$, and the shape parameter is $k=1.5$.
The expected number of seedlings is

$$
E[X]=4.
$$

The variance is

$$
\operatorname{Var}(X)
=
4+\frac{4^2}{1.5}
\approx 14.67.
$$

Across comparable quadrats, the model predicts four seedlings on average and a variance of
approximately $14.67$ seedlings squared. That is much more variation than the Poisson model would
allow at the same mean.

We can verify these calculations in R:

```{r}
# Calculate the negative binomial mean and variance from its parameters.
mean_seedlings = 4
negative_binomial_size = 1.5

expected_seedlings = mean_seedlings
variance_seedlings = mean_seedlings +
  mean_seedlings^2 / negative_binomial_size

c(mean = expected_seedlings, variance = variance_seedlings)
```

The probability of observing exactly 10 seedlings is:

```{r}
# Calculate the probability of one particular seedling count.
seedlings_to_check = 10

dnbinom(
  x = seedlings_to_check,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

The probability of observing at least 10 seedlings is:

```{r}
# Calculate the probability of at least the requested number of seedlings.
minimum_seedlings = 10

1 - pnbinom(
  q = minimum_seedlings - 1,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

Like the Poisson distribution, the negative binomial distribution has no fixed
upper limit. A count of 100 is mathematically possible, although very unlikely if your mean
is 4.

```{r}
#| echo: false
#| fig-cap: "Negative binomial probability mass for seedlings in a quadrat."
# Calculate and plot the negative binomial probability mass over a useful range.
mean_seedlings = 4
negative_binomial_size = 1.5
maximum_seedlings_to_plot = 20
seedling_values = seq.int(from = 0, to = maximum_seedlings_to_plot)
negative_binomial_mass = dnbinom(
  x = seedling_values,
  size = negative_binomial_size,
  mu = mean_seedlings
)

negative_binomial_mass_data = data.frame(
  seedlings = seedling_values,
  probability = negative_binomial_mass
)

# Vertical lines make the separate possible count probabilities visible.
plot(
  negative_binomial_mass_data$seedlings,
  negative_binomial_mass_data$probability,
  type = "h",
  xaxt = "n",
  xlab = "Number of seedlings",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: orchid stem counts with different dispersion"}

Suppose the mean number of flowering stems per plot is $\mu=3$. Calculate the negative
binomial variance for $k=1.5$ and $k=6$. Which value gives greater overdispersion? How
does increasing $k$ relate to the Poisson model?

::: {.callout-tip title="Solution" collapse="true"}

Using $\operatorname{Var}(X)=\mu+\mu^2/k$, the variances are 9 and 4.5, respectively.
The smaller value of $k$ gives greater overdispersion. As $k$ increases, the variance
approaches the mean, which is the Poisson mean–variance relationship.

:::
::::

# Chapter quiz

Probability only becomes useful when you can move comfortably between an ecological question,
the possible observations, and a model's assumptions. Try each question before opening its
solution.

:::: {.callout-important title="Question 1: lichen patches"}

On a fixed section of rock, let $X$ be the number of lichen patches found.
Suppose every count from zero through 18 is possible, but no larger count
is possible. What is the support of $X$? Is it finite or countably infinite?

::: {.callout-tip title="Answer" collapse="true"}

$X$ can take the integer values 0 through 18. Its support is finite because the survey
area has a maximum number of patches in this scenario.

:::
::::

:::: {.callout-important title="Question 2: freshwater larvae"}

A proposed distribution for the number of larvae in a sample assigns probabilities
0.15, 0.25, 0.40, and 0.30 to counts 0, 1, 2, and 3. Is this a valid probability
distribution? Explain.

::: {.callout-tip title="Answer" collapse="true"}

No. Although each probability is non-negative, they sum to 1.10 rather than 1.
Probabilities over all possible outcomes must sum to one.

:::
::::

:::: {.callout-important title="Question 3: coral spawning"}

For a model of the number of coral colonies spawning on a reef, which R function prefix
would you use to calculate (a) the probability of exactly five colonies, (b) the
probability of five or fewer, (c) the 0.9 quantile, and (d) a random sample?

::: {.callout-tip title="Answer" collapse="true"}

Use the `d` prefix for an exact probability, `p` for a cumulative probability, `q` for
a quantile, and `r` to generate a random sample.

:::
::::

:::: {.callout-important title="Question 4: acorn production"}

Suppose acorn production by two oak trees is modelled with independent random variables.
In this model, what does independence say about the probability of a particular joint
outcome? Give one ecological factor that might make the assumption questionable.

::: {.callout-tip title="Answer" collapse="true"}

Under independence, the probability of a particular pair of outcomes is the product of
their separate probabilities. Shared weather or soil conditions could make production
by the two trees dependent.

:::
::::

:::: {.callout-important title="Question 5: trout eggs"}

For a group of 12 trout eggs, suppose each egg hatches with probability 0.25 and the
outcomes are independent. Calculate the expected number hatched and the variance assuming
a Binomial distribution.

::: {.callout-tip title="Answer" collapse="true"}

For a binomial model, $E[X]=np=12(0.25)=3$ and
$\operatorname{Var}(X)=np(1-p)=12(0.25)(0.75)=2.25$.

:::
::::

:::: {.callout-important title="Question 6: mussel settlement"}

A model predicts an expected count (or mean) of 2.4 newly settled mussels per tile. Can one tile
have 2.4 mussels? What does the expected value describe?

::: {.callout-tip title="Answer" collapse="true"}

A tile cannot have a fractional mussel. The expected value describes the long-run average
count across comparable tiles under the model.

:::
::::

:::: {.callout-important title="Question 7: moth emergence"}

In a sample of 30 host plants, moths emerged from 9. Calculate the relative frequency
of emergence. How does this relate to the probability of emergence for all relevant host
plants?

::: {.callout-tip title="Answer" collapse="true"}

The relative frequency is $9/30=0.3$. It is a sample-based estimate of the probability,
not necessarily the exact model probability. Another sample could produce a
different relative frequency.

:::
::::

:::: {.callout-important title="Question 8: caddisfly larvae"}

You sample the number of caddisfly larvae in many equal-sized stream sections. As the
number of sections grows, what do you generally expect to happen to the empirical
distribution relative to the distribution that generated the data? Is exact agreement
expected for a finite sample?

::: {.callout-tip title="Answer" collapse="true"}

With more independent observations from the same distribution, the empirical distribution
generally becomes closer to the generating distribution. Exact agreement is not expected
in a finite sample, and individual simulations can vary.

:::
::::

:::: {.callout-important title="Question 9: choose a count model"}

Choose a plausible starting model for each scenario and briefly explain your choice:

1. The number of eggs hatching out of 15 eggs, with the same hatch probability for each.
2. The number of mayfly larvae in a standard sample, when the count is unbounded and
   its mean and variance are approximately equal.
3. The number of burrowing crabs per beach section, when counts are unbounded and the
   sample variance is substantially greater than the mean.

::: {.callout-tip title="Answer" collapse="true"}

1. A binomial model is plausible if the egg outcomes are independent, because
   there is a fixed number of trials and a common hatch probability.
2. A Poisson model is a plausible starting point because the support is non-negative
   integer counts and the mean and variance are approximately equal.
3. A negative binomial model is a plausible starting point because it allows an
   unbounded count with variance greater than the mean.

:::
::::

:::: {.callout-important title="Question 10: bumblebee nests"}

Across meadow plots, the sample mean number of bumblebee nests is 2 and the sample
variance is 6. Calculate the dispersion ratio and explain what it suggests relative to
a Poisson model. Does it identify the ecological cause?

::: {.callout-tip title="Answer" collapse="true"}

The ratio is $6/2=3$, suggesting greater variation than expected under a Poisson model
with the same mean. The ratio does not identify a cause; clustering, habitat differences,
or dependence could contribute.

:::
::::
