---
title: "Introduction to Probability (Count data)"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Introduction

In this chapter, we learn how to use probability distributions to describe that variation and connect a
model to the counts we observe. We focus on **discrete random variables**, which take
separate, countable values such as $0, 1, 2, \ldots$. [Chapter 3](../Chapter_3/Theory.qmd)
extends these ideas to continuous measurements.

# Learning goals

After studying this chapter, you should be able to:

- describe a discrete random variable and its probability distribution;
- calculate and interpret probability mass, cumulative probability, and quantiles;
- explain the independence assumption and when it may be doubtful;
- calculate and interpret means and variances;
- compare samples with the population and model they represent;
- simulate ecological counts and compare empirical and theoretical distributions;
- identify overdispersion relative to a specified count model;
- choose plausible binomial, Poisson, and negative binomial starting models.

# Random variables and probability distributions

A **random variable** is a quantity whose value is uncertain before an observation.
For a count, its possible values are non-negative integers. Suppose we place four seeds
in each quadrat, a marked area used for sampling, and later count how many survive in
one quadrat. Let $X$ be that count:

$$
X=\text{the number of seeds surviving in a quadrat}.
$$

If the quadrat contains four seeds, then $X$ can take the values

$$
0,1,2,3,4.
$$

These are the possible counts. The **support** is the set of values assigned positive
probability by a model. If every count from zero to four has positive probability,
the support is $\{0, 1, 2, 3, 4\}$. A discrete model can also have **countably
infinite** support, such as $\{0, 1, 2, 3, \ldots\}$ for counts with no fixed
upper limit.

::: {.callout-note title="Discrete random variable"}
A discrete random variable takes values from a finite or countably infinite set. Its
probability distribution assigns a probability to each possible value, and the
probabilities sum to one.
:::

For each possible value of $X$ we can assign a probability. For example:

| $x$ | $P(X=x)$ |
|---:|---:|
| 0 | 0.05 |
| 1 | 0.15 |
| 2 | 0.30 |
| 3 | 0.30 |
| 4 | 0.20 |

All probabilities must be **non-negative**:

$$
P(X=x)\geq 0,
$$

and they must add up to one:

$$
\sum_x P(X=x)=1.
$$

The notation $X$ refers to the random variable, while $x$ refers to one possible or observed value
of that variable.

We can also add probabilities for a set of counts. For example:

$$
P(X \leq 3) = \sum_{x=0}^{3} P(X=x).
$$

For larger supports, we use a **probability mass function** to calculate the
probability assigned to each count (see @sec-four_faces).

::: {.callout-note title="Probability distribution"}
A probability distribution assigns probabilities to the possible values of a random
variable. Its probability mass function and cumulative distribution function are
two ways to describe those probabilities (see @sec-four_faces).
:::

We will use a binomial model for seed survival to learn four related R operations.
In @sec-discrete_distributions, we examine its assumptions and probability mass
function alongside two other count models. [Chapter 7](../Chapter_7/Theory.qmd#sec-response-comparison)
introduces criteria for comparing fitted models.

## The four faces of a discrete distribution {#sec-four_faces}

### 1. Probability mass

The probability mass assigned to $X=x$ is

$$
P(X=x).
$$

For example, $P(X=4)$ is the probability that all four seeds in a quadrat survive.
The masses across possible counts show the model's shape. Later, we will use mass
functions to estimate parameters and check fitted models against observations.

Suppose each seed has survival probability $0.4$, independently of the others.
We can display the binomial probability mass with bars or vertical lines:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Binomial probability mass for the number of surviving seeds in a quadrat."
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
seed_mass = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)

par_val = par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)

barplot(
  height = seed_mass,
  names.arg = seed_values,
  xlab = "Surviving seeds per quadrat",
  ylab = "Probability",
  ylim = c(0, 0.4)
)

plot(
  x = seed_values,
  y = seed_mass,
  type = "h",
  lwd = 3,
  xlab = "Surviving seeds per quadrat",
  ylab = "Probability",
  ylim = c(0, 0.4)
)
par(par_val)
```

In R, distribution functions with the prefix `d` calculate probability mass for
discrete values. For example, `dbinom()` gives the probability that two seeds survive:

```{r}
dbinom(x = 2, size = seed_count, prob = seed_survival_probability)
```

### 2. Cumulative probability

The **cumulative distribution function** adds the probability masses up to and
including a count $x$:

$$
F_X(x)=P(X\leq x).
$$

For example, $F_X(3)$ is the probability that at most three seeds survive.
The complementary upper-tail probability $P(X>3)$ answers a different question.
We can draw the cumulative function for discrete counts as a step plot:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Cumulative probability for the number of surviving seeds in a quadrat."
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
seed_cumulative = pbinom(
  q = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)

plot(
  x = seed_values,
  y = seed_cumulative,
  type = "s",
  xlab = "Surviving seeds per quadrat",
  ylab = "Cumulative probability",
  ylim = c(0, 1)
)
```

The cumulative probability is always between zero and one, and it never decreases as $x$ increases.

In R, distribution functions with the prefix `p` calculate cumulative
probabilities. `pbinom()` gives the probability that at most two seeds survive:

```{r}
pbinom(q = 2, size = seed_count, prob = seed_survival_probability)
```

To calculate the probability that more than two seeds survive, set
`lower.tail = FALSE`:

```{r}
pbinom(
  q = 2,
  size = seed_count,
  prob = seed_survival_probability,
  lower.tail = FALSE
)
```

This is equivalent to subtracting the lower-tail probability from one:

```{r}
1 - pbinom(q = 2, size = seed_count, prob = seed_survival_probability)
```

### 3. Quantiles

A **quantile** gives a count associated with a specified cumulative probability.
For a discrete distribution, the cumulative function jumps, so it has no ordinary
inverse for every probability between zero and one. Instead, the $\alpha$-quantile
is the smallest count whose cumulative probability reaches or exceeds $\alpha$:

$$
q_\alpha=\min\{x:F_X(x)\geq\alpha\}.
$$

For example, the median is the $0.5$ quantile. Quantiles also enter confidence
interval calculations in [Chapter 4](../Chapter_4/Theory.qmd#sec-confidence_intervals).
They can be used to generate random values from a distribution, as we discuss below.

We can plot the quantile against its cumulative-probability input:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Binomial quantiles for the number of surviving seeds in a quadrat."
seed_count = 4
seed_survival_probability = 0.4
probability_values = seq(from = 0, to = 1, length.out = 501)
seed_quantiles = qbinom(
  p = probability_values,
  size = seed_count,
  prob = seed_survival_probability
)

plot(
  x = probability_values,
  y = seed_quantiles,
  type = "s",
  xlab = "Cumulative probability",
  ylab = "Surviving seeds per quadrat",
  ylim = c(0, seed_count)
)
```

In R, distribution functions with the prefix `q` calculate quantiles.
For the seed-survival model, `qbinom()` finds the smallest count with at
least $0.5$ cumulative probability:

```{r}
median_probability = 0.5
qbinom(
  p = median_probability,
  size = seed_count,
  prob = seed_survival_probability
)
```

### 4. Sampling

Simulating counts from a model helps us see how observations vary when the model
parameters are fixed. In a simulated sample, each count is one realisation of the
random variable. One way to generate each count is:

1. Generate a random number between zero and one.

2. Apply the distribution's quantile function to that number.

Repeat these steps for each observation. R's sampling functions carry out the
simulation directly. We compare samples with their model in @sec-sample_population.

A sampling function generates values rather than a curve to plot. We can visualise
the resulting counts through their frequencies.

In R, distribution functions with the prefix `r` generate random values.
We set a seed before this example so rerunning it with the same R settings
reproduces the sample. Changing the seed can change the sample. Here we
simulate ten quadrats, each containing four seeds:

```{r}
n_simulated_quadrats = 10
set.seed(seed = 42)
sample_survivors = rbinom(
  n = n_simulated_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
sample_survivors
```

The output contains ten simulated counts. Each is the number of surviving seeds
in one quadrat; these are simulated observations, not field measurements.

::: {.callout-note title="The four faces of distributions in R"}
For a distribution in R:

- `d*` functions calculate probability mass
- `p*` functions calculate cumulative probabilities
- `q*` functions calculate quantiles
- `r*` functions generate random samples

The first argument is conventionally named `x`, `q`, `p`, or `n`, respectively.
:::

:::: {.callout-important title="Practice: orchid fruit set"}

In a restoration plot, six tagged orchid flowers each have probability $0.5$ of setting
fruit. Let $X$ be the number of these flowers that set fruit.

1. What values can $X$ take?
2. Use R to calculate the probability that exactly four flowers set fruit and the
   probability that at most four set fruit.
3. Find the 0.75 quantile of $X$. In words, what does this quantile tell you?
4. Which R function prefix (`d`, `p`, `q`, or `r`) did you use for each calculation?

::: {.callout-tip title="Solution" collapse="true"}

The count can be any integer from 0 to 6. The exact-count probability uses `dbinom()`,
the cumulative probability uses `pbinom()`, and the quantile uses `qbinom()`.

```r
flower_count = 6
fruit_set_probability = 0.5

dbinom(x = 4, size = flower_count, prob = fruit_set_probability) # 0.23
pbinom(q = 4, size = flower_count, prob = fruit_set_probability) # 0.89
qbinom(p = 0.75, size = flower_count, prob = fruit_set_probability) # 4
```

Under this model, at least 75% of the probability is on counts of four or fewer.

:::
::::

## Independence and expectations {#sec-independence_expectations}

### Independence

Two random variables are **independent** if knowing the value of one does not
change the probabilities for the other. Consider germination of two planted seeds.
Let $X_1$ and $X_2$ indicate whether the first and second seed germinate:
each variable takes value $1$ for germination and $0$ otherwise. Their possible
values are $\{0, 1\}$; the variables themselves are $X_1$ and $X_2$.

We can ask for the probability of a pair of outcomes. For example,
$P(X_1=1, X_2=0)$ is the probability that the first seed germinates and the
second does not. If the germination outcomes are independent, the probability
of any pair is the product of its separate probabilities:

$$
P(X_1=x_1, X_2=x_2)=P(X_1=x_1)P(X_2=x_2).
$$

::: {.callout-warning title="Independence is a modelling assumption"}
Independence means that knowing the value of one random variable does not change the
probability distribution of another. It is not guaranteed merely because observations
were collected at different locations or times.
:::

Dependence can arise through spatial clustering, shared weather, competition,
or repeated measurements of the same individual. [Chapter 8](../Chapter_8/Theory.qmd)
examines grouped observations. Sampling seeds from different quadrats alone does
not establish independence.

#### Independent and identically distributed random variables

We often assume that observations are both independent and **identically
distributed**: each follows the same probability distribution. We call these
assumptions **independent and identically distributed (i.i.d.)**. They simplify
calculations used in later parameter estimation, but we must assess whether
they are reasonable for the ecological setting.

If both seeds have the same germination probability $p_g$ and their outcomes
are independent, the probability that both germinate is:

$$
P(X_1=1, X_2=1)=p_g\cdot p_g=p_g^2.
$$

### Expectations

An **expectation** is a probability-weighted average of a quantity calculated
from a random variable. The mean and variance describe a distribution's centre
and spread. We can compare these theoretical properties with sample summaries;
[Chapter 4](../Chapter_4/Theory.qmd#sec-estimates) uses that connection in
the method of moments.


#### Expected value

For a discrete random variable $X$, the expected value is a probability-weighted average
of all possible values. That is, we sum all possible values of $X$ weighted by their
individual probabilities.

$$
E[X] = \sum_x xP(X=x).
$$

For the binomial model of seed survival, calculate the expected count in R:

```{r}
seed_values = seq.int(from = 0, to = seed_count)
seed_probability = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
expected_value = sum(seed_values * seed_probability)
expected_value
```

The expected value is also the **mean** of the distribution, often denoted by $\mu$.

We can apply the same idea to a function $f$ of $X$. For each possible value
$x$, calculate $f(x)$ and weight it by $P(X=x)$:

$$
E[f(X)] = \sum_x f(x)P(X=x).
$$

::: {.callout-note title="Expected value is a long-run average"}
The expected value is a theoretical property of a probability distribution. It need not
be an observable value for an individual observation. For example, a quadrat cannot
contain 1.6 surviving seeds, even though the expected number may be 1.6.
:::

#### Variance and standard deviation

The **variance** is the expected squared distance from the mean. It measures
how widely a distribution spreads its probability across possible values:

$$
\operatorname{Var}(X) = E[(X-E[X])^2] = \sum_x P(X=x)(x-E[X])^2.
$$

For distributions with comparable means, a smaller variance generally means
counts are more concentrated near the mean. Variance for seed counts has units
of seeds squared.

Using the binomial masses calculated above, we get:

```{r}
variance = sum((seed_values - expected_value)^2 * seed_probability)
variance
```

The **standard deviation** is the square root of the variance:

$$
\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.
$$

It has units of seeds rather than seeds squared.

::: {.callout-note title="Variance and standard deviation"}
Variance measures the spread of a distribution in squared units. Standard deviation is
the square root of variance and has the same units as the original variable, making it
usually easier to interpret.
:::

:::: {.callout-important title="Practice: fruit set on orchid plants"}

Suppose eight tagged flowers on each of two orchid plants have the same fruit-set
probability, $p=0.6$. Let $T$ be the total number of flowers setting fruit across
both plants. For this exercise, model the outcomes of the 16 flowers as independent.

1. Use the probability mass function to find the probability of every possible value of
   $T$.
2. Calculate the mean and variance of $T$ using the probability mass function from point 1.
3. Explain what the mean represents. Give one ecological reason why independence among
   the flowers might be doubtful.

::: {.callout-tip title="Solution" collapse="true"}

The possible counts range from 0 to 16. First calculate the probability of each count,
then use the weighted-sum formulas for the mean and variance:

```r
fruit_count_values = seq.int(from = 0, to = 16)
fruit_probabilities = dbinom(
  x = fruit_count_values,
  size = 16,
  prob = 0.6
)

expected_fruit_count = sum(fruit_count_values * fruit_probabilities)
fruit_count_variance = sum(
  (fruit_count_values - expected_fruit_count)^2 * fruit_probabilities
)

expected_fruit_count
fruit_count_variance
```

The mean is 9.6 and the variance is 3.84. The mean is the long-run average fruit
count across comparable groups of 16 flowers; one group must have an integer
count. Flowers on the same plant may share physiological conditions or pollinator
visits, making their outcomes dependent. The binomial calculation is conditional
on the stated independence assumption; [Chapter 8](../Chapter_8/Theory.qmd)
introduces models for grouped observations.
:::
::::

# Samples and populations {#sec-sample_population}

So far, we have used probability distributions to describe possible counts.
How do those models relate to the counts collected in an ecological study?

::: {.callout-note title="Sample and population"}
The **population** is the set of units or observations about which we want to
learn. A **sample** contains the units or observations we collect. A probability
distribution is a model for variation in their values, not the population itself.
:::

For example, we might want to learn about seed survival across all quadrats in a
restoration site, while observing only some of them. To use a model to learn about
the site, we need assumptions about how the quadrats were selected and how their
counts vary. Random sampling can support this connection; a biased selection of
quadrats may not represent the population we care about.

A probability model represents variation and uncertainty without claiming that
ecological processes are fundamentally random. Its assumptions are approximations
that we check against the study design and observations.^[George E. P. Box
summarised this idea as "all models are wrong but some are useful".]

## From frequency to probability

For a sample of counts, the **relative frequency** of a value is the number
of times it appears divided by the sample size. Here we simulate ten quadrats
from the binomial model used earlier:

```{r}
set.seed(seed = 123)
n_quadrats = 10
sample_survivors = rbinom(
  n = n_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
```

`table()` counts how often each value appears. Divide by the number of quadrats
to calculate relative frequencies:

```{r}
seed_values = seq.int(from = 0, to = seed_count)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = count_survivors / length(sample_survivors)
relative_frequency
```

Some possible counts may be absent from a small sample. Setting all possible
values as `factor()` levels makes `table()` include those counts with frequency zero.

The relative frequencies lie between zero and one and sum to one. Compare them
with the theoretical probabilities under the assumed binomial model:

```{r}
data.frame(
  seed_count = seed_values,
  relative_frequency = as.numeric(x = relative_frequency),
  theoretical_probability = dbinom(
    x = seed_values,
    size = seed_count,
    prob = seed_survival_probability
  )
)
```

The two columns differ because the sample contains only ten quadrats. Under
independent sampling from the same fixed distribution, the relative frequency
of each count approaches its model probability as sample size grows. This is
an application of the **law of large numbers**.

::: {.callout-note title="Frequencies converge to probabilities"}
For independent observations from the same distribution, the relative
frequency of any specified value approaches its probability mass as sample
size grows. A larger realised sample need not be closer for every value.
:::

We can compare simulated relative frequencies at sample sizes from 10 to
100,000. The entries vary between simulations, even though the model is fixed:

```{r}
set.seed(123)
sample_sizes = 10^(seq_len(length.out = 5))
seed_values = seq.int(from = 0, to = seed_count)

all_frequencies = matrix(NA_real_, nrow = length(seed_values),
                         ncol = length(sample_sizes) + 1)
colnames(all_frequencies) = c(as.character(sample_sizes), "theory")

for (i in seq_along(sample_sizes)) {
  sample_survivors = rbinom(
    n = sample_sizes[i],
    size = seed_count,
    prob = seed_survival_probability
  )
  count_survivors = table(factor(
    x = sample_survivors,
    levels = seed_values
  ))
  all_frequencies[, i] = as.numeric(count_survivors) / sample_sizes[i]
}
all_frequencies[, length(sample_sizes) + 1] = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
round(all_frequencies, 3)
```

:::: {.callout-important title="Practice: fruit-set frequency"}
In a sample of 24 tagged orchid flowers, 15 set fruit. Calculate the relative frequency
of fruit set. Explain what this value estimates and why it need not equal the underlying
fruit-set probability exactly.

::: {.callout-tip title="Solution" collapse="true"}

The relative frequency is $15/24=0.625$. It estimates the probability that a flower sets
fruit in the population represented by the sample. A different sample could give a
different relative frequency, especially when the sample is small.
:::
::::

## Empirical distribution

The **empirical distribution** assigns each observed count its relative frequency.
It lets us compare a sample with the probability distribution assumed for the
population it represents.

::: {.callout-note title="Empirical distribution"}
A distribution whose mass at each count equals that count's relative frequency
in the sample.
:::

Like a theoretical distribution, an empirical distribution has a cumulative
function: the fraction of observations at or below each count. The figure compares
both functions for ten quadrats simulated from the seed-survival model. Here ten
is the number of quadrats sampled; four is the number of seeds in each quadrat.

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Model probabilities and sample frequencies for ten simulated quadrats."
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
n_quadrats = 10
seed_values = seq.int(from = 0, to = seed_count)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
cumulative_relative_frequency = cumsum(relative_frequency)

old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)
plot(
  x = seed_values, y = model_mass, type = "h", lwd = 2,
  xlab = "Surviving seeds per quadrat", ylab = "Mass / relative frequency",
  ylim = c(0, max(model_mass, relative_frequency))
)
points(x = seed_values, y = relative_frequency, col = "firebrick", pch = 16)
legend(
  x = "topright", legend = c("Model", "Sample"),
  col = c("black", "firebrick"), lty = c(1, NA), pch = c(NA, 16),
  bty = "n"
)
plot(
  x = seed_values, y = model_cumulative, type = "s",
  xlab = "Surviving seeds per quadrat", ylab = "Cumulative probability",
  ylim = c(0, 1)
)
lines(x = seed_values, y = cumulative_relative_frequency,
      type = "s", col = "firebrick", lty = 2)
legend(
  x = "bottomright", legend = c("Model", "Sample"),
  col = c("black", "firebrick"), lty = c(1, 2), bty = "n"
)
par(old_par)
```

### Mean, variance and quantiles of a sample

The mean of an empirical distribution is its relative-frequency-weighted
average. This equals the arithmetic mean of the sample:

```{r}
set.seed(123)
n_quadrats = 10
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
weighted_average = sum(relative_frequency * seed_values)
sample_mean = mean(sample_survivors)
c(weighted_average, sample_mean)
```

The variance of this empirical distribution uses divisor $n$, the number
of quadrats:

```{r}
empirical_variance = sum((seed_values - weighted_average)^2 * relative_frequency)
empirical_variance
```

R's `var()` uses divisor $n-1$, so it gives a different result:

```{r}
var(sample_survivors)
```

The $n-1$ version estimates the variance of the modelled population from a
sample. [Chapter 4](../Chapter_4/Theory.qmd) explains why that correction is used.

To match the discrete quantile definition above, use `type = 1` with
`quantile()`. This returns the smallest observed count whose empirical
cumulative probability reaches $0.5$:

```{r}
quantile(x = sample_survivors, probs = 0.5, type = 1)
```

The default `quantile()` method can interpolate between observations, so it
need not return a count in the empirical distribution's support.

### Asymptotic convergence of the empirical distribution

::: {.callout-note title="Asymptotic convergence of empirical distribution"}
For independent observations from the same fixed distribution, the empirical
cumulative probability at each count approaches the model's cumulative
probability as sample size grows. With bounded counts, as in our binomial
example, the empirical mean and variance also approach their model values.
A finite sample can sometimes agree exactly; agreement is not guaranteed
at any finite size.
:::

This is why larger samples can make empirical summaries more stable, provided
the model and sampling assumptions hold. It supports the method of moments in
[Chapter 4](../Chapter_4/Theory.qmd#sec-estimates). Quantiles at jumps in a
discrete cumulative function require separate care; the statement above does
not claim that every empirical quantile converges in the same way.

The next figure compares two simulated sample sizes with the same binomial
model. The larger sample generally follows the model more closely, although
individual values can move away from their targets in a particular simulation.

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Model probabilities and sample frequencies for 10 quadrats (top) and 1,000 quadrats (bottom)."
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_sizes = c(10, 1000)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)

old_par = par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)
for (n_quadrats in sample_sizes) {
  sample_survivors = rbinom(
    n = n_quadrats, size = seed_count, prob = seed_survival_probability
  )
  count_survivors = table(factor(x = sample_survivors, levels = seed_values))
  relative_frequency = as.numeric(count_survivors) / n_quadrats
  cumulative_relative_frequency = cumsum(relative_frequency)

  plot(
    x = seed_values, y = model_mass, type = "h", lwd = 2,
    xlab = "Surviving seeds per quadrat", ylab = "Mass / relative frequency",
    ylim = c(0, max(model_mass, relative_frequency)),
    main = paste(n_quadrats, "quadrats")
  )
  points(x = seed_values, y = relative_frequency, col = "firebrick", pch = 16)
  legend(
    x = "topright", legend = c("Model", "Sample"),
    col = c("black", "firebrick"), lty = c(1, NA), pch = c(NA, 16),
    bty = "n"
  )
  plot(
    x = seed_values, y = model_cumulative, type = "s",
    xlab = "Surviving seeds per quadrat", ylab = "Cumulative probability",
    ylim = c(0, 1), main = paste(n_quadrats, "quadrats")
  )
  lines(x = seed_values, y = cumulative_relative_frequency,
        type = "s", col = "firebrick", lty = 2)
  legend(
    x = "bottomright", legend = c("Model", "Sample"),
    col = c("black", "firebrick"), lty = c(1, 2), bty = "n"
  )
}
par(old_par)
```

:::: {.callout-important title="Practice: orchid flowering across plots"}

Suppose each restoration plot contains 12 orchid plants. Each plant flowers with
probability 0.25, independently of the others. Let $X$ be the number of flowering
plants in a plot. This is a binomial model. Simulate counts for 12 plots and for 500
plots. For each sample, calculate the empirical mean and plot the empirical distribution.
Compare both distributions with the theoretical probability mass function. What changes in
the result from the larger sample? How much can you rely on a single simulation?

::: {.callout-tip title="Solution" collapse="true"}

One possible simulation is:

```r
set.seed(246)

plants_per_plot = 12
flowering_probability = 0.25

small_plot_counts = rbinom(
  n = 12,
  size = plants_per_plot,
  prob = flowering_probability
)
large_plot_counts = rbinom(
  n = 500,
  size = plants_per_plot,
  prob = flowering_probability
)

mean(small_plot_counts)
mean(large_plot_counts)

count_values = seq.int(from = 0, to = plants_per_plot)
small_plot_table = table(factor(
  x = small_plot_counts,
  levels = count_values
))
large_plot_table = table(factor(
  x = large_plot_counts,
  levels = count_values
))

small_frequency = as.numeric(small_plot_table) / length(small_plot_counts)
large_frequency = as.numeric(large_plot_table) / length(large_plot_counts)
theoretical_mass = dbinom(
  x = count_values,
  size = plants_per_plot,
  prob = flowering_probability
)

plot(
  x = count_values,
  y = small_frequency,
  type = "h",
  lwd = 2,
  ylim = c(0, max(c(small_frequency, large_frequency, theoretical_mass))),
  xlab = "Flowering plants per plot",
  ylab = "Probability"
)
points(x = count_values, y = large_frequency, col = 2, pch = 16)
points(x = count_values, y = theoretical_mass, pch = 4)
legend(
  x = "topright",
  legend = c("12 plots", "500 plots", "Binomial model"),
  col = c(1, 2, 1),
  pch = c(NA, 16, 4),
  lty = c(1, NA, NA)
)
```

The larger sample will generally give a more stable empirical distribution and mean
closer to the theoretical values. A particular simulation need not be closer in every
summary, and the empirical distribution need not exactly match the theoretical one.

:::
::::

# Probability distributions for count data {#sec-discrete_distributions}

We now examine three models for counts: binomial, Poisson, and negative
binomial. For each, check which counts are possible, what its parameters mean,
how its mean and variance relate, and whether its assumptions suit the sampling
process. The [distribution supplement](../Supplements/distributions.qmd)
covers further examples.

The probability mass function specifies each model's probabilities. Support
and ecological interpretation help us choose plausible candidates; after
fitting them to the same observations, we can use the comparison criteria in
[Chapter 7](../Chapter_7/Theory.qmd#sec-response-comparison).

::: {.callout-warning title="Choosing a probability distribution"}
A distribution should be chosen according to the type and range of observed values, its
mean–variance relationship, and how well it describes the observed data. Ecological
interpretation is useful, but a distribution is not restricted to only the process from
which it was originally derived.
:::

## The binomial distribution

The binomial distribution models the number of successes among a fixed number
of trials. Each trial has two possible outcomes, the same success probability,
and an outcome independent of the other trials. For example, a success could
mean that a planted seed survives.

The binomial model has two parameters:

- $n$: the number of trials in one sampling unit;
- $p$: the probability of success on each trial, with $0\leq p\leq 1$.

For $0<p<1$, the support of $X$ is $\{0, 1, \ldots, n\}$. At $p=0$ or
$p=1$, only one count has positive probability. Here $n$ is the number
of trials in one unit, such as seeds in a quadrat; it is different from
the number of quadrats in a sample.

The probability mass function is

$$
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x}
=\frac{n!}{x!(n-x)!}p^x(1-p)^{n-x},
$$

where $\binom{n}{x}$ counts the ways to choose $x$ successes from $n$ trials
and $n!$ is a factorial, with $0!=1$.

The mean and variance are:

$$
E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$


### Example: seedling survival

Suppose 20 seedlings are planted in each plot and each survives independently
with probability $0.7$. Let $X$ be the number surviving in one plot. This is
an assumed model, not a set of observed field data.

The probability that exactly 14 survive is:

$$
P(X=14)
=
\binom{20}{14}(0.7)^{14}(1-0.7)^{20-14}
=
\frac{20!}{14! (20 - 14)!}(0.7)^{14}(0.3)^6
\approx 0.1916.
$$

We can verify our math with R:

```{r}
seedling_count = 20L
seedling_survival_probability = 0.7
survivors_to_check = 14L

dbinom(
  x = survivors_to_check,
  size = seedling_count,
  prob = seedling_survival_probability
)
```

The model gives an expected count of $20(0.7)=14$ seedlings per plot and
a variance of $20(0.7)(0.3)=4.2$ seedlings squared:

```{r}
expected_survivors = seedling_count * seedling_survival_probability
variance_survivors = seedling_count *
  seedling_survival_probability *
  (1 - seedling_survival_probability)
```

If a binomial mean and variance are known, we can invert their formulas.
Writing $m=E[X]$ and $v=\operatorname{Var}(X)$ gives

$$
\begin{aligned}
p &= 1-\frac{v}{m}, \\
n &= \frac{m}{p}.
\end{aligned}
$$

This conversion requires $m>0$, a probability $0<p\leq 1$, and a positive
integer result for $n$. For the model above, it recovers $p=0.7$ and $n=20$:

```{r}
mean_survivors = expected_survivors
success_probability = 1 - variance_survivors / mean_survivors
estimated_trial_count = mean_survivors / success_probability
```

Sample means and variances will not generally recover exact model parameters.
[Chapter 4](../Chapter_4/Theory.qmd#sec-estimates) uses sample summaries to
construct estimates and explains why estimates vary between samples.

## The Poisson distribution

The Poisson distribution is a starting model for counts in a fixed unit of
time, area, or sampling effort. For $\lambda>0$, its support is
$\{0, 1, 2, \ldots\}$; at $\lambda=0$, only zero has positive probability.
Under the model, the mean and variance are equal. A sample with similar
mean and variance is consistent with that feature, but does not establish
that the Poisson assumptions hold.

The Poisson distribution has one parameter:

$$
\lambda=\text{expected count per sampling unit}, \qquad \lambda\geq 0.
$$

Its probability mass function is

$$
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!},
$$

The mean and variance are both equal to $\lambda$:

$$
E[X]=\lambda
$$

and

$$
\operatorname{Var}(X)=\lambda.
$$


### Example: wildlife detections

Suppose a camera records an average of $0.8$ detections per night. If the
expected detection count scales with observation time, three nights have
expected count

$$
\lambda = 3(0.8) = 2.4.
$$

Under an assumed Poisson model, the probability of exactly two detections
over those three nights is

$$
P(X=2)
=
\frac{e^{-2.4}(2.4)^2}{2!}
\approx 0.2613.
$$

We can calculate it in R:

```{r}
rate_per_night = 0.8
n_nights = 3
expected_detections = rate_per_night * n_nights
detections_to_check = 2L

dpois(x = detections_to_check, lambda = expected_detections)
```

"At least five" means more than four, so we use `lower.tail = FALSE`:

```{r}
minimum_detections = 5
ppois(
  q = minimum_detections - 1,
  lambda = expected_detections,
  lower.tail = FALSE
)
```

Unlike a binomial count, a Poisson count has no fixed upper limit. A count
of 100 detections is possible under this model, although extremely unlikely
when $\lambda=2.4$. Detections need not correspond to distinct animals.

```{r}
#| echo: false
#| fig-cap: "Poisson probability mass for detections during three camera nights."
expected_detections = 2.4
maximum_detections_to_plot = 15
detection_values = seq.int(from = 0, to = maximum_detections_to_plot)
detection_mass = dpois(x = detection_values, lambda = expected_detections)

plot(
  x = detection_values,
  y = detection_mass,
  type = "h",
  lwd = 3,
  xlab = "Number of detections",
  ylab = "Probability",
  main = "Poisson probability mass"
)
```

:::: {.callout-important title="Practice: flowering stems per orchid plot"}

A survey covers two square metres of a restoration site. Assume flowering stems occur
at an average rate of 1.2 per square metre. Calculate the probability of finding exactly
three stems in the surveyed area. Why does this model allow counts greater than three,
even though three is the count in the question?

::: {.callout-tip title="Solution" collapse="true"}

The expected count across two square metres is $\lambda=2(1.2)=2.4$. The probability is:

```r
rate_per_square_metre = 1.2
survey_area = 2
expected_stems = rate_per_square_metre * survey_area

dpois(x = 3, lambda = expected_stems)
```

The Poisson model has support $0,1,2,\ldots$, so counts larger than three are possible.
The probability of exactly three is approximately 0.209, but the model does not set
three as a maximum.

:::
::::

## Overdispersion

The Poisson model has equal mean and variance. Counts that vary more than this
model predicts are **overdispersed relative to the Poisson model**. For example,
seedling counts may differ strongly among quadrats if some have better habitat.
The negative binomial distribution allows greater variance. Overdispersion
can also reflect detection differences or dependence, so we must examine
the ecological and observation processes before choosing a replacement model.

::: {.callout-warning title="Overdispersion is a statistical description"}
Overdispersion means that the observed variation is greater than the variation expected
under a reference model, often the Poisson model. It does not identify the ecological cause.
Clustering, habitat differences, omitted predictors, detection variation, and dependence
can all produce overdispersion.
:::

### A simple diagnostic for overdispersion

A first exploratory check compares the sample variance from `var()` with
the sample mean. If the observations are reasonably described as independent
Poisson counts with the same mean, these quantities tend to be similar.
The **dispersion ratio** summarises their comparison:

$$
\text{dispersion ratio}
=
\frac{\text{sample variance}}{\text{sample mean}}.
$$

Here we simulate quadrat counts from a negative binomial model to see what
the ratio can look like when variation exceeds a Poisson model's prediction:

```{r}
set.seed(seed = 123L)

n_quadrats = 100L
negative_binomial_size = 1.5
negative_binomial_mean = 4

counts = rnbinom(
  n = n_quadrats,
  size = negative_binomial_size,
  mu = negative_binomial_mean
)

sample_mean = mean(counts)
sample_variance = var(counts)
dispersion_ratio = sample_variance / sample_mean
c(sample_mean, sample_variance, dispersion_ratio)
```

A ratio close to one is consistent with the Poisson mean–variance relationship;
a substantially larger ratio suggests overdispersion relative to that model.
A ratio below one suggests less variation than the Poisson model predicts.
The ratio alone cannot identify the cause or confirm a distribution.
[Chapter 7](../Chapter_7/Theory.qmd#sec-response-comparison) introduces
criteria for comparing fitted candidate models.

:::: {.callout-important title="Practice: variation in orchid stem counts"}

Across a sample of restoration plots, the mean flowering-stem count is 3 and the sample
variance is 12. Calculate the dispersion ratio and describe what it suggests relative to
a Poisson model. Can this ratio tell you why the counts vary so much?

::: {.callout-tip title="Solution" collapse="true"}

The dispersion ratio is $12/3=4$. It suggests that the observed variation is greater
than expected under a Poisson model with the same mean. The ratio alone cannot identify
the cause; for example, plots may differ in habitat, or counts may be clustered.

:::
::::

## The negative binomial distribution

The negative binomial distribution is a starting model for counts with more
variation than a Poisson model with the same mean. Its support is
$\{0, 1, 2, \ldots\}$. Differences among sampling units are one possible
reason for the extra variation, though this distribution alone does not
identify its cause.

The negative binomial model has two parameters:

- $\mu>0$: the mean count per sampling unit;
- $k>0$: a shape parameter that controls the additional variation.

The probability mass function is

$$
P(X=x)
=
\frac{\Gamma(x+k)}{\Gamma(k)x!}
\left(\frac{k}{k+\mu}\right)^k
\left(\frac{\mu}{k+\mu}\right)^x,
$$

where $\Gamma$ is the gamma function, extending the factorial in this formula.

The mean and variance are:

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k}.
$$

For $\mu>0$ and finite $k$, the variance exceeds the mean. Smaller $k$
gives greater variance at a fixed $\mu$. As $k$ increases, the variance
approaches the mean:

$$
\mu+\frac{\mu^2}{k}\longrightarrow\mu.
$$

Thus, the negative binomial distribution approaches the Poisson distribution
as $k$ becomes large.

In R, `dnbinom()` takes $k$ through its `size` argument and $\mu$ through
`mu`. Its `pnbinom()` and `rnbinom()` counterparts provide cumulative
probabilities and simulated counts.

### Example: seedlings in quadrats

Suppose the mean number of seedlings per quadrat is $\mu=4$, and the
shape parameter is $k=1.5$. These are assumed model values.

The expected number of seedlings is

$$
E[X]=4.
$$

The variance is

$$
\operatorname{Var}(X)
=
4+\frac{4^2}{1.5}
\approx 14.67.
$$

Across comparable quadrats, the model predicts four seedlings on average
and variance of approximately $14.67$ seedlings squared. This variation
does not by itself identify its ecological cause.

We can verify these calculations in R:

```{r}
mean_seedlings = 4
negative_binomial_size = 1.5

expected_seedlings = mean_seedlings
variance_seedlings = mean_seedlings +
  mean_seedlings^2 / negative_binomial_size

c(mean = expected_seedlings, variance = variance_seedlings)
```

The probability of observing exactly 10 seedlings is:

```{r}
seedlings_to_check = 10

dnbinom(
  x = seedlings_to_check,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

"At least ten" means more than nine, so we use `lower.tail = FALSE`:

```{r}
minimum_seedlings = 10

pnbinom(
  q = minimum_seedlings - 1,
  size = negative_binomial_size,
  mu = mean_seedlings,
  lower.tail = FALSE
)
```

Like the Poisson distribution, the negative binomial distribution has no fixed
upper limit. A count of 100 is mathematically possible, although extremely
unlikely.

```{r}
#| echo: false
#| fig-cap: "Negative binomial probability mass for seedlings in a quadrat."
mean_seedlings = 4
negative_binomial_size = 1.5
maximum_seedlings_to_plot = 20
seedling_values = seq.int(from = 0, to = maximum_seedlings_to_plot)
negative_binomial_mass = dnbinom(
  x = seedling_values,
  size = negative_binomial_size,
  mu = mean_seedlings
)

plot(
  x = seedling_values,
  y = negative_binomial_mass,
  type = "h",
  lwd = 3,
  xlab = "Number of seedlings",
  ylab = "Probability",
  main = "Negative binomial probability mass"
)
```

:::: {.callout-important title="Practice: orchid stem counts with different dispersion"}

Suppose the mean number of flowering stems per plot is $\mu=3$. Calculate the negative
binomial variance for $k=1.5$ and $k=6$. Which value gives greater overdispersion? How
does increasing $k$ relate to the Poisson model?

::: {.callout-tip title="Solution" collapse="true"}

Using $\operatorname{Var}(X)=\mu+\mu^2/k$, the variances are 9 and 4.5, respectively.
The smaller value of $k$ gives greater overdispersion. As $k$ increases, the variance
approaches the mean, which is the Poisson mean–variance relationship.

:::
::::

# Chapter quiz

Try each question before opening its solution.

:::: {.callout-important title="Question 1: lichen patches"}

On a fixed section of rock, let $X$ be the number of lichen patches found.
Suppose every count from zero through 18 is possible, but no larger count
is possible. What is the support of $X$? Is it finite or countably infinite?

::: {.callout-tip title="Answer" collapse="true"}

$X$ can take the integer values 0 through 18. Its support is finite because the survey
area has a maximum number of patches in this scenario.

:::
::::

:::: {.callout-important title="Question 2: freshwater larvae"}

A proposed distribution for the number of larvae in a sample assigns probabilities
0.15, 0.25, 0.40, and 0.30 to counts 0, 1, 2, and 3. Is this a valid probability
distribution? Explain.

::: {.callout-tip title="Answer" collapse="true"}

No. Although each probability is non-negative, they sum to 1.10 rather than 1.
Probabilities over all possible outcomes must sum to one.

:::
::::

:::: {.callout-important title="Question 3: coral spawning"}

For a model of the number of coral colonies spawning on a reef, which R function prefix
would you use to calculate (a) the probability of exactly five colonies, (b) the
probability of five or fewer, (c) the 0.9 quantile, and (d) a random sample?

::: {.callout-tip title="Answer" collapse="true"}

Use the `d` prefix for an exact probability, `p` for a cumulative probability, `q` for
a quantile, and `r` to generate a random sample.

:::
::::

:::: {.callout-important title="Question 4: acorn production"}

Suppose acorn production by two oak trees is modelled with independent random variables.
In this model, what does independence say about the probability of a particular joint
outcome? Give one ecological factor that might make the assumption questionable.

::: {.callout-tip title="Answer" collapse="true"}

Under independence, the probability of a particular pair of outcomes is the product of
their separate probabilities. Shared weather or soil conditions could make production
by the two trees dependent.

:::
::::

:::: {.callout-important title="Question 5: trout eggs"}

For a group of 12 trout eggs, suppose each egg hatches with probability 0.25 and the
outcomes are independent. Calculate the expected number hatched and the variance.

::: {.callout-tip title="Answer" collapse="true"}

For a binomial model, $E[X]=np=12(0.25)=3$ and
$\operatorname{Var}(X)=np(1-p)=12(0.25)(0.75)=2.25$.

:::
::::

:::: {.callout-important title="Question 6: mussel settlement"}

A model predicts an expected count of 2.4 newly settled mussels per tile. Can one tile
have 2.4 mussels? What does the expected value describe?

::: {.callout-tip title="Answer" collapse="true"}

A tile cannot have a fractional mussel. The expected value describes the long-run average
count across comparable tiles under the model.

:::
::::

:::: {.callout-important title="Question 7: moth emergence"}

In a sample of 30 host plants, moths emerged from 9. Calculate the relative frequency
of emergence. Is this value necessarily the true probability of emergence for all
relevant host plants?

::: {.callout-tip title="Answer" collapse="true"}

The relative frequency is $9/30=0.3$. It is a sample-based estimate of the probability,
not necessarily the exact population probability. Another sample could produce a
different relative frequency.

:::
::::

:::: {.callout-important title="Question 8: caddisfly larvae"}

You sample the number of caddisfly larvae in many equal-sized stream sections. As the
number of sections grows, what do you generally expect to happen to the empirical
distribution relative to the distribution that generated the data? Is exact agreement
expected for a finite sample?

::: {.callout-tip title="Answer" collapse="true"}

With more independent observations from the same distribution, the empirical distribution
generally becomes closer to the generating distribution. Exact agreement is not expected
in a finite sample, and individual simulations can vary.

:::
::::

:::: {.callout-important title="Question 9: choose a count model"}

Choose a plausible starting model for each scenario and briefly explain your choice:

1. The number of eggs hatching out of 15 eggs, with the same hatch probability for each.
2. The number of mayfly larvae in a standard sample, when the count is unbounded and
   its mean and variance are approximately equal.
3. The number of burrowing crabs per beach section, when counts are unbounded and the
   sample variance is substantially greater than the mean.

::: {.callout-tip title="Answer" collapse="true"}

1. A binomial model is plausible if the egg outcomes are independent, because
   there is a fixed number of trials and a common hatch probability.
2. A Poisson model is a plausible starting point because the support is non-negative
   integer counts and the mean and variance are approximately equal.
3. A negative binomial model is a plausible starting point because it allows an
   unbounded count with variance greater than the mean. These are starting points; the
   data and assumptions still need to be checked.

:::
::::

:::: {.callout-important title="Question 10: bumblebee nests"}

Across meadow plots, the sample mean number of bumblebee nests is 2 and the sample
variance is 6. Calculate the dispersion ratio and explain what it suggests relative to
a Poisson model. Does it identify the ecological cause?

::: {.callout-tip title="Answer" collapse="true"}

The ratio is $6/2=3$, suggesting greater variation than expected under a Poisson model
with the same mean. The ratio does not identify a cause; clustering, habitat differences,
or dependence could contribute.

:::
::::
