---
title: "Introduction to Probability (Count data)"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Introduction

This chapter introduces random variables, probability distributions and how they related to
samples of data. We will focus only on  **discrete random variables** that are used in ecology
to represent count data. These are variables that take separate, countable values, such as $0,1,2,\ldots$.
In the next chapter we will extend this concept to continuous data.

# Learning goals

After studying this chapter, you should be able to:

- describe a discrete random variable and its probability distribution
- interpret probability mass, cumulative probability, samples, and quantiles
- explain independence of random variables
- calculate and interpret means and variances
- understand how samples related to populations and asymptotic convergence
- simulate ecological data and compare theoretical and empirical distributions
- recognize and describe overdispersion
- recognize situations where binomial, Poisson, and negative binomial distributions may be useful

# Random variables and probability distributions

A **random variable** is a variable whose value is uncertain before an observation is made.
As an example to help us understand these concepts, we will assume a field experiment on
seed survival. Imagine we place 4 seeds in specific locations on our experimental field
(called *quadrats*) and come back the next day to count how many seeds have survived. We
define $X$ as:

$$
X=\text{the number of seeds surviving in a quadrat}.
$$

If the quadrat contains four seeds, then $X$ can take the values

$$
0,1,2,3,4.
$$

The possible values form the **sample space** of $X$.

The **support** of a random variable is the set of values that have positive probability
under its model. For this seed-survival example, the support is the finite set
$\{0,1,2,3,4\}$. A discrete random variable can also have a **countably infinite**
support: its values can be listed one by one, but the list has no final value. For
example, a model for a count with no fixed upper limit may have support
$\{0,1,2,3,\ldots\}$. The values can be counted in order, even though there are
infinitely many of them.

::: {.callout-note title="Discrete random variable"}
A discrete random variable takes values from a finite or countably infinite set. Its
probability distribution assigns a probability to each possible value, and the
probabilities sum to one.
:::

For each possible value of $X$ we can assign a probability. For example:

| $x$ | $P(X=x)$ |
|---:|---:|
| 0 | 0.05 |
| 1 | 0.15 |
| 2 | 0.30 |
| 3 | 0.30 |
| 4 | 0.20 |

All probabilities must be **non-negative**:

$$
P(X=x)\geq 0
$$

and they must add up to one:

$$
\sum_x P(X=x)=1.
$$

The notation $X$ refers to the random variable, while $x$ refers to one possible or observed value
of that variable.

Probabilities can also be assigned to sets of possible values (see later @sec-independence_expectations for details). For example:

$$
P(X \leq 3) = \sum_{x=0}^{x=3} P(X=x)
$$

We generally do not work with tables of probabilities but instead use mathematical functions
that can calculate the probability for each individual value of $X$. These are called
probability mass functions (see @sec-four_faces).

::: {.callout-note title="Probability distribution"}
A probability distribution is a collection of related functions (see @sec-four_faces) that
describe a theoretical process by which a random variable can take different values. The
probability mass function is only one of the faces of a probability distribution, but often
people confuse the two.
:::

In the remainder of this section we will learn the different
main properties of discrete random variables and probability distributions for discrete
data, using the Binomial distribution as an example. In @sec-discrete_distributions we will
learn more about the Binomial distribution (including the formula for the probability mass)
and other distributions and some tips on how to choose a reasonable distribution for your
data (with more formal methods coming later in the course).

## The four faces of a discrete distribution {#sec-four_faces}

### 1. Probability mass

The probability mass assigned to $X=x$ is

$$
P(X=x).
$$

It answers a question such as:

> What is the probability of observing exactly four seeds?

This is the most importance face of a distribution and the one which *shape* we would
recognize. We will use probability mass functions throughout the course as the basis for
parameter estimation, comparing alternative models and checking how well a model fits to data.

A probability mass function can be displayed with a bar plot or a *histogram-like* plot.
For example, if we model the seed survival process with a Binomial distribution where each
seed has a probability of surviving of 0.4 it would look as follows.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Binomial probability mass (left: barplot, right: histogram-like)"
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
seed_mass = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)

par_val = par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)

barplot(
  height = seed_mass,
  names.arg = seed_values,
  xlab = "Number surviving",
  ylab = "Probability"
)

plot(
  x = seed_values,
  y = seed_mass,
  type = "h",
  lwd = 3,
  xlab = "Number surviving",
  ylab = "Probability"
)
par(par_val)
```

In R, the probability mass is implemented by functions that start with the prefix `d` followed
by the root name used for a distribution. For example, the probability mass assigned to $X = 2$
for the binomial model above can be calculated in R as:

```{r}
dbinom(x = 2, size = seed_count, prob = seed_survival_probability)
```

### 2. Cumulative probability

The cumulative probability function computes the probability for all possible values of $X = x$
(also know as **tail probability**):

$$
F_X(x)=P(X\leq x).
$$

It answers questions such as:

> What is the probability of observing four or fewer surviving seeds?

We will seldom use this type of function directly. However, the two types of functions
described below (*quantile* and *sampling* functions) are based on the cumulative probability function
so it is still important to understand it. It is also important in traditional statistical
procedures.^[A p-value is a tail probability calculated under a specified null model. We
will ignore p-values in this course.]

Cumulative probability functions of discrete variables are best visualized using a *staircase-like*
plot:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Binomial cumulative probability"
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
seed_cumulative = pbinom(
  q = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)

plot(
  x = seed_values,
  y = seed_cumulative,
  type = "s",
  xlab = "Number surviving",
  ylab = "Cumulative probability",
  ylim = c(0, 1)
)

```

The cumulative probability is always between zero and one, and it never decreases as $x$ increases.

We can compute cumulative probabilities in R by using functions with the prefix `p` followed
by the root name used for a distribution. For example, the cumulative probavility assigned to $X \leq 2$
for the binomial model above can be calculated in R as:

```{r}
pbinom(q = 2, size = seed_count, prob = seed_survival_probability)
```

We will often want to also compute the complement (i.e., cumulative probability that $X > x$).
R offers a shortcut to do that by setting `lower.tail = FALSE` inside the
relevant `p` function:

```{r}
pbinom(
  q = 2,
  size = seed_count,
  prob = seed_survival_probability,
  lower.tail = FALSE
)
```

which is equivalent to

```{r}
1 - pbinom(q = 2, size = seed_count, prob = seed_survival_probability)
```

### 3. Quantiles

A quantile function is the inverse of the cumulative probability function. That is, while the
cumulative probability function gives the cumulative probability for a given $x$ value, the
quantile function gives you the $x$ value that corresponds to a given cumulative probability.

In mathematical terms, the $\alpha$-quantile of a distribution is defined as

$$
q_\alpha=\min\{x:F_X(x)\geq\alpha\}.
$$

We will use quantile functions in one of the methods to estimate variability in parameter estimates
(we cover this in [Chapter 5](../Chapter_5/Theory.qmd)). It is also the basis for computing
random samples from a distribution (see below).

We can visualize the quantile function just like we did for the cumulative distribution but
inverting the axes. For the same example as before that would be:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Binomial quantiles"
x = 0:seed_count
cumulative = pbinom(x, size = seed_count, prob = seed_survival_probability)

plot(
  cumulative, x,
  type = "s",
  xlab = "Cumulative probability",
  ylab = "Number surviving",
  xlim = c(0, 1)
)

```

We can compute quantiles in R by using functions with the prefix `q` followed
by the root name used for a distribution. For example, the quantile that corresponds to
a cumulative probability of 0.5 for the binomial model of seed survival is computed as:

```{r}
median_probability = 0.5
qbinom(
  p = median_probability,
  size = seed_count,
  prob = seed_survival_probability
)
```

### 4. Sampling

A very important step in our analyses will be to generate possible samples of data
from our models (see below @sec-sample_population as well as [Chapter 5](../Chapter_5/Theory.qmd)).
In a random sample, each element is one possible realization of the random variable.
Conceptually the procedure to generate random samples is as follows:

1. We generate a random number between 0 and 1 using a standard random number generator.

2. We input this number into the quantile function of the distribution and take the result as
the realization of the random variable

We repeat this procedure for every element in the sample we want to generate. As long as we
are using a good quality random number generator (R will take care of that), the samples that
result from this approach will have the desired statistical properties (see @sec-sample_population).

A sampling function cannot be visualized directly, since it is a procedure that generate
random samples (rather than a function that relates two variables). We discuss how to
visualize samples in @sec-sample_population.

We can generate a random sample in R by using functions with the prefix `r` followed
by the root name used for a distribution. We should always preced any random sample generation
with a call to the function `set.seed` with any number inside. This will make sure that we
will always get the same sample so that our R scripts are reproducible (if you change the
seed number, a different random sample will be generated). For example, we can generate a
10 random values from the binomial distribution of seed survival as follows:

```{r}
n_simulated_quadrats = 10
set.seed(seed = 42)
sample_survivors = rbinom(
  n = n_simulated_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
sample_survivors
```

The sample contains 10 simulated values. Each value is a simulated quadrat with 4 seeds. In
the first simulated quadrat 3 seeds survived. In the third quadrat only 1 seed survived.

::: {.callout-note title="The four faces of distributions in R"}
For a distribution in R:

- `d*` functions calculate probability mass
- `p*` functions calculate cumulative probabilities
- `q*` functions calculate quantiles
- `r*` functions generate random samples

The first argument is conventionally named `x`, `q`, `p`, or `n`, respectively.
:::

:::: {.callout-important title="Practice: orchid fruit set"}

In a restoration plot, six tagged orchid flowers each have probability $0.5$ of setting
fruit. Let $X$ be the number of these flowers that set fruit.

1. What values can $X$ take?
2. Use R to calculate the probability that exactly four flowers set fruit and the
   probability that at most four set fruit.
3. Find the 0.75 quantile of $X$. In words, what does this quantile tell you?
4. Which R function prefix (`d`, `p`, `q`, or `r`) did you use for each calculation?

::: {.callout-tip title="Solution" collapse="true"}

The count can be any integer from 0 to 6. The exact-count probability uses `dbinom()`,
the cumulative probability uses `pbinom()`, and the quantile uses `qbinom()`.

```r
flower_count = 6
fruit_set_probability = 0.5

dbinom(x = 4, size = flower_count, prob = fruit_set_probability) # 0.23
pbinom(q = 4, size = flower_count, prob = fruit_set_probability) # 0.89
qbinom(p = 0.75, size = flower_count, prob = fruit_set_probability) # 4
```

Under this model, at least 75% of the probability is on counts of four or fewer.

:::
::::

## Independence and Expectations {#sec-independence_expectations}

### Independence

Two random variables are independent if knowing the value of one does not change the
probabilities for the other. For examples, let's assume that we have two seeds in a
germination experiment. We model
their germination on a given day of the experiment as an stochastic process. Each seed is
assigned a random variable with two possible values: $X_1 = \{0, 1\}$ and  $X_2 = \{0, 1\}$
where $0$ means dead and $1$ means alive. If we assume that the probability of one seed
germinating is independent of whether another seed has germinated we say that $X_1$ and $X_2$
are independent.

When we have two random variables we can ask what is the probability of a particular combination
of values. This is denoted with the symbol $\cap$. For example, the probability that seed 1
germinates while seed 2 does not would be written as $P(X_1 = 1 \cap X_2 = 0)$,

When two (or more) random variables are said to be independent, the probability of a combination
of possible values is equal to the product of probabilities of each variable. For our example,
the probability that $X_1 = x_1$ and $X_2 = x_2$ at the same time would be:
$$
P(X_1 = x_1 \cap X_2 = x_2) = P(X_1 = x_1)P(X_2 = x_2).
$$

::: {.callout-warning title="Independence is a modeling assumption"}
Independence means that knowing the value of one random variable does not change the
probability distribution of another. It is not guaranteed merely because observations
were collected at different locations or times.
:::

In many cases, we can model our data under the assumption of independence, though not always.
Ecological observations may be dependent because of:

    - spatial clustering;
    - temporal autocorrelation;
    - competition;
    - shared environmental conditions;
    - social behavior;
    - repeated measurements of the same individual or group.

We will address the last source of dependence in [Chapter 8](../Chapter_8/Theory.qmd)

#### Independent and identically distributed random variables.

We will often assume that different events or observations in our datasets are independent
but also that they can be modelled by the same distribution. In this case we say that the
random variables associated to the different events are *i.i.d* (independent and identically
distributed). This allows to simply greatly calculations of probability and will be the basis
for the methods of parameter estimation that we will see later.

Going back to our example of the seeds, if the probability of germination for each seed is $p_g$
the probability that on a given day they both germinate is:
$$
P(X_1 = 1 \cap X_2 = 1) = p_g\cdot p_g = p_g^2.
$$

### Expectations

Expectations are values that describe the *shape* of a distribution and they include properties
like the mean and variance. They are useful for several reasons:

- In statistics, often the goal is to estimate the mean and variance of a distribution from
observed data (assuming we choose a good distribution for the data).

- Many concepts in statistics are related to expectations of statistical estimators (see [Chapter 4](../Chapter_4/Theory.qmd)).

- The simplest method of parameter estimation match expectations of samples and distribution (see [Chapter 4](../Chapter_4/Theory.qmd))

- Expectations also underpin advanced methods for estimating model parameters.^[For example,
Markov Chain Monte Carlo methods use simulated samples to approximate expectations under a
probability model.]


#### Expected value

For a discrete random variable $X$, the expected value is a probability-weighted average
of all possible values. That is, we sum all possible values of $X$ weighted by their
individual probabilities.

$$
E[X] = \sum_x xP(X=x).
$$

For the binomial model of seed survival, the expected value can be calculated in R as

```{r}
seed_values = seq.int(from = 0, to = seed_count)
seed_probability = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
expected_value = sum(seed_values * seed_probability)
```

Often we will refer to the expected value as the **mean** of the distribution and denoted by
$\mu$.

We can generalize this approach to any function $f(X)$ into a general expectation *operator*
that computes weighted-average of the results of applying $f(X)$:

$$
E[f(X)] = \sum_x f(X=x)P(X=x).
$$

::: {.callout-note title="Expected value is a long-run average"}
The expected value is a theoretical property of a probability distribution. It need not
be an observable value for an individual observation. For example, a quadrat cannot
contain 1.6 surviving seeds, even though the expected number may be 1.6.
:::

#### Variance and standard deviation

A special family of expected values are those calculated for functions of the form

$$
f(X) = (X - E[X])^n
$$

That is, we subtract the expected value (mean, $E[x]$) from $X$ and raise the result to an integer
power ($n = 1, 2, ...$). These values are called the **moments** around the mean of the
distribution. They give information on how probabilities are spread around the mean of the distrbution.
Although one can have an infinite number of moments, we will focus mostly on the **variance** ($n = 2$):

$$
\operatorname{Var}(X) = E[(X-E[X])^2] = \sum_x P(X=x)(x-E[X])^2.
$$

The variance tells us about how probability spread around the mean. The lower the variance,
the more concentrated probability around the mean such that values far from the mean are
assigned very little probability.

We can calculate moments in R easily, following the sample example as before:

```{r}
variance = sum((seed_values - expected_value)^2 * seed_probability)
```

A common way to express variance is by taking its square root. This is know as the standard
deviation:
$$
\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.
$$

::: {.callout-note title="Variance and standard deviation"}
Variance measures the spread of a distribution in squared units. Standard deviation is
the square root of variance and has the same units as the original variable, making it
usually easier to interpret.
:::

:::: {.callout-important title="Practice: fruit set on orchid plants"}

Suppose eight tagged flowers on each of two orchid plants have the same fruit-set
probability, $p=0.6$. Let $T$ be the total number of flowers setting fruit across
both plants. For this exercise, model the outcomes of the 16 flower as independent.

1. Use the probability mass function to find the probability of every possible value of
   $T$.
2. Calculate the mean and variance of $T$ using the probability mass function from point 1.
3. Explain what the mean represents. Give one ecological reason why independence among
   the flowers might be doubtful.

::: {.callout-tip title="Solution" collapse="true"}

The possible counts range from 0 to 16. First calculate the probability of each count,
then use the weighted-sum formulas for the mean and variance:

```r
fruit_count_values = seq.int(from = 0, to = 16)
fruit_probabilities = dbinom(
  x = fruit_count_values,
  size = 16,
  prob = 0.6
)

expected_fruit_count = sum(fruit_count_values * fruit_probabilities)
fruit_count_variance = sum(
  (fruit_count_values - expected_fruit_count)^2 * fruit_probabilities
)

expected_fruit_count
fruit_count_variance
```

The mean is 9.6 and the variance is 3.84. The mean is the
long-run average fruit count across comparable groups of 16 flowers; a particular
group must have an integer count. Independence means one flower's outcome does not
change the probability distribution for another. However, the description indicates that
flowers are grouped as pairs within plants. The two flowers on a plant would be affected by
the physiological status, diseases or pollinator activity of that plant, but this would not
affect other flowers in other plants. This means that knowing something about one flower
gives us some information about the other flower on the same plant (at least more than about
other flowers in other plants). With this information, the assumption of independence no
longer seems suitable. We will learn how to deal with this type of dependence in [Chapter 8](../Chapter_8/Theory.qmd).
:::
::::

# Samples and populations {#sec-sample_population}

So far we have discussed random variables and distributions which are the mathematical tools
used in statistics to model variation and measurement errors in our data. But how does our
data related to these distributions?

::: {.callout-note title="Sample and population"}
In statistics **we analyze data as if it was a random sample from a distribution**. This
distribution is often called **population** or **data generating process**.
:::

We do not literally believe this because:

1. The data may have been collected in a non-random manner or only partially random. What
matters when collectind data is that the sample is representative of the system we are
analyzing and that biases are minimal, not so much whether it is random or not.

2. Randomness in a statistical model represents uncertainty or unexplained variation; it
need not be a claim that nature is fundamentally random.^[Many scientists believe that nature
is deterministic and that randomness represents our inability to predict the future. Others
regard randomness as fundamental, particularly in interpretations of quantum physics. This
philosophical question is not needed for using probability models in ecology.]


3. Even if we belief the data is an actual random sample, we will never know the distribution
it was generated from (reality is ultimately unknownable).

So.. why do we make this assumption? Because it is **useful and practical**. We know how random
samples relate to distributions (this section) having a distribution allows us to generalize
from the specific data we collected, that is, we can try to answer questions that go beyond the
specific circumstances of our survey, experiment or measurement campaign.

Statistical models are approximations to reality, but they can still be useful when they
capture aspects relevant to our question.^[P. E. Box, a statistician from the XX century,
emphasized that no statistical model is exactly true in his quote "all models are wrong but
some are useful". Models are useful when their approximations are adequate for the purpose
at hand.]

## From frequency to probability

Given a sample of count data, we can compute the relative frequency at which each value appears.
For example, let's generate a random sample from our binomial distribution:

```{r}
set.seed(seed = 123)
n_quadrats = 10
sample_survivors = rbinom(
  n = n_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
```

The `table` function will tally how often each value appears and we can normalize it by the
sample size to calculate the relative frequency.

```{r}
seed_values = seq.int(from = 0, to = seed_count)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = count_survivors/length(sample_survivors)
relative_frequency
```
Note that for small samples some values may be missing (because they have low probability we
may not get to generate them). A trick to force `table()` to fill all possible values is to
first turn it into a `factor` and assign all possible values to the argument `levels`.

The relative frequencies add up to 1 and they will always be between 0 and 1. They actually
look like probabilities, right? Let's compare it to the numbers we get from the probability
mass function:

```{r}
data.frame(
  seed_count = seed_values,
  relative_frequency = as.numeric(x = relative_frequency),
  theoretical_probability = dbinom(
    x = seed_values,
    size = seed_count,
    prob = seed_survival_probability
  )
)
```

They are actually quite similar but not the same. This is not by coincide: relative
frequencies in a sample will converge towards the probability mass values of the distribution
they were sampled from. This result is related to the *Law of Large Numbers*

::: {.callout-note title="Frequencies converge to probabilities"}
We given a sample of size $n$ of indenpendent random values from the same distribution, the
relative  frequency of each possible value in the sample converges to the probablity mass
of the distribution as $n$ grows towards infinity.
:::

Indeed, if we generate much larger samples the relative frequencies will get closer and closer
to probabilities. Let's compare the relative frequencies for samples from size 10 to 100,000:

```{r}
set.seed(123)
sample_sizes = 10^(seq_len(length.out = 5))

all_frequencies = matrix(NA, nrow = seed_count + 1, ncol = 6)
colnames(all_frequencies) = c(1,2,3,4,5,"theory")

for(i in 1:5) {
    sample_survivors = rbinom(n = sample_sizes[i],
                              size = seed_count,
                              prob = seed_survival_probability)
    relative_frequency = table(factor(sample_survivors, levels = seed_values))/sample_sizes[i]
    all_frequencies[,i] = as.numeric(relative_frequency)
}
all_frequencies[,6] = dbinom(x = seed_values,
                             size = seed_count,
                             prob = seed_survival_probability)
round(all_frequencies, 3)
```

:::: {.callout-important title="Practice: fruit-set frequency"}
In a sample of 24 tagged orchid flowers, 15 set fruit. Calculate the relative frequency
of fruit set. Explain what this value estimates and why it need not equal the underlying
fruit-set probability exactly.

::: {.callout-tip title="Solution" collapse="true"}

The relative frequency is $15/24=0.625$. It estimates the probability that a flower sets
fruit in the population represented by the sample. A different sample could give a
different relative frequency, especially when the sample is small.
:::
::::

## Empirical distribution

Given the relationship between frequencies and probabilities, we can define a so-called
**empirical distribution** that will helps us understand the relationship between samples
and its population or data generating process.

::: {.callout-note title="Empirical distribution"}
An ad hoc distribution whose probability mass function matches exactly the relative
frequencies in a sample of data.
:::

We can visualize the empirical distribution in an analogoues way to how we visualize the
different faces of theoretical distribution. For example, the probability mass function
would be visualized with a barplot, histogram-like or a simple scatter plot and the
cumulative probability function with a staircase-like plot.

This can be useful to compare the empirical and theoretical distributions and the idea of
convergence. Below is an example for a small (n = 10) sample from the Binomial model of
seed survival.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Empirical (red) and theoretical distributions for n = 10. Left panels: mass, Right panels: cumulative)"
set.seed(142)
mass = dbinom(x = seed_values,
              size = seed_count,
              prob = seed_survival_probability)
cumprob = pbinom(q = seed_values,
                 size = seed_count,
                 prob = seed_survival_probability)
sample_survivors = rbinom(n = n_quadrats,
                          size = seed_count,
                          prob = seed_survival_probability)
count_survivors = table(factor(sample_survivors,levels = seed_values))
relative_frequency = as.numeric(count_survivors)/n_quadrats
cumulative_relative_frequency = cumsum(relative_frequency)

# The four panels
par_val = par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)
# Small samples first
plot(x, relative_frequency, type = "p", col = 2, xlab = "Number surviving",
     ylab = "Probability", xlim = c(0,seed_count), ylim = range(mass, relative_frequency),
     pch = 16, cex = 1.5)
lines(x, mass, lwd = 3, type = "h")
plot(x, cumulative_relative_frequency, col = 2, type = "s", xlab = "Number surviving",
     ylab = "Cumulative probability", ylim = c(0,1))
lines(x, cumprob, type = "s")
par(par_val)
```

### Mean, variance and quantiles of a sample

We can also calculate expectations and moments just like we did for the theoretical distributions.
Remember that the expected value (the *mean*) of a distribution is the probability-weighted
sum. For the empirical distribution, it turns out that this is equivalent to the
arithmetic mean of the sample:

```{r}
set.seed(123)
sample_survivors = rbinom(n = n_quadrats,
                          size = seed_count,
                          prob = seed_survival_probability)
count_survivors = table(factor(sample_survivors, levels = seed_values))
relative_frequency = count_survivors/n_quadrats
weighted_average = sum(relative_frequency*seed_values)
sample_mean      = mean(sample_survivors)
c(weighted_average, sample_mean)
```

Similarly, we can calculate the variance using the same procedure as before:

```{r}
variance = sum((seed_values - weighted_average)^2 * relative_frequency)
```

Note that this is not equal to calling `var` on the original sample:

```{r}
var(sample_survivors)
```

Why is that? The function `var` in R does not actually calculate the the empirical distribution
but an estimator of the variance of the population, which uses a different formula. Why that
is the case and whether it matters will be covered on [Chapter 4](../Chapter_4/Theory.qmd).

We can also calculate the quantiles of the empirical distribution using the `quantile` function
applied to the sample. For example, the sample median corresponds to the 50% quantile on
the empirical distribution:

```{r}
quantile(sample_survivors, prob = 0.5)
```

### Asymptotic convergence of the empirical distribution

::: {.callout-note title="Asymptotic convergence of empirical distribution"}
As the sample size $n$ grows, the empirical distribution will converge to the true distribution
from which the sample was generated. All quantiles, expectations and moments of the empirical
distribution will converge to the analogous values in the true distribution. Exact equivalence
only happens for infinite sample sizes.
:::

The asymptotic convergence of the empirical distribution is fundamental for understanding how
samples relate to populations and why sample size matters.^[The method of moments uses sample
moments to estimate distributional parameters. Bootstrapping is another application: it
repeatedly resamples the observed data to approximate the sampling variability of an estimator.]
This idea will come back multiple times throughout the course.

In this course we will not learn the exact mathematical formalism behind this idea as that
is not needed for understanding the idea itself. If you are interested in deepening your
mathematical understanding of statistics you should look at the Central Limit Theorem and
the Weak and Strong Law of Large Numbers. Studying these theorems will tell you exactly under
which conditions this behavior holds.

For the fixed-structure models considered in this course, this is the convergence behavior
we will use. Additional care is needed when model complexity or the sampling design changes
with sample size.^[Flexible methods such as Gaussian processes, smoothing splines, and random
forests can change their effective complexity as more data become available. These methods
are not covered in this course.]^[Convergence can also be more difficult to interpret when
increasing the sample size adds new groups or clusters to the population being studied. In
that case, the target population may change as the sample grows. This topic is discussed in
Chapter 8.]

We can visualize asymptotic convergence by comparing the probability and cumulative probability
functions of the empirical and theoretical distribution for different sample sizes.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Empirical (red) and theoretical distributions for n = 10 (top) and n = 1000 (bottom). Left panels: mass, Right panels: cumulative)"
set.seed(142)
# Experiment with n = 10
n_quadrats = 10
mass = dbinom(x = seed_values,
              size = seed_count,
              prob = seed_survival_probability)
cumprob = pbinom(q = seed_values,
                 size = seed_count,
                 prob = seed_survival_probability)
sample_survivors = rbinom(n = n_quadrats,
                          size = seed_count,
                          prob = seed_survival_probability)
count_survivors = table(factor(sample_survivors,levels = seed_values))
relative_frequency = as.numeric(count_survivors/n_quadrats)
cumulative_relative_frequency = cumsum(relative_frequency)
# Experiment with n = 1000
n_quadrats = 1000
sample_survivors_large = rbinom(n = n_quadrats,
                                size = seed_count,
                                prob = seed_survival_probability)
count_survivors_large = table(factor(sample_survivors_large,levels = seed_values))
relative_frequency_large = as.numeric(count_survivors_large/n_quadrats)
cumulative_relative_frequency_large = cumsum(relative_frequency_large)

# The four panels
par_val = par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(5.1, 4.1, 2, 0.5), las = 1)
# Small samples first
plot(x, relative_frequency, type = "p", col = 2, xlab = "Number surviving",
     ylab = "Probability", xlim = c(0,4), ylim = (range(mass, relative_frequency)),
     pch = 16, cex = 1.5)
lines(x, mass, lwd = 3, t = "h")
plot(x, cumulative_relative_frequency, col = 2, type = "s", xlab = "Number surviving",
     ylab = "Cumulative probability", ylim = c(0,1))
lines(x, cumprob, type = "s")
# Large samples second
plot(x, relative_frequency_large, type = "p", col = 2, xlab = "Number surviving",
     ylab = "Probability", xlim = c(0,4), ylim = (range(mass, relative_frequency_large)),
     pch = 16, cex = 1.5)
lines(x, mass, lwd = 3, t = "h")
plot(x, cumulative_relative_frequency_large, col = 2, type = "s", xlab = "Number surviving",
     ylab = "Cumulative probability", ylim = c(0,1))
lines(x, cumprob, type = "s")
par(par_val)
```

:::: {.callout-important title="Practice: orchid flowering across plots"}

Suppose each restoration plot contains 12 orchid plants. Each plant flowers with
probability 0.25, independently of the others. Let $X$ be the number of flowering
plants in a plot. This is a binomial model. Simulate counts for 12 plots and for 500
plots. For each sample, calculate the empirical mean and plot the empirical distribution.
Compare both distributions with the theoretical probability mass function. What changes in
the result from the larger sample? How much can you rely on a single simulation?

::: {.callout-tip title="Solution" collapse="true"}

One possible simulation is:

```r
set.seed(246)

plants_per_plot = 12
flowering_probability = 0.25

small_plot_counts = rbinom(
  n = 12,
  size = plants_per_plot,
  prob = flowering_probability
)
large_plot_counts = rbinom(
  n = 500,
  size = plants_per_plot,
  prob = flowering_probability
)

mean(small_plot_counts)
mean(large_plot_counts)

count_values = seq.int(from = 0, to = plants_per_plot)
small_plot_table = table(factor(
  x = small_plot_counts,
  levels = count_values
))
large_plot_table = table(factor(
  x = large_plot_counts,
  levels = count_values
))

small_frequency = as.numeric(small_plot_table) / length(small_plot_counts)
large_frequency = as.numeric(large_plot_table) / length(large_plot_counts)
theoretical_mass = dbinom(
  x = count_values,
  size = plants_per_plot,
  prob = flowering_probability
)

plot(
  x = count_values,
  y = small_frequency,
  type = "h",
  lwd = 2,
  ylim = c(0, max(c(small_frequency, large_frequency, theoretical_mass))),
  xlab = "Flowering plants per plot",
  ylab = "Probability"
)
points(x = count_values, y = large_frequency, col = 2, pch = 16)
points(x = count_values, y = theoretical_mass, pch = 4)
legend(
  x = "topright",
  legend = c("12 plots", "500 plots", "Binomial model"),
  col = c(1, 2, 1),
  pch = c(NA, 16, 4),
  lty = c(1, NA, NA)
)
```

The larger sample will generally give a more stable empirical distribution and mean
closer to the theoretical values. A particular simulation need not be closer in every
summary, and the empirical distribution will not exactly match the theoretical one.

:::
::::

# Probability distributions for count data {#sec-discrete_distributions}

There are many distributions that can be used to model count data. In the examples above
we used the binomial distribution although we never explicitly describe what it is. In this
section we cover three important distributions but you can find many more in the [Supplement](../Supplements/distributions.qmd).

What do you need to know about each distribution?

- Its parameters and their typical ecological interpretation
- How the parameters are related to the mean and variance
- Any constraint on the values the random variable can take
- What kind of data is often used for

Usually distributions are presented by their probability mass function. This is because all
other faces of the distribution can be derived from this function and we often visualize
samples through their empirical probability mass function (i.e., barplots or similar).

The last point above is not a formal method for choosing distributions. We will learn formal methods
of model comparison in [Chapter 5](../Chapter_5/Theory.qmd) but it helps to have an idea of
what type of data a distribution tends to be used for as that is always a good start.

::: {.callout-warning title="Choosing a probability distribution"}
A distribution should be chosen according to the type and range of observed values, its
mean–variance relationship, and how well it describes the observed data. Ecological
interpretation is useful, but a distribution is not restricted to only the process from
which it was originally derived.
:::

## The binomial distribution

The binomial distribution is often used to model binary process (e.g., dead or alive,
germinated or not) from a fixed number of events or invidiuals and we expect the probability
of the process to be the same for all individuals or events.

The binomial model has two parameters:

- $n$: the number of trials;
- $p$: the probability of success on each trial.

The probability mass function is

$$
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x} = \frac{n!}{x! (n - x)!} p^x(1-p)^{n-x}
$$

where $!$ denotes a factorial operation ($n! = n(n - 1)(n - 2)...2$)

The mean and variance are:

$$
E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$


### Example: seedling survival

Suppose 20 seedlings are planted and each has probability $0.7$ of surviving.

The probability that exactly 14 survive is:

$$
P(X=14)
=
\binom{20}{14}(0.7)^{14}(1-0.7)^{20-14}
=
\frac{20!}{14! (20 - 14)!}(0.7)^{14}(0.3)^6
\approx 0.1916.
$$

We can verify our math with R:

```{r}
seedling_count = 20L
seedling_survival_probability = 0.7
survivors_to_check = 14L

dbinom(
  x = survivors_to_check,
  size = seedling_count,
  prob = seedling_survival_probability
)
```

We can easily derive the mean and variance of the distribution:

```{r}
expected_survivors = seedling_count * seedling_survival_probability
variance_survivors = seedling_count *
  seedling_survival_probability *
  (1 - seedling_survival_probability)
```

Imagine that we know the mean and variance of a Binomial distribution, how do we calculate
its parameters? We just invert the relationships:

$$
\begin{aligned}
p &= 1 - Var(X)/E[X] \\
n &= E[X]/p
\end{aligned}
$$

Which in code becomes:

```{r}
mean_survivors = expected_survivors
success_probability = 1 - variance_survivors / mean_survivors
estimated_trial_count = mean_survivors / success_probability
```

We will use this inverse parameterization a lot in this course because most of the time we
will be interested in how the mean of a distribution varies in response to ecologically
relevant variables.

## The Poisson distribution

The Poisson distribution is often used for counts of events occurring in a fixed unit of
time, space, area, or sampling effort. Bur you can also use it for any count data where the
mean and variance are approximately the same.

The Poisson distribution has one parameter:

$$
\lambda=\text{expected number of events per sampling unit}.
$$

Its probability mass function is

$$
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!},
$$

The mean and variance are both equal to $\lambda$:

$$
E[X]=\lambda
$$

and

$$
\operatorname{Var}(X)=\lambda.
$$


### Example: wildlife detections

Suppose a camera detects an average of $0.8$ animals per night. If we operate
the camera for three nights, the expected number of detections is

$$
\lambda = 3(0.8) = 2.4.
$$

The probability of observing exactly two detections is

$$
P(X=2)
=
\frac{e^{-2.4}(2.4)^2}{2!}
\approx 0.2613.
$$

We can verify this calculation in R:

```{r}
rate_per_night = 0.8
n_nights = 3
expected_detections = rate_per_night * n_nights
detections_to_check = 2L

dpois(x = detections_to_check, lambda = expected_detections)
```

The probability of observing at least five detections is:

```{r}
minimum_detections = 5
ppois(
  q = minimum_detections - 1,
  lambda = expected_detections,
  lower.tail = FALSE
)
```

Unlike the binomial distribution, the Poisson distribution has no fixed upper
limit. A count of 100 is mathematically possible, although extremely unlikely
when $\lambda=2.4$.

```{r}
maximum_detections_to_plot = 15
detection_values = seq.int(from = 0, to = maximum_detections_to_plot)
detection_mass = dpois(x = detection_values, lambda = expected_detections)

plot(
  x = detection_values,
  y = detection_mass,
  type = "h",
  lwd = 3,
  xlab = "Number of detections",
  ylab = "Probability",
  main = "Poisson probability mass"
)
```

:::: {.callout-important title="Practice: flowering stems per orchid plot"}

A survey covers two square metres of a restoration site. Assume flowering stems occur
at an average rate of 1.2 per square metre. Calculate the probability of finding exactly
three stems in the surveyed area. Why does this model allow counts greater than three,
even though three is the count in the question?

::: {.callout-tip title="Solution" collapse="true"}

The expected count across two square metres is $\lambda=2(1.2)=2.4$. The probability is:

```r
rate_per_square_metre = 1.2
survey_area = 2
expected_stems = rate_per_square_metre * survey_area

dpois(x = 3, lambda = expected_stems)
```

The Poisson model has support $0,1,2,\ldots$, so counts larger than three are possible.
The probability of exactly three is approximately 0.209, but the model does not set
three as a maximum.

:::
::::

## Overdispersion

The Poisson distribution assumes that the mean and variance are equal. Ecological counts
often have a variance larger than their mean because organisms are clustered or because
sampling units differ in quality. For example, if some quadrats are highly suitable habitat
and others are poor habitat, the counts may be more variable than a Poisson model predicts.

This phenomenon is called **overdispersion** and motivates switching away from the Poisson
and using distributions with higher variances than the means, including the negative
binomial.^[The beta-binomial distribution is an extension of the binomial distribution in
which the success probability varies among sampling units. See the Supplement for details
on this distribution.] Overdispersion can arise from a variety of ecological and
measurement processes.

::: {.callout-warning title="Overdispersion is a statistical description"}
Overdispersion means that the observed variation is greater than the variation expected
under a reference model, often the Poisson model. It does not identify the ecological cause.
Clustering, habitat differences, omitted predictors, detection variation, and dependence
can all produce overdispersion.
:::

### A simple diagnostic for overdispersion

A first diagnostic is to compare the sample variance with the sample mean. For
Poisson data, these quantities should be approximately equal. We can summarize
the comparison with the dispersion ratio:

$$
\text{dispersion ratio}
=
\frac{\text{sample variance}}{\text{sample mean}}.
$$

For example:

```{r}
set.seed(seed = 123L)

n_quadrats = 100L
negative_binomial_size = 1.5
negative_binomial_mean = 4

counts = rnbinom(
  n = n_quadrats,
  size = negative_binomial_size,
  mu = negative_binomial_mean
)

sample_mean = mean(counts)
sample_variance = var(counts)
dispersion_ratio = sample_variance / sample_mean
c(sample_mean, sample_variance, dispersion_ratio)
```

A dispersion ratio close to 1 is consistent with the mean–variance relationship
of a Poisson distribution. A ratio substantially greater than 1 suggests
overdispersion, while a ratio less than 1 suggests underdispersion or a
bounded-count process such as a binomial process. This is a useful exploratory diagnostic,
not a definitive test. We will more formal methods to make decisions about which models to
use on [Chapter 5](../Chapter_5/Theory.qmd).

:::: {.callout-important title="Practice: variation in orchid stem counts"}

Across a sample of restoration plots, the mean flowering-stem count is 3 and the sample
variance is 12. Calculate the dispersion ratio and describe what it suggests relative to
a Poisson model. Can this ratio tell you why the counts vary so much?

::: {.callout-tip title="Solution" collapse="true"}

The dispersion ratio is $12/3=4$. It suggests that the observed variation is greater
than expected under a Poisson model with the same mean. The ratio alone cannot identify
the cause; for example, plots may differ in habitat, or counts may be clustered.

:::
::::

## The negative binomial distribution

The negative binomial distribution is often used for count data with greater
variation than expected under a Poisson distribution. This overdispersion can
arise when individuals are clustered or when sampling units differ in quality.

The negative binomial model has two parameters:

- $\mu$: the mean count;
- $k$: the overdispersion parameter.

The probability mass function is

$$
P(X=x)
=
\frac{\Gamma(x+k)}{\Gamma(k)x!}
\left(\frac{k}{k+\mu}\right)^k
\left(\frac{\mu}{k+\mu}\right)^x,
$$

where $\Gamma$ is the gamma function.

The mean and variance are:

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k}.
$$

The variance is greater than the mean when $k$ is finite. Smaller values of $k$
produce greater overdispersion. As $k$ increases, the variance approaches the
mean:

$$
\mu+\frac{\mu^2}{k}\longrightarrow\mu.
$$

Thus, the negative binomial distribution approaches the Poisson distribution
as $k$ becomes large.

In R, the ecological parameter $k$ is supplied as the `size` argument, while
the mean is supplied as `mu`:

```r
dnbinom(x, size = k, mu = mu) # Not run
```

### Example: seedlings in quadrats

Suppose the mean number of seedlings per quadrat is $\mu=4$, and the
overdispersion parameter is $k=1.5$.

The expected number of seedlings is

$$
E[X]=4.
$$

The variance is

$$
\operatorname{Var}(X)
=
4+\frac{4^2}{1.5}
\approx 14.67.
$$

We can verify these calculations in R:

```{r}
mean_seedlings = 4
negative_binomial_size = 1.5

expected_seedlings = mean_seedlings
variance_seedlings = mean_seedlings +
  mean_seedlings^2 / negative_binomial_size

c(mean = expected_seedlings, variance = variance_seedlings)
```

The probability of observing exactly 10 seedlings is:

```{r}
seedlings_to_check = 10

dnbinom(
  x = seedlings_to_check,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

The probability of observing at least 10 seedlings is:

```{r}
minimum_seedlings = 10

pnbinom(
  q = minimum_seedlings - 1,
  size = negative_binomial_size,
  mu = mean_seedlings,
  lower.tail = FALSE
)
```

Like the Poisson distribution, the negative binomial distribution has no fixed
upper limit. A count of 100 is mathematically possible, although extremely
unlikely.

```{r}
maximum_seedlings_to_plot = 20
seedling_values = seq.int(from = 0, to = maximum_seedlings_to_plot)
negative_binomial_mass = dnbinom(
  x = seedling_values,
  size = negative_binomial_size,
  mu = mean_seedlings
)

plot(
  x = seedling_values,
  y = negative_binomial_mass,
  type = "h",
  lwd = 3,
  xlab = "Number of seedlings",
  ylab = "Probability",
  main = "Negative binomial probability mass"
)
```

:::: {.callout-important title="Practice: orchid stem counts with different dispersion"}

Suppose the mean number of flowering stems per plot is $\mu=3$. Calculate the negative
binomial variance for $k=1.5$ and $k=6$. Which value gives greater overdispersion? How
does increasing $k$ relate to the Poisson model?

::: {.callout-tip title="Solution" collapse="true"}

Using $\operatorname{Var}(X)=\mu+\mu^2/k$, the variances are 9 and 4.5, respectively.
The smaller value of $k$ gives greater overdispersion. As $k$ increases, the variance
approaches the mean, which is the Poisson mean–variance relationship.

:::
::::

# Chapter quiz

Try each question before opening its solution.

:::: {.callout-important title="Question 1: lichen patches"}

On a fixed section of rock, let $X$ be the number of lichen patches found. Give the
possible values of $X$ if the survey area can contain at most 18 patches. Is this a
finite or countably infinite support?

::: {.callout-tip title="Answer" collapse="true"}

$X$ can take the integer values 0 through 18. Its support is finite because the survey
area has a maximum number of patches in this scenario.

:::
::::

:::: {.callout-important title="Question 2: freshwater larvae"}

A proposed distribution for the number of larvae in a sample assigns probabilities
0.15, 0.25, 0.40, and 0.30 to counts 0, 1, 2, and 3. Is this a valid probability
distribution? Explain.

::: {.callout-tip title="Answer" collapse="true"}

No. Although each probability is non-negative, they sum to 1.10 rather than 1.
Probabilities over all possible outcomes must sum to one.

:::
::::

:::: {.callout-important title="Question 3: coral spawning"}

For a model of the number of coral colonies spawning on a reef, which R function prefix
would you use to calculate (a) the probability of exactly five colonies, (b) the
probability of five or fewer, (c) the 0.9 quantile, and (d) a random sample?

::: {.callout-tip title="Answer" collapse="true"}

Use the `d` prefix for an exact probability, `p` for a cumulative probability, `q` for
a quantile, and `r` to generate a random sample.

:::
::::

:::: {.callout-important title="Question 4: acorn production"}

Suppose acorn production by two oak trees is modelled with independent random variables.
In this model, what does independence say about the probability of a particular joint
outcome? Give one ecological factor that might make the assumption questionable.

::: {.callout-tip title="Answer" collapse="true"}

Under independence, the probability of a particular pair of outcomes is the product of
their separate probabilities. Shared weather or soil conditions could make production
by the two trees dependent.

:::
::::

:::: {.callout-important title="Question 5: trout eggs"}

For a group of 12 trout eggs, suppose each egg hatches with probability 0.25 and the
outcomes are independent. Calculate the expected number hatched and the variance.

::: {.callout-tip title="Answer" collapse="true"}

For a binomial model, $E[X]=np=12(0.25)=3$ and
$\operatorname{Var}(X)=np(1-p)=12(0.25)(0.75)=2.25$.

:::
::::

:::: {.callout-important title="Question 6: mussel settlement"}

A model predicts an expected count of 2.4 newly settled mussels per tile. Can one tile
have 2.4 mussels? What does the expected value describe?

::: {.callout-tip title="Answer" collapse="true"}

A tile cannot have a fractional mussel. The expected value describes the long-run average
count across comparable tiles under the model.

:::
::::

:::: {.callout-important title="Question 7: moth emergence"}

In a sample of 30 host plants, moths emerged from 9. Calculate the relative frequency
of emergence. Is this value necessarily the true probability of emergence for all
relevant host plants?

::: {.callout-tip title="Answer" collapse="true"}

The relative frequency is $9/30=0.3$. It is a sample-based estimate of the probability,
not necessarily the exact population probability. Another sample could produce a
different relative frequency.

:::
::::

:::: {.callout-important title="Question 8: caddisfly larvae"}

You sample the number of caddisfly larvae in many equal-sized stream sections. As the
number of sections grows, what do you generally expect to happen to the empirical
distribution relative to the distribution that generated the data? Is exact agreement
expected for a finite sample?

::: {.callout-tip title="Answer" collapse="true"}

With more independent observations from the same distribution, the empirical distribution
generally becomes closer to the generating distribution. Exact agreement is not expected
in a finite sample, and individual simulations can vary.

:::
::::

:::: {.callout-important title="Question 9: choose a count model"}

Choose a plausible starting model for each scenario and briefly explain your choice:

1. The number of eggs hatching out of 15 eggs, with the same hatch probability for each.
2. The number of mayfly larvae in a standard sample, when the count is unbounded and
   its mean and variance are approximately equal.
3. The number of burrowing crabs per beach section, when counts are unbounded and the
   sample variance is substantially greater than the mean.

::: {.callout-tip title="Answer" collapse="true"}

1. A binomial model, because there is a fixed number of trials and a common success
   probability.
2. A Poisson model is a plausible starting point because the support is non-negative
   integer counts and the mean and variance are approximately equal.
3. A negative binomial model is a plausible starting point because it allows an
   unbounded count with variance greater than the mean. These are starting points; the
   data and assumptions still need to be checked.

:::
::::

:::: {.callout-important title="Question 10: bumblebee nests"}

Across meadow plots, the sample mean number of bumblebee nests is 2 and the sample
variance is 6. Calculate the dispersion ratio and explain what it suggests relative to
a Poisson model. Does it identify the ecological cause?

::: {.callout-tip title="Answer" collapse="true"}

The ratio is $6/2=3$, suggesting greater variation than expected under a Poisson model
with the same mean. The ratio does not identify a cause; clustering, habitat differences,
or dependence could contribute.

:::
::::
