---
title: "Ecological Models and Data"
---


# Overview of the book

The purpose of this book is to help your learn how to ecological models and statistics to
answer ecological questions with data. A model is a simplification of reality that captures
some ecological relationship and are often derived from theory or accumulated knowledge in
a field. These include functional responses in predator-prey relationships, ecophysiological
responses to the environment, growth curves, allometric relationships, gradients or time
series. We learn these models in ecology courses, but they always remain rather theoretical
and mathematical. We learn about oscillations, hystereris, tipping points and other
exciting phenomena, but it always feels somewhat detached from the more experimental, boots
on the ground ecological research.

When it comes to the more experimental research, statistics tends to dominate. It is all
about p-values, statistical significance and testing those pesky assumptions. The statistics
that you often learn was developed in the first half of XX century, it focuses on factorial
experiments, linear responses and everything seems to follow Normal distribution.

So what is the issue? If you like math and beautiful graphs you do modelling and get some
data from the literature or your colleagues while you enjoy sitting in front of the computer
most of the time, at room temperature. If you like going out to the field and seeing
Nature, you do cool experiments, collect some data, test the assumptions of your ANOVA and...
it does not work. Ecological variables do not response linearly to their drivers, they do not
follow Normal distributions and a lot of your data will not come from nice, balanced,
factorial experiment (because things go wrong in the field or sometimes it is just not
logistically possible). What do to then?

The proposal of this book is to combine both disciplines. Take the knowledge that we have
from ecological modelling, all the weird distributions with confusing names that actually
describe how ecological variables behave and combine them. We combine them using the same
statistical machinery behind your classic linear models or the not so classic generalized
linear model. And we give on the idea of null hypothesis testing and p-values and use the
models to answer the research questions that we actually care about.

For example, instead of asking "Does tempearture affect insect development?" You would ask
"How quickly does development change across the observed temperature range, and is there
evidence of a thermal optimum?" Or for example, instead of asking "Is the effect of this
treatment on seedling survival statistically significant?" you would ask "How much does the
treatment change seedling survival, and is that change large enough to affect recruitment?"
In a nutshell, we turn away from oversimplistic binary question and focus on subtle,
quantitative questions that have direct relevant to the ecological context of interest.

This is not my own original idea and this view is slowly growing across ecologists, though
many communities are still dominated by the orthodoxy of (generalized) linear (mixed) models
and p-values. I am not sure if we can pinpoint an origin or seminal paper for this idea, but
in my case I learnt it by reading the book Ecological Models and Data in R by Benjamin Bolker
in 2010, and it changed how I looked at data. In fact, I used to not like statistics (but
loved modelling) and reading his book completely changed in my mind. I have read many other
books since then that share in this idea (though often from a Bayesian perspective where
rolling your own model is more the norm). In a sense, this book is my humble homage to that
original great book.

In the rest of this chapter I want to talk a bit about why probability matters in ecological
research and I elaborate a bit more on why the standard statistics that I assume you have
learnt are not enough. There is a roadmap at the end but do not worry too much about it, it
is more of a reference/summary that will make more sense once you start reading the different
chapters themselves.

# Why probability matters in ecological research

Probability is one of those words that we think we know what it means but it is actual
hard to agree on exact, more technical meanings. In the XXI century, the formal answer
to the question "What is probability?" is a mathematical answer. Probability is a mathematical
concept and is associated to functions that have specific properties. The more problematic
question si "What does probability mean?" Philosophers, statisticians and practitioners have
been debating the answer to this question for at least 250 years. And the debate is not
settled.

We will stay away from the more dogmatic view points and simply accept that probability is
a tool that can be used to model capture (i) unexplained variation, in all of its forms, and
(ii) uncertainty, which is the consequence of unexplained variation. In other words,
probability gives us a way to describe variation in the ecological process, imperfect
observations, and uncertainty about model parameters and predictions. Note that this view
of probability is not shared by all. But do not fear, the methods that we will use are very
orthodox and, in fact, reduce to classical statistics for the specific conditions where those
methods are valid. Its just that the approach I follow is more pragmatic and simplifies
things a bit, so that we can focus more the modeling and ecological research and less on
statistical purity.

## Observation error

**Observation error** is the difference between what actually happens in reality and what
we record. We may miss an individual or count it twice, instruments never give perfect readings
or the protocols are not followed perfectly. Observation error changes the data we analyse,
not the underlying ecological reality.

We use probability to characterize the error of a particular observation procedure. Having
repeated observations under the same conditions helps to determine this error more accurately,
though we will always have to make judicious assumptions about which model to use to
characterize it (these probability models are called **distributions**). We will learn that
different distributions will match different types of observations and that they way in which
the observations are collected also matters. There is no "one size fits all" and we in fact
we will seldom use the Normal distribution! You can see a couple of examples of ecological
data with non-Normal distributions fitted to them (@fig-observation-error). You will learn
in this course how to make these kind of plots and how to fit the distributions (and more).

```{r}
#| echo: false
#| label: fig-observation-error
#| fig-cap: "Examples of probability models fitted to observed ecological data. The fitted curves describe the observed data as simplified probability models."
#| fig-subcap:
#|   - "Relative frequencies of tobacco hornworm counts in the built-in `InsectSprays` dataset, with a fitted negative binomial distribution."
#|   - "Distribution of animal body masses in the `Animals` dataset, with a fitted lognormal density."
#| layout-ncol: 2
library(MASS)
library(ggplot2)

# Load and fit the pooled insect-count data.
data(InsectSprays, package = "datasets")
insect_counts = InsectSprays$count

negative_binomial_fit = fitdistr(
  x = insect_counts,
  densfun = "Negative Binomial"
)

# Calculate relative frequencies and fitted probabilities for every observed insect count.
count_values = 0:max(insect_counts)
count_table = table(factor(insect_counts, levels = count_values))
relative_frequency = as.numeric(count_table) / length(insect_counts)

fitted_negative_binomial = dnbinom(
  x = count_values,
  size = negative_binomial_fit$estimate["size"],
  mu = negative_binomial_fit$estimate["mu"]
)

# Load body masses and remove very large species for a readable plot.
data(Animals, package = "MASS")
body_mass = Animals$body
body_mass = body_mass[body_mass <= 2000]

lognormal_fit = fitdistr(
  x = body_mass,
  densfun = "lognormal"
)

# Put the values needed for the two plots into data frames.
insect_plot_data = data.frame(
  count = count_values,
  relative_frequency = relative_frequency,
  fitted_probability = fitted_negative_binomial
)

body_mass_plot_data = data.frame(body_mass = body_mass)
body_mass_values = seq(min(body_mass), max(body_mass), length.out = 200)
body_mass_density_data = data.frame(
  body_mass = body_mass_values,
  fitted_density = dlnorm(
    x = body_mass_values,
    meanlog = lognormal_fit$estimate["meanlog"],
    sdlog = lognormal_fit$estimate["sdlog"]
  )
)

# Plot the count frequencies and fitted negative binomial probabilities.
ggplot(insect_plot_data, aes(x = count, y = relative_frequency)) +
  geom_col(width = 0.9, fill = "grey90", colour = "grey40") +
  geom_line(aes(y = fitted_probability), colour = "#D55E00", linewidth = 0.8) +
  geom_point(aes(y = fitted_probability), colour = "#D55E00", size = 1.8) +
  scale_x_continuous(breaks = seq(0, max(count_values), by = 2)) +
  labs(x = "Number of insects", y = "Relative frequency") +
  theme_classic()

# Plot the body-mass density and fitted lognormal density.
ggplot(body_mass_plot_data, aes(x = body_mass)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 15,
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = body_mass_density_data,
    aes(x = body_mass, y = fitted_density),
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  coord_cartesian(ylim = c(0, 0.02)) +
  labs(x = "Body mass (kg)", y = "Density") +
  theme_classic()
```


## Process error

**Process error**, also called **process variation** or **stochasticity**, describes
variation in the ecological reality that is not explained by the deterministic part
of a model. It differs from obervation error because it will still be there even if we
had perfect measurements and had no uncertainty about the model (see below for this one).
There can be countless reasons for process error, including:

- **Environmental variability:** temperature, rainfall, nutrients, disturbance, or habitat conditions vary across space and time.
- **Unmeasured drivers:** factors such as predators, competitors, disease, or food availability are not included in the model.
- **Individual differences:** organisms differ in age, size, genetics, condition, or behaviour.
- **Random demographic events:** births, deaths, reproduction, dispersal, and survival involve chance.
- **Spatial and temporal dependence:** nearby observations or repeated observations are more similar than independent observations.
- **Interactions and nonlinearities:** the effect of one factor depends on another, or responses change across environmental conditions.
- **Model simplification:** a model represents only part of a complex ecological process.

Do not confuse inability to explain variation with true randomness. Many seemingly random
outcomes can be explained (and predicted) once you have a good enough model and measure all
the relevant variables with sufficient accuracy. My best example of this is the work by
Professor Persi Diaconis from Stanford University where he managed to build a model that
predicted coin flips, provided that the initial conditions were measured accurately enough.
There is a [video](https://youtu.be/AYnJv68T3MM?si=TmrdJjNlD6da9J8M) from Numberphile that
explains it really well.

A recorded value often reflects both process error and observation error. In simple
models we may describe their combined variation with one probability distribution. However,
if we can identify groups or clusters in the data that explain a significant part of the
variation we can use hierarchical models (*aka* mixed effect or multilevel models) to
disentangle the two or at least estimate a significant amount of process error. You will
also learn how to do this in the later part of the book.

To illustrate this idea of separating different sources of error, I plot again the data on
tobacco hornworm counts (@fig-process-error) but now I reveal that in the original experiments
the plants were sprayed with different products and the researchers kept track of this. We
can see in the plot that there is significant variation within the treatments and across the
treatments. The variation across treatments is part of process error as we do not expect our
measurements to have systematic errors depending on the treatment.

```{r}
#| echo: false
#| label: fig-process-error
#| fig-cap: "Boxplots and raw counts of tobacco hornworm (InsectSprays built-in dataset) by pooling all data together () and for each spray treatment separately."
library(ggplot2)

data(InsectSprays, package = "datasets")

# Combine the pooled counts with counts grouped by spray.
pooled_counts = data.frame(
  count = InsectSprays$count,
  group = "All data"
)

spray_counts = data.frame(
  count = InsectSprays$count,
  group = as.character(InsectSprays$spray)
)

plot_data = rbind(pooled_counts, spray_counts)

plot_data$group = factor(
  plot_data$group,
  levels = c("All data", levels(InsectSprays$spray))
)

# Spray explains some of the variation visible in the pooled data.
ggplot(plot_data, aes(x = group, y = count)) +
  geom_boxplot(
    width = 0.6,
    fill = "grey90",
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(width = 0.12, height = 0, seed = 20261001),
    alpha = 0.65,
    size = 2,
    colour = "#0072B2"
  ) +
  labs(
    x = "Group (spray treatment or all data)",
    y = "Number of insects"
  ) +
  theme_classic()
```


## Parameter uncertainty

A **model parameter** is a quantity that helps the model adjust itself to fit the data
better. The statistical view of the world is that, as far as parameter estimation is
concerned, we will treat the data as if it was a random sample from the model.
That randomness is precisely capturing the observation error and stochasticity or process
error/variation mentioned in th above.

This means that the same model is compatible with different samples of data (that look similar
but different) because of the random variations that we cannot explained with the model. In
other words, there isn't a one-to-one correspondence between a specific sample of data and
a model. If we invert this way of thinking, we will conclude that, if we fix the data, we
could have multiple models compatible with that particular sample of data and each of models
correspond to different values of the parameters. The fact that we do not exactly which
paramter values to use is what we call **uncertainty**.

We can see this idea in the relationship between mammal body mass and brain mass
(@fig-parameter-uncertainty). Even after taking logarithms, species with similar body masses
have different brain masses. Here we use a saturating response curve rather than asking R to
draw a flexible smooth line. The shaded band is an approximate 95% confidence interval around
the fitted mean response. It is not where we expect individual species to lie. It shows how
the observed variation means that several similar response curves could be compatible with
these data.

```{r}
#| echo: false
#| label: fig-parameter-uncertainty
#| fig-cap: "Brain mass in relation to body mass for mammals in the `Animals` dataset. A shifted and scaled Michaelis–Menten was fitted, shown are the mean model description and its 95% confidence interval. See inset for details."
library(MASS)
library(ggplot2)

# Log-transform both measurements so that the wide range of mammal sizes is visible.
data(Animals, package = "MASS")

animal_brain_data = data.frame(
  log10_body_mass = log10(Animals$body),
  log10_brain_mass = log10(Animals$brain)
)

# Shift and scale body mass to the interval from zero to one before fitting the response.
body_mass_minimum = min(animal_brain_data$log10_body_mass)
body_mass_range = diff(range(animal_brain_data$log10_body_mass))
animal_brain_data$scaled_body_mass =
  (animal_brain_data$log10_body_mass - body_mass_minimum) / body_mass_range

# Fit a Michaelis-Menten response. With normally distributed residuals, these estimates
# are also the maximum likelihood estimates of the response parameters.
response_fit = nls(
  formula = log10_brain_mass ~ lower_asymptote +
    increase * scaled_body_mass / (half_saturation + scaled_body_mass),
  data = animal_brain_data,
  start = list(
    lower_asymptote = min(animal_brain_data$log10_brain_mass),
    increase = diff(range(animal_brain_data$log10_brain_mass)),
    half_saturation = 0.2
  ),
  algorithm = "port",
  lower = c(-Inf, 0, 0.001)
)

# Calculate predictions and approximate 95% confidence intervals for the fitted mean response.
prediction_body_mass = seq(
  from = min(animal_brain_data$log10_body_mass),
  to = max(animal_brain_data$log10_body_mass),
  length.out = 200
)

prediction_data = data.frame(
  log10_body_mass = prediction_body_mass,
  scaled_body_mass = (prediction_body_mass - body_mass_minimum) / body_mass_range
)

prediction_data$fit = predict(response_fit, newdata = prediction_data)

prediction_gradient = cbind(
  lower_asymptote = 1,
  increase = prediction_data$scaled_body_mass /
    (coef(response_fit)["half_saturation"] + prediction_data$scaled_body_mass),
  half_saturation = -coef(response_fit)["increase"] * prediction_data$scaled_body_mass /
    (coef(response_fit)["half_saturation"] + prediction_data$scaled_body_mass)^2
)

prediction_standard_error = sqrt(
  rowSums((prediction_gradient %*% vcov(response_fit)) * prediction_gradient)
)

prediction_data$lower = prediction_data$fit - 1.96 * prediction_standard_error
prediction_data$upper = prediction_data$fit + 1.96 * prediction_standard_error

# Calculate profile confidence intervals and format them for the reporting inset.
parameter_confidence_intervals = confint(response_fit, level = 0.95)

format_parameter = function(display_name, parameter_name) {
  estimate = coef(response_fit)[parameter_name]
  confidence_interval = parameter_confidence_intervals[parameter_name, ]

  paste0(
    display_name,
    " = ", formatC(estimate, format = "f", digits = 2),
    " [", formatC(confidence_interval[1], format = "f", digits = 2),
    ", ", formatC(confidence_interval[2], format = "f", digits = 2), "]"
  )
}

model_inset = paste(
  "Model:",
  "\u03bc(x) = \u03b1 + \u03b2 x / (K + x)",
  "x = [log10(body mass) - minimum] / range",
  "",
  "Normal-error MLE (95% CI):",
  format_parameter("\u03b1", "lower_asymptote"),
  format_parameter("\u03b2", "increase"),
  format_parameter("K", "half_saturation"),
  sep = "\n"
)

# Plot the observations, fitted response, uncertainty band, and model-reporting inset.
ggplot(animal_brain_data, aes(x = log10_body_mass, y = log10_brain_mass)) +
  geom_ribbon(
    data = prediction_data,
    aes(x = log10_body_mass, ymin = lower, ymax = upper),
    inherit.aes = FALSE,
    fill = "#D55E00",
    alpha = 0.2
  ) +
  geom_point(colour = "#0072B2", alpha = 0.7, size = 2) +
  geom_line(
    data = prediction_data,
    aes(x = log10_body_mass, y = fit),
    inherit.aes = FALSE,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  annotate(
    geom = "label",
    x = 2,
    y = -2.15,
    label = model_inset,
    hjust = 0,
    vjust = 0,
    size = 3.2,
    fill = "grey99",
    colour = "black",
    label.size = 0.25
  ) +
  coord_cartesian(ylim = c(-2.3, 4)) +
  labs(
    x = expression(log[10] ~ "body mass (kg)"),
    y = expression(log[10] ~ "brain mass (g)")
  ) +
  theme_classic()
```

There are two different ways of formally defining parameter uncertainty in statistics: the
Frequentist and the Bayesian way. For large samples of data and simple models there is not much
difference between them. For small dataset the Bayesian way can be superior but it is
sensitive to prior assumptions that you need to bring to the table to complement what little
data you have. That can introduce biases in your conclusions and Bayesian methods are still
seen with suspicion by many researchers. So we will stick to the Frequentist intepretation in this
book but be aware that (i) our estimations of uncertainty will be only approximate when we
have little data and (ii) Bayesian methods are rising in popularity so you may end up having
to learn them anyways (and this book is meant to be a bridge between traditional and Bayesian
statistics).

## Other applications

Probability can also model uncertainty about a model's initial conditions
(its starting state), which model to use, and inputs such as future rainfall.
Alternative future scenarios add another source of uncertainty. Computers may also
introduce numerical approximation error when calculating model predictions.
Uncertain starting states and inputs are particularly relevant when a model predicts
how a system changes through time, because their effects can carry into later
predictions.

This book focuses on probability models for ecological observations and on fitted
response curves. We do not work with dynamical models, so we will not develop
these other sources of uncertainty in detail. We will still compare candidate models when we
fit ecological response curves in [Chapter 7](../Chapter_7/Theory.qmd) but we will not
assign probabilities to models.

# Why standard statistics are not enough

Most of us first learn statistics through linear models, ANOVA, and later perhaps
generalised linear and mixed models. These are useful tools and you should not
throw them away. But they are often taught as if the main challenge were choosing the right
formula, checking a few assumptions (which often you hope pass or otherwise you are stuck),
and seeing whether a coefficient is different from zero. That is not always the question we
care about.

Ecological responses are very rarely obliged to be straight lines. A feeding rate
may level off when a predator runs out of time to handle prey. Development may
increase with temperature up to an optimum and then decline. Survival may change
rapidly at low resource availability but barely at all once resources are abundant.
We often already know something about these shapes from ecological theory, previous
experiments, or simply from thinking about how organisms work. It seems a bit odd
to ignore that knowledge and fit a straight line just because it is convenient.

Historically, convenience was a good reason to use simple models: more complicated
models were difficult to fit as the math became untractable. That is no
longer such a strong limitation. Modern computers allow us to fit, visualise, and
check a much wider range of models by using numerical methods. The challenge is not whether
we *can* fit a complicated model. It is whether the model represents an ecological relationship
reasonably well, and whether the data can tell us anything useful about its parameters.

For example, imagine that we measure the feeding rate of a predator across a range
of prey densities. We could ask whether prey density has a statistically significant
effect on feeding rate. But a more useful set of questions might be: How quickly
does feeding rate increase when prey are scarce? Does it reach a maximum? What is
that maximum feeding rate? And how uncertain are those estimates?

A straight line might describe part of the observed range, but it cannot keep
increasing forever. A predator has limited time to search for, capture, and handle
prey. A saturating response curve can represent that biological constraint directly.
Its parameters describe features of the system that we can discuss, compare among
studies, and use in later ecological reasoning. Of course, those comparisons only
make sense when the model definitions, units, organisms, and study conditions are
sufficiently similar.

This is also why a small p-value is rarely the end of an ecological analysis (or the main
goal). A test against a zero-effect null hypothesis can sometimes be useful, but it does
not tell us whether an effect is large enough to matter, which ecological
explanation is most credible, or whether the fitted relationship makes biological
sense. In many cases, “nothing happens” is not even a plausible ecological model
to compare with. We are usually choosing among several ways in which something
might happen. In many cases, a large p-value is simply the result of not having enough
data to detect a clear signal from noisy data. It says more about the size of your
experiment or survey than whether there is an actual effect or not. Of course, the purist
would say that "you either reject the null hypothesis or fail to reject, never accept it",
meaning that either you get a p-value below the threshold or you get nothing. But turning research
into a success/failure story is not so useful and provides the wrong kind of incentives to
researchers. If you are interested in the current debates about "Moving beyonf p < 0.05" I
suggest starting by the editorial by Wasserstein et al. on the topic[^1]

[^1]: Wasserstein, R. L., Schirm, A. L., & Lazar, N. A. (2019). *Moving to a world beyond “p < 0.05”*. **The American Statistician, 73**(sup1), 1–19. <https://doi.org/10.1080/00031305.2019.1583913>

The approach in this book is to make those choices explicit. We combine an
ecological response function with a probability model for the observations. This
lets us estimate parameters, describe uncertainty, compare plausible candidate
models, and check whether the fitted model captures important features of the
data. The aim is not to replace every linear or mixed model with something more
elaborate. It is to use the simplest model that can answer the ecological question
we actually have. If that is a linear model that assumes normally distributed errors, then
we use that. But because we choose to and not because we have to.

These approaches are useful beyond pure research aimed at understanding ecological processes.
When ecological models are used to inform decisions about human impacts on
ecosystems, we need predictions but we also need to quantify their uncertainty. This book is
about analysing ecological data rather than developing
forecasting models, but stating a model clearly, estimating its parameters,
checking its fit, and being explicit about uncertainty are useful habits in either
case.


# The modelling cycle

Ecological modelling begins with a research question, but it does not move in a
straight line from question to answer. We use what we know about the system and the data to
propose a model, estimate its parameters, and compare plausible alternatives. Each result
may send us back to the question, the data, or the model. The aim is always to give a useful
answer to the ecological question, rather than to treat a fitted model as an end in itself.

![The ecological modelling cycle. Research questions guide the choice of data and model. The model and data are used to estimate parameters which are then intepreted to propose an ecological answer. Data can help select the most adequate model, but some ecological questions can also be framed as model comparison. The green boxes on the right show the steps involved in constructing a model, estimating its parameters with uncertainty, and the model comparison tools developed in this book; red numbers indicate in which chapter each tool is introduced.](ModelCycle.png)

- **Model:** We first express an ecological idea as a model that can generate the kind of
  data we observe. This includes choosing a probability distribution for discrete data
  ([Chapter 2](../Chapter_2/Theory.qmd)) and for continuous measurements
  ([Chapter 3](../Chapter_3/Theory.qmd)). We then describe how the parameters of the
  distribution vary as a function of continuous covariates/predictors ([Chapter 6](../Chapter_6/Theory.qmd) and
  [Chapter 7](../Chapter_7/Theory.qmd)) and how they vary across groups such as location, species or treatment
  ([Chapter 8](../Chapter_8/Theory.qmd)).

- **Estimation:** Given a model and data, we estimate the parameter values that make the
  model most compatible with the observations. [Chapter 4](../Chapter_4/Theory.qmd)
  introduces parameter estimates, their sampling variation, and confidence intervals.
  [Chapter 5](../Chapter_5/Theory.qmd) develops the method of maximum likelihood;
  [Chapter 9](../Chapter_9/Theory.qmd) deals with all the technical difficulties of
  finding the optimum in complex likelihood surfaces. [Chapter 8](../Chapter_8/Theory.qmd)
  covers the method of marginal likelihood necessary for hierarchical models.

- **Model comparison and answer:** More than one model may be plausible. In
  [Chapter 7](../Chapter_7/Theory.qmd), we compare fitted candidate models using
  information criteria and cross-validation, check how well they describe the data, and
  interpret their parameters and uncertainty in terms of the original ecological question.
  A comparison can reveal that we need a different model or additional data, which is why
  modelling is always a cycle.

If we zoom into estimation process we can see better what the relationship between models
and data is. Firstly, we need to abstract everything that happens in an experiment that can
affect the data we collect into a **data generating process**. This includes all the
experimental protocols, measurement procedures and actual ecological processes of relevance.
A statistical model is a mathematical approximation to that generating process. We often use
deterministic models to approximate the ecological process that we are most interested, but
a statistical model though needs to have at least one stochastic component (i.e., a
distribution). Some ecological questions can be framed in terms of variation and stochasticity
and if fact it is possible to have a model that is just made of nested distributions. Models
contain degrees of freedom that allow them a certain flexibility to match different possible
processes. These degrees of freedom are indexed by variables called **parameters**, which
are the variables that we are estimating when combined models and data.

By applying the data generating process we obtain a specific data sample. For the purpose of
parameter estimation, these data will be treated as if they were a random sample from the
model^[of course, in reality they emerge from the data-generating process, not the model,
but this assumption is needed to develop the relevant statistical procedures.]. In principle,
a parameter of a model can be estimated from data with any procedure we may want, and there
are in fact many procedures available. Frequentist statistics is a branch of statistics that
analyses the properties of the different procedures, including concepts such as bias and
means square error. In the course we will use the method of maximum likelihood which is the
most general procedure.

![The relationship between reality, model and data in the process of parameter estimation. Solid lines implies that some quantities are derived from data, reality or mathematical structures. Dashed lines imply an approximate statistical relationship that is assume for the purpose of parameter estimation.](Estimation.png)

# Roadmap of the course

This contains an overview of each chapter in more detail. It probably does not make much sense unless you
have read a particular chapter, so it is not meant as an introduction to each chapter but
rather as a reference that you check as you read, to remember where is what.

## Chapter 2: Probability models for count data

**Key concepts.** [Chapter 2](../Chapter_2/Theory.qmd) introduces discrete random
variables, their support, and probability distributions. You meet the four faces of a
distribution: probability mass, cumulative probability, quantiles, and random sampling.
The chapter also introduces independence and independent and identically distributed
(i.i.d.) observations, expected values, variance, samples and populations, empirical
distributions, and convergence as sample size increases. Binomial, Poisson, and negative
binomial distributions provide models for different types of ecological discrete data,
including overdispersion. Extra distributions are available in the [Supplement](../Supplements/distributions.qmd).

**R skills and tools.** You use the `d*`, `p*`, `q*`, and `r*` families of distribution
functions, including functions for binomial, Poisson, and negative binomial models. You
generate reproducible simulations with `set.seed()`, summarise samples with `table()`,
`mean()`, `var()`, and `quantile()`, and compare empirical frequencies with theoretical
probabilities using tables and plots. You will also become familiar with the formulae for
different distributions by doing exercises by hand.

**How it builds on earlier material.** Chapter 1 introduces the R skills needed to create
objects, call functions, and work with data. Chapter 2 uses those skills to establish the
probability language for the rest of the course. In later chapters, the same distributions
will be used to simulate data, estimate unknown parameters, construct likelihoods, and
describe variation around ecological relationships.

## Chapter 3: Probability models for continuous data

**Key concepts.** [Chapter 3](../Chapter_3/Theory.qmd) extends probability modelling to
continuous measurements such as tree height, body mass or vegetation cover. Probability
mass is replaced by probability density. You also learn about histograms and kernel density
estimates, and and about joint, marginal, and conditional distributions. Normal, lognormal,
and beta distributions provide models for mesaurements of different kinds.

**R skills and tools.** You apply the same `d*`, `p*`, `q*`, and `r*` pattern to continuous
distributions through functions such as `dnorm()`, `pnorm()`, `qnorm()`, and `rnorm()` and
their lognormal and beta equivalents. You compare samples with theoretical models using
`hist()`, `density()`, and `ecdf()`, and explore relationships between two variables with
scatterplots and contour plots. As with chapter 2, you also work on the formulae with pen
and paper.

**How it builds on earlier material.** The four faces of a distribution, random sampling,
empirical distributions, and independence all come from Chapter 2. Chapter 3 keeps that
framework but shows what changes when the response can vary continuously. Together,
Chapters 2 and 3 provide the probability models for the count and continuous data used in
the estimation chapters that follow.

## Chapter 4: Estimation and sampling distributions

**Key concepts.** [Chapter 4](../Chapter_4/Theory.qmd) changes the direction of the
question. Instead of calculating outcomes from known parameters, you use a sample to
estimate unknown parameters. You distinguish parameters, estimators, and estimates and
use the method of moments as a first estimation method. Repeated samples introduce
sampling distributions, bias, estimator variance, mean squared error, standard errors,
confidence intervals, consistency, and asymptotic normality. Simulation is used to study
these properties and to check confidence-interval coverage.

**R skills and tools.** You calculate estimates and standard errors with functions such as
`mean()`, `var()`, `sd()`, and `sqrt()`. You use `replicate()` with `rpois()` and `rnorm()`
to simulate repeated studies, visualise sampling distributions with histograms, obtain
critical values with `qnorm()` and `qt()`, and use simulation to check estimator bias,
variation, and the actual coverage of confidence intervals.

**How it builds on earlier material.** Chapters 2 and 3 describe distributions with known
parameters and show how finite samples vary around them. Chapter 4 uses those same models
and simulations to ask what sample summaries can tell us when the parameters are unknown.
It also develops the language of estimates and uncertainty needed to interpret parameter
estimates in all later chapters.

## Chapter 5: Maximum likelihood estimation

**Key concepts.** [Chapter 5](../Chapter_5/Theory.qmd) introduces likelihood as a way to
compare parameter values while holding the observed data fixed. You construct likelihoods
and log-likelihoods from probability masses or densities and find maximum
likelihood estimates. For models with several parameters, you explore likelihood surfaces,
slices, and profiles. The chapter connects likelihood curvature to standard errors and
compares quadratic and profile likelihood confidence intervals. An elaborate example
using a negative binomial distribution is used to ground the theory.

**R skills and tools.** You write negative log-likelihood functions, calculate log
probabilities or densitiesand fit models with `optim()`. You learn how to deal with
constraints on pararameters. You use grids and contour plots to explore likelihoods,
the Hessian to approximate standard errors, and `uniroot()` to compute likelihood
profiles.

**How it builds on earlier material.** Likelihood uses the discrete and continuous
distributions introduced in Chapters 2 and 3, combined with a sample of data. It addresses
the estimation problem from Chapter 4 by using the full probability model rather than only
matching moments. The sampling-distribution ideas from Chapter 4 remain essential for
understanding the properties of maximum likelihood estimators and the interpretation of
their confidence intervals.

## Chapter 6: Deterministic functions in ecology

**Key concepts.** [Chapter 6](../Chapter_6/Theory.qmd) focuses on deterministic functions
that describe how an expected ecological response changes with a predictor. Several
exampels are used to illustrate the concepts, including Holling type II, exponential, Ricker,
and logistic functions. Additional functions are given in the a [Supplement](../Supplements/deterministic-functions.qmd).
You learn to distinguish the inputs and outputs of a model, constraints on parameter values
but also how the same functions can be reparameterized as slighlty different formulae. The
properties of these functions are analyzed by taking limits and derivatives, to calculate
asymptotes, slopes and maxima. Special points like half-saturation, half-lives and
inflection points are also discussed. Finally, you learn how to modify functions by shifting
and scaling.

**R skills and tools.** You translate equations into vectorised R functions and evaluate
them for chosen inputs. You plot and compare response curves with `curve()`, `lines()`, and
`legend()`, and use `outer()` and `contour()` to explore functions that depend on two varying
quantities. R plots are used alongside algebra to check how parameter changes affect a
curve's shape. Many exercises are offered to practice the relevant analytical skills.

**How it builds on earlier material.** The book takes a turn at this point by presenting
the second pillar of ecological models for data analysis. Although there is little connection
to the probability and statistics of the previous chapters, the two will come together in
chapter 7.

## Chapter 7: Fitting ecological response curves

**Key concepts.** [Chapter 7](../Chapter_7/Theory.qmd) combines deterministic functions and
the distributions of previous chapter to construct more complex likelihood functions that
relate covariates to observations in ecologically meaningful ways. You estimate parameters
and confidence intervals with the techniques from chapter 5, but also compare candidate
models using information criteria and cross-validation, and assess goodness-of-fit using
simulation-based scaled quantile residuals. Repeated simulation is also used to investigate
bias and confidence-interval coverage.

**R skills and tools.** Write more complex likelihood functions, using RTMB and its built-in
automatic differentiation. . You calculate AIC and BIC, predictive RMSE, and create
diagnostic residuals. This chapter introduces the packages `RTMB` and `DHARMa`.

**How it builds on earlier material.** This chapter brings together the course so far. The
observation distributions come from Chapters 2 and 3; simulation, uncertainty, bias, and
coverage come from Chapter 4; the likelihood and optimisation workflow comes from Chapter
5; and the determinestic functions from Chapter 6. The result is a complete workflow for
specifying, fitting, comparing, checking, and interpreting ecological models.

## Chapter 8: Grouped ecological response curves

**Key concepts.** [Chapter 8](../Chapter_8/Theory.qmd) introduces repeated measurements
nested in groups. We learn the differences between complete, no, and partial pooling, then
use random and fixed effects to quantify how parameters vary across groups. We learn the
concept of maginal and conditional distribution and how to marginalize likelihoods over
random effects. The chapter introduces the concepts of shrinkage, regularisation, maximum
marginal likelihood, and Laplace approximation.

**R skills and tools.** You learn how to introduce groups of parameters into models,
declare random effects and perform maximum marginal likelihood with RTMB. You interpret
shared parameters, variance components, and group-specific fitted curves.

**How it builds on earlier material.** Chapter 8 applies Chapter 7's response curve,
observation model, and RTMB workflow to repeated measurements under groups. RTMB performs
the numerical marginalization, so you can focus on the modeling and interpreting its
assumptions.

## Chapter 9: Numerical optimization

**Key concepts.** [Chapter 9](../Chapter_9/Theory.qmd) examines the geometry of difficult
likelihood functions, the concept of identifiability and the different classes of optimization
algorithms and how to address convergence issues.

**R skills and tools.** Different functions for optimization (`optimize()`, `optim()`, `nlminb()`,
`DEoptim()` and `ga()`) to complex likelihood functions in ecologically relevant problems.

**How it builds on earlier material.** Chapters 5, 7, and 8 used optimizers to fit
likelihood and marginal-likelihood functions. Chapter 9 explains how those numerical
searches work and how to diagnose convergence and overcome technical difficulties.
