---
title: "Chapter 4: Maximum likelihood estimation"
date: today
---

Exercise 1 uses a laboratory experiment on the survival of tadpoles, Exercises 2 and 3 use
field counts of glacier lily seedlings, and Exercise 4 uses measurements of trees and of
penguins that come with R. Exercise 5 simulates hypothetical experiments from the models
fitted in Exercise 4. Exercises 1 to 4 are the core of the practical and Exercise 5 is extra
practice. The data of Exercises 1 to 3 are in the package `emdbook`. Work through the first
example before trying the remaining exercises. The ideas and notation are introduced in the
[Chapter 4 theory](../Theory.qmd).

As in the theory, every model is written and fitted with the packages RTMB and RTMBdist,
which you installed in the Chapter 1 practical together with `emdbook`. Load RTMB first and
RTMBdist second:

```{r}
#| message: false
# Load RTMB first and RTMBdist second.
library(RTMB)
library(RTMBdist)
```

# Exercise 1: Survival of reed frog tadpoles

The data set `ReedfrogPred` in the package `emdbook` contains the results of laboratory
experiments on the predation of tadpoles of the African reed frog *Hyperolius spinigularis*
(Vonesh and Bolker 2005). Each row is a tank of 1.2 × 0.8 × 0.4 m. The variable `density`
is the number of tadpoles that were placed in the tank at the start (10, 25 or 35) and
`surv` is the number of tadpoles that survived. Questions 1 to 5 use the 24 tanks with
predators, and question 6 also uses the 24 tanks without them:

```{r}
# Load the reed frog data from the emdbook package without attaching the package.
data(ReedfrogPred, package = "emdbook")

# Keep the 24 tanks with predators.
predator_tanks = ReedfrogPred[ReedfrogPred$pred == "pred", ]

# Initial number of tadpoles in each tank and number of survivors, and the number of tanks.
tadpoles = predator_tanks$density
survivors = predator_tanks$surv
n_tanks = length(survivors)
```

Let $X_i$ be the number of survivors in tank $i$, which started with $n_i$ tadpoles. We use
a [Binomial](../../Supplements/distributions.qmd#sec-binomial) model with the same
probability of survival $p$ in all the tanks,

$$
X_i \sim \operatorname{Binomial}(n_i, p),
$$

and assume that the tanks are independent. Note that the tanks are not identically
distributed, because $n_i$ differs among them. The likelihood is still the product of the
contributions of the tanks, since the product only requires independence. Each contribution
is calculated with the $n_i$ of its own tank.

1. On paper, write the likelihood contributions of the first three tanks (4, 9 and 7
   survivors out of 10 tadpoles each) and their likelihood as a function of $p$. Calculate
   the likelihood and the log-likelihood for $p=0.5$ and for $p=2/3$, and verify the
   results in R.
2. On paper, derive the maximum likelihood estimator of $p$ for all the tanks by
   differentiating the log-likelihood (the
   [rules of differentiation](../../Supplements/calculus.qmd#sec-calc-rules) are in the
   calculus refresher). Calculate the estimate.
3. Write the negative log-likelihood (NLL) with $\operatorname{logit}p$ as the parameter,
   process it with `MakeADFun()` and minimise it with `optim()`. Check the convergence and
   report the estimate of $p$. Calculate the standard error on the logit scale from the
   Hessian by hand and with `sdreport()`, and calculate a 95% Wald interval for $p$.
4. Calculate the 95% likelihood interval for $p$ by hand with `uniroot()` and with
   `TMB::tmbprofile()`, and compare it with the Wald interval.
5. Is the Binomial model adequate for these data? For each initial density, compare the
   variance of the number of survivors among tanks with the variance of the fitted model.
   What does the comparison imply for the intervals of questions 3 and 4?
6. The experiment also had 24 tanks without predators. Fit a model with one probability of
   survival for all 48 tanks and a model with one probability for each treatment, and
   compare them with AIC and AICc. How much does the presence of predators change the
   probability of survival, and what is the standard error of that change?

## Solution

### Question 1

The observational unit is one tank. The contribution of a tank to the likelihood is the
probability that the Binomial model assigns to its number of survivors,
$P(X_i=x_i\mid p)=\binom{n_i}{x_i}p^{x_i}(1-p)^{n_i-x_i}$. The first three tanks started
with 10 tadpoles each and had 4, 9 and 7 survivors, so their likelihood is

$$
\begin{aligned}
L(p\mid 4,9,7)
&= \binom{10}{4}p^{4}(1-p)^{6}\times
   \binom{10}{9}p^{9}(1-p)^{1}\times
   \binom{10}{7}p^{7}(1-p)^{3}\\
&= 210\times 10\times 120\;p^{20}(1-p)^{10}
= 252{,}000\;p^{20}(1-p)^{10}.
\end{aligned}
$$

The exponents are the total number of survivors (20) and the total number of tadpoles that
died (10) in the three tanks. For $p=0.5$ the likelihood is
$252{,}000/2^{30}\approx 2.35\times10^{-4}$ and for $p=2/3$ it is
$252{,}000\times 2^{20}/3^{30}\approx 1.28\times10^{-3}$. The log-likelihood is the
logarithm of the same expression,

$$
\ell(p)=\log(252{,}000)+20\log p+10\log(1-p),
$$

which gives $12.437-20.794=-8.357$ for $p=0.5$ and $12.437-8.109-10.986=-6.658$ for
$p=2/3$. We can verify our math with R:

```{r}
# The first three tanks: survivors and initial number of tadpoles.
first_survivors = survivors[1:3]
first_tadpoles = tadpoles[1:3]
first_survivors
first_tadpoles

# Likelihood contribution of each tank at p = 0.5 and at p = 2/3, and the likelihood of the
# three tanks, which is the product of the contributions.
dbinom(x = first_survivors, size = first_tadpoles, prob = 0.5)
dbinom(x = first_survivors, size = first_tadpoles, prob = 2 / 3)
prod(dbinom(x = first_survivors, size = first_tadpoles, prob = 0.5))
prod(dbinom(x = first_survivors, size = first_tadpoles, prob = 2 / 3))

# Log-likelihood at the same two values: the sum of the log probabilities.
sum(dbinom(x = first_survivors, size = first_tadpoles, prob = 0.5, log = TRUE))
sum(dbinom(x = first_survivors, size = first_tadpoles, prob = 2 / 3, log = TRUE))
```

R returns the same values. The difference between the two log-likelihoods is 1.70, so the
three counts are about 5.5 times more probable ($e^{1.70}$) under $p=2/3$ than under
$p=0.5$. This is no coincidence, as 20 of the 30 tadpoles in these three tanks survived.
The second tank, with 9 survivors out of 10, has a probability of only 0.0098 under
$p=0.5$.

### Question 2

For all the tanks the log-likelihood is the sum of the logarithms of the 24 contributions:

$$
\begin{aligned}
\ell(p)
&= \sum_{i=1}^{24}\log\binom{n_i}{x_i}
 +\log(p)\sum_{i=1}^{24}x_i
 +\log(1-p)\sum_{i=1}^{24}(n_i-x_i),\\
\frac{d\ell}{dp}
&= \frac{1}{p}\sum_{i=1}^{24}x_i-\frac{1}{1-p}\sum_{i=1}^{24}(n_i-x_i),\\
0 &= \frac{1}{\hat p}\sum_{i=1}^{24}x_i-\frac{1}{1-\hat p}\sum_{i=1}^{24}(n_i-x_i)
\quad\Longrightarrow\quad
\hat p=\frac{\sum_{i=1}^{24}x_i}{\sum_{i=1}^{24}n_i}.
\end{aligned}
$$

The first term of the log-likelihood does not contain $p$, so it disappears when we
differentiate. To solve the last line, multiply both sides by $\hat p(1-\hat p)$, which
gives $(1-\hat p)\sum x_i=\hat p\sum(n_i-x_i)$, and collect the terms with $\hat p$. The
second derivative, $-\sum x_i/p^2-\sum(n_i-x_i)/(1-p)^2$, is negative for every $p$, so
this is a maximum. In R:

```{r}
# Total number of survivors and of tadpoles, and the maximum likelihood estimate from
# the derivation: p_hat = sum(x_i) / sum(n_i).
total_survivors = sum(survivors)
total_tadpoles = sum(tadpoles)
p_hat = total_survivors / total_tadpoles
c(survivors = total_survivors, tadpoles = total_tadpoles, p_hat = p_hat)

# The derivative of the log-likelihood, sum(x) / p - (sum(n) - sum(x)) / (1 - p),
# is zero at that value.
total_survivors / p_hat - (total_tadpoles - total_survivors) / (1 - p_hat)
```

In total 264 of the 560 tadpoles survived, so $\hat p=264/560\approx 0.471$, and the
derivative of the log-likelihood is zero at that value. The maximum likelihood estimator
(MLE) is the fraction of all the tadpoles that survived. That is, the tanks are pooled, and
a tank with 35 tadpoles weighs more in the estimate than a tank with 10.

### Question 3

A probability lies between zero and one, so we let the optimiser search over
$\operatorname{logit}p$, which can take any real value, and convert it into a probability
with `plogis()` inside the function. The NLL has the form of the theory: its first argument
is a list of parameters, with one element (`logit_p`), and its second argument a list of
data, with two elements (`survivors` and `tadpoles`). `dbinom()` receives the vector with
the initial number of tadpoles of each tank in its argument `size`. The line `ADREPORT(p)`
asks for the standard error of $p$ itself, which we use at the end of this question. We
connect the NLL to the data with `cmb()`, process it with `MakeADFun()` and start the
search at $p=0.5$, whose logit is zero:

```{r}
# Negative log-likelihood of the Binomial model with one survival probability p for all tanks.
# The tanks differ in the number of tadpoles n_i, so size is a vector.
binomial_nll = function(parms, data) {
  # Make the elements of both lists available by name.
  getAll(parms, data)
  # logit_p can take any real value and plogis() turns it into a probability.
  p = plogis(logit_p)
  # Ask for the standard error of p itself.
  ADREPORT(p)
  -sum(dbinom(x = survivors, size = tadpoles, prob = p, log = TRUE))
}

# Connect a likelihood to a data set.
cmb = function(f, d) function(p) f(p, d)

# Process the NLL of the 24 tanks, starting from p = 0.5 (logit 0).
frog_obj = MakeADFun(
  cmb(binomial_nll, list(survivors = survivors, tadpoles = tadpoles)),
  list(logit_p = 0),
  silent = TRUE
)

# Fit with BFGS; hessian = TRUE is used for the standard error by hand.
frog_fit = optim(
  par = frog_obj$par,
  fn = frog_obj$fn,
  gr = frog_obj$gr,
  method = "BFGS",
  hessian = TRUE
)
frog_fit$convergence

# Estimate on the probability scale, and the minimum of the negative log-likelihood.
plogis(frog_fit$par)
frog_fit$value
```

The convergence code is 0. By invariance, `plogis()` of the estimate of
$\operatorname{logit}p$ is the MLE of $p$: 0.471, the value that we derived on paper. The
minimum of the NLL is 98.94, so the maximum log-likelihood is $-98.94$.

There are two routes to the standard error. The first is the calculation by hand of the
theory. The Hessian that `optim()` returns refers to the parameter that it searched over,
so its inverse is the approximate variance of the estimate of $\operatorname{logit}p$:

```{r}
# The Hessian is a 1 by 1 matrix; take its single element. For this model the second
# derivative of the NLL with respect to logit(p) is sum(n) * p * (1 - p).
frog_fit$hessian[1, 1]
total_tadpoles * p_hat * (1 - p_hat)

# Standard error of the estimate of logit(p).
se_logit_p = sqrt(1 / frog_fit$hessian[1, 1])
se_logit_p
```

The second route is `sdreport()`, which does this calculation for any model and adds a row
for the quantity that we asked for with `ADREPORT()`. We calculate the Wald interval on
the logit scale and transform its endpoints back with `plogis()`:

```{r}
# Standard errors from sdreport(): the row logit_p is on the scale that was optimised and
# the row p is the quantity in ADREPORT().
frog_sd = sdreport(frog_obj)
frog_summary = summary(frog_sd)
frog_summary

# Wald interval on the logit scale, back-transformed to survival probabilities.
critical_value = qnorm(p = 0.975)
wald_p = plogis(
  frog_summary["logit_p", "Estimate"] +
    c(-1, 1) * critical_value * frog_summary["logit_p", "Std. Error"]
)
wald_p

# Standard error on the probability scale, from the formula for a proportion.
sqrt(p_hat * (1 - p_hat) / total_tadpoles)
```

The Hessian is 139.5, which agrees with the second derivative of the NLL calculated by
hand. By both routes the standard error of the estimate of $\operatorname{logit}p$ is
$1/\sqrt{139.5}\approx 0.085$, and the 95% Wald interval for the probability of survival
goes from 0.430 to 0.513. As in the theory, we transformed the endpoints and not the
standard error, because a standard error on the logit scale cannot be converted into a
probability with `plogis()`. The standard error of $\hat p$ itself is in the row `p` of
the table, 0.021: RTMB calculates it with the delta method, which
[Chapter 6](../../Chapter_6/Theory.qmd#sec-delta-method) explains. For this model we can check it with a
formula: the observed information for $p$ is $\sum n_i/[\hat p(1-\hat p)]$, so the
standard error is $\sqrt{\hat p(1-\hat p)/\sum n_i}\approx 0.021$.

### Question 4

The model has one parameter, so the profile is the log-likelihood itself. The likelihood
interval contains the values of $p$ whose relative log-likelihood is above the cutoff of
$-1.92$ (see [Profile likelihood intervals](../Theory.qmd#sec-profile_intervals)). By hand,
we write the relative log-likelihood as a function of $p$ and use `uniroot()` to find the
two values at which it equals the cutoff. `uniroot()` needs an interval whose ends give
values of opposite sign. The function that we give it is positive at the estimate, so we
search between 0.2 and $\hat p$ and between $\hat p$ and 0.8, after checking that the
relative log-likelihood is below the cutoff at 0.2 and at 0.8:

```{r}
# Cutoff for a 95% interval: half the 95% quantile of a chi-squared distribution
# with one degree of freedom.
cutoff = qchisq(p = 0.95, df = 1) / 2

# Log-likelihood of the Binomial model as a function of p, relative to the maximum.
binomial_relative = function(p) {
  sum(dbinom(x = survivors, size = tadpoles, prob = p, log = TRUE)) +
    frog_fit$value
}

# Relative log-likelihood at the outer ends of the two search intervals.
binomial_relative(p = 0.2)
binomial_relative(p = 0.8)

# uniroot() finds where the relative log-likelihood equals -cutoff, below and above the
# estimate.
p_lower = uniroot(
  f = function(p) binomial_relative(p) + cutoff,
  interval = c(0.2, p_hat),
  tol = 1e-8
)$root
p_upper = uniroot(
  f = function(p) binomial_relative(p) + cutoff,
  interval = c(p_hat, 0.8),
  tol = 1e-8
)$root
c(lower = p_lower, upper = p_upper)
```

`TMB::tmbprofile()` calculates the profile of a parameter of the object, and `confint()`
finds the two crossings of the cutoff. The profile is calculated on the logit scale, so we
transform the endpoints with `plogis()`:

```{r}
# Profile of logit(p) and the endpoints of its 95% interval, transformed back.
frog_profile = TMB::tmbprofile(frog_obj, name = "logit_p", trace = FALSE)
plogis(confint(frog_profile))
```

The relative log-likelihood is $-103.7$ at $p=0.2$ and $-148.1$ at $p=0.8$, far below the
cutoff, so each search interval contains one crossing. By hand and with `tmbprofile()`
the likelihood interval goes from 0.430 to 0.513, the same as the Wald interval to three
decimals. The two methods agree because the sample is large: with 560 tadpoles the
log-likelihood is very close to a parabola on the logit scale, which is the approximation
behind the Wald interval. Under the Binomial model with independent tanks, we would report
that the probability of survival is 0.47 with a 95% confidence interval from 0.43 to 0.51.

### Question 5

The Binomial model gives the variance of the number of survivors, $n_ip(1-p)$, as well as
its mean, $n_ip$. The experiment has several tanks for each initial density, so we can
compare the variance of the number of survivors among the tanks of one density (calculated
with `var()`, which uses the divisor $n-1$) with the variance of the fitted model:

```{r}
# Variance of the number of survivors among tanks of the same initial density, and the
# variance that the Binomial model gives, n * p * (1 - p), with the estimated p.
densities = sort(unique(tadpoles))
variance_comparison = data.frame(
  density = densities,
  n_tanks = NA,
  observed_variance = NA,
  binomial_variance = NA
)
for (i in seq_along(densities)) {
  in_group = tadpoles == densities[i]
  variance_comparison$n_tanks[i] = sum(in_group)
  variance_comparison$observed_variance[i] = var(survivors[in_group])
  variance_comparison$binomial_variance[i] = densities[i] * p_hat * (1 - p_hat)
}
variance_comparison
```

There are eight tanks at each density. With 10 tadpoles the variance among tanks is 3.7,
against 2.5 under the model. With 25 tadpoles it is 24.8 against 6.2, and with 35 tadpoles
64.3 against 8.7. So the number of survivors varies more among tanks than the Binomial
model allows, up to seven times more: the counts are overdispersed relative to the Binomial
distribution (see [Chapter 2](../../Chapter_2/Theory.qmd#sec-discrete_distributions)). A
Binomial model with one $p$ assumes that every tadpole survives with the same probability
and independently of the others. Tadpoles in the same tank share their predators and their
conditions, so the probability of survival is likely to differ among tanks. The tanks also
differed in the size of the tadpoles (the variable `size`), which is a candidate predictor
of survival.

The consequence for questions 3 and 4 is that both intervals are too narrow. They are
calculated as if the model had generated the data, that is, as if we had 560 independent
tadpoles, while most of the variation is among the 24 tanks. The estimate of 0.47 is still
the overall fraction of survivors. A model in which $p$ varies among tanks is the
[Beta-Binomial](../../Supplements/distributions.qmd#sec-beta_binomial) distribution of the
supplement, and models for observations that are grouped are the subject of
[Chapter 7](../../Chapter_7/Theory.qmd).

### Question 6

The two treatments belong to the same experiment, so we can compare them with a model for
all 48 tanks. The first model is the one of question 3 applied to all the tanks, with one
probability of survival ($K=1$). In the second model the tanks with predators have a
probability $p_{\mathrm{pred}}$ and the tanks without predators a probability
$p_{\mathrm{no}}$ ($K=2$), and the contribution of a tank is calculated with the
probability of its treatment. For the first model we connect `binomial_nll()` of question
3 to the 48 tanks: the list of data that we give to `cmb()` must use the names that the
NLL expects (`survivors` and `tadpoles`), whatever the objects are called in our session.
For the second model we write a new NLL in which `logit_p` is a vector with the logit of
the two probabilities, and the data include the number of the treatment of each tank
(`group`), which selects the probability of that tank. Both models are compared with AIC
and with AICc, which now uses the 48 tanks as $n$:

```{r}
# All 48 tanks: survivors, initial number of tadpoles and treatment ("no" or "pred").
survivors_all = ReedfrogPred$surv
tadpoles_all = ReedfrogPred$density
treatment = ReedfrogPred$pred
n_tanks_all = length(survivors_all)
table(treatment)

# Number of the treatment of each tank: as.numeric() of a factor returns the position of
# each level, so 1 is without predators ("no") and 2 with predators ("pred").
group = as.numeric(treatment)

# NLL of the model with one probability for each treatment. The first element of logit_p is
# the logit of the survival probability without predators and the second with predators.
treatment_nll = function(parms, data) {
  getAll(parms, data)
  p = plogis(logit_p)
  # Ask for the standard error of the difference between the two probabilities.
  difference = p[1] - p[2]
  ADREPORT(difference)
  # p[group] is the probability of the treatment of each tank.
  -sum(dbinom(x = survivors, size = tadpoles, prob = p[group], log = TRUE))
}

# Model with one probability: binomial_nll() of question 3 connected to the 48 tanks.
pooled_obj = MakeADFun(
  cmb(binomial_nll, list(survivors = survivors_all, tadpoles = tadpoles_all)),
  list(logit_p = 0),
  silent = TRUE
)
pooled_fit = optim(
  par = pooled_obj$par,
  fn = pooled_obj$fn,
  gr = pooled_obj$gr,
  method = "BFGS"
)

# Model with one probability for each treatment: the parameter is a vector of two logits.
treatment_obj = MakeADFun(
  cmb(
    treatment_nll,
    list(survivors = survivors_all, tadpoles = tadpoles_all, group = group)
  ),
  list(logit_p = c(0, 0)),
  silent = TRUE
)
treatment_fit = optim(
  par = treatment_obj$par,
  fn = treatment_obj$fn,
  gr = treatment_obj$gr,
  method = "BFGS"
)
c(pooled = pooled_fit$convergence, treatment = treatment_fit$convergence)

# Estimates of the survival probabilities, and the fractions of survivors that they should
# equal: all the tadpoles, and the tadpoles of each treatment.
plogis(pooled_fit$par)
plogis(treatment_fit$par)
sum(survivors_all) / sum(tadpoles_all)
tapply(survivors_all, treatment, sum) / tapply(tadpoles_all, treatment, sum)

# AIC = 2K - 2 * log-likelihood and AICc = AIC + 2K(K + 1) / (n - K - 1), with n the
# number of tanks. The minimum of the NLL is minus the maximum log-likelihood.
k_models = c(pooled = 1, treatment = 2)
nll_models = c(pooled = pooled_fit$value, treatment = treatment_fit$value)
aic_tanks = 2 * k_models + 2 * nll_models
aicc_tanks = aic_tanks + 2 * k_models * (k_models + 1) / (n_tanks_all - k_models - 1)
rbind(aic = aic_tanks, aicc = aicc_tanks)

# Standard errors: rows 1 and 2 are the two logits and row 3 the difference between the
# probabilities, from ADREPORT().
treatment_summary = summary(sdreport(treatment_obj))
treatment_summary

# 95% Wald intervals: on the logit scale with the endpoints transformed back for the two
# probabilities, and on the original scale for their difference.
treatment_estimate = unname(treatment_summary[, "Estimate"])
treatment_se = unname(treatment_summary[, "Std. Error"])
treatment_lower = treatment_estimate - critical_value * treatment_se
treatment_upper = treatment_estimate + critical_value * treatment_se
rbind(
  no = plogis(c(lower = treatment_lower[1], upper = treatment_upper[1])),
  pred = plogis(c(lower = treatment_lower[2], upper = treatment_upper[2])),
  difference = c(lower = treatment_lower[3], upper = treatment_upper[3])
)
```

Both fits have convergence code 0. The model with one probability estimates 0.699, the
fraction of the 1,120 tadpoles that survived. The model with two probabilities estimates
0.927 without predators and 0.471 with predators. These are the fractions of survivors in
each group of tanks, as the derivation of question 2 predicts for a Binomial model, and the
second is the estimate of question 3. The AIC is 574.5 for one probability and 274.3 for
one probability per treatment, and the AICc is 574.6 and 274.5. The difference of about
300 units is far larger than the two units that the additional parameter costs, and the
AICc gives the same comparison.

The ecological question is about the size of the effect. The probability of survival is
0.47 with predators (95% Wald interval from 0.43 to 0.51) and 0.93 without them (from 0.90
to 0.95), so the presence of predators lowers it by 0.46, from 0.93 to 0.47. The difference
is a quantity calculated from the two parameters, so we asked for it with `ADREPORT()`: its
standard error is 0.024, which gives a 95% Wald interval from 0.41 to 0.50.

The evidence is weaker than the difference in AIC suggests. Question 5 showed that the
counts of the tanks with predators vary more than the Binomial model allows, so their
interval is too narrow and the difference in AIC overstates the evidence. The direction
and the size of the effect are still what the data show. This is the first model of the
book in which a parameter differs between groups. In [Chapter 6](../../Chapter_6/Theory.qmd)
parameters will depend on continuous predictors, and in
[Chapter 7](../../Chapter_7/Theory.qmd) on groups of observations.

# Exercise 2: Glacier lily seedlings in quadrats

The data set `Lily_sum` in the package `emdbook` contains counts of the glacier lily
*Erythronium grandiflorum* in a grid of 16 × 16 quadrats of 2 m × 2 m in Washington Gulch,
sampled in 1992 (Thomson et al. 1996). We use the column `seedlings`, the number of
seedlings in each of the 256 quadrats. Load the data with
`data(Lily_sum, package = "emdbook")`.

Let $X_i$ be the number of seedlings in quadrat $i$. We use a
[Negative Binomial](../../Supplements/distributions.qmd#sec-negative_binomial) model with
mean $\mu$ and shape $k$,

$$
X_i \sim \operatorname{NegBin}(\mu, k), \qquad \operatorname{Var}(X_i)=\mu+\frac{\mu^2}{k},
$$

and assume that the quadrats are i.i.d. Neighbouring quadrats of a grid are likely to
resemble each other, so this is a simplification, as it was for the visits to owl nests in
the Chapter 3 practical (models for observations that are not independent start in
[Chapter 7](../../Chapter_7/Theory.qmd)).

1. Estimate $\mu$ and $k$ with the method of moments, as in the Chapter 3 practical. Also
   calculate the fraction of quadrats without seedlings and the largest count.
2. Write the NLL of the model with $\log\mu$ and $\log k$ as parameters, process it with
   `MakeADFun()` and minimise it with `optim()`, using the method-of-moments estimates as
   starting values. Report the estimates on the original scale. Compare the log-likelihood
   at the maximum likelihood estimates with the log-likelihood at the method-of-moments
   estimates.
3. Calculate the log-likelihood on a grid of values of $\mu$ and $k$ with two nested `for`
   loops that call the function `fn` of the object, and draw the contours of the relative
   log-likelihood with `contour()`. Add both pairs of estimates and the contour of the
   approximate joint 95% confidence region. Is the method-of-moments estimate inside the
   region?
4. Calculate a 95% Wald interval for $k$ (on the log scale, with the endpoints transformed
   back) with the standard error from `sdreport()`. Then calculate the profile of $k$ by
   hand on a grid of values of $k$, as in the theory, and find the values of $k$ that are
   within 1.92 units of the maximum. Check the result with `TMB::tmbprofile()` and
   `confint()`, and compare the profile likelihood interval with the Wald interval.
5. Plot the relative frequencies of the counts together with the probabilities of the
   fitted Negative Binomial model and of a Poisson model. Compare the fraction of empty
   quadrats under each model with the observed fraction, and compare the two models with
   AIC and with its small-sample correction AICc (see
   [Model comparison with AIC](../Theory.qmd#sec-aic)).

# Exercise 3: Empty quadrats and zero inflation

About half of the quadrats of Exercise 2 have no seedlings. The Negative Binomial model
describes this with a small shape $k$. A different explanation is that a fraction of the
quadrats cannot have seedlings at all (for example, because no seeds arrive there), while
the counts in the rest follow a Poisson distribution. This is a zero-inflated Poisson
model (see
[How to make zero-inflated distributions](../../Supplements/distributions.qmd#sec-zero_inflated)
in the supplement). Its probability mass function is

$$
P(X=x\mid\lambda,p_0)=
\begin{cases}
p_0+(1-p_0)\,e^{-\lambda}, & x=0,\\[4pt]
(1-p_0)\,\dfrac{e^{-\lambda}\lambda^{x}}{x!}, & x>0,
\end{cases}
$$

where $p_0$ (`pzero` in the code of the supplement) is the probability that a quadrat
cannot have seedlings and $\lambda$ is the mean number of seedlings in the quadrats that
can have them. Use the same counts as in Exercise 2 and the same assumption of i.i.d.
quadrats.

1. Write a function `my_dzipois(x, lambda, pzero, log = FALSE)` that returns the
   probability mass of the zero-inflated Poisson distribution, or its logarithm when
   `log = TRUE`. Check that the probabilities add up to one over a wide range of counts,
   for two combinations of parameter values, and that your function returns the same
   probabilities as `dzipois()` of the package RTMBdist, in which $p_0$ is the argument
   `zeroprob`.
2. Write the NLL with `dzipois()` and with $\log\lambda$ and $\operatorname{logit}p_0$ as
   parameters. Choose starting values from the data (think of the fraction of empty
   quadrats and of the mean of the quadrats that are not empty) and fit the model.
   Interpret both estimates.
3. Compare the zero-inflated Poisson model with the Negative Binomial model of Exercise 2
   using AIC and AICc. What does the comparison come down to when both models have the
   same number of parameters?
4. Plot the probabilities of both fitted models over the relative frequencies of the
   counts. Under each model, calculate the probability of an empty quadrat and the
   probability of 20 or more seedlings, and compare them with the observed fractions. What
   does each model say about how the counts arise? Can the comparison identify the
   ecological cause of the empty quadrats?

# Exercise 4: Timber volume and body mass

This exercise fits models to continuous measurements, so the contributions to the
likelihood are densities. We use a [Gamma](../../Supplements/distributions.qmd#sec-gamma)
model with shape $k$ and rate $r$ (`shape` and `rate` in R), for which

$$
X_i \sim \operatorname{Gamma}(k, r), \qquad E[X_i]=\frac{k}{r},
\qquad \operatorname{Var}(X_i)=\frac{k}{r^2}.
$$

The first three questions use the data set `trees` of the Chapter 3 practical, with the
measurements of 31 felled black cherry trees. That practical used their heights. Here $X_i$
is the volume of timber of tree $i$ in cubic feet (ft³), the column `Volume`. The last two
questions use the data set `penguins`, which is part of the `datasets` package in R 4.5.0
and later.^[With an older version of R, install the package `palmerpenguins` and load the
data with `data(penguins, package = "palmerpenguins")`. It contains the same measurements,
but the column with the body mass is called `body_mass_g`.] It contains measurements of
adult penguins of three species near Palmer Station, Antarctica (Gorman, Williams and
Fraser 2014). There $X_i$ is the body mass of Adelie penguin $i$, the column `body_mass`,
in grams. In both cases we assume that the observations are i.i.d.

1. Load the trees with `data(trees, package = "datasets")`. Write the NLL of the Gamma
   model with `dgamma(..., log = TRUE)` and with $\log k$ and $\log r$ as parameters, and
   mark the data with `OBS()`, which Exercise 5 uses to simulate. Calculate
   method-of-moments estimates from the formulas of the mean and the variance, use them as
   starting values and fit the model. Report the estimates of $k$ and $r$ with their units
   and the maximum likelihood estimate of the mean volume.
2. Calculate the correlation between the estimates of $\log k$ and $\log r$ with
   `sdreport()`. Calculate the log-likelihood on a grid of values of $k$ and $r$ and draw
   the contours of the relative log-likelihood. What does the tilt of the contours mean?
3. Calculate a slice of the log-likelihood for the shape $k$ on a grid of values, and its
   profile with `TMB::tmbprofile()`, and plot both as relative log-likelihoods. Find the
   range of values of $k$ for which each curve is within 1.92 units of the maximum (for the
   profile, with `confint()`). Which of the two ranges is a confidence interval, and why is
   the slice narrower?
4. Load the penguins with `data(penguins, package = "datasets")` and keep the body masses
   of the Adelie penguins. One of them is missing: remove it explicitly. Fit the Gamma
   model to the masses in grams and again to the same masses in kilograms. Start both fits
   at a shape of 1 and at the rate that gives the mean of the sample, $1/\overline x$ (for
   these data the method-of-moments estimates are a poor starting point, as the solution
   shows; [Chapter 8](../../Chapter_8/Theory.qmd) explains how a fit depends on its
   starting values). Compare the estimates and the maximum log-likelihoods of the two
   fits. Explain the differences with the units of a density.
5. Fit a [LogNormal](../../Supplements/distributions.qmd#sec-lognormal) model to the body
   masses with `dlnorm()`, in grams and in kilograms, and compare it with the Gamma model
   using AIC and AICc. Does the comparison depend on the units?

# Exercise 5: Estimators and intervals across hypothetical experiments

This exercise is extra practice. With a single sample we cannot tell how far an estimate
is from the parameter, or whether a confidence interval contains it, because the parameter
is unknown. As in Exercise 3 of the Chapter 3 practical, we can study the behaviour of an
estimator by simulating hypothetical experiments from a model whose parameters we know.
Here we use the maximum likelihood estimates of Exercise 4 as the parameters of that model.
The simulation then shows how the estimators and the intervals would behave if the fitted
Gamma model had generated the data (see
[Properties of maximum likelihood estimators](../Theory.qmd#sec-mle_properties) and
[Uncertainty from the likelihood](../Theory.qmd#sec-likelihood-uncertainty)). This
exercise continues with the objects of Exercise 4. The function that runs the simulation
is given below the questions with three gaps, which you fill in.

1. For the shape $k$ of the Gamma model fitted to the tree volumes, calculate a 95% Wald
   interval (on the log scale, with the endpoints transformed back) with the standard
   error from `sdreport()`. Compare it with the profile likelihood interval of Exercise 4.
2. Complete the first gap of the function `simulate_gamma_experiments()` below. Set seed
   801 and simulate 2,000 hypothetical experiments, each with 31 trees from a Gamma model
   whose shape and rate are the maximum likelihood estimates of Exercise 4. For each
   experiment, the function estimates the shape by the method of moments and by maximum
   likelihood (the fit starts at a shape of 1 and at the rate that gives the mean of the
   simulated sample, as in question 4 of Exercise 4). Compare the mean, the standard
   deviation and the mean squared error of the two estimators with the shape that you used
   to simulate. Tip: the function has the sample size, the shape and the rate as arguments
   because you will need it again in question 4.
3. Complete the second and the third gap, which record for each experiment whether the
   Wald interval and the profile likelihood interval for the shape contain the shape that
   you used to simulate, and calculate the coverage of both intervals. Hint: you do not
   need `tmbprofile()` inside the simulation. A profile likelihood interval contains a
   value whenever the profile log-likelihood at that value is within
   `qchisq(p = 0.95, df = 1) / 2` of the maximum log-likelihood, so one extra fit with the
   shape fixed at that value is enough.
4. Set seed 802 and repeat questions 2 and 3 with experiments of 151 penguins from the
   Gamma model fitted to the body masses in grams. How do the bias and the coverage change?
   The two simulations differ in the sample size and in the model. How can you check which
   of the two explains the change? Do it, without resetting the seed.
5. Why do these results only approximate the behaviour of the estimators and the intervals
   for the real trees and penguins?

The function below needs `gamma_nll()` from Exercise 4, the helper `cmb()`, and the objects
`critical_value` and `cutoff` (the critical value of the standard Normal distribution and
the cutoff of the relative log-likelihood for a 95% interval). It draws every experiment
with the function `simulate()` of an object that has placeholder data of the right length,
as in the theory, and stores the results of every experiment in one row of a data frame.
In each gap, replace `NA` by your own code. The lines that follow a gap use the names of
the objects that the gap creates, so keep those names.

```{r}
#| eval: false
# One simulation of hypothetical experiments from a Gamma model whose parameters we know.
# For each experiment: the maximum likelihood and the method-of-moments estimates of the
# shape, and whether the Wald interval (log scale) and the profile interval contain the
# shape that was used to simulate.
simulate_gamma_experiments = function(n_obs, shape, rate, n_experiments) {
  results = data.frame(
    shape_ml = rep(NA, times = n_experiments),
    shape_mom = rep(NA, times = n_experiments),
    wald_covered = rep(NA, times = n_experiments),
    profile_covered = rep(NA, times = n_experiments),
    converged = rep(NA, times = n_experiments)
  )
  # An object with placeholder data of the right length, from which the experiments are
  # simulated: the values of the data are not used.
  placeholder_obj = MakeADFun(
    cmb(gamma_nll, list(x = rep(0, times = n_obs))),
    list(log_shape = 0, log_rate = 0),
    silent = TRUE
  )
  for (e in seq_len(n_experiments)) {
    # One hypothetical experiment (simulate() takes the parameters on the log scale) and
    # its method-of-moments estimate of the shape.
    x = placeholder_obj$simulate(log(c(shape, rate)))$x
    x_mean = mean(x)
    x_variance = mean((x - x_mean)^2)
    shape_mom = x_mean^2 / x_variance

    # GAP 1: connect gamma_nll() to x and process it with MakeADFun(), with a shape of 1
    # and a rate of 1 / x_mean as starting values (on the log scale), and store the result
    # in an object called sim_obj. Then fit it with optim() and method = "BFGS", and store
    # the result in an object called fit.
    sim_obj = NA
    fit = NA
    shape_ml = exp(fit$par[["log_shape"]])

    # The Wald interval contains the shape when the estimate is within the critical value
    # of standard errors on the log scale. The profile interval contains it when the
    # relative profile log-likelihood at that shape is within the cutoff.

    # GAP 2: the standard error of the estimate of log(shape), from sdreport() of sim_obj,
    # in an object called se_log_shape. Then TRUE in wald_covered when the Wald interval
    # contains the shape that was used to simulate.
    se_log_shape = NA
    wald_covered = NA

    # GAP 3: process gamma_nll() for x again with log_shape fixed at the logarithm of the
    # shape that was used to simulate (the argument map of MakeADFun(), as in the profile
    # by hand of Exercise 2) and log_rate starting at its estimate in fit, and store the
    # result in an object called fixed_obj. Fit it with optim(), which re-estimates the
    # rate only, and store the result in an object called fixed_fit. Then record in
    # profile_covered whether the relative profile log-likelihood at that shape is within
    # the cutoff.
    fixed_obj = NA
    fixed_fit = NA
    profile_covered = NA

    results$shape_ml[e] = shape_ml
    results$shape_mom[e] = shape_mom
    results$wald_covered[e] = wald_covered
    results$profile_covered[e] = profile_covered
    results$converged[e] = fit$convergence == 0 && fixed_fit$convergence == 0
  }
  results
}
```
