---
title: "Chapter 4: Maximum likelihood"
date: today
---

These exercises use illustrative or simulated ecological data, not field
observations. First calculate the small-sample likelihoods **on paper,
without R**. A calculator is fine for exponentials and logarithms.
Then use R for larger samples and numerical optimisation. The
[Chapter 4 theory](../Theory.qmd) introduces each method.

# Exercise 1: Seedlings in quadrats

Suppose the count $X_i$ in each equal-area quadrat follows an independent
Poisson model with an unknown common mean $\lambda$ seedlings per
quadrat. Four illustrative quadrats contain 0, 1, 2, and 3 seedlings.

1. Write the Poisson probability mass for one count. By hand, write
   each of the four likelihood contributions and multiply them to
   obtain $L(\lambda\mid0,1,2,3)$.
2. By hand, take the logarithm of the product and simplify
   $\ell(\lambda\mid0,1,2,3)$ into a sum. Calculate the likelihood
   and log-likelihood at $\lambda=1$ and $\lambda=2$. Which of these
   two candidate means gives the observations greater likelihood?
3. Calculate the sample mean. Use it to give the maximum likelihood
   estimate and its ecological units.
4. Check the hand calculations with `dpois()`. Write a negative
   log-likelihood using `log = TRUE`, fit $\lambda$ with `optim()`
   on the log scale, check convergence, and compare its answer with
   the sample mean.

## Solution

The observational unit is one quadrat. Under the assumed model,

$$
P(X_i=x_i\mid\lambda)
=\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}.
$$

Independence lets us multiply the four contributions. For the
observed counts, these are
$e^{-\lambda}\lambda^0/0!$,
$e^{-\lambda}\lambda^1/1!$,
$e^{-\lambda}\lambda^2/2!$, and
$e^{-\lambda}\lambda^3/3!$. Therefore,

$$
\begin{aligned}
L(\lambda\mid0,1,2,3)
&=\prod_{x\in\{0,1,2,3\}}
  \frac{e^{-\lambda}\lambda^x}{x!}\\
&=\frac{e^{-4\lambda}\lambda^{0+1+2+3}}
        {0!\,1!\,2!\,3!}
=\frac{e^{-4\lambda}\lambda^6}{12}.
\end{aligned}
$$

Taking logs changes the product into a sum:

$$
\begin{aligned}
\ell(\lambda\mid0,1,2,3)
&=\sum_{x\in\{0,1,2,3\}}
  [-\lambda+x\log\lambda-\log(x!)]\\
&=-4\lambda+6\log\lambda-\log(12).
\end{aligned}
$$

Substitute the two candidates:

| Candidate mean | Likelihood | Log-likelihood |
|---|---:|---:|
| $\lambda=1$ | $e^{-4}/12\approx0.001526$ | $-4-\log12\approx-6.485$ |
| $\lambda=2$ | $64e^{-8}/12\approx0.001789$ | $-8+6\log2-\log12\approx-6.326$ |

The likelihood is greater at $\lambda=2$ than at $\lambda=1$.
This comparison says nothing about the probability that either
parameter value is true. The MLE is the sample mean,
$(0+1+2+3)/4=1.5$ seedlings per quadrat. It can lie between the two
candidates we compared.

In R, sum the log probabilities directly. The negative sign gives
`optim()` a function to minimise. Exponentiating `log_lambda` keeps
the candidate Poisson mean positive.

```{r}
counts = c(0, 1, 2, 3)
candidate_values = c(1, 2)
likelihood = vapply(
  X = candidate_values,
  FUN = function(lambda) {
    prod(dpois(x = counts, lambda = lambda))
  },
  FUN.VALUE = numeric(1)
)
log_likelihood = vapply(
  X = candidate_values,
  FUN = function(lambda) {
    sum(dpois(x = counts, lambda = lambda, log = TRUE))
  },
  FUN.VALUE = numeric(1)
)
data.frame(lambda = candidate_values, likelihood, log_likelihood)

poisson_nll = function(log_lambda, counts) {
  lambda = exp(log_lambda)
  -sum(dpois(x = counts, lambda = lambda, log = TRUE))
}
fit = optim(
  par = log(1), fn = poisson_nll, counts = counts,
  method = "BFGS"
)
fit$convergence
c(optim_estimate = exp(fit$par), sample_mean = mean(counts))
```

The optimiser should return a convergence code of zero and an
estimate close to 1.5 seedlings per quadrat. The likelihood
comparison used only two candidates; optimisation searched beyond
them. With more counts, computing a product directly can underflow,
which is why the fitting function sums log probabilities.

# Exercise 2: Tree heights and two parameters

Assume independent tree heights follow
$H_i\sim\operatorname{Normal}(\mu,\sigma^2)$, where both $\mu$ and
$\sigma>0$ are measured in metres. For a first hand calculation,
consider the illustrative heights 21 m and 23 m.

1. Without R, write the normal **density** contribution for each
   height under candidate A, $(\mu,\sigma)=(22,1)$ m, and candidate B,
   $(22,3)$ m. Multiply each pair to obtain its likelihood. Calculate
   each log-likelihood as a sum. Which pair fits these two heights
   better? Explain why a density contribution is not the probability
   of an exact height.
2. Set seed 522 and simulate 30 independent heights with generating
   values $\mu=22$ m and $\sigma=4$ m. Write a negative log-likelihood
   for $\mu$ and $\log\sigma$. Fit both parameters with `optim()`,
   check convergence, and report the estimates in metres.
3. Draw a contour plot of the relative log-likelihood over a sensible
   grid of $\mu$ and $\sigma$. Mark the estimate and generating values.
   Explain why they need not coincide.
4. Describe how you would form a slice for $\mu$ and a profile for
   $\mu$. Which one re-estimates $\sigma$?

Hint: use `dnorm(x = heights, mean = mu, sd = sigma, log = TRUE)`
inside the fitting function. The generating values are known only
because you simulated the sample.

# Exercise 3: Overdispersed seedling counts

The [Negative Binomial model](../../Supplements/distributions.qmd#sec-negative_binomial)
has mean $\mu>0$ seedlings per quadrat and shape $k>0$.
Its variance is $\mu+\mu^2/k$. When $k=1$, the probability mass
simplifies to

$$
P(X=x\mid\mu,k=1)
=\frac{1}{1+\mu}
 \left(\frac{\mu}{1+\mu}\right)^x.
$$

1. Two illustrative quadrats contain 0 and 1 seedlings. **By hand**,
   write both likelihood contributions and their product with $k=1$.
   Evaluate the likelihood and log-likelihood at $\mu=1$ and $\mu=2$.
   Which candidate mean has greater likelihood for these counts?
2. Set seed 533. Simulate 80 independent quadrat counts with
   $\mu=4$ and $k=1.5$. Compare their sample mean and variance with
   the model mean and variance. Explain why this comparison alone
   cannot identify an ecological cause of overdispersion.
3. Write a negative log-likelihood using `dnbinom(..., log = TRUE)`.
   Fit $\mu$ and $k$ with `optim()` on log scales. Use sensible
   starting values, check convergence, and report both estimates
   on their original scales.
4. Compare observed relative frequencies with fitted negative
   binomial probabilities in a plot. What would make you question
   the fit, especially at zero and in the upper tail?

# Exercise 4: Uncertainty in mean tree height

Suppose 30 independent tree heights follow a normal model with
unknown mean $\mu$ and standard deviation $\sigma$. Generate an
illustrative sample with:

```{r}
set.seed(seed = 544)
heights = rnorm(n = 30, mean = 22, sd = 4)
```

1. Fit $\mu$ and $\log\sigma$ by maximum likelihood using
   `optim(..., hessian = TRUE)`. Check convergence and report both
   estimates in metres.
2. Invert the Hessian and obtain approximate standard errors on
   the fitted scale. Construct a 95% quadratic interval for $\mu$.
   Construct one for $\log\sigma$ and back-transform its endpoints.
   Explain why these intervals are approximate.
3. For each candidate $\mu$, maximise the normal log-likelihood over
   $\sigma$. Use the profile log-likelihood and the cutoff
   $-qchisq(0.95,1)/2$ relative to its maximum to find a 95% profile
   interval for $\mu$. A plot can help locate the two crossings.
4. Compare the intervals for $\mu$ and interpret them in ecological
   units. State one assumption that matters for their coverage.
   Would either interval give a 95% probability that the fixed
   population mean lies between its realised endpoints?

Hint: for a fixed candidate $\mu$ in a normal model, the value of
$\sigma$ that maximises likelihood is
$\sqrt{\operatorname{mean}[(h_i-\mu)^2]}$.
