---
title: "Chapter 4: Estimation and sampling distributions"
date: today
---

These exercises use illustrative counts and simulated ecological
measurements, not field observations. Work through the first example
before trying the remaining exercises. The ideas and notation are
introduced in the [Chapter 4 theory](../Theory.qmd).

# Exercise 1: Seedlings in quadrats

Suppose the count $X_i$ in quadrat $i$ follows a Poisson model with an
unknown mean $\lambda$ seedlings per quadrat. Assume equal-area,
independent quadrats with the same expected count. These illustrative
counts come from 12 quadrats:

```{r}
seedling_counts = c(2, 4, 3, 5, 1, 6, 4, 2, 3, 5, 4, 3)
```

1. Name the parameter, a method-of-moments estimator, and its estimate
   from these counts. State the estimate in ecological units.
2. Calculate the estimated standard error of the mean under the Poisson
   model. Explain how it differs from the standard deviation of the
   individual counts.
3. Construct an approximate 95% Wald confidence interval for
   $\lambda$ using `qnorm()`. Interpret the interval and its confidence
   level. Name one model assumption that matters for this interpretation.

## Solution

The observational unit is one quadrat. The unknown parameter
$\lambda=E[X_i]$ is the expected number of seedlings per quadrat.
The method of moments matches the mean of the Poisson model with
the mean of the sampled counts, giving the estimator
$\hat\lambda=\overline X$.

```{r}
n_quadrats = length(seedling_counts)
lambda_hat = mean(seedling_counts)
c(n_quadrats = n_quadrats, estimate_per_quadrat = lambda_hat)
```

The 12 counts sum to 42, so the estimate is $42/12=3.5$ seedlings
per quadrat. The parameter is fixed but unknown; the estimator is a
rule that could give another value for a different sample.

Under the independent Poisson model,
$\operatorname{SE}(\hat\lambda)=\sqrt{\lambda/n}$. We do not know
$\lambda$, so we insert $\hat\lambda$:

```{r}
estimated_se = sqrt(lambda_hat / n_quadrats)
count_sd = sd(seedling_counts)
c(estimated_se = estimated_se, count_sd = count_sd)
```

The estimated standard error is approximately 0.54 seedlings per
quadrat. It describes variation in estimated **mean counts** across
studies of 12 quadrats. The sample standard deviation describes
variation among the **individual quadrat counts**.

For a Wald interval, use the 97.5th percentile of a standard normal
distribution on each side of the estimate:

```{r}
critical_value = qnorm(p = 0.975)
lambda_interval = c(
  lower = lambda_hat - critical_value * estimated_se,
  upper = lambda_hat + critical_value * estimated_se
)
lambda_interval
```

The endpoints are approximately 2.44 and 4.56 seedlings per quadrat.
This is a range for the population **mean count**, not a range
containing 95% of individual counts. If we repeatedly sampled
quadrats and used this procedure, roughly 95% of the intervals
would contain the fixed $\lambda$ if the model and normal
approximation gave the procedure 95% coverage in this setting.
Independence and a common mean across equal-area quadrats are
important assumptions. Wald intervals can have poor coverage when
expected counts are small.

# Exercise 2: Breeding success at bird nests

Suppose each independently monitored nest either produces at least one
fledgling or none, with the same success probability $p=0.6$. The number of
successful nests in a study of $n$ nests follows a binomial model
$X\sim\operatorname{Binomial}(n,p)$. The true $p$ is known here because
you will simulate the studies.

1. Set seed 422. Simulate one study of 40 nests and estimate $p$ by
   matching the observed and model success proportions. Interpret the
   estimate.
2. Set seed 423. Simulate 2,000 studies of 20 nests and 2,000 studies
   of 100 nests. Calculate $\hat p=X/n$ for every study. Plot and
   compare the two sampling distributions.
3. For each study size, calculate the average estimate, simulated
   bias, variance, standard error, and mean squared error. Compare
   the simulated variance with $p(1-p)/n$. Explain what changes with
   sample size and why one study cannot establish bias.
4. Give one ecological reason why outcomes at different nests might
   fail to be independent.

Hint: one call to `rbinom()` can generate the success counts for all
2,000 studies at a given sample size.

# Exercise 3: Variation in tree heights

Suppose heights of independently sampled trees follow a normal model
with mean $\mu=22$ m and standard deviation $\sigma=4$ m. These are
the simulation parameters; in a field study they would be unknown.

1. Set seed 433 and simulate 25 heights. Estimate $\mu$ by method of
   moments. Estimate $\sigma^2$ using the mean squared deviation
   from the sample mean, with divisor $n$. Compare this variance
   estimate with `var()`. Give their units.
2. Set seed 434 and repeat the 25-tree study 2,000 times. For each
   study, calculate both variance estimates. Compare their average
   values and simulated biases with the true variance. The expected
   bias of the method-of-moments estimate is $-\sigma^2/n$.
3. Calculate the mean squared error of each variance estimator in
   the simulation. Does knowing which estimator is unbiased, by
   itself, tell you which has the smaller mean squared error?

Hint: `replicate()` can return two named estimates for each study.

# Exercise 4: Mean body mass and interval coverage

Suppose independent small-mammal body masses in grams follow a
lognormal model with `meanlog = log(30)` and `sdlog = 0.8`.
As in [Chapter 3](../../Chapter_3/Theory.qmd#sec-lognormal_distribution),
the model median on the original scale is $e^{\mu}$ and the
**mean mass** is $e^{\mu+\sigma^2/2}$, where $\mu$ and $\sigma$
describe log mass. The parameters are known because you will simulate
the studies.

1. Calculate the model mean and median mass in grams. Set seed 443,
   simulate one study of 12 animals, and calculate its sample mean,
   sample median, and estimated standard error of the sample mean.
2. For this study, form a 95% $t$ interval for the **original-scale
   mean mass** using the sample mean, `sd(mass) / sqrt(n)`, and
   `qt(p = 0.975, df = n - 1)`. Interpret the interval. Why is its
   coverage only approximate for these lognormal masses?
3. Set seed 444. Simulate 2,000 studies of 12 animals and 2,000
   studies of 100 animals. Recalculate the $t$ interval for every
   study and find the fraction covering the true original-scale mean
   at each size. Compare the fractions with 0.95 and explain why
   coverage might change with sample size.
4. State one reason why coverage in this simulation does not prove
   that the same procedure will work equally well for field data.

Hint: `replicate()` can return `TRUE` or `FALSE` for whether one
interval contains the model mean. `mean()` then gives the coverage
fraction.
