---
title: "Chapter 4: Maximum likelihood estimation — solutions"
date: today
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures on this page.
library(ggplot2)
```

<!-- PHASE 5: opening paragraph, as on the solutions page of Chapter 3 (Exercise 1 is solved on the student page) -->

# Exercise 2: Glacier lily seedlings in quadrats

<!-- PHASE 5: exercise statement -->

```{r}
# Load the glacier lily data from the emdbook package without attaching the package.
data(Lily_sum, package = "emdbook")

# Number of seedlings in each 2 m by 2 m quadrat, and the number of quadrats.
lily_counts = Lily_sum$seedlings
n_lily = length(lily_counts)
```

## Solution

### Question 1

```{r}
# Method-of-moments estimates, as in the practical of Chapter 3: the mean, the variance
# with divisor n, and the shape k from variance = mu + mu^2 / k.
mu_mom_lily = mean(lily_counts)
variance_lily = mean((lily_counts - mu_mom_lily)^2)
k_mom_lily = mu_mom_lily^2 / (variance_lily - mu_mom_lily)

c(
  n_quadrats = n_lily,
  mean = mu_mom_lily,
  variance = variance_lily,
  k_mom = k_mom_lily,
  fraction_empty = mean(lily_counts == 0),
  maximum = max(lily_counts)
)
```

<!-- VALUES: n_quadrats = 256; mean 3.96 seedlings per quadrat; variance 54.35 seedlings^2; k_mom = 0.311; fraction of empty quadrats 0.496; maximum count 72 -->

### Question 2

```{r}
# Negative log-likelihood of the Negative Binomial model with mean mu and shape k.
# par[1] is log(mu) and par[2] is log(k), so that both parameters stay positive.
nb_nll = function(par, counts) {
  mu = exp(par[1])
  k = exp(par[2])
  -sum(dnbinom(x = counts, mu = mu, size = k, log = TRUE))
}

# Start from the method-of-moments estimates, on the log scale.
start_lily = c(log_mu = log(mu_mom_lily), log_k = log(k_mom_lily))

# Fit with BFGS; hessian = TRUE is used for the standard errors in question 4.
lily_fit = optim(
  par = start_lily,
  fn = nb_nll,
  counts = lily_counts,
  method = "BFGS",
  hessian = TRUE
)
lily_fit$convergence

# Estimates on the original scale.
mu_mle_lily = unname(exp(lily_fit$par["log_mu"]))
k_mle_lily = unname(exp(lily_fit$par["log_k"]))
c(mu_mle = mu_mle_lily, k_mle = k_mle_lily)

# Log-likelihood at the maximum likelihood estimates and at the method-of-moments estimates.
loglik_mle_lily = -lily_fit$value
loglik_mom_lily = -nb_nll(par = start_lily, counts = lily_counts)
c(mle = loglik_mle_lily, mom = loglik_mom_lily)
```

<!-- VALUES: convergence code 0; mu_mle = 3.957 seedlings per quadrat; k_mle = 0.2615; log-likelihood at the MLE -564.23 (NLL 564.2); log-likelihood at the method-of-moments estimates -565.34 (difference 1.1) -->

### Question 3

```{r}
# Grids of candidate values for the mean (seedlings per quadrat) and the shape.
mu_grid_lily = seq(from = 2.5, to = 6, length.out = 71)
k_grid_lily = seq(from = 0.17, to = 0.42, length.out = 71)

# Fill a matrix with two nested loops: row i is the i-th mean and column j the j-th shape.
loglik_surface_lily = matrix(NA, nrow = length(mu_grid_lily), ncol = length(k_grid_lily))
for (i in seq_along(mu_grid_lily)) {
  for (j in seq_along(k_grid_lily)) {
    loglik_surface_lily[i, j] = sum(
      dnbinom(x = lily_counts, mu = mu_grid_lily[i], size = k_grid_lily[j], log = TRUE)
    )
  }
}

# Relative log-likelihood: the difference from the maximum found by optim().
relative_surface_lily = loglik_surface_lily - loglik_mle_lily

# Cutoff of the relative log-likelihood for a joint 95% region of two parameters:
# half the 95% quantile of a chi-squared distribution with two degrees of freedom.
joint_cutoff = qchisq(p = 0.95, df = 2) / 2
joint_cutoff
```

```{r}
#| label: fig-lily_surface
#| fig-cap: "Contours of the relative log-likelihood of the glacier lily seedling counts as a function of the mean and the shape of the Negative Binomial model. The circle is the maximum likelihood estimate, the triangle is the method-of-moments estimate and the thick contour is the approximate joint 95% confidence region."
# Contours of the relative log-likelihood with base R, then the joint region on top.
contour(
  x = mu_grid_lily,
  y = k_grid_lily,
  z = relative_surface_lily,
  levels = c(-12, -9, -6, -4, -2, -1),
  xlab = "Mean (seedlings per quadrat)",
  ylab = "Shape (k)"
)
contour(
  x = mu_grid_lily,
  y = k_grid_lily,
  z = relative_surface_lily,
  levels = -joint_cutoff,
  lwd = 2,
  col = "#D55E00",
  add = TRUE,
  drawlabels = FALSE
)

# Maximum likelihood estimate (circle) and method-of-moments estimate (triangle).
points(x = mu_mle_lily, y = k_mle_lily, pch = 16, col = "#D55E00")
points(x = mu_mom_lily, y = k_mom_lily, pch = 17)
```

<!-- VALUES: joint cutoff 2.996 (qchisq(0.95, 2) = 5.991); the method-of-moments estimate (mu 3.957, k 0.311) lies inside the joint region, near its upper edge in k (the region reaches k of about 0.35); the grids are wide enough to contain the region -->

### Question 4

```{r}
# 0.975 quantile of the standard Normal distribution and the cutoff for one parameter.
critical_value = qnorm(p = 0.975)
cutoff = qchisq(p = 0.95, df = 1) / 2

# Wald interval for k: the inverse Hessian approximates the covariance matrix of the
# estimates of log(mu) and log(k). Calculate on the log scale and back-transform the endpoints.
lily_se = sqrt(diag(solve(lily_fit$hessian)))
lily_se
wald_k_lily = exp(lily_fit$par["log_k"] + c(-1, 1) * critical_value * lily_se["log_k"])
wald_k_lily
```

```{r}
# Profile log-likelihood of k relative to the maximum: for each k the mean is
# re-estimated with optim() on log(mu), starting from the joint estimate.
nll_given_k_lily = function(log_mu, k, counts) {
  -sum(dnbinom(x = counts, mu = exp(log_mu), size = k, log = TRUE))
}
profile_relative_k_lily = function(k, counts) {
  inner_fit = optim(
    par = lily_fit$par["log_mu"],
    fn = nll_given_k_lily,
    k = k,
    counts = counts,
    method = "BFGS"
  )
  -inner_fit$value - loglik_mle_lily
}

# The profile is zero at the joint estimate.
profile_relative_k_lily(k = k_mle_lily, counts = lily_counts)

# uniroot() finds where the relative profile equals -cutoff, below and above the estimate.
# The ends of each search interval were chosen from the surface above.
k_lower_lily = uniroot(
  f = function(k) profile_relative_k_lily(k, counts = lily_counts) + cutoff,
  interval = c(0.05, k_mle_lily),
  tol = 1e-8
)$root
k_upper_lily = uniroot(
  f = function(k) profile_relative_k_lily(k, counts = lily_counts) + cutoff,
  interval = c(k_mle_lily, 1),
  tol = 1e-8
)$root

# Profile interval next to the Wald interval for the shape.
rbind(
  profile = c(lower = k_lower_lily, upper = k_upper_lily),
  wald = c(lower = wald_k_lily[[1]], upper = wald_k_lily[[2]])
)
```

<!-- VALUES: standard errors on the log scales 0.126 (mu) and 0.117 (k); Wald interval for k 0.208 to 0.329; profile interval for k 0.207 to 0.328 (no unit): the two agree to two decimals; the profile at the estimate is zero up to optimiser tolerance (3.6e-9) -->

### Question 5

```{r}
# Fitted probabilities of every count under the Negative Binomial and Poisson models,
# and the relative frequencies in the sample.
count_values_lily = 0:max(lily_counts)
observed_lily = tabulate(lily_counts + 1, nbins = length(count_values_lily)) / n_lily
nb_probability = dnbinom(x = count_values_lily, mu = mu_mle_lily, size = k_mle_lily)
poisson_probability = dpois(x = count_values_lily, lambda = mu_mom_lily)

# Fraction of empty quadrats: observed and under each model.
c(
  observed = mean(lily_counts == 0),
  negative_binomial = nb_probability[1],
  poisson = poisson_probability[1]
)

# Log-likelihood of the Poisson model at its maximum (the sample mean) and AIC of both
# models. AIC = 2K - 2 * log-likelihood, with K the number of estimated parameters.
loglik_poisson_lily = sum(dpois(x = lily_counts, lambda = mu_mom_lily, log = TRUE))
aic_lily = c(
  poisson = 2 * 1 - 2 * loglik_poisson_lily,
  negative_binomial = 2 * 2 - 2 * loglik_mle_lily
)
aic_lily
```

```{r}
#| label: fig-lily_fits
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of seedlings per quadrat (bars) and probabilities under the fitted Negative Binomial and Poisson models, up to 40 seedlings."
# Data frames for the sample (bars) and the models (lines and points).
observed_data_lily = data.frame(count = count_values_lily, frequency = observed_lily)
fitted_data_lily = rbind(
  data.frame(count = count_values_lily, probability = nb_probability, model = "Negative Binomial"),
  data.frame(count = count_values_lily, probability = poisson_probability, model = "Poisson")
)

ggplot() +
  geom_col(
    data = observed_data_lily,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data_lily,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data_lily,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(values = c("#D55E00", "grey20")) +
  coord_cartesian(xlim = c(-0.5, 40)) +
  labs(
    x = "Number of seedlings per quadrat",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: fraction of empty quadrats: observed 0.496, Negative Binomial 0.483, Poisson 0.019; log-likelihood of the Poisson model -1364.9; AIC Poisson 2731.8, Negative Binomial 1132.4 -->

# Exercise 3: Empty quadrats and zero inflation

<!-- PHASE 5: exercise statement -->

## Solution

### Question 1

```{r}
# Mass function of a zero-inflated Poisson distribution: with probability pzero a quadrat
# cannot have seedlings, otherwise the count follows a Poisson distribution.
dzipois = function(x, lambda, pzero, log = FALSE) {
  # Poisson probability scaled by the probability of belonging to the Poisson part.
  result = (1 - pzero) * dpois(x = x, lambda = lambda)
  # An empty quadrat can come from either part, so pzero is added to the probability of zero.
  result = ifelse(x == 0, pzero + result, result)
  if (log) log(result) else result
}

# The probabilities add up to one over a wide range of counts.
sum(dzipois(x = 0:200, lambda = 7.9, pzero = 0.5))
sum(dzipois(x = 0:200, lambda = 2, pzero = 0.1))
```

<!-- VALUES: both sums equal 1 -->

### Question 2

```{r}
# Negative log-likelihood of the zero-inflated Poisson model.
# par[1] is log(lambda) and par[2] is logit(pzero).
zip_nll = function(par, counts) {
  lambda = exp(par[1])
  pzero = plogis(par[2])
  -sum(dzipois(x = counts, lambda = lambda, pzero = pzero, log = TRUE))
}

# Starting values from the data: the fraction of empty quadrats for pzero and the mean
# of the non-empty quadrats for lambda.
start_zip = c(
  log_lambda = log(mean(lily_counts[lily_counts > 0])),
  logit_pzero = qlogis(p = mean(lily_counts == 0))
)

zip_fit = optim(
  par = start_zip,
  fn = zip_nll,
  counts = lily_counts,
  method = "BFGS"
)
zip_fit$convergence

# Estimates on the original scale.
lambda_zip = unname(exp(zip_fit$par["log_lambda"]))
pzero_zip = unname(plogis(zip_fit$par["logit_pzero"]))
c(lambda = lambda_zip, pzero = pzero_zip, loglik = -zip_fit$value)
```

<!-- VALUES: convergence code 0; lambda = 7.85 seedlings per quadrat; pzero = 0.496; NLL 848.0 (log-likelihood -848.0) -->

### Question 3

```{r}
# AIC of the zero-inflated Poisson model (K = 2) next to the Negative Binomial model of
# Exercise 2 (K = 2): with the same K, the comparison is a comparison of log-likelihoods.
aic_zip = 2 * 2 - 2 * (-zip_fit$value)
c(zero_inflated_poisson = aic_zip, negative_binomial = aic_lily[["negative_binomial"]])
```

<!-- VALUES: AIC zero-inflated Poisson 1700.0, Negative Binomial 1132.4 (difference 567.6) -->

### Question 4

```{r}
# Probability of zero and of 20 or more seedlings: observed, zero-inflated Poisson and
# Negative Binomial.
zip_probability = dzipois(x = count_values_lily, lambda = lambda_zip, pzero = pzero_zip)
c(
  zero_observed = mean(lily_counts == 0),
  zero_zip = zip_probability[1],
  zero_nb = nb_probability[1]
)
c(
  twenty_or_more_observed = mean(lily_counts >= 20),
  twenty_or_more_zip = sum(zip_probability[count_values_lily >= 20]),
  twenty_or_more_nb = pnbinom(q = 19, mu = mu_mle_lily, size = k_mle_lily, lower.tail = FALSE)
)
```

```{r}
#| label: fig-lily_zip
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of seedlings per quadrat (bars) and probabilities under the fitted zero-inflated Poisson and Negative Binomial models, up to 40 seedlings."
# Bars for the sample and lines with points for the two fitted models.
fitted_data_zip = rbind(
  data.frame(count = count_values_lily, probability = zip_probability, model = "Zero-inflated Poisson"),
  data.frame(count = count_values_lily, probability = nb_probability, model = "Negative Binomial")
)

ggplot() +
  geom_col(
    data = observed_data_lily,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data_zip,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data_zip,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(values = c("#D55E00", "grey20")) +
  coord_cartesian(xlim = c(-0.5, 40)) +
  labs(
    x = "Number of seedlings per quadrat",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: P(X = 0): observed 0.496, zero-inflated Poisson 0.496, Negative Binomial 0.483; P(X >= 20): observed 0.039, zero-inflated Poisson 0.0001, Negative Binomial 0.049 -->

# Exercise 4: Timber volume and body mass

<!-- PHASE 5: exercise statement -->

## Solution

### Question 1

```{r}
# Load the black cherry data from the datasets package and keep the timber volume (cubic feet).
data(trees, package = "datasets")
tree_volume = trees$Volume
n_trees = length(tree_volume)

# Negative log-likelihood of the Gamma model with shape k and rate r.
# par[1] is log(k) and par[2] is log(r). dgamma() returns a density, so these are
# log densities.
gamma_nll = function(par, x) {
  shape = exp(par[1])
  rate = exp(par[2])
  -sum(dgamma(x = x, shape = shape, rate = rate, log = TRUE))
}

# Method-of-moments starting values from E[X] = k / r and Var(X) = k / r^2,
# which give k = mean^2 / variance and r = mean / variance.
volume_mean = mean(tree_volume)
volume_variance = mean((tree_volume - volume_mean)^2)
start_volume = log(c(
  log_shape = volume_mean^2 / volume_variance,
  log_rate = volume_mean / volume_variance
))
start_volume

volume_fit = optim(
  par = start_volume,
  fn = gamma_nll,
  x = tree_volume,
  method = "BFGS",
  hessian = TRUE
)
volume_fit$convergence

# Estimates on the original scale and the mean of the fitted model, by invariance.
shape_volume = unname(exp(volume_fit$par["log_shape"]))
rate_volume = unname(exp(volume_fit$par["log_rate"]))
c(
  shape = shape_volume,
  rate_per_ft3 = rate_volume,
  mean_ft3 = shape_volume / rate_volume,
  sample_mean_ft3 = volume_mean,
  nll = volume_fit$value
)

# Method-of-moments estimate of the shape, for comparison.
exp(start_volume["log_shape"])
```

<!-- VALUES: convergence code 0; shape 3.886; rate 0.1288 per cubic foot; mean 30.17 cubic feet (equal to the sample mean); NLL 125.73; method-of-moments shape 3.48; the names in start_volume: check that they are log_shape and log_rate -->

### Question 2

```{r}
# Correlation between the estimates of log(k) and log(r), from the inverse Hessian.
cov2cor(solve(volume_fit$hessian))

# Grids of candidate values for the shape and the rate (per cubic foot).
shape_grid = seq(from = 1.5, to = 8, length.out = 71)
rate_grid = seq(from = 0.04, to = 0.3, length.out = 71)

# Fill the matrix of log-likelihoods with two nested loops.
loglik_surface_volume = matrix(NA, nrow = length(shape_grid), ncol = length(rate_grid))
for (i in seq_along(shape_grid)) {
  for (j in seq_along(rate_grid)) {
    loglik_surface_volume[i, j] = sum(
      dgamma(x = tree_volume, shape = shape_grid[i], rate = rate_grid[j], log = TRUE)
    )
  }
}
relative_surface_volume = loglik_surface_volume + volume_fit$value
```

```{r}
#| label: fig-volume_surface
#| fig-cap: "Contours of the relative log-likelihood of the tree volumes as a function of the shape and the rate of the Gamma model. The circle is the maximum likelihood estimate."
# Contours of the relative log-likelihood and the maximum likelihood estimate.
contour(
  x = shape_grid,
  y = rate_grid,
  z = relative_surface_volume,
  levels = c(-12, -9, -6, -4, -2, -1),
  xlab = "Shape (k)",
  ylab = "Rate (r, per cubic foot)"
)
points(x = shape_volume, y = rate_volume, pch = 16, col = "#D55E00")
```

<!-- VALUES: correlation between the estimates of log(k) and log(r): 0.937 (0.94); the contours are tilted: a larger shape goes with a larger rate; the outer contours are cut by the edges of the grids (shape 1.5 to 8, rate 0.04 to 0.30) -->

### Question 3

```{r}
# Cutoff for a 95% interval of one parameter: half the 95% quantile of a chi-squared
# distribution with one degree of freedom.
cutoff = qchisq(p = 0.95, df = 1) / 2

# Negative log-likelihood as a function of log(r) only, for a given shape.
gamma_nll_given_shape = function(log_rate, shape, x) {
  -sum(dgamma(x = x, shape = shape, rate = exp(log_rate), log = TRUE))
}

# Slice: the rate is fixed at its estimate. Profile: for each shape the rate is
# re-estimated with optim(), starting from the joint estimate.
shape_fine = seq(from = 1.5, to = 8, length.out = 261)
slice_loglik_volume = rep(NA, times = length(shape_fine))
profile_loglik_volume = rep(NA, times = length(shape_fine))
for (i in seq_along(shape_fine)) {
  slice_loglik_volume[i] = sum(
    dgamma(x = tree_volume, shape = shape_fine[i], rate = rate_volume, log = TRUE)
  )
  inner_fit = optim(
    par = volume_fit$par["log_rate"],
    fn = gamma_nll_given_shape,
    shape = shape_fine[i],
    x = tree_volume,
    method = "BFGS"
  )
  profile_loglik_volume[i] = -inner_fit$value
}

# Relative log-likelihoods and the range of shapes within the cutoff of the maximum.
slice_relative_volume = slice_loglik_volume + volume_fit$value
profile_relative_volume = profile_loglik_volume + volume_fit$value
range(shape_fine[slice_relative_volume >= -cutoff])
range(shape_fine[profile_relative_volume >= -cutoff])
```

```{r}
#| label: fig-volume_profile
#| fig-cap: "Slice (dashed line) and profile (solid line) of the relative log-likelihood for the shape of the Gamma model fitted to the tree volumes. The horizontal line is at minus half the 95% quantile of the chi-squared distribution with one degree of freedom."
# Relative log-likelihood along the slice and along the profile.
curve_volume = rbind(
  data.frame(shape = shape_fine, relative = slice_relative_volume, curve = "Slice"),
  data.frame(shape = shape_fine, relative = profile_relative_volume, curve = "Profile")
)
ggplot(curve_volume, aes(x = shape, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -cutoff, colour = "grey40") +
  coord_cartesian(ylim = c(-6, 0)) +
  labs(x = "Shape (k)", y = "Relative log-likelihood", linetype = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

<!-- VALUES: shape within the cutoff of the maximum: slice 3.26 to 4.56, profile 2.32 to 6.06 (grid of 261 values: the endpoints are accurate to about 0.025) -->

### Question 4

```{r}
# Load the penguin data (in R 4.5.0 and later) and keep the Adelie penguins.
data(penguins, package = "datasets")
adelie = penguins[penguins$species == "Adelie", ]

# One body mass is missing; remove it explicitly and keep the masses in grams.
sum(is.na(adelie$body_mass))
mass_g = adelie$body_mass[!is.na(adelie$body_mass)]
n_penguins = length(mass_g)

# The same masses in kilograms.
mass_kg = mass_g / 1000

# Fit the Gamma model to the masses in grams and in kilograms, with method-of-moments
# starting values calculated for each unit.
start_from_moments = function(x) {
  x_mean = mean(x)
  x_variance = mean((x - x_mean)^2)
  log(c(log_shape = x_mean^2 / x_variance, log_rate = x_mean / x_variance))
}

# The default stopping rule of optim() is relative to the value of the negative
# log-likelihood, which is large in grams, so the search could stop almost at the starting
# values. A smaller relative tolerance (reltol) makes both fits reach the maximum.
penguin_fit_g = optim(
  par = start_from_moments(mass_g),
  fn = gamma_nll,
  x = mass_g,
  method = "BFGS",
  hessian = TRUE,
  control = list(reltol = 1e-12)
)
penguin_fit_kg = optim(
  par = start_from_moments(mass_kg),
  fn = gamma_nll,
  x = mass_kg,
  method = "BFGS",
  hessian = TRUE,
  control = list(reltol = 1e-12)
)
c(grams = penguin_fit_g$convergence, kilograms = penguin_fit_kg$convergence)

# Estimates on the original scale: the shape, and the rate per gram and per kilogram.
shape_penguin = unname(exp(penguin_fit_g$par["log_shape"]))
rate_penguin_g = unname(exp(penguin_fit_g$par["log_rate"]))
rate_penguin_kg = unname(exp(penguin_fit_kg$par["log_rate"]))
c(
  n = n_penguins,
  shape_g = shape_penguin,
  shape_kg = unname(exp(penguin_fit_kg$par["log_shape"])),
  rate_per_g = rate_penguin_g,
  rate_per_kg = rate_penguin_kg
)

# Maximum log-likelihood in each unit and the change when the unit changes. A density is
# per unit of the measurement, so the data in kilograms have densities 1,000 times larger.
loglik_g = -penguin_fit_g$value
loglik_kg = -penguin_fit_kg$value
c(loglik_g = loglik_g, loglik_kg = loglik_kg, difference = loglik_kg - loglik_g)
n_penguins * log(1000)
```

<!-- VALUES: 1 missing body mass removed; 151 Adelie penguins; convergence codes 0; shape 66.09 in both units; rate 0.01786 per gram and 17.86 per kilogram; log-likelihood -1137.72 in grams and -94.65 in kilograms; difference 1043.07 = 151 * log(1000) -->

### Question 5

```{r}
# Negative log-likelihood of the LogNormal model; par[1] is meanlog and par[2] is log(sdlog).
lognormal_nll = function(par, x) {
  -sum(dlnorm(x = x, meanlog = par[1], sdlog = exp(par[2]), log = TRUE))
}

# Starting values: mean and standard deviation of the logarithms of the masses.
start_lognormal = function(x) {
  c(meanlog = mean(log(x)), log_sdlog = log(sd(log(x))))
}
lognormal_fit_g = optim(
  par = start_lognormal(mass_g),
  fn = lognormal_nll,
  x = mass_g,
  method = "BFGS",
  control = list(reltol = 1e-12)
)
lognormal_fit_kg = optim(
  par = start_lognormal(mass_kg),
  fn = lognormal_nll,
  x = mass_kg,
  method = "BFGS",
  control = list(reltol = 1e-12)
)
c(grams = lognormal_fit_g$convergence, kilograms = lognormal_fit_kg$convergence)

# AIC = 2K - 2 * log-likelihood; both models have K = 2 parameters. The differences in
# AIC are the same in grams and in kilograms.
aic_penguin = data.frame(
  unit = c("grams", "kilograms"),
  gamma = 2 * 2 + 2 * c(penguin_fit_g$value, penguin_fit_kg$value),
  lognormal = 2 * 2 + 2 * c(lognormal_fit_g$value, lognormal_fit_kg$value)
)
aic_penguin$difference = aic_penguin$gamma - aic_penguin$lognormal
aic_penguin
```

<!-- VALUES: convergence codes 0; NLL in grams: Gamma 1137.72, LogNormal 1137.48; in kilograms: Gamma 94.65, LogNormal 94.41; difference in AIC (Gamma minus LogNormal) about 0.5 in both units -->

# Exercise 5: Estimators and intervals across hypothetical experiments

<!-- PHASE 5: exercise statement -->

## Solution

### Question 1

```{r}
# Wald interval for the shape of the tree volumes: calculate on the log scale with the
# standard error from the inverse Hessian, and back-transform the endpoints.
critical_value = qnorm(p = 0.975)
se_log_shape_volume = sqrt(solve(volume_fit$hessian)[1, 1])
wald_shape_volume = exp(volume_fit$par["log_shape"] + c(-1, 1) * critical_value * se_log_shape_volume)

# Profile log-likelihood of the shape relative to the maximum, with the rate re-estimated.
gamma_profile_relative = function(shape, x, fit) {
  inner_fit = optim(
    par = fit$par["log_rate"],
    fn = gamma_nll_given_shape,
    shape = shape,
    x = x,
    method = "BFGS"
  )
  fit$value - inner_fit$value
}

# uniroot() finds where the relative profile equals -cutoff. The ends of the search
# intervals were chosen from the profile plotted in Exercise 4: below and above the
# estimate, where the profile is under -cutoff.
shape_lower_volume = uniroot(
  f = function(shape) gamma_profile_relative(shape, x = tree_volume, fit = volume_fit) + cutoff,
  interval = c(1, shape_volume),
  tol = 1e-8
)$root
shape_upper_volume = uniroot(
  f = function(shape) gamma_profile_relative(shape, x = tree_volume, fit = volume_fit) + cutoff,
  interval = c(shape_volume, 15),
  tol = 1e-8
)$root

rbind(
  wald = unname(wald_shape_volume),
  profile = c(shape_lower_volume, shape_upper_volume)
)
```

<!-- VALUES: Wald interval for the shape 2.41 to 6.27; profile interval 2.32 to 6.06 (no unit) -->

### Question 2

```{r}
# One simulation of hypothetical experiments from a Gamma model whose parameters we know.
# For each experiment: the maximum likelihood and the method-of-moments estimates of the
# shape, and whether the Wald interval (log scale) and the profile interval contain the
# shape that was used to simulate.
simulate_gamma_experiments = function(n_obs, shape, rate, n_experiments) {
  results = data.frame(
    shape_ml = rep(NA, times = n_experiments),
    shape_mom = rep(NA, times = n_experiments),
    wald_covered = rep(NA, times = n_experiments),
    profile_covered = rep(NA, times = n_experiments),
    converged = rep(NA, times = n_experiments)
  )
  for (e in seq_len(n_experiments)) {
    # One hypothetical experiment and its method-of-moments estimates.
    x = rgamma(n = n_obs, shape = shape, rate = rate)
    x_mean = mean(x)
    x_variance = mean((x - x_mean)^2)
    shape_mom = x_mean^2 / x_variance

    # Maximum likelihood fit, started at the method-of-moments estimates. The small
    # relative tolerance keeps the search from stopping too early when the negative
    # log-likelihood is large (the penguin masses are in grams).
    fit = optim(
      par = log(c(shape_mom, x_mean / x_variance)),
      fn = gamma_nll,
      x = x,
      method = "BFGS",
      hessian = TRUE,
      control = list(reltol = 1e-12)
    )
    shape_ml = exp(fit$par[1])
    se_log_shape = sqrt(solve(fit$hessian)[1, 1])

    # The Wald interval contains the shape when the estimate is within the critical value
    # of standard errors on the log scale. The profile interval contains it when the
    # relative profile log-likelihood at that shape is within the cutoff.
    wald_covered = abs(log(shape) - fit$par[1]) <= critical_value * se_log_shape
    inner_fit = optim(
      par = fit$par[2],
      fn = gamma_nll_given_shape,
      shape = shape,
      x = x,
      method = "BFGS",
      control = list(reltol = 1e-12)
    )
    profile_covered = fit$value - inner_fit$value >= -cutoff

    results$shape_ml[e] = shape_ml
    results$shape_mom[e] = shape_mom
    results$wald_covered[e] = wald_covered
    results$profile_covered[e] = profile_covered
    results$converged[e] = fit$convergence == 0 && inner_fit$convergence == 0
  }
  results
}

# Mean, standard deviation and mean squared error of both estimators of the shape,
# coverage of both intervals, and the number of experiments in which a fit failed.
summarise_experiments = function(results, shape) {
  data.frame(
    mean_ml = mean(results$shape_ml),
    sd_ml = sd(results$shape_ml),
    mse_ml = mean((results$shape_ml - shape)^2),
    mean_mom = mean(results$shape_mom),
    sd_mom = sd(results$shape_mom),
    mse_mom = mean((results$shape_mom - shape)^2),
    coverage_wald = mean(results$wald_covered),
    coverage_profile = mean(results$profile_covered),
    failed_fits = sum(!results$converged)
  )
}

# 2,000 hypothetical experiments of 31 trees from the Gamma model fitted to the volumes.
set.seed(seed = 801)
n_experiments = 2000
trees_31 = simulate_gamma_experiments(
  n_obs = n_trees,
  shape = shape_volume,
  rate = rate_volume,
  n_experiments = n_experiments
)
summarise_experiments(results = trees_31, shape = shape_volume)
```

<!-- VALUES (2,000 experiments, seed 801; time about 3 s): shape used in the simulation 3.886; ML mean 4.33, SD 1.20, MSE 1.63; method of moments mean 4.43, SD 1.30, MSE 1.99; no failed fits. The SD of both estimators changes with the seed because the sampling distribution is skewed: with 1,000 experiments and four other seeds the ML SD ranged from 1.09 to 1.22 (Appendix A: 4.26 and 1.09 for ML, 4.37 and 1.22 for the method of moments) -->

### Question 3

The coverage columns of the table in question 2 answer this question. The hint is already
used in the code of question 2: no `uniroot()` call is needed inside the simulation.

```{r}
# Coverage of the two intervals, from the experiments of question 2.
c(wald = mean(trees_31$wald_covered), profile = mean(trees_31$profile_covered))
```

<!-- VALUES: coverage about 0.93 for the Wald interval and 0.94 for the profile interval; simulation error of a coverage near 0.94 with 1,000 experiments: about 0.0075 -->

### Question 4

```{r}
# 2,000 hypothetical experiments of 151 penguins from the Gamma model fitted in Exercise 4
# (in grams), and of 151 trees from the model fitted to the tree volumes, to separate the
# effect of the sample size from the effect of the model.
set.seed(seed = 802)
penguins_151 = simulate_gamma_experiments(
  n_obs = n_penguins,
  shape = shape_penguin,
  rate = rate_penguin_g,
  n_experiments = n_experiments
)
trees_151 = simulate_gamma_experiments(
  n_obs = n_penguins,
  shape = shape_volume,
  rate = rate_volume,
  n_experiments = n_experiments
)

# Summary of the three simulations.
rbind(
  "31 trees" = summarise_experiments(results = trees_31, shape = shape_volume),
  "151 penguins" = summarise_experiments(results = penguins_151, shape = shape_penguin),
  "151 trees" = summarise_experiments(results = trees_151, shape = shape_volume)
)

# Relative bias of the maximum likelihood estimator of the shape in each simulation.
c(
  trees_31 = mean(trees_31$shape_ml) / shape_volume - 1,
  penguins_151 = mean(penguins_151$shape_ml) / shape_penguin - 1,
  trees_151 = mean(trees_151$shape_ml) / shape_volume - 1
)
```

<!-- VALUES (2,000 experiments each, seed 802; the chunk takes about 19 s): 151 penguins (shape 66.09): ML mean 67.17 (relative bias 1.6%), SD 7.91, MSE 63.7; method of moments mean 67.22, SD 7.97, MSE 64.8; coverage 0.947 for Wald and 0.946 for profile. 151 trees (shape 3.886): ML mean 3.945 (relative bias 1.5%), SD 0.442, MSE 0.199; method of moments mean 3.960, SD 0.507, MSE 0.263; coverage 0.951 and 0.956. 31 trees: relative bias 11.5%. No failed fits -->

### Question 5

<!-- NO CODE: reasoning only -->
