---
title: "Chapter 4: Estimation and sampling distributions — solutions"
date: today
---

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions use the
same models, tasks, and seeds as Exercises 2–4 on that page.

# Exercise 2: Breeding success at bird nests

One observational unit is a monitored nest. Under the assumed
binomial model, the expected proportion of successes is $p$, so
matching it to the observed proportion gives the method-of-moments
estimator $\hat p=X/n$.

```{r}
set.seed(seed = 422)
p_true = 0.6
n_nests = 40
successful_nests = rbinom(n = 1, size = n_nests, prob = p_true)
p_hat = successful_nests / n_nests
c(successful_nests = successful_nests, estimate = p_hat)
```

In this simulated study, 25 of 40 nests succeed, giving an estimated
success probability of 0.625. It need not equal the model probability of 0.6.
We know the latter only because we generated the data.

For the sampling distributions, each call to `rbinom()` generates
2,000 possible study totals. Dividing by the number of nests in
that study produces one estimated success probability per study.

```{r}
set.seed(seed = 423)
n_studies = 2000
n_small = 20
n_large = 100

p_estimates_small = rbinom(
  n = n_studies, size = n_small, prob = p_true
) / n_small
p_estimates_large = rbinom(
  n = n_studies, size = n_large, prob = p_true
) / n_large

old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))
hist(
  x = p_estimates_small, breaks = 20, probability = TRUE,
  xlim = c(0, 1), col = "grey85", border = "white",
  xlab = "Estimated nest success probability", ylab = "Density",
  main = "20 nests"
)
abline(v = p_true, lty = 2, lwd = 2)
hist(
  x = p_estimates_large, breaks = 20, probability = TRUE,
  xlim = c(0, 1), col = "grey85", border = "white",
  xlab = "Estimated nest success probability", ylab = "Density",
  main = "100 nests"
)
abline(v = p_true, lty = 2, lwd = 2)
par(old_par)
```

The distribution for 100 nests is more concentrated around 0.6.
The estimates are still variable and discrete because the number
of successful nests is an integer.

For independent nests with the same $p$, the estimator is unbiased
and has variance $p(1-p)/n$. Simulated bias is the average estimate
minus the known model value. Mean squared error (MSE) is the average
squared distance from that value.

```{r}
nest_summary = data.frame(
  n_nests = c(n_small, n_large),
  average_estimate = c(
    mean(p_estimates_small), mean(p_estimates_large)
  ),
  simulated_bias = c(
    mean(p_estimates_small) - p_true,
    mean(p_estimates_large) - p_true
  ),
  simulated_variance = c(
    var(p_estimates_small), var(p_estimates_large)
  ),
  simulated_se = c(
    sd(p_estimates_small), sd(p_estimates_large)
  ),
  simulated_mse = c(
    mean((p_estimates_small - p_true)^2),
    mean((p_estimates_large - p_true)^2)
  ),
  model_variance = p_true * (1 - p_true) /
    c(n_small, n_large)
)
round(nest_summary, digits = 4)
```

The average estimates are approximately 0.602 and 0.601, close to
the true 0.6. Simulated variance falls from about 0.0121 for 20
nests to 0.0023 for 100 nests; standard error and MSE fall too.
Their simulated values do not exactly equal the model values.
Bias concerns the estimator's average across possible studies; a
single estimate cannot reveal it.
Neighbouring nests could share weather, food supply, or predation
conditions, making their outcomes dependent.

# Exercise 3: Variation in tree heights

The first method-of-moments estimate matches the sample mean to
$E[X]=\mu$. The second averages squared deviations from that sample
mean with divisor $n$. R's `var()` uses $n-1$ instead.

```{r}
set.seed(seed = 433)
n_trees = 25
true_mu = 22
true_sigma = 4
true_variance = true_sigma^2
tree_heights = rnorm(
  n = n_trees, mean = true_mu, sd = true_sigma
)

mu_hat = mean(tree_heights)
variance_mm = mean((tree_heights - mu_hat)^2)
variance_with_var = var(tree_heights)
c(
  estimated_mean_m = mu_hat,
  variance_mm_m2 = variance_mm,
  variance_with_var_m2 = variance_with_var
)
```

The estimated mean is in metres; both variance estimates are in
square metres. They differ because `var()` adjusts the divisor.
Neither single-sample estimate needs to equal the model variance
of $4^2=16$ square metres.

To study bias, repeat the entire 25-tree study. Each repetition
returns a pair of variance estimates.

```{r}
set.seed(seed = 434)
n_studies = 2000
variance_estimates = replicate(
  n = n_studies,
  expr = {
    heights = rnorm(
      n = n_trees, mean = true_mu, sd = true_sigma
    )
    sample_mean = mean(heights)
    c(
      method_of_moments = mean((heights - sample_mean)^2),
      with_var = var(heights)
    )
  }
)

average_variance = rowMeans(variance_estimates)
simulated_bias = average_variance - true_variance
simulated_mse = rowMeans(
  (variance_estimates - true_variance)^2
)

variance_summary = data.frame(
  estimator = rownames(variance_estimates),
  average_variance_m2 = average_variance,
  simulated_bias_m2 = simulated_bias,
  simulated_mse_m4 = simulated_mse,
  row.names = NULL
)
variance_summary
```

The method-of-moments estimate averages about 15.14 square metres,
below the true 16. Its simulated bias is about $-0.86$ square
metres; its theoretical bias is $-16/25=-0.64$ square metres.
The estimate from `var()` averages about 15.77, with simulated
bias about $-0.23$ rather than exactly zero. The differences from
theoretical values reflect finite simulation.

MSE includes both variance across studies and squared bias. In
this simulation, the method-of-moments estimator has MSE about
$19.84\ \mathrm{m}^4$, compared with about $20.77\ \mathrm{m}^4$
for `var()`. Knowing that one estimator is unbiased does not,
by itself, settle which has smaller MSE.

# Exercise 4: Mean body mass and interval coverage

The lognormal parameters describe **log mass**. On the original
gram scale, the median is $e^\mu$ and the mean is
$e^{\mu+\sigma^2/2}$. Right skew makes the mean larger than the
median.

```{r}
mass_meanlog = log(30)
mass_sdlog = 0.8
true_median = exp(mass_meanlog)
true_mean = exp(mass_meanlog + mass_sdlog^2 / 2)
c(median_g = true_median, mean_g = true_mean)
```

The model median is 30 g and the model mean is approximately
41.3 g. Simulate one study and calculate the standard error of
its original-scale sample mean:

```{r}
set.seed(seed = 443)
n_animals = 12
body_mass = rlnorm(
  n = n_animals,
  meanlog = mass_meanlog, sdlog = mass_sdlog
)

sample_mean = mean(body_mass)
sample_median = median(body_mass)
estimated_se = sd(body_mass) / sqrt(n_animals)
c(
  sample_mean_g = sample_mean,
  sample_median_g = sample_median,
  estimated_se_g = estimated_se
)
```

This standard error describes uncertainty in the sample mean, not
variation among the 12 individual masses. The $t$ interval uses
the estimated standard error and $n-1$ degrees of freedom:

```{r}
critical_value = qt(p = 0.975, df = n_animals - 1)
mass_interval = c(
  lower = sample_mean - critical_value * estimated_se,
  upper = sample_mean + critical_value * estimated_se
)
mass_interval
```

The interval runs from approximately 14.1 to 54.6 g. Its endpoints
are candidate values for the **population mean mass in grams**,
not an interval containing 95% of animal masses.
This $t$ interval has exact 95% coverage for independent normal
observations. Lognormal masses are skewed, so here it is an
approximation; its actual coverage must be checked.

For each simulated study below, calculate a new interval and
record whether it contains the known model mean. The proportion
of `TRUE` results estimates coverage.

```{r}
coverage_for_size = function(n_animals, n_studies = 2000) {
  critical_value = qt(p = 0.975, df = n_animals - 1)
  covered = replicate(
    n = n_studies,
    expr = {
      mass = rlnorm(
        n = n_animals,
        meanlog = mass_meanlog, sdlog = mass_sdlog
      )
      estimate = mean(mass)
      se = sd(mass) / sqrt(n_animals)
      lower = estimate - critical_value * se
      upper = estimate + critical_value * se
      lower <= true_mean && true_mean <= upper
    }
  )
  mean(covered)
}

set.seed(seed = 444)
simulated_coverage = c(
  n_12 = coverage_for_size(n_animals = 12),
  n_100 = coverage_for_size(n_animals = 100)
)
simulated_coverage
```

The simulated coverage is 0.889 for 12 animals and 0.927 for 100
animals, both below 0.95. These fractions depend on the seed and
2,000 repetitions. With the more skewed sampling distribution at
12 animals, the usual $t$ interval misses the true mean more often
than its nominal 5% rate. At 100 animals, the sample mean tends
to have a more nearly normal sampling distribution, making this
approximation more plausible. These results describe performance
**under the specified lognormal model and independent sampling**.
Actual body masses or a field sampling design could depart from
both assumptions, so simulated coverage alone does not establish
coverage for a real study.
