---
title: "Chapter 4: Maximum likelihood estimation — solutions"
date: today
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures on this page.
library(ggplot2)
```

The fully worked first exercise is on the
[student practical page](no_solution.qmd). These solutions use the
same data, tasks and seeds as Exercises 2–5 on that page. Exercises 1 to 4 are the core of
the practical and Exercise 5 is extra practice.

# Exercise 2: Glacier lily seedlings in quadrats

The data set `Lily_sum` in the package `emdbook` contains counts of the glacier lily
*Erythronium grandiflorum* in a grid of 16 × 16 quadrats of 2 m × 2 m in Washington Gulch,
sampled in 1992 (Thomson et al. 1996). We use the column `seedlings`, the number of
seedlings in each of the 256 quadrats. Load the data with
`data(Lily_sum, package = "emdbook")`.

Let $X_i$ be the number of seedlings in quadrat $i$. We use a
[Negative Binomial](../../Supplements/distributions.qmd#sec-negative_binomial) model with
mean $\mu$ and shape $k$,

$$
X_i \sim \operatorname{NegBin}(\mu, k), \qquad \operatorname{Var}(X_i)=\mu+\frac{\mu^2}{k},
$$

and assume that the quadrats are i.i.d. Neighbouring quadrats of a grid are likely to
resemble each other, so this is a simplification, as it was for the visits to owl nests in
the Chapter 3 practical (models for observations that are not independent start in
[Chapter 7](../../Chapter_7/Theory.qmd)).

1. Estimate $\mu$ and $k$ with the method of moments, as in the Chapter 3 practical. Also
   calculate the fraction of quadrats without seedlings and the largest count.
2. Write the NLL of the model with $\log\mu$ and $\log k$ as parameters and minimise it
   with `optim()`, using the method-of-moments estimates as starting values and
   `hessian = TRUE`. Report the estimates on the original scale. Compare the log-likelihood
   at the maximum likelihood estimates with the log-likelihood at the method-of-moments
   estimates.
3. Calculate the log-likelihood on a grid of values of $\mu$ and $k$ with two nested `for`
   loops and draw the contours of the relative log-likelihood with `contour()`. Add both
   pairs of estimates and the contour of the approximate joint 95% confidence region. Is
   the method-of-moments estimate inside the region?
4. Calculate a 95% Wald interval for $k$ (on the log scale, with the endpoints transformed
   back) and a 95% profile likelihood interval for $k$ with `uniroot()`. Compare them.
5. Plot the relative frequencies of the counts together with the probabilities of the
   fitted Negative Binomial model and of a Poisson model. Compare the fraction of empty
   quadrats under each model with the observed fraction, and compare the two models with
   AIC and with its small-sample correction AICc (see
   [Model comparison with AIC](../Theory.qmd#sec-aic)).

```{r}
# Load the glacier lily data from the emdbook package without attaching the package.
data(Lily_sum, package = "emdbook")

# Number of seedlings in each 2 m by 2 m quadrat, and the number of quadrats.
lily_counts = Lily_sum$seedlings
n_lily = length(lily_counts)
```

## Solution

### Question 1

The observational unit is one quadrat. As in the Chapter 3 practical, the method of moments
equates the mean and the variance of the model with those of the sample, which gives
$\hat\mu=\overline x$ and $\hat k_{\mathrm{MM}}=\hat\mu^2/(\hat\sigma^2-\hat\mu)$, where
$\hat\sigma^2$ is the variance of the sample with divisor $n$:

```{r}
# Method-of-moments estimates, as in the practical of Chapter 3: the mean, the variance
# with divisor n, and the shape k from variance = mu + mu^2 / k.
mu_mom_lily = mean(lily_counts)
variance_lily = mean((lily_counts - mu_mom_lily)^2)
k_mom_lily = mu_mom_lily^2 / (variance_lily - mu_mom_lily)

c(
  n_quadrats = n_lily,
  mean = mu_mom_lily,
  variance = variance_lily,
  k_mom = k_mom_lily,
  fraction_empty = mean(lily_counts == 0),
  maximum = max(lily_counts)
)
```

There are 256 quadrats, with a mean of 3.96 seedlings per quadrat and a variance of 54.3,
almost 14 times the mean. The method-of-moments estimate of the shape is therefore small,
$\hat k_{\mathrm{MM}}\approx 0.31$. Half of the quadrats have no seedlings (127 of 256, a
fraction of 0.496), while the largest count is 72 seedlings.

### Question 2

The NLL is the one of the theory. `optim()` passes both parameters to the function in one
vector, and inside the function `par[1]` is $\log\mu$ and `par[2]` is $\log k$, so that
both parameters stay positive. We start the search at the method-of-moments estimates:

```{r}
# Negative log-likelihood of the Negative Binomial model with mean mu and shape k.
# par[1] is log(mu) and par[2] is log(k), so that both parameters stay positive.
nb_nll = function(par, counts) {
  mu = exp(par[1])
  k = exp(par[2])
  -sum(dnbinom(x = counts, mu = mu, size = k, log = TRUE))
}

# Start from the method-of-moments estimates, on the log scale.
start_lily = c(log_mu = log(mu_mom_lily), log_k = log(k_mom_lily))

# Fit with BFGS; hessian = TRUE is used for the standard errors in question 4.
lily_fit = optim(
  par = start_lily,
  fn = nb_nll,
  counts = lily_counts,
  method = "BFGS",
  hessian = TRUE
)
lily_fit$convergence

# Estimates on the original scale.
mu_mle_lily = unname(exp(lily_fit$par["log_mu"]))
k_mle_lily = unname(exp(lily_fit$par["log_k"]))
c(mu_mle = mu_mle_lily, k_mle = k_mle_lily)

# Log-likelihood at the maximum likelihood estimates and at the method-of-moments estimates.
loglik_mle_lily = -lily_fit$value
loglik_mom_lily = -nb_nll(par = start_lily, counts = lily_counts)
c(mle = loglik_mle_lily, mom = loglik_mom_lily)
```

The convergence code is 0. The maximum likelihood estimates are $\hat\mu\approx 3.96$
seedlings per quadrat, which is the mean of the sample (as for any Negative Binomial
model), and $\hat k\approx 0.26$, somewhat smaller than the method-of-moments estimate of
0.31. The log-likelihood is $-564.2$ at the maximum likelihood estimates and $-565.3$ at
the method-of-moments estimates. The difference of 1.1 units means that the observed counts
are about three times more probable ($e^{1.1}$) under the former. The two methods disagree
much less here than for the owls of the theory, where the maximum likelihood estimate of
the shape was 0.33 and the method-of-moments estimate was 0.60.

### Question 3

We choose grids that contain both pairs of estimates with some margin, and fill a matrix
that has one row per value of $\mu$ and one column per value of $k$. The cutoff for a
joint region of two parameters is half the 0.95 quantile of a chi-squared distribution
with two degrees of freedom (see
[Uncertainty from the likelihood](../Theory.qmd#sec-likelihood-uncertainty)):

```{r}
# Grids of candidate values for the mean (seedlings per quadrat) and the shape.
mu_grid_lily = seq(from = 2.5, to = 6, length.out = 71)
k_grid_lily = seq(from = 0.17, to = 0.42, length.out = 71)

# Fill a matrix with two nested loops: row i is the i-th mean and column j the j-th shape.
loglik_surface_lily = matrix(NA, nrow = length(mu_grid_lily), ncol = length(k_grid_lily))
for (i in seq_along(mu_grid_lily)) {
  for (j in seq_along(k_grid_lily)) {
    loglik_surface_lily[i, j] = sum(
      dnbinom(x = lily_counts, mu = mu_grid_lily[i], size = k_grid_lily[j], log = TRUE)
    )
  }
}

# Relative log-likelihood: the difference from the maximum found by optim().
relative_surface_lily = loglik_surface_lily - loglik_mle_lily

# Cutoff of the relative log-likelihood for a joint 95% region of two parameters:
# half the 95% quantile of a chi-squared distribution with two degrees of freedom.
joint_cutoff = qchisq(p = 0.95, df = 2) / 2
joint_cutoff
```

The function `contour()` draws the matrix, as in the Chapter 1 practical. A second call
with `add = TRUE` draws the contour of the joint region on top of the first plot, and
`points()` adds the two pairs of estimates:

```{r}
#| label: fig-lily_surface
#| fig-cap: "Contours of the relative log-likelihood of the glacier lily seedling counts as a function of the mean and the shape of the Negative Binomial model. The circle is the maximum likelihood estimate, the triangle is the method-of-moments estimate and the thick contour is the approximate joint 95% confidence region."
# Contours of the relative log-likelihood with base R, then the joint region on top.
contour(
  x = mu_grid_lily,
  y = k_grid_lily,
  z = relative_surface_lily,
  levels = c(-12, -9, -6, -4, -2, -1),
  xlab = "Mean (seedlings per quadrat)",
  ylab = "Shape (k)"
)
contour(
  x = mu_grid_lily,
  y = k_grid_lily,
  z = relative_surface_lily,
  levels = -joint_cutoff,
  lwd = 2,
  col = "#D55E00",
  add = TRUE,
  drawlabels = FALSE
)

# Maximum likelihood estimate (circle) and method-of-moments estimate (triangle).
points(x = mu_mle_lily, y = k_mle_lily, pch = 16, col = "#D55E00")
points(x = mu_mom_lily, y = k_mom_lily, pch = 17)
```

The cutoff is 3.00. The thick contour in @fig-lily_surface joins the combinations of $\mu$
and $k$ whose log-likelihood is 3.00 units below the maximum, and the area inside it is the
approximate joint 95% confidence region. It extends from about 2.9 to 5.5 seedlings per
quadrat in $\mu$ and from about 0.20 to 0.35 in $k$. As for the owls in the theory, the
contours are not tilted: with the mean as a parameter, the estimates of $\mu$ and $k$
hardly depend on each other. The method-of-moments estimate lies inside the region, near
its upper edge in $k$ (its relative log-likelihood is $-1.1$, as we saw in question 2). So
for these data the method-of-moments estimates are among the combinations of parameter
values that the likelihood supports.

### Question 4

For the Wald interval we invert the Hessian with `solve()` and take the square roots of its
diagonal, which are the standard errors of the estimates of $\log\mu$ and $\log k$. We
calculate the interval for $\log k$ and transform its endpoints with `exp()`:

```{r}
# 0.975 quantile of the standard Normal distribution and the cutoff for one parameter.
critical_value = qnorm(p = 0.975)
cutoff = qchisq(p = 0.95, df = 1) / 2

# Wald interval for k: the inverse Hessian approximates the covariance matrix of the
# estimates of log(mu) and log(k). Calculate on the log scale and back-transform the endpoints.
lily_se = sqrt(diag(solve(lily_fit$hessian)))
lily_se
wald_k_lily = exp(lily_fit$par["log_k"] + c(-1, 1) * critical_value * lily_se["log_k"])
wald_k_lily
```

The standard errors are 0.126 for the estimate of $\log\mu$ and 0.117 for the estimate of
$\log k$, and the Wald interval for $k$ goes from 0.208 to 0.329.

For the profile likelihood interval we write a function that returns the relative profile
log-likelihood for a given value of $k$. Inside it, `optim()` re-estimates $\log\mu$ with
$k$ held fixed, starting from the joint estimate. Then `uniroot()` finds the two values of
$k$ at which this function equals $-1.92$. We search between 0.05 and the estimate and
between the estimate and 1, which are values far outside the region of high likelihood in
@fig-lily_surface:

```{r}
# Profile log-likelihood of k relative to the maximum: for each k the mean is
# re-estimated with optim() on log(mu), starting from the joint estimate.
nll_given_k_lily = function(log_mu, k, counts) {
  -sum(dnbinom(x = counts, mu = exp(log_mu), size = k, log = TRUE))
}
profile_relative_k_lily = function(k, counts) {
  inner_fit = optim(
    par = lily_fit$par["log_mu"],
    fn = nll_given_k_lily,
    k = k,
    counts = counts,
    method = "BFGS"
  )
  -inner_fit$value - loglik_mle_lily
}

# The profile is zero at the joint estimate.
profile_relative_k_lily(k = k_mle_lily, counts = lily_counts)

# uniroot() finds where the relative profile equals -cutoff, below and above the estimate.
# The ends of each search interval were chosen from the surface above.
k_lower_lily = uniroot(
  f = function(k) profile_relative_k_lily(k, counts = lily_counts) + cutoff,
  interval = c(0.05, k_mle_lily),
  tol = 1e-8
)$root
k_upper_lily = uniroot(
  f = function(k) profile_relative_k_lily(k, counts = lily_counts) + cutoff,
  interval = c(k_mle_lily, 1),
  tol = 1e-8
)$root

# Profile interval next to the Wald interval for the shape.
rbind(
  profile = c(lower = k_lower_lily, upper = k_upper_lily),
  wald = c(lower = wald_k_lily[[1]], upper = wald_k_lily[[2]])
)
```

The profile is zero at the joint estimate (up to the precision of the optimiser), as it
should be. The profile likelihood interval for $k$ goes from 0.207 to 0.328 and the Wald
interval from 0.208 to 0.329, so they agree to two decimals. With 256 quadrats the profile
log-likelihood is close to a parabola on the log scale of $k$, and the fast approximation
gives the same answer as the reference method.

### Question 5

We calculate the probability of every count, from zero to the largest one in the sample,
under both models. For the Poisson model the MLE of $\lambda$ is the mean of the sample, so
no optimisation is needed. The AIC is $2K-2\ell(\hat\theta)$, with $K=1$ for the Poisson
model and $K=2$ for the Negative Binomial model:

```{r}
# Fitted probabilities of every count under the Negative Binomial and Poisson models,
# and the relative frequencies in the sample.
count_values_lily = 0:max(lily_counts)
observed_lily = tabulate(lily_counts + 1, nbins = length(count_values_lily)) / n_lily
nb_probability = dnbinom(x = count_values_lily, mu = mu_mle_lily, size = k_mle_lily)
poisson_probability = dpois(x = count_values_lily, lambda = mu_mom_lily)

# Fraction of empty quadrats: observed and under each model.
c(
  observed = mean(lily_counts == 0),
  negative_binomial = nb_probability[1],
  poisson = poisson_probability[1]
)

# Log-likelihood of the Poisson model at its maximum (the sample mean) and AIC of both
# models. AIC = 2K - 2 * log-likelihood, with K the number of estimated parameters.
loglik_poisson_lily = sum(dpois(x = lily_counts, lambda = mu_mom_lily, log = TRUE))
aic_lily = c(
  poisson = 2 * 1 - 2 * loglik_poisson_lily,
  negative_binomial = 2 * 2 - 2 * loglik_mle_lily
)
aic_lily

# AICc adds 2K(K + 1) / (n - K - 1) to the AIC, with n the number of quadrats.
k_lily = c(poisson = 1, negative_binomial = 2)
aicc_lily = aic_lily + 2 * k_lily * (k_lily + 1) / (n_lily - k_lily - 1)
aicc_lily
```

```{r}
#| label: fig-lily_fits
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of seedlings per quadrat (bars) and probabilities under the fitted Negative Binomial and Poisson models, up to 40 seedlings."
# Data frames for the sample (bars) and the models (lines and points).
observed_data_lily = data.frame(count = count_values_lily, frequency = observed_lily)
fitted_data_lily = rbind(
  data.frame(count = count_values_lily, probability = nb_probability, model = "Negative Binomial"),
  data.frame(count = count_values_lily, probability = poisson_probability, model = "Poisson")
)

ggplot() +
  geom_col(
    data = observed_data_lily,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data_lily,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data_lily,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(values = c("#D55E00", "grey20")) +
  coord_cartesian(xlim = c(-0.5, 40)) +
  labs(
    x = "Number of seedlings per quadrat",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

The observed fraction of empty quadrats is 0.496. The fitted Negative Binomial model gives
a probability of 0.483 to an empty quadrat and the Poisson model only 0.019.
@fig-lily_fits shows the same for the other counts (the horizontal axis ends at 40
seedlings, which leaves out the quadrat with 72). The Poisson distribution concentrates its
probability around its mean of 4 seedlings, whereas the Negative Binomial distribution
follows both the many empty quadrats and the long tail of large counts.

The AIC is 2731.8 for the Poisson model and 1132.5 for the Negative Binomial model. The
difference of 1,599 units is very large compared with the two units that the additional
parameter costs, so the Negative Binomial model is strongly preferred. AIC ranks the
candidates that we give it. Whether there is a better description of the empty quadrats
than the Negative Binomial model is the question of Exercise 3.

The AICc adds 0.016 units to the AIC of the Poisson model and 0.047 units to that of the
Negative Binomial model, so it is 2731.8 and 1132.5 (to one decimal) and the comparison is
the same. The correction is negligible because 256 quadrats are many more than one or two
parameters (the ratio $n/K$ is far above the value of about 40 below which AICc is
usually recommended).

# Exercise 3: Empty quadrats and zero inflation

About half of the quadrats of Exercise 2 have no seedlings. The Negative Binomial model
describes this with a small shape $k$. A different explanation is that a fraction of the
quadrats cannot have seedlings at all (for example, because no seeds arrive there), while
the counts in the rest follow a Poisson distribution. This is a zero-inflated Poisson
model (see
[How to make zero-inflated distributions](../../Supplements/distributions.qmd#sec-zero_inflated)
in the supplement). Its probability mass function is

$$
P(X=x\mid\lambda,p_0)=
\begin{cases}
p_0+(1-p_0)\,e^{-\lambda}, & x=0,\\[4pt]
(1-p_0)\,\dfrac{e^{-\lambda}\lambda^{x}}{x!}, & x>0,
\end{cases}
$$

where $p_0$ (`pzero` in the code of the supplement) is the probability that a quadrat
cannot have seedlings and $\lambda$ is the mean number of seedlings in the quadrats that
can have them. Use the same counts as in Exercise 2 and the same assumption of i.i.d.
quadrats.

1. Write a function `dzipois(x, lambda, pzero, log = FALSE)` that returns the probability
   mass of the zero-inflated Poisson distribution, or its logarithm when `log = TRUE`.
   Check that the probabilities add up to one over a wide range of counts, for two
   combinations of parameter values.
2. Write the NLL with $\log\lambda$ and $\operatorname{logit}p_0$ as parameters. Choose
   starting values from the data (think of the fraction of empty quadrats and of the mean
   of the quadrats that are not empty) and fit the model with `optim()`. Interpret both
   estimates.
3. Compare the zero-inflated Poisson model with the Negative Binomial model of Exercise 2
   using AIC and AICc. What does the comparison come down to when both models have the
   same number of parameters?
4. Plot the probabilities of both fitted models over the relative frequencies of the
   counts. Under each model, calculate the probability of an empty quadrat and the
   probability of 20 or more seedlings, and compare them with the observed fractions. What
   does each model say about how the counts arise? Can the comparison identify the
   ecological cause of the empty quadrats?

## Solution

### Question 1

The function follows the two cases of the mass function. It first calculates the Poisson
probability of every count multiplied by $1-p_0$, and then adds $p_0$ where the count is
zero. `ifelse()` applies its test to every element of `x` and returns the second argument
where the test is `TRUE` and the third where it is `FALSE`, so the function accepts a
vector of counts, like the `d*` functions of R. With `log = TRUE` it returns the logarithm
of the result:

```{r}
# Mass function of a zero-inflated Poisson distribution: with probability pzero a quadrat
# cannot have seedlings, otherwise the count follows a Poisson distribution.
dzipois = function(x, lambda, pzero, log = FALSE) {
  # Poisson probability scaled by the probability of belonging to the Poisson part.
  result = (1 - pzero) * dpois(x = x, lambda = lambda)
  # An empty quadrat can come from either part, so pzero is added to the probability of zero.
  result = ifelse(x == 0, pzero + result, result)
  if (log) log(result) else result
}

# The probabilities add up to one over a wide range of counts.
sum(dzipois(x = 0:200, lambda = 7.9, pzero = 0.5))
sum(dzipois(x = 0:200, lambda = 2, pzero = 0.1))
```

Both sums are 1, as they should be for a probability mass function. We used two different
pairs of parameter values, the first of which is close to the estimates of question 2.

### Question 2

Both parameters are constrained. The mean $\lambda$ is positive, so we search over
$\log\lambda$, and $p_0$ is a probability, so we search over $\operatorname{logit}p_0$ and
transform it with `plogis()`. For the starting values, suppose that every empty quadrat is
one that cannot have seedlings. Then $p_0$ is the fraction of empty quadrats and $\lambda$
is the mean of the quadrats that have seedlings:

```{r}
# Negative log-likelihood of the zero-inflated Poisson model.
# par[1] is log(lambda) and par[2] is logit(pzero).
zip_nll = function(par, counts) {
  lambda = exp(par[1])
  pzero = plogis(par[2])
  -sum(dzipois(x = counts, lambda = lambda, pzero = pzero, log = TRUE))
}

# Starting values from the data: the fraction of empty quadrats for pzero and the mean
# of the non-empty quadrats for lambda.
start_zip = c(
  log_lambda = log(mean(lily_counts[lily_counts > 0])),
  logit_pzero = qlogis(p = mean(lily_counts == 0))
)

zip_fit = optim(
  par = start_zip,
  fn = zip_nll,
  counts = lily_counts,
  method = "BFGS"
)
zip_fit$convergence

# Estimates on the original scale.
lambda_zip = unname(exp(zip_fit$par["log_lambda"]))
pzero_zip = unname(plogis(zip_fit$par["logit_pzero"]))
c(lambda = lambda_zip, pzero = pzero_zip, loglik = -zip_fit$value)

# Starting values on the original scale, and the probability of a count of zero under
# a Poisson distribution with the estimated lambda.
c(lambda_start = exp(start_zip[["log_lambda"]]), pzero_start = plogis(start_zip[["logit_pzero"]]))
dpois(x = 0, lambda = lambda_zip)
```

The convergence code is 0. The estimates are $\hat\lambda\approx 7.85$ seedlings per
quadrat and $\hat p_0\approx 0.496$, and the maximum log-likelihood is $-848.0$. According
to this model, about half of the quadrats cannot have seedlings and the other half have
7.85 seedlings on average. The estimates are almost identical to the starting values (7.85
and 0.496). The reason is that a Poisson distribution with a mean of 7.85 gives a
probability of only 0.0004 to a count of zero, so the model attributes practically all the
empty quadrats to the zero inflation.

### Question 3

The zero-inflated Poisson model has two parameters, like the Negative Binomial model:

```{r}
# AIC of the zero-inflated Poisson model (K = 2) next to the Negative Binomial model of
# Exercise 2 (K = 2): with the same K, the comparison is a comparison of log-likelihoods.
aic_zip = 2 * 2 - 2 * (-zip_fit$value)
c(zero_inflated_poisson = aic_zip, negative_binomial = aic_lily[["negative_binomial"]])

# AICc of both models: the correction 2K(K + 1) / (n - K - 1) is the same for both.
aicc_zip = aic_zip + 2 * 2 * (2 + 1) / (n_lily - 2 - 1)
c(zero_inflated_poisson = aicc_zip, negative_binomial = aicc_lily[["negative_binomial"]])
```

The AIC is 1700.1 for the zero-inflated Poisson model and 1132.5 for the Negative Binomial
model. Both models have $K=2$, so the term $2K$ cancels in the difference and the
comparison comes down to the maximum log-likelihoods: the difference of 567.6 units of AIC
is twice the difference of 283.8 units of log-likelihood ($-564.2$ against $-848.0$). The
Negative Binomial model is strongly preferred. The Poisson model without zero inflation had
an AIC of 2731.8 in Exercise 2, so allowing for quadrats that cannot have seedlings
improves the Poisson model a great deal, but not enough. The AICc is 1700.1 and 1132.5.
The correction of 0.047 units is the same for both models, because they have the same $K$
and the same $n$, so the difference does not change either.

### Question 4

We calculate the two probabilities under each model, together with the corresponding
fractions of the sample. For the Negative Binomial model, `pnbinom()` with
`lower.tail = FALSE` gives the probability of more than 19 seedlings:

```{r}
# Probability of zero and of 20 or more seedlings: observed, zero-inflated Poisson and
# Negative Binomial.
zip_probability = dzipois(x = count_values_lily, lambda = lambda_zip, pzero = pzero_zip)
c(
  zero_observed = mean(lily_counts == 0),
  zero_zip = zip_probability[1],
  zero_nb = nb_probability[1]
)
c(
  twenty_or_more_observed = mean(lily_counts >= 20),
  twenty_or_more_zip = sum(zip_probability[count_values_lily >= 20]),
  twenty_or_more_nb = pnbinom(q = 19, mu = mu_mle_lily, size = k_mle_lily, lower.tail = FALSE)
)
```

```{r}
#| label: fig-lily_zip
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Relative frequencies of the number of seedlings per quadrat (bars) and probabilities under the fitted zero-inflated Poisson and Negative Binomial models, up to 40 seedlings."
# Bars for the sample and lines with points for the two fitted models.
fitted_data_zip = rbind(
  data.frame(count = count_values_lily, probability = zip_probability, model = "Zero-inflated Poisson"),
  data.frame(count = count_values_lily, probability = nb_probability, model = "Negative Binomial")
)

ggplot() +
  geom_col(
    data = observed_data_lily,
    aes(x = count, y = frequency),
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_line(
    data = fitted_data_zip,
    aes(x = count, y = probability, colour = model, linetype = model),
    linewidth = 0.8
  ) +
  geom_point(
    data = fitted_data_zip,
    aes(x = count, y = probability, colour = model, shape = model)
  ) +
  scale_colour_manual(values = c("#D55E00", "grey20")) +
  coord_cartesian(xlim = c(-0.5, 40)) +
  labs(
    x = "Number of seedlings per quadrat",
    y = "Relative frequency or probability",
    colour = NULL,
    linetype = NULL,
    shape = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")
```

The zero-inflated Poisson model reproduces the fraction of empty quadrats exactly (0.496)
and the Negative Binomial model almost (0.483). The models differ at the other end of the
distribution (@fig-lily_zip). Ten of the 256 quadrats (3.9%) have 20 or more seedlings. The
Negative Binomial model gives a probability of 0.049 to such counts and the zero-inflated
Poisson model 0.0001. In the latter, the quadrats that are not empty follow a Poisson
distribution, whose variance equals its mean of 7.85, and that makes counts of 20 or more
almost impossible. Each of those ten quadrats then contributes a very low probability to
the likelihood, which is where the zero-inflated Poisson model loses most of its
log-likelihood.

The two models describe the origin of the counts differently. In the zero-inflated Poisson
model there are two kinds of quadrats: those that cannot have seedlings, and those in which
seedlings occur with one common mean and little variation around it. In the Negative
Binomial model there is one kind of quadrat, and the counts vary strongly around a common
mean, which produces both the many empty quadrats and the few crowded ones. AIC tells us
that the second description fits these counts much better. It does not tell us why the
seedlings are distributed in this way. As
[Chapter 2](../../Chapter_2/Theory.qmd#sec-discrete_distributions) explains, a distribution
does not identify the source of the overdispersion, and different processes (for example,
the dispersal of seeds around the parent plants, or differences in how suitable the
quadrats are for germination) can produce the same distribution of counts. The two
mechanisms could also act together, as in the zero-inflated Negative Binomial distribution
of the supplement, which we do not fit here. To learn about the cause we would need other
measurements on the quadrats in addition to the counts.

# Exercise 4: Timber volume and body mass

This exercise fits models to continuous measurements, so the contributions to the
likelihood are densities. We use a [Gamma](../../Supplements/distributions.qmd#sec-gamma)
model with shape $k$ and rate $r$ (`shape` and `rate` in R), for which

$$
X_i \sim \operatorname{Gamma}(k, r), \qquad E[X_i]=\frac{k}{r},
\qquad \operatorname{Var}(X_i)=\frac{k}{r^2}.
$$

The first three questions use the data set `trees` of the Chapter 3 practical, with the
measurements of 31 felled black cherry trees. That practical used their heights. Here $X_i$
is the volume of timber of tree $i$ in cubic feet (ft³), the column `Volume`. The last two
questions use the data set `penguins`, which is part of the `datasets` package in R 4.5.0
and later.^[With an older version of R, install the package `palmerpenguins` and load the
data with `data(penguins, package = "palmerpenguins")`. It contains the same measurements,
but the column with the body mass is called `body_mass_g`.] It contains measurements of
adult penguins of three species near Palmer Station, Antarctica (Gorman, Williams and
Fraser 2014). There $X_i$ is the body mass of Adelie penguin $i$, the column `body_mass`,
in grams. In both cases we assume that the observations are i.i.d.

1. Load the trees with `data(trees, package = "datasets")`. Write the NLL of the Gamma
   model with `dgamma(..., log = TRUE)` and with $\log k$ and $\log r$ as parameters.
   Calculate method-of-moments estimates from the formulas of the mean and the variance,
   use them as starting values and fit the model with `optim()` and `hessian = TRUE`.
   Report the estimates of $k$ and $r$ with their units and the maximum likelihood estimate
   of the mean volume.
2. Calculate the correlation between the estimates of $\log k$ and $\log r$ from the
   Hessian. Calculate the log-likelihood on a grid of values of $k$ and $r$ and draw the
   contours of the relative log-likelihood. What does the tilt of the contours mean?
3. Calculate a slice and a profile of the log-likelihood for the shape $k$ and plot both as
   relative log-likelihoods. Find the range of values of $k$ for which each curve is within
   1.92 units of the maximum. Which of the two ranges is a confidence interval, and why is
   the slice narrower?
4. Load the penguins with `data(penguins, package = "datasets")` and keep the body masses
   of the Adelie penguins. One of them is missing: remove it explicitly. Fit the Gamma
   model to the masses in grams and again to the same masses in kilograms. With the masses
   in grams the NLL is a large number, and the default stopping rule of `optim()` can end
   the search too early, so add the argument `control = list(reltol = 1e-12)` to both fits
   (if R prints "NaNs produced" during the search, see the callout
   [When a fit fails or prints warnings](../Theory.qmd#sec-maximum-likelihood)).
   Compare the estimates and the maximum log-likelihoods of the two fits. Explain the
   differences with the units of a density.
5. Fit a [LogNormal](../../Supplements/distributions.qmd#sec-lognormal) model to the body
   masses with `dlnorm()` and `optim()`, in grams and in kilograms, and compare it with the
   Gamma model using AIC and AICc. Does the comparison depend on the units?

## Solution

### Question 1

The observational unit is one tree, and the data are the 31 volumes in cubic feet.
`dgamma()` returns a density, so the NLL is minus the sum of the log densities. Both
parameters are positive and we search over their logarithms. For the starting values we
solve $E[X]=k/r$ and $\operatorname{Var}(X)=k/r^2$ for the parameters, which gives
$k=E[X]^2/\operatorname{Var}(X)$ and $r=E[X]/\operatorname{Var}(X)$, and we insert the mean
and the variance of the sample (with divisor $n$):

```{r}
# Load the black cherry data from the datasets package and keep the timber volume (cubic feet).
data(trees, package = "datasets")
tree_volume = trees$Volume
n_trees = length(tree_volume)

# Negative log-likelihood of the Gamma model with shape k and rate r.
# par[1] is log(k) and par[2] is log(r). dgamma() returns a density, so these are
# log densities.
gamma_nll = function(par, x) {
  shape = exp(par[1])
  rate = exp(par[2])
  -sum(dgamma(x = x, shape = shape, rate = rate, log = TRUE))
}

# Method-of-moments starting values from E[X] = k / r and Var(X) = k / r^2,
# which give k = mean^2 / variance and r = mean / variance.
volume_mean = mean(tree_volume)
volume_variance = mean((tree_volume - volume_mean)^2)
start_volume = log(c(
  log_shape = volume_mean^2 / volume_variance,
  log_rate = volume_mean / volume_variance
))
start_volume

volume_fit = optim(
  par = start_volume,
  fn = gamma_nll,
  x = tree_volume,
  method = "BFGS",
  hessian = TRUE
)
volume_fit$convergence

# Estimates on the original scale and the mean of the fitted model, by invariance.
shape_volume = unname(exp(volume_fit$par["log_shape"]))
rate_volume = unname(exp(volume_fit$par["log_rate"]))
c(
  shape = shape_volume,
  rate_per_ft3 = rate_volume,
  mean_ft3 = shape_volume / rate_volume,
  sample_mean_ft3 = volume_mean,
  nll = volume_fit$value
)

# Method-of-moments estimate of the shape, for comparison.
exp(start_volume["log_shape"])
```

The convergence code is 0. The estimates are $\hat k\approx 3.89$, which has no units, and
$\hat r\approx 0.129$/ft³. The rate has the units of one over the measurement,
so that $k/r$ is a volume. By invariance, the MLE of the mean volume is
$\hat k/\hat r\approx 30.2$ ft³. It equals the mean of the sample: as for the Negative
Binomial model, the MLE of the mean of a Gamma model is the sample mean. The
method-of-moments estimate of the shape is 3.48, so again the two methods agree on the
mean and disagree on the parameter that controls the variance. The minimum of the NLL is
125.7, that is, a maximum log-likelihood of $-125.7$.

### Question 2

`cov2cor()` converts the inverse of the Hessian into correlations. For the surface we use
the same nested loops as in Exercise 2, now with `dgamma()`:

```{r}
# Correlation between the estimates of log(k) and log(r), from the inverse Hessian.
cov2cor(solve(volume_fit$hessian))

# Grids of candidate values for the shape and the rate (1/ft³).
shape_grid = seq(from = 1.5, to = 8, length.out = 71)
rate_grid = seq(from = 0.04, to = 0.3, length.out = 71)

# Fill the matrix of log-likelihoods with two nested loops.
loglik_surface_volume = matrix(NA, nrow = length(shape_grid), ncol = length(rate_grid))
for (i in seq_along(shape_grid)) {
  for (j in seq_along(rate_grid)) {
    loglik_surface_volume[i, j] = sum(
      dgamma(x = tree_volume, shape = shape_grid[i], rate = rate_grid[j], log = TRUE)
    )
  }
}
relative_surface_volume = loglik_surface_volume + volume_fit$value
```

```{r}
#| label: fig-volume_surface
#| fig-cap: "Contours of the relative log-likelihood of the tree volumes as a function of the shape and the rate of the Gamma model. The circle is the maximum likelihood estimate."
# Contours of the relative log-likelihood and the maximum likelihood estimate.
contour(
  x = shape_grid,
  y = rate_grid,
  z = relative_surface_volume,
  levels = c(-12, -9, -6, -4, -2, -1),
  xlab = "Shape (k)",
  ylab = "Rate (r, 1/ft³)"
)
points(x = shape_volume, y = rate_volume, pch = 16, col = "#D55E00")
```

The correlation between the estimates of $\log k$ and $\log r$ is 0.94, and the contours of
@fig-volume_surface are strongly tilted: a larger shape goes with a larger rate (the outer
contours are cut by the edges of the grid). The reason is the same as for the parameters
`size` and `prob` of the Negative Binomial distribution in the theory. The data determine
the mean volume $k/r$ well, so a higher value of $k$ can be compensated by a higher value
of $r$ that keeps the mean close to 30 ft³, with little loss of likelihood. The tilt means
that the estimates of the two parameters depend on each other: a hypothetical experiment
that overestimates the shape would tend to overestimate the rate too.

### Question 3

The slice fixes the rate at its estimate and varies the shape. The profile re-estimates the
rate for every value of the shape, with an NLL in which the shape is given and only
$\log r$ is searched. Each of these inner searches starts at the joint estimate. We
evaluate both curves on a grid of 261 values of the shape:

```{r}
# Cutoff for a 95% interval of one parameter: half the 95% quantile of a chi-squared
# distribution with one degree of freedom.
cutoff = qchisq(p = 0.95, df = 1) / 2

# Negative log-likelihood as a function of log(r) only, for a given shape.
gamma_nll_given_shape = function(log_rate, shape, x) {
  -sum(dgamma(x = x, shape = shape, rate = exp(log_rate), log = TRUE))
}

# Slice: the rate is fixed at its estimate. Profile: for each shape the rate is
# re-estimated with optim(), starting from the joint estimate.
shape_fine = seq(from = 1.5, to = 8, length.out = 261)
slice_loglik_volume = rep(NA, times = length(shape_fine))
profile_loglik_volume = rep(NA, times = length(shape_fine))
for (i in seq_along(shape_fine)) {
  slice_loglik_volume[i] = sum(
    dgamma(x = tree_volume, shape = shape_fine[i], rate = rate_volume, log = TRUE)
  )
  inner_fit = optim(
    par = volume_fit$par["log_rate"],
    fn = gamma_nll_given_shape,
    shape = shape_fine[i],
    x = tree_volume,
    method = "BFGS"
  )
  profile_loglik_volume[i] = -inner_fit$value
}

# Relative log-likelihoods and the range of shapes within the cutoff of the maximum.
slice_relative_volume = slice_loglik_volume + volume_fit$value
profile_relative_volume = profile_loglik_volume + volume_fit$value
range(shape_fine[slice_relative_volume >= -cutoff])
range(shape_fine[profile_relative_volume >= -cutoff])
```

```{r}
#| label: fig-volume_profile
#| fig-cap: "Slice (dashed line) and profile (solid line) of the relative log-likelihood for the shape of the Gamma model fitted to the tree volumes. The horizontal line is at minus half the 95% quantile of the chi-squared distribution with one degree of freedom."
# Relative log-likelihood along the slice and along the profile.
curve_volume = rbind(
  data.frame(shape = shape_fine, relative = slice_relative_volume, curve = "Slice"),
  data.frame(shape = shape_fine, relative = profile_relative_volume, curve = "Profile")
)
ggplot(curve_volume, aes(x = shape, y = relative, linetype = curve)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_hline(yintercept = -cutoff, colour = "grey40") +
  coord_cartesian(ylim = c(-6, 0)) +
  labs(x = "Shape (k)", y = "Relative log-likelihood", linetype = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The slice is within 1.92 units of the maximum for shapes between 3.28 and 4.55, and the
profile for shapes between 2.33 and 6.05 (the grid has a step of 0.025, so the endpoints
are accurate to about that amount). The range of the profile is the approximate 95%
confidence interval for the shape. The slice is about a third as wide
(@fig-volume_profile). It describes what the data would say about the shape if we knew that
the rate is exactly 0.129/ft³. The rate is estimated from the same 31 trees, and
because the two estimates are strongly correlated, fixing one of them removes most of the
uncertainty about the other. The profile follows the ridge of the surface in
@fig-volume_surface and takes that uncertainty into account.

### Question 4

We select the Adelie penguins and count the missing body masses with `is.na()`. A missing
mass cannot contribute to the likelihood, so we remove it explicitly and report it. The
NLL is the function `gamma_nll()` of question 1, and a small function calculates the
method-of-moments starting values for each unit:

```{r}
# Load the penguin data (in R 4.5.0 and later) and keep the Adelie penguins.
data(penguins, package = "datasets")
adelie = penguins[penguins$species == "Adelie", ]

# One body mass is missing; remove it explicitly and keep the masses in grams.
sum(is.na(adelie$body_mass))
mass_g = adelie$body_mass[!is.na(adelie$body_mass)]
n_penguins = length(mass_g)

# The same masses in kilograms.
mass_kg = mass_g / 1000

# Fit the Gamma model to the masses in grams and in kilograms, with method-of-moments
# starting values calculated for each unit.
start_from_moments = function(x) {
  x_mean = mean(x)
  x_variance = mean((x - x_mean)^2)
  log(c(log_shape = x_mean^2 / x_variance, log_rate = x_mean / x_variance))
}

# The default stopping rule of optim() is relative to the value of the negative
# log-likelihood, which is large in grams, so the search could stop almost at the starting
# values. A smaller relative tolerance (reltol) makes both fits reach the maximum.
penguin_fit_g = optim(
  par = start_from_moments(mass_g),
  fn = gamma_nll,
  x = mass_g,
  method = "BFGS",
  hessian = TRUE,
  control = list(reltol = 1e-12)
)
penguin_fit_kg = optim(
  par = start_from_moments(mass_kg),
  fn = gamma_nll,
  x = mass_kg,
  method = "BFGS",
  hessian = TRUE,
  control = list(reltol = 1e-12)
)
c(grams = penguin_fit_g$convergence, kilograms = penguin_fit_kg$convergence)

# Estimates on the original scale: the shape, and the rate in 1/g and in 1/kg.
shape_penguin = unname(exp(penguin_fit_g$par["log_shape"]))
rate_penguin_g = unname(exp(penguin_fit_g$par["log_rate"]))
rate_penguin_kg = unname(exp(penguin_fit_kg$par["log_rate"]))
c(
  n = n_penguins,
  shape_g = shape_penguin,
  shape_kg = unname(exp(penguin_fit_kg$par["log_shape"])),
  rate_per_g = rate_penguin_g,
  rate_per_kg = rate_penguin_kg
)

# Mean of the fitted model, k / r, in grams and in kilograms.
c(mean_g = shape_penguin / rate_penguin_g, mean_kg = shape_penguin / rate_penguin_kg)

# Maximum log-likelihood in each unit and the change when the unit changes. A density is
# per unit of the measurement, so the data in kilograms have densities 1,000 times larger.
loglik_g = -penguin_fit_g$value
loglik_kg = -penguin_fit_kg$value
c(loglik_g = loglik_g, loglik_kg = loglik_kg, difference = loglik_kg - loglik_g)
n_penguins * log(1000)
```

One body mass is missing, which leaves 151 Adelie penguins. Both fits have convergence code
0. The estimate of the shape is 66.09 in both units. The estimate of the rate is
0.01786/g and 17.86/kg, a factor of 1,000, and the estimated mean $\hat k/\hat r$
is the same mass in both cases (3,701 g, or 3.70 kg). So the fitted model is the same and
only the units of the rate change.

The maximum log-likelihood does change. It is $-1137.72$ in grams and $-94.65$ in
kilograms, a difference of 1043.07. A density is a probability per unit of the
measurement. The probability that a penguin weighs between 3,700 and 3,701 g is the same
in any units, but in kilograms that interval has a width of 0.001, so the density must be
1,000 times larger. Every penguin therefore adds $\log(1000)\approx 6.91$ to the
log-likelihood, and $151\times\log(1000)=1043.07$ is the difference that we found. For this
reason the value of a log-likelihood means nothing by itself, and log-likelihoods
calculated from data in different units cannot be compared.

The argument `reltol` deserves a remark. `optim()` stops when a step reduces the function
by less than a small fraction of its value. With the masses in grams and the default
setting, the search stops after a single step at a shape of 65.56, practically at the
starting values, although the NLL can still decrease by 0.0025, and it reports a
convergence code of zero. A smaller `reltol` makes the search continue to the minimum. While
it searches, `optim()` may try a rate or a shape for which `dgamma()` returns `NaN`, and R
then prints "NaNs produced". These warnings are harmless when the search ends with
convergence code 0 and sensible estimates (see
[When a fit fails or prints warnings](../Theory.qmd#sec-maximum-likelihood)). How
optimisers decide when to stop is a topic of [Chapter 8](../../Chapter_8/Theory.qmd).

### Question 5

The LogNormal model has the parameters $\mu$ and $\sigma$ (`meanlog` and `sdlog` in R),
which are the mean and the standard deviation of the logarithm of the mass. The first can
take any real value and the second is positive, so we search over $\mu$ and $\log\sigma$.
As starting values we use the mean and the standard deviation of the logarithms of the
masses:

```{r}
# Negative log-likelihood of the LogNormal model; par[1] is meanlog and par[2] is log(sdlog).
lognormal_nll = function(par, x) {
  -sum(dlnorm(x = x, meanlog = par[1], sdlog = exp(par[2]), log = TRUE))
}

# Starting values: mean and standard deviation of the logarithms of the masses.
start_lognormal = function(x) {
  c(meanlog = mean(log(x)), log_sdlog = log(sd(log(x))))
}
lognormal_fit_g = optim(
  par = start_lognormal(mass_g),
  fn = lognormal_nll,
  x = mass_g,
  method = "BFGS",
  control = list(reltol = 1e-12)
)
lognormal_fit_kg = optim(
  par = start_lognormal(mass_kg),
  fn = lognormal_nll,
  x = mass_kg,
  method = "BFGS",
  control = list(reltol = 1e-12)
)
c(grams = lognormal_fit_g$convergence, kilograms = lognormal_fit_kg$convergence)

# AIC = 2K - 2 * log-likelihood; both models have K = 2 parameters. The differences in
# AIC are the same in grams and in kilograms.
aic_penguin = data.frame(
  unit = c("grams", "kilograms"),
  gamma = 2 * 2 + 2 * c(penguin_fit_g$value, penguin_fit_kg$value),
  lognormal = 2 * 2 + 2 * c(lognormal_fit_g$value, lognormal_fit_kg$value)
)
aic_penguin$difference = aic_penguin$gamma - aic_penguin$lognormal
aic_penguin

# AICc adds 2K(K + 1) / (n - K - 1) with K = 2 and n the number of penguins; the
# correction is the same constant for both models, so the difference does not change.
aicc_penguin = aic_penguin
aicc_penguin$gamma = aic_penguin$gamma + 2 * 2 * (2 + 1) / (n_penguins - 2 - 1)
aicc_penguin$lognormal = aic_penguin$lognormal + 2 * 2 * (2 + 1) / (n_penguins - 2 - 1)
aicc_penguin$difference = aicc_penguin$gamma - aicc_penguin$lognormal
aicc_penguin
```

All four fits have convergence code 0. In grams the AIC is 2279.4 for the Gamma model and
2279.0 for the LogNormal model, and in kilograms it is 193.3 and 192.8. The values depend
on the units, for the reason given in question 4. The difference between the two models is
0.49 in both units, because the change of units adds the same constant to the
log-likelihood of every model. So a comparison with AIC does not depend on the units,
provided that all the models are fitted to the same observations in the same units.

The AICc adds 0.081 units to both models (two parameters and 151 penguins), which gives
2279.5 and 2279.0 in grams and 193.4 and 192.9 in kilograms. The difference of 0.49 is
unchanged, so AICc leads to the same conclusion as AIC, and the units do not matter for
it either.

A difference of half a unit of AIC is very small. Both models have two parameters, so it
corresponds to a difference of 0.25 between their maximum log-likelihoods. These data
hardly distinguish a Gamma model from a LogNormal model for the body mass of Adelie
penguins.

# Exercise 5: Estimators and intervals across hypothetical experiments

This exercise is extra practice. With a single sample we cannot tell how far an estimate
is from the parameter, or whether a confidence interval contains it, because the parameter
is unknown. As in Exercise 3 of the Chapter 3 practical, we can study the behaviour of an
estimator by simulating hypothetical experiments from a model whose parameters we know.
Here we use the maximum likelihood estimates of Exercise 4 as the parameters of that model.
The simulation then shows how the estimators and the intervals would behave if the fitted
Gamma model had generated the data (see
[Properties of maximum likelihood estimators](../Theory.qmd#sec-mle_properties) and
[Uncertainty from the likelihood](../Theory.qmd#sec-likelihood-uncertainty)). This
exercise continues with the objects of Exercise 4. The function that runs the simulation
is given below the questions with three gaps, which you fill in.

1. For the shape $k$ of the Gamma model fitted to the tree volumes, calculate a 95% Wald
   interval (on the log scale, with the endpoints transformed back) and a 95% profile
   likelihood interval with `uniroot()`. Compare them.
2. Complete the first gap of the function `simulate_gamma_experiments()` below. Set seed
   801 and simulate 2,000 hypothetical experiments, each with 31 trees from a Gamma model
   whose shape and rate are the maximum likelihood estimates of Exercise 4. For each
   experiment, the function estimates the shape by the method of moments and by maximum
   likelihood (the fit starts at the method-of-moments estimates and uses
   `control = list(reltol = 1e-12)`, as in Exercise 4). Compare the mean, the standard
   deviation and the mean squared error of the two estimators with the shape that you used
   to simulate. Tip: the function has the sample size, the shape and the rate as arguments
   because you will need it again in question 4.
3. Complete the second and the third gap, which record for each experiment whether the
   Wald interval and the profile likelihood interval for the shape contain the shape that
   you used to simulate, and calculate the coverage of both intervals. Hint: you do not
   need `uniroot()` inside the simulation. A profile likelihood interval contains a value
   whenever the profile log-likelihood at that value is within
   `qchisq(p = 0.95, df = 1) / 2` of the maximum log-likelihood, so one extra fit with the
   shape fixed at that value is enough.
4. Set seed 802 and repeat questions 2 and 3 with experiments of 151 penguins from the
   Gamma model fitted to the body masses in grams. How do the bias and the coverage change?
   The two simulations differ in the sample size and in the model. How can you check which
   of the two explains the change? Do it, without resetting the seed.
5. Why do these results only approximate the behaviour of the estimators and the intervals
   for the real trees and penguins?

The function below needs `gamma_nll()` and `gamma_nll_given_shape()` from Exercise 4, and
the objects `critical_value` and `cutoff` (the critical value of the standard Normal
distribution and the cutoff of the relative log-likelihood for a 95% interval). It stores
the results of every experiment in one row of a data frame. In each gap, replace `NA` by
your own code. The lines that follow a gap use the names of the objects that the gap
creates, so keep those names.

```{r}
#| eval: false
# One simulation of hypothetical experiments from a Gamma model whose parameters we know.
# For each experiment: the maximum likelihood and the method-of-moments estimates of the
# shape, and whether the Wald interval (log scale) and the profile interval contain the
# shape that was used to simulate.
simulate_gamma_experiments = function(n_obs, shape, rate, n_experiments) {
  results = data.frame(
    shape_ml = rep(NA, times = n_experiments),
    shape_mom = rep(NA, times = n_experiments),
    wald_covered = rep(NA, times = n_experiments),
    profile_covered = rep(NA, times = n_experiments),
    converged = rep(NA, times = n_experiments)
  )
  for (e in seq_len(n_experiments)) {
    # One hypothetical experiment and its method-of-moments estimates.
    x = rgamma(n = n_obs, shape = shape, rate = rate)
    x_mean = mean(x)
    x_variance = mean((x - x_mean)^2)
    shape_mom = x_mean^2 / x_variance

    # GAP 1: maximum likelihood fit of the Gamma model to x with optim(), started at the
    # method-of-moments estimates of the shape and the rate (on the log scale), with
    # method = "BFGS", hessian = TRUE and control = list(reltol = 1e-12). Store the result
    # in an object called fit.
    fit = NA
    shape_ml = exp(fit$par[1])
    se_log_shape = sqrt(solve(fit$hessian)[1, 1])

    # The Wald interval contains the shape when the estimate is within the critical value
    # of standard errors on the log scale. The profile interval contains it when the
    # relative profile log-likelihood at that shape is within the cutoff.

    # GAP 2: TRUE when the Wald interval contains the shape that was used to simulate.
    wald_covered = NA

    # GAP 3: fit the model with the shape fixed at the shape that was used to simulate,
    # which re-estimates the rate only (use gamma_nll_given_shape() with method = "BFGS"
    # and control = list(reltol = 1e-12), starting at the estimate of log(rate) of fit),
    # and store it in an object called inner_fit. Then record in profile_covered whether the
    # relative profile log-likelihood at that shape is within the cutoff.
    inner_fit = NA
    profile_covered = NA

    results$shape_ml[e] = shape_ml
    results$shape_mom[e] = shape_mom
    results$wald_covered[e] = wald_covered
    results$profile_covered[e] = profile_covered
    results$converged[e] = fit$convergence == 0 && inner_fit$convergence == 0
  }
  results
}
```

## Solution

### Question 1

The Wald interval uses the standard error of the estimate of $\log k$, which is the square
root of the first diagonal element of the inverse Hessian. For the profile likelihood
interval we write the relative profile log-likelihood of the shape as a function, which
uses `gamma_nll_given_shape()` of Exercise 4, and find with `uniroot()` where it equals
$-1.92$. The profile plotted in Exercise 4 is below the cutoff at both ends of its grid
(1.5 and 8), so we search between 1 and the estimate and between the estimate and 15:

```{r}
# Wald interval for the shape of the tree volumes: calculate on the log scale with the
# standard error from the inverse Hessian, and back-transform the endpoints.
critical_value = qnorm(p = 0.975)
se_log_shape_volume = sqrt(solve(volume_fit$hessian)[1, 1])
wald_shape_volume = exp(volume_fit$par["log_shape"] + c(-1, 1) * critical_value * se_log_shape_volume)

# Profile log-likelihood of the shape relative to the maximum, with the rate re-estimated.
gamma_profile_relative = function(shape, x, fit) {
  inner_fit = optim(
    par = fit$par["log_rate"],
    fn = gamma_nll_given_shape,
    shape = shape,
    x = x,
    method = "BFGS"
  )
  fit$value - inner_fit$value
}

# uniroot() finds where the relative profile equals -cutoff. The ends of the search
# intervals were chosen from the profile plotted in Exercise 4: below and above the
# estimate, where the profile is under -cutoff.
shape_lower_volume = uniroot(
  f = function(shape) gamma_profile_relative(shape, x = tree_volume, fit = volume_fit) + cutoff,
  interval = c(1, shape_volume),
  tol = 1e-8
)$root
shape_upper_volume = uniroot(
  f = function(shape) gamma_profile_relative(shape, x = tree_volume, fit = volume_fit) + cutoff,
  interval = c(shape_volume, 15),
  tol = 1e-8
)$root

rbind(
  wald = unname(wald_shape_volume),
  profile = c(shape_lower_volume, shape_upper_volume)
)
```

The Wald interval goes from 2.41 to 6.27 and the profile likelihood interval from 2.32 to
6.06, which agrees with the range that we found on the grid in Exercise 4. The two
intervals differ in the first decimal, unlike those for the lily seedlings in Exercise 2:
with 31 trees the profile log-likelihood is further from a parabola. From a single sample
we cannot tell which of the two intervals behaves better, which is what the simulation is
for.

### Question 2

Here is the skeleton with its three gaps filled in, and a second function that summarises
the results. For each experiment the first function simulates the measurements with
`rgamma()`, calculates the method-of-moments estimates and uses them to start the maximum
likelihood fit (gap 1). It also records what question 3 asks for (gaps 2 and 3) and whether
the fits stopped normally. The results are stored in a data frame with one row per
experiment:

```{r}
# One simulation of hypothetical experiments from a Gamma model whose parameters we know.
# For each experiment: the maximum likelihood and the method-of-moments estimates of the
# shape, and whether the Wald interval (log scale) and the profile interval contain the
# shape that was used to simulate.
simulate_gamma_experiments = function(n_obs, shape, rate, n_experiments) {
  results = data.frame(
    shape_ml = rep(NA, times = n_experiments),
    shape_mom = rep(NA, times = n_experiments),
    wald_covered = rep(NA, times = n_experiments),
    profile_covered = rep(NA, times = n_experiments),
    converged = rep(NA, times = n_experiments)
  )
  for (e in seq_len(n_experiments)) {
    # One hypothetical experiment and its method-of-moments estimates.
    x = rgamma(n = n_obs, shape = shape, rate = rate)
    x_mean = mean(x)
    x_variance = mean((x - x_mean)^2)
    shape_mom = x_mean^2 / x_variance

    # Maximum likelihood fit, started at the method-of-moments estimates. The small
    # relative tolerance keeps the search from stopping too early when the negative
    # log-likelihood is large (the penguin masses are in grams).
    fit = optim(
      par = log(c(shape_mom, x_mean / x_variance)),
      fn = gamma_nll,
      x = x,
      method = "BFGS",
      hessian = TRUE,
      control = list(reltol = 1e-12)
    )
    shape_ml = exp(fit$par[1])
    se_log_shape = sqrt(solve(fit$hessian)[1, 1])

    # The Wald interval contains the shape when the estimate is within the critical value
    # of standard errors on the log scale. The profile interval contains it when the
    # relative profile log-likelihood at that shape is within the cutoff.
    wald_covered = abs(log(shape) - fit$par[1]) <= critical_value * se_log_shape
    inner_fit = optim(
      par = fit$par[2],
      fn = gamma_nll_given_shape,
      shape = shape,
      x = x,
      method = "BFGS",
      control = list(reltol = 1e-12)
    )
    profile_covered = fit$value - inner_fit$value >= -cutoff

    results$shape_ml[e] = shape_ml
    results$shape_mom[e] = shape_mom
    results$wald_covered[e] = wald_covered
    results$profile_covered[e] = profile_covered
    results$converged[e] = fit$convergence == 0 && inner_fit$convergence == 0
  }
  results
}

# Mean, standard deviation and mean squared error of both estimators of the shape,
# coverage of both intervals, and the number of experiments in which a fit failed.
summarise_experiments = function(results, shape) {
  data.frame(
    mean_ml = mean(results$shape_ml),
    sd_ml = sd(results$shape_ml),
    mse_ml = mean((results$shape_ml - shape)^2),
    mean_mom = mean(results$shape_mom),
    sd_mom = sd(results$shape_mom),
    mse_mom = mean((results$shape_mom - shape)^2),
    coverage_wald = mean(results$wald_covered),
    coverage_profile = mean(results$profile_covered),
    failed_fits = sum(!results$converged)
  )
}

# 2,000 hypothetical experiments of 31 trees from the Gamma model fitted to the volumes.
set.seed(seed = 801)
n_experiments = 2000
trees_31 = simulate_gamma_experiments(
  n_obs = n_trees,
  shape = shape_volume,
  rate = rate_volume,
  n_experiments = n_experiments
)
summarise_experiments(results = trees_31, shape = shape_volume)

# Relative bias of both estimators, and the median of the maximum likelihood estimates.
c(ml = mean(trees_31$shape_ml), mom = mean(trees_31$shape_mom)) / shape_volume - 1
median(trees_31$shape_ml)
```

The shape used to simulate is 3.89. Across the 2,000 hypothetical experiments, the maximum
likelihood estimates have a mean of 4.33, a standard deviation of 1.20 and a mean squared
error (MSE) of 1.63. The method-of-moments estimates have a mean of 4.43, a standard
deviation of 1.30 and an MSE of 1.99. No fit failed. So with 31 trees both estimators are
biased upwards, by 11% and 14%, and the MLE is the better of the two: it has the smaller
bias, the smaller standard deviation and the smaller MSE. The median of the maximum
likelihood estimates is 4.13, below their mean, because their sampling distribution is
skewed to the right. These summaries are themselves estimates from 2,000 experiments and
they change somewhat with the seed, in particular the standard deviations, which are
sensitive to a few very large estimates.

### Question 3

The simulation of question 2 already recorded what we need. For the Wald interval we do
not have to calculate the endpoints: the interval on the log scale contains $\log k$
exactly when the estimate of $\log k$ is within $z_{0.975}$ standard errors of it. For the
profile likelihood interval we followed the hint. One inner fit with the shape fixed at the
value used to simulate gives the profile log-likelihood at that value, and the interval
contains it when the difference from the maximum log-likelihood is above $-1.92$. The
coverage is the fraction of experiments in which each check is `TRUE`:

```{r}
# Coverage of the two intervals, from the experiments of question 2.
c(wald = mean(trees_31$wald_covered), profile = mean(trees_31$profile_covered))
```

The coverage is 0.931 for the Wald interval and 0.939 for the profile likelihood interval.
A coverage simulated with 2,000 experiments has a standard error of about 0.005
($\sqrt{0.94\times 0.06/2000}$), so both are below the nominal 0.95 by two standard errors
or more. With 31 trees both intervals are approximations, and they contain the shape
slightly less often than their confidence level states. The profile likelihood interval is
the closer of the two.

### Question 4

We call the same function with the sample size of the penguins and the estimates of the
Gamma model fitted to their masses in grams. To separate the effect of the sample size
from the effect of the model, we add a third simulation with the model of the trees and
the sample size of the penguins. If the sample size explains the change, 151 trees should
behave like 151 penguins and unlike 31 trees:

```{r}
# 2,000 hypothetical experiments of 151 penguins from the Gamma model fitted in Exercise 4
# (in grams), and of 151 trees from the model fitted to the tree volumes, to separate the
# effect of the sample size from the effect of the model.
set.seed(seed = 802)
penguins_151 = simulate_gamma_experiments(
  n_obs = n_penguins,
  shape = shape_penguin,
  rate = rate_penguin_g,
  n_experiments = n_experiments
)
trees_151 = simulate_gamma_experiments(
  n_obs = n_penguins,
  shape = shape_volume,
  rate = rate_volume,
  n_experiments = n_experiments
)

# Summary of the three simulations.
rbind(
  "31 trees" = summarise_experiments(results = trees_31, shape = shape_volume),
  "151 penguins" = summarise_experiments(results = penguins_151, shape = shape_penguin),
  "151 trees" = summarise_experiments(results = trees_151, shape = shape_volume)
)

# Relative bias of the maximum likelihood estimator of the shape in each simulation.
c(
  trees_31 = mean(trees_31$shape_ml) / shape_volume - 1,
  penguins_151 = mean(penguins_151$shape_ml) / shape_penguin - 1,
  trees_151 = mean(trees_151$shape_ml) / shape_volume - 1
)
```

No fit failed. With 151 penguins the relative bias of the MLE of the shape is 1.6% (a mean
of 67.2 against the 66.1 used to simulate), and the coverage is 0.947 for the Wald interval
and 0.946 for the profile likelihood interval, both within the simulation error of 0.95.
With 151 trees the relative bias is 1.5% and the coverages are 0.951 and 0.956. With 31
trees the bias was 11.5% and the coverages were 0.931 and 0.939. So the sample size
explains the change: the model of the tree volumes, which gave biased estimates and
intervals with a coverage below 0.95 for 31 trees, behaves as the large-sample results of
the theory predict when the sample has 151 trees.

How the two estimators compare does depend on the model. For the tree model with 151 trees
the MLE remains more precise than the method-of-moments estimator (a standard deviation of
0.44 against 0.51 and an MSE of 0.20 against 0.26), in agreement with its asymptotic
efficiency. For the penguin model the two estimators are almost equally precise (standard
deviations of 7.91 and 7.97).

### Question 5

The simulations describe the estimators and the intervals for models whose parameters we
chose: Gamma models with the estimates of Exercise 4 as parameters. The parameters of the
data generating processes of the real trees and penguins are unknown, and the estimates
differ from them. For the trees the difference can be considerable, since the confidence
interval for the shape goes from 2.3 to 6.1 and question 2 shows that the estimate tends to
be too large. The bias and the coverage would be somewhat different for another shape.

The simulated observations are also exactly i.i.d. from a Gamma distribution, whereas for
the real measurements the Gamma model is an approximation (in Exercise 4 a LogNormal model
described the body masses equally well), and the trees or the penguins need not be
independent of each other. With a misspecified model the properties of the estimators and
of the intervals change (see [Misspecified models](../Theory.qmd#sec-misspecified)), which
these simulations do not explore. So the results approximate the behaviour of the
estimators for the real data, and the approximation is better the closer the fitted model
is to the data generating process.
