---
title: "Probability distributions"
---

Before you start, read the entries for the [Binomial](../../Supplements/distributions.qmd#sec-binomial),
[Poisson](../../Supplements/distributions.qmd#sec-poisson),
[Negative Binomial](../../Supplements/distributions.qmd#sec-negative_binomial),
[Normal](../../Supplements/distributions.qmd#sec-normal),
[LogNormal](../../Supplements/distributions.qmd#sec-lognormal) and
[Beta](../../Supplements/distributions.qmd#sec-beta) distributions in the
[distribution supplement](../../Supplements/distributions.qmd). They give the probability mass or density
function, mean, variance and R functions of each distribution, which you will need in the exercises below.
The first two exercises are worked examples, one for a discrete and one for a continuous
distribution. Work through them before trying the remaining exercises.

# Exercise 1: Seedling survival

Assume that each of 25 seedlings has probability $0.65$ of surviving. Assume a Binomial
model to describe this process.

1. Calculate the probability that exactly 15 survive.
2. Calculate the probability that at least 15 survive.
3. Calculate the expected number and variance of survivors.
4. Simulate 1,000 groups of 25 seedlings and compare the observed frequencies with the model
probabilities.

## Solution

Assume that each of 25 seedlings has probability $0.65$ of surviving. Assume a Binomial
model to describe this process.

Let

$$
X=\text{the number of seedlings that survive}.
$$

The process has:

- a fixed number of trials, $n=25$;
- two possible outcomes for each seedling: survives or does not survive;
- the same survival probability, $p=0.65$;
- independent seedlings.

Therefore,

$$
X\sim\operatorname{Binomial}(n=25,p=0.65).
$$

In R, the Binomial distribution uses:

- `dbinom()` for probability mass;
- `pbinom()` for cumulative probabilities;
- `rbinom()` for simulation.

```{r}
n = 25
p_survive = 0.65
```

### 1. Probability that exactly 15 survive

We want to calculate

$$
P(X=15).
$$

The Binomial probability mass function is

$$
P(X=x)=
\binom{n}{x}p^x(1-p)^{n-x}.
$$

Substituting $x=15$, $n=25$, and $p=0.65$ gives

$$
\begin{aligned}
P(X=15)
&=
\binom{25}{15}(0.65)^{15}(1-0.65)^{25-15}\\
&=
\binom{25}{15}(0.65)^{15}(0.35)^{10}\\
&=
\frac{25!}{15!10!}(0.65)^{15}(0.35)^{10}\\
&\approx 0.14.
\end{aligned}
$$

Thus, the probability that exactly 15 seedlings survive is approximately **0.14**, or **14%**.

In R, `dbinom()` performs this calculation:

```{r}
prob_exactly_15 = dbinom(
  x = 15,
  size = n,
  prob = p_survive
)

prob_exactly_15
```

The combinatorial term can also be calculated directly in R:

```{r}
choose(25, 15)
```

### 2. Probability that at least 15 survive

“At least 15” means 15 or more:

$$
P(X\geq 15)=P(X=15)+P(X=16)+\cdots+P(X=25).
$$

By using the complement,

$$
P(X\geq 15)
=
1-P(X\leq 14).
$$

Using the Binomial formula, this is

$$
P(X\geq 15)
=
1-\sum_{x=0}^{14}
\binom{25}{x}(0.65)^x(0.35)^{25-x}.
$$

Evaluating this sum gives

$$
P(X\geq 15)\approx 0.77.
$$

Therefore, the probability that at least 15 seedlings survive is approximately **0.77**, or **77%**.

In R, `pbinom()` calculates $P(X\leq x)$ so we take the complement:

```{r}
prob_at_least_15 = 1 - pbinom(
  q = 14,
  size = n,
  prob = p_survive
)

prob_at_least_15
```

The value `q = 14` is used because

$$
P(X>14)=P(X\geq 15).
$$

### 3. Expected number and variance of survivors

For a Binomial random variable,

$$
\mu=E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$

The expected number of survivors is therefore

$$
\begin{aligned}
E[X]
&=25(0.65)\\
&=16.25.
\end{aligned}
$$

The variance is

$$
\begin{aligned}
\operatorname{Var}(X)
&=25(0.65)(1-0.65)\\
&=25(0.65)(0.35)\\
&=5.69.
\end{aligned}
$$

In R:

```{r}
expected_survivors = n * p_survive

variance_survivors = n * p_survive * (1 - p_survive)

c(
  expected = expected_survivors,
  variance = variance_survivors
)
```

The expected value, 16.25, does not need to be an integer. It represents the long-run
average number of survivors across many groups of 25 seedlings.

### 4. Simulating 1,000 groups

We can simulate 1,000 groups using `rbinom()`:

```{r}
set.seed(123)

simulated_survivors = rbinom(
  n = 1000,
  size = n,
  prob = p_survive
)

head(simulated_survivors)
```

Each value in `simulated_survivors` is the number of survivors in one simulated group.

To calculate the observed relative frequency for every possible number of survivors, use `table()`:

```{r}
support = 0:n

observed_frequency = as.numeric(
  table(factor(simulated_survivors, levels = support))
) / length(simulated_survivors)

model_probability = dbinom(
  x = support,
  size = n,
  prob = p_survive
)

comparison = data.frame(
  survivors = support,
  observed = observed_frequency,
  model = model_probability
)

round(comparison, 4)
```

The `factor()` step ensures that all possible values from 0 to 25 are included, even if a
value does not occur in the simulation.

We can compare the observed frequencies and the model probabilities graphically:

```{r}
barplot(
  rbind(observed_frequency, model_probability),
  beside = TRUE,
  names.arg = support,
  col = c("grey70", "steelblue"),
  xlab = "Number of seedlings surviving",
  ylab = "Probability",
  legend.text = c("Observed frequency", "Model probability"),
  args.legend = list(x = "topright"),
  main = "Observed and model seedling survival"
)
```

The observed frequencies will not exactly equal the model probabilities because only 1,000
groups were simulated. They should generally be similar. With a larger number of simulated
groups, the observed frequencies would usually become closer to the model probabilities.

We can also compare the simulated and model values for the first two questions:

```{r}
c(
  observed_exactly_15 = mean(simulated_survivors == 15),
  model_exactly_15 = prob_exactly_15,
  observed_at_least_15 = mean(simulated_survivors >= 15),
  model_at_least_15 = prob_at_least_15
)
```


# Exercise 2: Bill length under a Normal model

The data set `penguins`, which is built into R 4.5.0 and later^[With an older version of R,
install the package `palmerpenguins` and load the data with
`data(penguins, package = "palmerpenguins")`. It contains the same measurements, but the
column with the bill length is called `bill_length_mm`.], contains measurements of adult
penguins of three species near Palmer Station, Antarctica (Gorman, Williams and Fraser
2014). One of the measurements is the length of the bill (the column `bill_len`, in
millimetres). The theory used the flipper length of the Gentoo penguins. Here we use the
bill length of the 68 Chinstrap penguins (*Pygoscelis antarcticus*), which is well
described by a Normal distribution with a mean of 49 mm and a standard deviation of 3.3 mm
([Chapter 3](../../Chapter_3/Theory.qmd) explains how such values are obtained from a
sample).

Suppose the bill length $X$ of a Chinstrap penguin follows that Normal model:

$$
X\sim\operatorname{Normal}(49,3.3^2).
$$

1. Calculate the probability that a bill is between 45 mm and 52 mm long.
2. Find the length below which 95% of bills fall under this model.
3. Simulate 1,000 bill lengths. Compare a density-scaled histogram with the
   model density and explain any differences.

## Solution

Suppose the bill length $X$ of a Chinstrap penguin follows a Normal model with
mean 49 mm and standard deviation 3.3 mm.

The observational unit is one penguin and bill length is a continuous random variable, so
we calculate probabilities for intervals of length. Note that the model above is written
with the variance, $3.3^2$, whereas the `sd` argument in R takes the standard deviation,
3.3 mm.

In R, the Normal distribution uses:

- `dnorm()` for probability density;
- `pnorm()` for cumulative probabilities;
- `qnorm()` for quantiles;
- `rnorm()` for simulation.

```{r}
# Parameters of the bill length model (in millimetres)
bill_mean = 49
bill_sd = 3.3
```

### 1. Probability that a bill is between 45 mm and 52 mm long

The probability of an interval is the cumulative probability at its upper limit minus the
cumulative probability at its lower limit:

$$
P(45<X\leq52)=P(X\leq52)-P(X\leq45).
$$

In R, `pnorm()` calculates $P(X\leq x)$, so we subtract two calls:

```{r}
# Subtract the probability below 45 mm from the probability below 52 mm
prob_45_to_52 = pnorm(q = 52, mean = bill_mean, sd = bill_sd) -
  pnorm(q = 45, mean = bill_mean, sd = bill_sd)

prob_45_to_52
```

Thus, the probability that a bill is between 45 mm and 52 mm long is approximately
**0.71**, or **71%**. Including or excluding the endpoints makes no difference, because the
probability of any exact length is zero in a continuous model.

### 2. Length below which 95% of bills fall

We want the length $x$ such that

$$
P(X\leq x)=0.95.
$$

This is the 95th percentile (i.e., the 0.95 quantile) of the distribution. In R, `qnorm()`
calculates it:

```{r}
# Quantile of the model at a cumulative probability of 0.95
bill_95 = qnorm(p = 0.95, mean = bill_mean, sd = bill_sd)

bill_95
```

The length is approximately **54.4 mm**. That is, under this model 95% of bills are shorter
than 54.4 mm and 5% are longer.

### 3. Simulating 1,000 bill lengths

We can simulate 1,000 bill lengths using `rnorm()`:

```{r}
# Set a seed so that the simulated sample is reproducible
set.seed(seed = 101)

bill_lengths = rnorm(n = 1000, mean = bill_mean, sd = bill_sd)

head(bill_lengths)
```

Each value in `bill_lengths` is the bill length of one simulated penguin. We can now plot a
histogram of the sample and add the density of the model with `curve()`:

```{r}
# Histogram on the density scale (probability = TRUE) so that it is comparable to the curve
hist(
  x = bill_lengths, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Bill length (mm)", ylab = "Density per millimetre",
  main = "Simulated bill lengths and Normal model"
)

# Add the density of the model that generated the sample
curve(
  dnorm(x = x, mean = bill_mean, sd = bill_sd),
  add = TRUE, lwd = 2
)

legend(
  "topright", legend = "Normal density",
  lty = 1, lwd = 2, bty = "n"
)
```

The histogram uses density scaling because the curve also shows density (with counts on
the vertical axis the two would not be comparable). The bars do not match the curve
exactly because they summarise one sample of 1,000 simulated penguins, and their heights
also depend on the width of the bins. With a larger number of simulated penguins, the
histogram would usually become closer to the model density.

# Exercise 3: Checking for overdispersion

We are going to use a built-in dataset in R on counts of tobacco hornworm, in agricultural
experimental units treated with six different insecticides. We can load the data as follows:

```{r}
#| eval: false
data(InsectSprays, package = "datasets")
```

1. Calculate the mean and variance for `count`. Is there evidence of overdispersion?

2. Repeat the same calculations but now do it separately for each `spray` treatment (tip: use
the tools for data transformation we learnt in Chapter 1). Is there still evidence of overdispersion?

3. Which model would you use for `count` when ignoring the `spray` treatment and which model would
you try when doing it separately for each `spray` treatment?

# Exercise 4: Overdispersed quadrat counts

Let's assume a Negative Binomial distribution with a mean ($\mu = 6$) and shape parameter
($k = 0.8$). The formulae and R functions are in the
[Negative Binomial entry](../../Supplements/distributions.qmd#sec-negative_binomial) of the supplement.

1. Calculate the mean and variance of the model using the formulas. Then verify them
using `dnbinom()` and weighted sums.
2. Generate a sample of size 500.
3. Calculate the mean and variance of this sample in R.
4. Compare the Negative Binomial distribution with a Poisson distribution having the same mean.
How do they differ? What is the variance/mean ratio of the sample? Plot their probability mass functions.

# Exercise 5: Body mass under a LogNormal model

Suppose body mass $X$ in grams follows a LogNormal model with `meanlog = log(30)` and
`sdlog = 0.5`. These two parameters are the mean and the standard deviation of the
logarithm of body mass. The formulae for the median and the mean are in the
[LogNormal entry](../../Supplements/distributions.qmd#sec-lognormal) of the supplement.

1. Calculate the median and the mean body mass of the model. Explain why they differ.
2. Calculate the probability that body mass exceeds 50 g and the 95th percentile.
3. Simulate 500 body masses with seed 202. Make density-scaled histograms on the
   original and log scales. Describe their shapes.
4. Add the LogNormal density of the model to the original-scale histogram.
   Explain what the comparison can and cannot tell you about field data.

# Exercise 6: Vegetation cover under a Beta model

Suppose $X$ is the fraction of a quadrat covered by vegetation, measured as a continuous
variable, and that $X$ follows a Beta model with $a=2$ and $b=5$. In R these
parameters are `shape1` and `shape2`. The formulae for the mean and the variance are
in the [Beta entry](../../Supplements/distributions.qmd#sec-beta) of the supplement.

1. Calculate the mean and the variance of the model. Express the mean as a percentage of
   the quadrat.
2. Calculate $P(X<0.20)$ and the 95th percentile of cover. Explain both results.
3. Simulate 500 cover fractions with seed 303. Plot a density-scaled histogram
   and add the Beta density of the model.
4. A field survey also records some quadrats with exactly zero cover. Explain
   why the proposed Beta model cannot describe those observations directly.
