---
title: "Introduction to Probability (Discrete)"
---

# Exercise 1: Seedling survival

Assume that each of 25 seedlings has probability $0.65$ of surviving. Assume a Binomial
model to describe this process.

1. Calculate the probability that exactly 15 survive.
2. Calculate the probability that at least 15 survive.
3. Calculate the expected number and variance of survivors.
4. Simulate 1,000 groups of 25 seedlings and compare the observed frequencies with the model
probabilities.

## Solution

Assume that each of 25 seedlings has probability $0.65$ of surviving. Assume a Binomial
model to describe this process.

Let

$$
X=\text{the number of seedlings that survive}.
$$

The process has:

- a fixed number of trials, $n=25$;
- two possible outcomes for each seedling: survives or does not survive;
- the same survival probability, $p=0.65$;
- independent seedlings.

Therefore,

$$
X\sim\operatorname{Binomial}(n=25,p=0.65).
$$

In R, the Binomial distribution uses:

- `dbinom()` for probability mass;
- `pbinom()` for cumulative probabilities;
- `rbinom()` for simulation.

```{r}
n = 25
p_survive = 0.65
```

### 1. Probability that exactly 15 survive

We want to calculate

$$
P(X=15).
$$

The Binomial probability mass function is

$$
P(X=x)=
\binom{n}{x}p^x(1-p)^{n-x}.
$$

Substituting $x=15$, $n=25$, and $p=0.65$ gives

$$
\begin{aligned}
P(X=15)
&=
\binom{25}{15}(0.65)^{15}(1-0.65)^{25-15}\\
&=
\binom{25}{15}(0.65)^{15}(0.35)^{10}\\
&=
\frac{25!}{15!10!}(0.65)^{15}(0.35)^{10}\\
&\approx 0.14.
\end{aligned}
$$

Thus, the probability that exactly 15 seedlings survive is approximately **0.14**, or **14%**.

In R, `dbinom()` performs this calculation:

```{r}
prob_exactly_15 = dbinom(
  x = 15,
  size = n,
  prob = p_survive
)

prob_exactly_15
```

The combinatorial term can also be calculated directly in R:

```{r}
choose(25, 15)
```

### 2. Probability that at least 15 survive

“At least 15” means 15 or more:

$$
P(X\geq 15)=P(X=15)+P(X=16)+\cdots+P(X=25).
$$

By using the complement,

$$
P(X\geq 15)
=
1-P(X\leq 14).
$$

Using the Binomial formula, this is

$$
P(X\geq 15)
=
1-\sum_{x=0}^{14}
\binom{25}{x}(0.65)^x(0.35)^{25-x}.
$$

Evaluating this sum gives

$$
P(X\geq 15)\approx 0.77.
$$

Therefore, the probability that at least 15 seedlings survive is approximately **0.77**, or **77%**.

In R, `pbinom()` calculates $P(X\leq x)$ so we take the complement:

```{r}
prob_at_least_15 = 1 - pbinom(
  q = 14,
  size = n,
  prob = p_survive
)

prob_at_least_15
```

The value `q = 14` is used because

$$
P(X>14)=P(X\geq 15).
$$

### 3. Expected number and variance of survivors

For a Binomial random variable,

$$
\mu=E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$

The expected number of survivors is therefore

$$
\begin{aligned}
E[X]
&=25(0.65)\\
&=16.25.
\end{aligned}
$$

The variance is

$$
\begin{aligned}
\operatorname{Var}(X)
&=25(0.65)(1-0.65)\\
&=25(0.65)(0.35)\\
&=5.69.
\end{aligned}
$$

In R:

```{r}
expected_survivors = n * p_survive

variance_survivors = n * p_survive * (1 - p_survive)

c(
  expected = expected_survivors,
  variance = variance_survivors
)
```

The expected value, 16.25, does not need to be an integer. It represents the long-run
average number of survivors across many groups of 25 seedlings.

### 4. Simulating 1,000 groups

We can simulate 1,000 groups using `rbinom()`:

```{r}
set.seed(123)

simulated_survivors = rbinom(
  n = 1000,
  size = n,
  prob = p_survive
)

head(simulated_survivors)
```

Each value in `simulated_survivors` is the number of survivors in one simulated group.

To calculate the observed relative frequency for every possible number of survivors, use `table()`:

```{r}
support = 0:n

observed_frequency = as.numeric(
  table(factor(simulated_survivors, levels = support))
) / length(simulated_survivors)

model_probability = dbinom(
  x = support,
  size = n,
  prob = p_survive
)

comparison = data.frame(
  survivors = support,
  observed = observed_frequency,
  theoretical = model_probability
)

round(comparison, 4)
```

The `factor()` step ensures that all possible values from 0 to 25 are included, even if a
value does not occur in the simulation.

We can compare the observed and theoretical probabilities graphically:

```{r}
barplot(
  rbind(observed_frequency, model_probability),
  beside = TRUE,
  names.arg = support,
  col = c("grey70", "steelblue"),
  xlab = "Number of seedlings surviving",
  ylab = "Probability",
  legend.text = c("Observed frequency", "Theoretical probability"),
  args.legend = list(x = "topright"),
  main = "Observed and theoretical seedling survival"
)
```

The observed frequencies will not exactly equal the model probabilities because only 1,000
groups were simulated. They should generally be similar. With a larger number of simulated
groups, the observed frequencies would usually become closer to the theoretical probabilities.

We can also compare the simulated and theoretical values for the first two questions:

```{r}
c(
  observed_exactly_15 = mean(simulated_survivors == 15),
  theoretical_exactly_15 = prob_exactly_15,
  observed_at_least_15 = mean(simulated_survivors >= 15),
  theoretical_at_least_15 = prob_at_least_15
)
```

# Exercise 2: Checking for overdispersion

We are going to use a built-in dataset in R on counts of tobacco hornworm, in agricultural
experimental units treated with six different insecticides. We can load the data as follows:

```{r}
#| eval: false
data(InsectSprays, package = "datasets")
```

1. Calculate the mean and variance for `count`. Is there evidence of overdispersion?

2. Repeat the same calculations but not do it separately for each `spray` treatment (tip: use
the tools for data transformation we learnt in Chapter 1). Is there still evidence of overdispersion?

3. Which model would you use `count` when ignoring the `spray` treatment and which model would
you try when doing it separately for each `spray` treatment?

4. Calculate the parameters of the negative binomial using the mean and the variance of the
sample (tip: check the formulae in the theory chapter). Compare the probability mass of the
empirical distribution and the negative binomial with the parameters that you just calculated.
How well does the model capture the variation in the data?

# Exercise 3: Overdispersed quadrat counts

Let's assume a negative binomial distribution with a mean ($\mu = 6$) and shape parameters
($k = 0.8$).

1. Calculate the theoretical mean and variance using the formulas. Then verify them
using dnbinom() and weighted sums.
2. Generate a sample of size 500.
3. Calculate the mean and variance of this sample in R.
4. Compare the negative binomial distribution with a Poisson distribution having the same mean.
How do they differ? What is the variance/mean ratio of the sample? Plot their probability mass functions.


# Exercise 4: Asymptotic convergence

Simulate Poisson observations with $\lambda=3$ for two sample sizes: 10 and 1,000.

1. Calculate the mean, median and variance for each sample (use the actual variance of the
sample, not `var()`).
2. Repeat point 1 four times using different random seeds (tip: use a foor loop). How much
do the sample properties vary?
3. Compare these sample properties to the population values at the two sample sizes. How does
the median and mean differ in their behavior? What is happening to the variation across
repeated samples as sample size increases?
