---
title: "Probability distributions - Solutions"
---

The fully worked Exercises 1 and 2 are on the
[student practical page](no_solution.qmd). These solutions follow the same
assumed models, tasks, and seeds as Exercises 3–6 on that page.

# Exercise 3: Checking for overdispersion

We are going to use a built-in dataset in R on counts of tobacco hornworm, in agricultural
experimental units treated with six different insecticides. We can load the data as follows:

```{r}
#| eval: false
data(InsectSprays, package = "datasets")
```

1. Calculate the mean and variance for `count`. Is there evidence of overdispersion?

2. Repeat the same calculations but now do it separately for each `spray` treatment (tip: use
the tools for data transformation we learnt in Chapter 1). Is there still evidence of overdispersion?

3. Which model would you use for `count` when ignoring the `spray` treatment and which model would
you try when doing it separately for each `spray` treatment?

4. Calculate the parameters of the Negative Binomial using the mean and the variance of the
sample (tip: check the formulae in the
[Negative Binomial entry](../../Supplements/distributions.qmd#sec-negative_binomial) of the supplement). Compare the probability mass of the
empirical distribution and the Negative Binomial with the parameters that you just calculated.
How well does the model capture the variation in the data?

## Solution

### Question 1

Load the data and compare the sample variance with the mean. For a Poisson distribution, the
mean and variance are equal, so a variance much larger than the mean suggests overdispersion.

```{r}
data(InsectSprays, package = "datasets")

pooled_mean = mean(InsectSprays$count)
pooled_variance = var(InsectSprays$count)

c(
  mean = pooled_mean,
  variance = pooled_variance,
  dispersion_ratio = pooled_variance / pooled_mean
)
```

The mean is about 9.5, the variance is 51.9, and the variance-to-mean ratio is 5.5. This is
substantial overdispersion when the treatments are pooled.

### Question 2

Now calculate the summaries separately for each spray treatment:

```{r}
library(dplyr)

spray_groups = group_by(InsectSprays, spray)

spray_summary = summarise(
  spray_groups,
  n = n(),
  mean = mean(count),
  variance = var(count),
  dispersion_ratio = variance / mean,
  .groups = "drop"
)

spray_summary
```

Five of the six treatments have a variance-to-mean ratio greater than one, though the amount
of overdispersion is much smaller than when pooled. Notice how a big part of the overdispersion
could be explained by the fact that we were looking at different treatments pooled together.

### Question 3

Ignoring spray, a **Negative Binomial model** is a reasonable choice because the pooled
counts are strongly overdispersed. When accounting for spray, a Poisson model with spray as
a predictor is a useful starting point, though we might still end up using a Negative Binomial
(we will cover formal methods of model comparison later in the book).

### Question 4

For the Negative Binomial parameterisation used in R,

$$
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k},
$$

so the shape parameter is

$$
k=\frac{\mu^2}{\operatorname{Var}(X)-\mu}.
$$

To match the empirical distribution’s variance, calculate the variance of the sample. To emphasise
this I avoid using `var()` which is actually a statistical estimator and not the variance
of the sample (but in practice it is not a critical issue unless $n$ is very small, so do
not worry too much about it):

```{r}
counts = InsectSprays$count
mu_hat = mean(counts)
empirical_variance = mean((counts - mu_hat)^2)

k_hat = mu_hat^2 / (empirical_variance - mu_hat)

c(
  mean = mu_hat,
  empirical_variance = empirical_variance,
  k = k_hat
)
```

The empirical variance is about **51.2**, giving an estimated shape parameter of about **2.2**.

Compare the empirical probability mass with that of the estimated Negative Binomial:

```{r}
support = 0:max(counts)

empirical_mass = as.numeric(
  table(factor(counts, levels = support))
) / length(counts)

negative_binomial_mass = dnbinom(
  support,
  size = k_hat,
  mu = mu_hat
)

plot(
  support,
  empirical_mass,
  type = "h",
  lwd = 5,
  col = "grey65",
  xlab = "Count",
  ylab = "Probability",
  main = "Empirical and Negative Binomial probability mass"
)

lines(
  support,
  negative_binomial_mass,
  type = "b",
  pch = 16,
  col = "steelblue"
)

legend(
  "topright",
  legend = c("Empirical", "Negative Binomial"),
  col = c("grey65", "steelblue"),
  lty = 1,
  pch = c(NA, 16),
  lwd = c(5, 1)
)
```

The estimated Negative Binomial matches the empirical mean and variance by construction, and
allows more variation than a Poisson model. However, pooling treatments combines groups with
very different count levels. A single Negative Binomial distribution may therefore not capture
all the patterns in the empirical distribution and it would be better to account for the
structure of the data explicitly (we will do that in Chapter 8).


# Exercise 4: Overdispersed quadrat counts

Let's assume a Negative Binomial distribution with a mean ($\mu = 6$) and shape parameter
($k = 0.8$). The formulae and R functions are in the
[Negative Binomial entry](../../Supplements/distributions.qmd#sec-negative_binomial) of the supplement.

1. Calculate the theoretical mean and variance using the formulas. Then verify them
using `dnbinom()` and weighted sums.
2. Generate a sample of size 500.
3. Calculate the mean and variance of this sample in R.
4. Compare the Negative Binomial distribution with a Poisson distribution having the same mean.
How do they differ? What is the variance/mean ratio of the sample? Plot their probability mass functions.

## Solution

### Question 1

For a Negative Binomial distribution with $\mu=6$ and $k=0.8$,

$$
E[X]=\mu=6
$$

and

$$
\operatorname{Var}(X)
=
\mu+\frac{\mu^2}{k}
=
6+\frac{6^2}{0.8}
=
51.
$$

We can verify these values using the probability mass function and weighted sums. The
distribution’s support is infinite; using values from 0 to 1,000 makes the omitted tail
negligible.

```{r}
mu = 6
k = 0.8

x_theory = 0:1000
probability = dnbinom(x_theory, size = k, mu = mu)

c(
  probability_sum = sum(probability),
  weighted_mean = sum(x_theory * probability),
  weighted_variance = sum((x_theory - mu)^2 * probability)
)
```

### Question 2

Generate a sample of size 500, then calculate its mean and empirical variance. The empirical variance uses the divisor $n$, as in the empirical distribution.

```{r}
set.seed(123)

sample_counts = rnbinom(
  n = 500,
  size = k,
  mu = mu
)

sample_mean = mean(sample_counts)
sample_variance = mean((sample_counts - sample_mean)^2)

c(
  mean = sample_mean,
  empirical_variance = sample_variance,
  variance_mean_ratio = sample_variance / sample_mean,
  variance_using_var = var(sample_counts)
)
```

### Question 3

The sample summaries vary between simulations, but should be broadly consistent with a mean of 6 and a variance of 51. The theoretical variance-to-mean ratio is $51/6=8.5$; the sample ratio will not necessarily equal it.

### Question 4

A Poisson distribution with the same mean has variance 6, while the Negative Binomial has variance 51. The Negative Binomial therefore has more probability near zero and in the upper tail.

```{r}
x = 0:40

negative_binomial_mass = dnbinom(x, size = k, mu = mu)
poisson_mass = dpois(x, lambda = mu)

matplot(
  x,
  cbind(negative_binomial_mass, poisson_mass),
  type = "b",
  pch = c(16, 1),
  lty = c(1, 2),
  lwd = 2,
  col = c("firebrick", "steelblue"),
  xlab = "Count",
  ylab = "Probability",
  main = "Negative Binomial and Poisson probability mass"
)

legend(
  "topright",
  legend = c("Negative Binomial", "Poisson"),
  col = c("firebrick", "steelblue"),
  pch = c(16, 1),
  lty = c(1, 2),
  lwd = 2
)
```


# Exercise 5: Body mass under a LogNormal model

Suppose body mass $X$ in grams follows a LogNormal model with `meanlog = log(30)` and
`sdlog = 0.5`. These two parameters are the mean and the standard deviation of the
logarithm of body mass. The formulae for the median and the mean are in the
[LogNormal entry](../../Supplements/distributions.qmd#sec-lognormal) of the supplement.

1. Calculate the median and the mean body mass of the model. Explain why they differ.
2. Calculate the probability that body mass exceeds 50 g and the 95th percentile.
3. Simulate 500 body masses with seed 202. Make density-scaled histograms on the
   original and log scales. Describe their shapes.
4. Add the LogNormal density of the model to the original-scale histogram.
   Explain what the comparison can and cannot tell you about field data.

## Solution

### Question 1

The parameters of the LogNormal distribution describe $\log(X)$: `meanlog` is its mean,
$\mu$, and `sdlog` is its standard deviation, $\sigma$. On the original scale, the median
and the mean of body mass are

$$
\operatorname{median}(X)=e^{\mu}
$$

and

$$
E[X]=e^{\mu+\sigma^2/2}.
$$

In R:

```{r}
# Parameters of the model on the log scale
mass_meanlog = log(30)
mass_sdlog = 0.5

# Median and mean of body mass on the original scale (in grams)
median_mass = exp(mass_meanlog)
mean_mass = exp(mass_meanlog + mass_sdlog^2 / 2)

c(median_g = median_mass, mean_g = mean_mass)
```

The median is 30 g and the mean is approximately 34 g. They differ because the distribution
is skewed to the right: the large body masses in the upper tail pull the mean above the
median.

### Question 2

`plnorm()` calculates the cumulative probability $P(X\leq x)$. For a body mass above 50 g
we need the upper tail, which we obtain with `lower.tail = FALSE`. `qlnorm()` calculates the
mass below which 95% of the probability of the model falls.

```{r}
# Probability of a body mass above 50 g (upper tail)
prob_over_50 = plnorm(
  q = 50, meanlog = mass_meanlog, sdlog = mass_sdlog,
  lower.tail = FALSE
)

# Body mass at a cumulative probability of 0.95
mass_95 = qlnorm(p = 0.95, meanlog = mass_meanlog, sdlog = mass_sdlog)

c(prob_over_50 = prob_over_50, percentile_95_g = mass_95)
```

The probability that body mass exceeds 50 g is approximately **0.15**, and the 95th
percentile is approximately **68.3 g**. That is, under this model about 15% of the animals
are heavier than 50 g and 5% are heavier than 68.3 g.

### Question 3

Generate the sample and plot a histogram of the body masses and a histogram of their
logarithms. Both use density scaling (`probability = TRUE`):

```{r}
# Simulate 500 body masses (in grams) with the seed given in the exercise
set.seed(seed = 202)
body_mass = rlnorm(n = 500, meanlog = mass_meanlog, sdlog = mass_sdlog)

# Draw the two histograms side by side and restore the graphics settings afterwards
old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))

hist(
  x = body_mass, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Body mass (g)", ylab = "Density per gram",
  main = "Original scale"
)

hist(
  x = log(body_mass), breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Log body mass", ylab = "Density",
  main = "Log scale"
)

par(old_par)
```

On the original scale the body masses are all positive and skewed to the right, with a few
animals much heavier than the rest. Their logarithms are approximately symmetric, because
the logarithm of a LogNormal random variable follows a Normal distribution.

### Question 4

Plot the original-scale histogram again and add the density of the model with `curve()`:

```{r}
hist(
  x = body_mass, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Body mass (g)", ylab = "Density per gram",
  main = "Simulated body masses and LogNormal model"
)

# Add the density of the model that generated the sample
curve(
  dlnorm(x = x, meanlog = mass_meanlog, sdlog = mass_sdlog),
  add = TRUE, lwd = 2
)
```

The histogram follows the curve but does not match it exactly. Since the sample was
generated from this model, the differences show how much a histogram of 500 values varies
around the density of its model. To learn whether a LogNormal model describes body masses
measured in the field, we would have to compare the model with those measurements.

# Exercise 6: Vegetation cover under a Beta model

Suppose $X$ is the fraction of a quadrat covered by vegetation, measured as a continuous
variable, and that $X$ follows a Beta model with $a=2$ and $b=5$. In R these
parameters are `shape1` and `shape2`. The formulae for the mean and the variance are
in the [Beta entry](../../Supplements/distributions.qmd#sec-beta) of the supplement.

1. Calculate the mean and the variance of the model. Express the mean as a percentage of
   the quadrat.
2. Calculate $P(X<0.20)$ and the 95th percentile of cover. Explain both results.
3. Simulate 500 cover fractions with seed 303. Plot a density-scaled histogram
   and add the Beta density of the model.
4. A field survey also records some quadrats with exactly zero cover. Explain
   why the proposed Beta model cannot describe those observations directly.

## Solution

### Question 1

For a Beta distribution with shape parameters $a=2$ and $b=5$,

$$
E[X]=\frac{a}{a+b}=\frac{2}{7}\approx 0.286
$$

and

$$
\operatorname{Var}(X)
=
\frac{ab}{(a+b)^2(a+b+1)}
=
\frac{10}{392}
\approx 0.026.
$$

In R:

```{r}
# Shape parameters of the Beta model
cover_a = 2
cover_b = 5

# Mean and variance of the cover fraction
mean_cover = cover_a / (cover_a + cover_b)
variance_cover = cover_a * cover_b /
  ((cover_a + cover_b)^2 * (cover_a + cover_b + 1))

c(mean_fraction = mean_cover,
  mean_percent = 100 * mean_cover,
  variance = variance_cover)
```

The expected cover is approximately 0.286, or 28.6% of a quadrat. The variance is
approximately 0.026 and, like the cover fraction, it has no units.

### Question 2

`pbeta()` calculates the cumulative probability and `qbeta()` the quantile:

```{r}
# Probability of a cover fraction below 0.20
prob_below_20_percent = pbeta(
  q = 0.20, shape1 = cover_a, shape2 = cover_b
)

# Cover fraction at a cumulative probability of 0.95
cover_95 = qbeta(p = 0.95, shape1 = cover_a, shape2 = cover_b)

c(prob_below_20_percent = prob_below_20_percent,
  percentile_95_fraction = cover_95)
```

The first result is the probability that a quadrat has less than 20% cover, which is
approximately **0.34**. The second is the cover fraction below which 95% of the quadrats
fall under this model, approximately **0.58** (58% cover). Note that using $<$ or $\leq$
at 0.20 gives the same probability, because cover is a continuous random variable.

### Question 3

```{r}
# Simulate 500 cover fractions with the seed given in the exercise
set.seed(seed = 303)
cover_fraction = rbeta(n = 500, shape1 = cover_a, shape2 = cover_b)

# Histogram on the density scale so that it is comparable to the curve
hist(
  x = cover_fraction, breaks = 25, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Vegetation cover (fraction)", ylab = "Density",
  main = "Simulated cover and Beta model"
)

# Add the density of the model that generated the sample
curve(
  dbeta(x = x, shape1 = cover_a, shape2 = cover_b),
  add = TRUE, lwd = 2
)
```

The histogram shows one simulated sample, so its bars follow the density of the model
without matching it exactly.

### Question 4

The Beta distribution is continuous, so it assigns zero probability to any exact value,
including a cover of exactly zero or exactly one. A sample from this model will therefore
never contain a quadrat with zero cover, however large the sample is. Before modelling
field observations at those boundaries, we would ask whether they represent quadrats that
are really bare (or fully covered) and how they arose, and then extend the model to account
for them (see the section on
[zero-inflated distributions](../../Supplements/distributions.qmd#sec-zero_inflated) in the
supplement).
