---
title: "Fitting ecological response curves"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In this chapter, we learn how to fit an ecological response curve together
with a probability model for observations around that curve. This lets us use
maximum likelihood to estimate the curve's parameters, describe their
uncertainty, compare plausible models, and check whether the fitted model can
reproduce key features of the data.

In [Chapter 5](../Chapter_5/Theory.qmd), we studied the
[Michaelis–Menten](../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten)
function, known as the Holling type II functional response when it describes
predation. Here, we use it for the expected intake rate of a predator
consuming prey at different prey densities. Real intake measurements
will not all fall on the expected curve, so we need an observation model as
well as a response function. We write that likelihood and fit it with
`MakeADFun()` and `optim()`, as we have done since
[Chapter 4](../Chapter_4/Theory.qmd). We then say more about two things that
RTMB has been doing for us since that chapter: it calculates the slope of the
likelihood, and the standard error of a quantity calculated from the
parameters (the delta method). This also prepares us to add random effects to
nonlinear curves in Chapter 7.

The supplied practical datasets are simulated ecological examples. Their
generating models and parameter values are not given when you analyse them.

# Learning goals

After studying this chapter, you should be able to:

- specify a response function, observation distribution, parameter domain,
  and observational unit for a single-predictor dataset;
- write and optimise the resulting log-likelihood, check convergence, and
  interpret estimates in ecological units;
- explain what the gradient of a likelihood is, why an accurate gradient from
  RTMB helps when fitting curves, and recognise when automatic differentiation
  may be unsuitable;
- calculate and interpret approximate confidence intervals, explain how the
  delta method gives the standard error of a quantity calculated from the
  parameters, then use a simulation study to investigate bias and coverage;
- compare models fitted to the same observations using information criteria
  and held-out predictive performance;
- simulate from a fitted model and interpret scaled quantile residuals.

# From a curve to a model for observations {#sec-response-observation}

For one intake-rate measurement at a specified prey density, let $x_i$ be the
recorded prey density and let $Y_i$ be the recorded intake rate. The supplied
data use recorded units, so replace these labels with physical units when they
are available. A possible expected response is

$$
\mu_i=m(x_i;a,b)=\frac{a x_i}{b+x_i}, \qquad x_i\geq0,\ a,b>0.
$$

The parameter $a$ is the limiting intake rate; $b$ is the prey density at
half that rate. The function returns one expected rate for each density.
It does not say how far an observed $Y_i$ might lie from that expectation.

If rates vary approximately symmetrically around the curve, a first
observation model is

$$
Y_i\mid x_i\sim\operatorname{Normal}(\mu_i,\sigma^2),
\qquad \sigma>0.
$$

Here $\sigma$ describes variation in individual intake measurements, in the
same units as the rate. We also assume measurements are independent after
conditioning on their predictor values. With these assumptions, the model has
three unknown parameters: $a$, $b$, and $\sigma$. Because a normal distribution
also permits negative rates, check whether that part of its support matters
near low prey density.

::: {.callout-note title="Mean curve and observations"}
The curve describes $E[Y_i\mid x_i]$. The distribution describes possible
observations at that $x_i$. A close-looking curve alone does not specify a
likelihood or show that the chosen distribution is suitable.
:::

The points in this illustration were simulated from a known curve with
normal variation. Their spread shows why a likelihood needs both parts:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Simulated predator intake measurements around a Holling type II mean curve."
# Simulate four intake measurements at each prey density.
set.seed(seed = 710)
prey_density = rep(seq(from = 3, to = 30, by = 3), each = 4)
expected_intake = 12 * prey_density / (6 + prey_density)
observed_intake = rnorm(n = length(prey_density),
                        mean = expected_intake, sd = 1)
# Plot the observations and add their generating mean curve.
plot(prey_density, observed_intake,
     xlab = "Prey density (illustrative units)",
     ylab = "Intake rate (illustrative units)", pch = 16)
curve(12 * x / (6 + x), from = 0, to = 30, add = TRUE, lwd = 2)
legend("bottomright", legend = c("Simulated observations", "Mean curve"),
       pch = c(16, NA), lty = c(NA, 1), bty = "n")
```

The response determines which distributions are possible. A count of cones
can use a Poisson model with a positive mean curve. Presence or absence can
use a Bernoulli model with a probability curve between zero and one:

| Response | Example conditional model | What the curve must return |
|---|---|---|
| Continuous intake rate | $Y_i\mid x_i\sim\operatorname{Normal}(\mu_i,\sigma^2)$ | A mean in rate units |
| Number of cones | $Y_i\mid x_i\sim\operatorname{Poisson}(\mu_i)$ | A positive expected count |
| Presence or absence | $Y_i\mid x_i\sim\operatorname{Bernoulli}(p_i)$ | A probability in $[0,1]$ |

[Chapter 5](../Chapter_5/Theory.qmd#sec-function-composition) shows how to
give a curve the range that each of these responses requires, by composing it
with an exponential or an inverse logit function.

The observation model is a scientific assumption, not a label determined by
the response variable alone. Counts can be more variable than a Poisson
model allows; continuous variation can be asymmetric. Later diagnostics
help us find such disagreements.

:::: {.callout-important title="Try it: identify the random part"}
Suppose $p(x)=1/[1+e^{-k(x-c)}]$ predicts species presence. Which part of
the model says that two sites with the same $x$ may have different outcomes?

::: {.callout-tip title="Answer" collapse="true"}
The Bernoulli observation model does. The function gives the same $p(x)$
at both sites, while each presence/absence observation remains random.
:::
::::

# Fit a response curve by maximum likelihood {#sec-fit-response}

For independent normal observations, the log-likelihood is the sum of the
normal log densities at the observed rates. For $n$ intake measurements,
$f_{\mathrm N}$ denotes the normal density:

$$
\ell(a,b,\sigma)=\sum_{i=1}^n
\log f_{\mathrm N}\{y_i;m(x_i;a,b),\sigma\}.
$$

Here $y_i$ is the observed value of $Y_i$. As in
[Chapter 4](../Chapter_4/Theory.qmd), we write the negative log-likelihood
(NLL) with two named lists, connect it to the data with `cmb()`, process it
with `MakeADFun()` and minimise it with `optim()` and BFGS. We optimise the
logarithms of positive parameters and transform the estimates back afterwards.
The two `ADREPORT()` lines ask for the standard errors of two quantities
calculated from the parameters, which we use in @sec-delta-method.

The supplied predator data can be downloaded as `shapes2.csv` from the
[`Data` folder](https://github.com/AleMorales/EcologicalModelsData2026/tree/main/Data)
in the course GitHub repository. In this Quarto website, the file is one
directory above this theory page:

```{r}
library(RTMB)
library(RTMBdist)

# Load the supplied simulated intake data.
intake_data = read.csv("../Data/shapes2.csv")

# Define the Holling type II mean curve.
intake_curve = function(x, a, b) {
  a * x / (b + x)
}

# Negative log-likelihood of Normal observations around the curve.
intake_nll = function(parms, data) {
  # Make the elements of both lists available by name.
  getAll(parms, data)
  # Transform the parameters to the original scale.
  a = exp(log_a)
  b = exp(log_b)
  sigma = exp(log_sigma)
  mu = intake_curve(x = x, a = a, b = b)
  # Expected intake rate at a prey density of 10.
  intake_10 = intake_curve(x = 10, a = a, b = b)
  # Ask for the standard errors of a and of the expected intake rate.
  ADREPORT(a)
  ADREPORT(intake_10)
  -sum(dnorm(x = y, mean = mu, sd = sigma, log = TRUE))
}

# Connect a likelihood to a data set.
cmb = function(f, d) function(p) f(p, d)

# Process the NLL, with positive starting values on the log scale.
intake_obj = MakeADFun(
  cmb(intake_nll, list(x = intake_data$x, y = intake_data$y)),
  list(log_a = log(max(intake_data$y)),
       log_b = log(median(intake_data$x)),
       log_sigma = log(sd(intake_data$y))),
  silent = TRUE
)

# Minimise the NLL with BFGS.
intake_fit = optim(par = intake_obj$par, fn = intake_obj$fn,
                   gr = intake_obj$gr, method = "BFGS")

# Check convergence, then return estimates on their ecological scales.
intake_fit$convergence
estimates = c(a = exp(intake_fit$par[["log_a"]]),
              b = exp(intake_fit$par[["log_b"]]),
              sigma = exp(intake_fit$par[["log_sigma"]]))
estimates
```

Check the convergence code, the fitted curve against the points, and whether
the estimates make ecological sense. Try another plausible starting value
if the fit is unstable. A convergence code of zero is reassuring, but it
does not establish that the model describes the observations well.
[Chapter 5](../Chapter_5/Theory.qmd#sec-starting-values) shows how to read
starting values from a plot of the data, using the landmarks of the function.

## `MakeADFun()` and automatic differentiation {#sec-makeadfun}

Since [Chapter 4](../Chapter_4/Theory.qmd#sec-maximum-likelihood) we have given
every NLL to `MakeADFun()` before fitting it. There we said only that this step
processes the function so that the tools of the course can work with it, and
that RTMB also supplies the slope that BFGS uses. We can now say more about
that slope.

The **gradient** of the NLL is the vector of its partial derivatives with
respect to the parameters, that is, its slope in the direction of each
parameter ([Chapter 5](../Chapter_5/Theory.qmd#sec-parameter-derivatives)).
`MakeADFun()` processes the NLL in such a way that RTMB can calculate this
gradient accurately at any value of the parameters, with a method called
**automatic differentiation**. We do not have to derive any formula for it. The
element `gr` of the object returns the gradient, with one value per parameter,
and at the estimates every one of them is close to zero, as it should be at a
minimum:

```{r}
# The gradient of the NLL at the starting values and at the estimates.
intake_obj$gr(intake_obj$par)
intake_obj$gr(intake_fit$par)
```

An accurate gradient helps when fitting curves for three reasons. First, BFGS
needs the gradient at every step to decide in which direction to move, and
without `gr` the optimiser has to approximate it by evaluating the NLL several
more times for every parameter
([Chapter 8](../Chapter_8/Theory.qmd) explains how). A response curve adds its
own parameters to those of the distribution, so this cost grows. Second, the
parameters of a curve are often strongly correlated (here the limiting rate
$a$ and the half-maximum $b$), so that the NLL forms a narrow and curved
valley like the one of the owl model in
[Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-surface). In such a valley
the direction of the next step depends on small differences between the partial
derivatives, and an inexact gradient can stop the search before it reaches the
minimum. Third, the same derivatives are behind the other tools: `sdreport()`
calculates the Hessian from them, the delta method needs the derivatives of the
quantities requested with `ADREPORT()` (@sec-delta-method), and in Chapter 7 a
parameter marked with `random =` in `MakeADFun()` is integrated out of the
likelihood with an approximation that is built on them.

For RTMB to calculate these derivatives, the NLL must be written with
operations and functions that it knows. This is the reason why the density and
probability mass functions of the course come from RTMB and RTMBdist.

::: {.callout-caution collapse="true" icon=false title="Technical detail: how automatic differentiation works"}
[RTMB](https://cran.r-project.org/web/packages/RTMB/vignettes/RTMB-introduction.html)
is an R interface to the package Template Model Builder (TMB). `MakeADFun()`
runs the NLL once, at the starting values, and records every elementary
operation that the function carries out on the parameters (each sum, product,
division, exponential and logarithm), in the order in which they happen. This
record is called a **tape**. The derivative of each of these operations is
known (they are the
[rules of differentiation](../Supplements/calculus.qmd#sec-calc-rules)), and the
chain rule combines them along the tape into the derivative of the NLL with
respect to every parameter. In plain English, RTMB differentiates the code of
the NLL step by step, as we would differentiate a formula by hand. The result
is exact up to rounding error: it is neither an algebraic formula nor a
numerical approximation. The elements `fn`, `gr` and `he` (the Hessian) of the
object all work from the tape. Because the tape runs through TMB's compiled
code, it cannot be stepped through like ordinary R code.
:::

::: {.callout-warning title="When automatic differentiation needs care"}
RTMB works out the derivatives from the calculations that the NLL carried out
when `MakeADFun()` ran it at the starting values. If the NLL contains a branch
that depends on a parameter (an `if` that compares a parameter with a value),
only the side that was taken at the starting values is used afterwards, and a
curve with a sharp change at an *unknown* threshold can have an undefined or
misleading derivative at the change. Such
[piecewise calculations need special care](https://cran.r-project.org/web/packages/RTMB/vignettes/RTMB-tips.html).
Check the NLL on sensible parameter values in R first (a function written with
`getAll()` can be called directly with two lists), and inspect
`intake_obj$fn(intake_obj$par)` and `intake_obj$gr(intake_obj$par)` before
fitting.
:::

The practicals of this chapter and of Chapter 7 use a shorter notation of RTMB
for the same likelihood. In `y %~% dnorm(mean = mu, sd = sigma)` the left side
is the observed response and the right side its distribution, and RTMB adds the
negative log density of `y` to the NLL, so that no `-sum(...)` line is needed.
Their functions take one named list of parameters and select its elements with
`$`.

## Return to confidence intervals

As in [Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-uncertainty),
`sdreport()` calculates the standard errors of the estimates from the Hessian,
which describes curvature on the **log-parameter scale**. Under the regularity
conditions discussed in Chapter 4, its inverse approximates the covariance
matrix on that scale. An approximate 95% Wald interval for the limiting rate
$a$ is:

```{r}
# Standard errors on the log scale and for the two ADREPORT() quantities.
intake_sd = sdreport(intake_obj)
intake_summary = summary(intake_sd)
intake_summary

# Wald interval on the log scale, with the endpoints transformed back
# to intake-rate units.
z = qnorm(p = 0.975)
se_log_a = intake_summary["log_a", "Std. Error"]
a_interval = exp(intake_fit$par[["log_a"]] + c(-1, 1) * z * se_log_a)
a_interval
```

The endpoints are transformed back to rate units. For a profile interval,
hold a candidate value of $a$ fixed and re-optimise **both** $b$ and
$\sigma$ at every candidate value, which is what
`TMB::tmbprofile(intake_obj, name = "log_a")` does. The profile follows the best possible
fit at each $a$; a slice that fixes $b$ and $\sigma$ at their original
estimates is different. Both methods remain approximate and depend on the
observation model and independence assumption.

::: {.callout-warning title="A confidence interval is conditional"}
If several measurements come from the same leaf or one population followed
over time, they may be dependent. Treating every row as independent can
make an interval too narrow. A calculation from the Hessian cannot repair
an unsuitable observation model.
:::

## The delta method {#sec-delta-method}

The **delta method** approximates the standard error of a quantity that is
calculated from the estimates, using the standard errors of those estimates.
In [Chapter 4](../Chapter_4/Theory.qmd#sec-likelihood-uncertainty) we asked for
such a quantity with `ADREPORT()` and read its standard error from
`sdreport()`. Here we see how that standard error is calculated, first for a
quantity that depends on one parameter.

Let $g(\theta)$ be the quantity and $\hat\theta$ the estimate. Close to the
estimate we replace the function by its tangent line,
$g(\theta)\approx g(\hat\theta)+g'(\hat\theta)(\theta-\hat\theta)$. Across
hypothetical experiments the estimate varies around the parameter with a
standard deviation equal to its standard error, and a straight line multiplies
that variation by its slope. Therefore

$$
\operatorname{SE}\big(g(\hat\theta)\big)\approx
\left|g'(\hat\theta)\right|\,\operatorname{SE}(\hat\theta),
$$

where $g'$ is the derivative of $g$. In plain English, the standard error of
the new quantity is the standard error of the estimate multiplied by the slope
of the function that links the two. For the limiting rate, $a=e^{\log a}$,
the derivative of the exponential is the exponential itself, so
$\operatorname{SE}(\hat a)\approx\hat a\,\operatorname{SE}(\log\hat a)$. This is
the formula that Chapter 4 gave in a footnote.

A quantity may depend on several parameters. The expected intake rate at a prey
density of 10, $g=10a/(b+10)$, depends on $a$ and on $b$, whose estimates are
correlated. The tangent line becomes a tangent plane and the variance of the
quantity is approximately

$$
\operatorname{Var}\big(g(\hat\theta)\big)\approx
\nabla g(\hat\theta)^{\top}\,\hat\Sigma\,\nabla g(\hat\theta),
$$

where $\hat\theta$ is now the vector of estimates (here on the log scale),
$\nabla g$ is the vector of the partial derivatives of $g$ with respect to each
parameter ([Chapter 5](../Chapter_5/Theory.qmd#sec-parameter-derivatives)) and
$\hat\Sigma$ is the covariance matrix of the estimates, that is, the inverse of
the Hessian. In plain English, each parameter contributes its variance
multiplied by the square of the slope of $g$ in its direction, and each pair of
parameters contributes its covariance multiplied by the two slopes. RTMB
obtains $\nabla g$ by automatic differentiation as well, which is why
`ADREPORT()` is called inside the NLL: the quantity has to be calculated there
for RTMB to differentiate it. We calculate both standard errors by hand and
compare them with the rows `a` and `intake_10` of `intake_summary`:

```{r}
# Estimates on the original scale.
a_hat = exp(intake_fit$par[["log_a"]])
b_hat = exp(intake_fit$par[["log_b"]])

# By hand, one parameter: the estimate of a times the standard error of log(a).
a_hat * intake_summary["log_a", "Std. Error"]

# By hand, several parameters: partial derivatives of the expected intake rate
# at a density of 10 with respect to log(a), log(b) and log(sigma).
gradient_10 = c(
  a_hat * 10 / (b_hat + 10),
  -a_hat * 10 * b_hat / (b_hat + 10)^2,
  0
)
# Combine them with the covariance matrix of the estimates.
sqrt(sum(gradient_10 * (intake_sd$cov.fixed %*% gradient_10)))
```

The delta method is an approximation of its own, on top of the one behind the
standard errors of the estimates. It is good when $g$ is close to a straight
line over the range of values that the uncertainty of the estimates covers,
which is the case for large samples. For a positive parameter such as $a$, the
Wald interval calculated on the log scale with its endpoints transformed back
remains the one that we recommend, as in Chapter 4. The delta method is what we
need when we report a standard error on the original scale, or the uncertainty
of a quantity that is not itself a parameter, such as a point of the fitted
curve.

## Check an interval by simulation

When we simulate data, we choose the generating parameters and know their
values. We can repeatedly simulate responses at specified predictor values,
refit the model, and check how often each interval contains the fixed
generating value. That proportion estimates **coverage** for this simulation
setting. The mean of the fitted estimates minus the generating value
estimates **bias**.

For a simulation study, repeat the entire fitting and interval procedure in
each simulated dataset. Record failed fits rather than quietly removing them,
and vary features such as sample size or the range of predictor values. The
result describes the procedure under the chosen generating model; it is not
proof that the model generated the supplied data. Exercise 3 in the
[Chapter 6 practical](Practicals/no_solution.qmd) provides a full example.

# Compare candidate models {#sec-response-comparison}

An ecological question may support several curves. A Holling type II
([Michaelis–Menten](../Supplements/deterministic-functions.qmd#sec-fn-michaelis-menten))
function and a
[saturating exponential](../Supplements/deterministic-functions.qmd#sec-fn-monomolecular)
can both increase towards a plateau, but differ between low and high prey
density. Compare candidates that represent plausible explanations and use the
same observational unit and response. When one curve is a special case of
another, the two are nested, as described in
[Chapter 5](../Chapter_5/Theory.qmd#sec-nested-functions).

## Information criteria

A model with more free parameters can often fit the observed data more
closely. Information criteria balance fit against model complexity.
[Chapter 4](../Chapter_4/Theory.qmd#sec-aic) introduced the Akaike
information criterion (AIC); here we recall it and add the Bayesian
information criterion (BIC). For $K$ estimated parameters, $n$
observations and maximised log-likelihood $\hat\ell$,

$$
\mathrm{AIC}=2K-2\hat\ell,
\qquad
\mathrm{BIC}=K\log(n)-2\hat\ell.
$$

Both use the same fitted likelihood, but BIC's penalty per parameter
grows with sample size. Count observation-distribution parameters such as
$\sigma$ or a negative binomial shape parameter in $K$. On our `optim()`
fits, `intake_fit$value` is the *negative* log-likelihood, so
`2 * K + 2 * intake_fit$value` calculates AIC. A lower value is preferred by
that criterion **among the candidates compared**. The value does not
measure absolute model adequacy.

Only compare likelihoods computed for the same observed response on the
same scale, using the same rows and compatible probability masses or
densities. For example, dropping missing rows for one model but not
another makes an AIC comparison misleading. AIC and BIC can rank
misspecified candidates, so residual checks remain necessary. These
conditions also appear in [R's AIC documentation](https://search.r-project.org/R/refmans/stats/html/AIC.html).

::: {.callout-note title="What a small difference means"}
A small AIC or BIC difference gives little reason to claim that one
curve has a clearly preferable score. Neither criterion identifies a
biological mechanism or assigns a probability that a candidate is true.
:::

## Cross-validation

**Cross-validation** checks prediction on observations that were not used
to estimate a model's parameters. In five-fold cross-validation, divide
the observations into five parts. For each part, fit each candidate to
the other four parts, then score the held-out observations. Repeat until
every part has been held out once.

A useful score here is the sum of held-out log probabilities or densities.
It uses the whole predictive distribution, so it can assess continuous,
count, and binary models. Prefer the *larger* held-out log score. Refit
all parameters, including $\sigma$ or dispersion, within each training
set. Using the full-data estimates for held-out predictions would leak
information across the split.

Random folds suit observations that can reasonably be treated as
independent. For population measurements ordered in time, training on
later measurements to predict earlier ones answers a different question.
If the goal is future prediction, hold out later times and fit to earlier
times. [Time-based resampling guidance](https://rsample.tidymodels.org/articles/Common_Patterns.html)
uses this ordering for the same reason.

:::: {.callout-important title="Try it: score a held-out observation"}
A model assigns probability $0.8$ to a held-out species presence. Another
assigns probability $0.3$. Which model receives the higher log score
for this observation?

::: {.callout-tip title="Answer" collapse="true"}
The first: $\log(0.8)>\log(0.3)$. This one observation does not establish
which model predicts better across the whole dataset.
:::
::::

# Check the fitted observation model {#sec-response-residuals}

An ordinary residual such as $y_i-\hat\mu_i$ changes its meaning across
response distributions. It may also change its spread with the predictor
even when a count model is correct. **Simulation-based scaled quantile
residuals** provide a common scale: values between 0 and 1.

For each observation, simulate many possible responses from the fitted
model at its predictor value. Locate the observed response within those
simulations. An observation near the lower end gives a residual near 0;
one near the upper end gives a residual near 1. When the simulation
matches the observation process, the residuals should be approximately
uniform across 0 to 1 and show no systematic pattern against the
predictor. For discrete responses, randomisation handles ties. This is
the approach used by [DHARMa's scaled quantile residuals](https://search.r-project.org/CRAN/refmans/DHARMa/html/getQuantile.html).

Our fits are ordinary `optim()` results, for which DHARMa does not know
the response distribution. We therefore generate a matrix with one row
per observation and one column per simulation, then supply it to
`DHARMa::createDHARMa()`:

```r
# Fix the random simulation sequence for a reproducible residual check.
set.seed(seed = 701)
# Calculate fitted means, then simulate complete intake measurements.
mu_hat = intake_curve(
  x = intake_data$x,
  a = estimates["a"], b = estimates["b"]
)
sigma_hat = estimates["sigma"]
simulated_rates = replicate(
  n = 500,
  expr = rnorm(n = nrow(intake_data), mean = mu_hat, sd = sigma_hat)
)
# Create scaled residuals from the simulations and plot them against density.
scaled = DHARMa::createDHARMa(
  simulatedResponse = simulated_rates,
  observedResponse = intake_data$y,
  fittedPredictedResponse = mu_hat,
  integerResponse = FALSE
)
plot(scaled)
DHARMa::plotResiduals(scaled, predictor = intake_data$x)
```

Run this code after installing DHARMa in your R environment. For counts or
presence/absence, simulate with `rpois()` or `rbinom()` as specified by the
fitted model and set `integerResponse = TRUE`. The package's
[`createDHARMa()` documentation](https://search.r-project.org/CRAN/refmans/DHARMa/html/createDHARMa.html)
describes this interface for custom simulations.

Look for a non-uniform residual distribution, a trend against $x$, changing
spread, or unusual concentrations near 0 or 1. For time measurements,
also plot against time in order. A diagnostic indicates where the model
disagrees with observations; it does not, on its own, tell us whether the
curve, observation distribution, dependence assumption, or data quality
caused the disagreement. Randomisation and a finite number of simulations
also make the exact plot vary between runs, so set a seed.

::: {.callout-warning title="What was simulated?"}
Simulate complete observations from the fitted distribution at each
predictor value. Simulating only the fitted mean curve produces no
observation-level variation and cannot support this residual check.
:::

# A fitting workflow

1. Plot the observations and check their units, support, missing values,
   repeated measurements, and predictor range.
2. Choose a small set of plausible mean curves and observation models.
3. Write the likelihood, check it at sensible parameter values, process it
   with `MakeADFun()`, fit every parameter, inspect convergence, and plot
   fitted curves against the data.
4. Estimate confidence intervals for ecologically meaningful parameters, and
   ask for quantities calculated from them with `ADREPORT()`.
5. Compare compatible candidates by AIC or BIC and, when appropriate,
   cross-validated predictive scores.
6. Simulate observations from the fitted model and inspect scaled
   residuals. Revise the model or qualify the conclusion when a pattern
   remains.

# Chapter quiz

:::: {.callout-important title="Question 1: the missing component"}
You have written $m(x)=ax/(b+x)$ and plotted it over the data. What must
you add before calculating a likelihood?

::: {.callout-tip title="Answer" collapse="true"}
Specify a distribution for observations around the curve, its remaining
parameters, and the dependence assumptions for the observational units.
:::
::::

:::: {.callout-important title="Question 2: a positive parameter"}
Why might you estimate $\log b$ rather than $b$ directly? On which scale
should the final ecological interval be reported?

::: {.callout-tip title="Answer" collapse="true"}
Exponentiating an unconstrained estimate keeps $b$ positive. Calculate
the interval consistently on the fitted scale, then transform its
endpoints back to the units of the predictor.
:::
::::

:::: {.callout-important title="Question 3: comparing models"}
Two models use different subsets of rows because only one removed
missing responses. Can their AIC values be compared as calculated?

::: {.callout-tip title="Answer" collapse="true"}
No. Fit both to the same observed responses and rows before comparing
their AIC values.
:::
::::

:::: {.callout-important title="Question 4: held-out prediction"}
Why must an observation held out in cross-validation be absent from
*every* estimation step for its fold?

::: {.callout-tip title="Answer" collapse="true"}
Using it to choose parameters or dispersion would let information from
the assessment observation influence its own prediction.
:::
::::

:::: {.callout-important title="Question 5: scaled residuals"}
A residual plot slopes upwards as the predictor increases. What can you
conclude, and what can you not conclude?

::: {.callout-tip title="Answer" collapse="true"}
The simulated model does not reproduce how the observations change with
the predictor. The plot alone does not identify which assumption or
ecological process caused that pattern.
:::
::::

:::: {.callout-important title="Question 6: coverage"}
What must be known to estimate a confidence interval's coverage in a
simulation study, and why is this unavailable for an ordinary dataset?

::: {.callout-tip title="Answer" collapse="true"}
We must know the fixed generating parameter and repeat sampling and
interval construction. The generating parameter is unknown in an
ordinary dataset.
:::
::::

:::: {.callout-important title="Question 7: automatic differentiation"}
What does RTMB provide to the optimiser, and why might a response curve with
an unknown sharp threshold need extra care?

::: {.callout-tip title="Answer" collapse="true"}
RTMB calculates the gradient of the NLL accurately by automatic
differentiation, and `optim()` receives it as `gr`.
At a sharp threshold the derivative may be undefined, and of a branch
chosen according to a parameter only the side taken at the starting values is
used. Check the NLL in R and consider a smooth curve or a different
fitting strategy for that model.
:::
::::

:::: {.callout-important title="Question 8: the delta method"}
A fit gives $\log\hat b=2.40$ with a standard error of 0.11. What is the
approximate standard error of $\hat b$, and which line of the NLL makes
`sdreport()` report it?

::: {.callout-tip title="Answer" collapse="true"}
By the delta method it is the standard error on the log scale multiplied by
the slope of the exponential at the estimate, which is $\hat b=e^{2.40}\approx
11.0$: approximately $11.0\times0.11\approx1.2$, in the units of prey density.
The line `ADREPORT(b)` inside the NLL asks for it.
:::
::::
