---
title: "Deterministic functions in ecology"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

How does the number of prey eaten by one predator change as prey become more
abundant? At low prey density, encounters may limit consumption. At high density,
handling each prey item may limit it instead. We can describe this expected
response with a **deterministic function**: the same input and parameter values
always give the same output.

In [Chapter 5](../Chapter_5/Theory.qmd), we used probability models and
likelihood to estimate parameters from observations. Here we first study the
shapes of functions that could describe an ecological response. We will calculate
their slopes, limits, peaks, and other landmarks before deciding how observations
might vary around a curve. This gives us a way to interpret a fitted curve later,
rather than treating it as a line drawn by software.

The functions and numerical values in this chapter are illustrative. A curve
that looks plausible does not, by itself, establish an ecological mechanism.
A compact [reference to common functions](../Supplements/deterministic-functions.qmd)
is available alongside this chapter.

# Learning goals

After studying this chapter, you should be able to:

- define the input, output, domain, parameters, and units of an ecological
  response function;
- calculate and plot a function in R, and explain how changing a parameter
  changes its curve;
- use limits and derivatives to find intercepts, asymptotes, initial slopes,
  maxima, and inflection points;
- calculate half-lives and half-saturation points and interpret them in
  ecological units;
- shift, scale, and reparameterise a function without changing its intended
  meaning;
- use a transformation or a local approximation to understand a curve while
  recognising the range over which it is useful.

# From an ecological question to a function {#sec-deterministic-functions}

Let $N$ be prey density, measured in prey per square metre, and let $F(N)$ be
the expected number of prey eaten by one predator per hour. A possible response
is

$$
F(N)=\frac{aN}{b+N}, \qquad N\geq 0,\quad a>0,\quad b>0.
$$

Here $a$ has units of prey per hour for one predator, and $b$ has the same units as
$N$: prey per square metre. We will soon see that $a$ is the upper limit of
the response and $b$ is the density at which the response reaches half that
limit. This curve is called a **Holling type II functional response**; the
same mathematical shape appears in other ecological and biochemical settings.

For fixed $a$, $b$, and $N$, the equation returns one number. It contains no
random variable and gives no probability to a possible observation. A useful
general notation is $m(x;\boldsymbol\theta)$, where $x$ is a predictor and
$\boldsymbol\theta$ is a set of parameters. Whether the output represents a
mean count, a rate, a probability, or something else must be stated separately.

::: {.callout-note title="Domain and units"}
A mathematical formula may be defined for inputs that make no ecological
sense. Prey density cannot be negative, so our domain is $N\geq0$. The
denominator is positive on that domain when $b>0$. Stating the domain and
parameter restrictions is part of specifying the model.
:::

## Calculate and plot curves in R

We can translate the formula directly into an R function. The input comes
first, followed by the parameters:

```r
holling_type_ii = function(N, a, b) {
  a * N / (b + N)
}

holling_type_ii(N = c(0, 3, 6, 12), a = 12, b = 6)
```

Base R works on a vector of input values here because multiplication and
division are vectorised. To see the whole shape, use `curve()`:

```r
curve(
  holling_type_ii(N = x, a = 12, b = 6),
  from = 0, to = 30,
  xlab = "Prey density (prey per square metre)",
  ylab = "Prey eaten by one predator per hour"
)
```

The formal input to `curve()` is called `x`; inside the call we pass it to
the argument named `N`. Comparing curves while changing only $b$ helps us
see its effect:

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Holling type II responses with the same upper limit and different half-saturation densities."
holling_type_ii = function(N, a, b) {
  a * N / (b + N)
}
curve(
  holling_type_ii(N = x, a = 12, b = 3),
  from = 0, to = 30, ylim = c(0, 12),
  xlab = "Prey density (prey per square metre)",
  ylab = "Prey eaten by one predator per hour",
  lty = 1
)
curve(
  holling_type_ii(N = x, a = 12, b = 9),
  from = 0, to = 30, add = TRUE, lty = 2
)
legend(
  "bottomright", legend = c("b = 3", "b = 9"),
  lty = c(1, 2), bty = "n"
)
```

Both curves approach 12 prey per hour for one predator. The curve with larger
$b$ rises more slowly across the densities shown. We can make that statement
precise by calculating the half-saturation point below.

If a function depends on two varying quantities, we can evaluate it on a
grid. For example, we might ask how the predicted response varies with both
prey density $N$ and handling time $H$. Here $H$ is a parameter that we vary
to explore the model, not a measured predictor in a fitted analysis:

```r
response = function(N, H, alpha = 0.6) {
  alpha * N / (1 + alpha * H * N)
}

density_values = seq(from = 0, to = 30, length.out = 60)
handling_values = seq(from = 0.15, to = 0.5, length.out = 60)
response_grid = outer(
  X = density_values, Y = handling_values,
  FUN = response
)
contour(
  x = density_values, y = handling_values, z = response_grid,
  xlab = "Prey density (prey per square metre)",
  ylab = "Handling time (hours per prey)"
)
```

The grid and contour plot show how the same equation responds to combinations
of inputs. They do not tell us which parameter values describe any observed
data.

# Read the ends of a curve {#sec-function-limits}

Before taking a derivative, check where a function starts and what happens
as its input grows. A **limit** describes a value approached by a function,
whether or not the function reaches it at a finite input. A finite limiting
value as $x\to\infty$ is a horizontal **asymptote**.

Suppose $D(t)=Ae^{-kt}$ describes a quantity that declines with time, where
$A>0$, $k>0$, and $t\geq0$. At $t=0$, $D(0)=A$. As $t\to\infty$,
$e^{-kt}\to0$, so $D(t)\to0$. If the input instead represents canopy depth,
the rate parameter has units of inverse depth. Always attach its units
to the particular input.

For the Holling response, divide numerator and denominator by $N$ when
$N>0$:

$$
F(N)=\frac{a}{b/N+1}
\longrightarrow a
\quad\text{as }N\to\infty.
$$

Thus $a$ is the upper asymptote. At $N=0$, $F(0)=0$. For densities that are
small compared with $b$, $b+N\approx b$, giving
$F(N)\approx(a/b)N$. This is a *local* straight-line approximation: it
should not be extrapolated to arbitrarily high prey density.

The **Ricker function** provides a different shape. If $N$ is adult density
and $R(N)$ is recruit density, one form is

$$
R(N)=rNe^{-qN}, \qquad N\geq0,\quad r>0,\quad q>0.
$$

It begins at zero. As $N$ grows very large, exponential decline eventually
overcomes the factor $N$, so $R(N)\to0$. We will use a derivative to find
the peak between those two ends.

:::: {.callout-important title="Try it on paper"}

For $G(N)=8N/(4+N)$, calculate $G(0)$ and
$\lim_{N\to\infty}G(N)$. Then write the straight-line approximation
when $N$ is small compared with 4. Keep R closed until you have all three.

::: {.callout-tip title="Solution" collapse="true"}

$G(0)=0$. Dividing top and bottom by $N$ gives
$G(N)=8/(4/N+1)\to8$. For $N\ll4$, the denominator is
approximately 4, so $G(N)\approx2N$. The horizontal asymptote is 8.

:::
::::

# Slopes and derivatives {#sec-function-derivatives}

The **derivative** $f'(x)=df/dx$ is the slope of a curve at $x$. It
describes the change in the output per unit change in the input near that
point. Its units are output units divided by input units.

You will use four rules repeatedly. If $u(x)$ and $v(x)$ are functions,

| Rule | Derivative |
|---|---|
| Power | $(x^p)'=px^{p-1}$ |
| Chain | $(e^{u})'=e^u u'$ |
| Product | $(uv)'=u'v+uv'$ |
| Quotient | $(u/v)'=(u'v-uv')/v^2$, where $v\ne0$ |

For example, the chain rule gives the slope of exponential decay:

$$
\frac{d}{dt}\left(Ae^{-kt}\right)
=Ae^{-kt}(-k)
=-Ak e^{-kt}.
$$

The initial slope is $D'(0)=-Ak$, not $Ak$. The negative sign says the
quantity decreases as time increases. Since $D'(t)$ approaches zero from
below, the decline flattens towards its asymptote.

The Holling response uses the quotient rule. With $u(N)=aN$ and
$v(N)=b+N$, we have $u'(N)=a$ and $v'(N)=1$, so

$$
\begin{aligned}
F'(N)
&=\frac{a(b+N)-aN}{(b+N)^2}\\
&=\frac{ab}{(b+N)^2}.
\end{aligned}
$$

For $a,b>0$ and $N\geq0$, this slope is positive. It starts at
$F'(0)=a/b$ and approaches zero as $N$ grows. The response increases
and flattens; it never exceeds $a$ on its domain. Taking one more
derivative gives

$$
F''(N)=-\frac{2ab}{(b+N)^3}<0,
$$

so the curve bends down throughout its domain.

::: {.callout-warning title="Derivative and asymptote"}
Calculate the derivative's limit when you need its limiting slope.
A function approaching a finite value does not, by that fact alone, prove
that its derivative approaches zero.
:::

# Scale, half-saturation, and half-life {#sec-function-scales}

The position where a curve reaches half its asymptote is often easier to
read from a graph than the asymptote itself. For the Holling response,
set $F(N_{1/2})=a/2$:

$$
\begin{aligned}
\frac{aN_{1/2}}{b+N_{1/2}}&=\frac a2,\\
2N_{1/2}&=b+N_{1/2},\\
N_{1/2}&=b.
\end{aligned}
$$

$b$ is therefore the **half-saturation density**, not half the response
value. If $a=12$ prey per hour for one predator and $b=6$ prey per square
metre, then $F(6)=6$ prey per hour for one predator.

For exponential decay, a **half-life** $t_{1/2}$ solves
$D(t_{1/2})=A/2$:

$$
e^{-kt_{1/2}}=\frac12,
\qquad
t_{1/2}=\frac{\log 2}{k}.
$$

The characteristic input scale $1/k$ makes the quantity fall to $A/e$.
It has the same units as $t$. When a parameter multiplies an input, as
$k$ does in $kt$, its units must cancel the input units in the exponent.

# Peaks and changes in curvature {#sec-function-peaks}

A derivative of zero can mark a maximum, minimum, or a point where the
curve briefly flattens. A derivative may also fail to exist at a sharp
corner. Check the derivative's sign on either side, and check domain
boundaries, before calling a point a maximum or minimum.

For the Ricker function, use the product and chain rules:

$$
\begin{aligned}
R'(N)
&=r e^{-qN}+rN(-q)e^{-qN}\\
&=r e^{-qN}(1-qN).
\end{aligned}
$$

Because $r e^{-qN}>0$, the slope changes from positive to negative
when $1-qN=0$. The peak is at

$$
N_{\mathrm{peak}}=\frac1q,
\qquad
R(N_{\mathrm{peak}})=\frac{r}{qe}.
$$

The peak location has the units of adult density. Since $R(0)=0$ and
$R(N)\to0$ as $N\to\infty$, this is the global maximum for $N\geq0$.
This mathematical pattern alone does not tell us which ecological process
causes recruitment to decline at high density.

The sign of the **second derivative** describes curvature. A change from
$f''>0$ to $f''<0$ is an **inflection point**. Consider the logistic curve

$$
p(x)=\frac{1}{1+e^{-k(x-c)}}, \qquad k>0.
$$

It rises from 0 to 1, crosses $1/2$ at $x=c$, and has
$p'(x)=kp(x)[1-p(x)]$. Its slope is greatest at $p=1/2$, where
$p'(c)=k/4$. Differentiating again gives
$p''(x)=k^2p(x)[1-p(x)][1-2p(x)]$, which changes from positive to
negative at $x=c$. The midpoint is therefore an inflection point.

# What the parameters mean {#sec-function-parameters}

The same curve can be written with parameters that answer different
ecological questions. The Holling type II response is also often written

$$
F(N)=\frac{\alpha N}{1+\alpha HN},
$$

where $\alpha>0$ is an encounter or attack coefficient and $H>0$ is
handling time per prey. With $N$ in prey per square metre and $F$ in
prey per hour for one predator, $\alpha$ has units of square metres per hour
and $H$ has units of hours per prey. The denominator is dimensionless.

Divide the numerator and denominator by $\alpha H$:

$$
\frac{\alpha N}{1+\alpha HN}
=\frac{(1/H)N}{1/(\alpha H)+N}.
$$

Comparison with $aN/(b+N)$ gives

$$
a=\frac1H,\qquad b=\frac1{\alpha H},
\qquad \alpha=\frac ab,\qquad H=\frac1a.
$$

Both forms describe exactly the same curve. The $(a,b)$ form makes
the asymptote and half-saturation easy to read; the $(\alpha,H)$ form
emphasises an interpretation in terms of encounters and handling.
Such an interpretation depends on whether the assumptions about those
processes are appropriate for the ecological system.

For a Ricker curve, the form $rNe^{-qN}$ makes the initial slope $r$
clear. Writing $q=1/p$ instead gives $rNe^{-N/p}$, where $p$ directly
names the adult density at the peak. Reparameterising changes the labels
and numerical values of parameters, but not the predicted curve.

# Shift and scale a curve {#sec-function-shifts}

Suppose we know the shape of a basic function $g(z)$. We can construct

$$
f(x)=d+A\,g\left(\frac{x-c}{s}\right),
\qquad s>0.
$$

$c$ shifts the curve right by $c$ input units, $s$ stretches it
horizontally, $A$ scales its output, and $d$ shifts it vertically.
The order matters: we transform the input *inside* $g$ and the output
*outside* it.

For $g(z)=z/(1+z)$ on $z\geq0$, the transformed curve is

$$
f(x)=d+\frac{A(x-c)}{s+x-c},
\qquad x\geq c,\quad A>0.
$$

It begins at $f(c)=d$, approaches $d+A$, and reaches halfway
between those values when $x=c+s$. Its initial slope at $c$ is
$A/s$. Restricting the domain to $x\geq c$ prevents the shifted
formula from being used at ecologically unintended inputs.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "A saturating response shifted right by three input units and up by two output units."
shifted_response = function(x, A = 10, s = 4, c = 3, d = 2) {
  d + A * (x - c) / (s + x - c)
}
curve(
  shifted_response(x),
  from = 3, to = 27, ylim = c(0, 12),
  xlab = "Input x", ylab = "Response f(x)"
)
abline(h = c(2, 12), lty = 2)
points(x = c(3, 7), y = c(2, 7), pch = 19)
```

# Transform and approximate {#sec-function-approximations}

Transformations can expose a relationship hidden in a curved plot. For
$D(t)=Ae^{-kt}$, with $D(t)>0$, taking logarithms gives

$$
\log D(t)=\log A-kt.
$$

A graph of $\log D$ against $t$ is a line with intercept $\log A$ and
slope $-k$. Similarly, if $Y=ax^p$ with $x,Y>0$, then
$\log Y=\log a+p\log x$. These identities help us understand
parameters and find rough starting values. They do not automatically
justify fitting a straight line to transformed observations: doing so
changes the scale of the residual variation and can imply a different
statistical model.

For a logistic response $p(x)$ between zero and one, the
**logit** is $\log[p/(1-p)]$. If
$p(x)=1/(1+e^{-(u+vx)})$, then

$$
\log\left(\frac{p(x)}{1-p(x)}\right)=u+vx.
$$

This equation describes the deterministic curve. An observed binary
outcome is either zero or one; its logit is not defined at either endpoint.

A **Taylor approximation** replaces a smooth curve near $x_0$ by a
short polynomial:

$$
f(x)\approx
f(x_0)+f'(x_0)(x-x_0)
+\frac{f''(x_0)}{2}(x-x_0)^2.
$$

The first two terms give the tangent line. The quadratic term adds
local curvature. For the Holling response near $N=0$,
$F(0)=0$, $F'(0)=a/b$, and $F''(0)=-2a/b^2$, so

$$
F(N)\approx\frac abN-\frac{a}{b^2}N^2.
$$

This is useful only for $N$ small relative to $b$. A polynomial
approximation can behave implausibly far from the point where it was
constructed. The same local-curvature idea helps explain why
[Chapter 5](../Chapter_5/Theory.qmd#sec-likelihood-uncertainty)
used a quadratic approximation near a likelihood maximum.

# Summary

For a new ecological function, first state what its input and output
represent, their units, and the domain. Then calculate values at useful
points, limits, slopes, and any peaks or inflection points. Translate
those mathematical features into ecological claims only under the
assumptions that give the function that meaning. R plots can check and
communicate the shape, while the algebra explains it.

In a later analysis, a deterministic response such as $F(N)$ can describe
how a model's expected value changes with $N$. A probability distribution
would then describe variation among observations at a given $N$. The
function alone does not supply that variation.

The approach to exploring curve shapes in this chapter draws on
Benjamin Bolker's *Deterministic functions for ecological modeling*
(supplied 2007 PDF, especially sections 3 and 4).

# Chapter quiz

:::: {.callout-important title="Question 1: read a saturating curve"}

For $f(x)=9x/(3+x)$ on $x\geq0$, what are $f(0)$, its upper
asymptote, its half-saturation input, and its initial slope?

::: {.callout-tip title="Answer" collapse="true"}

$f(0)=0$, the upper asymptote is 9, the half-saturation input
is $x=3$, and $f'(0)=9/3=3$ output units per input unit.

:::
::::

:::: {.callout-important title="Question 2: check a stationary point"}

Why does a derivative of zero require an additional check before
you call a point a maximum?

::: {.callout-tip title="Answer" collapse="true"}

The point could be a minimum or a flat point. Check the
derivative's sign on either side and compare any domain
boundaries. A change from positive to negative indicates
a local maximum.

:::
::::

:::: {.callout-important title="Question 3: locate a Ricker peak"}

A Ricker curve is $R(N)=rNe^{-qN}$. Which parameter controls
its peak location, and what is that location?

::: {.callout-tip title="Answer" collapse="true"}

$q$ controls the peak location, $N=1/q$. The value of
$r$ scales the height without moving the peak.

:::
::::

:::: {.callout-important title="Question 4: shift a curve"}

If a function is shifted from $g(x)$ to $g(x-4)+2$, where
does its original point $(0,g(0))$ move?

::: {.callout-tip title="Answer" collapse="true"}

It moves to $(4,g(0)+2)$: four input units right and
two output units up.

:::
::::

:::: {.callout-important title="Question 5: add variation"}

What additional component would be needed before a deterministic
curve could assign a likelihood to observations?

::: {.callout-tip title="Answer" collapse="true"}

We need a probability distribution for observations
conditional on the predictor and the curve's parameters,
together with its assumptions. The deterministic curve
alone supplies only a predicted value.

:::
::::
