---
title: "From samples to estimates"
---

```{r}
#| include: false
# Load ggplot2 for the consistently formatted rendered figures in this chapter.
library(ggplot2)
```

# Introduction

In this chapter, we learn how to relate samples of data to probability distributions and how
to use a sample to estimate the parameters of a model. In
[Chapter 2](../Chapter_2/Theory.qmd) we chose the parameters of a distribution and
calculated probabilities, quantiles and expectations from them. In research we face the
reverse problem: we have a limited sample of observations and we want to learn about the
parameters of the model that we use to describe them.

The chapter has two parts. First we look at how samples and distributions are related,
through relative frequencies and the empirical distribution of a sample, both for counts
and for continuous measurements. We then define estimators and study how their estimates
vary across repeated samples, which leads to standard errors and confidence intervals. The
first part continues with the Binomial model of seed survival from Chapter 2. The second
part uses a Poisson model for counts of seedlings and a Normal model for tree heights. Most
of the data in this chapter are simulated, because in a simulation we know the parameters
that generated each sample and we can compare the estimates with them.

# Learning goals

In this chapter, we will cover:

- how samples and distributions are related (relative frequencies and the empirical
  distribution);
- histograms, density estimates and empirical cumulative distributions for continuous
  measurements;
- parameters, estimators and estimates;
- the method of moments;
- the sampling distribution of an estimator, its bias, variance and mean squared error;
- standard errors;
- confidence intervals and their coverage;
- consistency and asymptotic normality.

We will use the Binomial, Poisson and Normal distributions throughout the chapter. They are
described in the [Supplement](../Supplements/distributions.qmd).

# Samples and distributions {#sec-samples}

So far we have been working with mathematical models: we calculate probabilities and
expectations of distributions. But in research we collect limited, often noisy, samples of
data and want to generalise from these observations to the broader context in which they
were collected. As discussed in Chapter 1, we use statistical models as approximations for the
data generating process. This means that we treat the observations from our experiments as
if they were random samples from our statistical model. In this section we learn in more
detail about this relationship as it will be the basis for the rest of this
chapter.

For example, we might want to learn about seed survival across a restoration site while observing
only some quadrats. We cannot quietly turn those observed quadrats into a claim about the whole
site. To make that connection, we need assumptions about how quadrats were selected and how their
counts vary from quadrat to quadrat. We model this data generating process with a distribution
that is (at least qualitatively) compatible with that process.

## From frequency to probability

For a sample of counts, the **relative frequency** of a value is the number of times it appears
divided by the sample size. Let's simulate ten quadrats from the Binomial model of seed survival used
in [Chapter 2](../Chapter_2/Theory.qmd#sec-four_faces) (four seeds per quadrat, each surviving with probability 0.4):

```{r}
# Parameters of the Binomial model of seed survival used in Chapter 2.
seed_count = 4
seed_survival_probability = 0.4
# Simulate ten quadrats from the Binomial model.
set.seed(seed = 123)
n_quadrats = 10
sample_survivors = rbinom(
  n = n_quadrats,
  size = seed_count,
  prob = seed_survival_probability
)
```

`table()` counts how often each value appears. Divide by the number of quadrats to
calculate relative frequencies:

```{r}
# Count each possible value and divide by the sample size.
seed_values = seq.int(from = 0, to = seed_count)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = count_survivors / length(sample_survivors)
relative_frequency
```

Some possible counts may be absent from a small sample. Setting all possible
values as `factor()` levels makes `table()` include those counts with frequency zero.

The relative frequencies lie between zero and one and sum to one. Compare them
with the probabilities predicted by the Binomial model that we use to approximate the
data generating process:

```{r}
# Put the sample frequencies beside the model probabilities for comparison.
data.frame(
  seed_count = seed_values,
  relative_frequency = as.numeric(x = relative_frequency),
  probability = dbinom(
    x = seed_values,
    size = seed_count,
    prob = seed_survival_probability
  )
)
```

The two columns differ because the sample contains only ten quadrats. That is not a failure
of the model: it is what sampling variation looks like. For small sample sizes, the relative
frequencies and model probabilities can differ greatly. However, for independent and identically
distributed (i.i.d.) observations, the relative frequency of each count approaches its model
probability as sample size grows. This is an application of the **weak law of large numbers**.

For example, let $I_i=1$ when the $i$th quadrat has exactly three surviving seeds and
$I_i=0$ otherwise. The relative frequency of three survivors is
$\bar{I}_n=(I_1+\cdots+I_n)/n$. For any positive margin $\varepsilon$,

$$
\lim_{n \to \infty}
P\left(\left|\bar{I}_n - P(X=3)\right| \ge \varepsilon\right)=0.
$$

In plain English, as $n$ grows, the chance that the relative frequency differs from the
model probability by at least $\varepsilon$ becomes small. There are two probabilities here:
the probability assigned to a count by the model, and the probability of seeing a
relative frequency that is far from the model one. This convergence is
not deterministic: a larger realised sample need not be closer to the model probability
than every smaller realised sample.

::: {.callout-note title="Frequencies converge to probabilities"}
For independent observations from the same distribution, the relative frequency of any
specified value converges in probability to its probability mass as sample size grows.
:::

We can push the simulation further and compare relative frequencies at sample sizes
from 10 to 100,000 to see this convergence in action:

```{r}
# Compare relative frequencies for several sample sizes with the model.
set.seed(123)
sample_sizes = 10^(seq_len(length.out = 5))
seed_values = seq.int(from = 0, to = seed_count)

all_frequencies = matrix(NA, nrow = length(seed_values),
                         ncol = length(sample_sizes) + 1)
colnames(all_frequencies) = c(as.character(sample_sizes), "theory")

for (i in seq_along(sample_sizes)) {
  sample_survivors = rbinom(
    n = sample_sizes[i],
    size = seed_count,
    prob = seed_survival_probability
  )
  count_survivors = table(factor(
    x = sample_survivors,
    levels = seed_values
  ))
  all_frequencies[, i] = as.numeric(count_survivors) / sample_sizes[i]
}
all_frequencies[, length(sample_sizes) + 1] = dbinom(
  x = seed_values,
  size = seed_count,
  prob = seed_survival_probability
)
round(all_frequencies, 3)
```

One sequence of samples cannot show convergence directly. For example, the relative frequency
of three surviving seeds at 10,000 quadrats can happen to be closer to the model probability
than its value at 100,000 quadrats. To show what usually changes with sample size, repeat the
simulation 100 times at each size. This is shown in @fig-relative-frequency-convergence.
At small sample sizes, the repeated relative frequencies cover a wide range. As the sample
size grows, the relative frequencies concentrate around the probabilities of the model.
Individual samples can still generate values far from the line, but large deviations become
less likely.

```{r}
#| echo: false
#| eval: true
#| label: fig-relative-frequency-convergence
#| fig-cap: "Relative frequencies of exactly three surviving seeds for simulated samples of different sizes. The horizontal line is the probability assigned by the Binomial distribution."
# Repeatedly simulate samples and retain the relative frequency of three survivors.
set.seed(seed = 20261003)
seed_count = 4
seed_survival_probability = 0.4
target_count = 3
sample_sizes = 10^(seq_len(length.out = 5))
n_repetitions = 100

repeated_frequencies = data.frame(
  sample_size = numeric(),
  relative_frequency = numeric()
)

for (sample_size in sample_sizes) {
  relative_frequency = numeric(length = n_repetitions)

  for (repetition in seq_len(n_repetitions)) {
    sample_survivors = rbinom(
      n = sample_size,
      size = seed_count,
      prob = seed_survival_probability
    )
    relative_frequency[repetition] = mean(sample_survivors == target_count)
  }

  repeated_frequencies = rbind(
    repeated_frequencies,
    data.frame(
      sample_size = sample_size,
      relative_frequency = relative_frequency
    )
  )
}

# Calculate the model probability that the figure uses as a reference line.
probability = dbinom(
  x = target_count,
  size = seed_count,
  prob = seed_survival_probability
)

ggplot(
  repeated_frequencies,
  aes(
    x = sample_size,
    y = relative_frequency,
    group = sample_size
  )
) +
  geom_point(colour = "grey40",
             position = position_jitter(width = 0.05, height = 0.005)) +
  geom_hline(
    yintercept = probability,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  scale_x_log10(
    breaks = sample_sizes,
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    x = "Sample size (log10 scale)",
    y = "Relative frequency of three surviving seeds"
  ) +
  theme_classic()
```

:::: {.callout-important title="Practice: fruit-set frequency"}
In a sample of 24 tagged orchid flowers, 15 set fruit. Calculate the relative frequency
of fruit set. Explain what this value estimates and why it need not equal the model
fruit-set probability exactly.

::: {.callout-tip title="Solution" collapse="true"}

The relative frequency is $15/24=0.625$. It is not equal to the probability assigned by the
model to that count, but it is probably not too different for a sample of 24.
:::
::::

## Empirical distribution

Because of this relationship between the relative frequencies and probabilities, we often
define the **empirical distribution** of a sample. This is a procedure where we assign each
observed count its relative frequency and we treat it as if it were a probability mass
function (@fig-ecdf). In principle it is just a way to tabulate a (potentially long) collection of
observations, but we will see that this also facilitates calculating properties of a sample
such as quantiles or expectations that we can relate to the equivalent properties of the
model.

::: {.callout-note title="Empirical distribution"}
A distribution whose mass at each count equals the relative frequency in the sample.
:::

The empirical distribution is also important because, generally speaking^[this is not a
formal statistical statement, more of a heuristic derived from experience and common sense,
to build intuition], a sample whose empirical distribution matches well the model
distribution will generate reasonable estimates of the model's parameters with relatively
low uncertainty. In this we would say that the "sample is representative" of the process we
are trying to model. This is why it is crucial to avoid introducing biases in the way we
collect data from experiment (motivating procedures such as randomisation).

Like any distribution, an empirical distribution has a cumulative function: the fraction
of observations at or below each count. When comparing the empirical and model cumulative
probability functions, we often use step-plots, which easily allow both functions side by
side (@fig-ecdf). These types of visualisations are useful if you want to compare how well a
distribution matches a given sample of data (i.e., we want to test if a sample of observations
could be treated as a random sample from a given distribution). We will expand on this notion
in later chapters.

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 4
#| label: fig-ecdf
#| fig-cap: "Empirical and model distribution for a sample of ten quadrats in the example on seed survival."
# Simulate a sample and calculate its empirical distribution.
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
n_quadrats = 10
seed_values = seq.int(from = 0, to = seed_count)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
cumulative_relative_frequency = cumsum(relative_frequency)

model_comparison_data = rbind(
  data.frame(
    surviving_seeds = seed_values,
    value = model_mass,
    quantity = "Probability mass",
    series = "Model"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = relative_frequency,
    quantity = "Probability mass",
    series = "Sample"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = model_cumulative,
    quantity = "Cumulative probability",
    series = "Model"
  ),
  data.frame(
    surviving_seeds = seed_values,
    value = cumulative_relative_frequency,
    quantity = "Cumulative probability",
    series = "Sample"
  )
)
model_comparison_data$quantity = factor(
  model_comparison_data$quantity,
  levels = c("Probability mass", "Cumulative probability")
)

# Use different geometries as well as colours to distinguish the model and sample.
ggplot() +
  geom_linerange(
    data = subset(
      model_comparison_data,
      quantity == "Probability mass" & series == "Model"
    ),
    aes(x = surviving_seeds, ymin = 0, ymax = value, colour = series),
    linewidth = 0.8
  ) +
  geom_point(
    data = subset(
      model_comparison_data,
      quantity == "Probability mass" & series == "Sample"
    ),
    aes(x = surviving_seeds, y = value, colour = series),
    alpha = 0.7,
    size = 2
  ) +
  geom_step(
    data = subset(model_comparison_data, quantity == "Cumulative probability"),
    aes(x = surviving_seeds, y = value, colour = series, linetype = series),
    direction = "hv",
    linewidth = 0.8
  ) +
  facet_wrap(~quantity, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = c(Model = "#D55E00", Sample = "#0072B2")) +
  scale_linetype_manual(values = c(Model = "solid", Sample = "dashed")) +
  scale_x_continuous(breaks = seed_values) +
  labs(
    x = "Surviving seeds per quadrat",
    y = "Probability or relative frequency",
    colour = NULL,
    linetype = NULL
  ) +
  theme_classic()
```

### Mean, variance and quantiles of a data sample

Just like for any probability distribution described in
[Chapter 2](../Chapter_2/Theory.qmd#sec-expectations), we can compute
expectations such as means and variances from an empirical distribution, as well as quantiles
by inverting the empirical cumulative distribution. We will simply replace probabilities in
the formulae by relative frequencies of the sample.

Note that, strictly speaking, a data sample is just a collection of values. It does not have
a mean or a variance; those are properties of distributions. However, language in applied
statistics is often used lazily and inconsistently, which can lead to a lot of confusion.
Whenever we say *sample mean* or *sample variance* we actually imply the mean or variance of
the empirical distribution (but see below that the variance of a sample is an even more
confusing term).

It is trivial to see that the mean of the empirical distribution equals the arithmetic
mean of the sample:

```{r}
# Simulate a small sample and calculate its weighted and ordinary means.
set.seed(123)
n_quadrats = 10
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_survivors = rbinom(
  n = n_quadrats, size = seed_count, prob = seed_survival_probability
)
count_survivors = table(factor(x = sample_survivors, levels = seed_values))
relative_frequency = as.numeric(count_survivors) / n_quadrats
weighted_average = sum(relative_frequency * seed_values)
sample_mean = mean(sample_survivors)
c(weighted_average, sample_mean)
```

However, the variance of the empirical distribution is not exactly the same as what we normally
call *sample variance*:

```{r}
# Calculate the variance of the empirical distribution
empirical_variance = sum((seed_values - weighted_average)^2 * relative_frequency)
sample_variance = var(sample_survivors)
c(empirical_variance, sample_variance)
```

It turns out that what most people call sample variance is actually a statistical estimator
of the variance of the model distribution (i.e., in this case the variance of the Binomial
distribution), which is a completely different concept. For reasons that will become clear in
@sec-two_parameters, this
variance is calculated with a different formula^[To make matters worse, many individuals will
claim dogmatically that this second way of calculating the sample variance is the correct one,
without realising what the differences are between the two. Unfortunately, this type of
confusion and misunderstanding from sloppy use of language is very common in applied
statistics... And this has also polluted AI chatbots, so sometimes trying to understand
proper statistics becomes a very uphill battle.].

A quantile from the empirical distribution can be derived from the `quantile()` function. Note
that when dealing with discrete data you need to use `type = 1` as an argument. For example,
if we calculate the quantile for a cumulative probability of 0.5 (i.e. the median) it will
return the smallest observed count whose empirical cumulative probability reaches $0.5$:

```{r}
# Use type = 1 so the empirical quantile is one of the observed counts.
quantile(x = sample_survivors, probs = 0.5, type = 1)
```

The function `quantile()` appears in many areas of statistics, and there are up to nine different
ways of computing it in R. So the purpose for which we calculate the quantile really matters.

### Asymptotic convergence of the empirical distribution

::: {.callout-note title="Asymptotic convergence of empirical distribution"}
For independent observations from the same fixed distribution, the empirical
probability mass function converges to the model probability mass function as sample size
grows towards infinity. All the moments (means, variances, and others), as well as all
quantiles from the empirical distribution, will also converge to their equivalent values
in the model. With bounded counts, as in our Binomial example, the empirical mean and
variance also converge to the values in the model. A finite sample can sometimes agree
exactly; agreement is not guaranteed at any finite size.
:::

By defining the empirical distribution we can now better understand how samples and
distributions are related. That is, a collection of observations can be treated as a random
sample from a model if the empirical distribution converges to that model as sample size
increases. This does not imply that the observations are actual random samples from some
mathematical stochastic representation; it simply means that we have modelled the variation
across different observations correctly.

Note that in practice we can never verify this relationship as we will never have infinite
samples and for any finite size there can be deviations. But it does mean that the larger the
sample, the better we can judge if we have a good model (@fig-ecdf-binom). Also, larger
samples are more **stable**, meaning that there will be less variation across repeated samples,
which will increase our confidence (reduce our uncertainty) in the results (we cover this
in more detail in @sec-sampling_distribution).

```{r}
#| echo: false
#| eval: true
#| label: fig-ecdf-binom
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Model probabilities and sample frequencies for 10 quadrats (top) and 1,000 quadrats (bottom)."
# Compare empirical distributions at two sample sizes.
set.seed(142)
seed_count = 4
seed_survival_probability = 0.4
seed_values = seq.int(from = 0, to = seed_count)
sample_sizes = c(10, 1000)
model_mass = dbinom(
  x = seed_values, size = seed_count, prob = seed_survival_probability
)
model_cumulative = pbinom(
  q = seed_values, size = seed_count, prob = seed_survival_probability
)

# Store the model and empirical summaries for both sample sizes before plotting.
sample_comparison_data = data.frame()
for (n_quadrats in sample_sizes) {
  sample_survivors = rbinom(
    n = n_quadrats, size = seed_count, prob = seed_survival_probability
  )
  count_survivors = table(factor(x = sample_survivors, levels = seed_values))
  relative_frequency = as.numeric(count_survivors) / n_quadrats
  cumulative_relative_frequency = cumsum(relative_frequency)

  sample_size_label = paste(format(n_quadrats, big.mark = ","), "quadrats")
  sample_comparison_data = rbind(
    sample_comparison_data,
    data.frame(
      surviving_seeds = seed_values,
      value = model_mass,
      quantity = "Probability mass",
      series = "Model",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = relative_frequency,
      quantity = "Probability mass",
      series = "Sample",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = model_cumulative,
      quantity = "Cumulative probability",
      series = "Model",
      sample_size = sample_size_label
    ),
    data.frame(
      surviving_seeds = seed_values,
      value = cumulative_relative_frequency,
      quantity = "Cumulative probability",
      series = "Sample",
      sample_size = sample_size_label
    )
  )
}

sample_comparison_data$sample_size = factor(
  sample_comparison_data$sample_size,
  levels = c("10 quadrats", "1,000 quadrats")
)
sample_comparison_data$quantity = factor(
  sample_comparison_data$quantity,
  levels = c("Probability mass", "Cumulative probability")
)

# Compare each empirical distribution with the same model using matching panels.
ggplot() +
  geom_linerange(
    data = subset(
      sample_comparison_data,
      quantity == "Probability mass" & series == "Model"
    ),
    aes(x = surviving_seeds, ymin = 0, ymax = value, colour = series),
    linewidth = 0.8
  ) +
  geom_point(
    data = subset(
      sample_comparison_data,
      quantity == "Probability mass" & series == "Sample"
    ),
    aes(x = surviving_seeds, y = value, colour = series),
    alpha = 0.7,
    size = 2
  ) +
  geom_step(
    data = subset(sample_comparison_data, quantity == "Cumulative probability"),
    aes(x = surviving_seeds, y = value, colour = series, linetype = series),
    direction = "hv",
    linewidth = 0.8
  ) +
  facet_grid(sample_size ~ quantity, scales = "free_y") +
  scale_colour_manual(values = c(Model = "#D55E00", Sample = "#0072B2")) +
  scale_linetype_manual(values = c(Model = "solid", Sample = "dashed")) +
  scale_x_continuous(breaks = seed_values) +
  labs(
    x = "Surviving seeds per quadrat",
    y = "Probability or relative frequency",
    colour = NULL,
    linetype = NULL
  ) +
  theme_bw() + theme(panel.grid = element_blank())
```

:::: {.callout-important title="Practice: orchid flowering across plots"}

Suppose each restoration plot contains 12 orchid plants. Each plant flowers with
probability 0.25, independently of the others. Let $X$ be the number of flowering
plants in a plot. This is a Binomial model. Simulate counts for 12 plots and for 500
plots. For each sample, calculate the empirical mean and plot the empirical mass function.
Compare both distributions with the model probability mass function. What changes in
the result from the larger sample? How much can you rely on a single simulation?

::: {.callout-tip title="Solution" collapse="true"}

One possible simulation is:

```{r}
# Simulate the two sample sizes and compare each with the same model.
set.seed(246)

plants_per_plot = 12
flowering_probability = 0.25

small_plot_counts = rbinom(
  n = 12,
  size = plants_per_plot,
  prob = flowering_probability
)
large_plot_counts = rbinom(
  n = 500,
  size = plants_per_plot,
  prob = flowering_probability
)

mean(small_plot_counts)
mean(large_plot_counts)

count_values = seq.int(from = 0, to = plants_per_plot)
small_plot_table = table(factor(
  x = small_plot_counts,
  levels = count_values
))
large_plot_table = table(factor(
  x = large_plot_counts,
  levels = count_values
))

small_frequency = as.numeric(small_plot_table) / length(small_plot_counts)
large_frequency = as.numeric(large_plot_table) / length(large_plot_counts)
model_mass = dbinom(
  x = count_values,
  size = plants_per_plot,
  prob = flowering_probability
)

small_frequency_data = data.frame(
  flowering_plants = count_values,
  probability = small_frequency,
  series = "12 plots"
)
large_frequency_data = data.frame(
  flowering_plants = count_values,
  probability = large_frequency,
  series = "500 plots"
)
model_mass_data = data.frame(
  flowering_plants = count_values,
  probability = model_mass,
  series = "Binomial model"
)

# Use bars for one empirical distribution, points for the other, and lines for the model.
# Set common axis limits
all_probabilities = c(
  small_frequency_data$probability,
  large_frequency_data$probability,
  model_mass_data$probability
)

plot(
  NA,
  xlim = range(count_values),
  ylim = c(0, max(all_probabilities)),
  xaxt = "n",
  xlab = "Flowering plants per plot",
  ylab = "Probability or relative frequency"
)

axis(1, at = count_values)

# Draw bars for the small empirical distribution
with(
  small_frequency_data,
  rect(
    flowering_plants - 0.35,
    0,
    flowering_plants + 0.35,
    probability,
    col = "grey",
    border = 1
  )
)

# Add points for the large empirical distribution
with(
  large_frequency_data,
  points(
    flowering_plants,
    probability,
    pch = 16,
    col = 2
  )
)

# Add vertical lines for the model distribution
with(
  model_mass_data,
  segments(
    flowering_plants,
    0,
    flowering_plants,
    probability,
    col = 3
  )
)

# Add one legend for all three series
legend(
  "topright",
  legend = c("12 plots", "500 plots", "Binomial model"),
  col = c(1, 2, 3),
  fill = c("grey", NA, NA),
  border = c(1, NA, NA),
  pch = c(NA, 16, NA),
  lty = c(0, 0, 1)
)
```

The larger sample will generally give a more stable empirical distribution and mean
closer to the values in the model. A particular simulation need not be closer in every
summary, and the empirical distribution need not exactly match the model.

:::
::::

## Samples of continuous measurements {#sec-continuous_samples}

The relationship between samples and distributions is the same for continuous measurements,
but the tools to describe a sample need to be adapted. In a sample of continuous values we
do not expect any value to be repeated, so the relative frequency of every observed value
is $1/n$ and a table of relative frequencies tells us nothing. We will use the Normal model
of tree height from [Chapter 2](../Chapter_2/Theory.qmd#sec-continuous_four_faces), with a
mean of 22 m and a standard deviation of 4 m, to illustrate how to describe a sample of
continuous data. We simulate the heights of 200 trees from this model:

```{r}
# Simulate the heights (in m) of 200 trees from the Normal model.
set.seed(seed = 73)
height_sample = rnorm(n = 200, mean = 22, sd = 4)
```

A **histogram** is the continuous counterpart of the table of relative frequencies. We
divide the range of the measurements into intervals (called **bins**) and count the
observations that fall in each bin. Dividing the count by the sample size gives the relative
frequency of the bin, which we can compare with the probability that the model assigns to
that interval. For example, for heights between 20 and 22 m:

```{r}
# Relative frequency of heights between 20 and 22 m in the sample.
mean(height_sample > 20 & height_sample <= 22)
# Probability that the model assigns to the same interval.
pnorm(q = 22, mean = 22, sd = 4) - pnorm(q = 20, mean = 22, sd = 4)
```

That is, 39 of the 200 trees are in this bin, a relative frequency of $39/200 = 0.195$,
and the model probability is approximately $0.191$. As with counts, the weak law of large
numbers applies to each bin: its relative frequency converges in probability to the model
probability of the interval as sample size grows.

To compare a histogram with a probability density function we need to scale it in the same
units. In a **density-scaled histogram** the height of each bar is the relative frequency
divided by the width of the bin, so that the area of the bar equals the relative frequency
and the areas of all bars add up to one. For the bin above, the height of the bar is
$0.195 / 2 = 0.0975$ per metre (middle panel of @fig-histogram_density), which we can compare with the Normal density in the
middle of the bin (`dnorm(x = 21, mean = 22, sd = 4)` is approximately $0.097$ per metre).
In R, `hist()` draws a histogram of counts by default and a density-scaled histogram with
the argument `probability = TRUE`.

The width of the bins is our choice and it changes how the same sample looks
(@fig-histogram_density). Narrow bins contain few observations each, so the heights of the
bars vary a lot from bin to bin. Wide bins are more stable, but hide the shape of the
distribution within each bin. There is no correct bin width; the larger the sample, the
narrower the bins can be.

A **kernel density estimate** is a smooth alternative to the histogram. Instead of counting
observations in bins, it places a small smooth curve (the kernel, by default a Normal
density) on top of every observation and averages these curves. The standard deviation of
the kernel is called the **bandwidth** and it plays the same role as the bin width: a small
bandwidth produces a wiggly estimate and a large bandwidth a very smooth one. In R,
`density()` calculates the kernel density estimate and chooses a bandwidth from the sample:

```{r}
# Kernel density estimate of the sample; bw is the bandwidth that R chose (in m).
height_density = density(x = height_sample)
height_density$bw
```

Note that a kernel density estimate is calculated from the sample without assuming a
model, whereas `dnorm()` is the density of the model that we chose. The two curves in
@fig-histogram_density differ because of sampling variation, even though the sample was
generated from that model.

```{r}
#| echo: false
#| label: fig-histogram_density
#| fig-width: 10
#| fig-height: 3.5
#| fig-cap: "Density-scaled histograms of 200 simulated tree heights with bins of 0.5 m, 2 m and 6 m. The solid line is the density of the Normal model and the dashed line is the kernel density estimate of the sample."
library(patchwork)

# Density of the model and kernel density estimate of the sample on the same grid.
height_grid = seq(from = 8, to = 36, length.out = 300)
model_curve = data.frame(
  height = height_grid,
  density = dnorm(x = height_grid, mean = 22, sd = 4)
)
kernel_curve = data.frame(height = height_density$x, density = height_density$y)
height_data = data.frame(height = height_sample)

# Draw one density-scaled histogram for a given bin width.
plot_histogram = function(bin_width) {
  ggplot(height_data, aes(x = height)) +
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = bin_width,
      boundary = 0,
      fill = "grey90",
      colour = "grey40"
    ) +
    geom_line(
      data = model_curve,
      aes(x = height, y = density),
      colour = "#D55E00",
      linewidth = 0.8
    ) +
    geom_line(
      data = kernel_curve,
      aes(x = height, y = density),
      colour = "#0072B2",
      linetype = "dashed",
      linewidth = 0.8
    ) +
    coord_cartesian(xlim = c(8, 36), ylim = c(0, 0.2)) +
    labs(
      title = paste("Bin width:", bin_width, "m"),
      x = "Tree height (m)",
      y = "Density (per m)"
    ) +
    theme_classic()
}

plot_histogram(bin_width = 0.5) |
  plot_histogram(bin_width = 2) |
  plot_histogram(bin_width = 6)
```

The empirical cumulative distribution needs no adaptation: it is still the fraction of
observations at or below each value, and it requires neither bins nor a bandwidth. With
continuous data it increases by $1/n$ at each observed value, whereas the cumulative
probability function of the model is a smooth curve (@fig-ecdf_normal). In R, `ecdf()`
returns the empirical cumulative distribution of a sample as a function that we can evaluate
at any value:

```{r}
# ecdf() returns a function; evaluate it at a height of 26 m.
height_ecdf = ecdf(height_sample)
height_ecdf(26)
# Cumulative probability of the model at the same height.
pnorm(q = 26, mean = 22, sd = 4)
```

So 81% of the simulated trees are at most 26 m tall, while the model assigns a probability
of approximately $0.84$ to that event. The mean, variance and quantiles of the sample are
calculated as before (for continuous data the default `type` of `quantile()` is
appropriate), and the asymptotic convergence described above also holds: the empirical
cumulative distribution converges to the cumulative probability function of the model as
sample size grows.

```{r}
#| echo: false
#| label: fig-ecdf_normal
#| fig-cap: "Empirical cumulative distribution of 200 simulated tree heights (dashed steps) and cumulative probability function of the Normal model (solid line)."
# Cumulative probability of the model on the same grid of heights.
model_cumulative_curve = data.frame(
  height = height_grid,
  cumulative = pnorm(q = height_grid, mean = 22, sd = 4)
)

ggplot() +
  geom_line(
    data = model_cumulative_curve,
    aes(x = height, y = cumulative, colour = "Model", linetype = "Model"),
    linewidth = 0.8
  ) +
  stat_ecdf(
    data = height_data,
    aes(x = height, colour = "Sample", linetype = "Sample"),
    geom = "step",
    linewidth = 0.8
  ) +
  scale_colour_manual(values = c(Model = "#D55E00", Sample = "#0072B2")) +
  scale_linetype_manual(values = c(Model = "solid", Sample = "dashed")) +
  labs(
    x = "Tree height (m)",
    y = "Cumulative probability or relative frequency",
    colour = NULL,
    linetype = NULL
  ) +
  theme_classic()
```

:::: {.callout-important title="Practice: bins and seeds"}

You simulate 200 tree heights from the Normal model and draw a density-scaled histogram.
Would a different seed produce the same histogram? Would a different bin width change the
mean of the sample?

::: {.callout-tip title="Solution" collapse="true"}

A different seed produces a different sample, so the histogram will change (although with
200 trees its overall shape should remain similar). The bin width only changes how the
sample is displayed. The values in the sample remain the same and so does their mean.

:::
::::

# Parameters, estimators and estimates {#sec-estimates}

A **parameter** is a quantity that defines a model distribution, such as the probability
$p$ of a Binomial distribution or the mean $\mu$ and standard deviation $\sigma$ of a
Normal distribution. So far we have chosen the values of the parameters ourselves. With
real data the parameters are unknown and we want to learn about them from a sample. An
**estimator** is a rule (i.e., a function) that we apply to a sample to calculate a value
for a parameter. Since we treat the observations as random variables, an estimator is also
a random variable: a different sample would give a different result. An **estimate** is the
value that the estimator returns for the sample that we actually observed. We mark
estimators and estimates with a hat, e.g., $\hat\lambda$ for a parameter $\lambda$.

To illustrate these concepts we will use counts of seedlings in quadrats across a
restoration site. Let $X_i$ be the number of seedlings in quadrat $i$. We model the counts
with a [Poisson distribution](../Supplements/distributions.qmd#sec-poisson),

$$
X_i \sim \operatorname{Poisson}(\lambda),
$$

where $\lambda$ is the expected number of seedlings per quadrat. We assume that the counts
are i.i.d. (see [Chapter 2](../Chapter_2/Theory.qmd#sec-joint-independence)). That is, all
quadrats have the same area and the same expected count, and the count in one quadrat does
not change the probabilities for the others. The parameter $\lambda$ is unknown.

::: {.callout-note title="Parameter, estimator, and estimate"}
A **parameter** is a fixed, unknown quantity that defines the model distribution, such as
$\lambda$. An **estimator** is a rule applied to a random sample, so its result varies
across samples. An **estimate** is the number obtained by applying that rule to one
particular sample.
:::

Suppose that ten quadrats give us the following counts (these are illustrative values, not
field data):

```{r}
# Seedling counts from ten illustrative quadrats.
seedling_counts = c(3, 5, 2, 7, 4, 3, 4, 1, 6, 5)
# Mean of the sample (seedlings per quadrat).
mean(seedling_counts)
```

The mean of the sample is 4 seedlings per quadrat. As $\lambda$ is the mean of the Poisson
distribution, the mean of the sample seems a reasonable estimate of $\lambda$. The method
of moments turns this intuition into a general procedure.

## Method of moments

The **method of moments** constructs estimators by equating expectations of the model
distribution to the same expectations of the empirical distribution of the sample
(@sec-samples), and solving the resulting equations for the parameters. The expectations
that we equate are called **moments**: $E[X]$ is the first moment, $E[X^2]$ the second
moment, and so on. The justification for the method is the asymptotic convergence of the
empirical distribution: if the moments of the empirical distribution converge to the
moments of the model as sample size grows, then a parameter calculated from the former
should be close to the parameter of the model when the sample is large enough.

For a Poisson distribution the first moment is $E[X_i]=\lambda$. The first moment of the
empirical distribution is the mean of the $n$ counts in the sample,

$$
\overline X = \frac{1}{n}\sum_{i=1}^{n}X_i.
$$

Equating the two gives the method-of-moments estimator of $\lambda$:

$$
\hat\lambda = \overline X.
$$

Applied to our ten quadrats, the estimate is $\hat\lambda=4$ seedlings per quadrat. A model
with two parameters requires two moments, as we will see in @sec-two_parameters.

Maximum likelihood ([Chapter 4](../Chapter_4/Theory.qmd)) is the method of estimation that
we use in the rest of the course. We start with the method of moments because its
estimators are short formulae, which makes it easier to study how estimates vary across
samples. The concepts introduced in the rest of this chapter (sampling distribution,
standard error and confidence interval) apply to any estimator.

:::: {.callout-important title="Practice: estimate from quadrats"}

A second sample of five quadrats has counts 2, 4, 3, 5, and 1. What are the parameter, the
estimator, and the estimate under a Poisson model? Give the estimate with its units.

::: {.callout-tip title="Solution" collapse="true"}

The parameter is $\lambda$, the expected number of seedlings per quadrat. The estimator is
$\hat\lambda=\overline X$. For these five counts, the estimate is $(2+4+3+5+1)/5=3$
seedlings per quadrat.

:::
::::

# Sampling distributions {#sec-sampling_distribution}

The **sampling distribution** of an estimator is the distribution of its estimates across
repeated samples of the same size from the same model. It tells us how much an estimate
could change if we repeated the study, so we will use it to quantify the uncertainty in an
estimate (@sec-standard_errors and @sec-confidence_intervals). Note that there are two
distributions involved. The Poisson distribution describes the variation of the counts
across individual quadrats. The sampling distribution of $\hat\lambda$ describes the
variation of the mean of $n$ quadrats across repeated samples.

With real data we only have one sample, so we cannot observe the sampling distribution.
However, we can approximate it with a simulation, where we choose the parameter of the
model and generate as many samples as we want. Let's assume a Poisson model with
$\lambda=4$ and a sample size of 30 quadrats, and repeat the sampling 2,000 times. For each
sample we calculate the estimate of $\lambda$:

```{r}
# Parameter of the model, sample size and number of repeated samples.
set.seed(seed = 401)
true_lambda = 4
n_quadrats = 30
n_studies = 2000

# replicate() runs the expression n times and collects the results in a vector.
# Each run simulates one sample of 30 quadrats and returns its estimate of lambda.
lambda_estimates = replicate(
  n = n_studies,
  expr = mean(rpois(n = n_quadrats, lambda = true_lambda))
)

# Mean and standard deviation of the 2,000 estimates.
mean(lambda_estimates)
sd(lambda_estimates)
```

The estimates have a mean of approximately 4.00 and a standard deviation of approximately
0.37 seedlings per quadrat. Their distribution is shown in @fig-sampling_distribution. The
horizontal axis is a mean count per quadrat, which is why its values are not restricted to
integers.

```{r}
#| echo: false
#| label: fig-sampling_distribution
#| fig-cap: "Estimates of the mean seedling count from 2,000 simulated samples of 30 quadrats each. The vertical line is the value of the parameter used in the simulation."
# Plot the simulated sampling distribution and mark the parameter of the model.
ggplot(data.frame(estimate = lambda_estimates), aes(x = estimate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 0.1,
    center = true_lambda,
    fill = "grey90",
    colour = "grey40"
  ) +
  geom_vline(xintercept = true_lambda, colour = "#D55E00", linewidth = 0.8) +
  labs(
    x = "Estimated seedlings per quadrat",
    y = "Density"
  ) +
  theme_classic()
```

## Bias and variance

We describe a sampling distribution with its mean and its variance, which tell us whether
the estimator is centred on the parameter and how much the estimates vary around that
centre. The **bias** of an estimator is the difference between the mean of its sampling
distribution and the parameter that it estimates. Using $\theta$ for a parameter in general
and $\hat\theta$ for its estimator:

$$
\operatorname{Bias}(\hat\theta)=E[\hat\theta]-\theta.
$$

An estimator is **unbiased** when its bias is zero. For i.i.d. Poisson counts,
$E[\hat\lambda]=\lambda$, so the mean of the sample is an unbiased estimator of $\lambda$.
Bias is a property of the whole sampling distribution: individual estimates of an unbiased
estimator still fall above or below the parameter (@fig-sampling_distribution).

The **variance of an estimator** is the variance of its sampling distribution. For the mean
of $n$ i.i.d. Poisson counts,

$$
\operatorname{Var}(\hat\lambda)=\frac{\lambda}{n}.
$$

In other words, the variance of the estimator decreases as the sample size grows. The
variance of the counts in individual quadrats remains $\lambda$; what decreases is the
variation among the means of repeated samples. We can compare these results with the
simulation:

```{r}
# Compare the simulated bias and variance with the variance derived from the model.
c(
  simulated_bias = mean(lambda_estimates) - true_lambda,
  simulated_variance = var(lambda_estimates),
  model_variance = true_lambda / n_quadrats
)
```

The simulated bias is close to zero (approximately $-0.005$) and the simulated variance
(approximately $0.136$) is close to $4/30 \approx 0.133$. They do not match exactly because
2,000 repetitions are also a finite sample, in this case a sample of estimates.

## Mean squared error

The **mean squared error** (MSE) combines bias and variance into one measure of the
quality of an estimator. It is the expected squared distance between the estimator and the
parameter:

$$
\operatorname{MSE}(\hat\theta)
= E[(\hat\theta-\theta)^2]
= \operatorname{Var}(\hat\theta)
  + \operatorname{Bias}(\hat\theta)^2.
$$

For an unbiased estimator the MSE equals its variance, so for the mean of Poisson counts
it is $\lambda/n$. In the simulation:

```{r}
# Average squared distance between the estimates and the parameter.
mean((lambda_estimates - true_lambda)^2)
```

The MSE is useful to compare estimators of the same parameter. A biased estimator can have
a smaller MSE than an unbiased one if its variance is sufficiently smaller. We will see an
example with two estimators of a variance in @sec-two_parameters and in the practical.

:::: {.callout-important title="Practice: what varies?"}

Two independent studies each sample 30 quadrats and report means of 3.7 and 4.2 seedlings
per quadrat. What varies between the studies? Can we conclude from these two estimates that
the estimator is biased?

::: {.callout-tip title="Solution" collapse="true"}

The samples and therefore the estimates vary; the parameter $\lambda$ of the model is the
same in both studies. Bias is the difference between the mean of the sampling distribution
and the parameter, and two estimates are not enough to describe a sampling distribution
(also, in a real study we would not know $\lambda$).

:::
::::

# Standard errors {#sec-standard_errors}

The **standard error** (SE) of an estimator is the standard deviation of its sampling
distribution:

$$
\operatorname{SE}(\hat\theta)
= \sqrt{\operatorname{Var}(\hat\theta)}.
$$

We prefer it to the variance when reporting uncertainty because it has the same units as
the estimate. For the mean of Poisson counts, $\operatorname{SE}(\hat\lambda)=\sqrt{\lambda/n}$.
In the simulation above this is $\sqrt{4/30} \approx 0.365$, which is close to the standard
deviation of the 2,000 estimates (0.37).

With real data we do not know $\lambda$, so we cannot calculate the standard error. Instead
we replace the parameter with its estimate:

$$
\widehat{\operatorname{SE}}(\hat\lambda)
= \sqrt{\frac{\hat\lambda}{n}}.
$$

This is known as a **plug-in estimate**, and it means that the standard error that we
report is itself an estimate. For the ten quadrats of @sec-estimates:

```{r}
# Estimate of lambda and number of quadrats in the sample.
lambda_hat = mean(seedling_counts)
n_counts = length(seedling_counts)
# Plug the estimate into the formula of the standard error.
estimated_se = sqrt(lambda_hat / n_counts)
estimated_se
```

The estimated standard error is approximately 0.63 seedlings per quadrat. That is, if we
repeated the study with ten quadrats, we would expect the estimates of $\lambda$ to vary
with a standard deviation of about 0.63 seedlings per quadrat. This calculation relies on
the Poisson model: if the counts are overdispersed with respect to the Poisson distribution
(see [Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions)), this formula will
underestimate the variation across samples.

::: {.callout-warning title="Standard deviation and standard error"}
These two terms are often confused. The **standard deviation** describes the variation
among the observations (here, the counts in individual quadrats). For our ten quadrats
`sd(seedling_counts)` is approximately 1.83 seedlings per quadrat. The **standard error**
describes the variation of an estimate across repeated samples (here, 0.63 seedlings per
quadrat for the mean). A larger sample reduces the standard error, but it does not reduce
the standard deviation of the observations.
:::

:::: {.callout-important title="Practice: change the sample size"}

If the mean of the model is 4 seedlings per quadrat, what happens to the standard error of
$\hat\lambda$ when the sample grows from 25 to 100 quadrats? Does the standard deviation of
the individual counts change?

::: {.callout-tip title="Solution" collapse="true"}

The standard error changes from $\sqrt{4/25}=0.4$ to $\sqrt{4/100}=0.2$ seedlings per
quadrat. That is, a sample four times larger halves the standard error. The standard
deviation of an individual count remains $\sqrt{4}=2$ seedlings per quadrat.

:::
::::

# Confidence intervals {#sec-confidence_intervals}

A **confidence interval** is a range of values for a parameter, calculated from a sample
with a procedure that has a known behaviour across repeated samples. The **confidence
level** (e.g., 95%) is the fraction of repeated samples for which the interval calculated
with that procedure would contain the parameter. We will use confidence intervals to report
the uncertainty of parameter estimates in the rest of the course.

A simple procedure is available when the sampling distribution of the estimator is
approximately Normal (@sec-large_samples discusses when this is the case). An approximate
95% confidence interval is then

$$
\hat\theta \pm z_{0.975}\,\widehat{\operatorname{SE}}(\hat\theta),
$$

where $z_{0.975} \approx 1.96$ is the $0.975$ quantile of the standard Normal distribution
(mean 0 and standard deviation 1). This quantile leaves a probability of $0.025$ in each
tail, hence 95% in between. This type of interval is known as a **Wald interval**. For the
ten quadrats:

```{r}
# 0.975 quantile of the standard Normal distribution.
critical_value = qnorm(p = 0.975)
# Lower and upper endpoints of the interval.
interval = lambda_hat + c(-1, 1) * critical_value * estimated_se
interval
```

The interval goes from approximately 2.76 to 5.24 seedlings per quadrat. It is an interval
for the parameter $\lambda$ (the mean count), so it says nothing about the range of counts
in individual quadrats.

::: {.callout-note title="What 95% confidence means"}
The parameter is fixed and the interval is random: its endpoints change from sample to
sample. If we repeated the study many times and calculated an interval from each sample
with the same procedure, about 95% of those intervals would contain the parameter. Once an
interval has been calculated, it either contains the parameter or it does not, so we do not
say that the parameter has a probability of 95% of being inside it.
:::

## Coverage checked by simulation

The fraction of repeated samples for which the interval contains the parameter is called
the **coverage** of the procedure. The Wald interval relies on an approximation (and on a
plug-in standard error), so its coverage may differ from the confidence level that we
intended. We can check this with a simulation, where we know $\lambda$. For each simulated
sample we calculate the interval and record whether it contains the parameter:

```{r}
# Parameter of the model, sample size and number of repeated samples.
set.seed(seed = 402)
true_lambda = 4
n_quadrats = 30
n_studies = 2000
critical_value = qnorm(p = 0.975)

# Each repetition returns TRUE if the interval of that sample contains lambda.
covered = replicate(
  n = n_studies,
  expr = {
    # Simulate one sample and calculate its estimate and standard error.
    counts = rpois(n = n_quadrats, lambda = true_lambda)
    estimate = mean(counts)
    se = sqrt(estimate / n_quadrats)
    # Calculate the endpoints and check whether lambda lies between them.
    lower = estimate - critical_value * se
    upper = estimate + critical_value * se
    lower <= true_lambda && true_lambda <= upper
  }
)

# The mean of TRUE and FALSE values is the fraction of TRUE values.
mean(covered)
```

The simulated coverage is 0.944, close to the intended 0.95. This result is specific to a
mean of 4 and 30 quadrats. When the expected total count $n\lambda$ is small, the sampling
distribution of $\hat\lambda$ is far from Normal and the coverage of the Wald interval can
be noticeably lower than 95%. The lower endpoint can even be negative, which is not a
possible value for the mean of a Poisson distribution. The practical includes an exercise
on the coverage of an interval when the data are skewed.

:::: {.callout-important title="Practice: interpret an interval"}

A study reports a 95% confidence interval of 2.8 to 5.2 seedlings per quadrat. What does
the 95% refer to? Which assumptions would you check before using this interval?

::: {.callout-tip title="Solution" collapse="true"}

The 95% refers to the coverage of the procedure across repeated samples: about 95% of the
intervals calculated in this way would contain the parameter. For a Wald interval based on
a Poisson model, we would check whether the counts can be treated as i.i.d. Poisson
(e.g., whether they are overdispersed) and whether the sample is large enough for the
Normal approximation.

:::
::::

# Estimation of two parameters {#sec-two_parameters}

The concepts above apply in the same way to continuous measurements and to models with more
than one parameter. We return to the Normal model of tree height from
[Chapter 2](../Chapter_2/Theory.qmd#sec-continuous_four_faces). Let $X_i$ be the height
in metres of tree $i$ and assume that the heights are i.i.d. with

$$
X_i \sim \operatorname{Normal}(\mu,\sigma^2),
$$

where $\mu$ is the mean of the model (in m) and $\sigma^2$ its variance (in m²). Remember
that the argument `sd` of `rnorm()` and related functions takes the standard deviation
$\sigma$.

## Method of moments for the mean and the variance

The model has two parameters, so the method of moments needs two equations. The first
moment is $E[X]=\mu$, which gives $\hat\mu=\overline X$ as before. The second moment is
$E[X^2]=\mu^2+\sigma^2$. Equating it to the mean of the squared heights in the sample and
replacing $\mu$ by $\overline X$ gives

$$
\widehat{\sigma^2}_{\mathrm{MM}}
= \frac{1}{n}\sum_{i=1}^{n}X_i^2 - \overline X^2
= \frac{1}{n}\sum_{i=1}^{n}(X_i-\overline X)^2.
$$

That is, the method-of-moments estimators of $\mu$ and $\sigma^2$ are the mean and the
variance of the empirical distribution (@sec-samples). We can calculate them for a
simulated sample of 25 trees and compare the variance with the result of `var()`:

```{r}
# Simulate the heights (in m) of 25 trees from the Normal model.
set.seed(seed = 403)
tree_heights = rnorm(n = 25, mean = 22, sd = 4)

# Mean of the sample and the two estimates of the variance.
height_mean = mean(tree_heights)
variance_mm = mean((tree_heights - height_mean)^2)
variance_with_var = var(tree_heights)

c(
  mean_m = height_mean,
  variance_mm_m2 = variance_mm,
  variance_with_var_m2 = variance_with_var
)
```

The estimate of the mean is approximately 22.0 m. The two estimates of the variance differ:
18.3 m² with the method of moments and 19.1 m² with `var()`. We already found this
difference in @sec-samples and can now explain it. The method-of-moments estimator divides
the sum of squared deviations by $n$, whereas `var()` divides it by $n-1$. For i.i.d.
observations, the expectations of the two estimators are

$$
E[\widehat{\sigma^2}_{\mathrm{MM}}]
= \frac{n-1}{n}\sigma^2,
\qquad
E[s^2]=\sigma^2,
$$

where $s^2$ denotes the estimator calculated by `var()`. So the method-of-moments estimator
of the variance is biased: on average it underestimates $\sigma^2$ by $\sigma^2/n$. This
happens because the deviations are calculated with respect to $\overline X$, which is
estimated from the same sample and is therefore closer to the observations than $\mu$.
Dividing by $n-1$ removes the bias, and that is the estimator that most people call the
sample variance. The bias of the method-of-moments estimator decreases as sample size
grows ($16/25 = 0.64$ m² in this example, but only $0.16$ m² with 100 trees). Note that
being unbiased does not imply having the smallest MSE; in the practical you will compare
the MSE of the two estimators with a simulation.

## Confidence interval for the mean

The standard error of the mean of $n$ i.i.d. observations is $\sigma/\sqrt n$. As before,
we replace the unknown parameter with an estimate, $s/\sqrt n$, where $s$ is the square
root of $s^2$ (i.e., `sd(tree_heights)`). For a Normal model there is a procedure that
takes into account that $\sigma$ was estimated from the same sample. The 95% confidence
interval for $\mu$ is

$$
\overline X \pm t_{n-1,\,0.975}\frac{s}{\sqrt n},
$$

where $t_{n-1,\,0.975}$ is the $0.975$ quantile of a Student's $t$ distribution with $n-1$
degrees of freedom (`qt()` in R). This distribution resembles a standard Normal
distribution with heavier tails, so the interval is somewhat wider than a Wald interval,
and it approaches the standard Normal distribution as $n$ grows. For i.i.d. Normal
observations the coverage of this interval is exactly 95%. In R:

```{r}
# Sample size and estimated standard error of the mean height.
n_trees = length(tree_heights)
height_se = sd(tree_heights) / sqrt(n_trees)
# 0.975 quantile of the t distribution with n - 1 degrees of freedom.
t_critical = qt(p = 0.975, df = n_trees - 1)
# Lower and upper endpoints of the interval.
height_interval = height_mean + c(-1, 1) * t_critical * height_se
height_interval
```

The estimated standard error is approximately 0.87 m, the quantile is 2.06 (instead of
1.96) and the interval goes from approximately 20.2 to 23.8 m. This is an interval for the
mean height $\mu$. The range that contains 95% of the individual trees is a different
quantity, which depends on $\sigma$ and not on the standard error (for this model, from
`qnorm(p = 0.025, mean = 22, sd = 4)` to `qnorm(p = 0.975, mean = 22, sd = 4)`, i.e.,
approximately 14.2 to 29.8 m). When the measurements are skewed, as is common for body
mass, the coverage of the $t$ interval is only approximate. You will investigate this in
the practical.

:::: {.callout-important title="Practice: two kinds of spread"}

Which R expression estimates the variation among the 25 tree heights, and which one
estimates the standard error of their mean? Why does the confidence interval for the mean
use the second one?

::: {.callout-tip title="Solution" collapse="true"}

`sd(tree_heights)` estimates the variation among trees and
`sd(tree_heights) / sqrt(length(tree_heights))` estimates the standard error of the mean.
The interval describes the uncertainty in the estimate of the mean height, so it uses the
variation of the mean across repeated samples.

:::
::::

# Consistency and asymptotic normality {#sec-large_samples}

In @sec-samples we saw that the mean, variance and quantiles of the empirical distribution
converge to those of the model as sample size grows. Method-of-moments estimators are
calculated from the moments of the empirical distribution, so they inherit this behaviour.
An estimator $\hat\theta_n$ calculated from a sample of size $n$ is **consistent** if it
converges in probability to the parameter. That is, for any positive margin $\varepsilon$,

$$
\lim_{n \to \infty}
P\left(\left|\hat\theta_n - \theta\right| \ge \varepsilon\right)=0.
$$

This is the same statement that we made for relative frequencies, now applied to an
estimator, and it should be read in the same way: it describes the probability of large
errors and not the ordering of the estimates of two particular samples. The mean of i.i.d.
Poisson counts is a consistent estimator of $\lambda$. The method-of-moments estimator of
the variance of the Normal model is also consistent, although it is biased for any finite
sample.

The standard error tells us how fast an estimator approaches the parameter. For the mean
of Poisson counts it is $\sqrt{\lambda/n}$, so we need a sample four times larger to halve
the standard error.

Large samples have a second consequence. The mean of $n$ i.i.d. observations from a
distribution with finite variance has a sampling distribution that approaches a Normal
distribution as $n$ grows, regardless of the distribution of the individual observations.
This result is the **central limit theorem**, and an estimator with this behaviour is said
to be **asymptotically normal**. For the mean of Poisson counts:

$$
\hat\lambda \overset{\mathrm{approx}}{\sim}
\operatorname{Normal}\left(\lambda,\frac{\lambda}{n}\right)
\quad\text{for sufficiently large }n.
$$

This is the justification for the Wald interval in @sec-confidence_intervals. How large
$n$ needs to be depends on the model. In @fig-asymptotic_normality the sampling
distribution of $\hat\lambda$ for samples of 5 quadrats is discrete (the mean of five
counts can only change in steps of $0.2$, so each bar is one possible value) and slightly
asymmetric, with a longer tail to the right, whereas for 100 quadrats the Normal
approximation is close. Note that the counts in individual
quadrats still follow a Poisson distribution in both cases; it is the distribution of their
mean that approaches a Normal distribution.

```{r}
#| echo: false
#| label: fig-asymptotic_normality
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Sampling distributions of the mean seedling count for samples of 5 and 100 quadrats (2,000 simulated samples each) under a Poisson model with a mean of 4. The curves are the Normal approximations."
# Simulate the sampling distribution of the mean for two sample sizes.
set.seed(seed = 404)
true_lambda = 4
sample_sizes = c(5, 100)

# Draw the simulated sampling distribution with its Normal approximation.
plot_sampling_distribution = function(sample_size, bin_width) {
  # One estimate of lambda per simulated sample.
  estimates = replicate(
    n = 2000,
    expr = mean(rpois(n = sample_size, lambda = true_lambda))
  )
  # Normal approximation with the standard error derived from the model.
  model_se = sqrt(true_lambda / sample_size)
  estimate_grid = seq(
    from = true_lambda - 4 * model_se,
    to = true_lambda + 4 * model_se,
    length.out = 200
  )
  approximation = data.frame(
    estimate = estimate_grid,
    density = dnorm(x = estimate_grid, mean = true_lambda, sd = model_se)
  )
  ggplot(data.frame(estimate = estimates), aes(x = estimate)) +
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = bin_width,
      center = true_lambda,
      fill = "grey90",
      colour = "grey40"
    ) +
    geom_line(
      data = approximation,
      aes(x = estimate, y = density),
      colour = "#D55E00",
      linewidth = 0.8
    ) +
    labs(
      title = paste(sample_size, "quadrats"),
      x = "Estimated seedlings per quadrat",
      y = "Density"
    ) +
    theme_classic()
}

plot_sampling_distribution(sample_size = 5, bin_width = 0.2) |
  plot_sampling_distribution(sample_size = 100, bin_width = 0.05)
```

Consistency, coverage and asymptotic normality are all derived under the assumed model, in
our case i.i.d. Poisson counts. In a real restoration site the expected count may change
with habitat, seedlings may be clustered, or some seedlings may be missed. A larger sample
does not correct for this. Instead, we need a model that describes those sources of
variation, such as the distributions for overdispersed counts in
[Chapter 2](../Chapter_2/Theory.qmd#sec-discrete_distributions) or the models for grouped
observations in [Chapter 7](../Chapter_7/Theory.qmd).

:::: {.callout-important title="Practice: two studies of different size"}

A study of 20 quadrats estimates 3.9 seedlings per quadrat and another study of 100
quadrats estimates 4.2. The mean of the model is 4. Does this contradict the consistency
of the estimator?

::: {.callout-tip title="Solution" collapse="true"}

No. Consistency states that the probability of an error larger than any given margin
decreases towards zero as sample size grows. It does not state that every larger sample is
closer to the parameter than every smaller one, so the smaller study can produce the
closer estimate by chance.

:::
::::

# Summary

A sample is treated as a collection of random draws from a model distribution. The relative
frequencies of a sample of counts, or the density-scaled histogram of a sample of
continuous measurements, are compared with the probability mass or density of the model.
The empirical distribution assigns each observation its relative frequency, and its mean,
variance, quantiles and cumulative distribution converge to those of the model as sample
size grows.

An estimator is a rule that calculates a value for a parameter from a sample, and an
estimate is its result for one sample. The method of moments constructs estimators by
equating moments of the model to moments of the empirical distribution. The sampling
distribution of an estimator describes how its estimates vary across repeated samples. Its
mean determines the bias, its standard deviation is the standard error, and the mean
squared error combines both.

A confidence interval is calculated with a procedure that contains the parameter in a
given fraction of repeated samples (its coverage). Wald intervals rely on the asymptotic
normality of the estimator, and for the mean of a Normal model the $t$ interval has exact
coverage. Simulation allows us to check bias, standard errors and coverage, because we
know the parameters that generated the samples. All these properties are derived under the
assumed model.

In [Chapter 4](../Chapter_4/Theory.qmd) we introduce maximum likelihood, a general method
of estimation that we will use in the rest of the course. The estimators will change, but
we will describe their uncertainty with the same concepts.

# Chapter quiz

The questions below follow the order of the chapter. Try each question before opening its
solution.

:::: {.callout-important title="Question 1: moth emergence"}

In a sample of 30 host plants, moths emerged from 9. Calculate the relative frequency
of emergence. How does this relate to the probability of emergence for all relevant host
plants?

::: {.callout-tip title="Answer" collapse="true"}

The relative frequency is $9/30=0.3$. It is a sample-based estimate of the probability,
not necessarily the exact model probability. Another sample could produce a
different relative frequency.

:::
::::

:::: {.callout-important title="Question 2: caddisfly larvae"}

You sample the number of caddisfly larvae in many equal-sized stream sections. As the
number of sections grows, what do you generally expect to happen to the empirical
distribution relative to the distribution that generated the data? Is exact agreement
expected for a finite sample?

::: {.callout-tip title="Answer" collapse="true"}

With more independent observations from the same distribution, the empirical distribution
generally becomes closer to the generating distribution. Exact agreement is not expected
in a finite sample, and individual simulations can vary.

:::
::::

:::: {.callout-important title="Question 3: a sample and its model"}

Why might a density-scaled histogram of 100 simulated tree heights differ from the density
of the model used to generate them?

::: {.callout-tip title="Answer" collapse="true"}

The relative frequency of each bin varies from sample to sample around the probability
that the model assigns to that bin, and the shape of the histogram also depends on the bin
width that we choose.

:::
::::

:::: {.callout-important title="Question 4: an estimate"}

The mean of the counts from 20 quadrats is 5.1 seedlings per quadrat. Under a Poisson
model, name the parameter, the estimator, and the estimate.

::: {.callout-tip title="Answer" collapse="true"}

The parameter is the mean $\lambda$ of the model. The estimator is
$\hat\lambda=\overline X$. Its estimate for these counts is 5.1 seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 5: a histogram of means"}

A histogram contains the mean of each of 1,000 simulated samples. Does it describe the
empirical distribution of individual observations or the sampling distribution of an
estimator?

::: {.callout-tip title="Answer" collapse="true"}

It approximates the sampling distribution of the mean. Each value in the histogram is one
estimate calculated from a whole sample.

:::
::::

:::: {.callout-important title="Question 6: bias and variance"}

The mean of the sampling distribution of an estimator equals the parameter, but individual
estimates vary considerably. Is the estimator biased? Is its variance zero?

::: {.callout-tip title="Answer" collapse="true"}

It is unbiased, because the mean of its sampling distribution equals the parameter. Its
variance is larger than zero, because its estimates vary across samples.

:::
::::

:::: {.callout-important title="Question 7: standard error"}

For 100 i.i.d. quadrats under a Poisson model with $\lambda=4$, what is the standard error
of the estimated mean count? What is the standard deviation of the count in one quadrat?

::: {.callout-tip title="Answer" collapse="true"}

The standard error is $\sqrt{4/100}=0.2$ seedlings per quadrat. The standard deviation of
one count is $\sqrt{4}=2$ seedlings per quadrat.

:::
::::

:::: {.callout-important title="Question 8: meaning of a confidence interval"}

Does a calculated 95% confidence interval assign a probability of 95% to the parameter
lying inside it? Explain what the 95% describes.

::: {.callout-tip title="Answer" collapse="true"}

No. The 95% describes the procedure: across repeated samples, about 95% of the intervals
calculated in this way contain the parameter (if the coverage of the procedure equals its
confidence level). The parameter is fixed, so one calculated interval either contains it or
it does not.

:::
::::

:::: {.callout-important title="Question 9: estimating a variance"}

Which divisor does the method-of-moments estimator of the variance use, and which one does
`var()` use? What happens to the bias of the former as sample size grows?

::: {.callout-tip title="Answer" collapse="true"}

The method-of-moments estimator divides by $n$ and `var()` divides by $n-1$. For i.i.d.
observations, the bias of the method-of-moments estimator is $-\sigma^2/n$, which
approaches zero as $n$ grows.

:::
::::
