---
title: "Introduction to Probability for Continuous Data"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

Ecologists measure many different kinds of variables. We count seedlings, nests, or animal detections, but we also measure tree diameter, plant height, body mass, temperature, soil moisture, biomass, and vegetation cover.

The previous chapter focused on **discrete random variables**. These variables take separate, countable values such as $0,1,2,\ldots$.

This chapter introduces **continuous random variables**. A continuous variable can take any value within an interval. For example, a tree can have a diameter of 20 cm, 20.1 cm, 20.15 cm, or any other value within a plausible range.

We will focus on three continuous distributions:

- the Normal distribution;
- the Lognormal distribution;
- the Beta distribution.

These distributions are useful for different types of ecological measurements:

| Distribution | Typical ecological use | Range |
|---|---|---:|
| Normal | Symmetric measurements and measurement error | All real values |
| Lognormal | Positive, right-skewed measurements | Positive values |
| Beta | Proportions and fractional measurements | Between 0 and 1 |

All examples use base R.

## Learning goals

After studying this chapter, you should be able to:

- distinguish continuous and discrete random variables;
- explain the difference between probability density and probability mass;
- calculate probabilities over intervals;
- interpret density, cumulative probability, samples, and quantiles;
- use Normal, Lognormal, and Beta distributions in R;
- recognize when each distribution may be useful in ecology;
- interpret histograms and density plots;
- distinguish joint, marginal, and conditional distributions;
- interpret scatterplots and contour plots for two continuous variables.

# Continuous random variables

A **continuous random variable** is a variable that can take any value within an interval.

Examples include:

- the diameter of a tree;
- the height of a plant;
- the mass of an animal;
- the temperature at a field site;
- the concentration of nitrogen in soil;
- the proportion of a quadrat covered by vegetation.

Let

$$
X=\text{diameter of a randomly selected tree}.
$$

Unlike a count of seedlings, $X$ is not restricted to integer values. It could be 25 cm, 25.1 cm, 25.12 cm, or any other value within a plausible range.

In practice, measurements are recorded with limited precision. A value recorded as 25.1 cm may represent an underlying value somewhere between 25.05 cm and 25.15 cm. Treating the underlying measurement as continuous is usually a useful approximation.

## The probability of an exact value

The most important mathematical property of a continuous random variable is

$$
P(X=x)=0.
$$

This does not mean that a value such as 25.1 cm cannot be observed. It means that, for
continuous variables, probability can only be assigned to intervals rather than to individual
exact values.

For example, we may ask for the probability that a tree has a diameter between 20 and 25 cm:

$$
P(20<X\leq25).
$$

For a continuous variable, changing the boundaries from $20<X\leq25$ to $20\leq X\leq25$ does not change the probability because the probability of observing exactly 20 or exactly 25 is zero.

This differs from a discrete variable. For a discrete variable, the probability of exactly 5 seedlings may be positive:

$$
P(X=5)>0.
$$

# Density and probability

## Probability mass for discrete variables

For a discrete random variable, the probability mass function assigns probability to individual possible values:

$$
p_X(x)=P(X=x).
$$

For example, a binomial distribution may assign a probability of 0.15 to exactly five surviving seedlings.

The probabilities of all possible values add to one:

$$
\sum_x p_X(x)=1.
$$

## Probability density for continuous variables

For a continuous random variable, the equivalent function is the **probability density function**, written as $f_X(x)$.

A density is not itself a probability. Probabilities are calculated from areas under the density curve:

$$
P(a<X\leq b)=\int_a^b f_X(x)\,dx.
$$

The total area under the density curve is one:

$$
\int_{-\infty}^{\infty}f_X(x)\,dx=1.
$$

A density therefore has three important properties:

- $f_X(x)\geq0$;
- the total area under the curve is one;
- the probability of an interval is the area under the curve over that interval.

## Densities can be greater than one

Probabilities cannot be greater than one, but densities can be greater than one.

For example, consider a Normal distribution with a very small standard deviation:

```{r}
dnorm(
  1,
  mean = 1,
  sd = 0.01
)
```

The result is much greater than one. This is not a problem because the result is a density, not a probability.

The density is high because the distribution is concentrated in a very narrow interval. The total area under the curve is still one.

```{r}
x = seq(0.95, 1.05, length.out = 500)
density_values = dnorm(x, mean = 1, sd = 0.01)

plot(
  x,
  density_values,
  type = "l",
  lwd = 2,
  xlab = "Measurement",
  ylab = "Probability density",
  main = "A concentrated continuous distribution"
)
```

## Density and measurement units

Density values also depend on the units of measurement.

Suppose tree diameter is measured in metres. The density has units of probability per metre. If the same measurements are converted to centimetres, the density has units of probability per centimetre.

The numerical height of the density curve changes when the units change, but interval probabilities do not.

This is another reason that a density should not be interpreted as a probability at a point.

# The four faces of a continuous distribution

As in the previous chapter, a continuous distribution can be viewed in four complementary ways:

1. probability density;
2. cumulative probability;
3. random samples;
4. quantiles.

The R functions use the same naming system as before.

| Purpose | R function |
|---|---|
| Probability density | `d*` |
| Cumulative probability | `p*` |
| Quantile | `q*` |
| Random sample | `r*` |

For example, for the Normal distribution:

- `dnorm()` calculates density;
- `pnorm()` calculates cumulative probability;
- `qnorm()` calculates quantiles;
- `rnorm()` generates random observations.

## 1. Probability density

The density function describes the relative concentration of probability at different values.

For a continuous random variable, the probability of an interval is obtained from the area under the density curve.

For example:

```{r}
x = seq(-4, 4, length.out = 500)
density_values = dnorm(x)

plot(
  x,
  density_values,
  type = "l",
  lwd = 2,
  xlab = "Value",
  ylab = "Probability density",
  main = "Normal probability density"
)
```

## 2. Cumulative probability

The cumulative distribution function is

$$
F_X(x)=P(X\leq x).
$$

For a continuous distribution, it can also be written as

$$
F_X(x)=\int_{-\infty}^{x}f_X(t)\,dt.
$$

The CDF gives the probability of observing a value less than or equal to $x$.

It is always between zero and one, and it never decreases as $x$ increases.

```{r}
cumulative_values = pnorm(x)

plot(
  x,
  cumulative_values,
  type = "l",
  lwd = 2,
  ylim = c(0, 1),
  xlab = "Value",
  ylab = "Cumulative probability",
  main = "Normal cumulative distribution"
)
```

For a continuous variable, upper-tail probabilities can be calculated using `lower.tail = FALSE`.

```{r}
pnorm(
  1.96,
  lower.tail = FALSE
)
```

This calculates the probability that a standard Normal random variable is greater than 1.96.

## 3. Sampling

A distribution can be used to generate random observations.

```{r}
set.seed(42)

sample_values = rnorm(
  n = 1000,
  mean = 10,
  sd = 2
)

head(sample_values)
```

Each simulated value represents one possible realization of the random variable.

The simulated values will not have exactly the specified mean and standard deviation. They are one random sample from the distribution.

## 4. Quantiles

A quantile is a value that divides a distribution according to cumulative probability.

The $\alpha$-quantile is the value $q_\alpha$ satisfying

$$
P(X\leq q_\alpha)=\alpha.
$$

For example, the 95th percentile of a standard Normal distribution is:

```{r}
qnorm(0.95)
```

The middle 50% of a Normal distribution can be described using the 25th and 75th percentiles:

```{r}
qnorm(c(0.25, 0.75))
```

Quantiles are especially useful when distributions are skewed. For a strongly right-skewed distribution, the mean and median may be very different.

# The Normal distribution

The Normal distribution is a symmetric continuous distribution described by two parameters:

- $\mu$: the mean;
- $\sigma$: the standard deviation.

Its density is

$$
f_X(x)=
\frac{1}{\sigma\sqrt{2\pi}}
\exp\left(
-\frac{(x-\mu)^2}{2\sigma^2}
\right).
$$

The mean and variance are

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\sigma^2.
$$

The Normal distribution is symmetric around its mean. The mean, median, and mode are equal.

## The standard deviation

The standard deviation controls the spread of the distribution.

A small standard deviation produces a narrow distribution. A large standard deviation produces a wider distribution.

```{r}
x = seq(-10, 20, length.out = 500)

plot(
  x,
  dnorm(x, mean = 5, sd = 1),
  type = "l",
  lwd = 2,
  ylim = c(0, 0.45),
  xlab = "Value",
  ylab = "Probability density",
  main = "Normal distributions with different standard deviations"
)

lines(
  x,
  dnorm(x, mean = 5, sd = 3),
  lwd = 2,
  lty = 2
)

legend(
  "topright",
  legend = c("sd = 1", "sd = 3"),
  lwd = 2,
  lty = c(1, 2),
  bty = "n"
)
```

## The 68–95–99.7 rule

For a Normal distribution:

- approximately 68% of observations lie within one standard deviation of the mean;
- approximately 95% lie within two standard deviations;
- approximately 99.7% lie within three standard deviations.

For example, if tree heights have mean 20 m and standard deviation 3 m, approximately 95% of heights lie between 14 m and 26 m:

$$
20-2(3)=14
$$

and

$$
20+2(3)=26.
$$

The rule is only an approximation and applies to distributions that are reasonably Normal.

## Standardization

A Normal variable can be converted to a standard Normal variable using

$$
Z=\frac{X-\mu}{\sigma}.
$$

The standard Normal distribution has mean zero and standard deviation one.

Standardization is useful because probabilities for all Normal distributions can be calculated using the same standard distribution.

## Example: tree height

Suppose tree height is approximately Normal with mean 22 m and standard deviation 4 m.

```{r}
height_mean = 22
height_sd = 4
```

The probability that a randomly selected tree is between 18 and 26 m tall is:

```{r}
pnorm(
  26,
  mean = height_mean,
  sd = height_sd
) -
  pnorm(
    18,
    mean = height_mean,
    sd = height_sd
  )
```

The probability that a tree is taller than 30 m is:

```{r}
pnorm(
  30,
  mean = height_mean,
  sd = height_sd,
  lower.tail = FALSE
)
```

The 95th percentile is:

```{r}
qnorm(
  0.95,
  mean = height_mean,
  sd = height_sd
)
```

We can simulate tree heights and compare them with the theoretical density:

```{r}
set.seed(10)

tree_heights = rnorm(
  1000,
  mean = height_mean,
  sd = height_sd
)

hist(
  tree_heights,
  breaks = 30,
  probability = TRUE,
  col = "grey80",
  border = "white",
  xlab = "Tree height (m)",
  main = "Simulated tree heights"
)

curve(
  dnorm(
    x,
    mean = height_mean,
    sd = height_sd
  ),
  add = TRUE,
  col = "red",
  lwd = 2
)
```

## Limitations of the Normal distribution

The Normal distribution allows all real values, including negative values.

This is reasonable for some measurements or errors, but not for variables such as:

- body mass;
- biomass;
- tree diameter;
- survival time;
- concentrations.

A Normal model may assign some probability to negative values for these variables. When the distribution is strongly right-skewed or values must be positive, a Lognormal or Gamma model may be more appropriate.

# The Lognormal distribution

A positive random variable $X$ has a Lognormal distribution when its logarithm is Normally distributed:

$$
\log(X)\sim\operatorname{Normal}(\mu,\sigma^2).
$$

Equivalently, if $Y=\log(X)$ is Normal, then $X=e^Y$ is Lognormal.

The Lognormal density is

$$
f_X(x)=
\frac{1}{x\sigma\sqrt{2\pi}}
\exp\left(
-\frac{(\log x-\mu)^2}{2\sigma^2}
\right),
\qquad x>0.
$$

The Lognormal distribution is useful for positive quantities that vary multiplicatively.

Examples include:

- tree diameter;
- animal body mass;
- plant biomass;
- population size;
- growth measurements;
- positive concentrations.

## Why multiplicative variation matters

Suppose plant size changes through repeated proportional growth. A plant may become 10% larger during one period, then 5% larger during another period.

The final size is obtained by multiplying growth factors. Taking logarithms converts multiplication into addition:

$$
\log(ab)=\log(a)+\log(b).
$$

If many independent effects multiply together, the logarithm of the final measurement may be approximately Normal. The original measurement is then approximately Lognormal.

## Mean and median

If

$$
\log(X)\sim\operatorname{Normal}(\mu,\sigma^2),
$$

then the median of $X$ is

$$
\operatorname{Median}(X)=e^\mu.
$$

The mean is

$$
E[X]=e^{\mu+\sigma^2/2}.
$$

The mean is greater than the median when $\sigma>0$. This happens because the Lognormal distribution has a long right tail.

The variance is

$$
\operatorname{Var}(X)
=
e^{2\mu+\sigma^2}
\left(e^{\sigma^2}-1\right).
$$

## Lognormal parameters in R

R uses:

- `meanlog`: the mean of the logarithm;
- `sdlog`: the standard deviation of the logarithm.

These are not the mean and standard deviation on the original measurement scale.

```{r}
meanlog = log(30)
sdlog = 0.5

c(
  median = exp(meanlog),
  mean = exp(meanlog + sdlog^2 / 2)
)
```

## Example: animal body mass

Suppose body mass is modeled as Lognormal with `meanlog = log(30)` and `sdlog = 0.5`.

```{r}
set.seed(20)

body_mass = rlnorm(
  1000,
  meanlog = meanlog,
  sdlog = sdlog
)

hist(
  body_mass,
  breaks = 40,
  probability = TRUE,
  col = "grey80",
  border = "white",
  xlab = "Body mass",
  main = "Simulated Lognormal body mass"
)

curve(
  dlnorm(
    x,
    meanlog = meanlog,
    sdlog = sdlog
  ),
  add = TRUE,
  col = "red",
  lwd = 2
)
```

The distribution is positive and right-skewed.

The probability that body mass is greater than 50 units is:

```{r}
plnorm(
  50,
  meanlog = meanlog,
  sdlog = sdlog,
  lower.tail = FALSE
)
```

The median and 95th percentile are:

```{r}
qlnorm(
  c(0.5, 0.95),
  meanlog = meanlog,
  sdlog = sdlog
)
```

## Log transformation

A Lognormal variable becomes Normal after taking logarithms.

```{r}
par(mfrow = c(1, 2))

hist(
  body_mass,
  breaks = 30,
  probability = TRUE,
  col = "grey80",
  border = "white",
  main = "Original scale",
  xlab = "Body mass"
)

hist(
  log(body_mass),
  breaks = 30,
  probability = TRUE,
  col = "grey80",
  border = "white",
  main = "Log scale",
  xlab = "Log body mass"
)

par(mfrow = c(1, 1))
```

The original measurements are right-skewed, while their logarithms are approximately symmetric.

## Limitations of the Lognormal distribution

A Lognormal distribution cannot generate zero or negative values.

If a data set contains many zeros, the zeros may represent:

- true absence;
- a detection limit;
- a failed measurement;
- a separate ecological process.

In such cases, it may be necessary to model the zero process separately rather than adding a small number to every observation automatically.

# The Beta distribution

The Beta distribution is a continuous distribution with values between zero and one.

It is useful for variables such as:

- vegetation cover;
- canopy openness;
- fractional ground cover;
- the proportion of a diet made up of one food type;
- a probability such as germination probability;
- the fraction of habitat occupied.

The Beta density is

$$
f_X(x)=
\frac{\Gamma(a+b)}
{\Gamma(a)\Gamma(b)}
x^{a-1}(1-x)^{b-1},
\qquad 0<x<1.
$$

The parameters $a$ and $b$ control the shape of the distribution.

The mean is

$$
E[X]=\frac{a}{a+b}.
$$

The variance is

$$
\operatorname{Var}(X)
=
\frac{ab}
{(a+b)^2(a+b+1)}.
$$

## Shapes of the Beta distribution

Different parameter values produce very different shapes.

- If $a=b=1$, the density is flat.
- If $a>b$, values tend to be closer to one.
- If $a<b$, values tend to be closer to zero.
- If $a=b>1$, values concentrate near 0.5.
- If $a<1$ and $b<1$, values tend to concentrate near zero and one.

```{r}
x = seq(0.001, 0.999, length.out = 500)

plot(
  x,
  dbeta(x, shape1 = 2, shape2 = 5),
  type = "l",
  lwd = 2,
  ylim = c(0, 5),
  xlab = "Value",
  ylab = "Probability density",
  main = "Different Beta distributions"
)

lines(
  x,
  dbeta(x, shape1 = 5, shape2 = 2),
  lwd = 2,
  lty = 2
)

lines(
  x,
  dbeta(x, shape1 = 5, shape2 = 5),
  lwd = 2,
  lty = 3
)

lines(
  x,
  dbeta(x, shape1 = 0.5, shape2 = 0.5),
  lwd = 2,
  lty = 4
)

legend(
  "top",
  legend = c(
    "a = 2, b = 5",
    "a = 5, b = 2",
    "a = 5, b = 5",
    "a = 0.5, b = 0.5"
  ),
  lwd = 2,
  lty = 1:4,
  bty = "n"
)
```

## Example: vegetation cover

Suppose vegetation cover is modeled using a Beta distribution with $a=2$ and $b=5$.

```{r}
cover_a = 2
cover_b = 5

c(
  mean = cover_a / (cover_a + cover_b),
  variance =
    cover_a * cover_b /
    ((cover_a + cover_b)^2 * (cover_a + cover_b + 1))
)
```

The expected vegetation cover is

$$
E[X]=\frac{2}{2+5}=0.286.
$$

The probability that vegetation cover is less than 20% is:

```{r}
pbeta(
  0.20,
  shape1 = cover_a,
  shape2 = cover_b
)
```

The median and 95th percentile are:

```{r}
qbeta(
  c(0.5, 0.95),
  shape1 = cover_a,
  shape2 = cover_b
)
```

We can simulate cover values:

```{r}
set.seed(30)

vegetation_cover = rbeta(
  1000,
  shape1 = cover_a,
  shape2 = cover_b
)

hist(
  vegetation_cover,
  breaks = 30,
  probability = TRUE,
  col = "grey80",
  border = "white",
  xlab = "Vegetation cover",
  main = "Simulated Beta vegetation cover"
)

curve(
  dbeta(
    x,
    shape1 = cover_a,
    shape2 = cover_b
  ),
  add = TRUE,
  col = "darkgreen",
  lwd = 2
)
```

## Values equal to zero or one

The Beta distribution is intended for values strictly between zero and one.

Ecological proportions may nevertheless include exact zeros or ones. For example:

- a quadrat may contain no vegetation;
- a plot may have complete vegetation cover;
- a species may be absent from a site.

These boundary values should not automatically be replaced with tiny adjustments. Exact zeros and ones may indicate a separate ecological process.

Possible approaches include:

- a zero-one-inflated Beta model;
- a separate model for boundary values;
- a transformation when scientifically justified.

The appropriate choice depends on how the data were generated.

# Histograms and density plots

## Histograms

A histogram divides the range of observations into intervals called bins.

The height of each bar represents the number or proportion of observations in that interval.

```{r}
set.seed(40)

measurements = rlnorm(
  500,
  meanlog = 2,
  sdlog = 0.5
)

hist(
  measurements,
  breaks = 10,
  col = "grey80",
  border = "white",
  xlab = "Measurement",
  main = "A histogram with relatively wide bins"
)

hist(
  measurements,
  breaks = 40,
  col = "grey80",
  border = "white",
  xlab = "Measurement",
  main = "A histogram with relatively narrow bins"
)
```

The two histograms show the same observations, but the shapes appear different because the bin widths differ.

There is no universally correct bin width. The choice should reveal important features without producing unnecessary noise.

## Frequency and density histograms

A frequency histogram displays counts. A density histogram scales the bars so that their total area is approximately one.

```{r}
hist(
  measurements,
  breaks = 30,
  probability = TRUE,
  col = "grey80",
  border = "white",
  xlab = "Measurement",
  ylab = "Density",
  main = "A density histogram"
)
```

A density histogram can be compared directly with a probability density curve.

## Kernel density plots

A kernel density estimate is a smooth estimate of the distribution of observed values.

```{r}
density_estimate = density(measurements)

plot(
  density_estimate,
  lwd = 2,
  xlab = "Measurement",
  ylab = "Estimated density",
  main = "Kernel density estimate"
)
```

The amount of smoothing is controlled by the bandwidth.

- A small bandwidth produces a rougher curve.
- A large bandwidth produces a smoother curve.

A kernel density estimate is based on observed data. It is different from a theoretical density such as `dlnorm()`.

```{r}
hist(
  measurements,
  breaks = 30,
  probability = TRUE,
  col = "grey85",
  border = "white",
  xlab = "Measurement",
  ylab = "Density",
  main = "Observed and theoretical densities"
)

lines(
  density(measurements),
  lwd = 2
)

curve(
  dlnorm(
    x,
    meanlog = 2,
    sdlog = 0.5
  ),
  add = TRUE,
  col = "red",
  lwd = 2
)

legend(
  "topright",
  legend = c(
    "Kernel density estimate",
    "Theoretical Lognormal density"
  ),
  col = c("black", "red"),
  lwd = 2,
  bty = "n"
)
```

## Empirical cumulative distributions

The empirical cumulative distribution function shows the fraction of observed values less than or equal to each value.

For observations $x_1,\ldots,x_n$, the empirical CDF is

$$
\widehat{F}(x)
=
\frac{\text{number of observations less than or equal to }x}{n}.
$$

In R, `ecdf()` calculates the empirical CDF.

```{r}
plot(
  ecdf(measurements),
  do.points = FALSE,
  verticals = TRUE,
  xlab = "Measurement",
  ylab = "Cumulative probability",
  main = "Empirical cumulative distribution"
)
```

An empirical CDF can be compared with a theoretical CDF:

```{r}
plot(
  ecdf(measurements),
  do.points = FALSE,
  verticals = TRUE,
  xlab = "Measurement",
  ylab = "Cumulative probability",
  main = "Empirical and theoretical CDFs"
)

curve(
  plnorm(
    x,
    meanlog = 2,
    sdlog = 0.5
  ),
  add = TRUE,
  col = "red",
  lwd = 2
)

legend(
  "bottomright",
  legend = c(
    "Empirical CDF",
    "Theoretical CDF"
  ),
  col = c("black", "red"),
  lwd = 2,
  bty = "n"
)
```

# Multivariate continuous distributions

Ecological studies often measure several variables on the same individual, plot, or sampling occasion.

Examples include:

- tree height and diameter;
- animal body mass and body length;
- soil moisture and plant biomass;
- temperature and growth rate.

Let

$$
(X,Y)
$$

represent two continuous random variables.

The **joint distribution** describes how probability is distributed across combinations of $X$ and $Y$.

The joint density is written as

$$
f_{X,Y}(x,y).
$$

For a region $A$ in the two-dimensional plane, the probability that $(X,Y)$ falls in that region is

$$
P\big((X,Y)\in A\big)
=
\iint_A f_{X,Y}(x,y)\,dx\,dy.
$$

For a rectangular region, this becomes

$$
P(a<X\leq b,\ c<Y\leq d)
=
\int_a^b
\int_c^d
f_{X,Y}(x,y)\,dy\,dx.
$$

The mathematics is an extension of the one-dimensional case. Instead of calculating an area under a curve, we calculate a volume under a density surface.

## Marginal distributions

A **marginal distribution** describes one variable while ignoring the other.

The marginal density of $X$ is

$$
f_X(x)
=
\int_{-\infty}^{\infty}
f_{X,Y}(x,y)\,dy.
$$

Similarly, the marginal density of $Y$ is

$$
f_Y(y)
=
\int_{-\infty}^{\infty}
f_{X,Y}(x,y)\,dx.
$$

For example, the marginal distribution of tree height describes tree heights across all trees, regardless of their diameter.

A histogram or density plot of tree height is an estimate of its marginal distribution.

## Conditional distributions

A **conditional distribution** describes one variable given information about another variable.

The conditional density of $Y$ given $X=x$ is

$$
f_{Y\mid X}(y\mid x)
=
\frac{f_{X,Y}(x,y)}
{f_X(x)}.
$$

An ecological question involving a conditional distribution might be:

> What is the distribution of tree height among trees with a given diameter?

The distribution of tree height for all trees is marginal. The distribution of tree height for trees of a particular size is conditional.

Conditional distributions are central to regression and prediction.

## Independence

Two continuous variables are independent if knowing one variable does not provide information about the other.

For independent variables,

$$
f_{X,Y}(x,y)=f_X(x)f_Y(y).
$$

If $X$ and $Y$ are independent, the conditional distribution of $Y$ is the same for every value of $X$:

$$
f_{Y\mid X}(y\mid x)=f_Y(y).
$$

Ecological variables may be dependent because of:

- shared environmental conditions;
- biological relationships;
- spatial structure;
- temporal structure;
- common measurement processes.

# Simulating two related ecological variables

We can simulate correlated Normal variables and then exponentiate them to create positive, right-skewed measurements.

Suppose we want to simulate tree diameter and height.

```{r}
set.seed(50)

n_trees = 1000
rho = 0.75

z1 = rnorm(n_trees)
z2 = rnorm(n_trees)

log_diameter = 3.2 + 0.45 * z1

log_height = 3.0 +
  0.25 * (
    rho * z1 +
      sqrt(1 - rho^2) * z2
  )

diameter = exp(log_diameter)
height = exp(log_height)

c(
  correlation_on_log_scale =
    cor(log_diameter, log_height),
  correlation_on_original_scale =
    cor(diameter, height)
)
```

The variables are positively correlated on the logarithmic scale. They are also positively related on the original scale, but the numerical correlation can differ after exponentiation.

```{r}
par(mfrow = c(1, 2))

plot(
  log_diameter,
  log_height,
  pch = 16,
  col = rgb(0, 0, 0, 0.25),
  xlab = "Log diameter",
  ylab = "Log height",
  main = "Measurements on the log scale"
)

plot(
  diameter,
  height,
  pch = 16,
  col = rgb(0, 0, 0, 0.25),
  xlab = "Diameter",
  ylab = "Height",
  main = "Measurements on the original scale"
)

par(mfrow = c(1, 1))
```

The original-scale measurements are positive and right-skewed. The logarithmic transformation makes their distributions more symmetric and makes the relationship easier to describe with a linear correlation.

# Conditional distributions in an ecological example

We can compare the distribution of height for relatively small and relatively large trees.

```{r}
small_trees = log_diameter <= quantile(log_diameter, 0.25)
large_trees = log_diameter >= quantile(log_diameter, 0.75)

small_height_density = density(log_height[small_trees])
large_height_density = density(log_height[large_trees])

plot(
  small_height_density,
  lwd = 2,
  xlab = "Log height",
  ylab = "Estimated density",
  main = "Conditional height distributions",
  ylim = range(
    c(
      small_height_density$y,
      large_height_density$y
    )
  )
)

lines(
  large_height_density,
  lwd = 2,
  lty = 2
)

legend(
  "topright",
  legend = c(
    "Small trees",
    "Large trees"
  ),
  lwd = 2,
  lty = c(1, 2),
  bty = "n"
)
```

These are approximate conditional distributions. The first curve describes log height among trees in the lower quarter of the diameter distribution. The second describes log height among trees in the upper quarter.

The two distributions differ because height and diameter are related.

# Contour plots

A contour plot is a two-dimensional representation of a density surface.

A contour line connects points with equal density, similar to a line on a topographic map connecting points with equal elevation.

Contour plots can reveal:

- concentration;
- spread;
- correlation;
- the orientation of a relationship;
- multiple regions of high density.

Closely spaced contours indicate that density changes rapidly. However, the area between contour lines does not necessarily represent equal probability.

## Independent and correlated Normal variables

The following function calculates the density of a standard bivariate Normal distribution with correlation $\rho$:

$$
f(x,y)
=
\frac{1}{2\pi\sqrt{1-\rho^2}}
\exp\left[
-\frac{x^2-2\rho xy+y^2}
{2(1-\rho^2)}
\right].
$$

```{r}
bivariate_normal_density = function(x, y, rho) {
  1 /
    (2 * pi * sqrt(1 - rho^2)) *
    exp(
      -(
        x^2 -
          2 * rho * x * y +
          y^2
      ) /
        (2 * (1 - rho^2))
    )
}

grid = seq(-3.5, 3.5, length.out = 150)

par(mfrow = c(1, 2))

for (rho in c(0, 0.8)) {
  density_surface = outer(
    grid,
    grid,
    function(x, y) {
      bivariate_normal_density(x, y, rho)
    }
  )

  contour(
    grid,
    grid,
    density_surface,
    nlevels = 8,
    xlab = "X",
    ylab = "Y",
    main = paste("Correlation =", rho)
  )
}

par(mfrow = c(1, 1))
```

When $\rho=0$, the variables are independent and the contours are circular because both variables have the same variance.

When $\rho=0.8$, the contours are tilted ellipses. The positive slope indicates that large values of $X$ tend to occur with large values of $Y$.

A negative correlation would produce ellipses tilted in the opposite direction.

## Empirical two-dimensional density

For observed data, `smoothScatter()` provides a smoothed visualization of where observations are concentrated.

```{r}
smoothScatter(
  log_diameter,
  log_height,
  xlab = "Log diameter",
  ylab = "Log height",
  main = "Empirical two-dimensional density"
)
```

A scatterplot shows individual observations. A contour or smooth-density plot emphasizes regions containing many observations.

Both views are useful:

- the scatterplot shows individual variation and unusual observations;
- the density plot shows the overall concentration of observations.

# Choosing among the three distributions

The three continuous distributions introduced in this chapter are useful in different situations.

| Distribution | Typical use | Range | Important feature |
|---|---|---:|---|
| Normal | Symmetric measurements or errors | All real values | Symmetric |
| Lognormal | Positive, multiplicative measurements | $x>0$ | Right-skewed |
| Beta | Proportions and fractional measurements | $0<x<1$ | Flexible bounded shape |

A useful first set of questions is:

1. Can the variable take negative values?
2. Is the variable strongly right-skewed?
3. Is the variable strictly positive?
4. Is the variable a proportion between zero and one?
5. Are exact zeros or ones possible?
6. Is variation additive or multiplicative?
7. Are the observations approximately symmetric after transformation?

These questions do not automatically identify the correct distribution. They help identify models that are plausible enough to investigate.

# Summary

Continuous random variables can take any value within an interval. Unlike discrete variables, they assign probability to intervals rather than individual exact values:

$$
P(X=x)=0.
$$

A probability density is not itself a probability. Interval probabilities are areas under the density curve:

$$
P(a<X\leq b)=\int_a^b f_X(x)\,dx.
$$

The cumulative distribution function is

$$
F_X(x)=P(X\leq x).
$$

The four R function types are:

- `d*`: density;
- `p*`: cumulative probability;
- `q*`: quantile;
- `r*`: random sample.

The Normal distribution is useful for symmetric continuous measurements and errors.

The Lognormal distribution is useful for positive, right-skewed measurements that may result from multiplicative processes.

The Beta distribution is useful for continuous proportions between zero and one.

Histograms divide observations into intervals. Kernel density plots provide a smoothed estimate of the observed distribution. Empirical CDFs show cumulative observed probabilities.

For two continuous variables, the joint distribution describes combinations of values. Marginal distributions describe one variable while ignoring the other. Conditional distributions describe one variable given information about another.

Contour plots display two-dimensional density. Their shape and orientation can reveal the spread and dependence between variables.

# Exercises

## Exercise 1: Normal tree heights

Assume tree height follows a Normal distribution with mean 22 m and standard deviation 4 m.

1. Calculate the probability that a tree is between 18 and 26 m tall.
2. Calculate the probability that a tree is taller than 30 m.
3. Calculate the 5th, 50th, and 95th percentiles.
4. Simulate 1,000 tree heights.
5. Compare a histogram, kernel density estimate, and theoretical Normal density.

## Exercise 2: Lognormal body mass

Assume body mass follows a Lognormal distribution with `meanlog = log(30)` and `sdlog = 0.5`.

1. Calculate the theoretical median.
2. Calculate the theoretical mean.
3. Explain why the mean is greater than the median.
4. Calculate the probability that body mass exceeds 50 units.
5. Simulate 1,000 body masses.
6. Compare the distribution of body mass with the distribution of log body mass.

## Exercise 3: Beta vegetation cover

Assume vegetation cover follows a Beta distribution with $a=2$ and $b=5$.

1. Calculate the mean and variance.
2. Calculate the probability that vegetation cover is less than 20%.
3. Calculate the median and 95th percentile.
4. Simulate 1,000 cover values.
5. Plot the theoretical density and simulated observations.
6. Explain why a standard Beta model may be inappropriate if many observations are exactly zero.

## Exercise 4: Histograms and density estimates

1. Simulate 500 observations from a Lognormal distribution.
2. Plot histograms using at least three different bin widths.
3. Add a kernel density estimate to each plot.
4. Describe which features of the distribution are stable and which depend on the bin width.
5. Compare the empirical density with the theoretical Lognormal density.

## Exercise 5: Bivariate ecological measurements

Simulate correlated log diameter and log height values.

1. Produce a scatterplot on the logarithmic scale.
2. Produce a scatterplot on the original scale.
3. Calculate the correlation on both scales.
4. Create a smooth two-dimensional density plot.
5. Compare height distributions for relatively small and relatively large trees.
6. Explain the difference between a marginal distribution and a conditional distribution.

## Exercise 6: Contour plots

1. Create contour plots for bivariate Normal distributions with correlations of $0$, $0.5$, and $-0.5$.
2. Describe how the orientation of the contours changes.
3. Explain why circular contours correspond to independence only when the variables have equal variance.
4. Add simulated observations to one of the contour plots.
5. Explain why contours show density rather than individual observations.
