---
title: "Introduction to Probability (Continuous)"
---

# Introduction

In this chapter, we learn how to use probability distributions for the measurements
that refuse to stay neatly in whole numbers. In [Chapter 2](../Chapter_2/Theory.qmd),
we dealt with data such as counted surviving seeds and wildlife detections. But much of
ecological research deals with observations such as  height, body mass, leaf area, or
vegetation cover. A tree is not 22 m tall in the same sense that a quadrat contains four
seeds: its height can lie anywhere along a scale, and our measuring process adds its own
error.

That distinction changes what a probability model means. We cannot ask for the
probability of one exact height; we ask how much probability the model puts over a
meaningful range. This means we cannot define a probability mass function, so we will need
to use a different concept to build distributions. Still, we will have four faces of
probablility (some very familiar by now) and illustrate them with examples from the Normal,
LogNormal and Beta distributions.

# Learning goals

In this chapter, we will cover:

- continuous random variables and probability density;
- interval probabilities and quantiles;
- the four faces of a continuous distribution (density, cumulative probability,
  quantile and sampler);
- how samples and continuous distributions are related;
- joint and marginal distributions for two continuous variables;
- how to interpret scatterplots and contour plots;
- choosing a continuous distribution for an ecological measurement.

We will learn more about the normal, lognormal and beta distributions and
practise using them in R. Each is a starting model, not a label that we attach
to a variable automatically.

# Continuous measurements and probability {#sec-continuous_measurements}

A **continuous measurement** can vary across an interval, rather than
taking separate values such as whole-number counts. A probability model therefore treats
the value as uncertain before we observe it and assigns probability to *ranges* of
possible values.

Suppose $X$ is the height in metres of a randomly selected tree from a defined
population. We can treat $X$ as **continuous**: within a plausible range, it can take
any value, not just a whole number. A seedling count, by contrast, has to be an integer.
The distinction comes from what we are measuring, not from whether the spreadsheet has
decimal places.

Of course, instruments are not infinitely precise. A recorded height of 22.1 m might
represent an underlying height from 22.05 m up to 22.15 m. A continuous model is usually
a useful approximation to that underlying quantity, while the recorded values remind us
that every observation also comes through a measurement process.

::: {.callout-note title="Continuous random variable"}
A continuous random variable can take values throughout an interval. Its model
assigns probabilities to ranges of values. The values allowed by the model form
its **support**.
:::

## From probability mass to density

In [Chapter 2](../Chapter_2/Theory.qmd#sec-four_faces), the probability mass at a
particular count was $P(X=x)$. That made sense because a count of exactly four seedlings
is one distinct possible outcome. For an ideal continuous variable, the probability of
any one exact value is zero:

$$
P(X=x)=0.
$$

This sounds strange the first time you meet it, because we can certainly record a height
of 22.1 m. But the recorded number stands for a small range at the instrument's
precision. The questions that a continuous model answers concern intervals, such as
$P(20<X\leq25)$.

A **probability density function**, $f_X(x)$, shows where the model concentrates
probability along the measurement scale. Its curve is not a collection of exact-value
probabilities. Instead, the probability of an interval is the area under the curve:

$$
P(a<X\leq b)=\int_a^b f_X(x)\,dx.
$$

The density cannot be negative and its total area over the support is one. Because exact
points have probability zero, including or excluding an interval's endpoints does not
change its probability under a continuous model. That is a mathematical convenience, not
permission to forget how precisely the field measurement was made.

::: {.callout-warning title="Density is not an exact-value probability"}
The value $f_X(22)$ is a curve height, not $P(X=22)$. A density can exceed one if
probability is concentrated in a narrow interval. Its value also changes when,
for example, heights are expressed in centimetres rather than metres; the
probability for the same physical interval does not.
:::

:::: {.callout-important title="Practice: measuring tree height"}

A tree height is recorded to the nearest tenth of a metre as 22.1 m. What range
of underlying heights could have produced that record? Would a continuous model
assign positive probability to an underlying height of exactly 22.1 m?

::: {.callout-tip title="Solution" collapse="true"}

Rounding to the nearest tenth, heights from approximately 22.05 m up to 22.15 m
could be recorded as 22.1 m. Under an ideal continuous model,
$P(X=22.1)=0$, but the recording interval can have positive probability.

:::
::::

# The four faces of a continuous distribution {#sec-continuous_four_faces}

The [four faces of a distribution](../Chapter_2/Theory.qmd#sec-four_faces) still give
us a useful route through R. The prefixes are the same, but there is one crucial change:
for continuous variables, `d*` returns a **density**, not the probability of an exact
outcome.

To keep the ideas connected, suppose tree heights follow a normal model with mean 22 m
and standard deviation 4 m. These are assumed population values, not estimates from a
forest. We will examine what a normal model assumes later; for now, they give us something
concrete to calculate and question.

## 1. Probability density

Where the density is high, the model places more probability in a small interval around
that height. For the assumed model, `dnorm()` evaluates the density at chosen heights:

```{r}
# Calculate the density at one chosen height under the assumed model.
dnorm(x = 22, mean = 22, sd = 4)
```

The result has units of density per metre. It is not the probability that a tree is
exactly 22 m tall. The curve gives us a picture of where the model expects heights to be
concentrated.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Assumed normal density for tree height. Interval probability is area under the curve."
# Calculate the assumed density over a sequence of possible heights.
height_values = seq(from = 8, to = 36, length.out = 400)
height_density = dnorm(x = height_values, mean = 22, sd = 4)
# Plot the density with units that match the height scale.
plot(
  x = height_values, y = height_density, type = "l", lwd = 2,
  xlab = "Tree height (m)", ylab = "Density per metre"
)
```

## 2. Cumulative probability

Ecological questions are often about a threshold rather than one impossible exact value:
how many trees are above a canopy height, or how much of the population lies within a
size range? The **cumulative distribution function** is $F_X(x)=P(X\leq x)$. It gathers
the probability for all heights up to $x$. To calculate the probability that a tree is
between 18 m and 26 m tall, subtract two cumulative probabilities:

$$
P(18<X\leq26)=F_X(26)-F_X(18).
$$

In R:

```{r}
# Subtract cumulative probabilities to obtain an interval probability.
pnorm(q = 26, mean = 22, sd = 4) -
  pnorm(q = 18, mean = 22, sd = 4)
```

For a height above 30 m, use the upper tail:

```{r}
# Use the upper tail for the probability above the threshold.
pnorm(q = 30, mean = 22, sd = 4, lower.tail = FALSE)
```

The cumulative curve rises smoothly from zero towards one.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Cumulative probability for the assumed tree-height model."
# Evaluate and plot the cumulative probability over possible heights.
height_values = seq(from = 8, to = 36, length.out = 400)
plot(
  x = height_values,
  y = pnorm(q = height_values, mean = 22, sd = 4),
  type = "l", lwd = 2, ylim = c(0, 1),
  xlab = "Tree height (m)", ylab = "Cumulative probability"
)
```

## 3. Quantiles

Sometimes we turn the question around. Given a cumulative probability, which height
marks that point in the distribution? A **quantile** gives that height. The 0.95 quantile
is the height below which the model places 95% of tree heights:

```{r}
# Find the height at the requested cumulative probability.
qnorm(p = 0.95, mean = 22, sd = 4)
```

Because this model is continuous, $P(X\leq q_{0.95})=0.95$. It is a model prediction,
not the height of a particular observed tree. A 95th percentile only says something about
a real forest if the model and its parameters are reasonable for that forest.

## 4. Sampling

So far the normal curve may look rather too tidy. Sampling makes its variation visible:

```{r}
# Simulate one height for each of 100 trees from the assumed model.
set.seed(seed = 42)
tree_heights = rnorm(n = 100, mean = 22, sd = 4)
head(tree_heights)
```

Each value represents one simulated tree. A finite sample will not have exactly the
model's mean or standard deviation, and a new seed produces a different sample. That is
not a defect in the code; it is the sampling variation that we will later need to account
for when estimating parameters from field data.

::: {.callout-note title="The four R function types"}
For continuous distributions, `d*` gives density, `p*` cumulative probability,
`q*` a quantile, and `r*` a random sample. The first argument is conventionally
named `x`, `q`, `p`, or `n`, respectively.
:::

:::: {.callout-important title="Practice: a tree-height threshold"}

Under the assumed normal model, calculate the probability that a tree is taller
than 26 m. Find the height that marks the lower 10% of the model. Explain each
result in words.

::: {.callout-tip title="Solution" collapse="true"}

```{r}
# Calculate an upper-tail probability and a lower-tail quantile.
pnorm(q = 26, mean = 22, sd = 4, lower.tail = FALSE)
qnorm(p = 0.10, mean = 22, sd = 4)
```

The first result is approximately 0.159: about 16% of heights exceed 26 m under
this model. The second is approximately 16.9 m: about 10% fall below that height.

:::
::::

# From a model to observed measurements {#sec-continuous_samples}

Chapter 2 introduced the distinction between a
[population distribution and a sample](../Chapter_2/Theory.qmd#sec-sample_population).
The same distinction matters here, perhaps even more because a smooth density curve can
look deceptively like data. It is not. A theoretical density describes a proposed model;
a plot of measured heights or masses summarises one finite sample. We need both if we are
going to ask whether the model is a useful approximation.

## Histograms and density estimates

You will often start by plotting the measurements, which is sensible, but the plot itself
also makes choices. A **histogram** groups observations into intervals called bins. With
equal-width bins, a frequency histogram's bar height is the number of observations in a
bin. A density histogram instead has total *bar area* one, allowing us to compare it with
a theoretical density curve. Changing the bin width can make the same sample look more
or less lumpy without changing a single observed tree.

A **kernel density estimate**, calculated by `density()`, gives another smooth summary
of the sample. Its bandwidth controls how much smoothing it applies. It is an empirical
estimate, whereas `dnorm()` describes the normal model we chose before seeing this
simulation. Neither curve gets to be called “the truth” merely because it is smooth.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "A simulated sample of tree heights, its kernel density estimate, and the assumed normal density."
# Simulate a sample from the assumed normal model.
set.seed(seed = 73)
tree_heights = rnorm(n = 200, mean = 22, sd = 4)
# Compare the sample estimate with the density used to generate it.
hist(
  x = tree_heights, breaks = 16, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Density per metre",
  main = "Simulated tree heights"
)
lines(density(x = tree_heights), lwd = 2, lty = 2)
curve(dnorm(x = x, mean = 22, sd = 4), add = TRUE, lwd = 2)
legend(
  "topright", legend = c("Sample estimate", "Assumed model"),
  lty = c(2, 1), lwd = 2, bty = "n"
)
```

The two curves can differ through sampling variation even though the assumed model
generated the data. Conversely, a pleasing agreement in one plot does not establish that
a normal model is suitable for field measurements. It is evidence to weigh alongside the
measurement process and the ecological question, not a rubber stamp.

## Empirical cumulative distributions

Histograms require bins and density estimates require a smoothing choice. The **empirical
cumulative distribution** gives us a useful alternative: the fraction of sample values at
or below each threshold. `ecdf()` calculates it. The empirical curve rises in steps,
because it only changes when we encounter an observed height; a continuous model's
cumulative curve is smooth.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Empirical cumulative distribution of simulated heights and the assumed model's cumulative distribution."
# Simulate a sample and plot its empirical cumulative distribution.
set.seed(seed = 73)
tree_heights = rnorm(n = 200, mean = 22, sd = 4)
plot(
  ecdf(tree_heights), do.points = FALSE, verticals = TRUE,
  xlab = "Tree height (m)", ylab = "Cumulative proportion",
  main = "Simulated heights and model"
)
curve(pnorm(q = x, mean = 22, sd = 4), add = TRUE, lwd = 2, lty = 2)
legend(
  "bottomright", legend = c("Sample", "Assumed model"),
  lty = c(1, 2), lwd = 2, bty = "n"
)
```

:::: {.callout-important title="Practice: sample or model?"}

You simulate 200 heights from the assumed model and make a histogram. Would a
different seed produce exactly the same histogram? Would changing the bin width
change the sample mean?

::: {.callout-tip title="Solution" collapse="true"}

A different seed usually produces a different sample and histogram. Changing
the bin width changes the display, not the measured values, so the sample mean
stays the same.

:::
::::

# Two continuous variables {#sec-two_continuous_variables}

Ecological observations rarely arrive one variable at a time. When each observational
unit has two continuous measurements, a model must describe their individual variation
and how they vary together. Suppose $X$ is a tree's diameter in centimetres and $Y$ is
its height in metres. One row of data holds both measurements for the same tree. Breaking
that pairing accidentally would throw away the question we usually care about: how do
height and diameter occur together?

## Joint and marginal

The **joint distribution** describes combinations of diameter and height. Its joint
density, $f_{X,Y}(x,y)$, spreads density across a two-dimensional region. Probabilities
belong to regions, such as trees with diameters from 25 to 35 cm *and* heights above 20
m, not to one exact pair of measurements.

Sometimes we deliberately ignore one part of the pair. The **marginal distribution** of
height describes heights across all trees in the population, without selecting trees by
diameter. A histogram of all heights estimates that marginal distribution.

Chapter 2 introduced [independence](../Chapter_2/Theory.qmd#sec-independence_expectations).
Here, independence means that knowing a tree's diameter would not change the model's
distribution of its height. A relationship in a scatterplot may make that assumption hard
to defend, but it does not identify an ecological cause by itself. Older trees, species
composition, light, and crowding could all sit behind the same visible association.

## Scatterplots and contours

A scatterplot keeps the pair together and makes the association visible. Here we simulate
trees whose height and diameter tend to increase together. The parameters describe the
simulation, not a claim about a real forest. That distinction matters: a familiar-looking
scatterplot can suggest questions without supplying a causal answer.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Simulated tree heights and diameters with a positive association."
# Simulate two measurements per tree with a positive association.
set.seed(seed = 50)
n_trees = 300
z_height = rnorm(n = n_trees)
z_other = rnorm(n = n_trees)
# Transform the simulated values to the height and diameter scales.
tree_height = 22 + 4 * z_height
tree_diameter = 30 + 5 * (0.7 * z_height + sqrt(1 - 0.7^2) * z_other)
plot(
  x = tree_diameter, y = tree_height, pch = 16,
  col = grDevices::adjustcolor("black", alpha.f = 0.35),
  xlab = "Tree diameter (cm)", ylab = "Tree height (m)"
)
```

Larger diameters tend to accompany greater heights in this simulation, but individual
trees vary. A marginal height plot would hide the pairing. Each view
answers a different question, so there is no single “right” plot outside its purpose.

A **contour plot** is the model-based companion to a scatterplot. It joins combinations
of diameter and height with equal joint density, much as a topographic map joins places
of equal elevation. Contours show where the *model* concentrates outcomes. A contour line
is not a probability boundary, and the regions between successive lines need not contain
equal probability.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Contours of the assumed joint density for tree diameter and height."
# Set a grid of possible height and diameter combinations.
diameter_grid = seq(from = 10, to = 50, length.out = 100)
height_grid = seq(from = 8, to = 36, length.out = 100)
rho = 0.7
# Calculate the assumed joint density at each grid combination.
joint_density = outer(
  X = diameter_grid, Y = height_grid,
  FUN = function(diameter, height) {
    dx = (diameter - 30) / 5
    hy = (height - 22) / 4
    exp(-(dx^2 - 2 * rho * dx * hy + hy^2) / (2 * (1 - rho^2))) /
      (2 * pi * 5 * 4 * sqrt(1 - rho^2))
  }
)
contour(
  x = diameter_grid, y = height_grid, z = joint_density,
  nlevels = 8, xlab = "Tree diameter (cm)",
  ylab = "Tree height (m)"
)
```

These tilted contours show that larger diameter and greater height tend to occur together
under the assumed model. A scatterplot reveals individual observations; contours summarise
the joint density. Neither plot says that changing diameter would cause a change in height.
We need a study design and a model for that stronger ecological claim.

# The normal distribution {#sec-normal_distribution}

A **normal distribution** is a symmetric continuous model. It is often treated as the
default for measurements, which can encourage us to reach for it before asking what was
measured. A better habit is to start with its assumptions. Its mean $\mu$ gives the
centre and its standard deviation $\sigma>0$ sets the spread. We write
$X\sim\operatorname{Normal}(\mu,\sigma^2)$. The second quantity in this notation is
the **variance**, while R's `sd` argument takes the standard deviation $\sigma$.

The normal model has support over all real values. Its density is

$$
f_X(x)=\frac{1}{\sigma\sqrt{2\pi}}
\exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right).
$$

It has $E[X]=\mu$ and $\operatorname{Var}(X)=\sigma^2$. The mean, median, and mode
coincide because the curve is symmetric. Increasing $\sigma$ spreads the same total
probability over a wider range and lowers the curve's peak.

For the assumed tree heights, $\mu=22$ m and $\sigma=4$ m. The mean is the average
height across many trees generated by this model; the standard deviation describes
variation among those heights. Approximately 68% of a normal model falls within one
standard deviation of its mean, 95% within two, and 99.7% within three. Thus about 95%
of this model's heights fall between 14 m and 30 m. These percentages are approximations,
and they describe the model, not a guarantee about any one sample of trees.

We can express a height in standard deviation units with $Z=(X-\mu)/\sigma$. A 30 m
tree is two standard deviations above the assumed mean. R calculates probabilities
directly with `mean` and `sd`, so you do not need to standardise each observation by hand.

```{r}
# Calculate the probability of a height above 30 m.
pnorm(q = 30, mean = 22, sd = 4, lower.tail = FALSE)
```

::: {.callout-warning title="Check the support"}
A normal model allows negative values. For positive measurements such as body
mass, check whether it places meaningful probability below zero or misses
strong right skew. Symmetry alone does not make a model suitable.
:::

:::: {.callout-important title="Practice: spread in tree heights"}

Two proposed normal models have the same mean height, 22 m, but standard
deviations of 2 m and 5 m. Which assigns more probability to heights far from
22 m? What happens to the density's peak as spread increases?

::: {.callout-tip title="Solution" collapse="true"}

The model with standard deviation 5 m is more spread out and gives more
probability to heights far from 22 m. Its density peak is lower because each
density curve still has total area one.

:::
::::

# The lognormal distribution {#sec-lognormal_distribution}

Body mass gives us a familiar reason to hesitate before using a normal curve. It cannot
be negative, and a few large organisms can be much heavier than most of the sample. A
**lognormal distribution** is a continuous model for strictly positive values with this
kind of long right tail. A positive variable $X$ is lognormal when its logarithm is
normal:

$$
\log(X)\sim\operatorname{Normal}(\mu,\sigma^2).
$$

The model's support is $x>0$. Its density is

$$
f_X(x)=\frac{1}{x\sigma\sqrt{2\pi}}
\exp\left(-\frac{(\log x-\mu)^2}{2\sigma^2}\right),\qquad x>0.
$$

Animal body mass may have this right-skewed shape: most animals can be relatively light,
while a few are much heavier. The model has a biological-looking feature, but that does
not mean every body-mass dataset is lognormal. Different species, life stages, or a
sampling threshold can produce a quite different pattern.

In R, `meanlog` is $\mu$ and `sdlog` is $\sigma$. Both describe *log body mass*, not
body mass in its original units. This parameterisation is where it is easy to make a
plausible-looking but incorrect interpretation. On the original scale, the median is
$e^\mu$, the mean is $e^{\mu+\sigma^2/2}$, and the variance is
$e^{2\mu+\sigma^2}(e^{\sigma^2}-1)$. The mean exceeds the median when $\sigma>0$
because the right tail contains unusually large possible values.

Suppose body mass in grams follows a lognormal model with
`meanlog = log(30)` and `sdlog = 0.5`. Its median is 30 g, while its mean is
larger. We can calculate a probability and a quantile on the original scale:

```{r}
# Calculate an upper-tail probability and a high quantile on the original scale.
plnorm(q = 50, meanlog = log(30), sdlog = 0.5,
        lower.tail = FALSE)
qlnorm(p = 0.95, meanlog = log(30), sdlog = 0.5)
```

The log scale can make a multiplicative process easier to describe: successive
proportional changes multiply body mass, while their logarithms add. That is one possible
ecological motivation, not proof that a particular set of measurements follows a lognormal
model. We still have to confront the model with the observations and how they were collected.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Simulated lognormal body masses on the original and log scales."
# Simulate body masses from the assumed lognormal model.
set.seed(seed = 20)
body_mass = rlnorm(n = 500, meanlog = log(30), sdlog = 0.5)
# Show how the same values look on their original and log scales.
old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))
hist(body_mass, breaks = 25, col = "grey80", border = "white",
     xlab = "Body mass (g)", main = "Original scale")
hist(log(body_mass), breaks = 25, col = "grey80", border = "white",
     xlab = "Log body mass", main = "Log scale")
par(old_par)
```

The original-scale simulation is right-skewed; its logarithms are more
symmetric. The four functions are `dlnorm()`, `plnorm()`, `qlnorm()`, and
`rlnorm()`.

::: {.callout-warning title="Zero is outside a lognormal model"}
A lognormal model cannot generate zero or negative values. A recorded zero
could mean absence, a detection limit, or another process. Its meaning should
be investigated before changing the values or choosing a model.
:::

:::: {.callout-important title="Practice: mean and median mass"}

For `meanlog = log(30)` and `sdlog = 0.5`, calculate the median and mean body
mass. Why are they different?

::: {.callout-tip title="Solution" collapse="true"}

```{r}
# Convert the lognormal parameters to the original-scale median and mean.
median_mass = exp(log(30))
mean_mass = exp(log(30) + 0.5^2 / 2)
c(median_g = median_mass, mean_g = mean_mass)
```

The median is 30 g and the mean is approximately 34 g. Large possible masses
in the right tail pull the mean above the median.

:::
::::

# The beta distribution {#sec-beta_distribution}

Vegetation cover brings a different constraint. It has a natural lower bound of zero and
upper bound of one, so a normal or lognormal model would wander into impossible values.
A **beta distribution** is a continuous model for fractions strictly between zero and
one. Its two positive shape parameters, $a$ and $b$, allow many different density shapes
within those bounds. It is a model for a continuous fraction, not for the number of covered
points out of a fixed number of observations.

Suppose $X$ is the fraction of a quadrat covered by vegetation, measured on a continuous
scale. The bounds come from what a fraction means, not from a statistical convention.

Its density is

$$
f_X(x)=\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}
x^{a-1}(1-x)^{b-1},\qquad 0<x<1.
$$

Here $\Gamma$ is the gamma function, which provides the scaling needed for total area
one. We do not need to calculate it by hand to use the model. The mean and variance are

$$
E[X]=\frac{a}{a+b},\qquad
\operatorname{Var}(X)=\frac{ab}{(a+b)^2(a+b+1)}.
$$

If $a>b$, density tends towards higher cover; if $a<b$, it tends towards lower cover.
The pair also controls how concentrated or U-shaped the density is. R calls the parameters
`shape1` and `shape2`, which is more descriptive than memorable.

Suppose vegetation cover follows a beta model with $a=2$ and $b=5$.
Its expected cover is $2/7$, or about 29% of a quadrat. We can calculate
the chance of cover below 20% and the 95th percentile:

```r
# Calculate a lower-tail probability and a high quantile for the cover model.
pbeta(q = 0.20, shape1 = 2, shape2 = 5)
qbeta(p = 0.95, shape1 = 2, shape2 = 5)
```

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Beta densities with different shapes for a fraction between zero and one."
# Calculate density curves for three possible beta shapes.
cover_values = seq(from = 0.001, to = 0.999, length.out = 400)
plot(cover_values, dbeta(x = cover_values, shape1 = 2, shape2 = 5),
     type = "l", lwd = 2, ylim = c(0, 11),
     xlab = "Vegetation cover (fraction)", ylab = "Density")
lines(cover_values, dbeta(x = cover_values, shape1 = 5, shape2 = 2),
      lwd = 2, lty = 2)
lines(cover_values, dbeta(x = cover_values, shape1 = 0.5, shape2 = 0.5),
      lwd = 2, lty = 3)
legend("top", legend = c("a = 2, b = 5", "a = 5, b = 2",
                         "a = 0.5, b = 0.5"),
       lty = 1:3, lwd = 2, bty = "n")
```

The four R functions are `dbeta()`, `pbeta()`, `qbeta()`, and `rbeta()`.

::: {.callout-warning title="Exact zero and one values"}
The continuous beta model assigns no probability to an exact cover of zero
or one. Such observations may be ecologically meaningful. Do not replace
them with small adjustments automatically; first ask how cover was measured
and whether boundary values need a separate process in the model.
:::

:::: {.callout-important title="Practice: interpreting a cover model"}

Would a beta model with $a=2$ and $b=5$ assign positive probability to
cover between 0.2 and 0.4? Would it assign positive probability to cover
of exactly 0.2? Explain the difference.

::: {.callout-tip title="Solution" collapse="true"}

The interval has positive probability, calculated with
`pbeta(q = 0.4, shape1 = 2, shape2 = 5) -
pbeta(q = 0.2, shape1 = 2, shape2 = 5)`. The exact value has probability
zero under this continuous model. Its density at 0.2 is not an exact-value
probability.

:::
::::

# Choosing a continuous model {#sec-choosing_continuous_model}

The measured variable's support is a first check, followed by its shape and
the ecological process that produced it.

| Model | Support | Parameters in R | Starting question |
|---|---|---|---|
| Normal | All real values | `mean`, `sd` | Is a symmetric shape plausible, including its tails? |
| Lognormal | Strictly positive values | `meanlog`, `sdlog` | Is the positive measurement right-skewed? |
| Beta | Values strictly between 0 and 1 | `shape1`, `shape2` | Is this a continuous fraction without exact boundaries? |

These are starting models, not names we attach to variables automatically. A body-mass
column does not come with “lognormal” written on it, and a fraction does not make all the
awkward boundary values disappear. Compare the assumed density and cumulative distribution
with observations, check the measurement process, and remember that a finite sample will
differ from its generating model. The useful question is whether the model captures the
features needed to answer your ecological question.

# Summary

Continuous models give us a way to connect a measurement scale with questions about
ecologically meaningful ranges. Interval probabilities are areas under a density; a density
at one exact value is not a probability. R's `d`, `p`, `q`, and `r` functions connect
density, cumulative probability, quantiles, and simulation. We can then compare a proposed
model with a finite sample using density-scaled histograms and empirical cumulative
distributions.

When each observational unit has two measurements, joint distributions keep the pair,
marginal distributions describe one measurement across all units. Normal,
lognormal, and beta models have different supports and shapes. Choosing among them starts
with the measurement and the ecology, then continues with an honest look at the data.

# Chapter quiz

The quiz asks you to move between a measurement, the model, and the question it can answer.
Try each question before opening its answer.

:::: {.callout-important title="Question 1: exact height"}

Under a continuous tree-height model, what is $P(X=22)$? Can a height still be
recorded as 22 m?

::: {.callout-tip title="Answer" collapse="true"}

The probability of exactly 22 m is zero. A recorded 22 m represents a range
at the measurement's precision, so it can occur.

:::
::::

:::: {.callout-important title="Question 2: density or probability"}

Which R expression gives the probability of height between 18 m and 26 m
under a normal model with mean 22 m and standard deviation 4 m?

::: {.callout-tip title="Answer" collapse="true"}

`pnorm(q = 26, mean = 22, sd = 4) - pnorm(q = 18, mean = 22, sd = 4)`.
The difference gives the probability over an interval; `dnorm()` at one
height gives density.

:::
::::

:::: {.callout-important title="Question 3: a sample and its model"}

Why might a density-scaled histogram of 100 simulated heights differ from
the theoretical density used to generate them?

::: {.callout-tip title="Answer" collapse="true"}

A finite sample varies at random, and the histogram also depends on its
bins. Exact agreement is not expected.

:::
::::

:::: {.callout-important title="Question 5: choose a model"}

Give a plausible starting model for a positive, right-skewed body mass and
for vegetation cover measured strictly between zero and one. State one
limitation of each choice.

::: {.callout-tip title="Answer" collapse="true"}

A lognormal model is plausible for body mass but cannot include zero.
A beta model is plausible for continuous fractional cover but does not
include exact zero or one observations. Both choices need checking against
the measurement process and data.

:::
::::

:::: {.callout-important title="Question 6: a quantile"}

Under the assumed normal tree-height model, what does the value returned
by `qnorm(p = 0.95, mean = 22, sd = 4)` mean?

::: {.callout-tip title="Answer" collapse="true"}

It is the model height below which 95% of tree heights fall. About 5%
are above it. It is not the height of a particular tree.

:::
::::
