---
title: "Introduction to Probability for Continuous Data"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# Introduction

In [Chapter 2](../Chapter_2/Theory.qmd), we counted outcomes such as surviving
seedlings and wildlife detections. Ecologists also measure tree height, animal body
mass, and vegetation cover. What probability model could describe these measurements,
and what would its predictions mean?

We will start with the ideas shared by continuous measurements: density, probabilities
over intervals, samples, and relationships between two variables. We will then use
normal, lognormal, and beta distributions to describe different ecological settings.
Our examples are assumed models or simulations, not measurements from a field study.

# Learning goals

After studying this chapter, you should be able to:

- distinguish continuous measurements from discrete counts and explain density;
- calculate and interpret probabilities over intervals and quantiles;
- use density, cumulative probability, quantile, and sampling functions in R;
- compare a theoretical distribution with a histogram, density estimate, and
  empirical cumulative distribution;
- distinguish joint, marginal, and conditional distributions;
- interpret scatterplots and contour plots for two continuous variables;
- describe the support and parameters of normal, lognormal, and beta distributions;
- choose a plausible starting distribution for an ecological measurement and
  explain its limitations.

# Continuous measurements and probability {#sec-continuous_measurements}

Suppose $X$ is the height in metres of a randomly selected tree from a defined
population. Before selecting the tree, its height is uncertain. A model might treat
$X$ as **continuous**: within a plausible range, it can take any value, not just a
whole number. A seedling count, by contrast, takes separate integer values.

Recorded measurements have limited precision. A recorded height of 22.1 m might
represent an underlying height between 22.05 and 22.15 m. A continuous model is
often a useful approximation to that underlying measurement.

::: {.callout-note title="Continuous random variable"}
A continuous random variable can take values throughout an interval. Its model
assigns probabilities to ranges of values. The values allowed by the model form
its **support**.
:::

## From probability mass to density

For the discrete variables in [Chapter 2](../Chapter_2/Theory.qmd#sec-four_faces),
the probability mass at a particular count is $P(X=x)$. For an ideal continuous
variable, the probability of any one exact value is zero:

$$
P(X=x)=0.
$$

This does not prevent us from recording a height of 22.1 m. The recorded number
represents a range at the instrument's precision. Questions about a continuous
model concern intervals, such as $P(20<X\leq25)$.

A **probability density function**, $f_X(x)$, describes how probability is
concentrated along the measurement scale. The probability of an interval is the
area under its density curve:

$$
P(a<X\leq b)=\int_a^b f_X(x)\,dx.
$$

The density cannot be negative, and its total area over the support is one.
Because exact points have probability zero, including or excluding an interval's
endpoints does not change its probability under a continuous model.

::: {.callout-warning title="Density is not an exact-value probability"}
The value $f_X(22)$ is a curve height, not $P(X=22)$. A density can exceed one if
probability is concentrated in a narrow interval. Its value also changes when,
for example, heights are expressed in centimetres rather than metres; the
probability for the same physical interval does not.
:::

:::: {.callout-important title="Practice: measuring tree height"}

A tree height is recorded to the nearest tenth of a metre as 22.1 m. What range
of underlying heights could have produced that record? Would a continuous model
assign positive probability to an underlying height of exactly 22.1 m?

::: {.callout-tip title="Solution" collapse="true"}

Rounding to the nearest tenth, heights from approximately 22.05 m up to 22.15 m
could be recorded as 22.1 m. Under an ideal continuous model,
$P(X=22.1)=0$, but the recording interval can have positive probability.

:::
::::

# The four faces of a continuous distribution {#sec-continuous_four_faces}

We met the [four faces of a distribution](../Chapter_2/Theory.qmd#sec-four_faces)
for counts. The R prefixes are the same for continuous variables, but `d*` now
returns **density** rather than the probability of an exact outcome.

For a running example, suppose tree heights follow a normal model with mean 22 m
and standard deviation 4 m. These numbers describe an assumed population; we
will examine the normal model itself later in this chapter.

## 1. Probability density

The density is highest near heights the model considers most concentrated. For
the assumed model, `dnorm()` evaluates its density at chosen heights:

```r
dnorm(x = 22, mean = 22, sd = 4)
```

The result has units of density per metre. It is not the probability that a tree
is exactly 22 m tall. The curve helps us see where probability is concentrated.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Assumed normal density for tree height. Interval probability is area under the curve."
height_values = seq(from = 8, to = 36, length.out = 400)
height_density = dnorm(x = height_values, mean = 22, sd = 4)
plot(
  x = height_values, y = height_density, type = "l", lwd = 2,
  xlab = "Tree height (m)", ylab = "Density per metre"
)
```

## 2. Cumulative probability

The **cumulative distribution function** is $F_X(x)=P(X\leq x)$. It adds the
probability for all heights up to $x$. To calculate the probability that a tree
is between 18 m and 26 m tall, subtract two cumulative probabilities:

$$
P(18<X\leq26)=F_X(26)-F_X(18).
$$

In R:

```r
pnorm(q = 26, mean = 22, sd = 4) -
  pnorm(q = 18, mean = 22, sd = 4)
```

For a height above 30 m, use the upper tail:

```r
pnorm(q = 30, mean = 22, sd = 4, lower.tail = FALSE)
```

The cumulative curve rises smoothly from zero towards one.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Cumulative probability for the assumed tree-height model."
height_values = seq(from = 8, to = 36, length.out = 400)
plot(
  x = height_values,
  y = pnorm(q = height_values, mean = 22, sd = 4),
  type = "l", lwd = 2, ylim = c(0, 1),
  xlab = "Tree height (m)", ylab = "Cumulative probability"
)
```

## 3. Quantiles

A **quantile** reverses the cumulative question: given a cumulative probability,
which height marks that point in the distribution? The 0.95 quantile is the
height below which the model places 95% of tree heights:

```r
qnorm(p = 0.95, mean = 22, sd = 4)
```

Because this model is continuous, $P(X\leq q_{0.95})=0.95$. A quantile is a
model prediction, not the height of a particular observed tree.

## 4. Sampling

We can generate heights from the model to see the variation it predicts:

```r
set.seed(seed = 42)
tree_heights = rnorm(n = 100, mean = 22, sd = 4)
head(tree_heights)
```

Each value represents one simulated tree. A finite sample will not have exactly
the model's mean or standard deviation. A new seed produces a different sample.

::: {.callout-note title="The four R function types"}
For continuous distributions, `d*` gives density, `p*` cumulative probability,
`q*` a quantile, and `r*` a random sample. The first argument is conventionally
named `x`, `q`, `p`, or `n`, respectively.
:::

:::: {.callout-important title="Practice: a tree-height threshold"}

Under the assumed normal model, calculate the probability that a tree is taller
than 26 m. Find the height that marks the lower 10% of the model. Explain each
result in words.

::: {.callout-tip title="Solution" collapse="true"}

```r
pnorm(q = 26, mean = 22, sd = 4, lower.tail = FALSE)
qnorm(p = 0.10, mean = 22, sd = 4)
```

The first result is approximately 0.159: about 16% of heights exceed 26 m under
this model. The second is approximately 16.9 m: about 10% fall below that height.

:::
::::

# From a model to observed measurements {#sec-continuous_samples}

Chapter 2 introduced the distinction between a
[population distribution and a sample](../Chapter_2/Theory.qmd#sec-sample_population).
We use the same distinction for continuous measurements. A theoretical density
describes a proposed model; a plot of measurements summarises a finite sample.

## Histograms and density estimates

A **histogram** groups observations into intervals called bins. On a frequency
histogram, bar height is the number of observations in a bin when widths are
equal. On a density histogram, the *total bar area* is one, so it can be compared
with a theoretical density curve. Changing the bin width can change the visible
detail without changing the data.

A **kernel density estimate**, calculated by `density()`, smooths the observed
measurements. Its bandwidth controls the smoothing. It is an empirical estimate,
while `dnorm()` describes the assumed model.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "A simulated sample of tree heights, its kernel density estimate, and the assumed normal density."
set.seed(seed = 73)
tree_heights = rnorm(n = 200, mean = 22, sd = 4)
hist(
  x = tree_heights, breaks = 16, probability = TRUE,
  col = "grey85", border = "white",
  xlab = "Tree height (m)", ylab = "Density per metre",
  main = "Simulated tree heights"
)
lines(density(x = tree_heights), lwd = 2, lty = 2)
curve(dnorm(x = x, mean = 22, sd = 4), add = TRUE, lwd = 2)
legend(
  "topright", legend = c("Sample estimate", "Assumed model"),
  lty = c(2, 1), lwd = 2, bty = "n"
)
```

The differences between the two curves can arise from sampling variation even
when the assumed model generated the data. Agreement in one plot alone would not
establish that the model is suitable for field measurements.

## Empirical cumulative distributions

The **empirical cumulative distribution** gives the fraction of sample values at
or below a threshold. `ecdf()` calculates it. It rises in steps, whereas a
continuous model's cumulative curve is smooth.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Empirical cumulative distribution of simulated heights and the assumed model's cumulative distribution."
set.seed(seed = 73)
tree_heights = rnorm(n = 200, mean = 22, sd = 4)
plot(
  ecdf(tree_heights), do.points = FALSE, verticals = TRUE,
  xlab = "Tree height (m)", ylab = "Cumulative proportion",
  main = "Simulated heights and model"
)
curve(pnorm(q = x, mean = 22, sd = 4), add = TRUE, lwd = 2, lty = 2)
legend(
  "bottomright", legend = c("Sample", "Assumed model"),
  lty = c(1, 2), lwd = 2, bty = "n"
)
```

:::: {.callout-important title="Practice: sample or model?"}

You simulate 200 heights from the assumed model and make a histogram. Would a
different seed produce exactly the same histogram? Would changing the bin width
change the sample mean?

::: {.callout-tip title="Solution" collapse="true"}

A different seed usually produces a different sample and histogram. Changing
the bin width changes the display, not the measured values, so the sample mean
stays the same.

:::
::::

# Two continuous variables {#sec-two_continuous_variables}

Ecologists often measure two variables on each observational unit. Suppose $X$
is a tree's diameter in centimetres and $Y$ is its height in metres. One row of
data holds both measurements for the same tree. The model must describe their
individual variation and how they vary together.

## Joint, marginal, and conditional distributions

The **joint distribution** describes combinations of diameter and height. Its
joint density, $f_{X,Y}(x,y)$, places density across a two-dimensional region.
Probabilities belong to regions, such as trees with diameters from 25 to 35 cm
*and* heights above 20 m, not to one exact pair of measurements.

The **marginal distribution** of height describes heights across all trees in
the population, without selecting trees by diameter. A histogram of all heights
estimates that marginal distribution.

The **conditional distribution** of height describes heights among trees with
specified information about diameter. For example, we might compare heights
among trees with diameters of 20–25 cm with heights among trees with diameters
of 35–40 cm. These are distributions conditional on *diameter ranges*. A model
can also describe height conditional on an exact diameter, even though that
exact value has probability zero under a continuous model.

::: {.callout-note title="Three questions about two measurements"}
Joint: *Which combinations of height and diameter occur?* Marginal: *How do
heights vary across all trees?* Conditional: *How do heights vary among trees
in a stated diameter group?*
:::

Chapter 2 introduced [independence](../Chapter_2/Theory.qmd#sec-independence_expectations).
For two continuous variables, independence means that knowing diameter would
not change the model's distribution of height. A relationship in a scatterplot
may challenge that assumption, but it does not by itself identify its ecological
cause.

## Scatterplots and contours

A scatterplot shows the paired observations. Here we simulate trees whose height
and diameter tend to increase together. The parameters describe the simulation,
not a claim about a real forest.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Simulated tree heights and diameters with a positive association."
set.seed(seed = 50)
n_trees = 300
z_height = rnorm(n = n_trees)
z_other = rnorm(n = n_trees)
tree_height = 22 + 4 * z_height
tree_diameter = 30 + 5 * (0.7 * z_height + sqrt(1 - 0.7^2) * z_other)
plot(
  x = tree_diameter, y = tree_height, pch = 16,
  col = grDevices::adjustcolor("black", alpha.f = 0.35),
  xlab = "Tree diameter (cm)", ylab = "Tree height (m)"
)
```

Larger diameters tend to accompany greater heights in this simulation, but
individual trees vary. A marginal height plot would hide the pairing. Comparing
height distributions in two diameter groups would show how the conditional
distribution changes.

A **contour plot** joins combinations of diameter and height that have equal
joint density, like a topographic map joins places of equal elevation. Contours
show where the *model* concentrates outcomes. A contour line is not a probability
boundary, and the regions between successive lines do not necessarily contain
equal probability.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Contours of the assumed joint density for tree diameter and height."
diameter_grid = seq(from = 10, to = 50, length.out = 100)
height_grid = seq(from = 8, to = 36, length.out = 100)
rho = 0.7
joint_density = outer(
  X = diameter_grid, Y = height_grid,
  FUN = function(diameter, height) {
    dx = (diameter - 30) / 5
    hy = (height - 22) / 4
    exp(-(dx^2 - 2 * rho * dx * hy + hy^2) / (2 * (1 - rho^2))) /
      (2 * pi * 5 * 4 * sqrt(1 - rho^2))
  }
)
contour(
  x = diameter_grid, y = height_grid, z = joint_density,
  nlevels = 8, xlab = "Tree diameter (cm)",
  ylab = "Tree height (m)"
)
```

These tilted contours show that larger diameter and greater height tend to
occur together under the assumed model. A scatterplot reveals individual
observations; contours summarise the joint density. Neither plot establishes
that changing diameter would cause a change in height.

:::: {.callout-important title="Practice: two measurements on a tree"}

Suppose you select only trees with diameters between 35 and 40 cm and plot
their heights. Is that plot joint, marginal, or conditional? What would a plot
of the heights of *all* sampled trees show?

::: {.callout-tip title="Solution" collapse="true"}

The first plot estimates a conditional height distribution, given a diameter
range. The second estimates the marginal height distribution of the sampled
population.

:::
::::

# The normal distribution {#sec-normal_distribution}

We can now look more closely at the model used for tree height. A **normal
distribution** is symmetric around its mean $\mu$. Its standard deviation
$\sigma>0$ sets its spread. We write $X\sim\operatorname{Normal}(\mu,\sigma^2)$:
the second quantity in this notation is the **variance**, while R's `sd`
argument takes the standard deviation $\sigma$.

The normal model has support over all real values. Its density is

$$
f_X(x)=\frac{1}{\sigma\sqrt{2\pi}}
\exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right).
$$

It has $E[X]=\mu$ and $\operatorname{Var}(X)=\sigma^2$. The mean, median, and
mode coincide because the curve is symmetric. Increasing $\sigma$ spreads the
same total probability over a wider range and lowers the curve's peak.

For the assumed tree heights, $\mu=22$ m and $\sigma=4$ m. The mean is the
average height across many trees generated by this model; the standard deviation
describes variation among those heights. Approximately 68% of a normal model
falls within one standard deviation of its mean, 95% within two, and 99.7%
within three. Thus about 95% of this model's heights fall between 14 m and
30 m. These percentages are approximations, and they describe the model.

We can express a height in standard deviation units with
$Z=(X-\mu)/\sigma$. A 30 m tree is two standard deviations above the assumed
mean. R calculates probabilities directly with `mean` and `sd`, so you need not
standardise each observation by hand.

```r
pnorm(q = 30, mean = 22, sd = 4, lower.tail = FALSE)
```

::: {.callout-warning title="Check the support"}
A normal model allows negative values. For positive measurements such as body
mass, check whether it places meaningful probability below zero or misses
strong right skew. Symmetry alone does not make a model suitable.
:::

:::: {.callout-important title="Practice: spread in tree heights"}

Two proposed normal models have the same mean height, 22 m, but standard
deviations of 2 m and 5 m. Which assigns more probability to heights far from
22 m? What happens to the density's peak as spread increases?

::: {.callout-tip title="Solution" collapse="true"}

The model with standard deviation 5 m is more spread out and gives more
probability to heights far from 22 m. Its density peak is lower because each
density curve still has total area one.

:::
::::

# The lognormal distribution {#sec-lognormal_distribution}

Animal body mass is positive and may have a long right tail: most animals are
relatively light, while a few are much heavier. A **lognormal distribution**
can describe this shape. A positive variable $X$ is lognormal when its
logarithm is normal:

$$
\log(X)\sim\operatorname{Normal}(\mu,\sigma^2).
$$

The model's support is $x>0$. Its density is

$$
f_X(x)=\frac{1}{x\sigma\sqrt{2\pi}}
\exp\left(-\frac{(\log x-\mu)^2}{2\sigma^2}\right),\qquad x>0.
$$

In R, `meanlog` is $\mu$ and `sdlog` is $\sigma$. Both describe *log body
mass*, not body mass in its original units. The original-scale median is
$e^\mu$, the mean is $e^{\mu+\sigma^2/2}$, and the variance is
$e^{2\mu+\sigma^2}(e^{\sigma^2}-1)$. The mean exceeds the median when
$\sigma>0$ because the right tail includes unusually large values.

Suppose body mass in grams follows a lognormal model with
`meanlog = log(30)` and `sdlog = 0.5`. Its median is 30 g, while its mean is
larger. We can calculate a probability and a quantile on the original scale:

```r
plnorm(q = 50, meanlog = log(30), sdlog = 0.5,
        lower.tail = FALSE)
qlnorm(p = 0.95, meanlog = log(30), sdlog = 0.5)
```

The log scale can make a multiplicative process easier to describe: successive
proportional changes multiply body mass, while their logarithms add. This is
one possible ecological motivation, not proof that particular measurements
follow a lognormal model.

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Simulated lognormal body masses on the original and log scales."
set.seed(seed = 20)
body_mass = rlnorm(n = 500, meanlog = log(30), sdlog = 0.5)
old_par = par(no.readonly = TRUE)
par(mfrow = c(1, 2))
hist(body_mass, breaks = 25, col = "grey80", border = "white",
     xlab = "Body mass (g)", main = "Original scale")
hist(log(body_mass), breaks = 25, col = "grey80", border = "white",
     xlab = "Log body mass", main = "Log scale")
par(old_par)
```

The original-scale simulation is right-skewed; its logarithms are more
symmetric. The four functions are `dlnorm()`, `plnorm()`, `qlnorm()`, and
`rlnorm()`.

::: {.callout-warning title="Zero is outside a lognormal model"}
A lognormal model cannot generate zero or negative values. A recorded zero
could mean absence, a detection limit, or another process. Its meaning should
be investigated before changing the values or choosing a model.
:::

:::: {.callout-important title="Practice: mean and median mass"}

For `meanlog = log(30)` and `sdlog = 0.5`, calculate the median and mean body
mass. Why are they different?

::: {.callout-tip title="Solution" collapse="true"}

```r
median_mass = exp(log(30))
mean_mass = exp(log(30) + 0.5^2 / 2)
c(median_g = median_mass, mean_g = mean_mass)
```

The median is 30 g and the mean is approximately 34 g. Large possible masses
in the right tail pull the mean above the median.

:::
::::

# The beta distribution {#sec-beta_distribution}

Suppose $X$ is the fraction of a quadrat covered by vegetation, measured on a
continuous scale. Values inside $(0,1)$ have a natural lower and upper bound.
A **beta distribution** can describe different shapes within that interval.
It is a model for a continuous fraction, not for the number of covered points
out of a fixed number of observations.

The beta distribution has two positive shape parameters, $a$ and $b$. Its
density is

$$
f_X(x)=\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}
x^{a-1}(1-x)^{b-1},\qquad 0<x<1.
$$

Here $\Gamma$ is the gamma function, which provides the scaling needed for
total area one. The mean and variance are

$$
E[X]=\frac{a}{a+b},\qquad
\operatorname{Var}(X)=\frac{ab}{(a+b)^2(a+b+1)}.
$$

If $a>b$, density tends towards higher cover; if $a<b$, it tends towards
lower cover. The pair also controls how concentrated or U-shaped the density
is. R calls the parameters `shape1` and `shape2`.

Suppose vegetation cover follows a beta model with $a=2$ and $b=5$.
Its expected cover is $2/7$, or about 29% of a quadrat. We can calculate
the chance of cover below 20% and the 95th percentile:

```r
pbeta(q = 0.20, shape1 = 2, shape2 = 5)
qbeta(p = 0.95, shape1 = 2, shape2 = 5)
```

```{r}
#| echo: false
#| eval: true
#| fig-cap: "Beta densities with different shapes for a fraction between zero and one."
cover_values = seq(from = 0.001, to = 0.999, length.out = 400)
plot(cover_values, dbeta(x = cover_values, shape1 = 2, shape2 = 5),
     type = "l", lwd = 2, ylim = c(0, 11),
     xlab = "Vegetation cover (fraction)", ylab = "Density")
lines(cover_values, dbeta(x = cover_values, shape1 = 5, shape2 = 2),
      lwd = 2, lty = 2)
lines(cover_values, dbeta(x = cover_values, shape1 = 0.5, shape2 = 0.5),
      lwd = 2, lty = 3)
legend("top", legend = c("a = 2, b = 5", "a = 5, b = 2",
                         "a = 0.5, b = 0.5"),
       lty = 1:3, lwd = 2, bty = "n")
```

The four R functions are `dbeta()`, `pbeta()`, `qbeta()`, and `rbeta()`.

::: {.callout-warning title="Exact zero and one values"}
The continuous beta model assigns no probability to an exact cover of zero
or one. Such observations may be ecologically meaningful. Do not replace
them with small adjustments automatically; first ask how cover was measured
and whether boundary values need a separate process in the model.
:::

:::: {.callout-important title="Practice: interpreting a cover model"}

Would a beta model with $a=2$ and $b=5$ assign positive probability to
cover between 0.2 and 0.4? Would it assign positive probability to cover
of exactly 0.2? Explain the difference.

::: {.callout-tip title="Solution" collapse="true"}

The interval has positive probability, calculated with
`pbeta(q = 0.4, shape1 = 2, shape2 = 5) -
pbeta(q = 0.2, shape1 = 2, shape2 = 5)`. The exact value has probability
zero under this continuous model. Its density at 0.2 is not an exact-value
probability.

:::
::::

# Choosing a continuous model {#sec-choosing_continuous_model}

The measured variable's support is a first check, followed by its shape and
the ecological process that produced it.

| Model | Support | Parameters in R | Starting question |
|---|---|---|---|
| Normal | All real values | `mean`, `sd` | Is a symmetric shape plausible, including its tails? |
| Lognormal | Strictly positive values | `meanlog`, `sdlog` | Is the positive measurement right-skewed? |
| Beta | Values strictly between 0 and 1 | `shape1`, `shape2` | Is this a continuous fraction without exact boundaries? |

These are starting models. A variable's name alone does not determine its
distribution. Compare the assumed density and cumulative distribution with
observations, check the measurement process, and remember that a finite
sample will differ from its generating model.

# Summary

For continuous models, interval probabilities are areas under a density;
an exact-value density is not a probability. R's `d`, `p`, `q`, and `r`
functions connect density, cumulative probability, quantiles, and simulation.
Samples can be compared with models using density-scaled histograms and
empirical cumulative distributions.

With two measurements per unit, joint distributions describe pairs, marginal
distributions describe one measurement across all units, and conditional
distributions describe one measurement within a specified group of the other.
Normal, lognormal, and beta models describe different supports and shapes.

# Chapter quiz

Try each question before opening its answer.

:::: {.callout-important title="Question 1: exact height"}

Under a continuous tree-height model, what is $P(X=22)$? Can a height still be
recorded as 22 m?

::: {.callout-tip title="Answer" collapse="true"}

The probability of exactly 22 m is zero. A recorded 22 m represents a range
at the measurement's precision, so it can occur.

:::
::::

:::: {.callout-important title="Question 2: density or probability"}

Which R expression gives the probability of height between 18 m and 26 m
under a normal model with mean 22 m and standard deviation 4 m?

::: {.callout-tip title="Answer" collapse="true"}

`pnorm(q = 26, mean = 22, sd = 4) - pnorm(q = 18, mean = 22, sd = 4)`.
The difference gives the probability over an interval; `dnorm()` at one
height gives density.

:::
::::

:::: {.callout-important title="Question 3: a sample and its model"}

Why might a density-scaled histogram of 100 simulated heights differ from
the theoretical density used to generate them?

::: {.callout-tip title="Answer" collapse="true"}

A finite sample varies at random, and the histogram also depends on its
bins. Exact agreement is not expected.

:::
::::

:::: {.callout-important title="Question 4: paired measurements"}

You plot heights only for trees whose diameters are 30–35 cm. Which type
of distribution does this estimate? What plot would show paired height and
diameter observations?

::: {.callout-tip title="Answer" collapse="true"}

The height plot estimates a conditional distribution given a diameter range.
A scatterplot shows the paired observations.

:::
::::

:::: {.callout-important title="Question 5: choose a model"}

Give a plausible starting model for a positive, right-skewed body mass and
for vegetation cover measured strictly between zero and one. State one
limitation of each choice.

::: {.callout-tip title="Answer" collapse="true"}

A lognormal model is plausible for body mass but cannot include zero.
A beta model is plausible for continuous fractional cover but does not
include exact zero or one observations. Both choices need checking against
the measurement process and data.

:::
::::

:::: {.callout-important title="Question 6: a quantile"}

Under the assumed normal tree-height model, what does the value returned
by `qnorm(p = 0.95, mean = 22, sd = 4)` mean?

::: {.callout-tip title="Answer" collapse="true"}

It is the model height below which 95% of tree heights fall. About 5%
are above it. It is not the height of a particular tree.

:::
::::
