---
title: "Calculus refresher"
---

```{r}
#| include: false
# Load ggplot2 for the rendered figure of this page.
library(ggplot2)
```

This page recalls the rules of calculus that [Chapter 5](../Chapter_5/Theory.qmd) uses to
find out the shape of a function from its formula: exponents and logarithms, limits, and
derivatives. It is written for readers who have met these rules before and need to recall
them, so it is a reference to work through with pen and paper and it does not replace a
calculus course. Every rule comes with a worked case and each section ends with a short
exercise. A longer set of problems with solutions is at the end (@sec-calc-practice).

Throughout the page, $x$ is the input of a function and $y=f(x)$ is its output. The letters
$a$, $b$, $c$ and $n$ stand for constants, that is, for quantities that do not change when
$x$ changes. In Chapter 5 (onwards) these constants are the parameters we estimate from data.

# Exponents and logarithms {#sec-calc-exponents}

An **exponent** says how many times a quantity is multiplied by itself: $x^3=x\cdot x\cdot x$.
The definition is extended to exponents that are zero, negative or fractional in such a way
that the following rules hold for any positive $x$ and $y$:

$$
\begin{aligned}
x^m x^n &= x^{m+n}, & \frac{x^m}{x^n} &= x^{m-n}, & (x^m)^n &= x^{mn}, \\
x^0 &= 1, & x^{-n} &= \frac{1}{x^n}, & x^{1/n} &= \sqrt[n]{x}, \\
(xy)^n &= x^n y^n. & & & &
\end{aligned}
$$

For example, $x^2\sqrt{x}/x^3=x^{2+1/2-3}=x^{-1/2}=1/\sqrt{x}$. We will use these rules all
the time to simplify the functions of Chapter 5 and their derivatives.

The **exponential function** is $e^x$, where $e\approx2.718$ is a constant chosen because
it makes the derivative of the function as simple as it can be (see @sec-calc-rules). It
is also written $\exp(x)$, which is how R writes it (`exp()`). Because it is a power of a
positive number, $e^x$ is positive for any $x$, it equals 1 at $x=0$, and it follows the
rules of exponents: $e^{x}e^{y}=e^{x+y}$, $(e^{x})^n=e^{nx}$ and $e^{-x}=1/e^{x}$.

The **logarithm** is the inverse of the exponential function: $\log x$ is the exponent to
which $e$ must be raised to obtain $x$. That is,

$$
\log\left(e^{x}\right)=x \qquad \text{and} \qquad e^{\log x}=x \quad (x>0).
$$

The logarithm is only defined for positive values, with $\log 1=0$ and $\log e=1$. It is
negative for $0<x<1$ and positive for $x>1$. The rules of exponents become the rules of
logarithms:

$$
\log(xy)=\log x+\log y, \qquad \log\left(\frac{x}{y}\right)=\log x-\log y, \qquad
\log\left(x^n\right)=n\log x.
$$

In this book $\log$ always means the natural logarithm (base $e$), as in R, where `log()`
is the natural logarithm and `log10()` the logarithm in base 10. The two differ by a
constant factor, $\log_{10}x=\log x/\log 10$.

We will mostly use logarithms to solve an equation for a quantity that sits in an exponent.
There are two cases that return throughout Chapter 5.

In the first case the unknown is inside an exponential. Suppose a pesticide residue on
leaves declines with time as $20e^{-0.3t}$ mg/kg, where $t$ is the time in days. The
residue is half of its initial value when $20e^{-0.3t}=10$. We divide by 20 and take
logarithms on both sides:

$$
\begin{aligned}
e^{-0.3t} &= \tfrac{1}{2} \\
-0.3t &= \log\tfrac{1}{2} = -\log 2 \\
t &= \frac{\log 2}{0.3} \approx 2.31 \text{ days}.
\end{aligned}
$$

Note that the initial value cancelled: in general, $ae^{-bx}=a/2$ gives $x=\log 2/b$
whatever the value of $a$.

In the second case the unknown is the base or the exponent of a power, $y=ax^b$. Suppose
$y=2x^{1.5}$ and we want the value of $x$ for which $y=54$. We divide by 2 and raise both
sides to the power $1/1.5$, which gives $x=27^{1/1.5}=27^{2/3}=9$. If instead we knew that
$y=54$ at $x=9$ and that $a=2$, we could solve for the exponent by taking logarithms:
$\log 27=b\log 9$, so $b=\log 27/\log 9=1.5$. We can verify our math with R:

```{r}
# Time (days) at which 20 * exp(-0.3 * t) has fallen to half of its initial value,
# and the value of the function at that time (mg/kg).
t_half = log(2) / 0.3
t_half
20 * exp(-0.3 * t_half)

# Solve 54 = 2 * x^1.5 for x, and 54 = 2 * 9^b for b.
(54 / 2)^(1 / 1.5)
log(54 / 2) / log(9)
```

:::: {.callout-important title="Practice: exponents and logarithms"}

1. Write $\log\left(e^{3}x^{2}\right)$ without the exponential.
2. Solve $20e^{-0.5t}=5$ for $t$.

::: {.callout-tip title="Solution" collapse="true"}

1. The logarithm of a product is a sum of logarithms, so
   $\log\left(e^{3}x^{2}\right)=\log\left(e^{3}\right)+\log\left(x^{2}\right)=3+2\log x$.
2. Dividing by 20 gives $e^{-0.5t}=1/4$, so $-0.5t=-\log 4$ and
   $t=\log 4/0.5\approx2.77$.

:::
::::

# Limits {#sec-calc-limits}

The **limit** of a function $f(x)$ as $x$ approaches a value $x_0$ is the value that
$f(x)$ comes arbitrarily close to as $x$ comes close to $x_0$. We write it as
$\lim_{x\to x_0}f(x)$. In Chapter 5 we are mostly interested in the two ends of a function:
what happens as $x$ grows without bound, written $x\to\infty$, and what happens as $x$
approaches zero (see [Limits and asymptotes](../Chapter_5/Theory.qmd#sec-function-limits)).
When the function is defined at $x_0$ and has no jump there, its limit is the value
$f(x_0)$. For example, $ae^{-bx}\to a$ as $x\to0$, because $e^0=1$.

For limits at infinity with positive $b$ and $n$, three results are enough to start with:

$$
\lim_{x\to\infty}x^{n}=\infty, \qquad \lim_{x\to\infty}\frac{1}{x^{n}}=0, \qquad
\lim_{x\to\infty}e^{-bx}=0.
$$

When a function is a sum of several terms, its behaviour for large $x$ is set by the term
that grows fastest. Exponentials grow faster than any power of $x$, a higher power grows
faster than a lower power, and any power grows faster than a constant or a logarithm. In
other words, for large $x$ we can keep the fastest-growing term of a sum and ignore the
others. For small $x$ the ranking is reversed: the lowest power of $x$ is the one that
remains.

For a ratio of two sums, this idea is applied by dividing the numerator and the denominator
by the highest power of $x$. For the function $f(x)=ax/(b+x)$:

$$
\lim_{x\to\infty}\frac{ax}{b+x}=\lim_{x\to\infty}\frac{a}{b/x+1}=\frac{a}{0+1}=a.
$$

In the same way, $(2x^2+3x)/(x^2+5x)$ approaches $2/1=2$ as $x\to\infty$, because the
terms in $x^2$ dominate. As $x\to0$ the terms in $x$ dominate and the ratio approaches
$3/5$.

The ranking of terms is a consequence of **L'Hôpital's rule**.^[The rule is named after
de l'Hôpital, who printed it in his textbook of 1696; it is due to his teacher Johann
Bernoulli.] The rule applies to a ratio $f(x)/g(x)$ in which numerator and denominator
both approach zero or both approach infinity, so that the limit of the ratio cannot be
read directly. In that case,

$$
\lim_{x\to x_0}\frac{f(x)}{g(x)}=\lim_{x\to x_0}\frac{f'(x)}{g'(x)},
$$

where $f'$ and $g'$ are the derivatives of $f$ and $g$ (@sec-calc-derivatives), provided
that the limit on the right exists. The rule also holds for $x\to\infty$. For example, the
function $xe^{-bx}$ is the product of a term that grows and a term that approaches zero.
Written as a ratio it is $x/e^{bx}$, in which both parts approach infinity, so

$$
\lim_{x\to\infty}xe^{-bx}=\lim_{x\to\infty}\frac{x}{e^{bx}}
=\lim_{x\to\infty}\frac{1}{be^{bx}}=0.
$$

That is, the exponential wins. A limit can always be checked numerically by evaluating the
function at a large (or a small) value of $x$:

```{r}
# a * x / (b + x) with a = 10 and b = 2 approaches a = 10 for large x.
10 * 1e6 / (2 + 1e6)

# x * exp(-b * x) with b = 0.5 approaches zero for large x.
100 * exp(-0.5 * 100)
```

:::: {.callout-important title="Practice: limits"}

For the function $f(x)=\dfrac{5x+1}{2x+10}$, calculate $f(0)$ and
$\lim_{x\to\infty}f(x)$.

::: {.callout-tip title="Solution" collapse="true"}

At zero the function is defined, so $f(0)=1/10=0.1$. For large $x$ we divide numerator and
denominator by $x$, which gives $(5+1/x)/(2+10/x)$. Both $1/x$ and $10/x$ approach zero,
so the limit is $5/2=2.5$.

:::
::::

# The derivative {#sec-calc-derivatives}

The **derivative** of a function at a point is the slope of the function at that point:
the change in $y$ per unit change in $x$. For a straight line the slope is the same
everywhere. For a curve it changes from point to point, and it is defined as follows.
Take two points on the curve, at $x$ and at $x+h$. The straight line through them is a
**secant** and its slope is the change in $y$ divided by the change in $x$:

$$
\frac{f(x+h)-f(x)}{h}.
$$

As $h$ becomes smaller, the secant approaches the **tangent**, the straight line that
touches the curve at $x$ (@fig-secant). The derivative is the slope of the tangent, that
is, the limit of the slope of the secant:

$$
f'(x)=\frac{df}{dx}=\lim_{h\to0}\frac{f(x+h)-f(x)}{h}.
$$

The two notations mean the same.^[The notation $df/dx$ is due to Leibniz (1675) and $f'$
to Lagrange (1797); both are in use.] We write $f'(x)$ in running text and $df/dx$ when
we need to show with respect to which quantity the derivative is taken. Since a
derivative is a change in $y$ divided by a change in $x$, its units are the units of $y$
divided by the units of $x$. If $y$ is the height of a plant in cm and $x$ is time in days,
$f'(x)$ is a growth rate in cm per day.

```{r}
#| echo: false
#| label: fig-secant
#| fig-cap: "The function $f(x)=x^2$ with three secants through the point at $x=1$ (grey lines, for $h=1.5$, 1 and 0.5) and the tangent at that point (black line). The slopes of the secants are 3.5, 3 and 2.5, and they approach the slope of the tangent, which is 2."
# The function, the point of interest and three step sizes.
x_values = seq(0, 3, length.out = 200)
curve_data = data.frame(x = x_values, y = x_values^2)
x0 = 1
steps = c(1.5, 1, 0.5)

# Slope and intercept of each secant through (x0, x0^2) and (x0 + h, (x0 + h)^2).
secant_slope = ((x0 + steps)^2 - x0^2) / steps
secant_data = data.frame(
  step = factor(paste("h =", steps), levels = paste("h =", steps)),
  slope = secant_slope,
  intercept = x0^2 - secant_slope * x0,
  x_end = x0 + steps,
  y_end = (x0 + steps)^2
)

# Draw the curve, the secants, the tangent (slope 2 at x0 = 1) and the points.
ggplot(curve_data, aes(x = x, y = y)) +
  geom_abline(
    data = secant_data,
    aes(slope = slope, intercept = intercept, linetype = step),
    colour = "grey40"
  ) +
  geom_abline(slope = 2 * x0, intercept = x0^2 - 2 * x0 * x0, colour = "black") +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  geom_point(data = secant_data, aes(x = x_end, y = y_end), colour = "grey40", size = 2) +
  annotate(geom = "point", x = x0, y = x0^2, size = 2.5) +
  scale_linetype_manual(values = c("dotted", "dotdash", "dashed")) +
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 9)) +
  labs(x = "x", y = "f(x)", linetype = "Secant") +
  theme_classic()
```

We can calculate the derivative of $f(x)=x^2$ directly from the definition:

$$
\begin{aligned}
f'(x) &= \lim_{h\to0}\frac{(x+h)^2-x^2}{h} \\
&= \lim_{h\to0}\frac{x^2+2xh+h^2-x^2}{h} \\
&= \lim_{h\to0}\,(2x+h) = 2x.
\end{aligned}
$$

So the slope of $x^2$ is 2 at $x=1$, as in @fig-secant, and 6 at $x=3$. In practice we do
not go back to the definition: the rules of the next section give the derivative of any
function that is built from powers, exponentials and logarithms.

:::: {.callout-important title="Practice: a derivative from its definition"}

Use the definition to calculate the derivative of $f(x)=x^2+3x$.

::: {.callout-tip title="Solution" collapse="true"}

The numerator of the secant slope is
$(x+h)^2+3(x+h)-x^2-3x=2xh+h^2+3h$. Dividing by $h$ gives $2x+h+3$, and as $h\to0$ this
approaches $f'(x)=2x+3$.

:::
::::

# Rules of differentiation {#sec-calc-rules}

**Differentiation** is the calculation of a derivative. The rules below are all that
Chapter 5 needs (see
[Derivatives, slopes and curvature](../Chapter_5/Theory.qmd#sec-function-derivatives)). In
the table, $g$ and $h$ are functions of $x$, and $c$ and $n$ are constants.

| Rule | Function | Derivative |
|---|---|---|
| Constant | $c$ | $0$ |
| Constant multiple | $c\,g(x)$ | $c\,g'(x)$ |
| Sum | $g(x)+h(x)$ | $g'(x)+h'(x)$ |
| Power | $x^n$ | $nx^{n-1}$ |
| Exponential | $e^x$ | $e^x$ |
| Logarithm | $\log x$ | $1/x$ |
| Product | $g(x)\,h(x)$ | $g'(x)\,h(x)+g(x)\,h'(x)$ |
| Quotient | $\dfrac{g(x)}{h(x)}$ | $\dfrac{g'(x)\,h(x)-g(x)\,h'(x)}{h(x)^2}$ |
| Chain | $g(h(x))$ | $g'(h(x))\,h'(x)$ |

**Constant, constant multiple, sum and power.** A constant does not change, so its slope is
zero. A constant that multiplies a function stays in front of the derivative, and a sum is
differentiated term by term. Together with the power rule, this gives the derivative of
any polynomial:

$$
\frac{d}{dx}\left(a+bx+cx^2\right)=0+b+2cx=b+2cx.
$$

The power rule holds for any exponent, including negative and fractional ones. Hence, a
power law $ax^b$ has the derivative $abx^{b-1}$, and after writing $1/x=x^{-1}$ and
$\sqrt{x}=x^{1/2}$ we obtain

$$
\frac{d}{dx}\left(\frac{1}{x}\right)=-x^{-2}=-\frac{1}{x^2}
\qquad \text{and} \qquad
\frac{d}{dx}\sqrt{x}=\tfrac{1}{2}x^{-1/2}=\frac{1}{2\sqrt{x}}.
$$

**Exponential and logarithm.** The exponential function is its own derivative, and the
derivative of the logarithm is $1/x$. For example, the derivative of $3e^x+2\log x$ is
$3e^x+2/x$.

**Chain rule.** The chain rule is for a function of a function, $g(h(x))$, where $h$ is
the inner function and $g$ the outer function. We differentiate the outer function,
leaving the inner function in place, and multiply by the derivative of the inner function.
It is the rule we will use most, because the functions of Chapter 5 contain $x$ in the
form $-bx$, $x-c$ or $b+x$. For $e^{-bx}$ the outer function is the exponential and the
inner function is $-bx$, with derivative $-b$:

$$
\frac{d}{dx}e^{-bx}=e^{-bx}\cdot(-b)=-be^{-bx}.
$$

In the same way, the derivative of $\log(a+bx)$ is $b/(a+bx)$, and the derivative of
$(b+x)^{-1}$ is $-(b+x)^{-2}$, since the inner function $b+x$ has derivative 1.

**Product rule.** The derivative of a product is the derivative of the first factor times
the second, plus the first factor times the derivative of the second. For $xe^{-bx}$ the
factors are $x$ and $e^{-bx}$, and the chain rule gave us the derivative of the second:

$$
\frac{d}{dx}\left(xe^{-bx}\right)=1\cdot e^{-bx}+x\cdot\left(-be^{-bx}\right)
=(1-bx)\,e^{-bx}.
$$

**Quotient rule.** For a ratio, the order of the terms in the numerator matters because of
the minus sign: the derivative of the numerator comes first. For $x/(b+x)$:

$$
\frac{d}{dx}\left(\frac{x}{b+x}\right)=\frac{1\cdot(b+x)-x\cdot1}{(b+x)^2}
=\frac{b}{(b+x)^2}.
$$

A quotient can also be differentiated as the product of $g(x)$ and $h(x)^{-1}$, with the
chain rule for the second factor; the result is the same.

:::: {.callout-important title="Practice: rules of differentiation"}

Calculate the derivative of each function and name the rules that you used.

1. $f(x)=3x^2+\log x$
2. $f(x)=\left(1+x^2\right)^3$
3. $f(x)=x\log x$

::: {.callout-tip title="Solution" collapse="true"}

1. Sum, constant multiple, power and logarithm: $f'(x)=6x+1/x$.
2. Chain rule, with outer function $u^3$ and inner function $u=1+x^2$:
   $f'(x)=3\left(1+x^2\right)^2\cdot2x=6x\left(1+x^2\right)^2$.
3. Product rule: $f'(x)=1\cdot\log x+x\cdot(1/x)=\log x+1$.

:::
::::

# Second derivatives {#sec-calc-second}

The **second derivative** of a function is the derivative of its derivative, written
$f''(x)$ or $d^2f/dx^2$. It is the rate at which the slope changes, and its units are the
units of $y$ divided by the square of the units of $x$. Its sign describes the curvature of
the function. Where $f''(x)<0$ the slope decreases as $x$ increases and the function bends
downwards: it is **concave**. Where $f''(x)>0$ the slope increases and the function bends
upwards: it is **convex**. A straight line has $f''(x)=0$ everywhere.

For example, the first derivative of the exponential decay $f(x)=ae^{-bx}$ (with positive
$a$ and $b$) is $f'(x)=-abe^{-bx}$, by the chain rule. Applying the same rule again gives

$$
f''(x)=-ab\cdot\left(-be^{-bx}\right)=ab^2e^{-bx}.
$$

The first derivative is negative and the second derivative is positive for every $x$. So
the function decreases everywhere and it is convex: its slope becomes less and less
negative, which is how the function levels off towards zero.

We will use second derivatives in
[Chapter 5](../Chapter_5/Theory.qmd#sec-function-special-points) to tell a maximum from a
minimum and to find the point where a curve is steepest. Chapter 4 already used the second
derivative of the log-likelihood at its maximum to calculate
[standard errors](../Chapter_4/Theory.qmd#sec-likelihood-uncertainty).

:::: {.callout-important title="Practice: curvature"}

Calculate the first and second derivatives of $f(x)=\dfrac{x}{1+x}$ for $x\geq0$. Is the
function concave or convex?

::: {.callout-tip title="Solution" collapse="true"}

By the quotient rule, $f'(x)=\dfrac{(1+x)-x}{(1+x)^2}=(1+x)^{-2}$. By the power and chain
rules, $f''(x)=-2(1+x)^{-3}$. For $x\geq0$ the first derivative is positive and the second
is negative, so the function increases and is concave: it rises more and more slowly.

:::
::::

# Partial derivatives {#sec-calc-partial}

Many functions depend on more than one quantity. The output of $f(x,y)$ depends on two
inputs, and the output of a function such as $ax^b$ depends on its parameters $a$ and $b$
as well as on $x$; to show this we write $f(x;a,b)$. The **partial derivative** of a
function with respect to one of these quantities is the derivative that we obtain when all
the others are held fixed, that is, when they are treated as constants. It is written
with the symbol $\partial$ instead of $d$, as in $\partial f/\partial a$, and it tells us
how much the output changes per unit change in that quantity alone. No new rules are
needed.

For the power law $f(x;a,b)=ax^b$, the partial derivative with respect to $x$ is the
derivative of the previous sections, $abx^{b-1}$. With respect to $a$, the function is a
constant ($x^b$) times $a$. With respect to $b$, we first write $x^b=e^{b\log x}$ and then
apply the chain rule, with $\log x$ as a constant:

$$
\frac{\partial f}{\partial a}=x^b
\qquad \text{and} \qquad
\frac{\partial f}{\partial b}=a\,e^{b\log x}\log x=ax^b\log x.
$$

For example, with $a=2$ and $b=1.5$ the function equals 16 at $x=4$. There, increasing $a$
by a small amount increases the output by $4^{1.5}=8$ times that amount, and increasing
$b$ by a small amount increases it by $16\log4\approx22.2$ times that amount.

Chapter 5 uses partial derivatives for
[functions of two inputs](../Chapter_5/Theory.qmd#sec-function-surfaces) and to study how
a curve moves when one of its parameters changes
([Derivatives with respect to parameters](../Chapter_5/Theory.qmd#sec-parameter-derivatives)).

:::: {.callout-important title="Practice: partial derivatives"}

Calculate the partial derivatives of $f(x;a,b)=ae^{-bx}$ with respect to $a$ and to $b$.

::: {.callout-tip title="Solution" collapse="true"}

With respect to $a$, the exponential is a constant factor, so
$\partial f/\partial a=e^{-bx}$. With respect to $b$, we use the chain rule with inner
function $-bx$, whose derivative with respect to $b$ is $-x$:
$\partial f/\partial b=-axe^{-bx}$.

:::
::::

# Checking with R {#sec-calc-r}

R can differentiate a formula, which is a convenient way to check a derivative calculated
on paper. The function `expression()` stores a formula without evaluating it, `D()` returns
its derivative with respect to the quantity named in the second argument, and `eval()`
calculates the value of the result for the values given in a list:

```{r}
# Derivative of x * exp(-b * x) with respect to x. D() does not simplify the result.
slope_formula = D(expression(x * exp(-b * x)), "x")
slope_formula

# Value of the derivative at x = 1 for b = 0.5, and the formula derived by hand.
eval(slope_formula, list(x = 1, b = 0.5))
(1 - 0.5 * 1) * exp(-0.5 * 1)

# Partial derivative of the power law with respect to b, at x = 4, a = 2 and b = 1.5.
eval(D(expression(a * x^b), "b"), list(x = 4, a = 2, b = 1.5))
```

A second check needs no formula for the derivative at all. By definition, the derivative
is close to the slope of a secant with a small step $h$ (a **finite difference**):

```{r}
# Slope of the secant between x = 1 and x = 1 + h, for a small step h.
f = function(x, b = 0.5) {
  x * exp(-b * x)
}
h = 1e-6
(f(1 + h) - f(1)) / h
```

[Chapter 5](../Chapter_5/Theory.qmd#sec-function-derivatives) explains both checks in more
detail.

# Practice {#sec-calc-practice}

The problems below are grouped by topic and become harder within each group. Solve them
with pen and paper before opening the solutions.

:::: {.callout-important title="Practice: exponents and logarithms (problems 1 to 4)"}

1. Simplify $\dfrac{x^3x^{-1}}{\sqrt{x}}$ and $\left(e^{2x}\right)^3e^{-x}$.
2. Write $\log\left(ax^b\right)$ as a sum, and simplify $e^{\log a-bx}$.
3. A population of bacteria grows as $n(t)=50e^{0.2t}$ cells, where $t$ is the time in
   hours. After how many hours does it reach 400 cells?
4. A power law $y=ax^b$ passes through the points $(x,y)=(2,12)$ and $(8,96)$. Calculate
   $b$ and then $a$.

::: {.callout-tip title="Solution" collapse="true"}

1. $x^{3-1-1/2}=x^{3/2}$, and $e^{6x}e^{-x}=e^{5x}$.
2. $\log\left(ax^b\right)=\log a+b\log x$, and
   $e^{\log a-bx}=e^{\log a}e^{-bx}=ae^{-bx}$.
3. $50e^{0.2t}=400$ gives $e^{0.2t}=8$, so $t=\log 8/0.2\approx10.4$ hours.
4. Dividing the second equation by the first removes $a$: $96/12=(8/2)^b$, that is,
   $8=4^b$. Taking logarithms, $b=\log 8/\log 4=1.5$. Then $a=12/2^{1.5}\approx4.24$.

:::
::::

:::: {.callout-important title="Practice: limits (problems 5 to 7)"}

5. Calculate the limits of $f(x)=\dfrac{2x^2+x}{x^2+4x}$ as $x\to0$ and as $x\to\infty$.
6. For $f(x)=\dfrac{8}{1+3e^{-0.5x}}$, calculate $f(0)$ and the limits as $x\to\infty$ and
   as $x\to-\infty$.
7. Use L'Hôpital's rule to calculate $\lim_{x\to\infty}\dfrac{\log x}{x}$ and
   $\lim_{x\to0}\dfrac{e^{2x}-1}{x}$.

::: {.callout-tip title="Solution" collapse="true"}

5. For small $x$ the lowest powers remain: dividing by $x$ gives $(2x+1)/(x+4)$, which
   approaches $1/4$. For large $x$ we divide by $x^2$ to obtain $(2+1/x)/(1+4/x)$, which
   approaches 2.
6. $f(0)=8/(1+3)=2$. As $x\to\infty$, $e^{-0.5x}\to0$ and $f(x)\to8$. As $x\to-\infty$,
   $e^{-0.5x}\to\infty$, so the denominator grows without bound and $f(x)\to0$.
7. In the first ratio both parts approach infinity, and the ratio of the derivatives is
   $(1/x)/1$, which approaches 0: a logarithm grows more slowly than $x$. In the second
   ratio both parts approach zero, and the ratio of the derivatives is $2e^{2x}/1$, which
   approaches 2.

:::
::::

:::: {.callout-important title="Practice: one rule at a time (problems 8 to 10)"}

Calculate the derivative of each function.

8. $f(x)=4x^3-2x+7$
9. $f(x)=3\sqrt{x}+\dfrac{2}{x^2}$
10. $f(x)=5e^x-2\log x$

::: {.callout-tip title="Solution" collapse="true"}

8. Term by term with the power rule: $f'(x)=12x^2-2$.
9. Written with exponents, $f(x)=3x^{1/2}+2x^{-2}$, so
   $f'(x)=\tfrac{3}{2}x^{-1/2}-4x^{-3}=\dfrac{3}{2\sqrt{x}}-\dfrac{4}{x^3}$.
10. $f'(x)=5e^x-\dfrac{2}{x}$.

:::
::::

:::: {.callout-important title="Practice: chain, product and quotient rules (problems 11 to 13)"}

Calculate the derivative of each function.

11. $f(x)=e^{-x^2/2}$ and $f(x)=\log(1+3x)$
12. $f(x)=x^2\log x$
13. $f(x)=\dfrac{1+x}{1+2x}$

::: {.callout-tip title="Solution" collapse="true"}

11. Chain rule. For the first function the inner function is $-x^2/2$, with derivative
    $-x$, so $f'(x)=-xe^{-x^2/2}$. For the second the inner function is $1+3x$, with
    derivative 3, so $f'(x)=\dfrac{3}{1+3x}$.
12. Product rule: $f'(x)=2x\log x+x^2\cdot\dfrac{1}{x}=2x\log x+x$.
13. Quotient rule:
    $f'(x)=\dfrac{1\cdot(1+2x)-(1+x)\cdot2}{(1+2x)^2}=-\dfrac{1}{(1+2x)^2}$.

:::
::::

:::: {.callout-important title="Practice: rules combined (problems 14 to 16)"}

14. Calculate the derivative of $f(x)=\sqrt{x}\,e^{-x}$ for $x>0$. At which value of $x$ is
    the slope zero?
15. Calculate the derivative of $f(x)=\log\left(1+e^{x}\right)$ and show that it can be
    written as $1/\left(1+e^{-x}\right)$.
16. Calculate the derivative of $f(x)=\dfrac{e^{-x}}{1+x}$.

::: {.callout-tip title="Solution" collapse="true"}

14. Product rule, with the power rule for the first factor and the chain rule for the
    second:
    $$
    f'(x)=\frac{1}{2\sqrt{x}}e^{-x}-\sqrt{x}\,e^{-x}=\frac{(1-2x)\,e^{-x}}{2\sqrt{x}}.
    $$
    The exponential and the square root are positive, so the slope is zero where
    $1-2x=0$, at $x=1/2$.
15. Chain rule, with outer function $\log u$ and inner function $u=1+e^x$:
    $f'(x)=\dfrac{e^x}{1+e^x}$. Dividing numerator and denominator by $e^x$ gives
    $\dfrac{1}{e^{-x}+1}$.
16. Quotient rule, with the chain rule for the numerator:
    $$
    f'(x)=\frac{-e^{-x}(1+x)-e^{-x}\cdot1}{(1+x)^2}=-\frac{(2+x)\,e^{-x}}{(1+x)^2}.
    $$

:::
::::

:::: {.callout-important title="Practice: second derivatives (problems 17 and 18)"}

17. Calculate the second derivative of $f(x)=x^3-6x^2$. For which values of $x$ is the
    function concave, and for which is it convex?
18. Do the same for $f(x)=\log\left(1+x^2\right)$ with $x\geq0$.

::: {.callout-tip title="Solution" collapse="true"}

17. $f'(x)=3x^2-12x$ and $f''(x)=6x-12$. The second derivative is negative for $x<2$,
    where the function is concave, and positive for $x>2$, where it is convex.
18. By the chain rule, $f'(x)=\dfrac{2x}{1+x^2}$. By the quotient rule,
    $$
    f''(x)=\frac{2\left(1+x^2\right)-2x\cdot2x}{\left(1+x^2\right)^2}
    =\frac{2\left(1-x^2\right)}{\left(1+x^2\right)^2}.
    $$
    The denominator is positive, so the sign is that of $1-x^2$: the function is convex
    for $x<1$ and concave for $x>1$.

:::
::::

:::: {.callout-important title="Practice: partial derivatives (problems 19 and 20)"}

19. Calculate the partial derivatives of $f(x,y)=x^2y+3y^2$ with respect to $x$ and to
    $y$.
20. Calculate the partial derivatives of $f(x;a,b)=\dfrac{a}{b+x}$ with respect to $a$ and
    to $b$, and their values at $x=3$ for $a=10$ and $b=2$.

::: {.callout-tip title="Solution" collapse="true"}

19. With $y$ fixed, $\partial f/\partial x=2xy$. With $x$ fixed,
    $\partial f/\partial y=x^2+6y$.
20. With respect to $a$, the function is $a$ times a constant:
    $\partial f/\partial a=\dfrac{1}{b+x}$. With respect to $b$, we write the function as
    $a(b+x)^{-1}$ and use the power and chain rules:
    $\partial f/\partial b=-\dfrac{a}{(b+x)^2}$. At $x=3$ with $a=10$ and $b=2$ the values
    are $1/5=0.2$ and $-10/25=-0.4$.

:::
::::
