---
title: "Formula-based methods"
---

# Introduction

In this supplement, we learn how to fit linear and generalised linear models, including
their mixed-effects versions, with `glmmTMB`. These models are useful when a linear
predictor, combined with a suitable observation distribution and link function, can
describe the relationship we want to analyse. The package provides a formula interface
to TMB, so we can use the likelihood methods of the course without writing each
negative log-likelihood ourselves. Of course, you can write all of these models directly
in RTMB, but when a model can be fitted with `glmmTMB` it is a lot faster to implement, and
it may communicate better with others who are used to linear and generalised linear models.

We assume that you are familiar with maximum likelihood and profile likelihood intervals
from [Chapter 4](../Chapter_4/Theory.qmd), prediction and the delta method from
[Chapter 6](../Chapter_6/Theory.qmd), and random effects and marginal likelihood from
[Chapter 7](../Chapter_7/Theory.qmd). The statistical reasoning stays the same. This
supplement explains how to express those models and request those calculations through
`glmmTMB`, using penguin measurements, repeated measurements of chicks, and insect and
salamander counts.

# Learning goals

In this supplement, we will cover:

- linear predictors, formulae, and link functions;
- fixed effects, random effects, and grouping structures;
- linear and generalised linear models with `glmmTMB`;
- their mixed-effects versions and marginal likelihood;
- formulae for dispersion and zero-inflation;
- standard errors, profile likelihood intervals, and prediction uncertainty;
- model comparison and simulation-based checks.

# The formula interface

A **model formula** describes how the predictors enter a model. The variables in the
formula are columns of a data frame; their coefficients are parameters to be estimated.
The observation distribution and its link function are specified separately, through
the argument `family`.

## Linear predictors and link functions

A **linear predictor** is a sum of predictor values multiplied by coefficients. For one
continuous predictor $x_i$, it is

$$
\eta_i = a + b x_i,
$$

where $a$ is the intercept and $b$ is the slope. The formula `y ~ x` requests these two
coefficients. For a Normal linear model, the expected response is $\mu_i=\eta_i$. More
generally, a **link function** $g$ relates the mean to the linear predictor through
$g(\mu_i)=\eta_i$, so the inverse link gives the expected response. For example, a log
link gives $\mu_i=\exp(a+b x_i)$, while a logit link gives a probability
$p_i=\operatorname{logit}^{-1}(a+b x_i)$.

The word "linear" refers to the coefficients. A polynomial such as
$\eta_i=a+b x_i+c x_i^2$ is also linear in its coefficients, even though its graph is
curved. R's formula operators specify which terms receive coefficients:

| Formula | Terms in the linear predictor |
|---|---|
| `y ~ x` | Intercept and slope for `x` |
| `y ~ treatment` | Intercept and differences between treatment levels |
| `y ~ x + treatment` | Common slope and differences between treatment levels |
| `y ~ x * treatment` | `x + treatment + x:treatment`: intercepts and slopes can differ between treatments |
| `y ~ x + I(x^2)` | Intercept, slope, and a coefficient for the square of `x` |
| `y ~ 0 + treatment` | One coefficient for each treatment level |

With the default contrasts, a factor's first level is the reference level. Its expected
response on the link scale is included in the intercept, and the other coefficients are
differences from that level. Use `factor()` for a categorical predictor and `relevel()`
to choose its reference. The function `I()` tells R to interpret arithmetic such as
`x^2` literally inside a formula.

## Random effects

A **random effect** is a group deviation described by a distribution, as in Chapter 7.
In `glmmTMB`, `(1 | site)` adds a Normal random intercept for each site. With a continuous
predictor, the linear predictor becomes $\eta_{ij}=a+b x_{ij}+u_j$, where $u_j$ is the
deviation of site $j$ and $u_j\sim\operatorname{Normal}(0,\sigma^2_{\mathrm{site}})$.
The shared intercept, slope, and among-site variance are parameters of the model.

The formula `(1 + x | site)` also lets the slope vary among sites, and estimates the
covariance between the intercept and slope deviations. Grouping factors can be crossed
or nested. For example, `(1 | site) + (1 | year)` gives a site deviation and a year
deviation shared across sites. By comparison, `(1 | site/year)` expands to
`(1 | site) + (1 | site:year)`: the second deviation belongs to a particular year within
a particular site.

# From TMB to `glmmTMB` {#sec-glmmtmb}

The package **`glmmTMB`** constructs and fits a TMB likelihood from the formula, data,
and chosen family. You already know the calculations that follow: automatic
differentiation supplies derivatives for optimisation, the Hessian gives approximate
standard errors, and likelihood profiles refit the other parameters while fixing the
parameter of interest. For mixed models, TMB marginalises over the Normal random effects
with the Laplace approximation of Chapter 7. This approximation is exact for the Normal
linear mixed models considered here.

The difference is how we specify the model and read its results. In the course we write
an NLL and process it with `MakeADFun()`; here we call `glmmTMB()` with the model
description. The following methods request the corresponding calculations:

| Calculation from the course | Method for a `glmmTMB` fit |
|---|---|
| Estimates and standard errors | `summary()`, `fixef()`, and `vcov()` |
| Variance components and group deviations | `VarCorr()` and `ranef()` |
| Wald intervals | `confint(fit, method = "wald")` |
| Profile likelihood intervals | `confint(fit, method = "profile")` |
| Predictions and delta-method standard errors | `predict(fit, se.fit = TRUE)` |
| Hypothetical experiments | `simulate(fit)` |
| Likelihood and information criteria | `logLik()`, `AIC()`, and `BIC()` |

We use maximum likelihood throughout, which is the default (`REML = FALSE`).^[For Normal
linear mixed models, `REML = TRUE` requests restricted maximum likelihood, an alternative
for estimating variance components. Keep maximum likelihood when comparing models with
different fixed effects. See the [function documentation](https://glmmtmb.github.io/glmmTMB/reference/glmmTMB.html)
for the available arguments and options.]
Load the package once, before fitting the examples:

```{r}
#| message: false
# Load glmmTMB for the model fits of this supplement.
library(glmmTMB)
```

# Linear and generalised linear models

The same function fits all four model classes. Its main arguments are:

- `formula`: the response and the fixed and random effects;
- `data`: the data frame containing those variables;
- `family`: the observation distribution and link function.

## Linear models

A **linear model** describes independent Normal observations whose mean is a linear
predictor. For a continuous predictor $x_i$ and a two-level factor encoded as
$z_i\in\{0,1\}$, we can write

$$
Y_i\sim\operatorname{Normal}(\mu_i,\sigma^2),\qquad
\mu_i=a+b x_i+c z_i,
$$

where $a$ is the intercept for the reference level, $b$ is the common slope, $c$ is the
difference between levels at the same value of $x_i$, and $\sigma^2$ is the residual
variance. The family `gaussian(link = "identity")` specifies this observation model.

We return to the Gentoo penguins of Chapters 2 and 3, now asking how expected body mass
varies with flipper length and sex. The built-in `penguins` data require R 4.5.0 or later.
We remove birds missing any of these three measurements and choose females as the
reference level. Subtracting 217 mm from flipper length puts the intercept at a length
within the observed range:

```{r}
# Select the Gentoo penguins.
data(penguins, package = "datasets")
gentoo = penguins[penguins$species == "Gentoo", ]
# Count incomplete records before excluding them.
keep = complete.cases(gentoo[, c("body_mass", "flipper_len", "sex")])
sum(!keep)
gentoo = gentoo[keep, ]
# Make females the reference level and centre flipper length at 217 mm.
gentoo$sex = relevel(factor(gentoo$sex), ref = "female")
gentoo$flipper_mm = gentoo$flipper_len - 217
```

The formula requests a common slope and a difference between sexes. We fit the model
and read its estimates and standard errors from the conditional coefficient table:

```{r}
# Normal linear model: a common slope and a difference between sexes.
penguin_fit = glmmTMB(
  body_mass ~ flipper_mm + sex,
  data = gentoo,
  family = gaussian(link = "identity")
)
# Estimates and standard errors of the coefficients.
summary(penguin_fit)$coefficients$cond[, c("Estimate", "Std. Error")]
```

The intercept is the estimated mean body mass (g) of a female with a flipper length of
217 mm. The slope gives the change in expected mass (g/mm), holding sex constant, and
`sexmale` gives the male–female difference (g) at the same flipper length. The model
treats birds as independent conditional on these predictors. Adding an interaction,
`flipper_mm * sex`, would allow the slopes to differ between sexes.

## Linear mixed models

A **linear mixed model** adds Normal group deviations to a linear predictor with Normal
observations. For a random intercept, the observation model is

$$
Y_{ij}\mid u_j\sim\operatorname{Normal}(a+b x_{ij}+u_j,\sigma^2_{\mathrm{obs}}),
\qquad u_j\sim\operatorname{Normal}(0,\sigma^2_{\mathrm{group}}),
$$

where $i$ indexes observations within group $j$. Observations are assumed independent
conditional on their group deviations. Marginalising over a shared deviation makes
observations from the same group correlated.

The built-in `ChickWeight` data record repeated body masses of chicks receiving four
diets. We use measurements from days 10 to 20, over which a straight line is a candidate
description of growth. Each chick has its own intercept; diet changes the shared
intercept, with diet 1 (the first level of `Diet`) as the reference level. We express mass
in kg to keep the numerical scale small and centre age at day 10:

```{r}
# Keep the measurements from days 10 to 20.
data(ChickWeight, package = "datasets")
chicks = subset(ChickWeight, Time >= 10 & Time <= 20)
# Mass in kg and age in days since day 10.
chicks$mass_kg = chicks$weight / 1000
chicks$age_days = chicks$Time - 10
# Drop the chick that has no measurements in this period.
chicks$Chick = factor(chicks$Chick)
```

The term `(1 | Chick)` requests the random intercepts. `VarCorr()` reports the variance
components and `ranef()` returns the predicted group deviations (conditional modes):

```{r}
# Normal linear mixed model with a random intercept for each chick.
chick_fit = glmmTMB(
  mass_kg ~ age_days + Diet + (1 | Chick),
  data = chicks,
  family = gaussian(link = "identity")
)
summary(chick_fit)$coefficients$cond[, c("Estimate", "Std. Error")]
# Standard deviations among chicks and around each chick's line.
VarCorr(chick_fit)
# Predicted deviations of the first chicks.
head(ranef(chick_fit)$cond$Chick)
```

The slope is a shared growth rate (kg/day), and each diet coefficient is a difference
from diet 1 at day 10. The among-chick variance measures variation in intercepts (kg²);
the residual variance measures variation around each chick's line (kg²). Squaring a
reported standard deviation gives its variance. A random slope, `(1 + age_days | Chick)`,
would also allow growth rates to vary among chicks.

## Generalised linear models

A **generalised linear model** combines an observation distribution with a mean specified
through a link and a linear predictor. The observations are assumed independent given
the predictors. For counts under a Poisson model with a log link,

$$
Y_i\sim\operatorname{Poisson}(\mu_i),\qquad \log\mu_i=\eta_i.
$$

We return to the insect counts in the Chapter 2 practical, now estimating one mean for
each spray treatment. Spray A is the first level of `spray`, so it is the reference level:

```{r}
data(InsectSprays, package = "datasets")
# Poisson model with one mean for each spray.
insect_poisson_fit = glmmTMB(
  count ~ spray,
  data = InsectSprays,
  family = poisson(link = "log")
)
# Coefficients on the log scale.
fixef(insect_poisson_fit)$cond
```

The intercept is the log of the expected count for spray A. Each other coefficient is
a log ratio to that mean, so exponentiating it gives the corresponding ratio of expected
counts. For example, a coefficient of $\log 2$ would mean twice the expected count.

The family also specifies the relationship between the mean and variance. The Poisson
model has variance $\mu_i$; `nbinom2()` gives the Negative Binomial model of the course,
with variance $\mu_i+\mu_i^2/k$. It estimates the positive shape $k$, called the
dispersion parameter in `glmmTMB`. We can fit that alternative by changing the family.
For this family, `sigma()` returns the estimate of $k$:

```{r}
# Negative Binomial model with the same means.
insect_nb_fit = glmmTMB(
  count ~ spray,
  data = InsectSprays,
  family = nbinom2(link = "log")
)
# Estimated shape k.
sigma(insect_nb_fit)
```

Other observation models use the same interface. For Binomial counts, specify
`cbind(survived, total - survived) ~ treatment` with `family = binomial(link = "logit")`;
the two response columns contain successes and failures. `Gamma(link = "log")` and
`beta_family(link = "logit")` cover positive continuous measurements and fractions
strictly between zero and one. See the [family reference](https://glmmtmb.github.io/glmmTMB/reference/nbinom2.html)
for parameterisations, and the [distribution supplement](distributions.qmd) for the
probability models.

## Generalised linear mixed models

A **generalised linear mixed model** adds random effects to the linear predictor of a
generalised linear model. For a Negative Binomial random-intercept model,
$\log\mu_{ij}=\eta_{ij}+u_j$, with
$u_j\sim\operatorname{Normal}(0,\sigma^2_{\mathrm{site}})$. The observation variance is
$\mu_{ij}+\mu_{ij}^2/k$ conditional on the site effect; the Normal site distribution
describes an additional source of variation among sites.

The `Salamanders` data of Chapter 2 contain counts from four visits to each of 23 stream
sites. Here we use all species and life-stage categories in `spp`, with their different
mean counts, and ask how counts differ between mined and unmined sites. The counts are
those of Price et al. (2016).^[See the
[data documentation](https://glmmtmb.github.io/glmmTMB/reference/Salamanders.html) for the
study and the [data in Dryad](https://doi.org/10.5061/dryad.5m8f6).]
The shared site deviation represents dependence between counts from the same site. The
levels of `mined` are `yes` and `no`, in that order, so the mined sites are the reference
level:

```{r}
data(Salamanders, package = "glmmTMB")
# Negative Binomial mixed model with a random intercept for each site.
salamander_fit = glmmTMB(
  count ~ spp + mined + (1 | site),
  data = Salamanders,
  family = nbinom2(link = "log")
)
summary(salamander_fit)$coefficients$cond[, c("Estimate", "Std. Error")]
VarCorr(salamander_fit)
```

The coefficient `minedno` compares unmined with mined sites on the log scale, holding
species and site deviation constant. Its exponential is a ratio of expected counts.
These are observational data: that association alone does not identify the causal
effect of mining.

# Dispersion and zero-inflation

The argument **`dispformula`** models the dispersion parameter with its own linear
predictor and a log link. For a Normal model, it describes the residual standard
deviation: $\log\sigma_i=a_\sigma+b_\sigma x_i$. For `nbinom2()`, it describes the
shape: $\log k_i=a_k+b_k x_i$. Its default, `~ 1`, estimates one shared value. For example,
the penguin model can allow a different residual standard deviation for each sex:

```{r}
# The same linear model, with a residual standard deviation for each sex.
penguin_disp_fit = glmmTMB(
  body_mass ~ flipper_mm + sex,
  dispformula = ~ sex,
  data = gentoo,
  family = gaussian(link = "identity")
)
# Coefficients of the dispersion model, on the log scale.
fixef(penguin_disp_fit)$disp
```

The intercept is the log of the residual standard deviation (g) of females, and `sexmale`
is the difference between males and females on that scale, so its exponential is the ratio
of the two standard deviations.

The argument **`ziformula`** models a zero-inflation probability with a logit link,
as described in the [distribution supplement](distributions.qmd#sec-zero_inflated).
Its default, `~ 0`, adds no zero-inflation. We can allow that probability to differ
between mined and unmined sites while retaining the same conditional count model:

```{r}
# The same count model, with a zero-inflation probability for each type of site.
salamander_zi_fit = glmmTMB(
  count ~ spp + mined + (1 | site),
  ziformula = ~ mined,
  data = Salamanders,
  family = nbinom2(link = "log")
)
# Coefficients of the zero-inflation model, on the logit scale.
fixef(salamander_zi_fit)$zi
```

Thus the conditional mean, dispersion, and zero-inflation probability have separate
formulae, reported in `cond`, `disp`, and `zi`. Choose these components from the sampling
process and check them against the data; a large number of zeros may already be explained
by a Negative Binomial observation model.

# Uncertainty and prediction

## Standard errors and profile likelihood intervals

The uncertainty calculations are those of Chapter 4. `summary()` reports approximate
standard errors, and `vcov(fit)$cond` returns the covariance matrix of the conditional
fixed-effect estimates. For intervals, `confint()` defaults to the Wald approximation;
we request the reference method explicitly with `method = "profile"`. The argument
`parm = "beta_"` selects the fixed effects. For the penguins:

```{r}
# Wald intervals, then profile likelihood intervals, for the coefficients.
confint(penguin_fit, parm = "beta_", method = "wald")
confint(penguin_fit, parm = "beta_", method = "profile")
```

For mixed models, the profiles use the marginal likelihood. For example, we can profile
the among-site standard deviation with `parm = "theta_"`. This example has only a random
intercept, so the profiled parameter is its log standard deviation. We transform the
endpoints back, then square them to express the interval for the variance:

```{r}
# Profile likelihood interval of the log standard deviation among sites.
site_log_sd_interval = confint(
  salamander_fit, parm = "theta_", method = "profile"
)
# Transform the endpoints to the variance.
exp(site_log_sd_interval)^2
```

The scales returned for variance parameters depend on the interval method; see the
[interval documentation](https://glmmtmb.github.io/glmmTMB/reference/confint.glmmTMB.html).
As in Chapter 4, transforming interval endpoints is different from calculating the
standard error of a transformed quantity.

## Predictions and the delta method

The function **`predict()`** calculates the fitted response at predictor values in
`newdata`. With `se.fit = TRUE`, it also gives approximate standard errors, using TMB's
delta-method calculations. This is the same calculation as requesting a derived quantity
with `ADREPORT()` and reading its standard error from `sdreport()`, as we do from Chapter 4
onward; [Chapter 6](../Chapter_6/Theory.qmd#sec-delta-method) explains the delta method.

For the penguins, we request mean mass at a flipper length of 217 mm for each sex.
The result contains `fit` and `se.fit`; we use them to calculate approximate 95% Wald
intervals for the two expected masses:

```{r}
# One female and one male with a flipper length of 217 mm.
new_birds = data.frame(
  flipper_mm = c(0, 0),
  sex = factor(c("female", "male"), levels = levels(gentoo$sex))
)
mass_prediction = predict(
  penguin_fit, newdata = new_birds, type = "response", se.fit = TRUE
)
# Expected masses with their standard errors and Wald intervals.
data.frame(
  sex = new_birds$sex,
  mean_g = mass_prediction$fit,
  se_g = mass_prediction$se.fit,
  lower_g = mass_prediction$fit - 1.96 * mass_prediction$se.fit,
  upper_g = mass_prediction$fit + 1.96 * mass_prediction$se.fit
)
```

These intervals describe uncertainty in the expected mass. Prediction intervals for
individual birds must also include variation around that mean. For a log or logit link,
calculate the Wald interval with `type = "link"` and transform its endpoints with
`exp()` or `plogis()`. A response-scale standard error can be requested directly with
`type = "response", se.fit = TRUE`.

For a mixed model, `re.form = NULL` includes the predicted deviations of the observed
groups. Setting `re.form = NA` gives predictions with all random effects set to zero.
We can compare these two expected counts for the recorded salamander samples:

```{r}
# Expected counts with and without the predicted site deviations.
head(data.frame(
  site = Salamanders$site,
  fitted_site = predict(salamander_fit, type = "response", re.form = NULL),
  zero_site_effect = predict(salamander_fit, type = "response", re.form = NA)
))
```

With a log link, setting the random effect to zero gives $\exp(\eta_i)$, whereas averaging
the mean over the Normal site distribution gives
$\exp(\eta_i+\sigma^2_{\mathrm{site}}/2)$. So `re.form = NA` does not perform that
response-scale averaging. Marginalisation during fitting and averaging a prediction
over groups are separate calculations. For zero-inflated models, `type = "response"`
also includes the zero-inflation probability; `type = "conditional"` returns the mean
of the count component.^[The [prediction reference](https://glmmtmb.github.io/glmmTMB/reference/predict.glmmTMB.html)
describes these options and prediction for new groups.]

# Model comparison and checks

Information criteria use the fitted likelihood, as in Chapters 4 and 6. For example,
the two insect models use exactly the same observations and differ in their assumption
about the conditional variance. We can compare them with AIC and the small-sample
correction AICc. The `df` column returned by `AIC()` counts the estimated parameters,
including dispersion parameters:

```{r}
insect_comparison = AIC(insect_poisson_fit, insect_nb_fit)
# Number of parameters and of observations for the correction.
K = insect_comparison$df
n = nrow(InsectSprays)
insect_comparison$AICc = insect_comparison$AIC + 2 * K * (K + 1) / (n - K - 1)
insect_comparison
```

Lower AIC ranks a candidate more favourably; the differences compare its expected
predictive performance with that of the other candidates. AICc is the approximation
discussed in [Chapter 4](../Chapter_4/Theory.qmd#sec-aic). Model comparison is followed
by checks of convergence and of the observation model.

A `glmmTMB` fit stores the result of the optimiser in its element `fit` and the result of
`sdreport()` in its element `sdr`. As in Chapter 4, `convergence` should be zero after a
normal stop, and `pdHess` should be `TRUE` for the Hessian to give usable standard errors.
We check both for the salamander fit:

```{r}
# Convergence code of the optimiser and check of the Hessian.
salamander_fit$fit$convergence
salamander_fit$sdr$pdHess
```

The function **`simulate()`** draws hypothetical observations from a fitted model.
For a mixed model it draws new random effects as well as new observations. The result
is a data frame with one column per experiment. For example, we can compare the observed
fraction of zeros with that in three hypothetical experiments:

```{r}
# Simulate three hypothetical experiments from the fitted model.
set.seed(1234)
simulated_counts = simulate(salamander_fit, nsim = 3)
# Fraction of zeros in the data and in each simulated experiment.
mean(Salamanders$count == 0)
for (i in 1:3) {
  print(mean(simulated_counts[[i]] == 0))
}
```

For a fuller assessment, `glmmTMB` fits work directly with the simulation-based
residual tools of `DHARMa` used in
[Chapter 6](../Chapter_6/Theory.qmd#sec-response-residuals):

```{r}
#| eval: false
# Generate scaled residuals, then check their distribution and predictor patterns.
salamander_residuals = DHARMa::simulateResiduals(salamander_fit, n = 1000, seed = 1234)
plot(salamander_residuals)
DHARMa::plotResiduals(salamander_residuals, form = Salamanders$mined)
```

As with the models that we write ourselves, interpret the estimated effects and their
uncertainty in terms of the ecological question, and use the checks to decide whether the
observation distribution and grouping structure describe the data adequately.
