---
title: "Distributions"
toc: true
editor:
  render-on-save: true
format:
  html:
    include-in-header:
      text: |
        <style>
          mjx-container[display="true"] {
            text-align: left !important;
          }
        </style>
html-math-method:
  method: mathjax
  url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml-full.js
---

In this document I provide reference material for several distributions, including their
mathematical formulae, examples of code and tips on when they might be relevant. There is
also a section on how to construct zero-inflated distributions.

For each distribution I provide the default parameterization in R, but I also provide a
*inverse* parameterization where we use the mean ($\mu$) as input, as we normally model the
effect of covariates on the mean.

# Discrete distributions

Six distributions for count data that cover the most common scenarios in ecology. Some
guidance on how to choose (but also remember model selection as a general tool):

**Bernoulli** - Binary outcomes such as presence/absence or success/failure.

**Binomial** - Fixed number of binary outcomes.

**Poisson** - Counts of events in a fixed area, time, or sampling effort. Variance equal to mean.

**Conway-Maxwell-Poisson** - Count data with either underdispersion or overdispersion relative
to the Poisson distribution.

**Negative binomial** - Overdispersed count data (variance > mean).

**Beta binomial** -  Overdispersed count data (variance > mean). Also for repeated Binomial process
where the probaility of success varies randomly across trials.

## Bernoulli distribution

Base R uses the Binomial functions with `size = 1` to represent a Bernoulli distribution.

$$
\begin{aligned}
\text{R}: &\quad \texttt{dbinom, pbinom, qbinom, rbinom} \quad (\texttt{size = 1}) \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x \in \{0,1\} \\
\text{Probability}: &\quad p^x(1-p)^{1-x} \\
\text{Parameters}: &\quad p\;(\texttt{prob}) = \text{probability of success} \\
\text{Mean}: &\quad E[x] = p \\
\text{Variance}: &\quad Var(X) = p(1-p) \\
\text{Inverse}: &\quad p = \mu
\end{aligned}
$$

The Bernoulli distribution is the model for a single binary outcome: a site is occupied or
not, a seed germinates or not, an animal survives the winter or not. We code the outcome of
interest as $1$ (often called a *success*) and the alternative as $0$. The distribution has
one parameter, $p$, which is the probability of the outcome of interest, with
$0 \leq p \leq 1$.

Its probability mass function is

$$
P(X=x)=p^x(1-p)^{1-x}, \qquad x \in \{0, 1\},
$$

that is, $P(X=1)=p$ and $P(X=0)=1-p$. The mean and variance are:

$$
\mu=E[X]=p
$$

and

$$
\operatorname{Var}(X)=p(1-p).
$$

Note that the variance is fully determined by the mean. It is largest when $p=0.5$ and it
becomes zero when $p$ is either $0$ or $1$, as there is no variation left in the outcome.

The Bernoulli distribution is a Binomial distribution with a single trial ($n=1$), which is
why R does not have separate functions for it. In the other direction, the sum of $n$
independent Bernoulli random variables with the same $p$ follows a Binomial distribution.
The Bernoulli distribution is also one of the two components of a zero-inflated distribution
(see @sec-zero_inflated).

### Example: pond occupancy

Suppose the probability that a pond is occupied by a newt species is $0.3$. Let $X=1$ if the
pond is occupied and $X=0$ otherwise. The probability of an occupied pond is

$$
P(X=1) = (0.3)^1(1-0.3)^{1-1} = 0.3,
$$

the mean is $0.3$ and the variance is $0.3(0.7)=0.21$. In R:

```{r}
# A Bernoulli distribution is a Binomial distribution with size = 1.
occupancy_probability = 0.3

dbinom(x = 1, size = 1, prob = occupancy_probability)

# The variance follows from the probability of occupancy.
occupancy_probability * (1 - occupancy_probability)
```

A random sample from this model is a sequence of zeros and ones. For example, we can simulate
the occupancy of ten ponds:

```{r}
# Simulate whether each of ten ponds is occupied (1) or not (0).
set.seed(2026)
rbinom(n = 10, size = 1, prob = occupancy_probability)
```

```{r}
#| echo: false
#| fig-cap: "Bernoulli probability mass for the occupancy of a pond."
# Plot the probability of each of the two possible outcomes.
occupancy_values = 0:1
plot(
  occupancy_values,
  dbinom(x = occupancy_values, size = 1, prob = 0.3),
  type = "h",
  xaxt = "n",
  ylim = c(0, 1),
  xlab = "Pond occupied (0 = no, 1 = yes)",
  ylab = "Probability"
)
axis(side = 1, at = occupancy_values)
```

:::: {.callout-important title="Practice: nest survival"}

Suppose a nest survives until fledging with probability $0.85$. Calculate the mean and the
variance of this Bernoulli random variable. For which value of $p$ would the variance be
largest?

::: {.callout-tip title="Solution" collapse="true"}

The mean is $p=0.85$ and the variance is $0.85(1-0.85)=0.1275$. The variance $p(1-p)$ is
largest at $p=0.5$, where it equals $0.25$. That is the case in which the two outcomes are
equally likely, so the fate of a nest is hardest to predict.

:::
::::

## Binomial distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dbinom, pbinom, qbinom, rbinom} \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x = 0, 1, \ldots, n \\
\text{Probability}: &\quad \binom{n}{x}p^x(1-p)^{n-x} \\
\text{Parameters}: &\quad n\;(\texttt{size}) = \text{number of trials};\quad
p\;(\texttt{prob}) = \text{success probability} \\
\text{Mean}: &\quad E[x] = np \\
\text{Variance}: &\quad Var(X) = np(1-p) \\
\text{Inverse}: &\quad p = \mu/n \quad (\text{with } n \text{ fixed})
\end{aligned}
$$

The Binomial distribution is useful when every sampling unit contains a fixed number of binary
events or outcomes. It models the number of times a particular outcome occurs (e.g., survival,
germination, success, etc.). Each event is assumed to be independent of the others and has the same
probability of obtaining the outcome of interest.

As seen before, the probability mass function of the Binomial distribution is:

$$
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x}
=\frac{n!}{x!(n-x)!}p^x(1-p)^{n-x},
$$

where $\binom{n}{x}$ counts the ways to choose $x$ outcomes of interest (each with probability
$p$ of occurring) from $n$ events and $n!$ is a factorial.

Applying the relevant formulae, we can derive the mean and variance of a binomial distribution as:

$$
\mu=E[X]=np
$$

and

$$
\operatorname{Var}(X)=np(1-p).
$$


### Example: seedling survival

Suppose 20 seedlings are planted in each plot and each survives independently with probability
$0.7$. Let $X$ be the number surviving in one plot. The probability that exactly 14 survive
is:

$$
P(X=14)
=
\binom{20}{14}(0.7)^{14}(1-0.7)^{20-14}
=
\frac{20!}{14! (20 - 14)!}(0.7)^{14}(0.3)^6
\approx 0.1916.
$$

We can verify our math with R:

```{r}
# Set the binomial model parameters and the count we want to check.
seedling_count = 20
seedling_survival_probability = 0.7
survivors_to_check = 14

dbinom(
  x = survivors_to_check,
  size = seedling_count,
  prob = seedling_survival_probability
)
```

The model gives an expected count of $20(0.7)=14$ seedlings per plot and
a variance of $20(0.7)(0.3)=4.2$ seedlings squared:

```{r}
# Calculate the expected count and variance from the binomial formulas.
expected_survivors = seedling_count * seedling_survival_probability
variance_survivors = seedling_count *
  seedling_survival_probability *
  (1 - seedling_survival_probability)
```

If a binomial mean and variance are known, we can invert their formulas.
Writing $m=E[X]$ and $v=\operatorname{Var}(X)$ gives

$$
\begin{aligned}
p &= 1-\frac{v}{m}, \\
n &= \frac{m}{p}.
\end{aligned}
$$

This conversion requires $m>0$, a probability $0<p\leq 1$, and a positive
integer result for $n$. For the model above, it recovers $p=0.7$ and $n=20$:

```{r}
# Recover the binomial parameters from the model mean and variance.
mean_survivors = expected_survivors
success_probability = 1 - variance_survivors / mean_survivors
estimated_trial_count = mean_survivors / success_probability
```

:::: {.callout-important title="Practice: hatching success in a clutch"}

A clutch contains 12 eggs and each egg hatches independently with probability $0.6$.
Calculate the expected number of hatched eggs and its variance. What is the probability
that at least 10 eggs hatch?

::: {.callout-tip title="Solution" collapse="true"}

The expected number of hatched eggs is $12(0.6)=7.2$ and the variance is
$12(0.6)(0.4)=2.88$. The probability of at least 10 hatched eggs is the complement of
the cumulative probability for nine:

```{r}
# Calculate the probability of at least ten hatched eggs out of twelve.
1 - pbinom(q = 9, size = 12, prob = 0.6)
```

The probability is approximately $0.083$.

:::
::::

## Poisson distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dpois, ppois, qpois, rpois} \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x = 0, 1, 2, \ldots \\
\text{Probability}: &\quad \frac{e^{-\lambda}\lambda^x}{x!} \\
\text{Parameters}: &\quad \lambda\;(\texttt{lambda}) = \text{event rate} \\
\text{Mean}: &\quad E[x] = \lambda \\
\text{Variance}: &\quad Var(X) = \lambda \\
\text{Inverse}: &\quad \lambda = \mu
\end{aligned}
$$

The Poisson distribution is a common starting point when we count events in a fixed unit of time,
area, or sampling effort: insects caught per trap-night, seedlings per quadrat, or camera detections
per night. It is often the default distribution for count data when we do not have a fixed
upper limit to the number of observations. Its most important property is to always have the
same variance and mean, such that when this is not appropriate, we need to switch to other
distributions for count data.
The Poisson distribution has one parameter:

$$
\lambda=\text{expected count per sampling unit}, \qquad \lambda\geq 0.
$$

Its probability mass function is

$$
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!},
$$

The mean and variance are both equal to $\lambda$:

$$
\mu = E[X]=\lambda
$$

and

$$
\operatorname{Var}(X)=\lambda.
$$


### Example: wildlife detections

Suppose a camera records an average of $0.8$ detections per night. If the expected detection
count scales with observation time, three nights have expected count

$$
\lambda = 3(0.8) = 2.4.
$$

Under an assumed Poisson model, the probability of exactly two detections
over those three nights is

$$
P(X=2)
=
\frac{e^{-2.4}(2.4)^2}{2!}
\approx 0.2613.
$$

We can calculate it in R:

```{r}
# Calculate the Poisson mean for three nights and one exact probability.
rate_per_night = 0.8
n_nights = 3
expected_detections = rate_per_night * n_nights
detections_to_check = 2

dpois(x = detections_to_check, lambda = expected_detections)
```

We can also ask the probability of at least five detections in three nights. We can compute
this from the cumulative probability (note that we calculate the cumulative probability for
four and then take the complement):

```{r}
# Calculate the probability of at least the requested number of detections.
minimum_detections = 5
1 - ppois(
  q = minimum_detections - 1,
  lambda = expected_detections
)
```

Unlike a binomial count, a Poisson count has no fixed upper limit. A count of 100 detections is
possible under this model, although extremely unlikely when $\lambda=2.4$.

```{r}
#| echo: false
#| fig-cap: "Poisson probability mass for detections during three camera nights."
# Calculate and plot the Poisson probability mass over a useful range of counts.
expected_detections = 2.4
maximum_detections_to_plot = 15
detection_values = seq.int(from = 0, to = maximum_detections_to_plot)
detection_mass = dpois(x = detection_values, lambda = expected_detections)

detection_mass_data = data.frame(
  detections = detection_values,
  probability = detection_mass
)

# Plot each possible detection count as a vertical probability line.
plot(
  detection_mass_data$detections,
  detection_mass_data$probability,
  type = "h",
  xaxt = "n",
  xlab = "Number of detections",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: flowering stems per orchid plot"}

A survey covers two square metres of a restoration site. Assume flowering stems occur
at an average rate of 1.2 per square metre. Calculate the probability of finding exactly
three stems in the surveyed area. Why does this model allow counts greater than three,
even though three is the count in the question?

::: {.callout-tip title="Solution" collapse="true"}

The expected count across two square metres is $\lambda=2(1.2)=2.4$. The probability is:

```{r}
# Convert the observation effort into the Poisson mean before calculating a probability.
rate_per_square_metre = 1.2
survey_area = 2
expected_stems = rate_per_square_metre * survey_area

dpois(x = 3, lambda = expected_stems)
```

:::
::::

## Conway-Maxwell-Poisson distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{COMPoissonReg::dcmp, pcmp, qcmp, rcmp} \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x = 0, 1, 2, \ldots \\
\text{Probability}: &\quad
\frac{\lambda^x}{(x!)^\nu Z(\lambda, \nu)} \\
\text{Normalising constant}: &\quad
Z(\lambda, \nu) = \sum_{j = 0}^{\infty}\frac{\lambda^j}{(j!)^\nu} \\
\text{Parameters}: &\quad \lambda\;(\texttt{lambda}) > 0 = \text{rate parameter};\quad
\nu\;(\texttt{nu}) > 0 = \text{dispersion parameter} \\
\text{Mean}: &\quad
E[X] = \lambda\frac{\partial}{\partial\lambda}\log Z(\lambda, \nu) \\
\text{Variance}: &\quad
\operatorname{Var}(X) = \lambda\frac{\partial}{\partial\lambda}E[X] \\
\text{Inverse}: &\quad
\lambda \text{ must be found numerically for a chosen } \mu \text{ and } \nu
\end{aligned}
$$

The Conway-Maxwell-Poisson distribution, often abbreviated COM-Poisson or CMP,
extends the Poisson distribution by adding a dispersion parameter. It is useful when
counts are more regular than a Poisson model allows (**underdispersion**) as well as
when they are more variable (**overdispersion**). When $\nu = 1$, it is exactly the
Poisson distribution. Values of $\nu > 1$ produce underdispersion, whereas values of
$0 < \nu < 1$ produce overdispersion.

Its probability mass function is

$$
P(X=x)=\frac{\lambda^x}{(x!)^\nu Z(\lambda, \nu)},
\qquad
Z(\lambda, \nu) = \sum_{j = 0}^{\infty}\frac{\lambda^j}{(j!)^\nu},
$$

where $Z(\lambda, \nu)$ is the normalising constant that makes the probabilities add up
to one. A useful way to see what $\nu$ does is to compare the probabilities of two
consecutive counts:

$$
\frac{P(X=x)}{P(X=x-1)}=\frac{\lambda}{x^\nu}.
$$

For $\nu=1$ this ratio is $\lambda/x$, as in the Poisson distribution. When $\nu>1$ the
ratio decreases faster as $x$ grows, so the probability is concentrated on fewer counts.
When $\nu<1$ it decreases more slowly, which gives a longer right tail.

Unlike the Poisson distribution, `lambda` is not generally the mean. The normalising
constant $Z(\lambda, \nu)$ and therefore the mean have no simple closed form. The
`COMPoissonReg` package calculates these quantities numerically; use `ecmp()` and
`vcmp()` when you need the mean or variance for particular parameter values.

### Example: clutch size

Counts such as the number of eggs in a clutch often vary less than a Poisson model
predicts. Suppose clutch size follows a Conway-Maxwell-Poisson distribution with
$\lambda=8$ and $\nu=2$. We cannot calculate $Z(\lambda,\nu)$ by hand, but the ratios of
consecutive probabilities are simple:

$$
\frac{P(X=1)}{P(X=0)}=\frac{8}{1^2}=8, \qquad
\frac{P(X=2)}{P(X=1)}=\frac{8}{2^2}=2, \qquad
\frac{P(X=3)}{P(X=2)}=\frac{8}{3^2}\approx 0.89.
$$

The probabilities increase up to a clutch of two eggs and decrease afterwards, so two is
the most likely clutch size. We can verify these ratios in R:

```{r}
# Requires the COMPoissonReg package.
library(COMPoissonReg)

# nu > 1 gives a distribution that is less variable than a Poisson distribution.
clutch_lambda = 8
clutch_nu = 2

# Probabilities of clutches with zero to three eggs.
clutch_mass = dcmp(x = 0:3, lambda = clutch_lambda, nu = clutch_nu)
clutch_mass

# Ratios of consecutive probabilities.
clutch_mass[2:4] / clutch_mass[1:3]
```

The mean and variance are calculated numerically:

```{r}
# Calculate the mean and variance of the clutch size model.
ecmp(lambda = clutch_lambda, nu = clutch_nu)
vcmp(lambda = clutch_lambda, nu = clutch_nu)
```

The mean is approximately $2.56$ eggs and the variance is approximately $1.42$ eggs
squared. A Poisson distribution with the same mean would have a variance of $2.56$, so
this model is underdispersed relative to the Poisson distribution.

```{r}
#| echo: false
#| fig-cap: "Conway-Maxwell-Poisson probability mass for clutch size."
# Calculate and plot the probability mass up to a count that is very unlikely to be exceeded.
clutch_values = 0:qcmp(q = 0.999, lambda = 8, nu = 2)
plot(
  clutch_values,
  dcmp(x = clutch_values, lambda = 8, nu = 2),
  type = "h",
  xlab = "Number of eggs",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: dispersion in a count model"}

Calculate the mean and variance of a Conway-Maxwell-Poisson distribution with $\lambda=2$
and $\nu=0.5$. Is this model overdispersed or underdispersed relative to the Poisson
distribution?

::: {.callout-tip title="Solution" collapse="true"}

```{r}
# Calculate the mean and variance for a dispersion parameter smaller than one.
ecmp(lambda = 2, nu = 0.5)
vcmp(lambda = 2, nu = 0.5)
```

The mean is approximately $4.55$ and the variance is approximately $7.92$. The variance
is larger than the mean, so the model is overdispersed relative to the Poisson
distribution, as expected for $\nu<1$.

:::
::::

## Negative binomial distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dnbinom, pnbinom, qnbinom, rnbinom} \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x = 0, 1, 2, \ldots \\
\text{Probability}: &\quad
\frac{\Gamma(x+k)}{\Gamma(k)x!}p^k(1-p)^x \\
\text{Parameters}: &\quad k\;(\texttt{size}) = \text{dispersion};\quad
p\;(\texttt{prob}) = \text{prob success} \\
\text{Mean}: &\quad E[x] = k(1-p)/p \\
\text{Variance}: &\quad Var(X) = \mu + \mu^2/k \\
\text{Inverse}: &\quad p = k/(k+\mu)
\end{aligned}
$$

When overdispersion occurs, the negative binomial distribution is a useful starting model.
It has support $\{0, 1, 2, \ldots\}$ and can accommodate more variation
than a Poisson model with the same mean. The negative binomial model has two parameters:

- $\mu>0$: the mean count per sampling unit;
- $k>0$: a shape or dispersion parameter that controls the additional variation

The probability mass function is

$$
P(X=x)
=
\frac{\Gamma(x+k)}{\Gamma(k)x!}
\left(\frac{k}{k+\mu}\right)^k
\left(\frac{\mu}{k+\mu}\right)^x,
$$

where $\Gamma$ is the gamma function, which extends the factorial to real values.

The mean and variance are:

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\mu+\frac{\mu^2}{k}.
$$

For $\mu>0$ and finite $k$, the variance exceeds the mean. At a fixed $\mu$, smaller $k$ gives
greater variance. As $k$ increases, the extra variation fades and the variance approaches the mean:

$$
\mu+\frac{\mu^2}{k}\longrightarrow\mu.
$$

Thus, the negative binomial distribution approaches the Poisson distribution as $k$ becomes
large.

In R, `dnbinom()` and related functions, the argument `size` corresponds to $k$. This naming
can be confusing because `size` does not mean the sample size here; it is the negative
binomial shape parameter. The *Details* section of the documentation page for `dnbinom`
explains why it has that name.

### Example: seedlings in quadrats

Suppose the mean number of seedlings per quadrat is $\mu=4$, and the shape parameter is $k=1.5$.
The expected number of seedlings is

$$
E[X]=4.
$$

The variance is

$$
\operatorname{Var}(X)
=
4+\frac{4^2}{1.5}
\approx 14.67.
$$

Across comparable quadrats, the model predicts four seedlings on average and a variance of
approximately $14.67$ seedlings squared. That is much more variation than the Poisson model would
allow at the same mean.

We can verify these calculations in R:

```{r}
# Calculate the negative binomial mean and variance from its parameters.
mean_seedlings = 4
negative_binomial_size = 1.5

expected_seedlings = mean_seedlings
variance_seedlings = mean_seedlings +
  mean_seedlings^2 / negative_binomial_size

c(mean = expected_seedlings, variance = variance_seedlings)
```

The probability of observing exactly 10 seedlings is:

```{r}
# Calculate the probability of one particular seedling count.
seedlings_to_check = 10

dnbinom(
  x = seedlings_to_check,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

The probability of observing at least 10 seedlings is:

```{r}
# Calculate the probability of at least the requested number of seedlings.
minimum_seedlings = 10

1 - pnbinom(
  q = minimum_seedlings - 1,
  size = negative_binomial_size,
  mu = mean_seedlings
)
```

Like the Poisson distribution, the negative binomial distribution has no fixed
upper limit. A count of 100 is mathematically possible, although very unlikely if your mean
is 4.

```{r}
#| echo: false
#| fig-cap: "Negative binomial probability mass for seedlings in a quadrat."
# Calculate and plot the negative binomial probability mass over a useful range.
mean_seedlings = 4
negative_binomial_size = 1.5
maximum_seedlings_to_plot = 20
seedling_values = seq.int(from = 0, to = maximum_seedlings_to_plot)
negative_binomial_mass = dnbinom(
  x = seedling_values,
  size = negative_binomial_size,
  mu = mean_seedlings
)

negative_binomial_mass_data = data.frame(
  seedlings = seedling_values,
  probability = negative_binomial_mass
)

# Vertical lines make the separate possible count probabilities visible.
plot(
  negative_binomial_mass_data$seedlings,
  negative_binomial_mass_data$probability,
  type = "h",
  xaxt = "n",
  xlab = "Number of seedlings",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: orchid stem counts with different dispersion"}

Suppose the mean number of flowering stems per plot is $\mu=3$. Calculate the negative
binomial variance for $k=1.5$ and $k=6$. Which value gives greater overdispersion? How
does increasing $k$ relate to the Poisson model?

::: {.callout-tip title="Solution" collapse="true"}

Using $\operatorname{Var}(X)=\mu+\mu^2/k$, the variances are 9 and 4.5, respectively.
The smaller value of $k$ gives greater overdispersion. As $k$ increases, the variance
approaches the mean, which is the Poisson mean–variance relationship.

:::
::::

## Beta-Binomial distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dbetabinom.ab, pbetabinom.ab}\\
          &\quad \texttt{qbetabinom.ab, rbetabinom.ab} \\
\text{Type}: &\quad \text{Discrete} \\
\text{Range}: &\quad x = 0, 1, \ldots, n \\
\text{Probability}: &\quad
\binom{n}{x}\frac{B(x+a,n-x+b)}{B(a,b)} \\
\text{Parameters}: &\quad n\;(\texttt{size}) = \text{no. trials};\quad
a\;(\texttt{shape1}),\ b\;(\texttt{shape2}) = \text{shapes} \\
\text{Mean}: &\quad E[x] = n\frac{a}{a+b} \\
\text{Variance}: &\quad
Var(X) = \frac{nab(a+b+n)}
{(a+b)^2(a+b+1)} \\
\text{Inverse}: &\quad
\kappa = a+b;\quad
a = \kappa\frac{\mu}{n};\quad
b = \kappa\left(1-\frac{\mu}{n}\right)
\end{aligned}
$$

The Beta-Binomial distribution is useful when every sampling unit contains a fixed number
of binary events, as in the Binomial distribution, but the probability of the outcome of
interest varies from one sampling unit to another. It can be constructed in two steps.
First, the probability $p$ of each sampling unit is drawn from a Beta distribution with
shape parameters $a>0$ and $b>0$. Then, the count in that unit follows a Binomial
distribution with $n$ trials and that probability $p$. The result is a count between $0$
and $n$ that varies more than a Binomial count with the same mean, that is, it has
overdispersion relative to the Binomial distribution.

The probability mass function is

$$
P(X=x)=\binom{n}{x}\frac{B(x+a,\,n-x+b)}{B(a,b)},
$$

where $B(a,b)=\Gamma(a)\Gamma(b)/\Gamma(a+b)$ is the beta function.

These functions are available in the `VGAM` package. I recommend parameterising from mean,
size and a precision parameter ($\kappa$ inverse of a scale, the larger it is the narrower the
distribution). Writing $\bar{p}=a/(a+b)$ for the mean probability across sampling units and
$\kappa=a+b$ for the precision, the mean and variance are:

$$
\mu=E[X]=n\bar{p}
$$

and

$$
\operatorname{Var}(X)=n\bar{p}(1-\bar{p})\,\frac{\kappa+n}{\kappa+1}.
$$

The first part of the variance, $n\bar{p}(1-\bar{p})$, is the variance of a Binomial distribution.
The factor $(\kappa+n)/(\kappa+1)$ is larger than one whenever $n>1$ and it represents the
additional variation. As $\kappa$ increases this factor approaches one, so the Beta-Binomial
distribution approaches the Binomial distribution.

### Example: tadpole survival in tanks

Suppose 20 tadpoles are placed in each tank and the mean survival probability is $0.25$,
but the probability differs among tanks, which we describe with a precision of
$\kappa=12$. The shape parameters are

$$
a = 12(0.25) = 3, \qquad b = 12(1-0.25) = 9.
$$

The expected number of survivors per tank is $20(0.25)=5$ and the variance is

$$
\operatorname{Var}(X)
=
20(0.25)(0.75)\,\frac{12+20}{12+1}
=
3.75 \times \frac{32}{13}
\approx 9.23.
$$

A Binomial model with the same mean has a variance of only $3.75$. We can verify the
variance with the formula written in terms of $a$ and $b$:

```{r}
# Convert the mean and precision into the two shape parameters.
tadpole_count = 20
mean_survivors_per_tank = 5
survival_precision = 12
shape1 = survival_precision * mean_survivors_per_tank / tadpole_count
shape2 = survival_precision * (1 - mean_survivors_per_tank / tadpole_count)

# Calculate the variance from the shape parameters.
tadpole_count * shape1 * shape2 * (shape1 + shape2 + tadpole_count) /
  ((shape1 + shape2)^2 * (shape1 + shape2 + 1))
```

The additional variation is most visible in the tails. The probability that at least 10
tadpoles survive in a tank is much higher than under the Binomial model:

```{r}
#| message: false
# Requires the VGAM package
library(VGAM)

# Probability of at least ten survivors under the Beta-Binomial model.
1 - pbetabinom.ab(q = 9, size = tadpole_count, shape1 = shape1, shape2 = shape2)

# The same probability under a Binomial model with the same mean.
1 - pbinom(q = 9, size = tadpole_count, prob = mean_survivors_per_tank / tadpole_count)
```

```{r}
#| echo: false
#| fig-cap: "Beta-Binomial probability mass for surviving tadpoles in a tank (lines), compared with a Binomial distribution with the same mean (points)."
# Calculate the probability mass under both models for all possible counts.
survivor_values = 0:20
beta_binomial_mass = VGAM::dbetabinom.ab(
  x = survivor_values,
  size = 20,
  shape1 = 3,
  shape2 = 9
)
binomial_mass = dbinom(x = survivor_values, size = 20, prob = 0.25)

# Vertical lines show the Beta-Binomial model and points the Binomial model.
plot(
  survivor_values,
  beta_binomial_mass,
  type = "h",
  ylim = c(0, max(binomial_mass)),
  xlab = "Number of surviving tadpoles",
  ylab = "Probability"
)
points(survivor_values, binomial_mass, pch = 1)
legend(
  "topright",
  legend = c("Beta-Binomial", "Binomial"),
  lty = c(1, NA),
  pch = c(NA, 1),
  bty = "n"
)
```

:::: {.callout-important title="Practice: variation among tanks"}

For the tadpole example ($n=20$ and mean survival probability $0.25$), calculate the
variance of the number of survivors for $\kappa=2$ and for $\kappa=100$. Which of the
two models is closer to a Binomial distribution?

::: {.callout-tip title="Solution" collapse="true"}

The Binomial variance is $20(0.25)(0.75)=3.75$. For $\kappa=2$ it is multiplied by
$(2+20)/(2+1)\approx 7.33$, which gives a variance of $27.5$. For $\kappa=100$ it is
multiplied by $(100+20)/(100+1)\approx 1.19$, which gives a variance of approximately
$4.46$. The model with $\kappa=100$ is closer to the Binomial distribution, because a
large precision means that the survival probability hardly varies among tanks.

:::
::::

# Continuous distribution

Seven distributions for continous data that cover the common scenarios in ecology (note that
there is more diversity here than for count data). They are classified based on the values
the random variable can take:

**Unconstrained** - Any real value (from $-\infty$ to $\infty$).

**Positive** - Any positive real value (from $0$ to $\infty$). Sometimes $0$ is not included.

**Constrained** - Any real value within a range $[a,b]$ (from $a$ to $b$). Sometimes the edges
(i.e., $a$ and $b$) are not included.

Other distributions that show up in Ecology but I did not cover include Cauchy, Gompertz or
Power laws.

## Unconstrained values

**Normal** - Symmetric distribution. Arises from addition of many small,
independent random processes. The most common distribution, especially for measurement
error but a lot ecological variation is not Normal!

**T** - Like a Normal distribution but with thicker tails. Useful when dealing with a few
extreme values (i.e., robustness against *outliers*).

## Positive values

All distributions in this category are assymetric, especially when there is high probability
close to 0.

**Exponential** - Maximum at $0$. Distribution of waiting times
for a single event to happen if likelikood of the event per unit of time is constant.

**Log-Normal** -$X$ follows a Log-Normal if $\text{log}(X)$ follows
a Normal distribution. Arises from multiplication of many small, independent random processes.
Quite common for processes involving exponential growth.

**Gamma** - Distribution of waiting times for a certain number of
events to happen (same assumption as exponential). Very similar shape to Log-Normal.

**Weibull** - Distribution of waiting times for a a single event to
happen assuming the likelihood of the event per unit of time is not constant.

## Constrained values

**Uniform** - All values within the range are equally likely.

**Beta** - Values within the range are not equally likely. This distribution can be symmetric
or assymetric (left skewed or right skewed) depending on parameter values.

## Uniform distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dunif, punif, qunif, runif} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad \texttt{min} \leq x \leq \texttt{max} \\
\text{Density}: &\quad \frac{1}{b-a},\quad a \leq x \leq b \\
\text{Parameters}: &\quad a\;(\texttt{min}) = \text{lower bound};\quad
b\;(\texttt{max}) = \text{upper bound} \\
\text{Mean}: &\quad E[X] = (a+b)/2 \\
\text{Variance}: &\quad Var(X) = (b-a)^2/12 \\
\end{aligned}
$$

The Uniform distribution is the model for a continuous variable that is equally likely to
take any value between a lower bound $a$ and an upper bound $b$. Its density is constant
within those bounds and zero outside them:

$$
f_X(x)=\frac{1}{b-a}, \qquad a \leq x \leq b.
$$

Because the density is constant, the probability of an interval only depends on its length.
For any $c$ and $d$ with $a \leq c \leq d \leq b$,

$$
P(c \leq X \leq d)=\frac{d-c}{b-a}.
$$

The mean and variance are:

$$
\mu=E[X]=\frac{a+b}{2}
$$

and

$$
\operatorname{Var}(X)=\frac{(b-a)^2}{12}.
$$

The Uniform distribution is rarely a good model for an ecological measurement, but it is a
common building block in simulations (e.g., to choose random positions or directions). We
also use `runif()` to construct the random number generator of a zero-inflated distribution
(see @sec-zero_inflated).

### Example: position along a transect

Suppose a seedling is equally likely to establish anywhere along a 50 m transect. Let $X$ be
its position in metres, so $a=0$ and $b=50$. The density is $1/50=0.02$ per metre and the
probability that the seedling establishes between 10 m and 25 m is

$$
P(10 \leq X \leq 25)=\frac{25-10}{50-0}=0.3.
$$

The mean position is $(0+50)/2=25$ m and the variance is $50^2/12\approx 208.3$ m$^2$,
which corresponds to a standard deviation of approximately $14.4$ m.

We can verify our math with R:

```{r}
# Set the bounds of the transect in metres.
transect_start = 0
transect_end = 50

# Subtract cumulative probabilities to obtain the probability of an interval.
punif(q = 25, min = transect_start, max = transect_end) -
  punif(q = 10, min = transect_start, max = transect_end)

# Calculate the standard deviation of the position.
sqrt((transect_end - transect_start)^2 / 12)
```

```{r}
#| echo: false
#| fig-cap: "Uniform probability density for the position of a seedling along a 50 m transect."
# Calculate the density for positions inside and just outside the transect.
position_values = seq(from = -10, to = 60, length.out = 701)
plot(
  position_values,
  dunif(x = position_values, min = 0, max = 50),
  type = "l",
  xlab = "Position along the transect (m)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: direction of dispersal"}

Suppose the direction in which a seed leaves its parent plant is equally likely to be any
angle between 0 and 360 degrees. Calculate the probability that the direction falls between
45 and 135 degrees. What is the standard deviation of the direction?

::: {.callout-tip title="Solution" collapse="true"}

The interval is 90 degrees long, so the probability is $90/360=0.25$. The standard deviation
is $\sqrt{360^2/12}\approx 103.9$ degrees:

```{r}
# Calculate the probability of the interval and the standard deviation.
punif(q = 135, min = 0, max = 360) - punif(q = 45, min = 0, max = 360)
sqrt(360^2 / 12)
```

:::
::::

## Normal distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dnorm, pnorm, qnorm, rnorm} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad -\infty < x < \infty \\
\text{Density}: &\quad
\frac{1}{\sigma\sqrt{2\pi}}
\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right] \\
\text{Parameters}: &\quad \mu\;(\texttt{mean}) = \text{mean};\quad
\sigma\;(\texttt{sd}) = \text{standard deviation} \\
\text{Mean}: &\quad E[X] = \mu \\
\text{Variance}: &\quad Var(X) = \sigma^2 \\
\end{aligned}
$$

The Normal distribution is a symmetric model for a continuous variable that can take any
real value. It arises when a variable is the sum of many small, independent contributions,
which is why it is often used for measurement error. It has two parameters: the mean $\mu$,
which sets the centre of the distribution, and the standard deviation $\sigma>0$, which
sets its spread. We write $X\sim\operatorname{Normal}(\mu,\sigma^2)$. Note that the second
quantity in this notation is the variance, whereas the `sd` argument in R takes the
standard deviation $\sigma$.

Its probability density function is

$$
f_X(x)=\frac{1}{\sigma\sqrt{2\pi}}
\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right].
$$

The mean and variance are:

$$
E[X]=\mu
$$

and

$$
\operatorname{Var}(X)=\sigma^2.
$$

Because the density is symmetric around $\mu$, the mean, the median and the mode coincide.
Approximately 68% of the probability lies within one standard deviation of the mean, 95%
within two and 99.7% within three. We can express any value in units of standard deviations
with $Z=(X-\mu)/\sigma$, which follows a Normal distribution with mean $0$ and variance $1$
(the *standard* Normal distribution).

A Normal model always assigns some probability to negative values. For a positive
measurement this is only a concern when zero is within a few standard deviations of the
mean; otherwise that probability is negligible.

### Example: tree heights

As in [Chapter 3](../Chapter_3/Theory.qmd), suppose tree heights follow a Normal
distribution with $\mu=22$ m and $\sigma=4$ m. A height of 30 m is two standard deviations
above the mean:

$$
z=\frac{30-22}{4}=2.
$$

Approximately 95% of the probability lies between 14 m and 30 m, so about 5% lies outside
that interval. By symmetry, half of it is above 30 m:

$$
P(X>30)\approx \frac{1-0.95}{2}=0.025.
$$

We can calculate it in R:

```{r}
# Set the parameters of the tree height model in metres.
mean_height = 22
sd_height = 4

# Use the upper tail for the probability of a height above 30 m.
pnorm(q = 30, mean = mean_height, sd = sd_height, lower.tail = FALSE)
```

The result is approximately $0.0228$, slightly smaller than our approximation because the
interval that contains exactly 95% of the probability is a bit narrower than two standard
deviations on each side of the mean.

```{r}
#| echo: false
#| fig-cap: "Normal probability density for tree height."
# Calculate the density within four standard deviations of the mean.
height_values = seq(from = 6, to = 38, length.out = 400)
plot(
  height_values,
  dnorm(x = height_values, mean = 22, sd = 4),
  type = "l",
  xlab = "Tree height (m)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: short trees and a central interval"}

For the tree height model above ($\mu=22$ m and $\sigma=4$ m), calculate the probability
that a tree is shorter than 16 m. Between which two heights do we find the central 90% of
the distribution?

::: {.callout-tip title="Solution" collapse="true"}

The probability of a height below 16 m is a cumulative probability. The central 90% of the
distribution lies between the 5% and the 95% quantiles:

```{r}
# Calculate a lower-tail probability and the quantiles that enclose the central 90%.
pnorm(q = 16, mean = 22, sd = 4)
qnorm(p = c(0.05, 0.95), mean = 22, sd = 4)
```

The probability is approximately $0.067$ and the interval goes from approximately 15.4 m to
28.6 m.

:::
::::

## T distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dt, pt, qt, rt} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad -\infty < x < \infty \\
\text{Density}: &\quad
\frac{\Gamma((\nu+1)/2)}{s\sqrt{\nu\pi}\,\Gamma(\nu/2)}
\left[1+\frac{(x-\mu)^2}{\nu s^2}\right]^{-(\nu+1)/2} \\
\text{Parameters}: &\quad
\mu\;(\text{location});\quad s\;(\text{scale})>0;\quad
\nu\;(\texttt{df})>0 \\
\text{Mean}: &\quad \operatorname{E}[X]=\mu \quad (\nu>1) \\
\text{Variance}: &\quad \operatorname{Var}(X)=s^2\frac{\nu}{\nu-2}
\quad (\nu>2) \\
\text{Inverse}: &\quad
s=\sigma\sqrt{\frac{\nu-2}{\nu}}
\quad (\nu>2)
\end{aligned}
$$

The T distribution is symmetric and bell-shaped like the Normal distribution, but it has
thicker tails. That is, it assigns more probability to values that are far from the centre.
This makes it useful when the data contain a few extreme values that a Normal model would
treat as practically impossible. It has three parameters: a location $\mu$ that sets the
centre, a scale $s>0$ that sets the spread and the degrees of freedom $\nu>0$ that control
how thick the tails are. The smaller $\nu$ is, the thicker the tails. As $\nu$ increases,
the T distribution approaches the Normal distribution.

Its probability density function is

$$
f_X(x)=\frac{\Gamma((\nu+1)/2)}{s\sqrt{\nu\pi}\,\Gamma(\nu/2)}
\left[1+\frac{(x-\mu)^2}{\nu s^2}\right]^{-(\nu+1)/2},
$$

where $\Gamma$ is the gamma function. The mean and variance are:

$$
E[X]=\mu \qquad (\nu>1)
$$

and

$$
\operatorname{Var}(X)=s^2\frac{\nu}{\nu-2} \qquad (\nu>2).
$$

Note that the scale $s$ is not the standard deviation of the distribution. The standard
deviation is $\sigma=s\sqrt{\nu/(\nu-2)}$, which is always larger than $s$. So, if we want a
T distribution with a particular standard deviation $\sigma$, we need to use the scale
$s=\sigma\sqrt{(\nu-2)/\nu}$.

The T distribution functions built into R are standardised (i.e., $\mu = 0$ and $s = 1$)
and only depend on the degrees of freedom. If we want a different location or scale, we
define our random variable as $X = \mu + s \cdot T$, where $T$ follows the standardised T
distribution. In practice, this means that we standardise a value as $(x-\mu)/s$ before
passing it to `pt()`, and that the density is `dt((x - mu) / s, df) / s`. The standardised
T distribution is also the source of the critical values used for some of the confidence
intervals in [Chapter 4](../Chapter_4/Theory.qmd).

### Example: tree heights with thicker tails

Suppose tree heights have a mean of 22 m and a standard deviation of 4 m, as in the Normal
example, but follow a T distribution with $\nu=5$. The scale is

$$
s=4\sqrt{\frac{5-2}{5}}\approx 3.10 \text{ m}.
$$

To calculate the probability of a height above 30 m we first standardise that height:

$$
t=\frac{30-22}{3.10}\approx 2.58.
$$

The upper tail probability of the standardised T distribution at that value is
approximately $0.025$. We can calculate it in R:

```{r}
# Convert the standard deviation into the scale of the T distribution.
mean_height = 22
sd_height = 4
height_df = 5
height_scale = sd_height * sqrt((height_df - 2) / height_df)

# Standardise the height before calculating the upper-tail probability.
pt(q = (30 - mean_height) / height_scale, df = height_df, lower.tail = FALSE)
```

This probability is close to the $0.023$ of the Normal model. The difference between the
two models appears further out in the tail. For a height above 38 m (four standard
deviations above the mean):

```{r}
# Compare the probability of a very tall tree under the T and Normal models.
pt(q = (38 - mean_height) / height_scale, df = height_df, lower.tail = FALSE)
pnorm(q = 38, mean = mean_height, sd = sd_height, lower.tail = FALSE)
```

The T model gives a probability of approximately $0.0018$, whereas the Normal model gives
approximately $0.00003$.

```{r}
#| echo: false
#| fig-cap: "Probability density of tree height under a T distribution with five degrees of freedom (solid line) and a Normal distribution (dashed line) with the same mean and standard deviation."
# Calculate both densities over the same range of heights.
height_values = seq(from = 2, to = 42, length.out = 400)
t_scale = 4 * sqrt((5 - 2) / 5)
t_density = dt(x = (height_values - 22) / t_scale, df = 5) / t_scale
normal_density = dnorm(x = height_values, mean = 22, sd = 4)

# Use different line types to distinguish the two models.
plot(
  height_values,
  t_density,
  type = "l",
  xlab = "Tree height (m)",
  ylab = "Density"
)
lines(height_values, normal_density, lty = 2)
legend("topright", legend = c("T", "Normal"), lty = c(1, 2), bty = "n")
```

:::: {.callout-important title="Practice: degrees of freedom and extreme values"}

Consider two T models for tree height with a mean of 22 m and a standard deviation of 4 m,
one with $\nu=3$ and one with $\nu=30$. For each model, calculate the probability that a
height is more than three standard deviations away from the mean (i.e., below 10 m or
above 34 m). Compare the results with the Normal model.

::: {.callout-tip title="Solution" collapse="true"}

By symmetry, the probability is twice the upper tail probability at 34 m. For each model we
calculate the scale and standardise the height:

```{r}
# Calculate the two-sided tail probability for each number of degrees of freedom.
scale_df3 = 4 * sqrt((3 - 2) / 3)
2 * pt(q = (34 - 22) / scale_df3, df = 3, lower.tail = FALSE)

scale_df30 = 4 * sqrt((30 - 2) / 30)
2 * pt(q = (34 - 22) / scale_df30, df = 30, lower.tail = FALSE)

# The same probability under the Normal model.
2 * pnorm(q = 34, mean = 22, sd = 4, lower.tail = FALSE)
```

The probabilities are approximately $0.014$ for $\nu=3$, $0.004$ for $\nu=30$ and $0.003$
for the Normal model. The fewer the degrees of freedom, the more probability the model
assigns to extreme values, and with $\nu=30$ the model is already close to the Normal
model.

:::
::::

## Exponential distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dexp, pexp, qexp, rexp} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad \text{Positive} \\
\text{Density}: &\quad \lambda e^{-\lambda x} \\
\text{Parameters}: &\quad \lambda\;(\texttt{rate})=\ \text{Rate} \\
\text{Mean}: &\quad E[x] = 1/\lambda \\
\text{Variance}: &\quad Var(X) = 1/\lambda^2 \\
\text{Inverse}: &\quad \lambda = 1/\mu
\end{aligned}
$$

The Exponential distribution is the model for the waiting time until an event happens,
when the event occurs at a constant rate $\lambda>0$ per unit of time. A constant rate means
that the chance of the event happening in the next instant does not depend on how long we
have already waited. Under the same assumption, the number of events in a fixed period of
time follows a Poisson distribution.

Its probability density function is

$$
f_X(x)=\lambda e^{-\lambda x}, \qquad x \geq 0,
$$

which has its maximum at $x=0$ and decreases from there. The cumulative probability has a
simple formula, $P(X \leq x)=1-e^{-\lambda x}$, so the probability of waiting longer than
$x$ is $P(X>x)=e^{-\lambda x}$. The mean and variance are:

$$
\mu=E[X]=\frac{1}{\lambda}
$$

and

$$
\operatorname{Var}(X)=\frac{1}{\lambda^2}.
$$

That is, the standard deviation of an Exponential distribution is always equal to its
mean. Note that the rate has units of events per unit of time, whereas the mean has units
of time.

### Example: waiting for a pollinator

Suppose a flower is visited by a pollinator on average once every 2 minutes. The rate is
$\lambda=1/2=0.5$ visits per minute. The probability of waiting more than 5 minutes for the
next visit is

$$
P(X>5)=e^{-0.5(5)}=e^{-2.5}\approx 0.0821.
$$

We can verify our math with R:

```{r}
# Convert the mean waiting time (in minutes) into a rate.
mean_waiting_time = 2
visit_rate = 1 / mean_waiting_time

# Use the upper tail for the probability of waiting more than five minutes.
pexp(q = 5, rate = visit_rate, lower.tail = FALSE)
```

The median waiting time is the value at which the cumulative probability equals $0.5$, which
is $\log(2)/\lambda\approx 1.39$ minutes. It is shorter than the mean because the
distribution is skewed to the right:

```{r}
# Calculate the median waiting time with the quantile function.
qexp(p = 0.5, rate = visit_rate)
```

```{r}
#| echo: false
#| fig-cap: "Exponential probability density for the waiting time until the next pollinator visit."
# Calculate the density over a range that covers most of the probability.
waiting_values = seq(from = 0, to = 10, length.out = 400)
plot(
  waiting_values,
  dexp(x = waiting_values, rate = 0.5),
  type = "l",
  xlab = "Waiting time (minutes)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: time until a seed is found"}

Suppose a seed on the forest floor is found by a seed predator after 4 days on average and
that the rate at which seeds are found is constant. Calculate the probability that a seed
has not been found after 8 days. What is the median time until a seed is found?

::: {.callout-tip title="Solution" collapse="true"}

The rate is $\lambda=1/4=0.25$ per day. The probability that the seed has not been found
after 8 days is $e^{-0.25(8)}=e^{-2}\approx 0.135$ and the median is
$\log(2)/0.25\approx 2.77$ days:

```{r}
# Calculate the upper-tail probability and the median for a rate of 0.25 per day.
pexp(q = 8, rate = 1 / 4, lower.tail = FALSE)
qexp(p = 0.5, rate = 1 / 4)
```

:::
::::

## Gamma distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dgamma, pgamma, qgamma, rgamma} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad x \geq 0 \\
\text{Density}: &\quad
\frac{r^k}{\Gamma(k)}
x^{k-1}e^{-rx} \\
\text{Parameters}: &\quad k\;(\texttt{shape}) = \text{shape};\quad
r\;(\texttt{rate}) = \text{rate} \\
\text{Mean}: &\quad E[x] = k/r \\
\text{Variance}: &\quad Var(X) = k/r^2 \\
\text{Inverse}: &\quad r = k/\mu
\quad (\text{equivalently, } \texttt{scale}=\mu/k)
\end{aligned}
$$

The Gamma distribution is a model for positive continuous variables that are skewed to the
right. It can be derived as the waiting time until a certain number of events ($k$) have
happened, when the events occur at a constant rate $r$. That is, it is the distribution of
the sum of $k$ independent Exponential waiting times, and for $k=1$ it is the Exponential
distribution. However, $k$ does not need to be an integer and the Gamma distribution is
also used as a general model for positive measurements, in the same situations as the
LogNormal distribution.

Its probability density function is

$$
f_X(x)=\frac{r^k}{\Gamma(k)}x^{k-1}e^{-rx}, \qquad x \geq 0,
$$

where $k>0$ is the shape, $r>0$ is the rate and $\Gamma$ is the gamma function. The mean and
variance are:

$$
\mu=E[X]=\frac{k}{r}
$$

and

$$
\operatorname{Var}(X)=\frac{k}{r^2}=\frac{\mu^2}{k}.
$$

In other words, the standard deviation is proportional to the mean and the coefficient of
variation (the standard deviation divided by the mean) is $1/\sqrt{k}$. So the shape
controls the relative variation: the smaller $k$ is, the more variable and skewed the
distribution, whereas for large $k$ it becomes almost symmetric.

R's `rate` and `scale` are reciprocals: `scale = 1/rate`. I recommend to parameterise it from
the mean and either `rate` or `scale`.

### Example: waiting for the third pollinator

We continue with the flower that is visited by pollinators at a rate of $0.5$ visits per
minute. The waiting time until the third visit follows a Gamma distribution with $k=3$ and
$r=0.5$ per minute. The expected waiting time is

$$
E[X]=\frac{3}{0.5}=6 \text{ minutes}
$$

and the variance is

$$
\operatorname{Var}(X)=\frac{3}{0.5^2}=12 \text{ minutes}^2,
$$

which corresponds to a standard deviation of approximately $3.46$ minutes. The probability
of waiting more than 10 minutes for the third visit is:

```{r}
# Set the shape (number of visits) and the rate (visits per minute).
visit_shape = 3
visit_rate = 0.5

# Use the upper tail for the probability of waiting more than ten minutes.
pgamma(q = 10, shape = visit_shape, rate = visit_rate, lower.tail = FALSE)
```

If we know the mean instead of the rate, we calculate the rate as $r=k/\mu$. For a mean of
6 minutes and a shape of 3 we recover the same model:

```{r}
# Calculate the rate from the mean waiting time and the shape.
mean_waiting_time = 6
visit_rate_from_mean = visit_shape / mean_waiting_time

pgamma(q = 10, shape = visit_shape, rate = visit_rate_from_mean, lower.tail = FALSE)
```

```{r}
#| echo: false
#| fig-cap: "Gamma probability density for the waiting time until the third pollinator visit."
# Calculate the density over a range that covers most of the probability.
waiting_values = seq(from = 0, to = 20, length.out = 400)
plot(
  waiting_values,
  dgamma(x = waiting_values, shape = 3, rate = 0.5),
  type = "l",
  xlab = "Waiting time (minutes)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: seedling biomass"}

Suppose the dry biomass of a seedling follows a Gamma distribution with a mean of 2 g and a
shape of 3. Calculate the rate, the variance and the coefficient of variation. What is the
probability that a seedling weighs less than 1 g?

::: {.callout-tip title="Solution" collapse="true"}

The rate is $r=3/2=1.5$ per gram, the variance is $2^2/3\approx 1.33$ g$^2$ and the
coefficient of variation is $1/\sqrt{3}\approx 0.58$. The probability of a biomass below
1 g is a cumulative probability:

```{r}
# Calculate the rate from the mean and shape, then the lower-tail probability.
biomass_shape = 3
biomass_rate = biomass_shape / 2

pgamma(q = 1, shape = biomass_shape, rate = biomass_rate)
```

The probability is approximately $0.19$.

:::
::::

## LogNormal distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dlnorm, plnorm, qlnorm, rlnorm} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad x > 0 \\
\text{Density}: &\quad
\frac{1}{x\,\sigma\sqrt{2\pi}}
\exp\left[-\frac{(\log x-\mu)^2}{2\sigma^2}\right] \\
\text{Parameters}: &\quad \mu\;(\texttt{meanlog}) = \text{mean of log}(X);\quad
\sigma\;(\texttt{sdlog}) = \text{SD of log}(X) \\
\text{Mean}: &\quad E[x] = \exp(\mu+\sigma^2/2) \\
\text{Variance}: &\quad
Var(X) = \left[\exp(\sigma^2)-1\right]\exp(2\mu+\sigma^2) \\
\text{Inverse}: &\quad
\mu = \log(E[X])-\sigma^2/2
\end{aligned}
$$

A positive random variable $X$ follows a LogNormal distribution when its logarithm follows
a Normal distribution:

$$
\log(X)\sim\operatorname{Normal}(\mu,\sigma^2).
$$

It arises when a variable is the product of many small, independent factors, because the
logarithm turns that product into a sum. This makes it a common model for strictly positive
measurements that are skewed to the right, such as body mass. Note that $\mu$ and $\sigma$
are the mean and standard deviation of $\log(X)$ and not of $X$ itself. In R they are the
arguments `meanlog` and `sdlog`.

Its probability density function is

$$
f_X(x)=\frac{1}{x\sigma\sqrt{2\pi}}
\exp\left[-\frac{(\log x-\mu)^2}{2\sigma^2}\right], \qquad x>0.
$$

On the original scale, the median is $e^\mu$, whereas the mean and variance are:

$$
E[X]=e^{\mu+\sigma^2/2}
$$

and

$$
\operatorname{Var}(X)=\left(e^{\sigma^2}-1\right)e^{2\mu+\sigma^2}.
$$

The mean is always larger than the median because of the long right tail. If we want a
LogNormal distribution with a particular mean on the original scale, we need to solve for
$\mu$, which gives $\mu=\log(E[X])-\sigma^2/2$.

The LogNormal distribution cannot generate zeros. If the data contain zeros, a
zero-inflated distribution (see @sec-zero_inflated) is one possible alternative.

### Example: body mass

As in [Chapter 3](../Chapter_3/Theory.qmd), suppose body mass in grams follows a LogNormal
distribution with $\mu=\log(30)$ and $\sigma=0.5$. The median body mass is $e^{\log(30)}=30$
g and the mean is

$$
E[X]=e^{\log(30)+0.5^2/2}=30\,e^{0.125}\approx 33.99 \text{ g}.
$$

To calculate the probability of a body mass above 50 g we can work on the logarithmic
scale, where the model is a Normal distribution. In units of standard deviations, the
logarithm of 50 g is

$$
z=\frac{\log(50)-\log(30)}{0.5}\approx 1.02
$$

above the mean, and the probability of exceeding that value in a standard Normal
distribution is approximately $0.153$. We can verify our math with R:

```{r}
# Set the parameters on the logarithmic scale.
meanlog_mass = log(30)
sdlog_mass = 0.5

# Calculate the mean body mass on the original scale.
exp(meanlog_mass + sdlog_mass^2 / 2)

# The upper-tail probability is the same on the original and the logarithmic scale.
plnorm(q = 50, meanlog = meanlog_mass, sdlog = sdlog_mass, lower.tail = FALSE)
pnorm(q = (log(50) - meanlog_mass) / sdlog_mass, lower.tail = FALSE)
```

```{r}
#| echo: false
#| fig-cap: "LogNormal probability density for body mass."
# Calculate the density over a range that covers most of the probability.
mass_values = seq(from = 0.1, to = 120, length.out = 400)
plot(
  mass_values,
  dlnorm(x = mass_values, meanlog = log(30), sdlog = 0.5),
  type = "l",
  xlab = "Body mass (g)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: a model with a chosen mean"}

Suppose we want a LogNormal model for body mass with $\sigma=0.5$ and a mean (instead of a
median) of 30 g. Which value of $\mu$ do we need? What is the median body mass of that
model?

::: {.callout-tip title="Solution" collapse="true"}

We need $\mu=\log(30)-0.5^2/2\approx 3.28$. The median is $e^\mu\approx 26.5$ g, which is
smaller than the mean, as expected:

```{r}
# Calculate meanlog from the mean on the original scale, then the median and the mean.
meanlog_from_mean = log(30) - 0.5^2 / 2
meanlog_from_mean
exp(meanlog_from_mean)
exp(meanlog_from_mean + 0.5^2 / 2)
```

:::
::::

## Beta distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dbeta, pbeta, qbeta, rbeta} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad 0 \leq x \leq 1 \\
\text{Density}: &\quad
\frac{x^{a-1}(1-x)^{b-1}}{B(a,b)} \\
\text{Parameters}: &\quad a\;(\texttt{shape1}),\quad
b\;(\texttt{shape2}) = \text{shape parameters} \\
\text{Mean}: &\quad E[x] = a/(a+b) \\
\text{Variance}: &\quad
Var(X) = \frac{ab}{(a+b)^2(a+b+1)} \\
\text{Inverse}: &\quad
\kappa=a+b;\quad
a=\mu\kappa;\quad b=(1-\mu)\kappa
\end{aligned}
$$

The Beta distribution is a model for a continuous variable that takes values between $0$
and $1$, such as the fraction of a quadrat covered by vegetation or the fraction of a leaf
that has been eaten. When a proportion comes from counting outcomes in a known number of
trials, the Binomial distribution is the appropriate model. The Beta distribution has two
shape parameters, $a>0$ and $b>0$.

Its probability density function is

$$
f_X(x)=\frac{x^{a-1}(1-x)^{b-1}}{B(a,b)}, \qquad 0<x<1,
$$

where $B(a,b)=\Gamma(a)\Gamma(b)/\Gamma(a+b)$ is the beta function, which ensures that the
total probability is one. The mean and variance are:

$$
\mu=E[X]=\frac{a}{a+b}
$$

and

$$
\operatorname{Var}(X)=\frac{ab}{(a+b)^2(a+b+1)}.
$$

I recommend to parameterise it from the mean and a precision parameter (the larger it is
the narrower the distribution). Writing $\kappa=a+b$ for the precision, the shape
parameters are $a=\mu\kappa$ and $b=(1-\mu)\kappa$, and the variance becomes

$$
\operatorname{Var}(X)=\frac{\mu(1-\mu)}{\kappa+1}.
$$

The Beta distribution can take many shapes. It is symmetric when $a=b$, it places more
probability towards $1$ when $a>b$ and towards $0$ when $a<b$. When both parameters are
smaller than one the density is U-shaped, and for $a=b=1$ it is the Uniform distribution
between $0$ and $1$.

### Example: vegetation cover

As in [Chapter 3](../Chapter_3/Theory.qmd), suppose the fraction of a quadrat covered by
vegetation follows a Beta distribution with $a=2$ and $b=5$. The expected cover is

$$
E[X]=\frac{2}{2+5}\approx 0.286
$$

and the variance is

$$
\operatorname{Var}(X)=\frac{2(5)}{(2+5)^2(2+5+1)}=\frac{10}{392}\approx 0.0255,
$$

which corresponds to a standard deviation of approximately $0.16$. We can calculate the
probability of a cover below $0.2$ and the 95% quantile in R:

```{r}
# Set the two shape parameters of the cover model.
cover_shape1 = 2
cover_shape2 = 5

# Calculate a lower-tail probability and a high quantile.
pbeta(q = 0.2, shape1 = cover_shape1, shape2 = cover_shape2)
qbeta(p = 0.95, shape1 = cover_shape1, shape2 = cover_shape2)
```

If we start from a mean cover and a precision, we first calculate the shape parameters. For
example, a mean cover of $0.4$ with $\kappa=10$ corresponds to $a=0.4(10)=4$ and
$b=0.6(10)=6$:

```{r}
# Convert the mean and precision into the two shape parameters.
mean_cover = 0.4
cover_precision = 10
shape1_from_mean = mean_cover * cover_precision
shape2_from_mean = (1 - mean_cover) * cover_precision

c(shape1 = shape1_from_mean, shape2 = shape2_from_mean)
```

```{r}
#| echo: false
#| fig-cap: "Beta probability densities with different shape parameters for a fraction between zero and one."
# Calculate density curves for three combinations of shape parameters.
cover_values = seq(from = 0.001, to = 0.999, length.out = 400)
plot(
  cover_values,
  dbeta(x = cover_values, shape1 = 2, shape2 = 5),
  type = "l",
  lty = 1,
  ylim = c(0, 4),
  xlab = "Vegetation cover (fraction)",
  ylab = "Density"
)
lines(cover_values, dbeta(x = cover_values, shape1 = 5, shape2 = 2), lty = 2)
lines(cover_values, dbeta(x = cover_values, shape1 = 0.5, shape2 = 0.5), lty = 3)
legend(
  "top",
  legend = c("a = 2, b = 5", "a = 5, b = 2", "a = 0.5, b = 0.5"),
  lty = 1:3,
  bty = "n"
)
```

:::: {.callout-important title="Practice: precision of a cover model"}

Suppose the mean vegetation cover is $0.4$. Calculate the shape parameters and the standard
deviation of the cover for a precision of $\kappa=10$ and of $\kappa=50$. Which model is
narrower?

::: {.callout-tip title="Solution" collapse="true"}

For $\kappa=10$ the shape parameters are $a=4$ and $b=6$ and the variance is
$0.4(0.6)/11\approx 0.0218$, so the standard deviation is approximately $0.148$. For
$\kappa=50$ the shape parameters are $a=20$ and $b=30$ and the variance is
$0.4(0.6)/51\approx 0.0047$, so the standard deviation is approximately $0.069$. The model
with $\kappa=50$ is narrower, because a larger precision concentrates the probability
around the mean.

:::
::::

## Weibull distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dweibull, pweibull, qweibull, rweibull} \\
\text{Type}: &\quad \text{Continuous} \\
\text{Range}: &\quad x \geq 0 \\
\text{Density}: &\quad
\frac{k}{\lambda}\left(\frac{x}{\lambda}\right)^{k-1}
\exp\left[-\left(\frac{x}{\lambda}\right)^k\right] \\
\text{Parameters}: &\quad k\;(\texttt{shape})>0;\quad
\lambda\;(\texttt{scale})>0 \\
\text{Mean}: &\quad E[x] = \lambda\Gamma(1+1/k) \\
\text{Variance}: &\quad
Var(X) = \lambda^2\left[\Gamma(1+2/k)-\Gamma(1+1/k)^2\right] \\
\text{Inverse}: &\quad
\lambda = \frac{\mu}{\Gamma(1+1/k)}
\end{aligned}
$$

The Weibull distribution is a model for the waiting time until a single event happens when
the rate at which the event occurs changes with time. It is often used for lifespans and
survival times. It has a shape parameter $k>0$ and a scale parameter $\lambda>0$, where the
scale has the same units as the waiting time. The shape determines how the rate changes:
for $k<1$ the rate decreases with time (e.g., when mortality is highest in young
individuals), for $k>1$ it increases with time (e.g., when mortality increases with age)
and for $k=1$ it is constant, in which case we recover the Exponential distribution with
rate $1/\lambda$.

Its probability density function is

$$
f_X(x)=\frac{k}{\lambda}\left(\frac{x}{\lambda}\right)^{k-1}
\exp\left[-\left(\frac{x}{\lambda}\right)^k\right], \qquad x \geq 0.
$$

The probability that the event has not yet happened at time $x$ has a simple formula:

$$
P(X>x)=\exp\left[-\left(\frac{x}{\lambda}\right)^k\right].
$$

The mean and variance are:

$$
\mu=E[X]=\lambda\,\Gamma\left(1+\frac{1}{k}\right)
$$

and

$$
\operatorname{Var}(X)=\lambda^2\left[\Gamma\left(1+\frac{2}{k}\right)-
\Gamma\left(1+\frac{1}{k}\right)^2\right],
$$

where $\Gamma$ is the gamma function (`gamma()` in R). I show below how to parameterise from
the mean and shape.

### Example: leaf lifespan

Suppose the lifespan of a leaf has a mean of 2 years and follows a Weibull distribution
with shape $k=1.5$. The scale is

$$
\lambda=\frac{2}{\Gamma(1+1/1.5)}\approx\frac{2}{0.9027}\approx 2.215 \text{ years}.
$$

The probability that a leaf lives longer than 3 years is

$$
P(X>3)=\exp\left[-\left(\frac{3}{2.215}\right)^{1.5}\right]
\approx e^{-1.576}\approx 0.207.
$$

We can verify our math with R:

```{r}
# Calculate the scale from the mean lifespan (in years) and the shape.
mean_lifespan = 2
lifespan_shape = 1.5
lifespan_scale = mean_lifespan / gamma(1 + 1 / lifespan_shape)
lifespan_scale

# Use the upper tail for the probability of living longer than three years.
pweibull(q = 3, shape = lifespan_shape, scale = lifespan_scale, lower.tail = FALSE)
```

```{r}
#| echo: false
#| fig-cap: "Weibull probability density for the lifespan of a leaf."
# Calculate the density over a range that covers most of the probability.
lifespan_values = seq(from = 0, to = 7, length.out = 400)
plot(
  lifespan_values,
  dweibull(x = lifespan_values, shape = 1.5, scale = 2 / gamma(1 + 1 / 1.5)),
  type = "l",
  xlab = "Leaf lifespan (years)",
  ylab = "Density"
)
```

:::: {.callout-important title="Practice: seedling survival time"}

Suppose the survival time of a seedling follows a Weibull distribution with shape $k=0.8$
and scale $\lambda=2$ years. Does the mortality rate increase or decrease with age?
Calculate the probability that a seedling survives its first year and the mean survival
time.

::: {.callout-tip title="Solution" collapse="true"}

The shape is smaller than one, so the mortality rate decreases with age. The probability of
surviving the first year is $\exp[-(1/2)^{0.8}]\approx 0.563$ and the mean survival time is
$2\,\Gamma(1+1/0.8)\approx 2.27$ years:

```{r}
# Calculate the probability of surviving beyond one year and the mean survival time.
pweibull(q = 1, shape = 0.8, scale = 2, lower.tail = FALSE)
2 * gamma(1 + 1 / 0.8)
```

:::
::::

# Multivariate distributions

These are distributions of multiple random variables that are not independent. They appear
in two main context:

(i) Quantifying uncertainty or variation in parameters (typically the multivariate normal).

(ii) Modelling multiple types of observations that are ecologically correlated.

**Multinomial** - Generalizes the Binomial distribution to $K$ categories. The counts across
categories must sum to the fixed number of trials $N$. For example counts across several
species or habitat categories when the total number of observations is fixed.

**Multivariate Normal** - Generalizes the Normal distribution to $K$ correlated random variables.
Each variable is characterized by its mean and standard deviation but we also defined a
correlation coefficient for each pair of variables. The basis of multivariate statistics for
unconstrained, continuous data and also shows up when quantifying uncertainty/variation in
parameters or when modelling time series with autocorrelation.

**Dirichlet** - Generalizes the Beta distribution to $K$ correlated random variables. Each
of the random variables is constrained between 0 and 1 but also, the sum of all the variables
is always equal to 1. Useful to model proportions of a community across species or partitioning
of biomass across different compartments, soil texture, etc.

This area of probability is (relatively) less explored. If you are really interested in multivariate
modelling, I suggest to take a look into [*copulas*](https://www.sciencedirect.com/science/chapter/bookseries/abs/pii/S0065250420300039) but note that this is way beyond the scope of this course and a very rare
skill among ecological modelers, so it might be too hard to tackle now.

## Multinomial distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{dmultinom, rmultinom} \\
\text{Type}: &\quad \text{Discrete, multivariate} \\
\text{Range}: &\quad x_i \in \{0,1,\ldots,n\},\quad \sum_{i=1}^K x_i=n \\
\text{Probability}: &\quad
\frac{n!}{\prod_{i=1}^K x_i!}\prod_{i=1}^K p_i^{x_i} \\
\text{Parameters}: &\quad n\;(\texttt{size}) = \text{number of trials};\quad
p_i\;(\texttt{prob}) = \text{probability of category }i,\quad \sum_i p_i=1 \\
\text{Mean}: &\quad \mu_i = np_i \\
\text{Variance}: &\quad \operatorname{Var}(X_i)=np_i(1-p_i) \\
\text{Covariance}: &\quad \operatorname{Cov}(X_i,X_j)=-np_ip_j,\quad i\ne j \\
\text{Inverse}: &\quad p_i = \mu_i/n,\quad \sum_i\mu_i=n
\end{aligned}
$$

The Multinomial distribution generalises the Binomial distribution to events that have more
than two possible outcomes. Each of $n$ independent events falls into one of $K$ categories,
where category $i$ has probability $p_i$ and the probabilities add up to one. The random
variables $X_1, \ldots, X_K$ are the number of events in each category, so they always add
up to $n$.

The probability mass function is

$$
P(X_1=x_1,\ldots,X_K=x_K)=\frac{n!}{x_1!\cdots x_K!}\,p_1^{x_1}\cdots p_K^{x_K},
$$

where the first factor counts the number of ways in which the $n$ events can be assigned to
the categories. The mean and variance of each count are:

$$
E[X_i]=np_i
$$

and

$$
\operatorname{Var}(X_i)=np_i(1-p_i).
$$

These are the mean and variance of a Binomial distribution. In fact, each count on its own
follows a Binomial distribution with $n$ trials and probability $p_i$. However, the counts
are not independent of each other. Because the total is fixed, a higher count in one
category leaves fewer events for the others, which results in a negative covariance:

$$
\operatorname{Cov}(X_i,X_j)=-np_ip_j, \qquad i \neq j.
$$

### Example: species in a sample

Suppose we capture 20 individuals and each of them belongs to one of three species with
probabilities $0.25$, $0.4$ and $0.35$. The expected counts are $20(0.25)=5$, $20(0.4)=8$
and $20(0.35)=7$. The probability of capturing exactly those numbers is

$$
P(X_1=5,X_2=8,X_3=7)
=
\frac{20!}{5!\,8!\,7!}(0.25)^5(0.4)^8(0.35)^7
\approx 0.0411,
$$

and the covariance between the counts of the first two species is $-20(0.25)(0.4)=-2$. We
can verify the probability in R:

```{r}
# Set the number of captured individuals and the probability of each species.
captured_count = 20
species_probability = c(0.25, 0.4, 0.35)

# Probability of one particular combination of counts.
dmultinom(x = c(5, 8, 7), size = captured_count, prob = species_probability)
```

We can check the means and the covariance by simulating many samples. Note that
`rmultinom()` returns a matrix with one row per category and one column per sample:

```{r}
# Simulate 5000 samples of 20 individuals.
set.seed(2026)
species_counts = rmultinom(n = 5000, size = captured_count, prob = species_probability)

# Every sample adds up to the total number of individuals.
range(colSums(species_counts))

# Average count per species and covariance between the first two species.
rowMeans(species_counts)
cov(species_counts[1, ], species_counts[2, ])
```

The averages and the covariance of the simulated samples are close to the values of the
model.

```{r}
#| echo: false
#| fig-cap: "Probability mass for the number of individuals of the first species in a sample of 20, which is a Binomial distribution."
# Each count of a Multinomial distribution follows a Binomial distribution.
first_species_values = 0:20
plot(
  first_species_values,
  dbinom(x = first_species_values, size = 20, prob = 0.25),
  type = "h",
  xlab = "Number of individuals of the first species",
  ylab = "Probability"
)
```

:::: {.callout-important title="Practice: habitat use"}

An animal is observed on 10 independent occasions and on each occasion it is in one of
three habitats with probabilities $0.5$, $0.3$ and $0.2$. Calculate the probability of
observing it 5, 3 and 2 times in each habitat, respectively. What is the covariance between
the counts in the first two habitats?

::: {.callout-tip title="Solution" collapse="true"}

The probability is

$$
\frac{10!}{5!\,3!\,2!}(0.5)^5(0.3)^3(0.2)^2 = 2520 \times 0.00003375 \approx 0.085,
$$

and the covariance is $-10(0.5)(0.3)=-1.5$. In R:

```{r}
# Probability of one particular combination of counts across the three habitats.
dmultinom(x = c(5, 3, 2), size = 10, prob = c(0.5, 0.3, 0.2))
```

:::
::::

## Multivariate Normal distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{MASS::mvrnorm};\quad
\texttt{mvtnorm::dmvnorm, pmvnorm, rmvnorm} \\
\text{Type}: &\quad \text{Continuous, multivariate} \\
\text{Range}: &\quad \mathbf{x}\in\mathbb{R}^K \\
\text{Density}: &\quad
\frac{1}{(2\pi)^{K/2}|\boldsymbol{\Sigma}|^{1/2}}
\exp\left[
-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^\mathsf{T}
\boldsymbol{\Sigma}^{-1}(\mathbf{x}-\boldsymbol{\mu})
\right] \\
\text{Parameters}: &\quad
\boldsymbol{\mu}\;(\texttt{mu}) = \text{mean vector};\quad
\boldsymbol{\Sigma}\;(\texttt{Sigma}) = \text{covariance matrix} \\
\text{Mean}: &\quad \operatorname{E}[\mathbf{X}]=\boldsymbol{\mu} \\
\text{Covariance}: &\quad
\operatorname{Cov}(\mathbf{X})=\boldsymbol{\Sigma} \\
\text{Inverse}: &\quad
\Sigma_{ii}=\sigma_i^2;\quad
\Sigma_{ij}=\rho_{ij}\sigma_i \sigma_j\quad(i\ne j)
\end{aligned}
$$

The Multivariate Normal distribution generalises the Normal distribution to $K$ continuous
random variables that are correlated with each other. It has two parameters: a vector
$\boldsymbol{\mu}$ with the mean of each variable and a covariance matrix
$\boldsymbol{\Sigma}$ with $K$ rows and $K$ columns. The diagonal of the covariance matrix
contains the variance of each variable and the other elements contain the covariance
between each pair of variables.

Its probability density function is

$$
f_{\mathbf{X}}(\mathbf{x})=
\frac{1}{(2\pi)^{K/2}|\boldsymbol{\Sigma}|^{1/2}}
\exp\left[
-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^\mathsf{T}
\boldsymbol{\Sigma}^{-1}(\mathbf{x}-\boldsymbol{\mu})
\right],
$$

where $|\boldsymbol{\Sigma}|$ is the determinant of the covariance matrix and
$\boldsymbol{\Sigma}^{-1}$ is its inverse.

It is usually easier to think in terms of standard deviations and correlations than in
terms of covariances. If $\sigma_i$ is the standard deviation of variable $i$ and
$\rho_{ij}$ is the correlation between variables $i$ and $j$, the elements of the covariance
matrix are:

$$
\Sigma_{ii}=\sigma_i^2
$$

and

$$
\Sigma_{ij}=\rho_{ij}\sigma_i\sigma_j, \qquad i \neq j.
$$

Each variable on its own follows a Normal distribution with mean $\mu_i$ and variance
$\sigma_i^2$. When all the correlations are zero, the variables are independent.

### Example: tree diameter and height

Suppose the diameter and the height of a tree follow a Multivariate Normal distribution.
The diameter has a mean of 30 cm and a standard deviation of 8 cm, the height has a mean of
22 m and a standard deviation of 4 m, and the correlation between them is $0.6$. The
covariance between diameter and height is

$$
\Sigma_{12}=0.6(8)(4)=19.2,
$$

so the covariance matrix is

$$
\boldsymbol{\Sigma}=
\begin{pmatrix}
8^2 & 19.2 \\
19.2 & 4^2
\end{pmatrix}
=
\begin{pmatrix}
64 & 19.2 \\
19.2 & 16
\end{pmatrix}.
$$

We can construct this matrix in R from the standard deviations and the correlation matrix:

```{r}
# Set the means, standard deviations and correlation matrix of diameter and height.
tree_means = c(diameter = 30, height = 22)
tree_sds = c(8, 4)
tree_correlation = matrix(c(1, 0.6,
                            0.6, 1), nrow = 2)

# diag() creates a matrix with the standard deviations on its diagonal and
# %*% is matrix multiplication.
Sigma = diag(tree_sds) %*% tree_correlation %*% diag(tree_sds)
Sigma
```

Base R does not have functions for this distribution, but we can generate random samples
with the function `mvrnorm()` from the `MASS` package:

```{r}
library(MASS) # Install it first

# Simulate 5000 trees. Each row is a tree and each column a variable.
set.seed(2026)
tree_draws = mvrnorm(n = 5000, mu = tree_means, Sigma = Sigma)

# The means and covariances of the sample are close to those of the model.
colMeans(tree_draws)
cov(tree_draws)
```

```{r}
#| echo: false
#| fig-cap: "Diameter and height of 500 trees simulated from a Multivariate Normal distribution with a correlation of 0.6."
# Simulate a smaller sample so that individual trees remain visible in the plot.
set.seed(2026)
tree_plot_draws = MASS::mvrnorm(
  n = 500,
  mu = c(30, 22),
  Sigma = matrix(c(64, 19.2,
                   19.2, 16), nrow = 2)
)
plot(
  tree_plot_draws[, 1],
  tree_plot_draws[, 2],
  xlab = "Tree diameter (cm)",
  ylab = "Tree height (m)"
)
```

:::: {.callout-important title="Practice: a covariance matrix for two traits"}

Suppose the wing length of a bird has a standard deviation of 2 mm, its body mass has a
standard deviation of 3 g and the correlation between the two traits is $0.8$. Write down
the covariance matrix. What would the covariance matrix be if the two traits were
independent?

::: {.callout-tip title="Solution" collapse="true"}

The variances are $2^2=4$ and $3^2=9$ and the covariance is $0.8(2)(3)=4.8$, so

$$
\boldsymbol{\Sigma}=
\begin{pmatrix}
4 & 4.8 \\
4.8 & 9
\end{pmatrix}.
$$

If the traits were independent the correlation would be zero, so the covariance would also
be zero and only the variances on the diagonal would remain.

:::
::::

## Dirichlet distribution

$$
\begin{aligned}
\text{R}: &\quad \texttt{ddirichlet, rdirichlet} \\
\text{Type}: &\quad \text{Continuous, multivariate} \\
\text{Range}: &\quad x_i>0,\quad \sum_{i=1}^K x_i=1 \\
\text{Density}: &\quad
\frac{\Gamma(\alpha_0)}{\prod_{i=1}^K\Gamma(\alpha_i)}
\prod_{i=1}^K x_i^{\alpha_i-1},
\quad \alpha_0=\sum_{i=1}^K\alpha_i \\
\text{Parameters}: &\quad \alpha_i>0\;(\texttt{alpha}) = \text{shape parameters} \\
\text{Mean}: &\quad \mu_i=\alpha_i/\alpha_0 \\
\text{Variance}: &\quad
\operatorname{Var}(X_i)=\frac{\mu_i(1-\mu_i)}{\alpha_0+1} \\
\text{Covariance}: &\quad
\operatorname{Cov}(X_i,X_j)=\frac{-\mu_i\mu_j}{\alpha_0+1},\quad i\ne j \\
\text{Inverse}: &\quad
\alpha_0=\kappa;\quad \alpha_i=\kappa\mu_i
\end{aligned}
$$

The Dirichlet distribution generalises the Beta distribution to $K$ continuous random
variables that are each between $0$ and $1$ and that always add up to one. It is therefore
a model for how a total is divided among several components, such as the fractions of sand,
silt and clay in a soil. It has one shape parameter $\alpha_i>0$ per component and for
$K=2$ it reduces to the Beta distribution.

Its probability density function is

$$
f_{\mathbf{X}}(\mathbf{x})=
\frac{\Gamma(\alpha_0)}{\prod_{i=1}^K\Gamma(\alpha_i)}
\prod_{i=1}^K x_i^{\alpha_i-1},
\qquad \alpha_0=\sum_{i=1}^K\alpha_i,
$$

where $\Gamma$ is the gamma function. The sum of the shape parameters, $\alpha_0$, plays the
role of a precision (the larger it is, the less the fractions vary). The mean and variance
of each fraction are:

$$
\mu_i=E[X_i]=\frac{\alpha_i}{\alpha_0}
$$

and

$$
\operatorname{Var}(X_i)=\frac{\mu_i(1-\mu_i)}{\alpha_0+1}.
$$

As in the Multinomial distribution, the fractions are not independent. Because they add up
to one, a larger fraction for one component leaves less for the others, which results in a
negative covariance:

$$
\operatorname{Cov}(X_i,X_j)=\frac{-\mu_i\mu_j}{\alpha_0+1}, \qquad i \neq j.
$$

As with the Beta distribution, it is convenient to parameterise the Dirichlet distribution
from the mean fractions and a precision $\kappa=\alpha_0$, so that
$\alpha_i=\kappa\mu_i$.

Need to use the `gtools` package (only density and random sampler are available, but
that is enough for this course!).

### Example: biomass partitioning

Suppose a plant allocates on average 20% of its biomass to leaves, 30% to stems and 50% to
roots, and that these fractions vary among plants according to a Dirichlet distribution
with precision $\kappa=12$. The shape parameters are

$$
\alpha_1=12(0.2)=2.4, \qquad \alpha_2=12(0.3)=3.6, \qquad \alpha_3=12(0.5)=6.
$$

The variance of the fraction allocated to roots is

$$
\operatorname{Var}(X_3)=\frac{0.5(1-0.5)}{12+1}\approx 0.0192,
$$

which corresponds to a standard deviation of approximately $0.139$. We can check these
values by simulating many plants:

```{r}
library(gtools) # Install it first!

# Convert the mean fractions and the precision into the shape parameters.
fraction_means = c(leaves = 0.2, stems = 0.3, roots = 0.5)
fraction_precision = 12
alpha = fraction_precision * fraction_means

# Simulate 5000 plants. Each row is a plant and each column a fraction.
set.seed(2026)
biomass_draws = rdirichlet(n = 5000, alpha = alpha)

# The fractions of every plant add up to one.
range(rowSums(biomass_draws))

# Average fraction per component and variance of the fraction allocated to roots.
colMeans(biomass_draws)
var(biomass_draws[, 3])
```

```{r}
#| echo: false
#| fig-cap: "Fractions of biomass allocated to leaves and roots in 500 plants simulated from a Dirichlet distribution."
# Simulate a smaller sample so that individual plants remain visible in the plot.
set.seed(2026)
biomass_plot_draws = gtools::rdirichlet(n = 500, alpha = 12 * c(0.2, 0.3, 0.5))
plot(
  biomass_plot_draws[, 1],
  biomass_plot_draws[, 3],
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Fraction of biomass in leaves",
  ylab = "Fraction of biomass in roots"
)
```

:::: {.callout-important title="Practice: soil texture"}

Suppose the mean fractions of sand, silt and clay in the soils of a region are $0.6$, $0.3$
and $0.1$, and that they vary among sites according to a Dirichlet distribution with
precision $\kappa=50$. Calculate the shape parameters and the standard deviation of the
clay fraction.

::: {.callout-tip title="Solution" collapse="true"}

The shape parameters are $50(0.6)=30$, $50(0.3)=15$ and $50(0.1)=5$. The variance of the
clay fraction is $0.1(0.9)/51\approx 0.00176$, so its standard deviation is approximately
$0.042$.

:::
::::

# How to make zero-inflated distributions {#sec-zero_inflated}

A zero-inflated distribution is a mixture of two distributions:

(i) A Bernoulli distribution associated to a binary process.

(ii) Any other distribution (discrete of continuous) that may or may not contain zeros.

The idea is that each distribution in the mixture represents a different process. For example,
in the seed predation data we work with in the course, there is a probability that the
predators does not find the little dishes with seeds ($p_{0}$). That is represented by
the Bernoulli distribution mentioned above. If a predator finds a dish with seeds, it can eat
a variable number of seeds (from zero to as many seeds as there are available) depending on
how hungry it is and its foraging behaviour.

A zero-inflated version of a mass or density function $p(x)$ is going to look like this:

$$
\begin{aligned}
p_{\text{inf}}(x) &=
\begin{cases}
    p_{\text{0}} + &(1 - p_{\text{0}})p(0), & x = 0,\\
    &(1- p_{\text{0}})p(x), & x\gt 0.
\end{cases}
\end{aligned}
$$

Notice how there are two terms for $x = 0$, one comes from the Bernoulli distribution ($p_0$)
and the other one comes from the original (non-inflated) distribution ($p(0)$). Also notice
how the original distribution has to be scaled by a factor $(1- p_{\text{0}})$. This is to
ensure, as always, that probabilities add up to 1.

There might be packages in R that implement some zero-inflated distributions but we can
easily implement our own. Below is an example of how to define the `dpqr` functions for a
zero-inflated negative binomial.

```{r}
# The mass/density function you will use for the likelihood, including log
dzinbinom = function(x, size, ..., pzero, log = FALSE) {
  baseline = dnbinom(x, size = size, ..., log = FALSE)
  result = (1 - pzero) * baseline
  result = ifelse(x == 0, pzero + result, result)
  if (log) log(result) else result
}

# The random number generator for sampling
rzinbinom = function(n, size, ..., pzero) {
  # Original distribution
  result = rnbinom(n, size = size, ...)
  # Bernoulli distribution
  result[runif(length(result)) < pzero] = 0
  result
}

# Cumulative probability for tail probabilities and p-values
pzinbinom = function(q, size, ..., pzero,
                      lower.tail = TRUE, log.p = FALSE) {
  baseline = pnbinom(q, size = size, ...,
                      lower.tail = TRUE, log.p = FALSE)
  result = ifelse(q < 0, 0, pzero + (1 - pzero) * baseline)
  if (!lower.tail) result = 1 - result
  if (log.p) log(result) else result
}

# Quantile function (assuming 0 < pzero < 1)
qzinbinom = function(p, size, ..., pzero, lower.tail = TRUE) {
  # Switch between lower or upper tail
  if (!lower.tail) p = 1 - p

  # Quantile in the original distribution
  baseline_p = (p - pzero) / (1 - pzero)
  baseline_q = qnbinom(baseline_p, size = size, ...)

  # Adjust for zero inflation
  cdf_at_zero = pzero + (1 - pzero)*pnbinom(0, size = size, ..., lower.tail = TRUE)
  ifelse(p <= cdf_at_zero, 0, baseline_q)
}

# Example
mean = 8
size = 2
prob = size / (size + mean)
pzero = 0.25

x = 0:qzinbinom(0.999, size = size, prob = prob, pzero = pzero)
plot(x, dzinbinom(x, size = size, prob = prob, pzero = pzero),
     type = "h", xlab = "Count", ylab = "Probability")

pzinbinom(0:10, size = size, prob = prob, pzero = pzero)
qzinbinom(c(0.5, 0.9), size = size, prob = prob, pzero = pzero)
rzinbinom(10, size = size, prob = prob, pzero = pzero)
```
